### 2010-12-02 opencl/papers/foropencl.tex
Initial version

\begin{document}

\title{ForOpenCL: Examing Fortran as a base language for accelerated platforms:
I forgot what we decided upon}

%
% \author{}
%

\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

1. Goal is to write once, transform many. Otherwise potentially reprogram for every architecture.
2. Goal is to code for readability and maintainability, not performance
3. This goal requires expression in a high-level language.
4. But language must be simple enough for compiler to analyze.
5. Thus ideal if language maps well to accelerator architectures.
6. Data parallel constructs in Fortran 90 are chosen.
7. Up to 65 times speed up measured on automatically transformed code.

At Los Alamos we have a wide variety of computer platforms from which
to choose.  The common platform is made up of cluster of compute nodes
with standard multi-core processors.  Some nodes have accelerators
that range from the IBM Cell processor (Roadrunner) to a variety of
GPUs from NVIDIA and AMD.  Some nodes have have hardware with vector
instructions and others do not.  An important question is what programming
model and what languages are best suited for these platforms.

A data parallel programming model is suited for accelerators....

One possibility is to adopt a new language like CUDA or Chapel...
Another possibility is to use a mature language with OpenMP directives
or write OpenCL kernels.

The problem is that many of the language choices expose too much detail
about the machine architecture.  This is particularly true of CUDA and OpenCL.

At this point in time it is not really possible to write once and
run efficiently on the wide variety of computer platforms we have
available.

This papers addresses this issue by examining features in Fortran that
allow programmers to express algorithms at a very high level but yet
can be transformed by a compiler to run efficiently on a wide variety
of platforms.  In particular we consider nodes of GPUs.

We demonstrate that the array syntax of Fortran maps surprisingly well
onto GPUs when transformed to OpenCL kernels.  These Fortran language
features include pure and elemental functions and array constructs
like where and shift.  In addition we add a few functions that enable
a program to be deployed on machines with a hierarchy of processing elements,
such as nodes employing GPU acceleration,
\emph{without requiring explicit declaration of parallelism within the program.}
In addition the program uses entirely standard Fortran so it can be
on a single core without concurrency.

We study automatic transformations of Fortran for two applications examples.
The first is a simple shallow water model in two dimensions using
finite volume methods with .... for time update.  The second is based
on the primary compute kernels from PetaVision, a high-performance, 
neural science framework that is used in models of visual cortex.

\section{The programming model and Fortran-language subset}

The static analysis and source-to-source transformations used are not magic
and require the programmer to use a subset of Fortran that employs
a data-parallel programming model in this section we describe this
language subset.

Similar to ZPL.

\subsubsection{elemental functions}

\subsubsection{pure procedures}

Mention that don't require pure procedures as region functions are not pure.
But programmers should otherwise think in terms of writing pure procedures.

\subsection{Limitations}

Only Fortran procedures are transformed into OpenCL kernels.  The programmer
must currently explicitly call these kernels from Fortran using the ForOpenCL
library described below.  It is also possible using ROSE to modify the calling
site so that the entire program can be transformed but this functionality is
outside the scope of this paper.  Here we specifically examine transforming
Fortran procedures to OpenCL kernels.

Only elemental functions may be called from kernel functions.  These include
fortran functions that have an OpenCL analog and user-defined elemental functions.

Array sizes must be multiples of the local kernel size,
get_local_size(0)*get_local_size(1).  This may be relaxed in the future.

\subsection{ForOpenCL}

ForOpenCL is library of Fortran modules.  It contains Fortran 2003 interface
descriptions that allow language interoperability with the C OpenCL runtime.

\section{Transformation examples}

This sections show short Fortran code examples and OpenCL equivalent.
The notation uses uppercase for arrays and lowercase for scalar quantities.
Interior array subsections are denoted by an i preceeding the array, e.g. iA
is the interior subset of array A.

\subsubsection{array syntax}

PetaVision updates the action potential with the statement,

   V = V - iA*(V - v_rest)                  !! Fortran

   V[l] = V[l] - iA[lex]*(V[l] - v_rest);   // OpenCL equivalent

\subsubsection{where construct}

A neuron in PetaVision fires (activity is set to 1)
whenever the action potential is greater than some threshold.
This is easily expressed with a where construct.

   where (V > Vth)                        !! Fortran
      iA = 1
   elsewhere
      iA = 0
   end where

   iA[lex] = (V[l] > Vth[l]) ? 1 : 0;     // OpenCL equivalent

\subsection{New functions}

%\subsubsection{transfer_halo}
\subsubsection{region}

One of the state variables in the shallow water code is H, effectively the
height of the water.  This variable has a halo (ghost-cell region) surrounding
the interior of the grid to handle boundary conditions.  When using
MPI, the halo regions contains values from adjacent processors that must be
updated with new values at each time step.  This is accomplished with the
transfer_halo function.  The shallow water code assumes a five point stencil
so the state variables are extended by 2 in each dimension (e.g., by one to the
left, right, up, and down).

halo = [1,1,1,1]
iH => region(H, halo, transfer=false)

\section{Static Analysis}

1. Dependence analysis to insert thread group barriers

   barrier(CLK_LOCAL_MEM_FENCE)

2. subsection variables associated with array variables

\subsection{Analysis not required}

1. loop fusion
2. removal of array temporaries (shifts)

\section{Performance}

\end{document}

### 2010-12-03 opencl/papers/foropencl.tex
updating

@@
-\begin{document}
+\documentclass{article}
 
@@
 I forgot what we decided upon}
+\author{Awesome Fortran Programmers}
 
-%
-% \author{}
-%
+\begin{document}
 
@@
 \begin{abstract}
-
+I am abstract.
 \end{abstract}
@@
 \section{Introduction}
+
+This paper presents a compiler-level approach for targeting a single program to multiple fundamentally different low level execution and programming models while allowing the application programmer to adopt a single high level programming model.
 
@@
 
+\subsection{Why Fortran?}
+
+A likely question that one may pose is ``\emph{Why Fortran and not a more modern language like X?}''
+The recent rise in interest in concurrency and parallelism at the language level driven by multicore CPUs and
+manycore accelerators has driven a number of new language developments, both as novel languages and extensions
+on existing ones.
+
+For scientific users, these new languages and language extensions present a challenge: how do developers effectively use them while avoiding rewriting their code and potentially growing dependent on a transient technology that will vanish tomorrow?
+
+Fortran is unique in that it has contained language features that are well suited to modern architectures for a number of years.
+This should be unsurprising --- Fortran was a primary language used to target systems such as the vector supercomputers and massively parallel systems of the 1970s and 1980s.
+These are the systems in which architectural features were developed that have led to single chip high performance architectures of interest today.
+Given that these new systems have features very similar to their predecessors, it is clear that the language features within Fortran for them are still relevant.
+
 \section{The programming model and Fortran-language subset}
 
-The static analysis and source-to-source transformations used are not magic
-and require the programmer to use a subset of Fortran that employs
-a data-parallel programming model in this section we describe this
-language subset.
+The static analysis and source-to-source transformations used are very basic
+and simply require the programmer to use a subset of Fortran that employs
+a data-parallel programming model.  In particular, it encourages use of language features that
+were introduced and standardized in the Fortran 90 language specification.  In this section we describe the
+set of Fortran 90 features that our analysis and transformation method are based on.
 
 Similar to ZPL.
+
+%% cites: ZPL, CM5 stencil compilers
+
+% put something here on:
+%
+% shifts
+% array notation
+%
 
@@
 Array sizes must be multiples of the local kernel size,
-get_local_size(0)*get_local_size(1).  This may be relaxed in the future.
+{\tt get\_local\_size(0)*get\_local\_size(1)}.  This may be relaxed in the future.
 
@@
 
+\begin{verbatim}
    V = V - iA*(V - v_rest)                  !! Fortran
@@
    V[l] = V[l] - iA[lex]*(V[l] - v_rest);   // OpenCL equivalent
+\end{verbatim}
 
@@
 
+\begin{verbatim}
    where (V > Vth)                        !! Fortran
@@
    iA[lex] = (V[l] > Vth[l]) ? 1 : 0;     // OpenCL equivalent
+\end{verbatim}
 
@@
 updated with new values at each time step.  This is accomplished with the
-transfer_halo function.  The shallow water code assumes a five point stencil
+{\tt transfer\_halo} function.  The shallow water code assumes a five point stencil
 so the state variables are extended by 2 in each dimension (e.g., by one to the
@@
 
+\begin{verbatim}
 halo = [1,1,1,1]
 iH => region(H, halo, transfer=false)
+\end{verbatim}
 
@@
 
+\begin{verbatim}
    barrier(CLK_LOCAL_MEM_FENCE)
+\end{verbatim}
 

### 2010-12-03 opencl/papers/foropencl.tex
updates for conference format
more

@@
-\documentclass{article}
+\documentclass[10pt, conference, compsocconf]{IEEEtran}
 
-\title{ForOpenCL: Examing Fortran as a base language for accelerated platforms:
-I forgot what we decided upon}
+\usepackage{cite}
+\usepackage{graphicx}
+\usepackage[cmex10]{amsmath}
+\usepackage{array}
+\usepackage{url}
+
+\title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
 \author{Awesome Fortran Programmers}
@@
 
-This paper presents a compiler-level approach for targeting a single program to multiple fundamentally different low level execution and programming models while allowing the application programmer to adopt a single high level programming model.
+This paper presents a compiler-level approach for targeting a single
+program to multiple fundamentally different low level execution and
+programming models while allowing the application programmer to adopt
+a single high level programming model.
 
@@
 like where and shift.  In addition we add a few functions that enable
-a program to be deployed on machines with a hierarchy of processing elements,
-such as nodes employing GPU acceleration,
-\emph{without requiring explicit declaration of parallelism within the program.}
-In addition the program uses entirely standard Fortran so it can be
-on a single core without concurrency.
+a program to be deployed on machines with a hierarchy of processing
+elements, such as nodes employing GPU acceleration, \emph{without
+  requiring explicit declaration of parallelism within the program.}
+In addition the program uses entirely standard Fortran so it can be on
+a single core without concurrency.
 
@@
 
-A likely question that one may pose is ``\emph{Why Fortran and not a more modern language like X?}''
-The recent rise in interest in concurrency and parallelism at the language level driven by multicore CPUs and
-manycore accelerators has driven a number of new language developments, both as novel languages and extensions
-on existing ones.
+A likely question that one may pose is ``\emph{Why Fortran and not a
+  more modern language like X?}''  The recent rise in interest in
+concurrency and parallelism at the language level driven by multicore
+CPUs and manycore accelerators has driven a number of new language
+developments, both as novel languages and extensions on existing ones.
 
-For scientific users, these new languages and language extensions present a challenge: how do developers effectively use them while avoiding rewriting their code and potentially growing dependent on a transient technology that will vanish tomorrow?
+For scientific users, these new languages and language extensions
+present a challenge: how do developers effectively use them while
+avoiding rewriting their code and potentially growing dependent on a
+transient technology that will vanish tomorrow?
 
-Fortran is unique in that it has contained language features that are well suited to modern architectures for a number of years.
-This should be unsurprising --- Fortran was a primary language used to target systems such as the vector supercomputers and massively parallel systems of the 1970s and 1980s.
-These are the systems in which architectural features were developed that have led to single chip high performance architectures of interest today.
-Given that these new systems have features very similar to their predecessors, it is clear that the language features within Fortran for them are still relevant.
+Fortran is unique in that it has contained language features that are
+well suited to modern architectures for a number of years.  This
+should be unsurprising --- Fortran was a primary language used to
+target systems such as the vector supercomputers and massively
+parallel systems of the 1970s and 1980s.  These are the systems in
+which architectural features were developed that have led to single
+chip high performance architectures of interest today.  Given that
+these new systems have features very similar to their predecessors, it
+is clear that the language features within Fortran for them are still
+relevant.
 
@@
 
-The static analysis and source-to-source transformations used are very basic
-and simply require the programmer to use a subset of Fortran that employs
-a data-parallel programming model.  In particular, it encourages use of language features that
-were introduced and standardized in the Fortran 90 language specification.  In this section we describe the
-set of Fortran 90 features that our analysis and transformation method are based on.
+The static analysis and source-to-source transformations used are very
+basic and simply require the programmer to use a subset of Fortran
+that employs a data-parallel programming model.  In particular, it
+encourages use of language features that were introduced and
+standardized in the Fortran 90 language specification.  In this
+section we describe the set of Fortran 90 features that our analysis
+and transformation method are based on.
 
-Similar to ZPL.
+Fortran 90 introduced a rich array syntax that allows programmers to
+write code that is in terms of whole arrays or subarrays, with data
+parallel operators to compute on the arrays.  This has the benefit of
+avoiding explicit looping in the code and maintaining a high-level
+style that is closer to the original mathematical specification of the
+problem being solved.  More importantly from a compilation
+perspective, this defers decisions about how to implement these
+whole-array operations to a compilation tool.  When faced with a novel
+architecture such as modern GPUs that are ideally suited to data parallel
+programming models, the fit between Fortran arrays and these systems is
+quite clean.
+
+This paper uses four simple features of Fortran that form a language subset
+that does not require any language extensions and can be easily transformed
+to a lower level implementation in either CUDA or OpenCL.
+
+\begin{enumerate}
+\item Elemental functions
+\item Pure procedures
+\item Shift operators
+\item Array notation
+\end{enumerate}
+
+\subsection{Relationship to other languages}
+
+Similar to ZPL, CM5 stencil compilers, CMFortran stuff.
 

### 2010-12-03 opencl/papers/foropencl.tex
edits

@@
 programming models while allowing the application programmer to adopt
-a single high level programming model.
+a single high level programming model.  We suggest that features of
+Fortran 90 for data parallel programming are well suited to automatic
+transformation to generate code specifically tuned for different
+low-level programming models such as OpenCL and CUDA.
 
@@
 
-A data parallel programming model is suited for accelerators....
+A data parallel programming model is well suited to emerging architectures
+that are based on either vector processing or massively parallel collections
+of simple cores.  A popular trend for programming systems such as GPU
+accelerators is the CUDA language from NVidia, or new languages like Chapel
+from Cray.  A less intrusive approach that is well suited to legacy 
+applications and languages are language extensions or libraries that allow
+programmers to avoid adopting entirely new languages.  The recent OpenCL
+specification is intended to support this programming model, as are
+directive-based methods such as OpenMP or the ``Accelerate'' product from
+the Portland Group.
 
-One possibility is to adopt a new language like CUDA or Chapel...
-Another possibility is to use a mature language with OpenMP directives
-or write OpenCL kernels.
-
-The problem is that many of the language choices expose too much detail
-about the machine architecture.  This is particularly true of CUDA and OpenCL.
+The problem with many of these choices is that they expose too much detail
+about the machine architecture to the programmer.  
+This is particularly true of CUDA and OpenCL.  In CUDA, programmers must
+adapt their codes to fit the threading model used by NVidia GPUs, while
+OpenCL requires programmers to provide specially tuned versions of
+their code for different classes of machine.  In both cases, the programmer
+is also responsible for explicitly managing memory, including staging of
+data back and forth from the host CPU and the accelerator device memory.
+While these models have been attractive as a method for early adopters
+to utilize these new architectures, they are less attractive to programmers
+who do not have the time or resources to manually port their code to every
+new architecture and programming model that emerges.
 
@@
 run efficiently on the wide variety of computer platforms we have
-available.
+available.  For some classes of applications, we believe that this goal
+is possible using language constructs already present in a popular
+mainstream scientific programming language -- Fortran.
+
+\subsection{Approach}
 
@@
 
+\section{Junk}
+
+\cite{chamberlain04zpl}
+
+\bibliographystyle{IEEEtran}
+\bibliography{IEEEabrv,foropencl}
+
 \end{document}

### 2010-12-03 opencl/papers/foropencl.tex
PGI
more

@@
 
+%%
+%%
+%%
+
 \section{The programming model and Fortran-language subset}
@@
 
-\begin{enumerate}
-\item Elemental functions
-\item Pure procedures
-\item Shift operators
-\item Array notation
-\end{enumerate}
+\subsubsection*{Elemental functions}
+
+An elemental function consumes and produces scalar values, but can be
+applied to variables of array type such that the function is
+applied to each and every element of the array.  This allows programmers
+to avoid explicit looping and instead simply state that they intend a
+function to be applied to every element of an array in parallel, deferring
+the choice of implementation technique to the compiler.  Elemental
+functions are intended to be used for data parallel programming, and
+as a result must be side effect free (or, \emph{pure}).
+
+\subsubsection*{Pure procedures}
+
+Pure procedures, like elemental functions, must be free of side effects.
+The absence of side effects removes ordering constraints from the compiler
+allowing it to invoke pure functions out of order.  Procedures and functions
+of this sort are also common in pure functional languages like Haskell,
+and are exploited by compilers in order to emit parallel code automatically
+due to their suitability for compiler-level analysis.
+
+\subsubsection*{Shift operators}
+
+Many array-based algorithms require the same operation to be performed
+on each element of the array using the value of that element and some small
+set of neighboring cells.  Often programmers implement these operations that
+are local to each element within the inner-loop of a set of nested FOR-
+or DO-loops using offsets relative to the current array index.  An alternative
+to this local-view of the algorithm is to take a global view and write the
+algorithm in terms of the whole array.  For example, consider a 1D array in
+which we wish to subtract the $(i-1)$th element from the $i$th for all
+elements.  One way to look at this is that we are subtracting the entire array
+shifted by one element from itself.
+
+Fortran provides a set of shifting operators that allow programmers to
+define operations based on shifted arrays.  These intrinsic operators take
+an array, a dimension, and the amount by which it should be shifted (using
+the sign to indicate direction).  By defining operations on entire arrays
+based on a global view of them shifted relative to each other, programmers can
+avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
+analysis of the extent of the set of shifted arrays in a given expression
+allows analysis tools to determine the amount of temporary or buffer storage
+necessary to hold intermediate values during whole array operations.  With
+explicit loops, programmers must maintain this temporary storage manually.
+
+\subsubsection*{Array notation}
+
+The final piece that we exploit is the array notation present in modern
+versions of Fortran.  ....
 
@@
 
-\cite{chamberlain04zpl}
+\cite{chamberlain04zpl,pgi10accelerator}
 

### 2010-12-04 opencl/papers/foropencl.tex
more text, more refs

@@
 \title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
-\author{Awesome Fortran Programmers}
+
+%%
+%% reorder appropriately later
+%%
+\author{\IEEEauthorblockN{Matthew J. Sottile}
+\IEEEauthorblockA{Galois, Inc.\\
+421 SW 6th Ave. Suite 300 \\
+Portland, OR 97204\\
+Email: matt@galois.com}
+\and
+\IEEEauthorblockN{Craig E Rasmussen}
+\IEEEauthorblockA{Los Alamos National Laboratory\\
+CCS-7, MS B287\\
+Los Alamos, NM\\
+Email: rasmussn@lanl.gov}
+}
 
@@
 \begin{abstract}
-I am abstract.
+Emerging architectures for high performance computing often are well
+suited to a data parallel programming model.  This paper presents a
+simple programming methodology based on existing languages and compiler
+tools that allows programmers to take advantage of these systems.
+We will work with the array features of Fortran 90 to show how this
+infrequently exploited, standardized language feature is easily
+transformed to lower level accelerator code.  Our transformations are
+based on a mapping from Fortran 90 to C++ code with OpenCL extensions.
 \end{abstract}
@@
 programming models while allowing the application programmer to adopt
-a single high level programming model.  We suggest that features of
+a single high level programming model.  We show that features of
 Fortran 90 for data parallel programming are well suited to automatic
 transformation to generate code specifically tuned for different
-low-level programming models such as OpenCL and CUDA.
+low-level programming models such as OpenCL and CUDA.  For algorithms
+that can be easily expressed in terms of whole array, data parallel operations,
+writing code in Fortran and transforming it automatically to specific low
+level implementations removes the burden from the programmer of working
+with tedious, error prone, low-level tools.
 
@@
 
-A data parallel programming model is well suited to emerging architectures
-that are based on either vector processing or massively parallel collections
-of simple cores.  A popular trend for programming systems such as GPU
-accelerators is the CUDA language from NVidia, or new languages like Chapel
-from Cray.  A less intrusive approach that is well suited to legacy 
-applications and languages are language extensions or libraries that allow
-programmers to avoid adopting entirely new languages.  The recent OpenCL
-specification is intended to support this programming model, as are
-directive-based methods such as OpenMP or the ``Accelerate'' product from
-the Portland Group.
+A data parallel programming model is well suited to emerging
+architectures that are based on either vector processing or massively
+parallel collections of simple cores.  A popular trend for programming
+systems such as GPU accelerators is the CUDA language from NVidia, or
+new languages like Chapel from Cray.  A less intrusive approach that
+is well suited to legacy applications and languages are language
+extensions or libraries that allow programmers to avoid adopting
+entirely new languages.  The recent OpenCL specification is intended
+to support this programming model, as are directive-based methods such
+as OpenMP or the Accelerator programming model from the Portland
+Group~\cite{pgi10accelerator}.
 
@@
 
+Techniques based on whole-array programming being mapped to high performance,
+parallel implementations are not new.
+
 Similar to ZPL, CM5 stencil compilers, CMFortran stuff.
@@
 
-\cite{chamberlain04zpl,pgi10accelerator}
+\cite{chamberlain04zpl, roth97stencils}
 

### 2010-12-10 opencl/papers/foropencl.tex
Added a lot of text from proposal and the book.  Still need to weed it down.

@@
+%
+% I'm adding an outline here so I can see how everything fits together
+
+%
+% \title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
+%
+
+%
+% \abstract
+%
+
+%
+% \section{Introduction}
+%    \subsection{Approach}
+%    \subsection{Why Fortran?}
+%    \subsection{Other Languages}
+%
+
+%
+% \section{Programming model}
+%    \subsection{Fortran Syntax}   
+%       \subsubsection*{Array notation}
+%       \subsubsection*{Elemental functions}
+%       \subsubsection*{Pure procedures}
+%       \subsubsection*{Shift operators}
+%       \subsubsection*{Regions}
+%    \subsection{Parallelism}
+%    \subsection{New Functions}
+%    \subsection{Limitations}
+%
+
+%
+% \section{Source-To-Source Transformations}
+%    \subsection{ForOpenCL}
+%       \subsubsection{array syntax}
+%       \subsubsection{where construct}
+%    \subsection{New functions}
+%       \subsubsection{region}
+%    \subsection{Static Analysis}
+%       \subsubsection{Analysis not required}
+%    \subsection{Simplifying Assumptions}
+%
+
+%
+% \section{Shallow Water Model}
+%    \subsection{Equations}
+%
+
+%
+% \section{Performance}
+%
+
+%
+% \section{Conclusions}
+%
+
 \documentclass[10pt, conference, compsocconf]{IEEEtran}
@@
 
-At Los Alamos we have a wide variety of computer platforms from which
-to choose.  The common platform is made up of cluster of compute nodes
-with standard multi-core processors.  Some nodes have accelerators
-that range from the IBM Cell processor (Roadrunner) to a variety of
-GPUs from NVIDIA and AMD.  Some nodes have have hardware with vector
-instructions and others do not.  An important question is what programming
-model and what languages are best suited for these platforms.
+At Los Alamos National Laboratory (LANL) we have a wide variety of computer
+platforms from which to choose.  The common platform is made up of clusters of
+compute nodes with standard multi-core processors.  Some nodes also have
+accelerators that range from the IBM Cell processor (Roadrunner) to a variety
+of GPUs from NVIDIA and AMD.  Some nodes have hardware with vector
+instructions and others do not. The peak
+performance of the accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
+processor to XXX for NVIDIA Fermi. %% Fermi has another name
+
+%% NEW - CER
+However, the performance that these accelerator architectures offer comes at a
+cost as processor architectures are trending toward multiple cores with
+instances of integrated accelerator units and less of a reliance on memory
+caches.  These changes place a heavy burden on application programmers as they
+are forced to adapt to the new systems.  An especially challenging problem
+faced by application programmers is not only how to program to these new
+architectures (considering the massive scale of concurrency available), but
+also how to design programs that are portable across the changing landscape of
+computer architectures.  The fundamental question we address is what
+programming model and what languages are best suited for these platforms.
 
@@
 
-This papers addresses this issue by examining features in Fortran that
+This paper addresses this issue by examining features in Fortran that
 allow programmers to express algorithms at a very high level but yet
@@
 
-We study automatic transformations of Fortran for two applications examples.
-The first is a simple shallow water model in two dimensions using
-finite volume methods with .... for time update.  The second is based
-on the primary compute kernels from PetaVision, a high-performance, 
-neural science framework that is used in models of visual cortex.
+%% NEW - CER
+The sheer complexity of programming for clusters of many or multi-core
+processors with tens of millions threads of execution make the simplicity of
+the data parallel model attractive.  Furthermore, the increasing complexity of
+todays applications (especially when convolved with the increasing complexity
+of the hardware) and the need for portability across hardware architectures
+make a higher-level and simpler programming model like data parallel
+attractive.
+
+%% NEW - CER
+The goal of this work is to exploit source-to-source transformations that
+allow programmers to develop and maintain programs at a high-level of
+abstraction, without coding to a specific hardware architecture.
+Furthermore these transformations allow multiple hardware architectures
+to be targeted without changing the high-level source.  It also removes the
+necessity for application programmers to understand details of the accelerator
+architecture or to know OpenCL.
+
+We use the ROSE compiler infrastructure \cite{} to develop these
+transformations.  ROSE uses the Open Fortran Parser \cite{} to parse Fortran
+2008 syntax and can output (unparse) to OpenCL.  Since ROSE's intermediate
+representation (IR) was constructed to represent multiple languages, it is
+relatively straightforward to transform high-level Fortran IR nodes to OpenCL
+nodes.  Furthermore, the Fortran array syntax maps directly to one, two, and
+three-dimensional thread groups in OpenCL.
+
+%%We study automatic transformations of Fortran for two applications examples.
+The first is a simple shallow water model in two dimensions using finite volume methods with ....
+for time update.
+The second is based on the primary compute kernels from PetaVision, a high-performance, neural science framework that is used in models of visual cortex.
+
+%% To simplify things, lets save PetaVision for the local stuff as it requires convolutions
+
+We study automatic transformations of Fortran for an application example that
+is typical of many applications that are based on finite-difference or
+finite-volume methods in computational fluid dynamics (CFD).  The example is a
+simple shallow water model in two dimensions using finite volume methods with
+.... for time update.
+
+%% NEW - CER
+PAGOSA is a production code at LANL completely written in data-parallel
+Fortran.  In a prior work we investigated automatically transforming this code
+to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
+and IBM Cell processors).  We demonstrated that a source-to-source compiler
+can automatically vectorize and parallelize (both within the Cell processor
+and across nodes) a small section of this code.  Preliminary results showed a
+12 times performance gain on the Cell processor over a traditional single core
+processor.  A factor of 9 is gained from vectorizing and parallelizing the
+code to run on the Cell and the rest is gained from automatically overlapping
+inter-node communication with computation on the Cell.
 
@@
 
-A likely question that one may pose is ``\emph{Why Fortran and not a
-  more modern language like X?}''  The recent rise in interest in
-concurrency and parallelism at the language level driven by multicore
-CPUs and manycore accelerators has driven a number of new language
-developments, both as novel languages and extensions on existing ones.
+%% NEW - CER
+Fortran \footnote{Originally named FORTRAN for FORmula TRANslation language.}
+is the oldest high-level programming language, and was developed to facilitate
+the translation of math formulae into machine code. Fortran was the first
+major language to use a compiler to translate from a high-level program
+representation to assembly language. Due to its age, it carries certain arcane
+baggage that later programming languages have evolved away from. As a result,
+Fortran has fallen into disfavor in certain programming circles.
+
+%% NEW - CER
+However, with the introduction of Fortran 90 (and later revisions to the
+standard), Fortran is a truly modern programming language. Much of the arcane
+baggage associated with it is actually no longer part of the language used by
+modern programmers, and exists in a deprecated form simply to support legacy
+codes. It is now modular and has many object-oriented features. Furthermore,
+Fortran should be of interest to those studying parallel programming because
+of its functional and data parallel constructs and because of the coarray
+notation introduced after Fortran 2003. Unlike languages like C and C++, which
+are considered to be more modern, Fortran has become a truly parallel language
+with features added to recent language standards. With coarrays, Fortran
+provides a mechanism to run replicas (called images) of a program on many
+different processors (or cores) at the same time. It also provides language
+constructs that allow the program to both read from and write to data residing
+on {\it remote} processors and to synchronize the different program images so
+that access to the data by the multiple images can be done atomically.
+
+A likely question that one may pose is ``\emph{Why Fortran and not a more
+  modern language like X?}''  The recent rise in interest in concurrency and
+parallelism at the language level driven by multicore CPUs and manycore
+accelerators has driven a number of new language developments, both as novel
+languages and extensions on existing ones.
+
+%% NEW - CER
+So perhaps it is time to replace Fortran with yet another computer language.  The
+problems with replacing Fortran with an entirely new language are two fold:
+the economics of replacing the existing application base and the difficulty in
+obtaining programmer acceptance.  It is estimated that replacing a major
+production application at Los Alamos National Laboratory would cost between 50
+and 150 million dollars.  In terms of programmer acceptance, there is always
+the "chicken and egg problem": programmers won't use a new language until they
+can expect good performance across a variety of platforms, and compiler
+vendors can't afford to produce quality compilers until there is a reasonable
+expectation of a market.
 
@@
 transient technology that will vanish tomorrow?
+
+%% NEW - CER
+History has also shown that an investment in rewriting code does not guarantee
+success either, as seen in an effort at LANL to modernize a legacy Fortran
+code with the the newer C++ POOMA framework.  This massive overhaul effort led
+to a code that was both slower and less flexible than the original Fortran
+\cite{}.  Similar experiences have occurred in the past, notably during the
+devlopment of the Sisal language.
 
@@
 
+%% NEW - CER
+We will examine how the existing data-parallel constructs in Fortran can be
+combined with coarrays or MPI to provide effectively a new parallel programming
+language, one that is evolutionary in nature and provides complete compatibility
+with existing applications and libraries.  Data parallelism is a high-level
+abstraction that is, at the same time, both easier to program and gives the
+compiler more leeway (if fully exploited) in retargeting a program to different
+computer architectures.
+
+%% NEW - CER
+An example of this style of programming in Fortran is shown in
+
+\begin{verbatim}
+
+  Bz = Bz + dt * (cshift(Ex,dim=2,shift=+1) - Ex) / dy  &
+          - dt * (cshift(Ey,dim=1,shift=+1) - Ey) / dx
+
+\end{verbatim}
+
+%% NEW - CER
+This example is a solution to Maxwell's equations for the $z$ component of the
+magnetic field using Fortran array syntax.  Note that there are no explicit
+loops in this example.  The operators {\tt +}, {\tt -}, and {\tt *} are
+applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt
+  Ex}, and {\tt Ey}, individually.  This is why data parallelism has been
+called collection-oriented programming by
+Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift} function
+and the array-valued expressions all semantically returns a value, this style
+of programming is also similar to functional programming (or value-oriented
+programming). %%~\cite{simonpeytonjones}).  The heart of the solution to
+Maxwell's equations is the statement shown in Listing~\ref{lst:dpexample} and
+five similar, simple equations.
+
+%% NEW - CER
+The power in this notation is that the compiler is free to distribute the
+computation in these expressions over any hardware processing elements available
+to it, such as the vector units on an Intel or AMD processor or the Synergistic
+Processing Units on an IBM Cell processor.  If the arrays are declared as
+coarrays, this includes spreading the computation over \emph{all of the nodes in
+  a cluster as well.}  In this case, communication occurs within the
+\texttt{cshift} functions, though the compiler is free to overlap communication
+and computation by scheduling the communication early and performing the
+computation on the interior of the arrays while waiting for the communication to
+complete.
+
+%% NEW - CER
+Complete and very concise and elegant programs can be built with statements
+similar to that shown in Listing~\ref{lst:dpexample}, using intrinsic
+functions like the array constructors (\texttt{CSHIFT}, \texttt{MERGE},
+\texttt{TRANSPOSE}, ...), the array location functions (\texttt{MAXLOC} and
+\texttt{MINLOC}), and the array reduction functions (\texttt{ANY},
+\texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM}, \texttt{PRODUCT},
+...)~\cite{guydp}.  It should be emphasized, however, that {\it no} compiler
+currently supplies coarray versions of these intrinsic functions as they are
+not defined by the proposed Fortran 2008 standard (we have submitted a paper
+proposing the integration of coarrays with these intricics functions to the J3
+Fortran committee~\cite{rasmussen-caf-intrinsics}).  With coarray or MPI
+implementations of these intrinsic functions, programs written in this
+data-parallel subset of Fortran are implicitly parallel.  Furthermore, in
+instances it is necessary to implement algorithms that don't lend themselves
+to expressibility in this data-parallel subset, coarray notation can be used
+for complete access to individual (including remote) data elements.  It is the
+combination of data parallel and coarray constructs, with the transformational
+properties of a source-to-source compiler (transforming array notation to
+optimized Fortran 95 and library calls), that make this possibility such a
+promising avenue for research.
+
+%% NEW - CER
+This style of programming meets the requirements we have set for a programming
+model for developing petascale applications.  It allows the programmer to
+program at a very-high level of abstraction while providing the compiler with
+maximum flexibility in targeting the application for a particular hardware
+architecture.  The data parallel programming model simultaneously meets the
+seemingly conflicting goals of maintainability, portability, and performance.
+
+%% NEW - CER
+Unfortunately, this style of programming has never really caught on because
+when Fortran 90 was first introduced, performance was relatively poor and thus
+programmers shied away from using array syntax (even now, some are actively
+counseling against its usage because of performance issues~\cite{Cray-SC07}).
+Thus the Fortran community was caught in a classic ``chicken-and-egg''
+conundrum: (1) programmers didn't use it because it was slow; and (2)
+compilers vendors didn't improve it because programmers didn't use it.
+
+%% NEW - CER
+The goal of this research is to demonstrate that parallel programs written in
+this style of Fortran are maintainable and can achieve good performance on a
+variety of processor architectures, including possibly, future terascale
+architectures from Intel (such as the Larrabee architecture).  Using a code
+segment from a production code at Los Alamos (written entirely in this style
+of Fortran), we have demonstrated the potential for source-to-source
+transformations to target the IBM Cell architecture used in Los Alamos
+National Laboratories' Roadrunner computer~\footnote{Roadrunner is the first
+  computer to break the petaflop barrier.}.  Initial results show that
+transformed code using OFP and ROSE can achieve a speedup of up to 13 times on
+sixteen Roadrunner nodes, compared with the equivalent number of single-core
+Opteron nodes.
+
 This paper uses four simple features of Fortran that form a language subset
 that does not require any language extensions and can be easily transformed
-to a lower level implementation in either CUDA or OpenCL.
+to a lower-level implementation in either CUDA or OpenCL.
 
@@
 
+%% NEW - CER
+The use of array notation allows one to program in a data-parallel subset of
+Fortran. This style of programming makes use of array notation and pure and
+elemental functions to operate on array elements.
+
+%% NEW - CER
+Because elemental functions return scalar values and are free from side
+effects, the compiler is free to distribute the computation over any hardware
+processing elements available to it, such as the multiple cores and vector
+units on an Intel or AMD processor or the Synergistic Processing Units on an
+IBM Cell processor.
+
+%% NEW - CER
+An example of an elemental function is shown in Listing~\ref{lst:elemental}.
+This example simply returns its input divided by 2. While this may seem like a
+trivial example, such simple functions may be composed with other elemental
+functions to perform powerful computations, especially when applied to arrays.
+
 \subsubsection*{Pure procedures}
@@
 due to their suitability for compiler-level analysis.
+
+%% NEW - CER
+The use of an elemental function is shown in the pure procedure {\tt laplace}
+in Listing~\ref{lst:pure}. This function (for example) can be used to solve
+the steady-state temperature along a one-dimensional surface. The temperature
+at any point along the surface is equal to the average of the temperature at
+neighboring points. Rather than writing explicit loops and using indexing to
+reference neighboring points, one may simply shift the array {\tt T} to the
+left and right, add the results and divide by 2. The intrinsic function {\tt
+EOSHIFT} (end-off shift) semantically returns an array copy with the value at
+any element shifted by the amount given by the {\tt shift} argument. The value
+shifted in at the boundary is supplied by the {\tt boundary} argument. The
+elemental function {\tt one\_half} is used to divide by 2.
+
+%% NEW - CER
+Note that semantically the elemental functions return their array results by
+value. This is important because had loops been used to express the {\tt
+laplace} procedure, temporary variables would have been required. Otherwise,
+{\tt T(i)} would be updated with new values of {\tt T(i-1)}.
+
+%% NEW - CER
+ A ``pure'' function is one
+without side effects (though it may change the value of array elements in an
+array argument). An elemental function is a pure function with the additional
+constraint that it takes only scalar arguments and must return a scalar.
+Though defined in terms of scalars, surprisingly, an elemental function may
+take an array as an actual argument, with the compiler applying the function
+to each element of the array, and returning an array result. The plus operator
+as described in the previous paragraph can be seen as an elemental function,
+as the plus operator is defined in terms of scalars but can be applied to
+whole arrays and can return an array result.
 
@@
 
+%% NEW - CER
+Arrays are first class citizens in Fortran as they are part of the language
+type system rather than just a pointer to raw memory as in C. A Fortran array
+contains metadata associated with it that describes the array's shape and
+size. This metadata is lacking in other languages, such as C, C++, and Java.
+Even in the case where multidimensional array data structures are available
+(such as in C++ or Java), they exist outside the language standard, limiting
+the ability of the compiler to analyze them fully. The existence of this
+metadata allows the use of array notation whereby explicit indexing of arrays
+is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
+of the same rank and shape, then the statement {\tt C = A + B} results in the
+sum of elements of {\tt A} and {\tt B} being stored in the corresponding
+elements of {\tt C}. The first element of {\tt C} will contain the value of
+the first element of {\tt A} added to the first element of {\tt B}.
+
 The final piece that we exploit is the array notation present in modern
-versions of Fortran.  ....
+versions of Fortran.  For example,
+
+\begin{equation}
+\label{eq:dataparallel}
+A = B + s * C
+\end{equation}
+
+is a array assignment statement where A, B, and C are arrays and s is a
+scalar.  Note that no explicit iteration over array indices is needed and that
+the individual operators, plus, times, and assignment are applied by the
+compiler to individual elements of the arrays independently.  Thus the
+compiler is able to spread the computation in~\ref{eq:dataparallel} across any
+hardware threads under its control.
+
+There are several advantages to this style of programming:
+
+\begin{itemize}
+	\item There are no loops or index variables to keep track of.  Off by one
+index errors are a common programming mistake.
+        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
+	\item Semantically the intrinsic functions return arrays by value.  This is usually what the algorithm requires.
+	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
+\end{itemize}
+
+Complete and very concise and elegant programs can be built with procedures
+similar to the {\tt laplace} example shown above. To aid this effort, Fortran
+supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
+\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
+functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
+functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
+\texttt{PRODUCT}, ...).
 

### 2010-12-11 opencl/papers/foropencl.tex
Cleaned up after massive copy and paste operation.

@@
 %    \subsection{Why Fortran?}
-%    \subsection{Other Languages}
+%    \subsection{Comparison to Other Languages}
 %
@@
 %       \subsubsection*{Pure procedures}
-%       \subsubsection*{Shift operators}
+%       \subsubsection*{Shift functions}
 %       \subsubsection*{Regions}
+%    \subsection{New Functions}
 %    \subsection{Parallelism}
-%    \subsection{New Functions}
 %    \subsection{Limitations}
@@
 computer architectures.  The fundamental question we address is what
-programming model and what languages are best suited for these platforms.
+programming model and language constructs are best suited for these platforms.
+
+%% NEW - CER
+%%%We will examine how the existing data-parallel constructs in Fortran can be combined with coarrays or MPI to provide effectively a new parallel programming language, one that is evolutionary in nature and provides complete compatibility with existing applications and libraries.
+Data parallelism is a high-level abstraction that is, at the same time, both easier to program and gives the compiler more leeway (if fully exploited) in retargeting a program to different computer architectures.
 
@@
 
-%% NEW - CER
-The sheer complexity of programming for clusters of many or multi-core
-processors with tens of millions threads of execution make the simplicity of
-the data parallel model attractive.  Furthermore, the increasing complexity of
-todays applications (especially when convolved with the increasing complexity
-of the hardware) and the need for portability across hardware architectures
-make a higher-level and simpler programming model like data parallel
-attractive.
+We provide (via Fortran interfaces in the ForOpenCL library) a mechanism to
+call the C OpenCL runtime and enable Fortran programmers to access OpenCL
+kernels.  Transformations are supplied that provide a mechanism for converting
+Fortran procedures written in the Fortran subset described in this paper to
+OpenCL kernels.  We use the ROSE compiler infrastructure \cite{} to develop
+these transformations.  ROSE uses the Open Fortran Parser \cite{} to parse
+Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
+intermediate representation (IR) was constructed to represent multiple
+languages, it is relatively straightforward to transform high-level Fortran IR
+nodes to OpenCL nodes.  Furthermore, the Fortran array syntax maps directly to
+one, two, and three-dimensional thread groups in OpenCL.
 
-%% NEW - CER
-The goal of this work is to exploit source-to-source transformations that
-allow programmers to develop and maintain programs at a high-level of
-abstraction, without coding to a specific hardware architecture.
-Furthermore these transformations allow multiple hardware architectures
-to be targeted without changing the high-level source.  It also removes the
-necessity for application programmers to understand details of the accelerator
-architecture or to know OpenCL.
+Transformations for general Fortran procedures are not provided.  Furthermore,
+a mechanism to transform the calling site to automatically invoke
+OpenCL kernels is not provided at this time.  While it is possible to
+accomplish this task within ROSE, it is considered outside the scope of this
+paper.
 
-We use the ROSE compiler infrastructure \cite{} to develop these
-transformations.  ROSE uses the Open Fortran Parser \cite{} to parse Fortran
-2008 syntax and can output (unparse) to OpenCL.  Since ROSE's intermediate
-representation (IR) was constructed to represent multiple languages, it is
-relatively straightforward to transform high-level Fortran IR nodes to OpenCL
-nodes.  Furthermore, the Fortran array syntax maps directly to one, two, and
-three-dimensional thread groups in OpenCL.
+We examine the performance of the Fortran data-parallel abstraction when
+transformed to OpenCL to run on GPU architectures.  Since single node performance
+is often given as a reason for not using data-parallel constructs within Fortran,
+we consider the performance of serial data-parallel codes compared with the
+usage of explicit loop constructs.
 
-%%We study automatic transformations of Fortran for two applications examples.
-The first is a simple shallow water model in two dimensions using finite volume methods with ....
-for time update.
-The second is based on the primary compute kernels from PetaVision, a high-performance, neural science framework that is used in models of visual cortex.
+We study automatic transformations and the performance for an
+application example that is typical of many applications that are based on
+finite-difference or finite-volume methods in computational fluid dynamics
+(CFD).  The example is a simple shallow water model in two dimensions using
+finite volume methods with .... for time update.
 
-%% To simplify things, lets save PetaVision for the local stuff as it requires convolutions
-
-We study automatic transformations of Fortran for an application example that
-is typical of many applications that are based on finite-difference or
-finite-volume methods in computational fluid dynamics (CFD).  The example is a
-simple shallow water model in two dimensions using finite volume methods with
-.... for time update.
-
-%% NEW - CER
-PAGOSA is a production code at LANL completely written in data-parallel
-Fortran.  In a prior work we investigated automatically transforming this code
+%% NEW - CER - copied from proposal
+An initial study was made for an important procedure in PAGOSA,
+production code at LANL completely written in data-parallel
+Fortran.  We investigated automatically transforming this code
 to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
@@
 inter-node communication with computation on the Cell.
+
+%%We study automatic transformations of Fortran for two applications examples.
+The first is a simple shallow water model in two dimensions using finite volume methods with ....
+for time update.
+The second is based on the primary compute kernels from PetaVision, a high-performance, neural science framework that is used in models of visual cortex.
+
+%% To simplify things, lets save PetaVision for the local stuff as it requires convolutions
 
@@
 are considered to be more modern, Fortran has become a truly parallel language
-with features added to recent language standards. With coarrays, Fortran
-provides a mechanism to run replicas (called images) of a program on many
-different processors (or cores) at the same time. It also provides language
-constructs that allow the program to both read from and write to data residing
-on {\it remote} processors and to synchronize the different program images so
-that access to the data by the multiple images can be done atomically.
+with features added to recent language standards.
 
-A likely question that one may pose is ``\emph{Why Fortran and not a more
+Yet, a likely question that one may pose is ``\emph{Why Fortran and not a more
   modern language like X?}''  The recent rise in interest in concurrency and
@@
 
-%%
-%%
-%%
+\section{Comparison to Other Languages}
 
-\section{The programming model and Fortran-language subset}
+%% Matt you should provide an historical (and current) context here
+
+Techniques based on whole-array programming being mapped to high performance,
+parallel implementations are not new.
+
+Similar to ZPL, CM5 stencil compilers, CMFortran stuff.
+
+%% cites: ZPL, CM5 stencil compilers POOMA, HPL OpenMP, CUDA
+
+% put something here on: regions, array notation, ...
+
+\section{Programming Model}
 
 The static analysis and source-to-source transformations used are very
-basic and simply require the programmer to use a subset of Fortran
+basic but require the programmer to use a subset of Fortran
 that employs a data-parallel programming model.  In particular, it
@@
 and transformation method are based on.
+
+This paper uses four simple features of Fortran that form a language subset
+that does not require any language extensions and can be easily transformed
+to a lower-level implementation in either CUDA or OpenCL.
+
+\subsubsection*{Array notation}
 
@@
 %% NEW - CER
-We will examine how the existing data-parallel constructs in Fortran can be
-combined with coarrays or MPI to provide effectively a new parallel programming
-language, one that is evolutionary in nature and provides complete compatibility
-with existing applications and libraries.  Data parallelism is a high-level
-abstraction that is, at the same time, both easier to program and gives the
-compiler more leeway (if fully exploited) in retargeting a program to different
-computer architectures.
+Arrays are first class citizens in Fortran as they are part of the language
+type system rather than just a pointer to raw memory as in C. A Fortran array
+contains metadata associated with it that describes the array's shape and
+size. This metadata is lacking in other languages, such as C, C++, and Java.
+Even in the case where multidimensional array data structures are available
+(such as in C++ or Java), they exist outside the language standard, limiting
+the ability of the compiler to analyze them fully. The existence of this
+metadata allows the use of array notation whereby explicit indexing of arrays
+is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
+of the same rank and shape and {\tt s} is a scalar, then the statement
+
+\begin{verbatim}
+
+ C = A + s*B
+
+\end{verbatim}
+
+results in the sum of elements of {\tt A} and {\tt s} times the elements of
+{\tt B} being stored in the corresponding elements of {\tt C}. The first
+element of {\tt C} will contain the value of the first element of {\tt A}
+added to the first element of {\tt c*B}.  Note that no explicit iteration over
+array indices is needed and that the individual operators, plus, times, and
+assignment are applied by the compiler to individual elements of the arrays
+independently.  Thus the compiler is able to spread the computation in the
+example across any hardware threads under its control.
+
+%% NEW - CER
+%%% The use of array notation allows one to program in a data-parallel subset of Fortran.
+This style of programming makes use of array notation and pure and elemental functions to operate on array elements.
+
+\subsubsection*{Elemental functions}
+
+An elemental function consumes and produces scalar values, but can be applied
+to variables of array type such that the function is applied to each and every
+element of the array.  This allows programmers to avoid explicit looping and
+instead simply state that they intend a function to be applied to every
+element of an array in parallel, deferring the choice of implementation
+technique to the compiler.  Elemental functions are intended to be used for
+data parallel programming, and as a result must be side effect free (or,
+\emph{pure}).
+
+%% NEW - CER
+%%% Because elemental functions return scalar values and are free from side effects, the compiler is free to distribute the computation over any hardware processing elements available to it, such as the multiple cores and vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.
+
+%% NEW - CER
+An example of shown above using array syntax could be refactored to use
+the elemental function,
+
+\begin{verbatim}
+   pure elemental real function elem_example(a, b, s)
+      real, intent(in) :: a, b, s
+      elem_example = a + s*b
+   end function
+\end{verbatim}
+
+and called with
+
+\begin{verbatim}
+   C = elem_example(A, B, s)
+\end{verbatim}
+
+Note that while {\tt elem_example} is defined in terms of purely scalar
+quantities, it can be \emph{applied} to arrays as shown.
+
+While this may seem like a trivial example, such simple functions may be
+composed with other elemental functions to perform powerful computations,
+especially when applied to arrays.  ForOpenCL transforms elemental functions to inline OpenCL
+functions.  Thus there is no penalty for usage of elemental functions and provide
+a convient mechanism to express algorithms in simpler segments.
+
+Also note that the plus and times operators shown in the data-statement
+example can be seen as an elemental functions, as plus and times are defined
+in terms of scalars but can be applied to whole arrays and can return an array
+result.
+
+\subsubsection*{Pure procedures}
+
+Pure procedures, like elemental functions, must be free of side effects.
+(athough it may change the value of array elements in an array argument,
+unlike elemental arguments which must have the intent(in) attribute).
+The absence of side effects removes ordering constraints from the compiler
+allowing it to invoke pure functions out of order.  Procedures and functions
+of this sort are also common in pure functional languages like Haskell,
+and are exploited by compilers in order to emit parallel code automatically
+due to their suitability for compiler-level analysis.
+
+Since pure procedures don't have side effects they a candidates for running on
+accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
+OpenCL kernels that emph{do not} call other procedures, except for elemental
+functions.
+
+\subsubsection*{Shift functions}
+
+Many array-based algorithms require the same operation to be performed
+on each element of the array using the value of that element and some small
+set of neighboring cells.  Often programmers implement these operations that
+are local to each element within the inner-loop of a set of nested FOR-
+or DO-loops using offsets relative to the current array index.  An alternative
+to this local-view of the algorithm is to take a global view and write the
+algorithm in terms of the whole array.  For example, consider a 1D array in
+which we wish to subtract the $(i-1)$th element from the $i$th for all
+elements.  One way to look at this is that we are subtracting the entire array
+shifted by one element from itself.
+
+Fortran provides a set of shifting operators that allow programmers to
+define operations based on shifted arrays.  These intrinsic operators take
+an array, a dimension, and the amount by which it should be shifted (using
+the sign to indicate direction).  By defining operations on entire arrays
+based on a global view of them shifted relative to each other, programmers can
+avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
+analysis of the extent of the set of shifted arrays in a given expression
+allows analysis tools to determine the amount of temporary or buffer storage
+necessary to hold intermediate values during whole array operations.  With
+explicit loops, programmers must maintain this temporary storage manually.
+
+\subsubsection*{Regions}
+
+Borrowing from ZPL, we introduce the concept of regions to Fortran.  In
+Fortran, regions are expressed as pointers to a subsection of an existing array.
+Regions may refer to an interior portion of an array or the entire array.
+
+The use of regions implies the use of the halo (or ghost) cell pattern 
+where the size of an array is increased to provide extra array elements
+surrounding the interior portion of the array.  
+%parlab.eecs.berkeley.edu/wiki/_media/patterns/ghostcell.pdf
+
+Regions are similar to the shift operator as they can be used to reference
+portions of the array that are shifted with respect to the interior portion.
+However, unlike the shift operator, regions are not expressed in terms of
+boundary conditions and thus don't explicitly \emph{require} a knowledge of
+not the application of boundary conditions locally.  Thus, as will be shown
+below, regions are more suitable for usage by OpenCL thread groups which
+access only local subsections of an array stored in global memory.
+
+Unfortunately, as regions functions return pointers to arrays, they cannot
+be used in pure procedures in current Fortran.  We relax this constraint in
+ForOpenCL for region pointers only.
+
+\subsection{New Functions}
+
+% TODO - Craig - describe these functions
+%   region
+%   interior (can be written in terms of regions) but also specifies the thread
+%       group domain)
+%   more ...
+%
+
+\subsection{Parallelism}
+
+% TODO - Matt - do you want a try at this.
+
+% need to describe processor hierarchy of nodes and accelerators
+% need to describe memory hierarchy as well in regards to global array view,
+% node array view (with a copy is GPU global memory) and local GPU memory which
+% further subdivides the node array memory
+
+% figures on slicing and dicing of memory region go here
+
+There are several advantages to this style of programming using array syntax, shifts, regions,
+and pure and elemental functions:
+
+\begin{itemize}
+	\item There are no loops or index variables to keep track of.  Off by one
+index errors are a common programming mistake.
+        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
+	\item Semantically the intrinsic functions return arrays by value (though not regions).  This is usually what the algorithm requires.
+	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
+\end{itemize}
 
@@
 %% NEW - CER
-The power in this notation is that the compiler is free to distribute the
-computation in these expressions over any hardware processing elements available
-to it, such as the vector units on an Intel or AMD processor or the Synergistic
-Processing Units on an IBM Cell processor.  If the arrays are declared as
-coarrays, this includes spreading the computation over \emph{all of the nodes in
-  a cluster as well.}  In this case, communication occurs within the
-\texttt{cshift} functions, though the compiler is free to overlap communication
-and computation by scheduling the communication early and performing the
-computation on the interior of the arrays while waiting for the communication to
-complete.
+%%% The power in this notation is that the compiler is free to distribute the computation in these expressions over any hardware processing elements available to it, such as the vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.
+If the arrays are declared as coarrays, this includes spreading the computation over \emph{all of the nodes in   a cluster as well.}  In this case, communication occurs within the \texttt{cshift} functions, though the compiler is free to overlap communication and computation by scheduling the communication early and performing the computation on the interior of the arrays while waiting for the communication to complete.
 
 %% NEW - CER
-Complete and very concise and elegant programs can be built with statements
-similar to that shown in Listing~\ref{lst:dpexample}, using intrinsic
-functions like the array constructors (\texttt{CSHIFT}, \texttt{MERGE},
-\texttt{TRANSPOSE}, ...), the array location functions (\texttt{MAXLOC} and
-\texttt{MINLOC}), and the array reduction functions (\texttt{ANY},
-\texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM}, \texttt{PRODUCT},
-...)~\cite{guydp}.  It should be emphasized, however, that {\it no} compiler
-currently supplies coarray versions of these intrinsic functions as they are
-not defined by the proposed Fortran 2008 standard (we have submitted a paper
-proposing the integration of coarrays with these intricics functions to the J3
-Fortran committee~\cite{rasmussen-caf-intrinsics}).  With coarray or MPI
-implementations of these intrinsic functions, programs written in this
-data-parallel subset of Fortran are implicitly parallel.  Furthermore, in
-instances it is necessary to implement algorithms that don't lend themselves
-to expressibility in this data-parallel subset, coarray notation can be used
-for complete access to individual (including remote) data elements.  It is the
-combination of data parallel and coarray constructs, with the transformational
-properties of a source-to-source compiler (transforming array notation to
-optimized Fortran 95 and library calls), that make this possibility such a
-promising avenue for research.
+Complete and very concise and elegant programs can be built with procedures
+similar to the example shown above. To aid this effort, Fortran
+supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
+\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
+functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
+functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
+\texttt{PRODUCT}, ...).  To this set we add region functions described above.
 
@@
 This style of programming meets the requirements we have set for a programming
-model for developing petascale applications.  It allows the programmer to
+model for developing applications suitable for acceleration.  It allows the programmer to
 program at a very-high level of abstraction while providing the compiler with
@@
 %% NEW - CER
-The goal of this research is to demonstrate that parallel programs written in
+A goal of this paper is to demonstrate that parallel programs written in
 this style of Fortran are maintainable and can achieve good performance on a
 variety of processor architectures, including possibly, future terascale
-architectures from Intel (such as the Larrabee architecture).  Using a code
-segment from a production code at Los Alamos (written entirely in this style
-of Fortran), we have demonstrated the potential for source-to-source
-transformations to target the IBM Cell architecture used in Los Alamos
-National Laboratories' Roadrunner computer~\footnote{Roadrunner is the first
-  computer to break the petaflop barrier.}.  Initial results show that
-transformed code using OFP and ROSE can achieve a speedup of up to 13 times on
-sixteen Roadrunner nodes, compared with the equivalent number of single-core
-Opteron nodes.
-
-This paper uses four simple features of Fortran that form a language subset
-that does not require any language extensions and can be easily transformed
-to a lower-level implementation in either CUDA or OpenCL.
-
-\subsubsection*{Elemental functions}
-
-An elemental function consumes and produces scalar values, but can be
-applied to variables of array type such that the function is
-applied to each and every element of the array.  This allows programmers
-to avoid explicit looping and instead simply state that they intend a
-function to be applied to every element of an array in parallel, deferring
-the choice of implementation technique to the compiler.  Elemental
-functions are intended to be used for data parallel programming, and
-as a result must be side effect free (or, \emph{pure}).
-
-%% NEW - CER
-The use of array notation allows one to program in a data-parallel subset of
-Fortran. This style of programming makes use of array notation and pure and
-elemental functions to operate on array elements.
-
-%% NEW - CER
-Because elemental functions return scalar values and are free from side
-effects, the compiler is free to distribute the computation over any hardware
-processing elements available to it, such as the multiple cores and vector
-units on an Intel or AMD processor or the Synergistic Processing Units on an
-IBM Cell processor.
-
-%% NEW - CER
-An example of an elemental function is shown in Listing~\ref{lst:elemental}.
-This example simply returns its input divided by 2. While this may seem like a
-trivial example, such simple functions may be composed with other elemental
-functions to perform powerful computations, especially when applied to arrays.
-
-\subsubsection*{Pure procedures}
-
-Pure procedures, like elemental functions, must be free of side effects.
-The absence of side effects removes ordering constraints from the compiler
-allowing it to invoke pure functions out of order.  Procedures and functions
-of this sort are also common in pure functional languages like Haskell,
-and are exploited by compilers in order to emit parallel code automatically
-due to their suitability for compiler-level analysis.
-
-%% NEW - CER
-The use of an elemental function is shown in the pure procedure {\tt laplace}
-in Listing~\ref{lst:pure}. This function (for example) can be used to solve
-the steady-state temperature along a one-dimensional surface. The temperature
-at any point along the surface is equal to the average of the temperature at
-neighboring points. Rather than writing explicit loops and using indexing to
-reference neighboring points, one may simply shift the array {\tt T} to the
-left and right, add the results and divide by 2. The intrinsic function {\tt
-EOSHIFT} (end-off shift) semantically returns an array copy with the value at
-any element shifted by the amount given by the {\tt shift} argument. The value
-shifted in at the boundary is supplied by the {\tt boundary} argument. The
-elemental function {\tt one\_half} is used to divide by 2.
-
-%% NEW - CER
-Note that semantically the elemental functions return their array results by
-value. This is important because had loops been used to express the {\tt
-laplace} procedure, temporary variables would have been required. Otherwise,
-{\tt T(i)} would be updated with new values of {\tt T(i-1)}.
-
-%% NEW - CER
- A ``pure'' function is one
-without side effects (though it may change the value of array elements in an
-array argument). An elemental function is a pure function with the additional
-constraint that it takes only scalar arguments and must return a scalar.
-Though defined in terms of scalars, surprisingly, an elemental function may
-take an array as an actual argument, with the compiler applying the function
-to each element of the array, and returning an array result. The plus operator
-as described in the previous paragraph can be seen as an elemental function,
-as the plus operator is defined in terms of scalars but can be applied to
-whole arrays and can return an array result.
-
-\subsubsection*{Shift operators}
-
-Many array-based algorithms require the same operation to be performed
-on each element of the array using the value of that element and some small
-set of neighboring cells.  Often programmers implement these operations that
-are local to each element within the inner-loop of a set of nested FOR-
-or DO-loops using offsets relative to the current array index.  An alternative
-to this local-view of the algorithm is to take a global view and write the
-algorithm in terms of the whole array.  For example, consider a 1D array in
-which we wish to subtract the $(i-1)$th element from the $i$th for all
-elements.  One way to look at this is that we are subtracting the entire array
-shifted by one element from itself.
-
-Fortran provides a set of shifting operators that allow programmers to
-define operations based on shifted arrays.  These intrinsic operators take
-an array, a dimension, and the amount by which it should be shifted (using
-the sign to indicate direction).  By defining operations on entire arrays
-based on a global view of them shifted relative to each other, programmers can
-avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
-analysis of the extent of the set of shifted arrays in a given expression
-allows analysis tools to determine the amount of temporary or buffer storage
-necessary to hold intermediate values during whole array operations.  With
-explicit loops, programmers must maintain this temporary storage manually.
-
-\subsubsection*{Array notation}
-
-%% NEW - CER
-Arrays are first class citizens in Fortran as they are part of the language
-type system rather than just a pointer to raw memory as in C. A Fortran array
-contains metadata associated with it that describes the array's shape and
-size. This metadata is lacking in other languages, such as C, C++, and Java.
-Even in the case where multidimensional array data structures are available
-(such as in C++ or Java), they exist outside the language standard, limiting
-the ability of the compiler to analyze them fully. The existence of this
-metadata allows the use of array notation whereby explicit indexing of arrays
-is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
-of the same rank and shape, then the statement {\tt C = A + B} results in the
-sum of elements of {\tt A} and {\tt B} being stored in the corresponding
-elements of {\tt C}. The first element of {\tt C} will contain the value of
-the first element of {\tt A} added to the first element of {\tt B}.
-
-The final piece that we exploit is the array notation present in modern
-versions of Fortran.  For example,
-
-\begin{equation}
-\label{eq:dataparallel}
-A = B + s * C
-\end{equation}
-
-is a array assignment statement where A, B, and C are arrays and s is a
-scalar.  Note that no explicit iteration over array indices is needed and that
-the individual operators, plus, times, and assignment are applied by the
-compiler to individual elements of the arrays independently.  Thus the
-compiler is able to spread the computation in~\ref{eq:dataparallel} across any
-hardware threads under its control.
-
-There are several advantages to this style of programming:
-
-\begin{itemize}
-	\item There are no loops or index variables to keep track of.  Off by one
-index errors are a common programming mistake.
-        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
-	\item Semantically the intrinsic functions return arrays by value.  This is usually what the algorithm requires.
-	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
-\end{itemize}
-
-Complete and very concise and elegant programs can be built with procedures
-similar to the {\tt laplace} example shown above. To aid this effort, Fortran
-supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
-\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
-functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
-functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
-\texttt{PRODUCT}, ...).
-
-\subsection{Relationship to other languages}
-
-Techniques based on whole-array programming being mapped to high performance,
-parallel implementations are not new.
-
-Similar to ZPL, CM5 stencil compilers, CMFortran stuff.
-
-%% cites: ZPL, CM5 stencil compilers
-
-% put something here on:
-%
-% shifts
-% array notation
-%
-
-\subsubsection{elemental functions}
-
-\subsubsection{pure procedures}
-
-Mention that don't require pure procedures as region functions are not pure.
-But programmers should otherwise think in terms of writing pure procedures.
+architectures from Intel (such as the XXX architecture).
 
@@
 
+\section{Source-To-Source Transformations}
+
+% TODO - craig - I need to go through this section and replace PetaVision
+% with the shallow water code
+
 \subsection{ForOpenCL}
@@
 The notation uses uppercase for arrays and lowercase for scalar quantities.
-Interior array subsections are denoted by an i preceeding the array, e.g. iA
-is the interior subset of array A.
+Interior array subsections are denoted by an i succeeding the array, e.g. Ai
+is the interior region of array A.
 
@@
 
+\section{Conclusions}
+
+%% Explain what we have accomplished (copied from proposal)
+
+%% NEW - CER
+The sheer complexity of programming for clusters of many or multi-core
+processors with tens of millions threads of execution make the simplicity of
+the data parallel model attractive.  Furthermore, the increasing complexity of
+todays applications (especially when convolved with the increasing complexity
+of the hardware) and the need for portability across hardware architectures
+make a higher-level and simpler programming model like data parallel
+attractive.
+
+%% NEW - CER
+The goal of this work has been to exploit source-to-source transformations that
+allow programmers to develop and maintain programs at a high-level of
+abstraction, without coding to a specific hardware architecture.
+Furthermore these transformations allow multiple hardware architectures
+to be targeted without changing the high-level source.  It also removes the
+necessity for application programmers to understand details of the accelerator
+architecture or to know OpenCL.
+
+% Summarize results
+
 \section{Junk}

### 2010-12-12 opencl/papers/foropencl.tex
splitting into separate files

@@
 
-\section{Introduction}
-
-This paper presents a compiler-level approach for targeting a single
-program to multiple fundamentally different low level execution and
-programming models while allowing the application programmer to adopt
-a single high level programming model.  We show that features of
-Fortran 90 for data parallel programming are well suited to automatic
-transformation to generate code specifically tuned for different
-low-level programming models such as OpenCL and CUDA.  For algorithms
-that can be easily expressed in terms of whole array, data parallel operations,
-writing code in Fortran and transforming it automatically to specific low
-level implementations removes the burden from the programmer of working
-with tedious, error prone, low-level tools.
-
-1. Goal is to write once, transform many. Otherwise potentially reprogram for every architecture.
-2. Goal is to code for readability and maintainability, not performance
-3. This goal requires expression in a high-level language.
-4. But language must be simple enough for compiler to analyze.
-5. Thus ideal if language maps well to accelerator architectures.
-6. Data parallel constructs in Fortran 90 are chosen.
-7. Up to 65 times speed up measured on automatically transformed code.
-
-At Los Alamos National Laboratory (LANL) we have a wide variety of computer
-platforms from which to choose.  The common platform is made up of clusters of
-compute nodes with standard multi-core processors.  Some nodes also have
-accelerators that range from the IBM Cell processor (Roadrunner) to a variety
-of GPUs from NVIDIA and AMD.  Some nodes have hardware with vector
-instructions and others do not. The peak
-performance of the accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
-processor to XXX for NVIDIA Fermi. %% Fermi has another name
-
-%% NEW - CER
-However, the performance that these accelerator architectures offer comes at a
-cost as processor architectures are trending toward multiple cores with
-instances of integrated accelerator units and less of a reliance on memory
-caches.  These changes place a heavy burden on application programmers as they
-are forced to adapt to the new systems.  An especially challenging problem
-faced by application programmers is not only how to program to these new
-architectures (considering the massive scale of concurrency available), but
-also how to design programs that are portable across the changing landscape of
-computer architectures.  The fundamental question we address is what
-programming model and language constructs are best suited for these platforms.
-
-%% NEW - CER
-%%%We will examine how the existing data-parallel constructs in Fortran can be combined with coarrays or MPI to provide effectively a new parallel programming language, one that is evolutionary in nature and provides complete compatibility with existing applications and libraries.
-Data parallelism is a high-level abstraction that is, at the same time, both easier to program and gives the compiler more leeway (if fully exploited) in retargeting a program to different computer architectures.
-
-A data parallel programming model is well suited to emerging
-architectures that are based on either vector processing or massively
-parallel collections of simple cores.  A popular trend for programming
-systems such as GPU accelerators is the CUDA language from NVidia, or
-new languages like Chapel from Cray.  A less intrusive approach that
-is well suited to legacy applications and languages are language
-extensions or libraries that allow programmers to avoid adopting
-entirely new languages.  The recent OpenCL specification is intended
-to support this programming model, as are directive-based methods such
-as OpenMP or the Accelerator programming model from the Portland
-Group~\cite{pgi10accelerator}.
-
-The problem with many of these choices is that they expose too much detail
-about the machine architecture to the programmer.  
-This is particularly true of CUDA and OpenCL.  In CUDA, programmers must
-adapt their codes to fit the threading model used by NVidia GPUs, while
-OpenCL requires programmers to provide specially tuned versions of
-their code for different classes of machine.  In both cases, the programmer
-is also responsible for explicitly managing memory, including staging of
-data back and forth from the host CPU and the accelerator device memory.
-While these models have been attractive as a method for early adopters
-to utilize these new architectures, they are less attractive to programmers
-who do not have the time or resources to manually port their code to every
-new architecture and programming model that emerges.
-
-At this point in time it is not really possible to write once and
-run efficiently on the wide variety of computer platforms we have
-available.  For some classes of applications, we believe that this goal
-is possible using language constructs already present in a popular
-mainstream scientific programming language -- Fortran.
-
-\subsection{Approach}
-
-This paper addresses this issue by examining features in Fortran that
-allow programmers to express algorithms at a very high level but yet
-can be transformed by a compiler to run efficiently on a wide variety
-of platforms.  In particular we consider nodes of GPUs.
-
-We demonstrate that the array syntax of Fortran maps surprisingly well
-onto GPUs when transformed to OpenCL kernels.  These Fortran language
-features include pure and elemental functions and array constructs
-like where and shift.  In addition we add a few functions that enable
-a program to be deployed on machines with a hierarchy of processing
-elements, such as nodes employing GPU acceleration, \emph{without
-  requiring explicit declaration of parallelism within the program.}
-In addition the program uses entirely standard Fortran so it can be on
-a single core without concurrency.
-
-We provide (via Fortran interfaces in the ForOpenCL library) a mechanism to
-call the C OpenCL runtime and enable Fortran programmers to access OpenCL
-kernels.  Transformations are supplied that provide a mechanism for converting
-Fortran procedures written in the Fortran subset described in this paper to
-OpenCL kernels.  We use the ROSE compiler infrastructure \cite{} to develop
-these transformations.  ROSE uses the Open Fortran Parser \cite{} to parse
-Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
-intermediate representation (IR) was constructed to represent multiple
-languages, it is relatively straightforward to transform high-level Fortran IR
-nodes to OpenCL nodes.  Furthermore, the Fortran array syntax maps directly to
-one, two, and three-dimensional thread groups in OpenCL.
-
-Transformations for general Fortran procedures are not provided.  Furthermore,
-a mechanism to transform the calling site to automatically invoke
-OpenCL kernels is not provided at this time.  While it is possible to
-accomplish this task within ROSE, it is considered outside the scope of this
-paper.
-
-We examine the performance of the Fortran data-parallel abstraction when
-transformed to OpenCL to run on GPU architectures.  Since single node performance
-is often given as a reason for not using data-parallel constructs within Fortran,
-we consider the performance of serial data-parallel codes compared with the
-usage of explicit loop constructs.
-
-We study automatic transformations and the performance for an
-application example that is typical of many applications that are based on
-finite-difference or finite-volume methods in computational fluid dynamics
-(CFD).  The example is a simple shallow water model in two dimensions using
-finite volume methods with .... for time update.
-
-%% NEW - CER - copied from proposal
-An initial study was made for an important procedure in PAGOSA,
-production code at LANL completely written in data-parallel
-Fortran.  We investigated automatically transforming this code
-to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
-and IBM Cell processors).  We demonstrated that a source-to-source compiler
-can automatically vectorize and parallelize (both within the Cell processor
-and across nodes) a small section of this code.  Preliminary results showed a
-12 times performance gain on the Cell processor over a traditional single core
-processor.  A factor of 9 is gained from vectorizing and parallelizing the
-code to run on the Cell and the rest is gained from automatically overlapping
-inter-node communication with computation on the Cell.
-
-%%We study automatic transformations of Fortran for two applications examples.
-The first is a simple shallow water model in two dimensions using finite volume methods with ....
-for time update.
-The second is based on the primary compute kernels from PetaVision, a high-performance, neural science framework that is used in models of visual cortex.
-
-%% To simplify things, lets save PetaVision for the local stuff as it requires convolutions
-
-\subsection{Why Fortran?}
-
-%% NEW - CER
-Fortran \footnote{Originally named FORTRAN for FORmula TRANslation language.}
-is the oldest high-level programming language, and was developed to facilitate
-the translation of math formulae into machine code. Fortran was the first
-major language to use a compiler to translate from a high-level program
-representation to assembly language. Due to its age, it carries certain arcane
-baggage that later programming languages have evolved away from. As a result,
-Fortran has fallen into disfavor in certain programming circles.
-
-%% NEW - CER
-However, with the introduction of Fortran 90 (and later revisions to the
-standard), Fortran is a truly modern programming language. Much of the arcane
-baggage associated with it is actually no longer part of the language used by
-modern programmers, and exists in a deprecated form simply to support legacy
-codes. It is now modular and has many object-oriented features. Furthermore,
-Fortran should be of interest to those studying parallel programming because
-of its functional and data parallel constructs and because of the coarray
-notation introduced after Fortran 2003. Unlike languages like C and C++, which
-are considered to be more modern, Fortran has become a truly parallel language
-with features added to recent language standards.
-
-Yet, a likely question that one may pose is ``\emph{Why Fortran and not a more
-  modern language like X?}''  The recent rise in interest in concurrency and
-parallelism at the language level driven by multicore CPUs and manycore
-accelerators has driven a number of new language developments, both as novel
-languages and extensions on existing ones.
-
-%% NEW - CER
-So perhaps it is time to replace Fortran with yet another computer language.  The
-problems with replacing Fortran with an entirely new language are two fold:
-the economics of replacing the existing application base and the difficulty in
-obtaining programmer acceptance.  It is estimated that replacing a major
-production application at Los Alamos National Laboratory would cost between 50
-and 150 million dollars.  In terms of programmer acceptance, there is always
-the "chicken and egg problem": programmers won't use a new language until they
-can expect good performance across a variety of platforms, and compiler
-vendors can't afford to produce quality compilers until there is a reasonable
-expectation of a market.
-
-For scientific users, these new languages and language extensions
-present a challenge: how do developers effectively use them while
-avoiding rewriting their code and potentially growing dependent on a
-transient technology that will vanish tomorrow?
-
-%% NEW - CER
-History has also shown that an investment in rewriting code does not guarantee
-success either, as seen in an effort at LANL to modernize a legacy Fortran
-code with the the newer C++ POOMA framework.  This massive overhaul effort led
-to a code that was both slower and less flexible than the original Fortran
-\cite{}.  Similar experiences have occurred in the past, notably during the
-devlopment of the Sisal language.
-
-Fortran is unique in that it has contained language features that are
-well suited to modern architectures for a number of years.  This
-should be unsurprising --- Fortran was a primary language used to
-target systems such as the vector supercomputers and massively
-parallel systems of the 1970s and 1980s.  These are the systems in
-which architectural features were developed that have led to single
-chip high performance architectures of interest today.  Given that
-these new systems have features very similar to their predecessors, it
-is clear that the language features within Fortran for them are still
-relevant.
+\input{intro}
 
@@
 
-Note that while {\tt elem_example} is defined in terms of purely scalar
+Note that while {\tt elem\_example} is defined in terms of purely scalar
 quantities, it can be \emph{applied} to arrays as shown.

### 2010-12-12 opencl/papers/intro.tex
splitting into separate files

\section{Introduction}

This paper presents a compiler-level approach for targeting a single
program to multiple fundamentally different low level execution and
programming models while allowing the application programmer to adopt
a single high level programming model.  We show that features of
Fortran 90 for data parallel programming are well suited to automatic
transformation to generate code specifically tuned for different
low-level programming models such as OpenCL and CUDA.  For algorithms
that can be easily expressed in terms of whole array, data parallel operations,
writing code in Fortran and transforming it automatically to specific low
level implementations removes the burden from the programmer of working
with tedious, error prone, low-level tools.

1. Goal is to write once, transform many. Otherwise potentially reprogram for every architecture.
2. Goal is to code for readability and maintainability, not performance
3. This goal requires expression in a high-level language.
4. But language must be simple enough for compiler to analyze.
5. Thus ideal if language maps well to accelerator architectures.
6. Data parallel constructs in Fortran 90 are chosen.
7. Up to 65 times speed up measured on automatically transformed code.

At Los Alamos National Laboratory (LANL) we have a wide variety of computer
platforms from which to choose.  The common platform is made up of clusters of
compute nodes with standard multi-core processors.  Some nodes also have
accelerators that range from the IBM Cell processor (Roadrunner) to a variety
of GPUs from NVIDIA and AMD.  Some nodes have hardware with vector
instructions and others do not. The peak
performance of the accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
processor to XXX for NVIDIA Fermi. %% Fermi has another name

%% NEW - CER
However, the performance that these accelerator architectures offer comes at a
cost as processor architectures are trending toward multiple cores with
instances of integrated accelerator units and less of a reliance on memory
caches.  These changes place a heavy burden on application programmers as they
are forced to adapt to the new systems.  An especially challenging problem
faced by application programmers is not only how to program to these new
architectures (considering the massive scale of concurrency available), but
also how to design programs that are portable across the changing landscape of
computer architectures.  The fundamental question we address is what
programming model and language constructs are best suited for these platforms.

%% NEW - CER
%%%We will examine how the existing data-parallel constructs in Fortran can be combined with coarrays or MPI to provide effectively a new parallel programming language, one that is evolutionary in nature and provides complete compatibility with existing applications and libraries.  Data parallelism is a high-level abstraction that is, at the same time, both easier to program and gives the compiler more leeway (if fully exploited) in retargeting a program to different computer architectures.

A data parallel programming model is well suited to emerging
architectures that are based on either vector processing or massively
parallel collections of simple cores.  A popular trend for programming
systems such as GPU accelerators is the CUDA language from NVidia, or
new languages like Chapel from Cray.  A less intrusive approach that
is well suited to legacy applications and languages are language
extensions or libraries that allow programmers to avoid adopting
entirely new languages.  The recent OpenCL specification is intended
to support this programming model, as are directive-based methods such
as OpenMP or the Accelerator programming model from the Portland
Group~\cite{pgi10accelerator}.

The problem with many of these choices is that they expose too much detail
about the machine architecture to the programmer.  
This is particularly true of CUDA and OpenCL.  In CUDA, programmers must
adapt their codes to fit the threading model used by NVidia GPUs, while
OpenCL requires programmers to provide specially tuned versions of
their code for different classes of machine.  In both cases, the programmer
is also responsible for explicitly managing memory, including staging of
data back and forth from the host CPU and the accelerator device memory.
While these models have been attractive as a method for early adopters
to utilize these new architectures, they are less attractive to programmers
who do not have the time or resources to manually port their code to every
new architecture and programming model that emerges.

At this point in time it is not really possible to write once and
run efficiently on the wide variety of computer platforms we have
available.  For some classes of applications, we believe that this goal
is possible using language constructs already present in a popular
mainstream scientific programming language -- Fortran.

\subsection{Approach}

This paper addresses this issue by examining features in Fortran that
allow programmers to express algorithms at a very high level but yet
can be transformed by a compiler to run efficiently on a wide variety
of platforms.  In particular we consider nodes of GPUs.

We demonstrate that the array syntax of Fortran maps surprisingly well
onto GPUs when transformed to OpenCL kernels.  These Fortran language
features include pure and elemental functions and array constructs
like where and shift.  In addition we add a few functions that enable
a program to be deployed on machines with a hierarchy of processing
elements, such as nodes employing GPU acceleration, \emph{without
  requiring explicit declaration of parallelism within the program.}
In addition the program uses entirely standard Fortran so it can be on
a single core without concurrency.

We provide (via Fortran interfaces in the ForOpenCL library) a mechanism to
call the C OpenCL runtime and enable Fortran programmers to access OpenCL
kernels.  Transformations are supplied that provide a mechanism for converting
Fortran procedures written in the Fortran subset described in this paper to
OpenCL kernels.  We use the ROSE compiler infrastructure \cite{} to develop
these transformations.  ROSE uses the Open Fortran Parser \cite{} to parse
Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
intermediate representation (IR) was constructed to represent multiple
languages, it is relatively straightforward to transform high-level Fortran IR
nodes to OpenCL nodes.  Furthermore, the Fortran array syntax maps directly to
one, two, and three-dimensional thread groups in OpenCL.

Transformations for general Fortran procedures are not provided.  Furthermore,
a mechanism to transform the calling site to automatically invoke
OpenCL kernels is not provided at this time.  While it is possible to
accomplish this task within ROSE, it is considered outside the scope of this
paper.

We examine the performance of the Fortran data-parallel abstraction when
transformed to OpenCL to run on GPU architectures.  Since single node performance
is often given as a reason for not using data-parallel constructs within Fortran,
we consider the performance of serial data-parallel codes compared with the
usage of explicit loop constructs.

We study automatic transformations and the performance for an
application example that is typical of many applications that are based on
finite-difference or finite-volume methods in computational fluid dynamics
(CFD).  The example is a simple shallow water model in two dimensions using
finite volume methods with .... for time update.

%% NEW - CER - copied from proposal
An initial study was made for an important procedure in PAGOSA,
production code at LANL completely written in data-parallel
Fortran.  We investigated automatically transforming this code
to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
and IBM Cell processors).  We demonstrated that a source-to-source compiler
can automatically vectorize and parallelize (both within the Cell processor
and across nodes) a small section of this code.  Preliminary results showed a
12 times performance gain on the Cell processor over a traditional single core
processor.  A factor of 9 is gained from vectorizing and parallelizing the
code to run on the Cell and the rest is gained from automatically overlapping
inter-node communication with computation on the Cell.

%%We study automatic transformations of Fortran for two applications examples.  The first is a simple shallow water model in two dimensions using finite volume methods with .... for time update.  The second is based on the primary compute kernels from PetaVision, a high-performance, neural science framework that is used in models of visual cortex.

%% To simplify things, lets save PetaVision for the local stuff as it requires convolutions

\subsection{Why Fortran?}

%% NEW - CER
Fortran \footnote{Originally named FORTRAN for FORmula TRANslation language.}
is the oldest high-level programming language, and was developed to facilitate
the translation of math formulae into machine code. Fortran was the first
major language to use a compiler to translate from a high-level program
representation to assembly language. Due to its age, it carries certain arcane
baggage that later programming languages have evolved away from. As a result,
Fortran has fallen into disfavor in certain programming circles.

%% NEW - CER
However, with the introduction of Fortran 90 (and later revisions to the
standard), Fortran is a truly modern programming language. Much of the arcane
baggage associated with it is actually no longer part of the language used by
modern programmers, and exists in a deprecated form simply to support legacy
codes. It is now modular and has many object-oriented features. Furthermore,
Fortran should be of interest to those studying parallel programming because
of its functional and data parallel constructs and because of the coarray
notation introduced after Fortran 2003. Unlike languages like C and C++, which
are considered to be more modern, Fortran has become a truly parallel language
with features added to recent language standards.

Yet, a likely question that one may pose is ``\emph{Why Fortran and not a more
  modern language like X?}''  The recent rise in interest in concurrency and
parallelism at the language level driven by multicore CPUs and manycore
accelerators has driven a number of new language developments, both as novel
languages and extensions on existing ones.

%% NEW - CER
So perhaps it is time to replace Fortran with yet another computer language.  The
problems with replacing Fortran with an entirely new language are two fold:
the economics of replacing the existing application base and the difficulty in
obtaining programmer acceptance.  It is estimated that replacing a major
production application at Los Alamos National Laboratory would cost between 50
and 150 million dollars.  In terms of programmer acceptance, there is always
the "chicken and egg problem": programmers won't use a new language until they
can expect good performance across a variety of platforms, and compiler
vendors can't afford to produce quality compilers until there is a reasonable
expectation of a market.

For scientific users, these new languages and language extensions
present a challenge: how do developers effectively use them while
avoiding rewriting their code and potentially growing dependent on a
transient technology that will vanish tomorrow?

%% NEW - CER
History has also shown that an investment in rewriting code does not guarantee
success either, as seen in an effort at LANL to modernize a legacy Fortran
code with the the newer C++ POOMA framework.  This massive overhaul effort led
to a code that was both slower and less flexible than the original Fortran
\cite{}.  Similar experiences have occurred in the past, notably during the
devlopment of the Sisal language.

Fortran is unique in that it has contained language features that are
well suited to modern architectures for a number of years.  This
should be unsurprising --- Fortran was a primary language used to
target systems such as the vector supercomputers and massively
parallel systems of the 1970s and 1980s.  These are the systems in
which architectural features were developed that have led to single
chip high performance architectures of interest today.  Given that
these new systems have features very similar to their predecessors, it
is clear that the language features within Fortran for them are still
relevant.

### 2010-12-12 opencl/papers/foropencl.tex
a bit of history

@@
 \input{intro}
-
-\section{Comparison to Other Languages}
-
-%% Matt you should provide an historical (and current) context here
-
-Techniques based on whole-array programming being mapped to high performance,
-parallel implementations are not new.
-
-Similar to ZPL, CM5 stencil compilers, CMFortran stuff.
-
-%% cites: ZPL, CM5 stencil compilers POOMA, HPL OpenMP, CUDA
-
-% put something here on: regions, array notation, ...
-
-\section{Programming Model}
-
-The static analysis and source-to-source transformations used are very
-basic but require the programmer to use a subset of Fortran
-that employs a data-parallel programming model.  In particular, it
-encourages use of language features that were introduced and
-standardized in the Fortran 90 language specification.  In this
-section we describe the set of Fortran 90 features that our analysis
-and transformation method are based on.
-
-This paper uses four simple features of Fortran that form a language subset
-that does not require any language extensions and can be easily transformed
-to a lower-level implementation in either CUDA or OpenCL.
-
-\subsubsection*{Array notation}
-
-Fortran 90 introduced a rich array syntax that allows programmers to
-write code that is in terms of whole arrays or subarrays, with data
-parallel operators to compute on the arrays.  This has the benefit of
-avoiding explicit looping in the code and maintaining a high-level
-style that is closer to the original mathematical specification of the
-problem being solved.  More importantly from a compilation
-perspective, this defers decisions about how to implement these
-whole-array operations to a compilation tool.  When faced with a novel
-architecture such as modern GPUs that are ideally suited to data parallel
-programming models, the fit between Fortran arrays and these systems is
-quite clean.
-
-%% NEW - CER
-Arrays are first class citizens in Fortran as they are part of the language
-type system rather than just a pointer to raw memory as in C. A Fortran array
-contains metadata associated with it that describes the array's shape and
-size. This metadata is lacking in other languages, such as C, C++, and Java.
-Even in the case where multidimensional array data structures are available
-(such as in C++ or Java), they exist outside the language standard, limiting
-the ability of the compiler to analyze them fully. The existence of this
-metadata allows the use of array notation whereby explicit indexing of arrays
-is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
-of the same rank and shape and {\tt s} is a scalar, then the statement
-
-\begin{verbatim}
-
- C = A + s*B
-
-\end{verbatim}
-
-results in the sum of elements of {\tt A} and {\tt s} times the elements of
-{\tt B} being stored in the corresponding elements of {\tt C}. The first
-element of {\tt C} will contain the value of the first element of {\tt A}
-added to the first element of {\tt c*B}.  Note that no explicit iteration over
-array indices is needed and that the individual operators, plus, times, and
-assignment are applied by the compiler to individual elements of the arrays
-independently.  Thus the compiler is able to spread the computation in the
-example across any hardware threads under its control.
-
-%% NEW - CER
-%%% The use of array notation allows one to program in a data-parallel subset of Fortran.
-This style of programming makes use of array notation and pure and elemental functions to operate on array elements.
-
-\subsubsection*{Elemental functions}
-
-An elemental function consumes and produces scalar values, but can be applied
-to variables of array type such that the function is applied to each and every
-element of the array.  This allows programmers to avoid explicit looping and
-instead simply state that they intend a function to be applied to every
-element of an array in parallel, deferring the choice of implementation
-technique to the compiler.  Elemental functions are intended to be used for
-data parallel programming, and as a result must be side effect free (or,
-\emph{pure}).
-
-%% NEW - CER
-%%% Because elemental functions return scalar values and are free from side effects, the compiler is free to distribute the computation over any hardware processing elements available to it, such as the multiple cores and vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.
-
-%% NEW - CER
-An example of shown above using array syntax could be refactored to use
-the elemental function,
-
-\begin{verbatim}
-   pure elemental real function elem_example(a, b, s)
-      real, intent(in) :: a, b, s
-      elem_example = a + s*b
-   end function
-\end{verbatim}
-
-and called with
-
-\begin{verbatim}
-   C = elem_example(A, B, s)
-\end{verbatim}
-
-Note that while {\tt elem\_example} is defined in terms of purely scalar
-quantities, it can be \emph{applied} to arrays as shown.
-
-While this may seem like a trivial example, such simple functions may be
-composed with other elemental functions to perform powerful computations,
-especially when applied to arrays.  ForOpenCL transforms elemental functions to inline OpenCL
-functions.  Thus there is no penalty for usage of elemental functions and provide
-a convient mechanism to express algorithms in simpler segments.
-
-Also note that the plus and times operators shown in the data-statement
-example can be seen as an elemental functions, as plus and times are defined
-in terms of scalars but can be applied to whole arrays and can return an array
-result.
-
-\subsubsection*{Pure procedures}
-
-Pure procedures, like elemental functions, must be free of side effects.
-(athough it may change the value of array elements in an array argument,
-unlike elemental arguments which must have the intent(in) attribute).
-The absence of side effects removes ordering constraints from the compiler
-allowing it to invoke pure functions out of order.  Procedures and functions
-of this sort are also common in pure functional languages like Haskell,
-and are exploited by compilers in order to emit parallel code automatically
-due to their suitability for compiler-level analysis.
-
-Since pure procedures don't have side effects they a candidates for running on
-accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
-OpenCL kernels that emph{do not} call other procedures, except for elemental
-functions.
-
-\subsubsection*{Shift functions}
-
-Many array-based algorithms require the same operation to be performed
-on each element of the array using the value of that element and some small
-set of neighboring cells.  Often programmers implement these operations that
-are local to each element within the inner-loop of a set of nested FOR-
-or DO-loops using offsets relative to the current array index.  An alternative
-to this local-view of the algorithm is to take a global view and write the
-algorithm in terms of the whole array.  For example, consider a 1D array in
-which we wish to subtract the $(i-1)$th element from the $i$th for all
-elements.  One way to look at this is that we are subtracting the entire array
-shifted by one element from itself.
-
-Fortran provides a set of shifting operators that allow programmers to
-define operations based on shifted arrays.  These intrinsic operators take
-an array, a dimension, and the amount by which it should be shifted (using
-the sign to indicate direction).  By defining operations on entire arrays
-based on a global view of them shifted relative to each other, programmers can
-avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
-analysis of the extent of the set of shifted arrays in a given expression
-allows analysis tools to determine the amount of temporary or buffer storage
-necessary to hold intermediate values during whole array operations.  With
-explicit loops, programmers must maintain this temporary storage manually.
-
-\subsubsection*{Regions}
-
-Borrowing from ZPL, we introduce the concept of regions to Fortran.  In
-Fortran, regions are expressed as pointers to a subsection of an existing array.
-Regions may refer to an interior portion of an array or the entire array.
-
-The use of regions implies the use of the halo (or ghost) cell pattern 
-where the size of an array is increased to provide extra array elements
-surrounding the interior portion of the array.  
-%parlab.eecs.berkeley.edu/wiki/_media/patterns/ghostcell.pdf
-
-Regions are similar to the shift operator as they can be used to reference
-portions of the array that are shifted with respect to the interior portion.
-However, unlike the shift operator, regions are not expressed in terms of
-boundary conditions and thus don't explicitly \emph{require} a knowledge of
-not the application of boundary conditions locally.  Thus, as will be shown
-below, regions are more suitable for usage by OpenCL thread groups which
-access only local subsections of an array stored in global memory.
-
-Unfortunately, as regions functions return pointers to arrays, they cannot
-be used in pure procedures in current Fortran.  We relax this constraint in
-ForOpenCL for region pointers only.
-
-\subsection{New Functions}
-
-% TODO - Craig - describe these functions
-%   region
-%   interior (can be written in terms of regions) but also specifies the thread
-%       group domain)
-%   more ...
-%
-
-\subsection{Parallelism}
-
-% TODO - Matt - do you want a try at this.
-
-% need to describe processor hierarchy of nodes and accelerators
-% need to describe memory hierarchy as well in regards to global array view,
-% node array view (with a copy is GPU global memory) and local GPU memory which
-% further subdivides the node array memory
-
-% figures on slicing and dicing of memory region go here
-
-There are several advantages to this style of programming using array syntax, shifts, regions,
-and pure and elemental functions:
-
-\begin{itemize}
-	\item There are no loops or index variables to keep track of.  Off by one
-index errors are a common programming mistake.
-        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
-	\item Semantically the intrinsic functions return arrays by value (though not regions).  This is usually what the algorithm requires.
-	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
-\end{itemize}
-
-%% NEW - CER
-An example of this style of programming in Fortran is shown in
-
-\begin{verbatim}
-
-  Bz = Bz + dt * (cshift(Ex,dim=2,shift=+1) - Ex) / dy  &
-          - dt * (cshift(Ey,dim=1,shift=+1) - Ey) / dx
-
-\end{verbatim}
-
-%% NEW - CER
-This example is a solution to Maxwell's equations for the $z$ component of the
-magnetic field using Fortran array syntax.  Note that there are no explicit
-loops in this example.  The operators {\tt +}, {\tt -}, and {\tt *} are
-applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt
-  Ex}, and {\tt Ey}, individually.  This is why data parallelism has been
-called collection-oriented programming by
-Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift} function
-and the array-valued expressions all semantically returns a value, this style
-of programming is also similar to functional programming (or value-oriented
-programming). %%~\cite{simonpeytonjones}).  The heart of the solution to
-Maxwell's equations is the statement shown in Listing~\ref{lst:dpexample} and
-five similar, simple equations.
-
-%% NEW - CER
-%%% The power in this notation is that the compiler is free to distribute the computation in these expressions over any hardware processing elements available to it, such as the vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.
-If the arrays are declared as coarrays, this includes spreading the computation over \emph{all of the nodes in   a cluster as well.}  In this case, communication occurs within the \texttt{cshift} functions, though the compiler is free to overlap communication and computation by scheduling the communication early and performing the computation on the interior of the arrays while waiting for the communication to complete.
-
-%% NEW - CER
-Complete and very concise and elegant programs can be built with procedures
-similar to the example shown above. To aid this effort, Fortran
-supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
-\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
-functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
-functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
-\texttt{PRODUCT}, ...).  To this set we add region functions described above.
-
-%% NEW - CER
-This style of programming meets the requirements we have set for a programming
-model for developing applications suitable for acceleration.  It allows the programmer to
-program at a very-high level of abstraction while providing the compiler with
-maximum flexibility in targeting the application for a particular hardware
-architecture.  The data parallel programming model simultaneously meets the
-seemingly conflicting goals of maintainability, portability, and performance.
-
-%% NEW - CER
-Unfortunately, this style of programming has never really caught on because
-when Fortran 90 was first introduced, performance was relatively poor and thus
-programmers shied away from using array syntax (even now, some are actively
-counseling against its usage because of performance issues~\cite{Cray-SC07}).
-Thus the Fortran community was caught in a classic ``chicken-and-egg''
-conundrum: (1) programmers didn't use it because it was slow; and (2)
-compilers vendors didn't improve it because programmers didn't use it.
-
-%% NEW - CER
-A goal of this paper is to demonstrate that parallel programs written in
-this style of Fortran are maintainable and can achieve good performance on a
-variety of processor architectures, including possibly, future terascale
-architectures from Intel (such as the XXX architecture).
-
-\subsection{Limitations}
-
-Only Fortran procedures are transformed into OpenCL kernels.  The programmer
-must currently explicitly call these kernels from Fortran using the ForOpenCL
-library described below.  It is also possible using ROSE to modify the calling
-site so that the entire program can be transformed but this functionality is
-outside the scope of this paper.  Here we specifically examine transforming
-Fortran procedures to OpenCL kernels.
-
-Only elemental functions may be called from kernel functions.  These include
-fortran functions that have an OpenCL analog and user-defined elemental functions.
-
-Array sizes must be multiples of the local kernel size,
-{\tt get\_local\_size(0)*get\_local\_size(1)}.  This may be relaxed in the future.
-
-\section{Source-To-Source Transformations}
-
-% TODO - craig - I need to go through this section and replace PetaVision
-% with the shallow water code
-
-\subsection{ForOpenCL}
-
-ForOpenCL is library of Fortran modules.  It contains Fortran 2003 interface
-descriptions that allow language interoperability with the C OpenCL runtime.
-
-\section{Transformation examples}
-
-This sections show short Fortran code examples and OpenCL equivalent.
-The notation uses uppercase for arrays and lowercase for scalar quantities.
-Interior array subsections are denoted by an i succeeding the array, e.g. Ai
-is the interior region of array A.
-
-\subsubsection{array syntax}
-
-PetaVision updates the action potential with the statement,
-
-\begin{verbatim}
-   V = V - iA*(V - v_rest)                  !! Fortran
-
-   V[l] = V[l] - iA[lex]*(V[l] - v_rest);   // OpenCL equivalent
-\end{verbatim}
-
-\subsubsection{where construct}
-
-A neuron in PetaVision fires (activity is set to 1)
-whenever the action potential is greater than some threshold.
-This is easily expressed with a where construct.
-
-\begin{verbatim}
-   where (V > Vth)                        !! Fortran
-      iA = 1
-   elsewhere
-      iA = 0
-   end where
-
-   iA[lex] = (V[l] > Vth[l]) ? 1 : 0;     // OpenCL equivalent
-\end{verbatim}
-
-\subsection{New functions}
-
-%\subsubsection{transfer_halo}
-\subsubsection{region}
-
-One of the state variables in the shallow water code is H, effectively the
-height of the water.  This variable has a halo (ghost-cell region) surrounding
-the interior of the grid to handle boundary conditions.  When using
-MPI, the halo regions contains values from adjacent processors that must be
-updated with new values at each time step.  This is accomplished with the
-{\tt transfer\_halo} function.  The shallow water code assumes a five point stencil
-so the state variables are extended by 2 in each dimension (e.g., by one to the
-left, right, up, and down).
-
-\begin{verbatim}
-halo = [1,1,1,1]
-iH => region(H, halo, transfer=false)
-\end{verbatim}
-
-\section{Static Analysis}
-
-1. Dependence analysis to insert thread group barriers
-
-\begin{verbatim}
-   barrier(CLK_LOCAL_MEM_FENCE)
-\end{verbatim}
-
-2. subsection variables associated with array variables
-
-\subsection{Analysis not required}
-
-1. loop fusion
-2. removal of array temporaries (shifts)
+\input{programming-model}
+\input{source-to-source}
+\input{static-analysis}
 

### 2010-12-12 opencl/papers/intro.tex
a bit of history

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 low-level programming models such as OpenCL and CUDA.  For algorithms
-that can be easily expressed in terms of whole array, data parallel operations,
-writing code in Fortran and transforming it automatically to specific low
-level implementations removes the burden from the programmer of working
-with tedious, error prone, low-level tools.
+that can be easily expressed in terms of whole array, data parallel
+operations, writing code in Fortran and transforming it automatically
+to specific low level implementations removes the burden from the
+programmer of working with tedious, error prone, low-level tools.
 
@@
 relevant.
+
+\subsection{Comparison to Other Languages}
+
+A number of previous efforts have exploited data parallel programming
+at the language level to utilize novel architectuers, particularly
+in previous decades during the reign of vector and massively
+parallel computers in the high performance computing world.  The
+origin of the array syntax that was adopted in Fortran 90 can be found
+in the APL language\cite{iverson79apl}.  Fortran 90 differed from
+previous extensions of Fortran in that parallelism within whole-array operations
+was expressed at the expression level instead of via parallelism within explicit
+DO-loops (such as within IVTRAN for the Illiac IV).
+
+The High Performance Fortran (HPF) extension of Fortran 90 was
+proposed to add features to the language that would enhance the
+ability of compilers to emit fast parallel code for distributed and
+shared memory parallel computers\cite{koelbel94hpf}.  One of the
+notable additions to the language in HPF was syntax to specify the
+distribution of data structures amongst a set of parallel processors.
+HPF also introduced an alternative looping construct to the
+traditional DO-loop called \emph{FORALL} that was better suited for
+parallel compilation.  An additional keyword, \emph{INDEPENDENT}, was
+added to allow the programmer to indicate when the order of execution
+of the program (such as a sequence of loop iterations) can be flexible
+in order to allow parallel execution.  Interestingly, the parallelism
+features introduced in HPF did not exploit the new array features
+introduced in 1990 in any significant way, relying instead on explicit
+loop-based parallelism.  This was likely in order to support parallel
+programming that wasn't easily mapped onto a pure data parallel model.
+
+In some instances though, a purely data parallel model is appropriate
+for part or all of the major computations within a program.  One of
+the systems where programmers relied heavily on higher level
+operations instead of explicit looping constructs was the Thinking
+Machines Connection Machine 5 (CM-5).  A common programming pattern
+used on the CM-5 that we exploit in this paper was to write
+whole-array operations from a global perspective in which computations
+are expressed in terms of operations over the entire array instead of
+a single local index.  The use of the array shift intrinsic functions
+(like {\tt CSHIFT}) were used to build computations in which arrays
+were combined by shifting the entire arrays instead of working based
+on local offsets from singe indices.  A simple 1D example is one in
+which an element is replaced with the average of its own value with
+that of its two direct neighbors.  Ignoring boundary indices that wrap
+around, explicit indexing will result in a loop such as:
+
+\begin{verbatim}
+do i = 2,(n-1)
+  X(i) = (X(i-1) + X(i) + X(i+1)) / 3
+end do
+\end{verbatim}
+
+When shifts are employed, this can be expressed as:
+
+\begin{verbatim}
+X = (cshift(X,-1) + X + cshift(X,1)) / 3
+\end{verbatim}
+
+Similar whole array shifting was used in higher dimensions for finite
+different codes within the computational physics community, especially
+at Los Alamos for codes targetting the CM-5 system that resided there until the
+late 1990s.
+
+The whole-array model was attractive because it deferred
+responsibility for optimally implementing the computations to the
+compiler.  Instead of relying on a compiler to infer parallelism from
+a set of explicit loops, the choice for how to implement loops was
+left entirely up to the tool.  Unfortunately, this had two side
+effects that have limited broad acceptance of the whole-array
+programming model in Fortran.  First, programmers must translate their
+algorithms into a set of global operations.  Finite difference
+stencils and similar computations are traditionally defined in terms
+of offsets from some central index.  Shifting, while conceptually
+analogous, can be awkward to think about for high dimensional stencils
+with many points.  Second, the semantics of these operations are such
+that all elements of an array operation are updated as if they were
+updated simultaneously.  In a program where the programmer explicitly
+manages arrays and loops, double buffering techniques and user managed
+temporaries are used to maintain these semantics.  When the compiler
+is responsible for managing this intermediate storage, it has
+historically proven that they are inefficient and generate code that
+requires far more temporary storage than really necessary.  This is
+not a flaw of the language constructs, but a sign of the lack of
+sophistication of the compilers with respect to their internal
+analysis to determine how to optimally generate this intermediate
+storage.
+
+An interesting line of language research that grew out of the early work
+with HPF was that associated with the ZPL language work at the University
+of Washington~\ref{chamberlain04zpl}.  In ZPL, programmers adopt a simimlar
+global view of computation over arrays, but define their computations based
+on the local view of indices that participate in the update of each element of
+an array.  For example, in the 1D averaging case stated above, a programmer would
+define the computation in terms of the update that occurs at each index in terms
+of relative offsets from that index:
+
+\begin{verbatim}
+Write some ZPL
+\end{verbatim}
+
+This would then be executed over the entire array. {\bf more...}
+
+Stuff on stencil compilers?  No?

### 2010-12-12 opencl/papers/intro.tex
more

@@
 mainstream scientific programming language -- Fortran.
+A common, long-standing tongue-in-cheek response to new language developments
+in the scientific and high performance computing community is that a new language
+will arise to answer the needs of new systems, and it will be called Fortran.  
+We believe that the work presented in this paper validates that notion -- 
+we need a new language to work with, and that language is Fortran.
 
@@
 
-This paper addresses this issue by examining features in Fortran that
-allow programmers to express algorithms at a very high level but yet
-can be transformed by a compiler to run efficiently on a wide variety
-of platforms.  In particular we consider nodes of GPUs.
+This paper addresses the accelerator programmign problem by examining
+features in Fortran that allow programmers to express algorithms at a
+very high level that can be easily transformed by a compiler to run
+efficiently on a wide variety of platforms.  In particular we consider
+computers based on GPUs and related accelerator processors.
 
@@
 In addition the program uses entirely standard Fortran so it can be on
-a single core without concurrency.
+a single core without concurrency.  This work also is applicable
+to vendor-specific languages similar to OpenCL such as the NVidia
+CUDA language.
 
@@
 
-%% NEW - CER - copied from proposal
-An initial study was made for an important procedure in PAGOSA,
-production code at LANL completely written in data-parallel
-Fortran.  We investigated automatically transforming this code
-to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
-and IBM Cell processors).  We demonstrated that a source-to-source compiler
-can automatically vectorize and parallelize (both within the Cell processor
-and across nodes) a small section of this code.  Preliminary results showed a
-12 times performance gain on the Cell processor over a traditional single core
-processor.  A factor of 9 is gained from vectorizing and parallelizing the
-code to run on the Cell and the rest is gained from automatically overlapping
+An initial study was made for an important procedure in PAGOSA, a
+non-research, production-grade code at LANL completely written in
+data-parallel Fortran.  We investigated automatically transforming
+this code to run on LANL's Petaflop Roadrunner computer (a hybrid
+mixture of AMD Opterons and IBM Cell processors).  We demonstrated that a
+source-to-source compiler can automatically vectorize and parallelize
+(both within the Cell processor and across nodes) a small section of
+this code.  Preliminary results showed a 12 times performance gain on
+the Cell processor over a traditional single core processor.  A factor
+of 9 is gained from vectorizing and parallelizing the code to run on
+the Cell and the rest is gained from automatically overlapping
 inter-node communication with computation on the Cell.
@@
 
-%% NEW - CER
-Fortran \footnote{Originally named FORTRAN for FORmula TRANslation language.}
-is the oldest high-level programming language, and was developed to facilitate
-the translation of math formulae into machine code. Fortran was the first
-major language to use a compiler to translate from a high-level program
-representation to assembly language. Due to its age, it carries certain arcane
-baggage that later programming languages have evolved away from. As a result,
-Fortran has fallen into disfavor in certain programming circles.
+Fortran (originally named FORTRAN, the FORmula TRANslation
+language\footnote{The capitalization convention for referring the
+  language was changed in 1990 to \emph{Fortran}}) is the oldest
+high-level programming language in continuous use since its
+introduction, and was developed to facilitate the translation of math
+formulae into machine code. Fortran was the first major language to
+use a compiler to translate from a high-level program representation
+to assembly language. Due to its age, it carries certain arcane
+baggage that later programming languages have evolved away from. As a
+result, Fortran has fallen into disfavor in certain programming
+circles.  Modern versions of the language standardized in 1990, 1995, 2003, and 2008
+have removed much of this legacy baggage, but these changes are not
+widely known.  Modern Fortran exhibits features that are similar to 
+other modern programming languages, and does not mandate legacy features
+from decades old, deprecated versions of the language.
 
-%% NEW - CER
-However, with the introduction of Fortran 90 (and later revisions to the
-standard), Fortran is a truly modern programming language. Much of the arcane
-baggage associated with it is actually no longer part of the language used by
-modern programmers, and exists in a deprecated form simply to support legacy
-codes. It is now modular and has many object-oriented features. Furthermore,
-Fortran should be of interest to those studying parallel programming because
-of its functional and data parallel constructs and because of the coarray
-notation introduced after Fortran 2003. Unlike languages like C and C++, which
-are considered to be more modern, Fortran has become a truly parallel language
-with features added to recent language standards.
+With the introduction of Fortran 90 (and later revisions to the
+standard), Fortran became a truly modern programming language. Much of
+the arcane baggage associated with it is actually no longer part of
+the language used by modern programmers, and exists in a deprecated
+form simply to support legacy codes. It is now modular and has many
+object-oriented features.  It now includes a type system in which rich
+first-class array data types and corresponding syntax are part of the
+language, something that languages like C continue to lack.
+Furthermore, Fortran should be of interest to those studying parallel
+programming because of its functional and data parallel constructs and
+because of the coarray notation introduced in Fortran 2008. Unlike
+languages like C and C++, which are considered to be more modern,
+Fortran has become a truly parallel language with features added to
+recent language standards.
 

### 2010-12-13 opencl/papers/programming-model.tex
forgot to add these

\section{Programming Model}

The static analysis and source-to-source transformations used are very
basic but require the programmer to use a subset of Fortran
that employs a data-parallel programming model.  In particular, it
encourages use of language features that were introduced and
standardized in the Fortran 90 language specification.  In this
section we describe the set of Fortran 90 features that our analysis
and transformation method are based on.

This paper uses four simple features of Fortran that form a language subset
that does not require any language extensions and can be easily transformed
to a lower-level implementation in either CUDA or OpenCL.

\subsubsection*{Array notation}

Fortran 90 introduced a rich array syntax that allows programmers to
write code that is in terms of whole arrays or subarrays, with data
parallel operators to compute on the arrays.  This has the benefit of
avoiding explicit looping in the code and maintaining a high-level
style that is closer to the original mathematical specification of the
problem being solved.  More importantly from a compilation
perspective, this defers decisions about how to implement these
whole-array operations to a compilation tool.  When faced with a novel
architecture such as modern GPUs that are ideally suited to data parallel
programming models, the fit between Fortran arrays and these systems is
quite clean.

%% NEW - CER
Arrays are first class citizens in Fortran as they are part of the language
type system rather than just a pointer to raw memory as in C. A Fortran array
contains metadata associated with it that describes the array's shape and
size. This metadata is lacking in other languages, such as C, C++, and Java.
Even in the case where multidimensional array data structures are available
(such as in C++ or Java), they exist outside the language standard, limiting
the ability of the compiler to analyze them fully. The existence of this
metadata allows the use of array notation whereby explicit indexing of arrays
is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
of the same rank and shape and {\tt s} is a scalar, then the statement

\begin{verbatim}

 C = A + s*B

\end{verbatim}

results in the sum of elements of {\tt A} and {\tt s} times the elements of
{\tt B} being stored in the corresponding elements of {\tt C}. The first
element of {\tt C} will contain the value of the first element of {\tt A}
added to the first element of {\tt c*B}.  Note that no explicit iteration over
array indices is needed and that the individual operators, plus, times, and
assignment are applied by the compiler to individual elements of the arrays
independently.  Thus the compiler is able to spread the computation in the
example across any hardware threads under its control.

%% NEW - CER
%%% The use of array notation allows one to program in a data-parallel subset of Fortran. This style of programming makes use of array notation and pure and elemental functions to operate on array elements.

\subsubsection*{Elemental functions}

An elemental function consumes and produces scalar values, but can be applied
to variables of array type such that the function is applied to each and every
element of the array.  This allows programmers to avoid explicit looping and
instead simply state that they intend a function to be applied to every
element of an array in parallel, deferring the choice of implementation
technique to the compiler.  Elemental functions are intended to be used for
data parallel programming, and as a result must be side effect free (or,
\emph{pure}).

%% NEW - CER
%%% Because elemental functions return scalar values and are free from side effects, the compiler is free to distribute the computation over any hardware processing elements available to it, such as the multiple cores and vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.

%% NEW - CER
An example of shown above using array syntax could be refactored to use
the elemental function,

\begin{verbatim}
   pure elemental real function elem_example(a, b, s)
      real, intent(in) :: a, b, s
      elem_example = a + s*b
   end function
\end{verbatim}

and called with

\begin{verbatim}
   C = elem_example(A, B, s)
\end{verbatim}

Note that while {\tt elem\_example} is defined in terms of purely scalar
quantities, it can be \emph{applied} to arrays as shown.

While this may seem like a trivial example, such simple functions may be
composed with other elemental functions to perform powerful computations,
especially when applied to arrays.  ForOpenCL transforms elemental functions to inline OpenCL
functions.  Thus there is no penalty for usage of elemental functions and provide
a convient mechanism to express algorithms in simpler segments.

Also note that the plus and times operators shown in the data-statement
example can be seen as an elemental functions, as plus and times are defined
in terms of scalars but can be applied to whole arrays and can return an array
result.

\subsubsection*{Pure procedures}

Pure procedures, like elemental functions, must be free of side effects.
(athough it may change the value of array elements in an array argument,
unlike elemental arguments which must have the intent(in) attribute).
The absence of side effects removes ordering constraints from the compiler
allowing it to invoke pure functions out of order.  Procedures and functions
of this sort are also common in pure functional languages like Haskell,
and are exploited by compilers in order to emit parallel code automatically
due to their suitability for compiler-level analysis.

Since pure procedures don't have side effects they a candidates for running on
accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
OpenCL kernels that emph{do not} call other procedures, except for elemental
functions.

\subsubsection*{Shift functions}

Many array-based algorithms require the same operation to be performed
on each element of the array using the value of that element and some small
set of neighboring cells.  Often programmers implement these operations that
are local to each element within the inner-loop of a set of nested FOR-
or DO-loops using offsets relative to the current array index.  An alternative
to this local-view of the algorithm is to take a global view and write the
algorithm in terms of the whole array.  For example, consider a 1D array in
which we wish to subtract the $(i-1)$th element from the $i$th for all
elements.  One way to look at this is that we are subtracting the entire array
shifted by one element from itself.

Fortran provides a set of shifting operators that allow programmers to
define operations based on shifted arrays.  These intrinsic operators take
an array, a dimension, and the amount by which it should be shifted (using
the sign to indicate direction).  By defining operations on entire arrays
based on a global view of them shifted relative to each other, programmers can
avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
analysis of the extent of the set of shifted arrays in a given expression
allows analysis tools to determine the amount of temporary or buffer storage
necessary to hold intermediate values during whole array operations.  With
explicit loops, programmers must maintain this temporary storage manually.

\subsubsection*{Regions}

Borrowing from ZPL, we introduce the concept of regions to Fortran.  In
Fortran, regions are expressed as pointers to a subsection of an existing array.
Regions may refer to an interior portion of an array or the entire array.

The use of regions implies the use of the halo (or ghost) cell pattern 
where the size of an array is increased to provide extra array elements
surrounding the interior portion of the array.  
%parlab.eecs.berkeley.edu/wiki/_media/patterns/ghostcell.pdf

Regions are similar to the shift operator as they can be used to reference
portions of the array that are shifted with respect to the interior portion.
However, unlike the shift operator, regions are not expressed in terms of
boundary conditions and thus don't explicitly \emph{require} a knowledge of
not the application of boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which
access only local subsections of an array stored in global memory.

Unfortunately, as regions functions return pointers to arrays, they cannot
be used in pure procedures in current Fortran.  We relax this constraint in
ForOpenCL for region pointers only.

\subsection{New Functions}

% TODO - Craig - describe these functions
%   region
%   interior (can be written in terms of regions) but also specifies the thread
%       group domain)
%   more ...
%

\subsection{Parallelism}

% TODO - Matt - do you want a try at this.

% need to describe processor hierarchy of nodes and accelerators
% need to describe memory hierarchy as well in regards to global array view,
% node array view (with a copy is GPU global memory) and local GPU memory which
% further subdivides the node array memory

% figures on slicing and dicing of memory region go here

There are several advantages to this style of programming using array syntax, shifts, regions,
and pure and elemental functions:

\begin{itemize}
	\item There are no loops or index variables to keep track of.  Off by one
index errors are a common programming mistake.
        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
	\item Semantically the intrinsic functions return arrays by value (though not regions).  This is usually what the algorithm requires.
	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

%% NEW - CER
An example of this style of programming in Fortran is shown in

\begin{verbatim}

  Bz = Bz + dt * (cshift(Ex,dim=2,shift=+1) - Ex) / dy  &
          - dt * (cshift(Ey,dim=1,shift=+1) - Ey) / dx

\end{verbatim}

%% NEW - CER
This example is a solution to Maxwell's equations for the $z$ component of the
magnetic field using Fortran array syntax.  Note that there are no explicit
loops in this example.  The operators {\tt +}, {\tt -}, and {\tt *} are
applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt
  Ex}, and {\tt Ey}, individually.  This is why data parallelism has been
called collection-oriented programming by
Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift} function
and the array-valued expressions all semantically returns a value, this style
of programming is also similar to functional programming (or value-oriented
programming). %%~\cite{simonpeytonjones}).  The heart of the solution to
Maxwell's equations is the statement shown in Listing~\ref{lst:dpexample} and
five similar, simple equations.

%% NEW - CER
%%% The power in this notation is that the compiler is free to distribute the computation in these expressions over any hardware processing elements available to it, such as the vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.  If the arrays are declared as coarrays, this includes spreading the computation over \emph{all of the nodes in   a cluster as well.}  In this case, communication occurs within the \texttt{cshift} functions, though the compiler is free to overlap communication and computation by scheduling the communication early and performing the computation on the interior of the arrays while waiting for the communication to complete.

%% NEW - CER
Complete and very concise and elegant programs can be built with procedures
similar to the example shown above. To aid this effort, Fortran
supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
\texttt{PRODUCT}, ...).  To this set we add region functions described above.

%% NEW - CER
This style of programming meets the requirements we have set for a programming
model for developing applications suitable for acceleration.  It allows the programmer to
program at a very-high level of abstraction while providing the compiler with
maximum flexibility in targeting the application for a particular hardware
architecture.  The data parallel programming model simultaneously meets the
seemingly conflicting goals of maintainability, portability, and performance.

%% NEW - CER
Unfortunately, this style of programming has never really caught on because
when Fortran 90 was first introduced, performance was relatively poor and thus
programmers shied away from using array syntax (even now, some are actively
counseling against its usage because of performance issues~\cite{Cray-SC07}).
Thus the Fortran community was caught in a classic ``chicken-and-egg''
conundrum: (1) programmers didn't use it because it was slow; and (2)
compilers vendors didn't improve it because programmers didn't use it.

%% NEW - CER
A goal of this paper is to demonstrate that parallel programs written in
this style of Fortran are maintainable and can achieve good performance on a
variety of processor architectures, including possibly, future terascale
architectures from Intel (such as the XXX architecture).

\subsection{Limitations}

Only Fortran procedures are transformed into OpenCL kernels.  The programmer
must currently explicitly call these kernels from Fortran using the ForOpenCL
library described below.  It is also possible using ROSE to modify the calling
site so that the entire program can be transformed but this functionality is
outside the scope of this paper.  Here we specifically examine transforming
Fortran procedures to OpenCL kernels.

Only elemental functions may be called from kernel functions.  These include
fortran functions that have an OpenCL analog and user-defined elemental functions.

Array sizes must be multiples of the local kernel size,
{\tt get\_local\_size(0)*get\_local\_size(1)}.  This may be relaxed in the future.

### 2010-12-13 opencl/papers/source-to-source.tex
forgot to add these

\section{Source-To-Source Transformations}

% TODO - craig - I need to go through this section and replace PetaVision
% with the shallow water code

\subsection{ForOpenCL}

ForOpenCL is library of Fortran modules.  It contains Fortran 2003 interface
descriptions that allow language interoperability with the C OpenCL runtime.

\section{Transformation examples}

This sections show short Fortran code examples and OpenCL equivalent.
The notation uses uppercase for arrays and lowercase for scalar quantities.
Interior array subsections are denoted by an i succeeding the array, e.g. Ai
is the interior region of array A.

\subsubsection{array syntax}

PetaVision updates the action potential with the statement,

\begin{verbatim}
   V = V - iA*(V - v_rest)                  !! Fortran

   V[l] = V[l] - iA[lex]*(V[l] - v_rest);   // OpenCL equivalent
\end{verbatim}

\subsubsection{where construct}

A neuron in PetaVision fires (activity is set to 1)
whenever the action potential is greater than some threshold.
This is easily expressed with a where construct.

\begin{verbatim}
   where (V > Vth)                        !! Fortran
      iA = 1
   elsewhere
      iA = 0
   end where

   iA[lex] = (V[l] > Vth[l]) ? 1 : 0;     // OpenCL equivalent
\end{verbatim}

\subsection{New functions}

%\subsubsection{transfer_halo}
\subsubsection{region}

One of the state variables in the shallow water code is H, effectively the
height of the water.  This variable has a halo (ghost-cell region) surrounding
the interior of the grid to handle boundary conditions.  When using
MPI, the halo regions contains values from adjacent processors that must be
updated with new values at each time step.  This is accomplished with the
{\tt transfer\_halo} function.  The shallow water code assumes a five point stencil
so the state variables are extended by 2 in each dimension (e.g., by one to the
left, right, up, and down).

\begin{verbatim}
halo = [1,1,1,1]
iH => region(H, halo, transfer=false)
\end{verbatim}

### 2010-12-14 opencl/papers/intro.tex
text massaging

@@
 
-1. Goal is to write once, transform many. Otherwise potentially reprogram for every architecture.
-2. Goal is to code for readability and maintainability, not performance
-3. This goal requires expression in a high-level language.
-4. But language must be simple enough for compiler to analyze.
-5. Thus ideal if language maps well to accelerator architectures.
-6. Data parallel constructs in Fortran 90 are chosen.
-7. Up to 65 times speed up measured on automatically transformed code.
+In the ideal situation, application programmers would like to adopt
+a programming model in which they write their application once and
+use automated tools to retarget it to many architectures.  This has
+proven to be very challenging historically due to the subtle balance
+between high-level expressiveness of code and the performance of the
+lower-level code that is emitted by a compiler.  This ideal high-level
+model that programmers work with should emphasize readability,
+maintainability, and close proximity in abstraction to the problem being
+solved.  It should not be corrupted with details of specific target
+architectures solely for the purpose of single-system performance.
+For certain classes of applications, specifically those that map well
+onto a data-parallel programming model, we show that Fortran 90 contains
+language features that encourage high-level programming abstractions
+without sacrificing performance during low-level code generation.
 
-At Los Alamos National Laboratory (LANL) we have a wide variety of computer
-platforms from which to choose.  The common platform is made up of clusters of
-compute nodes with standard multi-core processors.  Some nodes also have
-accelerators that range from the IBM Cell processor (Roadrunner) to a variety
-of GPUs from NVIDIA and AMD.  Some nodes have hardware with vector
-instructions and others do not. The peak
-performance of the accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
-processor to XXX for NVIDIA Fermi. %% Fermi has another name
+%1. Goal is to write once, transform many. Otherwise potentially reprogram for every architecture.
+%2. Goal is to code for readability and maintainability, not performance
+%3. This goal requires expression in a high-level language.
+%4. But language must be simple enough for compiler to analyze.
+%5. Thus ideal if language maps well to accelerator architectures.
+%6. Data parallel constructs in Fortran 90 are chosen.
+%7. Up to 65 times speed up measured on automatically transformed code.
 
-%% NEW - CER
-However, the performance that these accelerator architectures offer comes at a
-cost as processor architectures are trending toward multiple cores with
-instances of integrated accelerator units and less of a reliance on memory
-caches.  These changes place a heavy burden on application programmers as they
-are forced to adapt to the new systems.  An especially challenging problem
-faced by application programmers is not only how to program to these new
-architectures (considering the massive scale of concurrency available), but
-also how to design programs that are portable across the changing landscape of
-computer architectures.  The fundamental question we address is what
-programming model and language constructs are best suited for these platforms.
+At Los Alamos National Laboratory (LANL), as with many supercomputing
+facilities today, users have a wide variety of computer platforms from
+which to choose.  The most common platform is made up of clusters of
+compute nodes with standard multi-core processors.  An increasingly
+common feature is that some nodes also have accelerators that range
+from the IBM Cell processor (such as the LANL Roadrunner system) to a
+variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
+vector instructions and others do not.  The peak performance of these
+accelerated nodes often resides in the hundreds of gigaflops.
+
+%The peak performance of the
+%accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
+%processor to XXX for NVIDIA Fermi. %% Fermi has another name
+
+However, the performance that these accelerator architectures offer
+comes at a cost as processor architectures are trending toward
+multiple cores with instances of integrated accelerator units with
+user managed memory and less of a reliance on superscalar instruction
+level parallelism and hardware managed memory hierarchies (such as
+traditional caches).  These changes place a heavy burden on
+application programmers as they are forced to adapt to the new
+systems.  An especially challenging problem faced by application
+programmers is not only how to program to these new architectures
+(considering the massive scale of concurrency available), but also how
+to design programs that are portable across the changing landscape of
+computer architectures with unique memory systems and programming
+models.  The fundamental question we address is what programming model
+and language constructs are best suited to span this set of new
+hardware designs.
 
@@
 
-A data parallel programming model is well suited to emerging
-architectures that are based on either vector processing or massively
-parallel collections of simple cores.  A popular trend for programming
-systems such as GPU accelerators is the CUDA language from NVidia, or
-new languages like Chapel from Cray.  A less intrusive approach that
-is well suited to legacy applications and languages are language
+A common theme amongst new processors is the emphasis on data parallel
+programming.  This model is well suited to emerging architectures that
+are based on either vector processing or massively parallel
+collections of simple cores.  A popular trend for programming systems
+such as GPU accelerators is the CUDA language from NVidia, or new
+languages like Chapel from Cray.  A less intrusive approach that is
+well suited to legacy applications and languages are language
 extensions or libraries that allow programmers to avoid adopting
@@
 
-The problem with many of these choices is that they expose too much detail
-about the machine architecture to the programmer.  
-This is particularly true of CUDA and OpenCL.  In CUDA, programmers must
-adapt their codes to fit the threading model used by NVidia GPUs, while
+The problem with many of these choices is that they expose too much
+detail about the machine architecture to the programmer.  This is
+particularly true of CUDA and OpenCL.  In CUDA, programmers must adapt
+their codes to fit the threading model used by NVidia GPUs, while
 OpenCL requires programmers to provide specially tuned versions of
-their code for different classes of machine.  In both cases, the programmer
-is also responsible for explicitly managing memory, including staging of
-data back and forth from the host CPU and the accelerator device memory.
-While these models have been attractive as a method for early adopters
-to utilize these new architectures, they are less attractive to programmers
-who do not have the time or resources to manually port their code to every
-new architecture and programming model that emerges.
+their code for different classes of machine.  In both cases, the
+programmer is also responsible for explicitly managing memory,
+including staging of data back and forth from the host CPU and the
+accelerator device memory.  While these models have been attractive as
+a method for early adopters to utilize these new architectures, they
+are less attractive to programmers who do not have the time or
+resources to manually port their code to every new architecture and
+programming model that emerges.
 
-At this point in time it is not really possible to write once and
-run efficiently on the wide variety of computer platforms we have
-available.  For some classes of applications, we believe that this goal
-is possible using language constructs already present in a popular
-mainstream scientific programming language -- Fortran.
-A common, long-standing tongue-in-cheek response to new language developments
-in the scientific and high performance computing community is that a new language
-will arise to answer the needs of new systems, and it will be called Fortran.  
-We believe that the work presented in this paper validates that notion -- 
-we need a new language to work with, and that language is Fortran.
+At this point in time it is not really possible to write once and run
+efficiently on the wide variety of computer platforms we have
+available.  For some classes of applications, we believe that this
+goal is possible using language constructs already present in a
+popular mainstream scientific programming language -- Fortran.  A
+common, long-standing tongue-in-cheek response to new language
+developments in the scientific and high performance computing
+community is that a new language will arise to answer the needs of new
+systems, and it will be called Fortran.  We believe that the work
+presented in this paper validates that notion -- we need a new
+language to work with, and that language is Fortran.
 

### 2010-12-14 opencl/papers/source-to-source.tex
Backup copy

@@
 
-% TODO - craig - I need to go through this section and replace PetaVision
-% with the shallow water code
+This section describes the transformations that take Fortran elemental and pure procedures
+as input and generate OpenCL kernels.
+
+\subsection{OpenCL}
+
+OpenCL~\cite{opencl:standard} is an open language standard for developing
+applications for accelerators.  The C-based language provides extensions for
+programming kernels that run on accelerator processing elements.  The kernels
+are run by calling a C runtime library from the OpenCL host (normally the
+CPU).  Efforts to standardize a C++ runtime are underway and Fortran
+interfaces to the C runtime are described later.
+
+An important concept in OpenCL is that of a thread and a thread group.  Thread
+groups are used to run an OpenCL kernel concurrently on several processor
+elements on the OpenCL device (often a GPU).  Consider a data-parallel
+statement written in terms of an elemental fuction as discussed above.  The
+act of running an OpenCL kernel can be thought of as having a particular
+thread assigned to each instance of the call to the elemental function as it
+is mapped across the arrays in the data-parallel statement.  In practice,
+these threads are packaged into thread groups when they are run on the device
+hardware.
+
+Device memory is separated hierarchically.  A thread instance has access to
+its own memory, thread groups to OpenCL local memory, and all thread groups
+have access to OpenCL global memory.  When multiple members of a thread group
+access the same memory elements, for example if {\tt region} or {\tt shift}
+functions are called, for performance reasons it is often best if global
+memory accessed by a thread group is copied into local memory.
+
+The \emph{region} and \emph{halo} constructs easily map onto the OpenCL memory
+hierarchy.  A schematic of this mapping is shown in Figure~\ref{fig:cl-memory}
+for a two-dimensional array with a 2x2 array of 4 thread groups.  The memory
+for the array and its halo are stored in global memory on the device as shown
+in the background layer of the figure.  The array copy in local memory is
+shown in the foreground divided into 4 \emph{local} tiles that partition the
+array.  Halo regions in global memory are shown in dark gray and halo regions
+in local memory are shown in light gray.
+
+We emphasize that the hierarchical distribution of memory used on the
+OpenCL device shown in Figure~\ref{fig:cl-memory} can be extended to include memory
+across MPI nodes as well.  In this case, the virtual global array is represented by the background
+layer (with its halo) and its partitions stored in the 4 MPI nodes shown in the
+foreground.
+
+Halo regions are constrained semantically so that they can not be written to
+by an OpenCL kernel because the \emph{region} and \emph{interior} functions
+return copies of the global memory.  Thus once memory for a halo region has
+been transferred into global device memory by all of the host nodes running
+MPI (before the OpenCL kernel is run), memory is in a consistent state so that
+the kernels are free to read from global device memory.  Because the local
+memory is a copy, it functions as a software cache for the local thread group.
+Thus the compiler must insert OpenCL barriers at proper locations in the code
+to insure that all threads have written to the local memory cache before any
+thread can read from the cache.  On exit from a kernel, the local memory cache
+is copied back to global memory for all \emph{interior} regions leaving global
+memory in a consistent state again.
 
@@
 
-ForOpenCL is library of Fortran modules.  It contains Fortran 2003 interface
-descriptions that allow language interoperability with the C OpenCL runtime.
+This work describes transformations that automatically create OpenCL kernel
+functions from Fortran pure and elemental procedures.  These transformations
+will not transform an entire program.  Users, for now, must explicitly replace
+calls to Fortran kernel procedures that run on the device, with calls to the
+OpenCL runtime that will run the kernel on the OpenCL host.  While these host
+transformations are straightforward using ROSE, they are outside the scope of
+this paper.
+
+In addition to the Fortran to OpenCL transformations, the
+ForOpenCL~\cite{http:ofp} library provides programmers with the ability to
+call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
+providing Fortran 2003 interface descriptions and classes that allow language
+interoperability with the OpenCL runtime.
 
@@
 
-This sections show short Fortran code examples and OpenCL equivalent.
-The notation uses uppercase for arrays and lowercase for scalar quantities.
-Interior array subsections are denoted by an i succeeding the array, e.g. Ai
-is the interior region of array A.
+This sections show short Fortran code examples and OpenCL equivalent.  The
+notation uses uppercase for arrays and lowercase for scalar quantities.
+Interior and region array copies are denoted by an i or an r preceeding the array.
+For example, {\tt iH = interior(H, halo)} is a local copy of interior region
+of array {\tt H} (representing height) in the shallow water code.
+
+\subsubsection{interior and region functions}
+
+While the serial versions of the interior and regions functions return an
+array copy, in OpenCL, these functions return a scalar quantity based on
+the location of a thread in a thread group and the relationship of its location
+to the array copy in local memory.  Because we assume there is a thread for
+every element in the interior, the array index is just the thread index adjusted
+for the halo.  Thus {\tt interior} and {\tt region} are just inline OpenCL
+functions provided by the ForOpenCL library.
 
 \subsubsection{array syntax}
-
-PetaVision updates the action potential with the statement,
 

### 2010-12-14 opencl/papers/shallow-water.tex
Initial version

\section{Shallow Water Model}

A short description of the shallow water model goes here.

\subsection{Serial code examples}

The Fortran kernel procedure {\tt wave_advance} for the shallow water code
is declared as:
\begin{verbatim}
   pure subroutine wave_advance(H, U, V, dx, dy, dt)
      real, dimension(:,:), intent(inout) :: H, U, V
      real, intent(in) :: dx, dy, dt
      !$OFP CONTIGUOUS :: H, U, V
      !$OFP KERNEL :: wave_advance
   end subroutine
\end{verbatim}
where {\tt H, U}, and {\tt V} are start variables for height and
$x$ and $y$ momentum respectively.  The \emph{OFP} compiler directives
{\tt CONTIGUOUS} and {\tt KERNEL} are added because contiguous is Fortran
2008 syntax (not in current compilers) and kernel is an \emph{OFP} extension.

Temporary arrays are required for the interior copies {\tt iH, iU}, and
{\tt iV} of the state variables and for the flux quantities {\tt Hx, Hy, Ux,
Vx}, and {\tt Vy} defined on cell faces.  These temporary arrays are declared as,
\begin{verbatim}
   real, allocatable, dimension(:,:) :: iH, iU, iV
   real, allocatable, dimension(:,:) :: Hx, Hy, Ux, Uy, Vx, Vy
\end{verbatim}

Halos variables for the interior and cell faces are defined as
\begin{verbatim}
   integer, dimension(4) :: halo, face_lt, face_rt, face_up, face_dn
   halo    = [1,1,1,1]
   face_lt = [0,1,1,1]; face_rt = [1,0,1,1]
   face_dn = [1,1,0,1]; face_up = [1,1,1,0]
\end{verbatim}
Note that the halos for the four faces each have a 0 in the definition. Thus the returned
array copy will have size that is larger than the interior regions that use {\tt [1,1,1,1]}.
This is because there is one more cell face quantity than there are cells in a given direction.

The Lax Wendrof (???) update algorithm for the shallow-water model has two steps.
The fluxes are first updated on the cell faces, for example,
\begin{verbatim}
   Hx =           0.5 * ( region(H, face_lt) + region(H, face_rt) ) &
        + (0.5*dt/dx) * ( region(U, face_lt) - region(U, face_rt) )
\end{verbatim}
updates the $x$-direction flux for the height state variable.

Then second step uses these fluxes to update the interior regions for the state varaibles.
For example,
\begin{verbatim}
   face_lt = [0,1,0,0];  face_rt = [1,0,0,0]
   face_dn = [0,0,0,1];  face_up = [0,0,1,0]
   iH = iH + (dt/dx) * ( region(Ux, face_lt) - region(Ux, face_rt) ) &
           + (dt/dy) * ( region(Vy, face_dn) - region(Vy, face_up) )
\end{verbatim}
Note that face halos have been redefined so that the array copy returned by
now has the same size as the interior region.

These simple code segments show how the shallow water model is
implemented in standard Fortran (2003) using the data-parallel
programming model described above.  The resulting code is simple,
consise, and easy to understand.  However it does \emph{not}
necessarily perform well because of the temporary array variables,
especially those produced by the {\tt region} function.  This is
generally true of algorithms that use Fortran shift functions as well,
as some Fortran compilers (e.g., gfortran) do not generate optimal
code for shifts.  We note (as seen below) that these temporary array
copies are replaced by scalars in the transformed Fortran code so
there is no performance penalty for using data-parallel statements
as outlined.

### 2010-12-14 opencl/papers/source-to-source.tex
Rough draft 0 stage

@@
 
-This section describes the transformations that take Fortran elemental and pure procedures
-as input and generate OpenCL kernels.
+This section describes the transformations that take Fortran elemental and
+pure procedures as input and generate OpenCL kernels.
 
@@
 
-We emphasize that the hierarchical distribution of memory used on the
-OpenCL device shown in Figure~\ref{fig:cl-memory} can be extended to include memory
-across MPI nodes as well.  In this case, the virtual global array is represented by the background
-layer (with its halo) and its partitions stored in the 4 MPI nodes shown in the
-foreground.
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.5in]{cl-memory.pdf}
+\caption{A schematic of global memory for an array and its copy stored in local memory
+for four thread groups.}
+\label{fig:cl-memory}
+\end{figure}
+
+We emphasize that the hierarchical distribution of memory used on the OpenCL
+device shown in Figure~\ref{fig:cl-memory} can be extended to include memory
+across MPI nodes as well.  In this case, the virtual global array is
+represented by the background layer (with its halo) and its partitions stored
+in the 4 MPI nodes shown in the foreground.
 
@@
 
-This sections show short Fortran code examples and OpenCL equivalent.  The
-notation uses uppercase for arrays and lowercase for scalar quantities.
-Interior and region array copies are denoted by an i or an r preceeding the array.
-For example, {\tt iH = interior(H, halo)} is a local copy of interior region
-of array {\tt H} (representing height) in the shallow water code.
+This sections outlines the OpenCL equivalent syntax for portions of the
+Fortran shallow-water code described in the previous section.  The notation
+uses uppercase for arrays and lowercase for scalar quantities.  Interior and
+region array copies are denoted by an i or an r preceeding the array.  For
+example, {\tt iH = interior(H, halo)} is a local copy of interior region of
+array {\tt H} (representing height) in the shallow water code.
 
@@
 While the serial versions of the interior and regions functions return an
-array copy, in OpenCL, these functions return a scalar quantity based on
-the location of a thread in a thread group and the relationship of its location
-to the array copy in local memory.  Because we assume there is a thread for
-every element in the interior, the array index is just the thread index adjusted
-for the halo.  Thus {\tt interior} and {\tt region} are just inline OpenCL
+array copy, in OpenCL, these functions return a scalar quantity based on the
+location of a thread in a thread group and the relationship of its location to
+the array copy in local memory.  Because we assume there is a thread for every
+element in the interior, the array index is just the thread index adjusted for
+the halo.  Thus {\tt interior} and {\tt region} are just inline OpenCL
 functions provided by the ForOpenCL library.
+
+\subsubsection{function and variable declarations}
+
+Fortran kernel procedures have direct correspondence with OpenCL equivalents.
+For example, {\tt wave_advance} is transformed as {\tt __kernel void
+  wave_advance(__ global float * H, ..., float dt);}.  The global state
+variables have local equivalents, e.g., {\tt __local float
+  H_local[MAX_LOCAL_SIZE]}, declared with the appropriate size and are copied
+to local memory by the compiler with inlined library functions.  The flux
+array temporaries (region variables) are declared similarly but initialized to
+zero with inlined library functions.  Interior variables are simple scalars,
+e.g., {\tt float iH;}.  Region variables cannot be scalar objects because
+regions are shifted and thus \emph{shared} by threads within a thread group.
 
@@
 
+Array syntax transforms nearly directly to OpenCL code.  For example, interior
+variables are particularly straightforward as they as scalar quantities in
+OpenCL,
 \begin{verbatim}
-   V = V - iA*(V - v_rest)                  !! Fortran
-
-   V[l] = V[l] - iA[lex]*(V[l] - v_rest);   // OpenCL equivalent
+   iH = iH + (dt/dx) * ( region(Ux, face_lt) - region(Ux, face_rt) )
+           + (dt/dy) * ( region(Vy, face_dn) - region(Vy, face_up) );
 \end{verbatim}
 
-\subsubsection{where construct}
+Region variables are more complicated because they are arrays.
+\begin{verbatim}
+   Hx[l] = 0.5 * ( region(H_local, face_lt) + ... );
+\end{verbatim}
+where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
+  LX = get_local_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
+  get_local_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
+  get_local_size(0)} is the size of the thread group in the $x$ dimension, and
+the {\tt get_local_id} and {\tt get_local_size} functions are defined by the
+OpenCL compiler.
 
-A neuron in PetaVision fires (activity is set to 1)
-whenever the action potential is greater than some threshold.
-This is easily expressed with a where construct.
+\subsection{Performance measurements}
 
-\begin{verbatim}
-   where (V > Vth)                        !! Fortran
-      iA = 1
-   elsewhere
-      iA = 0
-   end where
+Performance measurements were made comparing the transformed code with
+different versions of the serial shallow-water code.  The serial versions
+include the original written in C and two Fortran versions: one using copy
+semantics for the regions and one using reference semantics.  We also compare
+with a hand written OpenCL implementation that was optimized for local memory
+usage (no array temporaries).  The measurments were made using an NVIDIA Fermi
+GPU (TODO - need model numbers) running at XXX Hz and a VENDOR hexacore CPU
+running at XXX Hz.  The compilers were gfortran and gcc VERSIONS with
+optimization of -O3.
 
-   iA[lex] = (V[l] > Vth[l]) ? 1 : 0;     // OpenCL equivalent
-\end{verbatim}
+The performance comparison is shown in Figure~\ref{fig:cl-performance} for
+varying array sizes (of the state variables).  The time represents an average
+of 100 iterations of the outer time-advance loop calling the OpenCL kernel.
+This tight loop keeps the OpenCL kernel supplied with threads to take
+advantage of potential latency hiding by the NVIDIA GPU.  Any serial code
+within this loop would reduce the measured values.
 
-\subsection{New functions}
+As can be seen in Figure~\ref{fig:cl-performance}, the transformed get very
+good results of roughly 65 times speedup over the best Fortran code and YYY
+speedup over the original C code.  It comes with XXX \% of the hand coded
+OpenCL version.
 
-%\subsubsection{transfer_halo}
-\subsubsection{region}
-
-One of the state variables in the shallow water code is H, effectively the
-height of the water.  This variable has a halo (ghost-cell region) surrounding
-the interior of the grid to handle boundary conditions.  When using
-MPI, the halo regions contains values from adjacent processors that must be
-updated with new values at each time step.  This is accomplished with the
-{\tt transfer\_halo} function.  The shallow water code assumes a five point stencil
-so the state variables are extended by 2 in each dimension (e.g., by one to the
-left, right, up, and down).
-
-\begin{verbatim}
-halo = [1,1,1,1]
-iH => region(H, halo, transfer=false)
-\end{verbatim}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.5in]{cl-performance.pdf}
+\caption{Performance comparison for varying array size.}
+\label{fig:cl-performance}
+\end{figure}

### 2010-12-15 opencl/papers/programming-model.tex
Added function discussion.

@@
 Pure procedures, like elemental functions, must be free of side effects.
-(athough it may change the value of array elements in an array argument,
+(athough they may change the value of array elements in an array argument,
 unlike elemental arguments which must have the intent(in) attribute).
@@
 accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
-OpenCL kernels that emph{do not} call other procedures, except for elemental
+OpenCL kernels that \emph{do not} call other procedures, except for elemental
 functions.
@@
 boundary conditions and thus don't explicitly \emph{require} a knowledge of
-not the application of boundary conditions locally.  Thus, as will be shown
+nor the application of boundary conditions locally.  Thus, as will be shown
 below, regions are more suitable for usage by OpenCL thread groups which
@@
 
-Unfortunately, as regions functions return pointers to arrays, they cannot
-be used in pure procedures in current Fortran.  We relax this constraint in
-ForOpenCL for region pointers only.
+\subsection{New Procedures}
 
-\subsection{New Functions}
+Unlike OpenMP, we require limited use of compiler directives, although
+two are used to enforce the semantics required by the programming
+model.  These directives are:
+\begin{itemize}
+   \item !$OFP CONTIGUOUS: specifies that an dummy array variable is
+   contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.
+   \item !$OFP KERNEL: specifies that a pure subroutine can be
+     transformed to an OpenCL kernel.
+\end{itemize}
 
-% TODO - Craig - describe these functions
-%   region
-%   interior (can be written in terms of regions) but also specifies the thread
-%       group domain)
-%   more ...
-%
+Three new procedures (in addition to the intrinsic shift function) are
+defined in Fortran that are used in array-syntax operations.  Each
+procedure takes a integer array halo argument that specifies the
+number of ghost cells on either side of a region, for each dimension.
+For example {\tt halo = [left, right, down, up]} specifies a halo for
+a two-dimensional region.  These functions are:
+\begin{itemize}
+   \item transfer_halo(array, halo): an impure subroutine that
+     exchanges halo regions between nodes using MPI or Fortran
+     coarrays. Array is the memory stored on the node representing the
+     node-local portion of a virtual global array. Not used in this work.
+   \item interior(array, halo): a pure function that returns a copy of
+     the interior portion of the array specified by halo. Array is
+     local to a node.
+   \item region(array, halo): a pure function that returns a copy of
+     the portion of the array specified by halo.  Array is local to a node.
+\end{itemize}
+
+It should be noted that the two functions {\tt interior} and {\tt
+  region} are pure and thus can be called from within a pure kernel
+procedure.  These two functions are part of the language recognized by
+the compiler and though the two functions return a copy of a portion
+of an array \emph{semantically}, the compiler is not forced to
+actually make a copy and is free to enforce copy semantics through
+other means.
 
@@
 
-% TODO - Matt - do you want a try at this.
+% TODO - Matt - do you want a try at this.  Not sure we have room for
+% this subsection. I briefly mention node parallelism in other sections.
 
@@
         \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
-	\item Semantically the intrinsic functions return arrays by value (though not regions).  This is usually what the algorithm requires.
+	\item Semantically the intrinsic functions return arrays by value.  This is usually what the algorithm requires.
 	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
@@
 
-Array sizes must be multiples of the local kernel size,
-{\tt get\_local\_size(0)*get\_local\_size(1)}.  This may be relaxed in the future.
+A kernel procedure (specified by the !$OFP KERNEL directive) must be
+pure and all array variables must be declared as contiguous.  A kernel
+procedure may not call other procedures except for limited intrisic
+functions (primarily math), user-defined elemental function, and the
+{\tt interior} and {\tt region} functions.
+
+Array sizes must be multiples of the local kernel size, {\tt
+  get\_local\_size(0)*get\_local\_size(1)}.  This may be relaxed in
+the future.

### 2010-12-15 opencl/papers/source-to-source.tex
minor change

@@
 with a hand written OpenCL implementation that was optimized for local memory
-usage (no array temporaries).  The measurments were made using an NVIDIA Fermi
-GPU (TODO - need model numbers) running at XXX Hz and a VENDOR hexacore CPU
-running at XXX Hz.  The compilers were gfortran and gcc VERSIONS with
+usage (no array temporaries).  The measurments were made using an NVIDIA Tesla C2050
+GPU (TODO - need model numbers) running at XXX Hz and an Intel XXX hexacore CPU
+running at XXX Hz with 2.56 GB of RAM.  The compilers were gfortran and gcc VERSIONS with
 optimization of -O3.

### 2010-12-15 opencl/papers/programming-model.tex
cleaning up after Craig....

@@
 \begin{itemize}
-   \item !$OFP CONTIGUOUS: specifies that an dummy array variable is
+   \item {\tt \!\$OFP CONTIGUOUS}: specifies that an dummy array variable is
    contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.
-   \item !$OFP KERNEL: specifies that a pure subroutine can be
+   \item {\tt \!\$OFP KERNEL}: specifies that a pure subroutine can be
      transformed to an OpenCL kernel.
@@
 \begin{itemize}
-   \item transfer_halo(array, halo): an impure subroutine that
+   \item {\tt transfer\_halo(array, halo)}: an impure subroutine that
      exchanges halo regions between nodes using MPI or Fortran
@@
      node-local portion of a virtual global array. Not used in this work.
-   \item interior(array, halo): a pure function that returns a copy of
+   \item {\tt interior(array, halo)}: a pure function that returns a copy of
      the interior portion of the array specified by halo. Array is
      local to a node.
-   \item region(array, halo): a pure function that returns a copy of
+   \item {\tt region(array, halo)}: a pure function that returns a copy of
      the portion of the array specified by halo.  Array is local to a node.
@@
 
-A kernel procedure (specified by the !$OFP KERNEL directive) must be
+A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
 pure and all array variables must be declared as contiguous.  A kernel

### 2010-12-15 opencl/papers/source-to-source.tex
cleaning up after Craig....

@@
 Fortran kernel procedures have direct correspondence with OpenCL equivalents.
-For example, {\tt wave_advance} is transformed as {\tt __kernel void
-  wave_advance(__ global float * H, ..., float dt);}.  The global state
-variables have local equivalents, e.g., {\tt __local float
-  H_local[MAX_LOCAL_SIZE]}, declared with the appropriate size and are copied
+For example, {\tt wave\_advance} is transformed as {\tt \_\_kernel void
+  wave\_advance(\_\_ global float * H, ..., float dt);}.  The global state
+variables have local equivalents, e.g., {\tt \_\_local float
+  H\_local[MAX\_LOCAL\_SIZE]}, declared with the appropriate size and are copied
 to local memory by the compiler with inlined library functions.  The flux
@@
 where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
-  LX = get_local_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
-  get_local_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
-  get_local_size(0)} is the size of the thread group in the $x$ dimension, and
-the {\tt get_local_id} and {\tt get_local_size} functions are defined by the
+  LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
+  get\_local\_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
+  get\_local\_size(0)} is the size of the thread group in the $x$ dimension, and
+the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
 OpenCL compiler.
@@
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=2.5in]{cl-performance.pdf}
-\caption{Performance comparison for varying array size.}
-\label{fig:cl-performance}
-\end{figure}
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=2.5in]{cl-performance.pdf}
+%\caption{Performance comparison for varying array size.}
+%\label{fig:cl-performance}
+%\end{figure}

### 2010-12-15 opencl/papers/foropencl.tex
Added shallow-water.tex
cleaning up verbatim mess

@@
 %
+% \section{Shallow Water Model}
+%    \subsection{Equations}
+%
+
+%
 % \section{Source-To-Source Transformations}
@@
 %    \subsection{Simplifying Assumptions}
-%
-
-%
-% \section{Shallow Water Model}
-%    \subsection{Equations}
 %
@@
 \usepackage{url}
+\usepackage{listings}
+
+\lstdefinestyle{FortranLike}{float,frame=lines,language=Fortran,commentstyle=\ttfamily,basicstyle=\ttfamily}
+\lstdefinestyle{CLike}{float,frame=lines,language=C,commentstyle=\ttfamily,basicstyle=\ttfamily}
+\lstdefinestyle{NoFloatCLike}{frame=lines,language=C,commentstyle=\ttfamily,basicstyle=\ttfamily}
 
@@
 \input{programming-model}
+\input{shallow-water}
 \input{source-to-source}

### 2010-12-15 opencl/papers/programming-model.tex
cleaning up verbatim mess

@@
 
+{\small
 \begin{verbatim}
-
  C = A + s*B
-
 \end{verbatim}
+}
 
@@
 
+{\small
 \begin{verbatim}
-   pure elemental real function elem_example(a, b, s)
+  pure elemental real function foo(a, b, s)
     real, intent(in) :: a, b, s
-      elem_example = a + s*b
+    foo = a + s*b
   end function
 \end{verbatim}
+}
 
-and called with
+\noindent and called with
 
+{\small
 \begin{verbatim}
-   C = elem_example(A, B, s)
+  C = foo(A, B, s)
 \end{verbatim}
+}
 
-Note that while {\tt elem\_example} is defined in terms of purely scalar
+Note that while {\tt foo} is defined in terms of purely scalar
 quantities, it can be \emph{applied} to arrays as shown.
@@
 
+{\small
 \begin{verbatim}
-
-  Bz = Bz + dt * (cshift(Ex,dim=2,shift=+1) - Ex) / dy  &
-          - dt * (cshift(Ey,dim=1,shift=+1) - Ey) / dx
-
+  Bz = Bz &
+     + dt*(cshift(Ex,dim=2,shift=+1)-Ex)/dy &
+     - dt*(cshift(Ey,dim=1,shift=+1)-Ey)/dx
 \end{verbatim}
+}
 

### 2010-12-15 opencl/papers/shallow-water.tex
cleaning up after Craig....
cleaning up verbatim mess

@@
 
-The Fortran kernel procedure {\tt wave_advance} for the shallow water code
+The Fortran kernel procedure {\tt wave\_advance} for the shallow water code
 is declared as:
+
+{\small
 \begin{verbatim}
-   pure subroutine wave_advance(H, U, V, dx, dy, dt)
-      real, dimension(:,:), intent(inout) :: H, U, V
-      real, intent(in) :: dx, dy, dt
-      !$OFP CONTIGUOUS :: H, U, V
+pure subroutine wave_advance(H,U,V,dx,dy,dt)
+  real, dimension(:,:), intent(inout) :: H,U,V
+  real, intent(in) :: dx,dy,dt
+  !$OFP CONTIGUOUS :: H,U,V
   !$OFP KERNEL :: wave_advance
@@
 \end{verbatim}
+}
+
 where {\tt H, U}, and {\tt V} are start variables for height and
@@
 Vx}, and {\tt Vy} defined on cell faces.  These temporary arrays are declared as,
+
+{\small
 \begin{verbatim}
-   real, allocatable, dimension(:,:) :: iH, iU, iV
-   real, allocatable, dimension(:,:) :: Hx, Hy, Ux, Uy, Vx, Vy
+real, allocatable, dimension(:,:) :: iH,iU,iV
+real, allocatable, dimension(:,:) :: Hx,Hy,Ux
+real, allocatable, dimension(:,:) :: Uy,Vx,Vy
 \end{verbatim}
+}
 
 Halos variables for the interior and cell faces are defined as
+
+{\small
 \begin{verbatim}
-   integer, dimension(4) :: halo, face_lt, face_rt, face_up, face_dn
+integer, dimension(4) :: halo,face_lt,face_rt
+integer, dimension(4) :: face_up,face_dn
 halo    = [1,1,1,1]
@@
 \end{verbatim}
-Note that the halos for the four faces each have a 0 in the definition. Thus the returned
-array copy will have size that is larger than the interior regions that use {\tt [1,1,1,1]}.
-This is because there is one more cell face quantity than there are cells in a given direction.
+}
 
-The Lax Wendrof (???) update algorithm for the shallow-water model has two steps.
-The fluxes are first updated on the cell faces, for example,
+Note that the halos for the four faces each have a 0 in the
+definition. Thus the returned array copy will have size that is larger
+than the interior regions that use {\tt [1,1,1,1]}.  This is because
+there is one more cell face quantity than there are cells in a given
+direction.
+
+The Lax-Wendroff update algorithm for the shallow-water model has
+two steps.  The fluxes are first updated on the cell faces, for
+example,
+
+{\small
 \begin{verbatim}
-   Hx =           0.5 * ( region(H, face_lt) + region(H, face_rt) ) &
-        + (0.5*dt/dx) * ( region(U, face_lt) - region(U, face_rt) )
+Hx = 0.5*(region(H,face_lt)+ &
+          region(H,face_rt)) &
+   + (0.5*dt/dx) &
+   * (region(U,face_lt)-region(U,face_rt))
 \end{verbatim}
+}
+
 updates the $x$-direction flux for the height state variable.
 
-Then second step uses these fluxes to update the interior regions for the state varaibles.
-For example,
+Then second step uses these fluxes to update the interior regions for
+the state variables.  For example,
+
+{\small
 \begin{verbatim}
@@
 face_dn = [0,0,0,1];  face_up = [0,0,1,0]
-   iH = iH + (dt/dx) * ( region(Ux, face_lt) - region(Ux, face_rt) ) &
-           + (dt/dy) * ( region(Vy, face_dn) - region(Vy, face_up) )
+iH = iH + (dt/dx) * ( region(Ux, face_lt) - &
+                      region(Ux, face_rt) ) &
+        + (dt/dy) * ( region(Vy, face_dn) - &
+                      region(Vy, face_up) )
 \end{verbatim}
-Note that face halos have been redefined so that the array copy returned by
-now has the same size as the interior region.
+}
+
+Note that face halos have been redefined so that the array copy
+returned by now has the same size as the interior region.
 
@@
 copies are replaced by scalars in the transformed Fortran code so
-there is no performance penalty for using data-parallel statements
-as outlined.
+there is no performance penalty for using data-parallel statements as
+outlined.

### 2010-12-15 opencl/papers/source-to-source.tex
more verbatim

@@
 OpenCL,
+
+{\small
 \begin{verbatim}
-   iH = iH + (dt/dx) * ( region(Ux, face_lt) - region(Ux, face_rt) )
-           + (dt/dy) * ( region(Vy, face_dn) - region(Vy, face_up) );
+iH = iH + (dt/dx) * (region(Ux, face_lt) - 
+                     region(Ux, face_rt) )
+        + (dt/dy) * (region(Vy, face_dn) - 
+                     region(Vy, face_up) );
 \end{verbatim}
+}
 
 Region variables are more complicated because they are arrays.
+
+{\small
 \begin{verbatim}
-   Hx[l] = 0.5 * ( region(H_local, face_lt) + ... );
+Hx[l] = 0.5 * (region(H_local, face_lt)+ ...);
 \end{verbatim}
+}
+
 where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt

### 2010-12-15 opencl/papers/intro.tex
cleaning up verbatim mess
if someone doesn't know what FORTRAN stands for, they shouldn't be     reviewing papers.

@@
 
-Fortran (originally named FORTRAN, the FORmula TRANslation
-language\footnote{The capitalization convention for referring the
-  language was changed in 1990 to \emph{Fortran}}) is the oldest
+Fortran 
+%(originally named FORTRAN, the FORmula TRANslation
+%language\footnote{The capitalization convention for referring the
+%  language was changed in 1990 to \emph{Fortran}}) 
+is the oldest
 high-level programming language in continuous use since its
@@
 
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 

### 2010-12-15 opencl/papers/intro.tex
edits

@@
 Fortran procedures written in the Fortran subset described in this paper to
-OpenCL kernels.  We use the ROSE compiler infrastructure \cite{} to develop
-these transformations.  ROSE uses the Open Fortran Parser \cite{} to parse
+OpenCL kernels.  We use the ROSE compiler infrastructure \cite{rose-web} to develop
+these transformations.  ROSE uses the Open Fortran Parser \cite{ofp-web} to parse
 Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
@@
 to a code that was both slower and less flexible than the original Fortran
-\cite{}.  Similar experiences have occurred in the past, notably during the
+\cite{pooma-cite}.  Similar experiences have occurred in the past, notably during the
 devlopment of the Sisal language.

### 2010-12-15 opencl/papers/programming-model.tex
edits

@@
 The static analysis and source-to-source transformations used are very
-basic but require the programmer to use a subset of Fortran
-that employs a data-parallel programming model.  In particular, it
+basic and only require the programmer to use a subset of Fortran that
+employs a data-parallel programming model.  In particular, it
 encourages use of language features that were introduced and
 standardized in the Fortran 90 language specification.  In this
-section we describe the set of Fortran 90 features that our analysis
-and transformation method are based on.
+section we describe the set of four Fortran 90 features that our analysis
+and transformation method are based on.  From these language constructs,
+we are able to easily transform them to a lower-level CUDA or OpenCL
+implementation.
 
-This paper uses four simple features of Fortran that form a language subset
-that does not require any language extensions and can be easily transformed
-to a lower-level implementation in either CUDA or OpenCL.
+%This paper uses four simple features of Fortran that form a language subset
+%that does not require any language extensions and can be easily transformed
+%to a lower-level implementation in either CUDA or OpenCL.
 
@@
 Fortran 90 introduced a rich array syntax that allows programmers to
-write code that is in terms of whole arrays or subarrays, with data
-parallel operators to compute on the arrays.  This has the benefit of
-avoiding explicit looping in the code and maintaining a high-level
-style that is closer to the original mathematical specification of the
-problem being solved.  More importantly from a compilation
-perspective, this defers decisions about how to implement these
-whole-array operations to a compilation tool.  When faced with a novel
-architecture such as modern GPUs that are ideally suited to data parallel
-programming models, the fit between Fortran arrays and these systems is
-quite clean.
+write statements that are in terms of whole arrays or subarrays, with
+data parallel operators to compute on the arrays.  This has the
+benefit of avoiding explicit looping in the code and maintaining a
+high-level style that is closer to the original mathematical
+specification of the problem being solved.  More importantly from a
+compilation perspective, this defers decisions about how to implement
+these whole-array operations to a compilation tool.  Sophisticated
+loop analysis to identify parallelism within the loop are not
+necessary.  When faced with a novel architecture such as modern GPUs
+that are ideally suited to data parallel programming models, the fit
+between Fortran arrays and these systems is quite clean.
 
 %% NEW - CER
-Arrays are first class citizens in Fortran as they are part of the language
-type system rather than just a pointer to raw memory as in C. A Fortran array
-contains metadata associated with it that describes the array's shape and
-size. This metadata is lacking in other languages, such as C, C++, and Java.
-Even in the case where multidimensional array data structures are available
-(such as in C++ or Java), they exist outside the language standard, limiting
-the ability of the compiler to analyze them fully. The existence of this
-metadata allows the use of array notation whereby explicit indexing of arrays
-is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
-of the same rank and shape and {\tt s} is a scalar, then the statement
+Arrays are first class citizens in Fortran instead of a basic pointer
+to raw memory as in many other compiled languages.  A Fortran array
+contains metadata associated with it that describes the array's shape
+and size. It should be noted that other languages that provide rich
+array data types and whole array operations may also be suitable
+targets for the transformations described in this paper.  
+In addition to metadata about the structure of the arrays, operators
+are provided that allow programmers to write expressions based on the
+whole array instead of explicit indexing and application of operations
+to individual elements.  For example, if {\tt A}, {\tt B}, and {\tt C}
+are all arrays of the same rank and shape and {\tt s} is a scalar,
+then the statement
 
@@
 
-results in the sum of elements of {\tt A} and {\tt s} times the elements of
+results in the element-wise sum of {\tt A} and {\tt s} times the elements of
 {\tt B} being stored in the corresponding elements of {\tt C}. The first
@@
 %% NEW - CER
-An example of shown above using array syntax could be refactored to use
-the elemental function,
+For example, the basic array operation shown above could be refactored into
+an elemental function,
 
@@
 Note that while {\tt foo} is defined in terms of purely scalar
-quantities, it can be \emph{applied} to arrays as shown.
+quantities, it can be \emph{applied} to arrays as shown.  While this
+may seem like a trivial example, such simple functions may be composed
+with other elemental functions to perform powerful computations,
+especially when applied to arrays.  Our prototype tool transforms
+elemental functions to inline OpenCL functions.  Thus there is no
+penalty for usage of elemental functions and provide a convient
+mechanism to express algorithms in simpler segments.
 
-While this may seem like a trivial example, such simple functions may be
-composed with other elemental functions to perform powerful computations,
-especially when applied to arrays.  ForOpenCL transforms elemental functions to inline OpenCL
-functions.  Thus there is no penalty for usage of elemental functions and provide
-a convient mechanism to express algorithms in simpler segments.
-
-Also note that the plus and times operators shown in the data-statement
-example can be seen as an elemental functions, as plus and times are defined
-in terms of scalars but can be applied to whole arrays and can return an array
-result.
+%Also note that the plus and times operators shown in the data-statement
+%example can be seen as an elemental functions, as plus and times are defined
+%in terms of scalars but can be applied to whole arrays and can return an array
+%result.
 
@@
 
-Pure procedures, like elemental functions, must be free of side effects.
-(athough they may change the value of array elements in an array argument,
-unlike elemental arguments which must have the intent(in) attribute).
-The absence of side effects removes ordering constraints from the compiler
-allowing it to invoke pure functions out of order.  Procedures and functions
-of this sort are also common in pure functional languages like Haskell,
-and are exploited by compilers in order to emit parallel code automatically
-due to their suitability for compiler-level analysis.
+Pure procedures, like elemental functions, must be free of side
+effects.  Unlike elemental functions that require arguments to have an
+{\tt intent(in)} attribute, they may change the contents of array
+arguments that are passed to them.  The absence of side effects
+removes ordering constraints that could restrict the freedom of the
+compiler to invoke pure functions out of order and possibly in parallel.
+Procedures and functions of this sort are also common in pure
+functional languages like Haskell, and are exploited by compilers in
+order to emit parallel code automatically due to their suitability for
+compiler-level analysis.
 
-Since pure procedures don't have side effects they a candidates for running on
-accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
-OpenCL kernels that \emph{do not} call other procedures, except for elemental
-functions.
+Since pure procedures don't have side effects they a candidates for
+running on accelerators in OpenCL.  Currently our prototype tool
+transforms pure procedures to OpenCL kernels that \emph{do not} call
+other procedures, except for elemental functions.
 
@@
 
-Unlike OpenMP, we require limited use of compiler directives, although
+Unlike OpenCL, we require limited use of compiler directives, although
 two are used to enforce the semantics required by the programming
 model.  These directives are:
+
 \begin{itemize}
+
 \item {\tt \!\$OFP CONTIGUOUS}: specifies that an dummy array variable is
   contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.
+
 \item {\tt \!\$OFP KERNEL}: specifies that a pure subroutine can be
   transformed to an OpenCL kernel.
+
 \end{itemize}
@@
 a two-dimensional region.  These functions are:
+
 \begin{itemize}
+
 \item {\tt transfer\_halo(array, halo)}: an impure subroutine that
@@
   node-local portion of a virtual global array. Not used in this work.
+
 \item {\tt interior(array, halo)}: a pure function that returns a copy of
@@
   local to a node.
+
 \item {\tt region(array, halo)}: a pure function that returns a copy of
   the portion of the array specified by halo.  Array is local to a node.
+
 \end{itemize}
@@
 
-There are several advantages to this style of programming using array syntax, shifts, regions,
-and pure and elemental functions:
+There are several advantages to this style of programming using array
+syntax, shifts, regions, and pure and elemental functions:
 
 \begin{itemize}
-	\item There are no loops or index variables to keep track of.  Off by one
-index errors are a common programming mistake.
-        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
-	\item Semantically the intrinsic functions return arrays by value.  This is usually what the algorithm requires.
-	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
+\item There are no loops or index variables to keep track of.  Off by
+  one index errors and improper handling of array boundaries are a
+  common programming mistake.
+\item The written code is closer to the algorithm, easier to
+  understand, and is usually substantially shorter.
+\item Semantically the intrinsic functions return arrays by value.
+  This is usually what the algorithm requires.
+\item Because pure and elemental function are free from side effects,
+  it is easier for a compiler to schedule the work to be done in
+  parallel.
 \end{itemize}
+
+%% MJS - mention that these features were designed for parallelism in
+%% the first place?
 
@@
 %% NEW - CER
-This example is a solution to Maxwell's equations for the $z$ component of the
-magnetic field using Fortran array syntax.  Note that there are no explicit
-loops in this example.  The operators {\tt +}, {\tt -}, and {\tt *} are
-applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt
-  Ex}, and {\tt Ey}, individually.  This is why data parallelism has been
-called collection-oriented programming by
-Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift} function
-and the array-valued expressions all semantically returns a value, this style
-of programming is also similar to functional programming (or value-oriented
-programming). %%~\cite{simonpeytonjones}).  The heart of the solution to
-Maxwell's equations is the statement shown in Listing~\ref{lst:dpexample} and
-five similar, simple equations.
+
+This example is a solution to Maxwell's equations for the $z$
+component of the magnetic field using Fortran array syntax.  Note that
+there are no explicit loops in this example.  The operators {\tt +},
+{\tt -}, and {\tt *} are applied to all of the elements of the
+three-dimensional arrays {\tt Bz}, {\tt Ex}, and {\tt Ey},
+individually.  This is why data parallelism has been called
+collection-oriented programming by
+Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift}
+function and the array-valued expressions all semantically returns a
+value, this style of programming is also similar to functional
+programming (or value-oriented
+programming).
+
+%%~\cite{simonpeytonjones}).  The heart of the solution
+%to Maxwell's equations is the statement shown in
+%Listing~\ref{lst:dpexample} and five similar, simple equations.
 

### 2010-12-15 opencl/papers/programming-model.tex
LA-UR version

@@
 collection-oriented programming by
-Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift}
+Blelloch~\cite{blelloch90,rajopadhye93}.  As the {\tt cshift}
 function and the array-valued expressions all semantically returns a
@@
 programmers shied away from using array syntax (even now, some are actively
-counseling against its usage because of performance issues~\cite{Cray-SC07}).
+counseling against its usage because of performance issues~\cite{Levesque:SC08}).
 Thus the Fortran community was caught in a classic ``chicken-and-egg''
@@
 compilers vendors didn't improve it because programmers didn't use it.
-
-%% NEW - CER
 A goal of this paper is to demonstrate that parallel programs written in
-this style of Fortran are maintainable and can achieve good performance on a
-variety of processor architectures, including possibly, future terascale
-architectures from Intel (such as the XXX architecture).
+this style of Fortran are maintainable and can achieve good performance on
+accelerator architectures.
 

### 2010-12-15 opencl/papers/shallow-water.tex
LA-UR version

@@
 
-A short description of the shallow water model goes here.
+The numerical code used for this study is from a presentation at the
+NM Supercomputing Challenge~\cite{Robey07}.
+The algorithm solves the standard 2D Shallow Water equations. This
+algorithm is typical of a wide range of modeling equations based on
+conservation laws such as compressible fluid dynamics (CFD), elastic
+material waves, acoustics, electromagnetic waves and even traffic
+flow~\cite{Leveque02}. For the shallow water problem there are
+three equations with one based on conservation of mass and the other
+two on conservation of momentum.\begin{eqnarray*}
+h_{t}+(hu)_{x}+(hv)_{y} & = & 0\quad\mbox{(mass)}\\
+(hu)_{t}+(h{u}^{2}+\tfrac{1}{2}gh^{2})_{x}+(huv)_{y} & = & 0\mbox{\quad($x$-momentum)}\\
+(hv)_{t}+(huv)_{x}+(h{v}^{2}+\tfrac{1}{2}gh^{2})_{y} & = & 0\mbox{\quad($y$-momentum)}\end{eqnarray*}
+
+%h_{t}+(hu)_{x}+(hv)_{y} & = & 0\qquad\mbox{(conservation of mass)}\\
+%(hu)_{t}+(h{u}^{2}+\tfrac{1}{2}gh^{2})_{x}+(huv)_{y} & = & 0\mbox{\qquad(conservation of x-momentum)}\\
+%(hv)_{t}+(huv)_{x}+(h{v}^{2}+\tfrac{1}{2}gh^{2})_{y} & = & 0\mbox{\qquad(conservation of y-momentum)}\end{eqnarray*}
+
+where h = height of water column (mass), $u$ = x velocity, $v$ =
+y velocity, and $g$ = gravity. The height $h$ can be used for mass
+because of the simplification of a unit cell size and a uniform water
+density. Another simplifying assumption is that the water depth is
+small in comparison to length and width and so velocities in the z-direction
+can be ignored. A fixed time step is used for simplicity though it
+must be less than $dt\leqq(\sqrt{gh}+|u|)/dx$.
+
+The numerical method is a two-step Lax-Wendroff scheme. The method
+has some numerical oscillations with sharp gradients but is adequate
+for simulating smooth shallow-water flows. In the following explanation,
+$U$ is the conserved state variable at the center of the cell . This
+state variable, $U=(h,hu,hv)$ in the first term in the equations
+above. $F$ is the flux quantity that crosses the boundary of the cell
+and is subtracted from one cell and added to the other. The remaining
+terms after the first term are the flux terms in the equations above
+with one term for the flux in the x-direction and the next term for
+the flux in the y-direction. The first step estimates the values a
+half-step advanced in time and space on each face, using loops on
+the faces.\begin{eqnarray*}
+U_{i+\frac{1}{2},j}^{n+\frac{1}{2}} & = & (U_{i+1,j}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle x}\left(F_{i+1,j}^{n}-F_{i,j}^{n}\right)\\
+U_{i,j+\frac{1}{2}}^{n+\frac{1}{2}} & = & (U_{i,j+1}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle y}\left(F_{i,j+1}^{n}-F_{i,j}^{n}\right)\end{eqnarray*}
+
+The second step uses the estimated values from step 1 to compute the
+values at the next time step in a dimensionally unsplit loop.\[
+U_{i,j}^{n+1}=U_{i,j}^{n}-\frac{\triangle t}{\triangle x}(F_{i+\frac{1}{2},j}^{n+\frac{1}{2}}-F_{i-\frac{1}{2},j}^{n+\frac{1}{2}})-\frac{\triangle t}{\triangle y}(F_{i,j+\frac{1}{2}}^{n+\frac{1}{2}}-F_{i,j-\frac{1}{2}}^{n+\frac{1}{2}})\]
 
@@
 pure subroutine wave_advance(H,U,V,dx,dy,dt)
-  real, dimension(:,:), intent(inout) :: H,U,V
+  real, dimension(:,:),intent(inout):: H,U,V
   real, intent(in) :: dx,dy,dt
@@
 The Lax-Wendroff update algorithm for the shallow-water model has
-two steps.  The fluxes are first updated on the cell faces, for
+two steps.  The fluxes are first updated on cell faces, for
 example,
@@
 
+\noindent
 updates the $x$-direction flux for the height state variable.
@@
 
+\noindent
 Note that face halos have been redefined so that the array copy

### 2010-12-15 opencl/papers/performance.tex
Initial version

--- 2010-12-15 opencl/papers/source-to-source.tex
@@
-\section{Source-To-Source Transformations}
-
-This section describes the transformations that take Fortran elemental and
-pure procedures as input and generate OpenCL kernels.
-
-\subsection{OpenCL}
-
-OpenCL~\cite{opencl:standard} is an open language standard for developing
-applications for accelerators.  The C-based language provides extensions for
-programming kernels that run on accelerator processing elements.  The kernels
-are run by calling a C runtime library from the OpenCL host (normally the
-CPU).  Efforts to standardize a C++ runtime are underway and Fortran
-interfaces to the C runtime are described later.
-
-An important concept in OpenCL is that of a thread and a thread group.  Thread
-groups are used to run an OpenCL kernel concurrently on several processor
-elements on the OpenCL device (often a GPU).  Consider a data-parallel
-statement written in terms of an elemental fuction as discussed above.  The
-act of running an OpenCL kernel can be thought of as having a particular
-thread assigned to each instance of the call to the elemental function as it
-is mapped across the arrays in the data-parallel statement.  In practice,
-these threads are packaged into thread groups when they are run on the device
-hardware.
-
-Device memory is separated hierarchically.  A thread instance has access to
-its own memory, thread groups to OpenCL local memory, and all thread groups
-have access to OpenCL global memory.  When multiple members of a thread group
-access the same memory elements, for example if {\tt region} or {\tt shift}
-functions are called, for performance reasons it is often best if global
-memory accessed by a thread group is copied into local memory.
-
-The \emph{region} and \emph{halo} constructs easily map onto the OpenCL memory
-hierarchy.  A schematic of this mapping is shown in Figure~\ref{fig:cl-memory}
-for a two-dimensional array with a 2x2 array of 4 thread groups.  The memory
-for the array and its halo are stored in global memory on the device as shown
-in the background layer of the figure.  The array copy in local memory is
-shown in the foreground divided into 4 \emph{local} tiles that partition the
-array.  Halo regions in global memory are shown in dark gray and halo regions
-in local memory are shown in light gray.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=2.5in]{cl-memory.pdf}
-\caption{A schematic of global memory for an array and its copy stored in local memory
-for four thread groups.}
-\label{fig:cl-memory}
-\end{figure}
-
-We emphasize that the hierarchical distribution of memory used on the OpenCL
-device shown in Figure~\ref{fig:cl-memory} can be extended to include memory
-across MPI nodes as well.  In this case, the virtual global array is
-represented by the background layer (with its halo) and its partitions stored
-in the 4 MPI nodes shown in the foreground.
-
-Halo regions are constrained semantically so that they can not be written to
-by an OpenCL kernel because the \emph{region} and \emph{interior} functions
-return copies of the global memory.  Thus once memory for a halo region has
-been transferred into global device memory by all of the host nodes running
-MPI (before the OpenCL kernel is run), memory is in a consistent state so that
-the kernels are free to read from global device memory.  Because the local
-memory is a copy, it functions as a software cache for the local thread group.
-Thus the compiler must insert OpenCL barriers at proper locations in the code
-to insure that all threads have written to the local memory cache before any
-thread can read from the cache.  On exit from a kernel, the local memory cache
-is copied back to global memory for all \emph{interior} regions leaving global
-memory in a consistent state again.
-
-\subsection{ForOpenCL}
-
-This work describes transformations that automatically create OpenCL kernel
-functions from Fortran pure and elemental procedures.  These transformations
-will not transform an entire program.  Users, for now, must explicitly replace
-calls to Fortran kernel procedures that run on the device, with calls to the
-OpenCL runtime that will run the kernel on the OpenCL host.  While these host
-transformations are straightforward using ROSE, they are outside the scope of
-this paper.
-
-In addition to the Fortran to OpenCL transformations, the
-ForOpenCL~\cite{http:ofp} library provides programmers with the ability to
-call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
-providing Fortran 2003 interface descriptions and classes that allow language
-interoperability with the OpenCL runtime.
-
-\section{Transformation examples}
-
-This sections outlines the OpenCL equivalent syntax for portions of the
-Fortran shallow-water code described in the previous section.  The notation
-uses uppercase for arrays and lowercase for scalar quantities.  Interior and
-region array copies are denoted by an i or an r preceeding the array.  For
-example, {\tt iH = interior(H, halo)} is a local copy of interior region of
-array {\tt H} (representing height) in the shallow water code.
-
-\subsubsection{interior and region functions}
-
-While the serial versions of the interior and regions functions return an
-array copy, in OpenCL, these functions return a scalar quantity based on the
-location of a thread in a thread group and the relationship of its location to
-the array copy in local memory.  Because we assume there is a thread for every
-element in the interior, the array index is just the thread index adjusted for
-the halo.  Thus {\tt interior} and {\tt region} are just inline OpenCL
-functions provided by the ForOpenCL library.
-
-\subsubsection{function and variable declarations}
-
-Fortran kernel procedures have direct correspondence with OpenCL equivalents.
-For example, {\tt wave_advance} is transformed as {\tt __kernel void
-  wave_advance(__ global float * H, ..., float dt);}.  The global state
-variables have local equivalents, e.g., {\tt __local float
-  H_local[MAX_LOCAL_SIZE]}, declared with the appropriate size and are copied
-to local memory by the compiler with inlined library functions.  The flux
-array temporaries (region variables) are declared similarly but initialized to
-zero with inlined library functions.  Interior variables are simple scalars,
-e.g., {\tt float iH;}.  Region variables cannot be scalar objects because
-regions are shifted and thus \emph{shared} by threads within a thread group.
-
-\subsubsection{array syntax}
-
-Array syntax transforms nearly directly to OpenCL code.  For example, interior
-variables are particularly straightforward as they as scalar quantities in
-OpenCL,
-\begin{verbatim}
-   iH = iH + (dt/dx) * ( region(Ux, face_lt) - region(Ux, face_rt) )
-           + (dt/dy) * ( region(Vy, face_dn) - region(Vy, face_up) );
-\end{verbatim}
-
-Region variables are more complicated because they are arrays.
-\begin{verbatim}
-   Hx[l] = 0.5 * ( region(H_local, face_lt) + ... );
-\end{verbatim}
-where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
-  LX = get_local_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
-  get_local_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
-  get_local_size(0)} is the size of the thread group in the $x$ dimension, and
-the {\tt get_local_id} and {\tt get_local_size} functions are defined by the
-OpenCL compiler.
-
-\subsection{Performance measurements}
+\section{Performance Measurements}
 
 Performance measurements were made comparing the transformed code with
-different versions of the serial shallow-water code.  The serial versions
-include the original written in C and two Fortran versions: one using copy
-semantics for the regions and one using reference semantics.  We also compare
-with a hand written OpenCL implementation that was optimized for local memory
-usage (no array temporaries).  The measurments were made using an NVIDIA Tesla C2050
-GPU (TODO - need model numbers) running at XXX Hz and an Intel XXX hexacore CPU
-running at XXX Hz with 2.56 GB of RAM.  The compilers were gfortran and gcc VERSIONS with
-optimization of -O3.
+different versions of the serial shallow-water code.  The serial
+versions include the original written in C and two Fortran versions:
+one using copy semantics for the regions and one using reference
+semantics.  We also compare with a hand written OpenCL implementation
+that was optimized for local memory usage (no array temporaries).  The
+accelerated measurments were made using an NVIDIA Tesla C2050 (Fermi) cGPU with
+2.625 GB GDDR5 memory, and 448 processor cores.  The serial
+measurements were made using an Intel Xeon X5650 hexacore CPU with 96 GB of RAM
+running at 2.67 GHz.  The compilers were gfortran and gcc version 4.4.3
+with an optimization level of -O3.
+
+%%TODO All problems were run on a 1280x1280 mesh for 10000 cycles.
+
+%%TODO Timing results were done on a Tesla 1050 class system with an AMD processor (insert info from cat /etc/procinfo and with the DEBUG flag turned on in ezcl.c).
 

### 2010-12-16 opencl/f90/shallow_water.f90
Added usage of dy

@@ -5,16 +5,18 @@ program shallow_water
 
    integer :: status
 
+   integer, parameter :: G_WIDTH = 1280
+
    integer,           parameter :: NPAD = 1
-   integer(c_size_t), parameter :: NX  = 1280
-   integer(c_size_t), parameter :: NY  = 1280
+   integer(c_size_t), parameter :: NX  = G_WIDTH
+   integer(c_size_t), parameter :: NY  = G_WIDTH
    integer(c_size_t), parameter :: NXL = 16
    integer(c_size_t), parameter :: NYL = 16
 
    integer(c_size_t), parameter :: SIZE_ELEMENT = 4
 
    real(c_float), target, dimension(NX+2*NPAD,NY+2*NPAD) :: H, U, V
-   real(c_float) :: dx, dt
+   real(c_float) :: dx, dy, dt
 
    type(CLDevice) :: device
    type(CLKernel) :: kernel
@@ -47,7 +49,7 @@ program shallow_water
    ! initialize memory
    !
 
-   dx = 1.0;  dt = 0.01;
+   dx = 1.0;  dy = 1.0;  dt = 0.01;
 
    H = 1.0;  U = 0.0;  V = 0.0;
 
@@ -69,8 +71,10 @@ program shallow_water
    status = setKernelArgMem (kernel, 1, clMemObject(d_U))
    status = setKernelArgMem (kernel, 2, clMemObject(d_V))
    status = setKernelArgReal(kernel, 3, dx)
-   status = setKernelArgReal(kernel, 4, dt)
-   status = setKernelArgLoc (kernel, 5, 9*tile_mem_size)
+   status = setKernelArgReal(kernel, 4, dy)
+   status = setKernelArgReal(kernel, 5, dt)
+   status = setKernelArgLoc (kernel, 6, 9*tile_mem_size)
+!   status = setKernelArgLoc (kernel, 6, 1*tile_mem_size)
 
 print*, "warmup"
    ! warmup the kernel

### 2010-12-16 opencl/f90/shallow_water_f.f90
Moved declaration of interior variables above allocated ones.

@@ -97,11 +97,11 @@ subroutine wave_advance(H, U, V, dx, dt)
 
    integer, dimension(4) :: halo, face_lt, face_rt, face_up, face_dn
 
-   ! explicit temporaries
-   real, allocatable, dimension(:,:) :: Hx, Hy, Ux, Uy, Vx, Vy
-
    ! pointers for interior and shifted regions
    real, pointer, dimension(:,:) :: iH, iU, iV
+
+   ! explicit temporaries
+   real, allocatable, dimension(:,:) :: Hx, Hy, Ux, Uy, Vx, Vy
 
    ! scalar quantities
    real :: cs, gs, dtdx, dtdx2

### 2010-12-16 opencl/f90/test_shift.f90
Added h_time

@@ -57,7 +57,7 @@ program test_shift
    integer(c_size_t) :: global_ex_mem_size = (NX +2*NPAD)*(NY +2*NPAD) * SIZE_ELEMENT
    integer(c_size_t) :: local_ex_mem_size  = (NXL+2*NPAD)*(NYL+2*NPAD) * SIZE_ELEMENT
 
-   type(CPUTimer) :: timer
+    type(CPUTimer) :: timer
    real(c_double) :: h_time
 
    integer :: nxGlobal=NX, nyGlobal=NY

### 2010-12-16 opencl/c/opencl_utils.c
Added additional error conditions

@@ -103,6 +103,9 @@ void stop_on_error(int code)
    switch (code) {
       case CL_SUCCESS:
          return;
+      case CL_INVALID_ARG_INDEX:
+         sprintf(msg, "%s (%d)", "CL_INVALID_ARG_INDEX", code);
+         break;
       case CL_INVALID_ARG_SIZE:
          sprintf(msg, "%s (%d)", "CL_INVALID_ARG_SIZE", code);
          break;
@@ -142,11 +145,14 @@ void stop_on_error(int code)
       case CL_INVALID_QUEUE_PROPERTIES:
          sprintf(msg, "%s (%d)", "CL_INVALID_QUEUE_PROPERTIES", code);
          break;
+      case CL_INVALID_VALUE:
+         sprintf(msg, "%s (%d)", "CL_INVALID_VALUE", code);
+         break;
       case CL_INVALID_WORK_GROUP_SIZE:
          sprintf(msg, "%s (%d)", "CL_INVALID_WORK_GROUP_SIZE", code);
          break;
-      case CL_INVALID_VALUE:
-         sprintf(msg, "%s (%d)", "CL_INVALID_VALUE", code);
+      case CL_OUT_OF_RESOURCES:
+         sprintf(msg, "%s (%d)", "CL_OUT_OF_RESOURCES", code);
          break;
       default:
          sprintf(msg, "%s (%d)\n", "UNKNOWN_CODE", code);

### 2010-12-16 opencl/papers/intro.tex
LA-UR version

--- 2010-12-15 opencl/papers/intro.tex
@@
 This paper presents a compiler-level approach for targeting a single
-program to multiple fundamentally different low level execution and
+program to multiple fundamentally different low-level execution and
 programming models while allowing the application programmer to adopt
-a single high level programming model.  We show that features of
+a single high-level programming model.  We show that features of
 Fortran 90 for data parallel programming are well suited to automatic
@@
 operations, writing code in Fortran and transforming it automatically
-to specific low level implementations removes the burden from the
+to specific low-level implementations removes the burden from the
 programmer of working with tedious, error prone, low-level tools.
 
-In the ideal situation, application programmers would like to adopt
-a programming model in which they write their application once and
-use automated tools to retarget it to many architectures.  This has
-proven to be very challenging historically due to the subtle balance
-between high-level expressiveness of code and the performance of the
+In the ideal situation, application programmers would like to adopt a
+programming model in which they write their application once and use
+automated tools to retarget it to many architectures.  This has proven
+to be very challenging historically due to the subtle balance between
+high-level expressiveness of code and the performance of the
 lower-level code that is emitted by a compiler.  This ideal high-level
 model that programmers work with should emphasize readability,
-maintainability, and close proximity in abstraction to the problem being
-solved.  It should not be corrupted with details of specific target
-architectures solely for the purpose of single-system performance.
-For certain classes of applications, specifically those that map well
-onto a data-parallel programming model, we show that Fortran 90 contains
-language features that encourage high-level programming abstractions
-without sacrificing performance during low-level code generation.
+maintainability, and close proximity in abstraction to the problem
+being solved --- in this instance, the abstraction that we care about
+are mathematical formulae.  It should not be corrupted with details of
+specific target architectures solely for the purpose of single-system
+performance.  For certain classes of applications, specifically those
+that map well onto a data-parallel programming model, we show that
+Fortran 90 contains language features that encourage high-level
+programming abstractions without sacrificing performance during
+low-level code generation.
 
@@
 
-At Los Alamos National Laboratory (LANL), as with many supercomputing
+At the Los Alamos National Laboratory (LANL), as with many supercomputing
 facilities today, users have a wide variety of computer platforms from
@@
 are based on either vector processing or massively parallel
-collections of simple cores.  A popular trend for programming systems
-such as GPU accelerators is the CUDA language from NVidia, or new
-languages like Chapel from Cray.  A less intrusive approach that is
-well suited to legacy applications and languages are language
+collections of simple cores.  A proposed approach for programming that
+is well suited to legacy applications and languages are language
 extensions or libraries that allow programmers to avoid adopting
@@
 
-At this point in time it is not really possible to write once and run
-efficiently on the wide variety of computer platforms we have
-available.  For some classes of applications, we believe that this
-goal is possible using language constructs already present in a
-popular mainstream scientific programming language -- Fortran.  A
-common, long-standing tongue-in-cheek response to new language
-developments in the scientific and high performance computing
-community is that a new language will arise to answer the needs of new
-systems, and it will be called Fortran.  We believe that the work
-presented in this paper validates that notion -- we need a new
-language to work with, and that language is Fortran.
+%At this point in time it is not really possible to write once and run
+%efficiently on the wide variety of computer platforms we have
+%available.  For some classes of applications, we believe that this
+%goal is possible using language constructs already present in a
+%popular mainstream scientific programming language -- Fortran.  A
+%common, long-standing tongue-in-cheek response to new language
+%developments in the scientific and high performance computing
+%community is that a new language will arise to answer the needs of new
+%systems, and it will be called Fortran.  We believe that the work
+%presented in this paper validates that notion -- we need a new
+%language to work with, and that language is Fortran.
 
@@
 
-This paper addresses the accelerator programmign problem by examining
-features in Fortran that allow programmers to express algorithms at a
-very high level that can be easily transformed by a compiler to run
-efficiently on a wide variety of platforms.  In particular we consider
-computers based on GPUs and related accelerator processors.
+%This paper addresses the accelerator programmign problem by examining
+%features in Fortran that allow programmers to express algorithms at a
+%very high level that can be easily transformed by a compiler to run
+%efficiently on a wide variety of platforms.  In particular we consider
+%computers based on GPUs and related accelerator processors.
 
@@
 features include pure and elemental functions and array constructs
-like where and shift.  In addition we add a few functions that enable
-a program to be deployed on machines with a hierarchy of processing
-elements, such as nodes employing GPU acceleration, \emph{without
-  requiring explicit declaration of parallelism within the program.}
-In addition the program uses entirely standard Fortran so it can be on
-a single core without concurrency.  This work also is applicable
-to vendor-specific languages similar to OpenCL such as the NVidia
-CUDA language.
+like {\tt where} and {\tt cshift}.  In addition we add a few functions
+that enable a program to be deployed on machines with a hierarchy of
+processing elements, such as nodes employing GPU acceleration,
+\emph{without requiring explicit declaration of parallelism within the
+  program.}  In addition the program uses entirely standard Fortran so
+it can be compiled for and executed on a single core without concurrency.
+This work also is applicable to vendor-specific languages similar to
+OpenCL such as the NVidia CUDA language.
 
-We provide (via Fortran interfaces in the ForOpenCL library) a mechanism to
-call the C OpenCL runtime and enable Fortran programmers to access OpenCL
-kernels.  Transformations are supplied that provide a mechanism for converting
-Fortran procedures written in the Fortran subset described in this paper to
-OpenCL kernels.  We use the ROSE compiler infrastructure \cite{} to develop
-these transformations.  ROSE uses the Open Fortran Parser \cite{} to parse
-Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
-intermediate representation (IR) was constructed to represent multiple
-languages, it is relatively straightforward to transform high-level Fortran IR
-nodes to OpenCL nodes.  Furthermore, the Fortran array syntax maps directly to
-one, two, and three-dimensional thread groups in OpenCL.
+%We provide (via Fortran interfaces in the ForOpenCL library) a
+%mechanism to call the C OpenCL runtime and enable Fortran programmers
+%to access OpenCL kernels.  
+Transformations are supplied that provide a mechanism for converting
+Fortran procedures written in the Fortran subset described in this
+paper to OpenCL kernels.  We use the ROSE compiler infrastructure
+\cite{rose-web} to develop these transformations.  ROSE uses the Open
+Fortran Parser \cite{ofp:sf} to parse Fortran 2008 syntax and can
+generate C-based OpenCL.  Since ROSE's intermediate representation
+(IR) was constructed to represent multiple languages, it is relatively
+straightforward to transform high-level Fortran IR nodes to C OpenCL
+nodes.
 
-Transformations for general Fortran procedures are not provided.  Furthermore,
-a mechanism to transform the calling site to automatically invoke
-OpenCL kernels is not provided at this time.  While it is possible to
-accomplish this task within ROSE, it is considered outside the scope of this
-paper.
+%  Furthermore, the Fortran array syntax maps directly to one,
+%two, and three-dimensional thread groups in OpenCL.
 
-We examine the performance of the Fortran data-parallel abstraction when
-transformed to OpenCL to run on GPU architectures.  Since single node performance
-is often given as a reason for not using data-parallel constructs within Fortran,
-we consider the performance of serial data-parallel codes compared with the
-usage of explicit loop constructs.
+Transformations for general Fortran procedures are not provided.
+Furthermore, a mechanism to transform the calling site to
+automatically invoke OpenCL kernels is not provided at this time.
+While it is possible to accomplish this task within ROSE, it is
+considered outside the scope of this paper.
+
+We examine the performance of the Fortran data-parallel abstraction
+when transformed to OpenCL to run on GPU architectures.  Since single
+node performance is often given as a reason for not using
+data-parallel constructs within Fortran, we consider the performance
+of serial data-parallel codes compared with the usage of explicit loop
+constructs.
 
 We study automatic transformations and the performance for an
-application example that is typical of many applications that are based on
-finite-difference or finite-volume methods in computational fluid dynamics
-(CFD).  The example is a simple shallow water model in two dimensions using
-finite volume methods with .... for time update.
+application example that is typical of many applications that are
+based on finite-difference or finite-volume methods in computational
+fluid dynamics (CFD).  The example described later in this paper is a
+simple shallow water model in two dimensions using finite volume
+methods.
 
@@
 the Cell and the rest is gained from automatically overlapping
-inter-node communication with computation on the Cell.
+inter-node communication MPI with computation on the Cell.
 
@@
 
-Fortran 
-%(originally named FORTRAN, the FORmula TRANslation
-%language\footnote{The capitalization convention for referring the
-%  language was changed in 1990 to \emph{Fortran}}) 
-is the oldest
-high-level programming language in continuous use since its
-introduction, and was developed to facilitate the translation of math
-formulae into machine code. Fortran was the first major language to
-use a compiler to translate from a high-level program representation
-to assembly language. Due to its age, it carries certain arcane
-baggage that later programming languages have evolved away from. As a
-result, Fortran has fallen into disfavor in certain programming
-circles.  Modern versions of the language standardized in 1990, 1995, 2003, and 2008
-have removed much of this legacy baggage, but these changes are not
-widely known.  Modern Fortran exhibits features that are similar to 
-other modern programming languages, and does not mandate legacy features
-from decades old, deprecated versions of the language.
+Fortran is the oldest high-level programming language in continuous
+use since its introduction, and was developed to facilitate the
+translation of math formulae into machine code. Fortran was the first
+major language to use a compiler to translate from a high-level
+program representation to assembly language. Due to its age, it
+carries certain arcane baggage that later programming languages have
+evolved away from. As a result, Fortran has fallen into disfavor in
+certain programming circles.  Modern versions of the language
+standardized in 1990, 1995, 2003, and 2008 have removed much of this
+legacy baggage, but these changes are not widely known.  Modern
+Fortran exhibits features that are similar to other modern programming
+languages, and does not mandate the use of legacy features from
+decades old, deprecated versions of the language.  Compilers for
+Fortran (like those for other languages that have removed features
+over time) can prohibit the use of archaic features such as fixed
+format code or constructs that have been removed from the language.
 
@@
 form simply to support legacy codes. It is now modular and has many
-object-oriented features.  It now includes a type system in which rich
-first-class array data types and corresponding syntax are part of the
-language, something that languages like C continue to lack.
-Furthermore, Fortran should be of interest to those studying parallel
-programming because of its functional and data parallel constructs and
-because of the coarray notation introduced in Fortran 2008. Unlike
-languages like C and C++, which are considered to be more modern,
-Fortran has become a truly parallel language with features added to
-recent language standards.
+object-oriented features.  Most importantly for this work, it now
+includes a type system in which rich first-class array data types and
+corresponding syntax are part of the language, something that
+languages like C continue to lack.  Furthermore, Fortran should be of
+interest to those studying parallel programming because of its
+functional and data parallel constructs and because of the coarray
+notation introduced in Fortran 2008. Unlike languages like C and C++,
+which are considered to be more modern, Fortran has become a truly
+parallel language with features added to recent language standards.
 
-Yet, a likely question that one may pose is ``\emph{Why Fortran and not a more
-  modern language like X?}''  The recent rise in interest in concurrency and
-parallelism at the language level driven by multicore CPUs and manycore
-accelerators has driven a number of new language developments, both as novel
-languages and extensions on existing ones.
+Yet, a likely question that one may pose is ``\emph{Why Fortran and
+  not a more modern language like X?}''  The recent rise in interest
+in concurrency and parallelism at the language level driven by
+multicore CPUs and manycore accelerators has driven a number of new
+language developments, both as novel languages and extensions on
+existing ones.  For scientific users, new languages and language
+extensions to use novel new architectures present a challenge: how do
+developers effectively use them while avoiding rewriting their code
+and potentially growing dependent on a transient technology that will
+vanish tomorrow?
 
 %% NEW - CER
-So perhaps it is time to replace Fortran with yet another computer language.  The
-problems with replacing Fortran with an entirely new language are two fold:
-the economics of replacing the existing application base and the difficulty in
-obtaining programmer acceptance.  It is estimated that replacing a major
-production application at Los Alamos National Laboratory would cost between 50
-and 150 million dollars.  In terms of programmer acceptance, there is always
-the "chicken and egg problem": programmers won't use a new language until they
-can expect good performance across a variety of platforms, and compiler
-vendors can't afford to produce quality compilers until there is a reasonable
-expectation of a market.
-
-For scientific users, these new languages and language extensions
-present a challenge: how do developers effectively use them while
-avoiding rewriting their code and potentially growing dependent on a
-transient technology that will vanish tomorrow?
+%So perhaps it is time to replace Fortran with yet another computer language.  The
+%problems with replacing Fortran with an entirely new language are two fold:
+%the economics of replacing the existing application base and the difficulty in
+%obtaining programmer acceptance.  It is estimated that replacing a major
+%production application at Los Alamos National Laboratory would cost between 50
+%and 150 million dollars.  In terms of programmer acceptance, there is always
+%the "chicken and egg problem": programmers won't use a new language until they
+%can expect good performance across a variety of platforms, and compiler
+%vendors can't afford to produce quality compilers until there is a reasonable
+%expectation of a market.
 
@@
 to a code that was both slower and less flexible than the original Fortran
-\cite{}.  Similar experiences have occurred in the past, notably during the
+\cite{basili08hpc}.  Similar experiences have occurred in the past, notably during the
 devlopment of the Sisal language.
@@
 A number of previous efforts have exploited data parallel programming
-at the language level to utilize novel architectuers, particularly
-in previous decades during the reign of vector and massively
-parallel computers in the high performance computing world.  The
-origin of the array syntax that was adopted in Fortran 90 can be found
-in the APL language\cite{iverson79apl}.  Fortran 90 differed from
-previous extensions of Fortran in that parallelism within whole-array operations
-was expressed at the expression level instead of via parallelism within explicit
-DO-loops (such as within IVTRAN for the Illiac IV).
+at the language level to utilize novel architectuers, particularly in
+previous decades during the reign of vector and massively parallel
+computers in the high performance computing world.  The origin of the
+array syntax that was adopted in Fortran 90 can be found in the APL
+language\cite{iverson79apl}.  Fortran 90 differed from previous
+extensions of Fortran in that parallelism within whole-array
+operations was expressed at the expression level instead of via
+parallelism within explicit DO-loops (such as within IVTRAN for the
+Illiac IV).
 
@@
 HPF also introduced an alternative looping construct to the
-traditional DO-loop called \emph{FORALL} that was better suited for
-parallel compilation.  An additional keyword, \emph{INDEPENDENT}, was
+traditional DO-loop called {\tt FORALL} that was better suited for
+parallel compilation.  An additional keyword, {\tt INDEPENDENT}, was
 added to allow the programmer to indicate when the order of execution
@@
 
-When shifts are employed, this can be expressed as:
+\noindent When shifts are employed, this can be expressed as:
 
@@
 
-\begin{verbatim}
-Write some ZPL
-\end{verbatim}
-
-This would then be executed over the entire array. {\bf more...}
-
-Stuff on stencil compilers?  No?
+%%TODO
+%%\begin{verbatim}
+%%Write some ZPL
+%%\end{verbatim}
+%%
+%%This would then be executed over the entire array. {\bf more...}
+%%
+%%Stuff on stencil compilers?  No?

### 2010-12-16 opencl/papers/programming-model.tex

@@
 employs a data-parallel programming model.  In particular, it
-encourages use of language features that were introduced and
-standardized in the Fortran 90 language specification.  In this
-section we describe the set of four Fortran 90 features that our analysis
-and transformation method are based on.  From these language constructs,
-we are able to easily transform them to a lower-level CUDA or OpenCL
-implementation.
+encourages use of four language features introduced in Fortran 90 ---
+array notation, shift functions, elemental functions, and pure
+procedures.  From these language constructs, we are able to easily
+transform them to a lower-level CUDA or OpenCL implementation.
 
@@
 these whole-array operations to a compilation tool.  Sophisticated
-loop analysis to identify parallelism within the loop are not
-necessary.  When faced with a novel architecture such as modern GPUs
-that are ideally suited to data parallel programming models, the fit
-between Fortran arrays and these systems is quite clean.
+loop analysis to identify parallelism within the sequential loop code
+are not necessary.  When faced with a novel architecture such as
+modern GPUs that are ideally suited to data parallel programming
+models, the fit between Fortran arrays and these systems is quite
+clean.
 
-%% NEW - CER
-Arrays are first class citizens in Fortran instead of a basic pointer
-to raw memory as in many other compiled languages.  A Fortran array
-contains metadata associated with it that describes the array's shape
-and size. It should be noted that other languages that provide rich
-array data types and whole array operations may also be suitable
-targets for the transformations described in this paper.  
-In addition to metadata about the structure of the arrays, operators
-are provided that allow programmers to write expressions based on the
-whole array instead of explicit indexing and application of operations
-to individual elements.  For example, if {\tt A}, {\tt B}, and {\tt C}
-are all arrays of the same rank and shape and {\tt s} is a scalar,
-then the statement
+A Fortran array combines a base pointer to memory with metadata that
+describes the array shape, stride and size. It should be noted that
+other languages that provide rich array data types and whole array
+operations may also be suitable targets for the transformations
+described in this paper.  Array variables can be used in expressions
+based on whole-array operations.  For example, if {\tt A}, {\tt B},
+and {\tt C} are all arrays of the same rank and shape and {\tt s} is a
+scalar, then the statement
 
@@
 
-results in the element-wise sum of {\tt A} and {\tt s} times the elements of
-{\tt B} being stored in the corresponding elements of {\tt C}. The first
-element of {\tt C} will contain the value of the first element of {\tt A}
-added to the first element of {\tt c*B}.  Note that no explicit iteration over
-array indices is needed and that the individual operators, plus, times, and
-assignment are applied by the compiler to individual elements of the arrays
-independently.  Thus the compiler is able to spread the computation in the
-example across any hardware threads under its control.
+\noindent results in the element-wise sum of {\tt A} and {\tt s} times
+the elements of {\tt B} being stored in the corresponding elements of
+{\tt C}. The first element of {\tt C} will contain the value of the
+first element of {\tt A} added to the first element of {\tt c*B}.
+Note that no explicit iteration over array indices is needed and that
+the individual operators, plus, times, and assignment are applied by
+the compiler to individual elements of the arrays independently.  Thus
+the compiler is able to spread the computation in the example across
+any hardware threads under its control.
 
@@
 technique to the compiler.  Elemental functions are intended to be used for
-data parallel programming, and as a result must be side effect free (or,
-\emph{pure}).
+data parallel programming, and as a result must be side effect free and
+mandate an {\tt intent(in)} attribute for all arguments.
 
@@
 elemental functions to inline OpenCL functions.  Thus there is no
-penalty for usage of elemental functions and provide a convient
+penalty for usage of elemental functions and they provide a convient
 mechanism to express algorithms in simpler segments.
@@
 %% NEW - CER
-Complete and very concise and elegant programs can be built with procedures
-similar to the example shown above. To aid this effort, Fortran
-supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
-\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
-functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
-functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
-\texttt{PRODUCT}, ...).  To this set we add region functions described above.
+%Complete and very concise and elegant programs can be built with procedures
+%similar to the example shown above. To aid this effort, Fortran
+%supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
+%\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
+%functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
+%functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
+%\texttt{PRODUCT}, ...).  To this set we add region functions described above.
 

### 2010-12-16 opencl/papers/source-to-source.tex

@@
 
-OpenCL~\cite{opencl:standard} is an open language standard for developing
+OpenCL~\cite{opencl08} is an open language standard for developing
 applications for accelerators.  The C-based language provides extensions for
@@
 This work describes transformations that automatically create OpenCL kernel
-functions from Fortran pure and elemental procedures.  These transformations
+functions from Fortran pure and elemental procedures.  At present, these transformations
 will not transform an entire program.  Users, for now, must explicitly replace
@@
 In addition to the Fortran to OpenCL transformations, the
-ForOpenCL~\cite{http:ofp} library provides programmers with the ability to
+ForOpenCL~\cite{ofp:sf} library provides programmers with the ability to
 call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
@@
 OpenCL compiler.
-
-\subsection{Performance measurements}
-
-Performance measurements were made comparing the transformed code with
-different versions of the serial shallow-water code.  The serial versions
-include the original written in C and two Fortran versions: one using copy
-semantics for the regions and one using reference semantics.  We also compare
-with a hand written OpenCL implementation that was optimized for local memory
-usage (no array temporaries).  The measurments were made using an NVIDIA Tesla C2050
-GPU (TODO - need model numbers) running at XXX Hz and an Intel XXX hexacore CPU
-running at XXX Hz with 2.56 GB of RAM.  The compilers were gfortran and gcc VERSIONS with
-optimization of -O3.
-
-The performance comparison is shown in Figure~\ref{fig:cl-performance} for
-varying array sizes (of the state variables).  The time represents an average
-of 100 iterations of the outer time-advance loop calling the OpenCL kernel.
-This tight loop keeps the OpenCL kernel supplied with threads to take
-advantage of potential latency hiding by the NVIDIA GPU.  Any serial code
-within this loop would reduce the measured values.
-
-As can be seen in Figure~\ref{fig:cl-performance}, the transformed get very
-good results of roughly 65 times speedup over the best Fortran code and YYY
-speedup over the original C code.  It comes with XXX \% of the hand coded
-OpenCL version.
-
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{cl-performance.pdf}
-%\caption{Performance comparison for varying array size.}
-%\label{fig:cl-performance}
-%\end{figure}

### 2010-12-19 opencl/papers/intro.tex
Nearly at page limit.

@@
 This paper presents a compiler-level approach for targeting a single
-program to multiple fundamentally different low-level execution and
-programming models while allowing the application programmer to adopt
-a single high-level programming model.  We show that features of
-Fortran 90 for data parallel programming are well suited to automatic
+program to multiple, fundamentally different low-level execution
+models.  This technique allows the application programmer to adopt
+a single high-level programming model without sacrificing performance.
+We show that features of
+Fortran 90 for data-parallel programming are well suited to automatic
 transformation to generate code specifically tuned for different
-low-level programming models such as OpenCL and CUDA.  For algorithms
-that can be easily expressed in terms of whole array, data parallel
-operations, writing code in Fortran and transforming it automatically
-to specific low-level implementations removes the burden from the
-programmer of working with tedious, error prone, low-level tools.
+hardware architectures using low-level programming models such as
+OpenCL and CUDA.  For algorithms that can be easily expressed in terms
+of whole array, data-parallel operations, writing code in Fortran and
+transforming it automatically to specific low-level implementations
+removes the burden from the programmer of working with tedious, error
+prone, low-level tools.
 
@@
 performance.  For certain classes of applications, specifically those
-that map well onto a data-parallel programming model, we show that
+that map onto a data-parallel programming model, we show that
 Fortran 90 contains language features that encourage high-level
@@
 However, the performance that these accelerator architectures offer
-comes at a cost as processor architectures are trending toward
-multiple cores with instances of integrated accelerator units with
-user managed memory and less of a reliance on superscalar instruction
+comes at a cost, as processor architectures are trending toward
+multiple cores with instances of integrated accelerator units (with
+user managed memory) and less of a reliance on superscalar instruction
 level parallelism and hardware managed memory hierarchies (such as
@@
 are based on either vector processing or massively parallel
-collections of simple cores.  A proposed approach for programming that
-is well suited to legacy applications and languages are language
-extensions or libraries that allow programmers to avoid adopting
-entirely new languages.  The recent OpenCL specification is intended
-to support this programming model, as are directive-based methods such
-as OpenMP or the Accelerator programming model from the Portland
-Group~\cite{pgi10accelerator}.
+collections of simple cores.  The recent CUDA and OpenCL programming
+languages are are intended to support this programming model, as are
+directive-based methods such as OpenMP or the Accelerator programming
+model from the Portland Group~\cite{pgi10accelerator}.
+
+%% Deleted by CER
+%%A proposed approach for programming that is well suited to legacy applications and languages are language extensions or libraries that allow programmers to avoid adopting entirely new languages.
 
@@
 their code for different classes of machine.  In both cases, the
-programmer is also responsible for explicitly managing memory,
+programmer is responsible for explicitly managing memory,
 including staging of data back and forth from the host CPU and the
@@
 
-Transformations for general Fortran procedures are not provided.
+Transformations for general Fortran procedures are not attempted.
 Furthermore, a mechanism to transform the calling site to
@@
 source-to-source compiler can automatically vectorize and parallelize
-(both within the Cell processor and across nodes) a small section of
-this code.  Preliminary results showed a 12 times performance gain on
-the Cell processor over a traditional single core processor.  A factor
-of 9 is gained from vectorizing and parallelizing the code to run on
-the Cell and the rest is gained from automatically overlapping
-inter-node communication MPI with computation on the Cell.
-
-%%We study automatic transformations of Fortran for two applications examples.
-The first is a simple shallow water model in two dimensions using finite volume methods with ....
-for time update.
-The second is based on the primary compute kernels from PetaVision, a high-performance, neural science framework that is used in models of visual cortex.
-
-%% To simplify things, lets save PetaVision for the local stuff as it requires convolutions
+a small section of this code for the Cell processor.  Preliminary results
+showed a 9 times performance gain of the transformed code when compared
+with the original serial version on a traditional single-core processor.
 
@@
 program representation to assembly language. Due to its age, it
-carries certain arcane baggage that later programming languages have
-evolved away from. As a result, Fortran has fallen into disfavor in
-certain programming circles.  Modern versions of the language
-standardized in 1990, 1995, 2003, and 2008 have removed much of this
-legacy baggage, but these changes are not widely known.  Modern
-Fortran exhibits features that are similar to other modern programming
-languages, and does not mandate the use of legacy features from
-decades old, deprecated versions of the language.  Compilers for
-Fortran (like those for other languages that have removed features
-over time) can prohibit the use of archaic features such as fixed
-format code or constructs that have been removed from the language.
-
-With the introduction of Fortran 90 (and later revisions to the
-standard), Fortran became a truly modern programming language. Much of
-the arcane baggage associated with it is actually no longer part of
-the language used by modern programmers, and exists in a deprecated
-form simply to support legacy codes. It is now modular and has many
+carries certain arcane baggage.  However
+with the introduction of Fortran 90 (and later revisions to the
+standard), Fortran became a truly modern programming language. 
+It is now modular and has many
 object-oriented features.  Most importantly for this work, it now
@@
 notation introduced in Fortran 2008. Unlike languages like C and C++,
-which are considered to be more modern, Fortran has become a truly
+Fortran has become a truly
 parallel language with features added to recent language standards.
+
+%% - deleted by CER
+%% that later programming languages have evolved away from.
+As a result, Fortran has fallen into disfavor in certain programming circles.
+Modern versions of the language standardized in 1990, 1995, 2003, and 2008 have removed much of this legacy baggage, but these changes are not widely known.
+Modern Fortran exhibits features that are similar to other modern programming languages, and does not mandate the use of legacy features from decades old, deprecated versions of the language.
+Compilers for Fortran (like those for other languages that have removed features over time) can prohibit the use of archaic features such as fixed format code or constructs that have been removed from the language.
 
@@
 extensions to use novel new architectures present a challenge: how do
-developers effectively use them while avoiding rewriting their code
+developers effectively use them while avoiding rewriting code
 and potentially growing dependent on a transient technology that will
@@
 A number of previous efforts have exploited data parallel programming
-at the language level to utilize novel architectuers, particularly in
+at the language level to utilize novel architectures, particularly in
 previous decades during the reign of vector and massively parallel

### 2010-12-19 opencl/papers/programming-model.tex
Nearly at page limit.

@@
 The static analysis and source-to-source transformations used are very
-basic and only require the programmer to use a subset of Fortran that
+basic and only require the programmer to use a language subset that
 employs a data-parallel programming model.  In particular, it
@@
 Fortran 90 introduced a rich array syntax that allows programmers to
-write statements that are in terms of whole arrays or subarrays, with
-data parallel operators to compute on the arrays.  This has the
-benefit of avoiding explicit looping in the code and maintaining a
-high-level style that is closer to the original mathematical
-specification of the problem being solved.  More importantly from a
-compilation perspective, this defers decisions about how to implement
-these whole-array operations to a compilation tool.  Sophisticated
-loop analysis to identify parallelism within the sequential loop code
-are not necessary.  When faced with a novel architecture such as
-modern GPUs that are ideally suited to data parallel programming
-models, the fit between Fortran arrays and these systems is quite
-clean.
-
-A Fortran array combines a base pointer to memory with metadata that
-describes the array shape, stride and size. It should be noted that
-other languages that provide rich array data types and whole array
-operations may also be suitable targets for the transformations
-described in this paper.  Array variables can be used in expressions
+write statements in terms of whole arrays or subarrays, with
+data parallel operators to compute on the arrays.
+Array variables can be used in expressions
 based on whole-array operations.  For example, if {\tt A}, {\tt B},
@@
 scalar, then the statement
+
+%% Deleted by CER
+%%This has the benefit of avoiding explicit looping in the code and maintaining a high-level style that is closer to the original mathematical specification of the problem being solved.
+More importantly from a compilation perspective, this defers decisions about how to implement these whole-array operations to a compilation tool.
+Sophisticated loop analysis to identify parallelism within the sequential loop code are not necessary.
+When faced with a novel architecture such as modern GPUs that are ideally suited to data parallel programming models, the fit between Fortran arrays and these systems is quite clean.
+
+%% Deleted by CER
+%% A Fortran array combines a base pointer to memory with metadata that describes the array shape, stride and size.
+It should be noted that other languages that provide rich array data types and whole array operations may also be suitable targets for the transformations described in this paper.
 
@@
 on each element of the array using the value of that element and some small
-set of neighboring cells.  Often programmers implement these operations that
-are local to each element within the inner-loop of a set of nested FOR-
-or DO-loops using offsets relative to the current array index.  An alternative
-to this local-view of the algorithm is to take a global view and write the
-algorithm in terms of the whole array.  For example, consider a 1D array in
-which we wish to subtract the $(i-1)$th element from the $i$th for all
-elements.  One way to look at this is that we are subtracting the entire array
-shifted by one element from itself.
-
+set of neighboring cells. As described above, this can often be done with
+shift operations.
 Fortran provides a set of shifting operators that allow programmers to
@@
 
+%% Deleted by CER
+%% Often programmers implement these operations that are local to each element within the inner-loop of a set of nested FOR- or DO-loops using offsets relative to the current array index.
+An alternative to this local-view of the algorithm is to take a global view and write the algorithm in terms of the whole array.
+For example, consider a 1D array in which we wish to subtract the $(i-1)$th element from the $i$th for all elements.
+One way to look at this is that we are subtracting the entire array shifted by one element from itself.
+
 \subsubsection*{Regions}
 
-Borrowing from ZPL, we introduce the concept of regions to Fortran.  In
-Fortran, regions are expressed as pointers to a subsection of an existing array.
-Regions may refer to an interior portion of an array or the entire array.
-
-The use of regions implies the use of the halo (or ghost) cell pattern 
+Borrowing ideas from ZPL, we introduce region functions to Fortran.
+In Fortran, these functions return a copy of an
+existing array or array section.  This is unlike ZPL where regions
+are first-class citizens and have special syntax.
+These functions imploy a halo (or ghost) cell pattern,
 where the size of an array is increased to provide extra array elements
@@
 
-Regions are similar to the shift operator as they can be used to reference
+Region functions are similar to the shift operator as they can be used to reference
 portions of the array that are shifted with respect to the interior portion.
@@
 
-Unlike OpenCL, we require limited use of compiler directives, although
+Unlike OpenMP, we require limited use of compiler directives, although
 two are used to enforce the semantics required by the programming
@@
 
-\subsection{Parallelism}
-
-% TODO - Matt - do you want a try at this.  Not sure we have room for
-% this subsection. I briefly mention node parallelism in other sections.
-
-% need to describe processor hierarchy of nodes and accelerators
-% need to describe memory hierarchy as well in regards to global array view,
-% node array view (with a copy is GPU global memory) and local GPU memory which
-% further subdivides the node array memory
-
-% figures on slicing and dicing of memory region go here
+\subsection{Advantages}
 
@@
 
-%% NEW - CER
-An example of this style of programming in Fortran is shown in
+%% Deleted by CER
+%% An example of this style of programming in Fortran is shown in
+%
+%{\small
+%\begin{verbatim}
+%  Bz = Bz &
+%     + dt*(cshift(Ex,dim=2,shift=+1)-Ex)/dy &
+%     - dt*(cshift(Ey,dim=1,shift=+1)-Ey)/dx
+%\end{verbatim}
+%}
 
-{\small
-\begin{verbatim}
-  Bz = Bz &
-     + dt*(cshift(Ex,dim=2,shift=+1)-Ex)/dy &
-     - dt*(cshift(Ey,dim=1,shift=+1)-Ey)/dx
-\end{verbatim}
-}
+%%This example is a solution to Maxwell's equations for the $z$ component of the magnetic field using Fortran array syntax.
+Note that there are no explicit loops in this example.
+The operators {\tt +}, {\tt -}, and {\tt *} are applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt Ex}, and {\tt Ey}, individually.
 
-%% NEW - CER
-
-This example is a solution to Maxwell's equations for the $z$
-component of the magnetic field using Fortran array syntax.  Note that
-there are no explicit loops in this example.  The operators {\tt +},
-{\tt -}, and {\tt *} are applied to all of the elements of the
-three-dimensional arrays {\tt Bz}, {\tt Ex}, and {\tt Ey},
-individually.  This is why data parallelism has been called
+Data parallelism has been called
 collection-oriented programming by
@@
 value, this style of programming is also similar to functional
-programming (or value-oriented
-programming).
+programming (or value-oriented programming).
 
@@
 
-%% NEW - CER
-This style of programming meets the requirements we have set for a programming
-model for developing applications suitable for acceleration.  It allows the programmer to
-program at a very-high level of abstraction while providing the compiler with
-maximum flexibility in targeting the application for a particular hardware
-architecture.  The data parallel programming model simultaneously meets the
-seemingly conflicting goals of maintainability, portability, and performance.
+%% Deleted by CER
+%% This style of programming meets the requirements we have set for a programming model for developing applications suitable for acceleration.
+It allows the programmer to program at a very-high level of abstraction while providing the compiler with maximum flexibility in targeting the application for a particular hardware architecture.
+The data parallel programming model simultaneously meets the seemingly conflicting goals of maintainability, portability, and performance.
 
-%% NEW - CER
 Unfortunately, this style of programming has never really caught on because
@@
 
-\subsection{Limitations}
+\subsection{Current Limitations}
 
 Only Fortran procedures are transformed into OpenCL kernels.  The programmer
-must currently explicitly call these kernels from Fortran using the ForOpenCL
-library described below.  It is also possible using ROSE to modify the calling
-site so that the entire program can be transformed but this functionality is
+must explicitly call these kernels from Fortran using the ForOpenCL
+library described below.  It is also possible, using ROSE, to modify the calling
+site so that the entire program can be transformed, but this functionality is
 outside the scope of this paper.  Here we specifically examine transforming
-Fortran procedures to OpenCL kernels.
+Fortran procedures to OpenCL kernels.  Because OpenCL support is relatively
+new to ROSE, some generated code must be modified.  For example, the
+{\tt \_\_global} attribute for kernel arguments was added by hand.
+
+It is assumed that memory for all arrays reside on the device.  The programmer must
+copy memory to and from the device.  In addition, array size (neglecting ghost cell
+regions) must be multiples of the global OpenCL kernel size.
 
@@
 {\tt interior} and {\tt region} functions.
-
-Array sizes must be multiples of the local kernel size, {\tt
-  get\_local\_size(0)*get\_local\_size(1)}.  This may be relaxed in
-the future.

### 2010-12-19 opencl/papers/shallow-water.tex
fixed spelling error
Nearly at page limit.

@@
 
+\noindent
 where h = height of water column (mass), $u$ = x velocity, $v$ =
@@
 can be ignored. A fixed time step is used for simplicity though it
-must be less than $dt\leqq(\sqrt{gh}+|u|)/dx$.
+must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$.
 
@@
 
-The Fortran kernel procedure {\tt wave\_advance} for the shallow water code
+The coding style to implement the serial version of the shallow water model
+is now shown.  The Fortran kernel procedure {\tt wave\_advance} 
 is declared as:
@@
 
-where {\tt H, U}, and {\tt V} are start variables for height and
+\noindent
+where {\tt H, U}, and {\tt V} are state variables for height and
 $x$ and $y$ momentum respectively.  The \emph{OFP} compiler directives
@@
 Temporary arrays are required for the interior copies {\tt iH, iU}, and
-{\tt iV} of the state variables and for the flux quantities {\tt Hx, Hy, Ux,
+{\tt iV} of the state variables and for the quantities {\tt Hx, Hy, Ux,
 Vx}, and {\tt Vy} defined on cell faces.  These temporary arrays are declared as,
@@
 
-Halos variables for the interior and cell faces are defined as
+Halo variables for the interior and cell faces are
 
@@
 
+\noindent
 Note that the halos for the four faces each have a 0 in the
@@
 
-The Lax-Wendroff update algorithm for the shallow-water model has
-two steps.  The fluxes are first updated on cell faces, for
-example,
+The first Lax-Wendroff step updates state variables on the cell
+faces, for example,
 
@@
 \noindent
-updates the $x$-direction flux for the height state variable.
-
-Then second step uses these fluxes to update the interior regions for
-the state variables.  For example,
+updates the $x$-direction height variable.  The second step then uses
+these face variables to update the interior region, for example,
 
@@
 Note that face halos have been redefined so that the array copy
-returned by now has the same size as the interior region.
+returned has the same size as the interior region.
 
@@
 programming model described above.  The resulting code is simple,
-consise, and easy to understand.  However it does \emph{not}
+concise, and easy to understand.  However it does \emph{not}
 necessarily perform well because of the temporary array variables,

### 2010-12-19 opencl/papers/source-to-source.tex
Nearly at page limit.

@@
 have access to OpenCL global memory.  When multiple members of a thread group
-access the same memory elements, for example if {\tt region} or {\tt shift}
-functions are called, for performance reasons it is often best if global
+access the same memory elements (if {\tt region} or {\tt shift}
+functions are called), for performance reasons it is often best if global
 memory accessed by a thread group is copied into local memory.
@@
 This sections outlines the OpenCL equivalent syntax for portions of the
-Fortran shallow-water code described in the previous section.  The notation
-uses uppercase for arrays and lowercase for scalar quantities.  Interior and
-region array copies are denoted by an i or an r preceeding the array.  For
+Fortran shallow-water code described previously.  The notation
+uses uppercase for arrays and lowercase for scalar quantities.  Interior
+variables are denoted by an i preceeding the array name.  For
 example, {\tt iH = interior(H, halo)} is a local copy of interior region of
-array {\tt H} (representing height) in the shallow water code.
+array {\tt H} (representing height) in the shallow water model.
 
@@
 variables have local equivalents, e.g., {\tt \_\_local float
-  H\_local[MAX\_LOCAL\_SIZE]}, declared with the appropriate size and are copied
-to local memory by the compiler with inlined library functions.  The flux
-array temporaries (region variables) are declared similarly but initialized to
-zero with inlined library functions.  Interior variables are simple scalars,
-e.g., {\tt float iH;}.  Region variables cannot be scalar objects because
-regions are shifted and thus \emph{shared} by threads within a thread group.
+  H\_local[MAX\_LOCAL\_SIZE]}, declared with the appropriate size and are
+copied to local memory by the compiler with inlined library functions.  The
+array temporaries defined on cell faces are declared similarly while interior
+variables are simple scalars, e.g., {\tt float iH;}.  Region variables cannot
+be scalar objects because regions may be shifted and thus \emph{shared} by
+threads within a thread group.
 
@@
 
+\noindent
 Region variables are more complicated because they are arrays.
@@
 
+\noindent
 where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt

### 2010-12-19 opencl/papers/performance.tex
Nearly at page limit.

@@
 different versions of the serial shallow-water code.  The serial
-versions include the original written in C and two Fortran versions:
-one using copy semantics for the regions and one using reference
-semantics.  We also compare with a hand written OpenCL implementation
+versions include two separate Fortran versions:
+one using data-parallel notation and the other using explicit looping
+constructs.
+We also compare with a hand-written OpenCL implementation
 that was optimized for local memory usage (no array temporaries).  The
@@
 
-As can be seen in Figure~\ref{fig:cl-performance}, the transformed get very
-good results of roughly 65 times speedup over the best Fortran code and YYY
-speedup over the original C code.  It comes with XXX \% of the hand coded
-OpenCL version.
+As can be seen in Figure~\ref{fig:cl-performance}, the transformed code achieves very
+good results.  It is up to 40 times faster than the best serial Fortran code.
 

### 2010-12-19 opencl/papers/foropencl.tex
if someone doesn't know what FORTRAN stands for, they shouldn't be     reviewing papers.
edits

@@
 \usepackage[cmex10]{amsmath}
+\usepackage{amssymb}
 \usepackage{array}
@@
 
-\title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
+%% Changed by CER
+%%\title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
+
+\title{Exploiting Array Syntax in Fortran for Accelerator Programming\footnote*{Foo!}}
 
@@
 \IEEEauthorblockA{Galois, Inc.\\
-421 SW 6th Ave. Suite 300 \\
-Portland, OR 97204\\
+%421 SW 6th Ave. Suite 300 \\
+%Portland, OR 97204\\
 Email: matt@galois.com}
 \and
-\IEEEauthorblockN{Craig E Rasmussen}
+\IEEEauthorblockN{Craig E Rasmussen,\\
+                  Wayne N. Weseloh,\\
+                  Robert W. Robey}
 \IEEEauthorblockA{Los Alamos National Laboratory\\
-CCS-7, MS B287\\
-Los Alamos, NM\\
+%CCS-7, MS B287\\
+%Los Alamos, NM\\
 Email: rasmussn@lanl.gov}
+
+\and
+\IEEEauthorblockN{Daniel Quinlan}
+\IEEEauthorblockA{Lawrence Livermore\\ National Laboratory}
+
+\and
+\IEEEauthorblockN{Jeffrey Overbey}
+\IEEEauthorblockA{Indiana University}
+
 }
@@
 \begin{abstract}
-Emerging architectures for high performance computing often are well
-suited to a data parallel programming model.  This paper presents a
-simple programming methodology based on existing languages and compiler
-tools that allows programmers to take advantage of these systems.
-We will work with the array features of Fortran 90 to show how this
+Emerging architectures for high performance computing are well suited
+to a data-parallel programming model.  This paper presents a simple
+programming methodology that allows programmers to take advantage of
+these systems.  We use array constructs in Fortran 90 to show how this
 infrequently exploited, standardized language feature is easily
-transformed to lower level accelerator code.  Our transformations are
-based on a mapping from Fortran 90 to C++ code with OpenCL extensions.
+transformed to lower-level accelerator code.  These transformations
+are based on a simple mapping from Fortran 90 to OpenCL.  We
+demonstrate, using a shallow-water code, that an acceleration up to 40
+times can be achieved on GPU hardware with these compiler-generated
+transformations.
 \end{abstract}
@@
 \input{source-to-source}
-\input{static-analysis}
-
-\section{Performance}
+%\input{static-analysis}
+\input{performance}
 
@@
 
-%% Explain what we have accomplished (copied from proposal)
-
-%% NEW - CER
 The sheer complexity of programming for clusters of many or multi-core
@@
 
-%% NEW - CER
 The goal of this work has been to exploit source-to-source transformations that
@@
 
-% Summarize results
-
-\section{Junk}
-
-\cite{chamberlain04zpl, roth97stencils}
+%%TODO \cite{chamberlain04zpl, roth97stencils}
 

### 2010-12-19 opencl/papers/intro.tex
edits

@@
 
-At the Los Alamos National Laboratory (LANL), as with many supercomputing
-facilities today, users have a wide variety of computer platforms from
-which to choose.  The most common platform is made up of clusters of
-compute nodes with standard multi-core processors.  An increasingly
-common feature is that some nodes also have accelerators that range
-from the IBM Cell processor (such as the LANL Roadrunner system) to a
-variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
-vector instructions and others do not.  The peak performance of these
-accelerated nodes often resides in the hundreds of gigaflops.
+%At the Los Alamos National Laboratory (LANL), as with many supercomputing
+%facilities today, users have a wide variety of computer platforms from
+%which to choose.  The most common platform is made up of clusters of
+%compute nodes with standard multi-core processors.  An increasingly
+%common feature is that some nodes also have accelerators that range
+%from the IBM Cell processor (such as the LANL Roadrunner system) to a
+%variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
+%vector instructions and others do not.  The peak performance of these
+%accelerated nodes often resides in the hundreds of gigaflops.
 
@@
 
-However, the performance that these accelerator architectures offer
-comes at a cost, as processor architectures are trending toward
-multiple cores with instances of integrated accelerator units (with
-user managed memory) and less of a reliance on superscalar instruction
-level parallelism and hardware managed memory hierarchies (such as
+The performance that new accelerator architectures offer comes at a
+cost, as processor architectures are trending toward multiple cores
+with instances of integrated accelerator units (with user managed
+memory) and less of a reliance on superscalar instruction level
+parallelism and hardware managed memory hierarchies (such as
 traditional caches).  These changes place a heavy burden on
@@
 computer architectures with unique memory systems and programming
-models.  The fundamental question we address is what programming model
-and language constructs are best suited to span this set of new
-hardware designs.
+models.  Can a programmer easily write one program that can run on a
+conventional multicore CPU, graphics processing unit, Cell processor,
+and one of many emerging many core architectures?  The fundamental
+question we address is what programming model and language constructs
+are best suited to span this set of new hardware designs.
 
@@
 
-Transformations for general Fortran procedures are not attempted.
+Transformations for arbitrary Fortran procedures are not attempted.
 Furthermore, a mechanism to transform the calling site to
@@
 program representation to assembly language. Due to its age, it
-carries certain arcane baggage.  However
-with the introduction of Fortran 90 (and later revisions to the
-standard), Fortran became a truly modern programming language. 
-It is now modular and has many
+carries certain arcane baggage.  However with the introduction of
+Fortran 90 (and later revisions to the standard), Fortran became a
+truly modern programming language.  It is now modular and has many
 object-oriented features.  Most importantly for this work, it now
@@
 corresponding syntax are part of the language, something that
-languages like C continue to lack.  Furthermore, Fortran should be of
-interest to those studying parallel programming because of its
-functional and data parallel constructs and because of the coarray
-notation introduced in Fortran 2008. Unlike languages like C and C++,
-Fortran has become a truly
-parallel language with features added to recent language standards.
+languages like C continue to lack\footnote{The recent Intel 12.0 C/C++
+  compiler supports extensions that provide array notation for C/C++
+  code, as detailed here: http://intel.ly/g454zp}.  Furthermore,
+Fortran should be of interest to those studying parallel programming
+because of its functional and data parallel constructs and because of
+the coarray notation introduced in Fortran 2008. Unlike languages like
+C and C++, Fortran has become a truly parallel language with features
+added to recent language standards.
 
@@
 %% NEW - CER
-History has also shown that an investment in rewriting code does not guarantee
-success either, as seen in an effort at LANL to modernize a legacy Fortran
-code with the the newer C++ POOMA framework.  This massive overhaul effort led
-to a code that was both slower and less flexible than the original Fortran
-\cite{basili08hpc}.  Similar experiences have occurred in the past, notably during the
-devlopment of the Sisal language.
+History has also shown that an investment in rewriting code does not
+guarantee success either, as seen in an effort at LANL to modernize a
+legacy Fortran code with the the newer C++ POOMA framework.  This
+massive overhaul effort led to a code that was both slower and less
+flexible than the original Fortran \cite{basili08hpc}.  Similar
+experiences have occurred in the past, notably during the devlopment
+of the functional SISAL language in the early 1990s.
 

### 2010-12-19 opencl/papers/programming-model.tex
edits

@@
 
-\subsubsection*{Regions}
-
-Borrowing ideas from ZPL, we introduce region functions to Fortran.
-In Fortran, these functions return a copy of an
-existing array or array section.  This is unlike ZPL where regions
-are first-class citizens and have special syntax.
-These functions imploy a halo (or ghost) cell pattern,
-where the size of an array is increased to provide extra array elements
-surrounding the interior portion of the array.
-%parlab.eecs.berkeley.edu/wiki/_media/patterns/ghostcell.pdf
-
-Region functions are similar to the shift operator as they can be used to reference
-portions of the array that are shifted with respect to the interior portion.
-However, unlike the shift operator, regions are not expressed in terms of
-boundary conditions and thus don't explicitly \emph{require} a knowledge of
-nor the application of boundary conditions locally.  Thus, as will be shown
-below, regions are more suitable for usage by OpenCL thread groups which
-access only local subsections of an array stored in global memory.
-
 \subsection{New Procedures}
 
-Unlike OpenMP, we require limited use of compiler directives, although
+Unlike OpenCL, we require limited use of compiler directives, although
 two are used to enforce the semantics required by the programming
@@
 \end{itemize}
+
+Borrowing ideas from ZPL, we introduce a concept of a region to
+Fortran with a set of functions that allow programmers to work with
+subarrays in expressions.  In Fortran, these functions return a copy
+of an existing array or array section.  This is unlike ZPL where
+regions are analogous to index sets and are used primarily for address
+resolution within an array without dictating storage related behavior.
+The functions that we propose are similar in that they allow a
+programmer to deal with index regions that are meaningful to their
+algorithm, and automatically induce a halo (or ghost) cell pattern as
+needed in the implementation generated by the compiler, where the size
+of an array is increased to provide extra array elements surrounding
+the interior portion of the array.  This is critical to reducing the
+amount of code that the programmer is forced to write, as the halo cells
+are often where boundary errors and off-by-one errors may occur in code
+that is manually generated.
+
+Region functions are similar to the shift operator as they can be used
+to reference portions of the array that are shifted with respect to
+the interior portion.  However, unlike the shift operator, regions are
+not expressed in terms of boundary conditions and thus don't
+explicitly \emph{require} a knowledge of nor the application of
+boundary conditions locally.  Thus, as will be shown below, regions
+are more suitable for usage by OpenCL thread groups which access only
+local subsections of an array stored in global memory.
 
@@
   This is usually what the algorithm requires.
-\item Because pure and elemental function are free from side effects,
-  it is easier for a compiler to schedule the work to be done in
-  parallel.
+\item Pure and elemental function are free from side effects, so it is
+  easier for a compiler to schedule the work to be done in parallel.
 \end{itemize}
@@
 
-Unfortunately, this style of programming has never really caught on because
-when Fortran 90 was first introduced, performance was relatively poor and thus
-programmers shied away from using array syntax (even now, some are actively
-counseling against its usage because of performance issues~\cite{Levesque:SC08}).
-Thus the Fortran community was caught in a classic ``chicken-and-egg''
+Unfortunately, this style of programming has never really caught on
+because when Fortran 90 was first introduced, performance of these
+features was relatively poor and thus programmers shied away from
+using array syntax (even recently, some are actively counseling against its
+usage because of performance issues~\cite{Levesque:SC08}).  Thus the
+Fortran community was caught in a classic ``chicken-and-egg''
 conundrum: (1) programmers didn't use it because it was slow; and (2)
 compilers vendors didn't improve it because programmers didn't use it.
-A goal of this paper is to demonstrate that parallel programs written in
-this style of Fortran are maintainable and can achieve good performance on
-accelerator architectures.
+A goal of this paper is to demonstrate that parallel programs written
+in this style of Fortran are maintainable and can achieve good
+performance on accelerator architectures.
 
@@
 
-Only Fortran procedures are transformed into OpenCL kernels.  The programmer
-must explicitly call these kernels from Fortran using the ForOpenCL
-library described below.  It is also possible, using ROSE, to modify the calling
-site so that the entire program can be transformed, but this functionality is
-outside the scope of this paper.  Here we specifically examine transforming
-Fortran procedures to OpenCL kernels.  Because OpenCL support is relatively
-new to ROSE, some generated code must be modified.  For example, the
-{\tt \_\_global} attribute for kernel arguments was added by hand.
+Only pure Fortran procedures are transformed into OpenCL kernels.  The
+programmer must explicitly call these kernels from Fortran using the
+ForOpenCL library described below.  It is also possible, using ROSE,
+to modify the calling site so that the entire program can be
+transformed, but this functionality is outside the scope of this
+paper.  Here we specifically examine transforming Fortran procedures
+to OpenCL kernels.  Because OpenCL support is relatively new to ROSE,
+some generated code must be modified.  For example, the {\tt
+  \_\_global} attribute for kernel arguments was added by hand.
 
-It is assumed that memory for all arrays reside on the device.  The programmer must
-copy memory to and from the device.  In addition, array size (neglecting ghost cell
-regions) must be multiples of the global OpenCL kernel size.
+It is assumed that memory for all arrays reside on the device.  The
+programmer must copy memory to and from the device.  In addition,
+array size (neglecting ghost cell regions) must be multiples of the
+global OpenCL kernel size.
 
-Only elemental functions may be called from kernel functions.  These include
-fortran functions that have an OpenCL analog and user-defined elemental functions.
+Only elemental functions may be called from kernel functions.  These
+include Fortran functions that have an OpenCL analog and user-defined
+elemental functions.
 
@@
 functions (primarily math), user-defined elemental function, and the
-{\tt interior} and {\tt region} functions.
+{\tt interior} and {\tt region} functions.  We plan to address
+non-contiguous arrays (such as those that result from strided access)
+by mapping array strides to gather/scatter-style memory accessors.

### 2010-12-19 opencl/papers/shallow-water.tex
edits

@@
 \section{Shallow Water Model}
+\label{sec:shallow-water}
 
-The numerical code used for this study is from a presentation at the
+The numerical code used for this work is from a presentation at the
 NM Supercomputing Challenge~\cite{Robey07}.
-The algorithm solves the standard 2D Shallow Water equations. This
+The algorithm solves the standard 2D shallow water equations. This
 algorithm is typical of a wide range of modeling equations based on
@@
 three equations with one based on conservation of mass and the other
-two on conservation of momentum.\begin{eqnarray*}
+two on conservation of momentum.
+
+\begin{eqnarray*}
 h_{t}+(hu)_{x}+(hv)_{y} & = & 0\quad\mbox{(mass)}\\
 (hu)_{t}+(h{u}^{2}+\tfrac{1}{2}gh^{2})_{x}+(huv)_{y} & = & 0\mbox{\quad($x$-momentum)}\\
-(hv)_{t}+(huv)_{x}+(h{v}^{2}+\tfrac{1}{2}gh^{2})_{y} & = & 0\mbox{\quad($y$-momentum)}\end{eqnarray*}
+(hv)_{t}+(huv)_{x}+(h{v}^{2}+\tfrac{1}{2}gh^{2})_{y} & = & 0\mbox{\quad($y$-momentum)}
+\end{eqnarray*}
 
@@
 can be ignored. A fixed time step is used for simplicity though it
-must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$.
+must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$ to fulfil the CFL
+condition.
 

### 2010-12-19 opencl/papers/source-to-source.tex
edits

@@
 
-This sections outlines the OpenCL equivalent syntax for portions of the
-Fortran shallow-water code described previously.  The notation
-uses uppercase for arrays and lowercase for scalar quantities.  Interior
-variables are denoted by an i preceeding the array name.  For
-example, {\tt iH = interior(H, halo)} is a local copy of interior region of
-array {\tt H} (representing height) in the shallow water model.
+This sections outlines the OpenCL equivalent syntax for portions of
+the Fortran shallow-water code described in
+Section~\ref{sec:shallow-water}.  The notation uses uppercase for
+arrays and lowercase for scalar quantities.  Interior variables are
+denoted by an {\tt i} preceeding the array name.  For example, {\tt iH =
+  interior(H, halo)} is a local copy of interior region of array {\tt
+  H} (representing height) in the shallow water model.
 

### 2010-12-19 opencl/papers/intro.tex
spelling

@@
 
-%This paper addresses the accelerator programmign problem by examining
+%This paper addresses the accelerator programming problem by examining
 %features in Fortran that allow programmers to express algorithms at a
@@
 flexible than the original Fortran \cite{basili08hpc}.  Similar
-experiences have occurred in the past, notably during the devlopment
+experiences have occurred in the past, notably during the development
 of the functional SISAL language in the early 1990s.
@@
 different codes within the computational physics community, especially
-at Los Alamos for codes targetting the CM-5 system that resided there until the
+at Los Alamos for codes targeting the CM-5 system that resided there until the
 late 1990s.
@@
 with HPF was that associated with the ZPL language work at the University
-of Washington~\ref{chamberlain04zpl}.  In ZPL, programmers adopt a simimlar
+of Washington~\ref{chamberlain04zpl}.  In ZPL, programmers adopt a similar
 global view of computation over arrays, but define their computations based

### 2010-12-19 opencl/papers/programming-model.tex
Spelling

@@
 elemental functions to inline OpenCL functions.  Thus there is no
-penalty for usage of elemental functions and they provide a convient
+penalty for usage of elemental functions and they provide a convenient
 mechanism to express algorithms in simpler segments.
@@
 pure and all array variables must be declared as contiguous.  A kernel
-procedure may not call other procedures except for limited intrisic
+procedure may not call other procedures except for limited intrinsic
 functions (primarily math), user-defined elemental function, and the

### 2010-12-19 opencl/papers/source-to-source.tex
Spelling

@@
 elements on the OpenCL device (often a GPU).  Consider a data-parallel
-statement written in terms of an elemental fuction as discussed above.  The
+statement written in terms of an elemental function as discussed above.  The
 act of running an OpenCL kernel can be thought of as having a particular
@@
 arrays and lowercase for scalar quantities.  Interior variables are
-denoted by an {\tt i} preceeding the array name.  For example, {\tt iH =
+denoted by an {\tt i} preceding the array name.  For example, {\tt iH =
   interior(H, halo)} is a local copy of interior region of array {\tt

### 2010-12-20 opencl/papers/foropencl.tex
oops
Nearly final draft.

@@
 
-%% Changed by CER
-%%\title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
-
-\title{Exploiting Array Syntax in Fortran for Accelerator Programming\footnote*{Foo!}}
+\title{Exploiting Array Syntax in Fortran for Accelerator Programming}
 
@@
 %Los Alamos, NM\\
-Email: rasmussn@lanl.gov}
+Email: crasmussen@lanl.gov}
 
@@
 programming methodology that allows programmers to take advantage of
-these systems.  We use array constructs in Fortran 90 to show how this
+these emerging systems.  We use array constructs in Fortran 90 to show how this
 infrequently exploited, standardized language feature is easily
@@
 are based on a simple mapping from Fortran 90 to OpenCL.  We
-demonstrate, using a shallow-water code, that an acceleration up to 40
+demonstrate, using a shallow-water code, that an acceleration of up to 40
 times can be achieved on GPU hardware with these compiler-generated
@@
 processors with tens of millions threads of execution make the simplicity of
-the data parallel model attractive.  Furthermore, the increasing complexity of
+the data-parallel model attractive.  Furthermore, the increasing complexity of
 todays applications (especially when convolved with the increasing complexity
 of the hardware) and the need for portability across hardware architectures
-make a higher-level and simpler programming model like data parallel
+make a higher-level and simpler programming model like data-parallel
 attractive.

### 2010-12-20 opencl/papers/intro.tex
Nearly final draft.

@@
 
-A common theme amongst new processors is the emphasis on data parallel
+A common theme amongst new processors is the emphasis on data-parallel
 programming.  This model is well suited to emerging architectures that
@@
 particularly true of CUDA and OpenCL.  In CUDA, programmers must adapt
-their codes to fit the threading model used by NVidia GPUs, while
+their codes to fit the threading model used by NVIDIA GPUs, while
 OpenCL requires programmers to provide specially tuned versions of
@@
 This work also is applicable to vendor-specific languages similar to
-OpenCL such as the NVidia CUDA language.
+OpenCL such as the NVIDIA CUDA language.
 
@@
 Fortran should be of interest to those studying parallel programming
-because of its functional and data parallel constructs and because of
+because of its functional and data-parallel constructs and because of
 the coarray notation introduced in Fortran 2008. Unlike languages like
@@
 guarantee success either, as seen in an effort at LANL to modernize a
-legacy Fortran code with the the newer C++ POOMA framework.  This
+legacy Fortran code with the newer C++ POOMA framework.  This
 massive overhaul effort led to a code that was both slower and less
@@
 
-A number of previous efforts have exploited data parallel programming
+A number of previous efforts have exploited data-parallel programming
 at the language level to utilize novel architectures, particularly in
@@
 loop-based parallelism.  This was likely in order to support parallel
-programming that wasn't easily mapped onto a pure data parallel model.
+programming that wasn't easily mapped onto a pure data-parallel model.
 
-In some instances though, a purely data parallel model is appropriate
+In some instances though, a purely data-parallel model is appropriate
 for part or all of the major computations within a program.  One of
@@
 were combined by shifting the entire arrays instead of working based
-on local offsets from singe indices.  A simple 1D example is one in
+on local offsets from single indices.  A simple 1D example is one in
 which an element is replaced with the average of its own value with
@@
 with HPF was that associated with the ZPL language work at the University
-of Washington~\ref{chamberlain04zpl}.  In ZPL, programmers adopt a similar
+of Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar
 global view of computation over arrays, but define their computations based
@@
 of relative offsets from that index:
-
-%%TODO
-%%\begin{verbatim}
-%%Write some ZPL
-%%\end{verbatim}
-%%
-%%This would then be executed over the entire array. {\bf more...}
-%%
-%%Stuff on stencil compilers?  No?

### 2010-12-20 opencl/papers/programming-model.tex
Nearly final draft.

@@
 
-Unlike OpenCL, we require limited use of compiler directives, although
+Unlike OpenMP, we require \emph{limited} use of compiler directives, although
 two are used to enforce the semantics required by the programming
@@
 subarrays in expressions.  In Fortran, these functions return a copy
-of an existing array or array section.  This is unlike ZPL where
+of an existing array or array section.  This is unlike ZPL, where
 regions are analogous to index sets and are used primarily for address
@@
 
-Data parallelism has been called
-collection-oriented programming by
-Blelloch~\cite{blelloch90,rajopadhye93}.  As the {\tt cshift}
-function and the array-valued expressions all semantically returns a
+Blelloch~\cite{blelloch90,rajopadhye93}.
+
+Data parallelism has been called collection-oriented programming by
+Blelloch~\cite{blelloch90}.  As the {\tt cshift}
+function and the array-valued expressions all semantically return a
 value, this style of programming is also similar to functional
@@
 
-Only elemental functions may be called from kernel functions.  These
-include Fortran functions that have an OpenCL analog and user-defined
-elemental functions.
+%% Duplicated CER
+%%Only elemental functions may be called from kernel functions.  These include Fortran functions that have an OpenCL analog and user-defined elemental functions.
 
@@
 procedure may not call other procedures except for limited intrinsic
-functions (primarily math), user-defined elemental function, and the
+functions (primarily math), user-defined elemental functions, and the
 {\tt interior} and {\tt region} functions.  We plan to address

### 2010-12-20 opencl/papers/source-to-source.tex
Nearly final draft.

@@
 
-This sections outlines the OpenCL equivalent syntax for portions of
+This section outlines the OpenCL equivalent syntax for portions of
 the Fortran shallow-water code described in
@@
 the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
-OpenCL compiler.
+OpenCL language standard.

### 2010-12-20 opencl/papers/intro.tex
edits

@@
 array syntax that was adopted in Fortran 90 can be found in the APL
-language\cite{iverson79apl}.  Fortran 90 differed from previous
+language.  Fortran 90 differed from previous % \cite{iverson79apl}
 extensions of Fortran in that parallelism within whole-array

### 2010-12-20 opencl/papers/foropencl.tex
edits

@@
 %Portland, OR 97204\\
-Email: matt@galois.com}
+matt@galois.com}
 \and
@@
 %Los Alamos, NM\\
-Email: crasmussen@lanl.gov}
+\{crasmussen, weseloh, brobey\}@lanl.gov}
 
@@
 \IEEEauthorblockN{Daniel Quinlan}
-\IEEEauthorblockA{Lawrence Livermore\\ National Laboratory}
+\IEEEauthorblockA{Lawrence Livermore\\ National Laboratory\\
+dquinlan@llnl.gov}
 
@@
 \IEEEauthorblockN{Jeffrey Overbey}
-\IEEEauthorblockA{Indiana University}
+\IEEEauthorblockA{Indiana University\\
+overbey2@illinois.edu}
 
@@
 processors with tens of millions threads of execution make the simplicity of
-the data-parallel model attractive.  Furthermore, the increasing complexity of
-todays applications (especially when convolved with the increasing complexity
-of the hardware) and the need for portability across hardware architectures
+the data-parallel model attractive.  The increasing complexity of
+todays applications (especially in light of the increasing complexity
+of the hardware) and the need for portability across architectures
 make a higher-level and simpler programming model like data-parallel

### 2010-12-20 opencl/papers/performance.tex
Spelling
Added new text regarding hand-optimized kernel.

@@
 that was optimized for local memory usage (no array temporaries).  The
-accelerated measurments were made using an NVIDIA Tesla C2050 (Fermi) cGPU with
+accelerated measurements were made using an NVIDIA Tesla C2050 (Fermi) cGPU with
 2.625 GB GDDR5 memory, and 448 processor cores.  The serial
@@
 As can be seen in Figure~\ref{fig:cl-performance}, the transformed code achieves very
-good results.  It is up to 40 times faster than the best serial Fortran code.
+good results.  It is up to 40 times faster than the best serial Fortran code and
+within 25\% of the hand-optimized OpenCL kernel.
 
@@
 \centering
-\includegraphics[width=2.5in]{cl-performance.pdf}
+\includegraphics[width=3in]{cl-performance.pdf}
 \caption{Performance comparison for varying array size.}

### 2010-12-20 opencl/papers/intro.tex
updates
tweaks to shrink to 8 pages

@@
 Fortran procedures written in the Fortran subset described in this
-paper to OpenCL kernels.  We use the ROSE compiler infrastructure
-\cite{rose-web} to develop these transformations.  ROSE uses the Open
-Fortran Parser \cite{ofp:sf} to parse Fortran 2008 syntax and can
-generate C-based OpenCL.  Since ROSE's intermediate representation
-(IR) was constructed to represent multiple languages, it is relatively
-straightforward to transform high-level Fortran IR nodes to C OpenCL
-nodes.
+paper to OpenCL kernels.  We use the ROSE compiler
+infrastructure\footnote{\url{http://www.rosecompiler.org/}} to
+develop these transformations.  ROSE uses the Open Fortran
+Parser\footnote{\url{http://fortran-parser.sf.net/}} to parse
+Fortran 2008 syntax and can generate C-based OpenCL.  Since ROSE's
+intermediate representation (IR) was constructed to represent multiple
+languages, it is relatively straightforward to transform high-level
+Fortran IR nodes to C OpenCL nodes.
 
@@
 massive overhaul effort led to a code that was both slower and less
-flexible than the original Fortran \cite{basili08hpc}.  Similar
-experiences have occurred in the past, notably during the development
-of the functional SISAL language in the early 1990s.
+flexible than the original Fortran \cite{basili08hpc}.
+
+%Similar
+%experiences have occurred in the past, notably during the development
+%of the functional SISAL language in the early 1990s.
 
@@
 at Los Alamos for codes targeting the CM-5 system that resided there until the
-late 1990s.
+late 1990s.  A body of research in compilation of stencil-based codes
+that use shift operators targeting these systems is related to the
+work we present here~\cite{stencil-compiler}.
 

### 2010-12-20 opencl/papers/source-to-source.tex
edits
tweaks to shrink to 8 pages

@@
 OpenCL~\cite{opencl08} is an open language standard for developing
-applications for accelerators.  The C-based language provides extensions for
-programming kernels that run on accelerator processing elements.  The kernels
+applications for accelerators.
+The kernels
 are run by calling a C runtime library from the OpenCL host (normally the
@@
 \centering
-\includegraphics[width=2.5in]{cl-memory.pdf}
+\includegraphics[width=1.75in]{cl-memory.pdf}
 \caption{A schematic of global memory for an array and its copy stored in local memory
@@
 
-\subsection{ForOpenCL}
+%\subsection{ForOpenCL}
 
-This work describes transformations that automatically create OpenCL kernel
-functions from Fortran pure and elemental procedures.  At present, these transformations
-will not transform an entire program.  Users, for now, must explicitly replace
-calls to Fortran kernel procedures that run on the device, with calls to the
-OpenCL runtime that will run the kernel on the OpenCL host.  While these host
-transformations are straightforward using ROSE, they are outside the scope of
-this paper.
+%This work describes transformations that automatically create OpenCL kernel
+%functions from Fortran pure and elemental procedures.  At present, these transformations
+%will not transform an entire program.  Users, for now, must explicitly replace
+%calls to Fortran kernel procedures that run on the device, with calls to the
+%OpenCL runtime that will run the kernel on the OpenCL host.  While these host
+%transformations are straightforward using ROSE, they are outside the scope of
+%this paper.
 
-In addition to the Fortran to OpenCL transformations, the
-ForOpenCL~\cite{ofp:sf} library provides programmers with the ability to
-call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
-providing Fortran 2003 interface descriptions and classes that allow language
-interoperability with the OpenCL runtime.
+%In addition to the Fortran to OpenCL transformations, the
+%ForOpenCL library available from the Open Fortran Parser web page provides programmers with the ability to
+%call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
+%providing Fortran 2003 interface descriptions and classes that allow language
+%interoperability with the OpenCL runtime.
 
@@
 
-\subsubsection{interior and region functions}
+\subsubsection{Interior and region functions}
 
@@
 
-\subsubsection{function and variable declarations}
+\subsubsection{Function and variable declarations}
 
@@
 
-\subsubsection{array syntax}
+\subsubsection{Array syntax}
 
 Array syntax transforms nearly directly to OpenCL code.  For example, interior
-variables are particularly straightforward as they as scalar quantities in
+variables are particularly straightforward as they are scalar quantities in
 OpenCL,
@@
 
-\noindent
-Region variables are more complicated because they are arrays.
+\noindent Region variables are more complicated because they are
+arrays.
 
@@
 \begin{verbatim}
-Hx[l] = 0.5 * (region(H_local, face_lt)+ ...);
+Hx[i] = 0.5 * (region(H_local, face_lt)+ ...);
 \end{verbatim}
@@
 \noindent
-where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
+where {\tt i = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
   LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =

### 2010-12-21 opencl/papers/foropencl.tex
updates
ack!  error!

@@
 \IEEEauthorblockN{Jeffrey Overbey}
-\IEEEauthorblockA{Indiana University\\
+\IEEEauthorblockA{University of Illinois\\at Urbana-Champaign\\
 overbey2@illinois.edu}
@@
 The sheer complexity of programming for clusters of many or multi-core
-processors with tens of millions threads of execution make the simplicity of
+processors with tens of millions threads of execution makes the simplicity of
 the data-parallel model attractive.  The increasing complexity of

### 2011-04-09 opencl/paper-local/programming-model.tex
Initial version based on SciDAC proposal.

\section{Program Restructuring}

Major challenges face application programmers on exescale platforms.  Among
these challenges are of course performance, but in addition, programmers are
also challenged by the need for their applications to be maintainable and
portable across a wide variety of hardware platforms.  Unfortunately,
performance often is at odds with maintainability and portability.  For example,
hardware accelerators like GPUs are rapidly evolving and require separate coding
for each GPU vendor and even for a specific vendor's range of models.  A
synthetic cognition application at LANL has six different software versions just
to support a subset of the possible hardware combinations that exist today.

To meet these challenges we propose three main thrust areas:

1. Guidance to users on the choice of a programming model for their
applications.  This guidance will be based on an extensive analysis by proposal
team members on the models that would work well on existing and future exoscale
hardware architectures along with input from the user communities.  A high-level
programming model will provide application programmers with the ability to
incrementally raise the level of abstraction within their applications.  Only by
raising the level of abstraction can codes be retargeted by compilers across the
expected range of exoscale platforms.  Discussions with scientists in various
scientific disciplines (e.g., atmospheric science) suggest that programmers
would prefer that their codes just work \emph{as is} on exoscale platforms,
primarily because of the huge effort envolved in recoding a major scientific
code.  However experience on porting codes to LANL's petascale Roadrunner
computer suggests this may not be entirely possible, at least on early
encarnations of exoscale platforms.  Every code ported to Roadrunner was
completely rewritten to perform well on Roadrunner's many-core processors.  It
is our goal to provide user-guided refactoring tools to incrementally modify
codes in order to ease the porting effort for programmers, without a complete
hand recoding as was required for the initial applications that were ported to
Roadrunner.

2. Once a user or user community has chosen a programming model, tools can be
provided that restructure an application to adopt the chosen programming model.
These code refactorings need not be complicated.  For example, OpenMP uses
compiler directives to provide high-level information to the compiler.  Thus
analysis and refactoring tools guiding users in porting codes to an OpenMP
threading model would be appropriate.  Other tools supporting a data parallel
programming model would require a different set of tools.

3. Finally, given functions (or code segments) that have been restructured to
take advantage the chosen, high-level programming model, compiler tools are
needed to generate code for different system architectures.  We intent to
develop tools that are complimentary to and exploit a hardware vendor's C, C++,
OpenCL, and Fortran compilers.  We propose to develop source-to-source
transformational tools that take in high-level code and produce an intermidiate
code that can be further optimized by the vendor's backend compiler.

\subsection{Programming Models}

While it is impossible to predict the exact system architecture that an
exoscale system will have, we can anticipate the general direction that a
programming model must support.  We anticipate the need to support over 100K
nodes supported by a message passing communication system.  Each node can be
expected to contain a few 10s of heterogeneous cores with the possibility of
order 1K additional simple cores like those seen in existing GPUs.  Thus the
programming model must support this hierarchical distribution of
memory and processing elements.

Replicated programs using message passing with MPI is the current standard for
parallel programming using distributed nodes.  Parallel programmers have become
used to this environment and we expect to support this programming environment.  While
OpenMP is gathering acceptance, unfortunately there is currently not a
widely-adopted programming model that exploits on-node parallelism like MPI does
for distributed nodes.  Two evolving cross-platform standards for developing
multi-threaded applications are OpenMP and OpenCL.  We intend to develop tools
that support both.  OpenMP as a way to express parallelism directly within an
application and OpenCL as an intermediate language generated by our
source-to-source compiler tools.  OpenCL is too low-level a language to achieve our
code-maintainability goal but supports a data-parallel programming model as a
high level of abstraction.

\subsection{Source-Code Abstractions}

Since a majority of scientific codes are written in Fortran, a large effort will
be given to supporting the Fortran programming language.  Fortunately Fortran is
an ideal language for programming at a high level of abstraction, especially for
procedures chosen as targets for threaded architectures like GPUs, as it has a
well-developed array abstraction as part of the base language.  While not always
exploited in existing Fortran applications, most of these abstractions are
available in todays compilers (although not currently targeted for GPUs).
Fortran provides a data-parallel programming model much like CUDA and is well
suited for intermedicate code generation to, for example, OpenCL (or CUDA if it
becomes a standard supported by multiple vendors).  Many of these abstractions
can be extended to C and C++ by array abstractions and compiler directives.
Currently Intel is working on adding array extensions to their C compiler.

In Fortran a user can choose between several source-code abstractions for
exploiting parallelism.  The goal of this proposal is to provide tools transform
these abstractions to low-level code for final compilation.  Fortran supports a
data-parallel programming model with array syntax, e.g., C = A + B, where A, B,
C are arrays.  Or the user can explicitly use loop constructs like DO CONCURRENT
to expose parallelism (a Fortran 2008 addition) or by using similar constructs
available in OpenMP like, {\tt #pragma omp parallel for}.

\subsection{Locally-Oriented Programming}

We highlight the Fortran elemental source-code abstraction because it provides a
local orientation similar to that of OpenCL or CUDA kernel functions that were
designed to exploit streaming, highly-threaded architectures like GPUs.
Elemental functions are pure in that they are guaranteed to be free of side
effects such as I/O.  Consider a convolution where a 3x3 filter is applied to
individual pixel elements in a photograph to average of blur the original photograph.
A convolution is similar to stencil operations in computational fluid dynamics that
are used to obtain spatial derivatives of state variables.  Simple code is shown
with extentions to Fortran shown in capital letters.  The extensions could be
provided to C or Fortran functions via compiler directives rather than explicit 
language syntax.

\begin{verbatim}
    CONCURRENT real elemental function convolve(a, filter)
       real, intent(in) :: a, filter(3,3)
       real :: a_region(3,3)

       a_region = HALO(a, [1,1,1,1])
       convolve = sum( filter*a_region )

    end subroutine
\end{verbatim}

The {\tt convolve} function is called with
\begin{verbatim}
    A = convolve(A, filter)
\end{verbatim}
where {\tt A} and {\tt Avg} are arrays, but where the {\tt a} argument in the
{\tt convolve} definition is declared as a scalar.  The {\tt HALO} function
returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.  The
{\tt convolve} function is free of race conditions because of the copy semantics
of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute,
whereby no output variables can be updated before all threads have completed
execution.  In addition, while any thread may load from an extended region
about \emph{its} element with the {\tt HALO} function, it may only store into
its own elemental location.

\subsection{Code Restructuring Tools}

This example is provided in some detail as it highlights several of the tools
that are useful in aiding the user in the task of restructuring source code.
The goal of the restructuring is to transform it into a high-level, maintainable
form, that can be automatically transformed into a source form that can
ultimately be compiled into an executable on a particular hardware platform.  We
assume that the starting point is a traditional serial code using loop
constructs; the convolution example would require four nested loops.

\begin{itemize}

\item \emph{Code outliner.}  This tool would replace a block of code chosen by
the user with a function and a call to the function.  In the above example the
loops over the array elements would be moved to the elemental {\tt convolve}
function and replaced by a call to the newly created function.  The tool may
require input from the user as it completes its task (size of ghost/halo region
for example).

\item \emph{User modifications.} At this point the user may want to restructure
the outlined function {\tt convolve} by hand, by replacing the loops over the
filter with array syntax as shown in the example.

\item \emph{Insertion of memory placement hints.}  Any compiler generated code
will have to be conservative regarding where memory resides, whether on the
accelerator device or on the host CPU.  For the above example, memory for {\tt
A} will need to be copied to and from the device on entry and exit of function
{\tt convolve}.  However, memory for {\tt A} may already reside on the device
from a previous function invocation and therefore need not be copied.  Users may
provide hints to the compiler as to whether to copy the memory and if so, where
to begin the memory copy so that it can be done asynchronously with other work.
Tools will be created to guide the user in making this placement based on
dependence analysis.

\item \emph{Insertion of MPI communication.}  Memory placement hints can also be
used so that calls to MPI can be inserted into the code for the communication of
ghost cells between nodes.  Users may start with a version of the code that
already uses MPI, but the same information needed to generate OpenCL kernels can
be used to generate MPI communication; off-node memory is just another level in
the memory hierarchy.

%%% NOTE, I can provide a figure here showing the memory hierarchy.

\item \emph{Insertion of OpenMP directives.}  As an alternative to outlining a
function (the opposite of inlining) a tool will be created to restructure user
code by inserting OpenMP directives to indicate that loop iterations can be
performed concurrently, for example.  This tool will provide guidance to the
user as to where it may be appropriate to place these directives (and why) but
the ultimate responsibility for where and what directives to be used is a user's
responsibility.

\item \emph{Replace with DO CONCURRENT.}  Another alternative to code outlining
is to replace loops in Fortran with the DO CONCURRENT construct.  As with
compiler directives, this restructuring would require user input and can also
suggest loops that are candidates for converting.

\end{itemize}

At this point it is assumed that the user has finished restructuring the
original source into a form such that that parallelism is exposed to the
source-to-source compiler through language syntax or compiler directives.  This
is the high-level, maintainable, and portable source form that is stored in CVS
repository.  Tools are also required that can take the high-level source and
retargets it to a specific hardware platform.  These tools produce a lower-level
source, through source-to-source transformations, that a vendor's compiler will
use to generate executable code.  While described as separate tools, they will
likely be aggragated into a single tool.

The transformations may produce code for following targets (a combination of
language and hardware platform): 1. Serial version to be run on a single core.
This is necessary because some language constructs like DO CONCURRENT may not be
available for a period of time; 2. OpenCL version optimized for GPUs; 3. OpenCL
for optimzed for multiple CPU cores; and 4. OpenMP version for final compilation
with an OpenMP compiler.

\begin{itemize}

\item \emph{Thread initialization.} This tool generates code for calls to the
thread runtime library to initialize all thread devices and thread memory.

\item \emph{Code inliner.} This tool inlines calls to elemental functions,
effectively undoing the effect of the outliner. It is not needed for Fortran as
the compiler already performs this task.

\item \emph{Thread execution.} This tool generates code for execution of an
elemental or kernel function by a thread runtime library like OpenCL.

\item \emph{Memory transfer.}  This tool generates code for all memory transfer
and synchronization points.

\end{itemize}

Programming Models
------------------

Fortran currently supports:

1. Array syntax: e.g., C = A + B, where A, B, C are arrays.  Note that this is implicitly a loop structure, but that no loop indices need be provided.  Also the programmer need not specify where in memory these arrays reside.  Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardward threads).  This is a simple example and it is a research question as to what compiler directives would be useful for memory placement and other directions to the compiler for efficient code generation.

2. Pure procedures:  Fortran has syntax for specifying procedures that have no side effects during execution.  Specifying code that is side-effect free code is important information to provide to the compiler so that it can generate efficient multi-threaded code.

3. Pure elemental procedures: Fortran has syntax for specifying procedures that take only scalar arguments, but may be applied across array elements.  Elemental procedures are ideal for writing code to be executed within a hardware thread.  They resemble OpenCL kernels, but are simpler because they leave all indexing up to the compiler.

We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.  This additional syntax is provided by functions that return a copy of a small region of memory surrounding an array element (as seen within an elemental procedure) and with functions for thread synchronization.  This additional syntax will allow pure procedures to perform stencil and other convolution-like operations on a copy of memory, synchronize, then store the computed results back to the array element associated with the given thread.

In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.  This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.  Fortran also has excellent facilities for interoperability with C so that programming in a mixed language invironment is easily accomplished, including interoperability with native Fortran arrays.

### 2011-06-20 papers/shallow-water.tex
Renamed opencl/papers to papers.

--- 2010-12-19 opencl/papers/shallow-water.tex
@@
 can be ignored. A fixed time step is used for simplicity though it
-must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$ to fulfil the CFL
+must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$ to fulfill the CFL
 condition.
@@
 as some Fortran compilers (e.g., gfortran) do not generate optimal
-code for shifts.  We note (as seen below) that these temporary array
+code for shifts.  We note (as shown below) that these temporary array
 copies are replaced by scalars in the transformed Fortran code so

### 2011-06-20 papers/GPUScA-2011/intro.tex
Copied IPDPS 2011 paper to GPUScA 2011 workshop (in Galveston)

--- 2010-12-20 opencl/papers/intro.tex
@@
 Similar whole array shifting was used in higher dimensions for finite
-different codes within the computational physics community, especially
+difference codes within the computational physics community, especially
 at Los Alamos for codes targeting the CM-5 system that resided there until the
@@
 on the local view of indices that participate in the update of each element of
-an array.  For example, in the 1D averaging case stated above, a programmer would
-define the computation in terms of the update that occurs at each index in terms
-of relative offsets from that index:
+an array.

### 2011-06-20 papers/GPUScA-2011/foropencl.tex
Simplifications as discussed in Portland

--- 2010-12-21 opencl/papers/foropencl.tex
@@
 \begin{abstract}
-Emerging architectures for high performance computing are well suited
-to a data-parallel programming model.  This paper presents a simple
-programming methodology that allows programmers to take advantage of
-these emerging systems.  We use array constructs in Fortran 90 to show how this
-infrequently exploited, standardized language feature is easily
-transformed to lower-level accelerator code.  These transformations
-are based on a simple mapping from Fortran 90 to OpenCL.  We
-demonstrate, using a shallow-water code, that an acceleration of up to 40
-times can be achieved on GPU hardware with these compiler-generated
-transformations.
+Emerging GPU architectures for high performance computing are well suited to a
+data-parallel programming model.  This paper presents a simple programming
+methodology that allows programmers to take advantage of these emerging systems.
+We use array constructs in Fortran 90 to show how this infrequently exploited,
+standardized language feature is easily transformed to lower-level accelerator
+code.  These transformations are based on a simple mapping from Fortran 90 to
+OpenCL.  We demonstrate, using a stencil code solving the shallow-water fluid
+equations, that the performance of our compiler-generated transformations is
+comparable with that of hand-optimized OpenCL code.
 \end{abstract}

### 2011-06-20 papers/GPUScA-2011/intro.tex
Simplifications as discussed in Portland

@@
 
-This paper presents a compiler-level approach for targeting a single
-program to multiple, fundamentally different low-level execution
-models.  This technique allows the application programmer to adopt
-a single high-level programming model without sacrificing performance.
-We show that features of
-Fortran 90 for data-parallel programming are well suited to automatic
-transformation to generate code specifically tuned for different
-hardware architectures using low-level programming models such as
-OpenCL and CUDA.  For algorithms that can be easily expressed in terms
-of whole array, data-parallel operations, writing code in Fortran and
-transforming it automatically to specific low-level implementations
-removes the burden from the programmer of working with tedious, error
-prone, low-level tools.
+This paper presents a compiler-level approach for targeting a single program to
+multiple and possibly fundamentally different processor architectures.  This
+technique allows the application programmer to adopt a single, high-level
+programming model without sacrificing performance.  We suggest that existing
+data-parallel features in Fortran are well suited to applying automatic
+transformations that generate code specifically tuned for different hardware
+architectures using low-level programming models such as OpenCL.  For algorithms
+that can be easily expressed in terms of whole array, data-parallel operations,
+writing code in Fortran and transforming it automatically to specific low-level
+implementations removes the burden of creating and maintaining multiple versions
+of architecuture specific code.
 
-In the ideal situation, application programmers would like to adopt a
-programming model in which they write their application once and use
-automated tools to retarget it to many architectures.  This has proven
-to be very challenging historically due to the subtle balance between
-high-level expressiveness of code and the performance of the
-lower-level code that is emitted by a compiler.  This ideal high-level
-model that programmers work with should emphasize readability,
-maintainability, and close proximity in abstraction to the problem
-being solved --- in this instance, the abstraction that we care about
-are mathematical formulae.  It should not be corrupted with details of
-specific target architectures solely for the purpose of single-system
-performance.  For certain classes of applications, specifically those
-that map onto a data-parallel programming model, we show that
-Fortran 90 contains language features that encourage high-level
-programming abstractions without sacrificing performance during
-low-level code generation.
+%%%In the ideal situation, application programmers would like to adopt a programming model in which they write their application once and use automated tools to retarget it to many architectures.
+This has proven to be very challenging historically due to the subtle balance between high-level expressiveness of code and the performance of the lower-level code that is emitted by a compiler.
+This ideal high-level model that programmers work with should emphasize readability, maintainability, and close proximity in abstraction to the problem being solved --- in this instance, the abstraction that we care about are mathematical formulae.
+It should not be corrupted with details ofspecific target architectures solely for the purpose of single-system performance.
+For certain classes of applications, specifically those that map onto a data-parallel programming model, we show that Fortran 90 contains language features that encourage high-level programming abstractions without sacrificing performance during low-level code generation.
 
@@
 
-The performance that new accelerator architectures offer comes at a
-cost, as processor architectures are trending toward multiple cores
-with instances of integrated accelerator units (with user managed
-memory) and less of a reliance on superscalar instruction level
-parallelism and hardware managed memory hierarchies (such as
-traditional caches).  These changes place a heavy burden on
-application programmers as they are forced to adapt to the new
-systems.  An especially challenging problem faced by application
-programmers is not only how to program to these new architectures
-(considering the massive scale of concurrency available), but also how
-to design programs that are portable across the changing landscape of
-computer architectures with unique memory systems and programming
-models.  Can a programmer easily write one program that can run on a
-conventional multicore CPU, graphics processing unit, Cell processor,
-and one of many emerging many core architectures?  The fundamental
-question we address is what programming model and language constructs
-are best suited to span this set of new hardware designs.
+Unfortunately the performance that the new accelerator architectures offer comes
+at a cost.  Architectural changes are trending toward multiple heterogeneous
+cores and less of a reliance on superscalar instruction level parallelism and
+hardware managed memory hierarchies (such as traditional caches).  These changes
+place a heavy burden on application programmers as they are forced to adapt to
+these new systems.  An especially challenging problem is not only how to program
+to these new architectures (considering the massive scale of concurrency
+available), but also how to design programs that are portable across the
+changing landscape of computer architectures with unique memory systems and
+programming models.  How can a programmer easily write one program that can
+perform well on both a conventional multicore CPU and a GPU (or one of the
+emerging many-core architectures)?  A directive-based approach, such as OpenMP
+or the Accelerator programming model from the Portland
+Group~\cite{pgi10accelerator}, is one solution to this problem.
+
+%%%The fundamental question we address is what programming model and language constructs are best suited to span this set of new hardware designs.
 
@@
 
-A common theme amongst new processors is the emphasis on data-parallel
-programming.  This model is well suited to emerging architectures that
-are based on either vector processing or massively parallel
-collections of simple cores.  The recent CUDA and OpenCL programming
-languages are are intended to support this programming model, as are
-directive-based methods such as OpenMP or the Accelerator programming
-model from the Portland Group~\cite{pgi10accelerator}.
+In this paper we take a somewhat different approach.  A common theme amongst the new
+processors is the emphasis on data-parallel programming.  This model is well
+suited to the emerging architectures that are based on either vector processing
+or massively parallel collections of simple cores.  The recent CUDA and OpenCL
+programming languages are are intended to support this programming model,
 
@@
 
-The problem with many of these choices is that they expose too much
-detail about the machine architecture to the programmer.  This is
-particularly true of CUDA and OpenCL.  In CUDA, programmers must adapt
-their codes to fit the threading model used by NVIDIA GPUs, while
-OpenCL requires programmers to provide specially tuned versions of
-their code for different classes of machine.  In both cases, the
-programmer is responsible for explicitly managing memory,
-including staging of data back and forth from the host CPU and the
-accelerator device memory.  While these models have been attractive as
-a method for early adopters to utilize these new architectures, they
-are less attractive to programmers who do not have the time or
-resources to manually port their code to every new architecture and
-programming model that emerges.
+The problem with OpenCL and CUDA is that they expose too much detail about the
+machine architecture to the programmer.  The programmer is responsible for
+explicitly managing memory (including the staging of data back and forth between
+the host CPU and the accelerator device) and specifically taking into account
+architectural differences (such as whether the architecture contains vector
+units).  While these languages have been attractive as a method for early
+adopters to utilize these new architectures, they are less attractive to
+programmers who do not have the time or resources to manually port their code to
+every new architecture and programming model that emerges.
 
@@
 
-We demonstrate that the array syntax of Fortran maps surprisingly well
-onto GPUs when transformed to OpenCL kernels.  These Fortran language
-features include pure and elemental functions and array constructs
-like {\tt where} and {\tt cshift}.  In addition we add a few functions
+We demonstrate that the data-parallel features of Fortran map surprisingly well
+onto GPUs when transformed to OpenCL kernels.  These features include: array syntax
+employing binary operators, array constructs like {\tt WHERE} and {\tt DO CONCURRENT},
+and the use of pure and elemental functions.
+In addition we add functions that explicitly take advantage of the
+stencil geometry.
+
 that enable a program to be deployed on machines with a hierarchy of

### 2011-06-21 opencl/f90/shallow_water.f90
Added code go return the results.

@@ -115,4 +115,8 @@ program shallow_water
    flops = (1.0e-9 * 1000) * nLoops * (100*NX*NY/h_time)
    print *, "   flops        ==    ", flops, "GFlops"
 
+   ! get the results
+   !
+   status = readBuffer(d_H, c_loc(H), global_ex_mem_size)
+
 end program shallow_water

### 2011-06-21 pagosa/Advect_Basic.f90
Fixed wierd DOS formatting.

@@ -1,92 +1,93 @@
 ! +   Basic advection routine
 
-Subroutine Advect_Basic  (id,  Array,  Eps,  dFlux,  TmpA,  TmpB)
+Subroutine Advect_Basic (id, Array, Eps, dFlux, TmpA, TmpB)
 
-!=======================================================================
-!   Description:  The basic advection routine for all variables, except the
-!                           three hydrodynamic variables.
+!==============================================================================
+!  Description:  The basic advection routine for all variables, except the
+!                three hydrodynamic variables.
 !
-!   Method:  The donor, upwind, and downwind advection quantities are computed.
-!                    The advection is performed and put back into "Array"
+!  Method:  The donor, upwind, and downwind advection quantities are computed.
+!           The advection is performed and put back into "Array"
 !
-!   Reference: PAGOSA physics Manual,  LA-14425-M,  August 2010,  Los Alamos 
-!                        National Laoratory,  p. 42-48 and p. 183-193, p. 197-202.
+!  Reference: PAGOSA physics Manual,  LA-14425-M,  August 2010,  Los Alamos 
+!             National Laoratory,  p. 42-48 and p. 183-193, p. 197-202.
 !
-!   Externals:
-!       global_eoshift
+!  Externals:
+!      global_eoshift
 ! 
-!   History:                  Date                  Author
-!   Version
-!   - - - - - - - - - -         - - - - - - - - -      - - - - - - - - - - - - - - 
-!   17.0                       07/07/2009       Wayne Weseloh
+!  History:              Date            Author
+!  Version
+!  ----------         ----------      -------------
+!   17.0              07/07/2009      Wayne Weseloh
 !
-!=======================================================================
+!=============================================================================
 
-Use constants,            only:  zero
-Use Kind_Module,     only:  knit,  kreal
-Use Param_Module,  only:  mx, my, mz
-Use Shift_Module,      only:  global_eoshift
+Use constants,     only: zero
[…]

### 2011-06-21 pagosa/Advect_Basic_local.f90
Minor formatting and name changes.

@@ -1,6 +1,7 @@
 ! +   Basic advection routine
 
-Subroutine Advect_Basic (id, Array, Eps, dFlux, TmpA, TmpB)
+Pure Elemental &
+Subroutine Advect_Basic (id, Array, Eps, dFlux, lVol, lUpw, lDnw)
 
 !==============================================================================
 !  Description:  The basic advection routine for all variables, except the
@@ -22,7 +23,7 @@ Subroutine Advect_Basic (id, Array, Eps, dFlux, TmpA, TmpB)
 !
 !=============================================================================
 
-!$ OFP kernel Advect_Basic
+!$OFP concurrent :: Advect_Basic
 
 Use Kind_Module,   only: kint, kreal
 Use Param_Module,  only: mx, my, mz
@@ -34,11 +35,18 @@ Save
 Integer(kind=kint),  intent(in) :: id    !  advection sweep direction id
 
 !... Arrays:
-Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz), intent(inout) :: Array ! advection array section
-Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz)  intent(in)    :: Eps   ! residual donor mass/volume fraction
-Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz)  intent(in)    :: dFlux ! cell face mass/volume flux
-Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz)  intent(in)    :: TmpA  !
-Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz)  intent(in)    :: TmpB  !
+!----------------------------------------------------------------------------
+!Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz), intent(inout) :: Array ! advection array section
+!Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz)  intent(in)    :: Eps   ! residual donor mass/volume fraction
+!Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz)  intent(in)    :: dFlux ! cell face mass/volume flux
+!Real(kind=kreal),  dimension(0:mx, 0:my, 0:mz)  intent(in)    :: TmpA  !
[…]

### 2011-06-21 papers/IPDPS-2011/cl-performance.py
I think these were last minute changes for paper.

@@ -1,6 +1,7 @@
 import random
 import numpy as np
 import pylab as pl
+from math import *
 
 '''from scikits.learn import svm, datasets'''
 '''from scikits.learn.metrics import roc, auc'''
@@ -12,6 +13,7 @@ import matplotlib.pyplot as plt
 import matplotlib.mlab as mlab
 import matplotlib.cm as cm
 import matplotlib.lines as ln
+import matplotlib.scale as scale
 
 npts = 8
 
@@ -28,10 +30,13 @@ cl     = np.zeros(npts)
 
 x = [64.,128.,256.,512.,1024.,1280.,2048.,4096.]
 
-fdp    = [.2  , .76 , 3.02, 12.1, 49.5, 77.7, 199.1, 794.7]
+fdp    = [.25 , .78 , 3.16, 12.7, 52.2, 81.8, 209.8, 839.1]
 floops = [.19 , .71 , 2.82, 11.2, 45.0, 70.1, 178.7, 714.1]
 cl     = [.020, .036, .092, .32 , 1.22, 1.89, 4.82 , 19.29]
 
+#   fdp[i]    = log10(1.e6 * fdp[i]   /(x[i]*x[i]))
+#   floops[i] = log10(1.e6 * floops[i]/(x[i]*x[i]))
+#   cl[i]     = log10(1.e6 * cl[i]    /(x[i]*x[i]))
 for i in range(npts):
    fdp[i]    = 1.e6 * fdp[i]   /(x[i]*x[i])
    floops[i] = 1.e6 * floops[i]/(x[i]*x[i])
@@ -40,12 +45,18 @@ for i in range(npts):
 fig = plt.figure()
 ax = fig.add_subplot(111)
 
-ax.plot(x, fdp,    '-o', color='k', linewidth=3.0, ms=10.0)
-ax.plot(x, floops, '^' , color='k', linewidth=3.0, ms=10.0)
-ax.plot(x, cl,     '+' , color='k', linewidth=3.0, ms=10.0)
+ax.plot(x, fdp,    '--', color='k', linewidth=3.0, ms=10.0)
+ax.plot(x, floops, '-.' , color='k', linewidth=3.0, ms=10.0)
+ax.plot(x, cl,     '-' , color='k', linewidth=3.0, ms=10.0)
+ax.plot([1280], 1.e6*1.38/(1280*1280), 'o' , color='k', linewidth=2.0, ms=10.0)
 
[…]

### 2011-06-21 papers/local-orientation/code/elemental_add.f90
Added extensive comments.

@@ -1,15 +1,36 @@
-pure subroutine elemental_add(A, B, C)
-   real, intent(in)  :: A, B
+!
+! Eventual form of high-level representation.  This will first be turned
+! into syntax for OFP and ROSE and then, hopefully it can make its way
+! into the Fortran standard.
+!
+! pure elemental CONCURRENt subroutine elemental_add(A, B, C)
+!    TYPE(*), intent(COPYIN)  :: A, B
+!    TYPE(*), intent(COPYOUT) :: C
+!
+!   C = A + B   
+!
+!end subroutine elemental_add
+
+
+!
+! Note that this function is dimensionless.  The function parameters are scalars
+! but the compiler will generate code for arrays of a particular dimension based
+! on the actual parameters at the calling site.  Also note that the parameters
+! are typeless (see TYPE(*) above); the compiler will create an instance of
+! function according to the called type.
+!
+pure elemental subroutine elemental_add(A, B, C, s)
+!   !
+!   ! compiler directives that will eventually be turned into syntax
+!   ! 
+!   !$OFP CONCURRENT :: elemental_add
+!   !$OFP intent(COPYIN)   :: A, B
+!   !$OFP intent(COPYOUT)  :: C
+!   !
+!
+   real, intent(in), value :: A, B, s
    real, intent(out) :: C
 
-   !
-   ! Add compiler directives that will be turned into syntax
-   ! 
-   !$OFP thread elemental :: elemental_add
-   !$OFP intent(COPYIN)   :: A, B
-   !$OFP intent(COPYOUT)  :: C
-   !
-
-   C = A + B   
+   C = A + s*B   
 
 end subroutine elemental_add

### 2011-06-21 papers/local-orientation/codegen/elemental_add.f90
Moved to ../code

@@ -1,15 +1,0 @@
-pure subroutine elemental_add(A, B, C)
-   real, intent(in)  :: A, B
-   real, intent(out) :: C
-
-   !
-   ! Add compiler directives that will be turned into syntax
-   ! 
-   !$OFP thread elemental :: elemental_add
-   !$OFP intent(COPYIN)   :: A, B
-   !$OFP intent(COPYOUT)  :: C
-   !
-
-   C = A + B   
-
-end subroutine elemental_add

### 2011-06-21 papers/local-orientation/codegen/template.c
Testing with float4

@@ -2,22 +2,23 @@
  * This kernel is automatically generated by Open Fortran Project tools.
  */
 
-__kernel void Retina_update_state (
-    float time,
-    float dt,
-    int nx,
-    int ny,
-    int nf,
-    int nb,
-    int params,
-    __global float * phiExc,
-    __global float * phiInh,
-    __global float * activity)
+static inline int get_global_idx_1d()
 {
-    __global float * time, dt
-  integer, intent(in) :: nx, ny, nf, nb
-  integer, intent(in) :: params
-  integer :: params1
-    __global float * phiExc, phiInh, activity
+   return get_global_id(0);
+}
 
+static inline int get_global_idx_2d()
+{
+   return get_global_id(0) + get_global_size(0)*get_global_id(1);
+}
+
+__kernel void elemental_add (
+    __global float4 * A,
+    __global float4 * B,
+    __global float4 * C )
+{
+
+   int idx = get_global_idx_2d();
+
+   C[idx] = A[idx] + B[idx];
 }

### 2011-06-22 papers/GPUScA-2011/intro.tex
Cleaned up and shortened by moving some language discussion to the next section.

@@
 This paper presents a compiler-level approach for targeting a single program to
-multiple and possibly fundamentally different processor architectures.  This
+multiple, and possibly fundamentally different, processor architectures.  This
 technique allows the application programmer to adopt a single, high-level
@@
 
-%%%In the ideal situation, application programmers would like to adopt a programming model in which they write their application once and use automated tools to retarget it to many architectures.
-This has proven to be very challenging historically due to the subtle balance between high-level expressiveness of code and the performance of the lower-level code that is emitted by a compiler.
-This ideal high-level model that programmers work with should emphasize readability, maintainability, and close proximity in abstraction to the problem being solved --- in this instance, the abstraction that we care about are mathematical formulae.
-It should not be corrupted with details ofspecific target architectures solely for the purpose of single-system performance.
-For certain classes of applications, specifically those that map onto a data-parallel programming model, we show that Fortran 90 contains language features that encourage high-level programming abstractions without sacrificing performance during low-level code generation.
+The peak performance of the newer accelerator architectures can be substantial.
+Intel expects a teraflop for the SGEMM benchmark with their Knights Ferry
+processor while the performance of the M2090 NVIDIA Tesla processor is in the
+same neighborhood~\cite{hpcwire11manycore}.  Unfortunately the performance that
+many of the new accelerator architectures offer comes at a cost.
+Architectural changes are trending toward multiple heterogeneous cores and less
+of a reliance on superscalar instruction level parallelism and hardware managed
+memory hierarchies (such as traditional caches).  These changes place a heavy
+burden on application programmers as they are forced to adapt to these new
+systems.  An especially challenging problem is not only how to program to these
+new architectures (considering the massive scale of concurrency available), but
+also how to design programs that are portable across the changing landscape of
+computer architectures with unique memory systems and programming models.  How
+can a programmer easily write one program that can perform well on both a
+conventional multicore CPU and a GPU (or one of the emerging many-core
+architectures)?  A directive-based approach, such as OpenMP or the Accelerator
+programming model from the Portland Group~\cite{pgi10accelerator}, is one
+solution to this problem.
 
-%1. Goal is to write once, transform many. Otherwise potentially reprogram for every architecture.
-%2. Goal is to code for readability and maintainability, not performance
-%3. This goal requires expression in a high-level language.
-%4. But language must be simple enough for compiler to analyze.
-%5. Thus ideal if language maps well to accelerator architectures.
-%6. Data parallel constructs in Fortran 90 are chosen.
-%7. Up to 65 times speed up measured on automatically transformed code.
-
-%At the Los Alamos National Laboratory (LANL), as with many supercomputing
-%facilities today, users have a wide variety of computer platforms from
-%which to choose.  The most common platform is made up of clusters of
-%compute nodes with standard multi-core processors.  An increasingly
-%common feature is that some nodes also have accelerators that range
-%from the IBM Cell processor (such as the LANL Roadrunner system) to a
-%variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
-%vector instructions and others do not.  The peak performance of these
-%accelerated nodes often resides in the hundreds of gigaflops.
-
-%The peak performance of the
-%accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
-%processor to XXX for NVIDIA Fermi. %% Fermi has another name
-
-Unfortunately the performance that the new accelerator architectures offer comes
-at a cost.  Architectural changes are trending toward multiple heterogeneous
-cores and less of a reliance on superscalar instruction level parallelism and
-hardware managed memory hierarchies (such as traditional caches).  These changes
-place a heavy burden on application programmers as they are forced to adapt to
-these new systems.  An especially challenging problem is not only how to program
-to these new architectures (considering the massive scale of concurrency
-available), but also how to design programs that are portable across the
-changing landscape of computer architectures with unique memory systems and
-programming models.  How can a programmer easily write one program that can
-perform well on both a conventional multicore CPU and a GPU (or one of the
-emerging many-core architectures)?  A directive-based approach, such as OpenMP
-or the Accelerator programming model from the Portland
-Group~\cite{pgi10accelerator}, is one solution to this problem.
-
-%%%The fundamental question we address is what programming model and language constructs are best suited to span this set of new hardware designs.
-
-%% NEW - CER
-%%%We will examine how the existing data-parallel constructs in Fortran can be combined with coarrays or MPI to provide effectively a new parallel programming language, one that is evolutionary in nature and provides complete compatibility with existing applications and libraries.
-Data parallelism is a high-level abstraction that is, at the same time, both easier to program and gives the compiler more leeway (if fully exploited) in retargeting a program to different computer architectures.
-
-In this paper we take a somewhat different approach.  A common theme amongst the new
-processors is the emphasis on data-parallel programming.  This model is well
+In this paper we take a somewhat different approach.  A common theme amongst the
+new processors is the emphasis on data-parallel programming.  This model is well
 suited to the emerging architectures that are based on either vector processing
@@
 
-%% Deleted by CER
-%%A proposed approach for programming that is well suited to legacy applications and languages are language extensions or libraries that allow programmers to avoid adopting entirely new languages.
-
 The problem with OpenCL and CUDA is that they expose too much detail about the
-machine architecture to the programmer.  The programmer is responsible for
-explicitly managing memory (including the staging of data back and forth between
-the host CPU and the accelerator device) and specifically taking into account
-architectural differences (such as whether the architecture contains vector
-units).  While these languages have been attractive as a method for early
-adopters to utilize these new architectures, they are less attractive to
-programmers who do not have the time or resources to manually port their code to
-every new architecture and programming model that emerges.
-
-%At this point in time it is not really possible to write once and run
-%efficiently on the wide variety of computer platforms we have
-%available.  For some classes of applications, we believe that this
-%goal is possible using language constructs already present in a
-%popular mainstream scientific programming language -- Fortran.  A
-%common, long-standing tongue-in-cheek response to new language
-%developments in the scientific and high performance computing
-%community is that a new language will arise to answer the needs of new
-%systems, and it will be called Fortran.  We believe that the work
-%presented in this paper validates that notion -- we need a new
-%language to work with, and that language is Fortran.
+machine architecture to the programmer~\cite{wolfeSC10}.  The programmer is
+responsible for explicitly managing memory (including the staging of data back
+and forth between the host CPU and the accelerator device) and specifically
+taking into account architectural differences (such as whether the architecture
+contains vector units).  While these languages have been attractive as a method
+for early adopters to utilize these new architectures, they are less attractive
+to programmers who do not have the time or resources to manually port their code
+to every new architecture and programming model that emerges.
 
@@
 
-%This paper addresses the accelerator programming problem by examining
-%features in Fortran that allow programmers to express algorithms at a
-%very high level that can be easily transformed by a compiler to run
-%efficiently on a wide variety of platforms.  In particular we consider
-%computers based on GPUs and related accelerator processors.
+We demonstrate that the data-parallel features of Fortran map surprisingly well
+onto GPUs when transformed to OpenCL kernels.  These features include: array
+syntax (e.g., assignment and binary operators), array constructs like {\tt
+  WHERE} and {\tt DO CONCURRENT}, and the use of pure and elemental functions.
+In addition, we provide new functions that explicitly take advantage of the
+stencil geometry of the problem domain we are addressing.  Note that this subset
+of the Fortran language is implicitly parallel.  This programming model
+\emph{does not require explicit declaration of parallelism within the program.}
+In addition, programs are expressed using entirely standard Fortran so it can be
+compiled for and executed on a single core without concurrency.
 
-We demonstrate that the data-parallel features of Fortran map surprisingly well
-onto GPUs when transformed to OpenCL kernels.  These features include: array syntax
-employing binary operators, array constructs like {\tt WHERE} and {\tt DO CONCURRENT},
-and the use of pure and elemental functions.
-In addition we add functions that explicitly take advantage of the
-stencil geometry.
-
-that enable a program to be deployed on machines with a hierarchy of
-processing elements, such as nodes employing GPU acceleration,
-\emph{without requiring explicit declaration of parallelism within the
-  program.}  In addition the program uses entirely standard Fortran so
-it can be compiled for and executed on a single core without concurrency.
-This work also is applicable to vendor-specific languages similar to
-OpenCL such as the NVIDIA CUDA language.
-
-%We provide (via Fortran interfaces in the ForOpenCL library) a
-%mechanism to call the C OpenCL runtime and enable Fortran programmers
-%to access OpenCL kernels.  
-Transformations are supplied that provide a mechanism for converting
-Fortran procedures written in the Fortran subset described in this
-paper to OpenCL kernels.  We use the ROSE compiler
-infrastructure\footnote{\url{http://www.rosecompiler.org/}} to
-develop these transformations.  ROSE uses the Open Fortran
-Parser\footnote{\url{http://fortran-parser.sf.net/}} to parse
-Fortran 2008 syntax and can generate C-based OpenCL.  Since ROSE's
-intermediate representation (IR) was constructed to represent multiple
-languages, it is relatively straightforward to transform high-level
-Fortran IR nodes to C OpenCL nodes.
-
-%  Furthermore, the Fortran array syntax maps directly to one,
-%two, and three-dimensional thread groups in OpenCL.
+Transformations are supplied that provide a mechanism for converting Fortran
+procedures written in the Fortran subset described in this paper to OpenCL
+kernels.  We use the ROSE compiler
+infrastructure\footnote{\url{http://www.rosecompiler.org/}} to develop these
+transformations.  ROSE uses the Open Fortran
+Parser\footnote{\url{http://fortran-parser.sf.net/}} to parse Fortran 2008
+syntax and can generate C-based OpenCL.  Since ROSE's intermediate
+representation (IR) was constructed to represent multiple languages, it is
+relatively straightforward to transform high-level Fortran IR nodes to C OpenCL
+nodes.  This work is also applicable to transformations to vendor-specific
+languages, similar to OpenCL, such as the NVIDIA CUDA language.
 
 Transformations for arbitrary Fortran procedures are not attempted.
-Furthermore, a mechanism to transform the calling site to
-automatically invoke OpenCL kernels is not provided at this time.
-While it is possible to accomplish this task within ROSE, it is
-considered outside the scope of this paper.
+Furthermore, a mechanism to transform the calling site to automatically invoke
+OpenCL kernels is not provided at this time.  While it is possible to accomplish
+this task within ROSE, it is considered outside the scope of this paper.
+However, we do provide (via Fortran interfaces) a mechanism to call the C OpenCL
+runtime and enable Fortran programmers to access OpenCL kernels generated by the
+automatic transformations.
 
-We examine the performance of the Fortran data-parallel abstraction
-when transformed to OpenCL to run on GPU architectures.  Since single
-node performance is often given as a reason for not using
-data-parallel constructs within Fortran, we consider the performance
-of serial data-parallel codes compared with the usage of explicit loop
-constructs.
+We study the automatic transformations for an application example that is
+typical of stencil codes that update array elements according to a fixed
+pattern.  Stencil codes are often employed in applications based on
+finite-difference or finite-volume methods in computational fluid dynamics
+(CFD).  The example described later in this paper is a simple shallow-water
+model in two dimensions using finite volume methods.
 
-We study automatic transformations and the performance for an
-application example that is typical of many applications that are
-based on finite-difference or finite-volume methods in computational
-fluid dynamics (CFD).  The example described later in this paper is a
-simple shallow water model in two dimensions using finite volume
-methods.
-
-An initial study was made for an important procedure in PAGOSA, a
-non-research, production-grade code at LANL completely written in
-data-parallel Fortran.  We investigated automatically transforming
-this code to run on LANL's Petaflop Roadrunner computer (a hybrid
-mixture of AMD Opterons and IBM Cell processors).  We demonstrated that a
-source-to-source compiler can automatically vectorize and parallelize
-a small section of this code for the Cell processor.  Preliminary results
-showed a 9 times performance gain of the transformed code when compared
-with the original serial version on a traditional single-core processor.
-
-\subsection{Why Fortran?}
-
-Fortran is the oldest high-level programming language in continuous
-use since its introduction, and was developed to facilitate the
-translation of math formulae into machine code. Fortran was the first
-major language to use a compiler to translate from a high-level
-program representation to assembly language. Due to its age, it
-carries certain arcane baggage.  However with the introduction of
-Fortran 90 (and later revisions to the standard), Fortran became a
-truly modern programming language.  It is now modular and has many
-object-oriented features.  Most importantly for this work, it now
-includes a type system in which rich first-class array data types and
-corresponding syntax are part of the language, something that
-languages like C continue to lack\footnote{The recent Intel 12.0 C/C++
-  compiler supports extensions that provide array notation for C/C++
-  code, as detailed here: http://intel.ly/g454zp}.  Furthermore,
-Fortran should be of interest to those studying parallel programming
-because of its functional and data-parallel constructs and because of
-the coarray notation introduced in Fortran 2008. Unlike languages like
-C and C++, Fortran has become a truly parallel language with features
-added to recent language standards.
-
-%% - deleted by CER
-%% that later programming languages have evolved away from.
-As a result, Fortran has fallen into disfavor in certain programming circles.
-Modern versions of the language standardized in 1990, 1995, 2003, and 2008 have removed much of this legacy baggage, but these changes are not widely known.
-Modern Fortran exhibits features that are similar to other modern programming languages, and does not mandate the use of legacy features from decades old, deprecated versions of the language.
-Compilers for Fortran (like those for other languages that have removed features over time) can prohibit the use of archaic features such as fixed format code or constructs that have been removed from the language.
-
-Yet, a likely question that one may pose is ``\emph{Why Fortran and
-  not a more modern language like X?}''  The recent rise in interest
-in concurrency and parallelism at the language level driven by
-multicore CPUs and manycore accelerators has driven a number of new
-language developments, both as novel languages and extensions on
-existing ones.  For scientific users, new languages and language
-extensions to use novel new architectures present a challenge: how do
-developers effectively use them while avoiding rewriting code
-and potentially growing dependent on a transient technology that will
-vanish tomorrow?
-
-%% NEW - CER
-%So perhaps it is time to replace Fortran with yet another computer language.  The
-%problems with replacing Fortran with an entirely new language are two fold:
-%the economics of replacing the existing application base and the difficulty in
-%obtaining programmer acceptance.  It is estimated that replacing a major
-%production application at Los Alamos National Laboratory would cost between 50
-%and 150 million dollars.  In terms of programmer acceptance, there is always
-%the "chicken and egg problem": programmers won't use a new language until they
-%can expect good performance across a variety of platforms, and compiler
-%vendors can't afford to produce quality compilers until there is a reasonable
-%expectation of a market.
-
-%% NEW - CER
-History has also shown that an investment in rewriting code does not
-guarantee success either, as seen in an effort at LANL to modernize a
-legacy Fortran code with the newer C++ POOMA framework.  This
-massive overhaul effort led to a code that was both slower and less
-flexible than the original Fortran \cite{basili08hpc}.
-
-%Similar
-%experiences have occurred in the past, notably during the development
-%of the functional SISAL language in the early 1990s.
-
-Fortran is unique in that it has contained language features that are
-well suited to modern architectures for a number of years.  This
-should be unsurprising --- Fortran was a primary language used to
-target systems such as the vector supercomputers and massively
-parallel systems of the 1970s and 1980s.  These are the systems in
-which architectural features were developed that have led to single
-chip high performance architectures of interest today.  Given that
-these new systems have features very similar to their predecessors, it
-is clear that the language features within Fortran for them are still
-relevant.
-
-\subsection{Comparison to Other Languages}
-
-A number of previous efforts have exploited data-parallel programming
-at the language level to utilize novel architectures, particularly in
-previous decades during the reign of vector and massively parallel
-computers in the high performance computing world.  The origin of the
-array syntax that was adopted in Fortran 90 can be found in the APL
-language.  Fortran 90 differed from previous % \cite{iverson79apl}
-extensions of Fortran in that parallelism within whole-array
-operations was expressed at the expression level instead of via
-parallelism within explicit DO-loops (such as within IVTRAN for the
-Illiac IV).
-
-The High Performance Fortran (HPF) extension of Fortran 90 was
-proposed to add features to the language that would enhance the
-ability of compilers to emit fast parallel code for distributed and
-shared memory parallel computers\cite{koelbel94hpf}.  One of the
-notable additions to the language in HPF was syntax to specify the
-distribution of data structures amongst a set of parallel processors.
-HPF also introduced an alternative looping construct to the
-traditional DO-loop called {\tt FORALL} that was better suited for
-parallel compilation.  An additional keyword, {\tt INDEPENDENT}, was
-added to allow the programmer to indicate when the order of execution
-of the program (such as a sequence of loop iterations) can be flexible
-in order to allow parallel execution.  Interestingly, the parallelism
-features introduced in HPF did not exploit the new array features
-introduced in 1990 in any significant way, relying instead on explicit
-loop-based parallelism.  This was likely in order to support parallel
-programming that wasn't easily mapped onto a pure data-parallel model.
-
-In some instances though, a purely data-parallel model is appropriate
-for part or all of the major computations within a program.  One of
-the systems where programmers relied heavily on higher level
-operations instead of explicit looping constructs was the Thinking
-Machines Connection Machine 5 (CM-5).  A common programming pattern
-used on the CM-5 that we exploit in this paper was to write
-whole-array operations from a global perspective in which computations
-are expressed in terms of operations over the entire array instead of
-a single local index.  The use of the array shift intrinsic functions
-(like {\tt CSHIFT}) were used to build computations in which arrays
-were combined by shifting the entire arrays instead of working based
-on local offsets from single indices.  A simple 1D example is one in
-which an element is replaced with the average of its own value with
-that of its two direct neighbors.  Ignoring boundary indices that wrap
-around, explicit indexing will result in a loop such as:
-
-{\small
-\begin{verbatim}
-  do i = 2,(n-1)
-    X(i) = (X(i-1) + X(i) + X(i+1)) / 3
-  end do
-\end{verbatim}
-}
-
-\noindent When shifts are employed, this can be expressed as:
-
-{\small
-\begin{verbatim}
-  X = (cshift(X,-1) + X + cshift(X,1)) / 3
-\end{verbatim}
-}
-
-Similar whole array shifting was used in higher dimensions for finite
-difference codes within the computational physics community, especially
-at Los Alamos for codes targeting the CM-5 system that resided there until the
-late 1990s.  A body of research in compilation of stencil-based codes
-that use shift operators targeting these systems is related to the
-work we present here~\cite{stencil-compiler}.
-
-The whole-array model was attractive because it deferred
-responsibility for optimally implementing the computations to the
-compiler.  Instead of relying on a compiler to infer parallelism from
-a set of explicit loops, the choice for how to implement loops was
-left entirely up to the tool.  Unfortunately, this had two side
-effects that have limited broad acceptance of the whole-array
-programming model in Fortran.  First, programmers must translate their
-algorithms into a set of global operations.  Finite difference
-stencils and similar computations are traditionally defined in terms
-of offsets from some central index.  Shifting, while conceptually
-analogous, can be awkward to think about for high dimensional stencils
-with many points.  Second, the semantics of these operations are such
-that all elements of an array operation are updated as if they were
-updated simultaneously.  In a program where the programmer explicitly
-manages arrays and loops, double buffering techniques and user managed
-temporaries are used to maintain these semantics.  When the compiler
-is responsible for managing this intermediate storage, it has
-historically proven that they are inefficient and generate code that
-requires far more temporary storage than really necessary.  This is
-not a flaw of the language constructs, but a sign of the lack of
-sophistication of the compilers with respect to their internal
-analysis to determine how to optimally generate this intermediate
-storage.
-
-An interesting line of language research that grew out of the early work
-with HPF was that associated with the ZPL language work at the University
-of Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar
-global view of computation over arrays, but define their computations based
-on the local view of indices that participate in the update of each element of
-an array.
+Finally, we examine the performance of the Fortran data-parallel abstraction
+when transformed to OpenCL to run on GPU architectures.  The performance of
+automatically transformed code is compared with a hand-optimized OpenCL
+version of the shallow-water code.

### 2011-06-22 papers/GPUScA-2011/programming-model.tex
Moved comparison from intro to programming model section

--- 2010-12-20 opencl/papers/programming-model.tex
@@
 \section{Programming Model}
+
+A question that one may pose is ``\emph{Why Fortran and
+  not a more modern language like X?}''  The recent rise in interest
+in concurrency and parallelism at the language level driven by
+multicore CPUs and manycore accelerators has driven a number of new
+language developments, both as novel languages and extensions on
+existing ones.  For scientific users, new languages and language
+extensions to use novel new architectures present a challenge: how do
+developers effectively use them while avoiding rewriting code
+and potentially growing dependent on a transient technology that will
+vanish tomorrow?
+
+\subsection{Comparison to Other Languages}
+
+A number of previous efforts have exploited data-parallel programming
+at the language level to utilize novel architectures, particularly in
+previous decades during the reign of vector and massively parallel
+computers in the high performance computing world.  The origin of the
+array syntax that was adopted in Fortran 90 can be found in the APL
+language.  Fortran 90 differed from previous % \cite{iverson79apl}
+extensions of Fortran in that parallelism within whole-array
+operations was expressed at the expression level instead of via
+parallelism within explicit DO-loops (such as within IVTRAN for the
+Illiac IV).
+
+The High Performance Fortran (HPF) extension of Fortran was
+proposed to add features to the language that would enhance the
+ability of compilers to emit fast parallel code for distributed and
+shared memory parallel computers\cite{koelbel94hpf}.  One of the
+notable additions to the language in HPF was syntax to specify the
+distribution of data structures amongst a set of parallel processors.
+HPF also introduced an alternative looping construct to the
+traditional DO-loop called {\tt FORALL} that was better suited for
+parallel compilation.  An additional keyword, {\tt INDEPENDENT}, was
+added to allow the programmer to indicate when the order of execution
+of the program (such as a sequence of loop iterations) can be flexible
+in order to allow parallel execution.  Interestingly, the parallelism
+features introduced in HPF did not exploit the new array features
+introduced in 1990 in any significant way, relying instead on explicit
+loop-based parallelism.  This was likely in order to support parallel
+programming that wasn't easily mapped onto a pure data-parallel model.
+
+In some instances though, a purely data-parallel model is appropriate
+for part or all of the major computations within a program.  One of
+the systems where programmers relied heavily on higher level
+operations instead of explicit looping constructs was the Thinking
+Machines Connection Machine 5 (CM-5).  A common programming pattern
+used on the CM-5 that we exploit in this paper was to write
+whole-array operations from a global perspective in which computations
+are expressed in terms of operations over the entire array instead of
+a single local index.  The use of the array shift intrinsic functions
+(like {\tt CSHIFT}) were used to build computations in which arrays
+were combined by shifting the entire arrays instead of working based
+on local offsets from single indices.  A simple 1D example is one in
+which an element is replaced with the average of its own value with
+that of its two direct neighbors.  Ignoring boundary indices that wrap
+around, explicit indexing will result in a loop such as:
+
+{\small
+\begin{verbatim}
+  do i = 2,(n-1)
+    X(i) = (X(i-1) + X(i) + X(i+1)) / 3
+  end do
+\end{verbatim}
+}
+
+\noindent When shifts are employed, this can be expressed as:
+
+{\small
+\begin{verbatim}
+  X = (cshift(X,-1) + X + cshift(X,1)) / 3
+\end{verbatim}
+}
+
+Similar whole array shifting was used in higher dimensions for finite
+difference codes within the computational physics community, especially
+at Los Alamos for codes targeting the CM-5 system that resided there until the
+late 1990s.  A body of research in compilation of stencil-based codes
+that use shift operators targeting these systems is related to the
+work we present here~\cite{stencil-compiler}.
+
+The whole-array model was attractive because it deferred
+responsibility for optimally implementing the computations to the
+compiler.  Instead of relying on a compiler to infer parallelism from
+a set of explicit loops, the choice for how to implement loops was
+left entirely up to the tool.  Unfortunately, this had two side
+effects that have limited broad acceptance of the whole-array
+programming model in Fortran.  First, programmers must translate their
+algorithms into a set of global operations.  Finite difference
+stencils and similar computations are traditionally defined in terms
+of offsets from some central index.  Shifting, while conceptually
+analogous, can be awkward to think about for high dimensional stencils
+with many points.  Second, the semantics of these operations are such
+that all elements of an array operation are updated as if they were
+updated simultaneously.  In a program where the programmer explicitly
+manages arrays and loops, double buffering techniques and user managed
+temporaries are used to maintain these semantics.  When the compiler
+is responsible for managing this intermediate storage, it has
+historically proven that they are inefficient and generate code that
+requires far more temporary storage than really necessary.  This is
+not a flaw of the language constructs, but a sign of the lack of
+sophistication of the compilers with respect to their internal
+analysis to determine how to optimally generate this intermediate
+storage.
+
+An interesting line of language research that grew out of the early work
+with HPF was that associated with the ZPL language work at the University
+of Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar
+global view of computation over arrays, but define their computations based
+on the local view of indices that participate in the update of each element of
+an array.
 
@@
 
-Since pure procedures don't have side effects they a candidates for
+Since pure procedures don't have side effects they are candidates for
 running on accelerators in OpenCL.  Currently our prototype tool
@@
 
-Blelloch~\cite{blelloch90,rajopadhye93}.
+%Blelloch~\cite{blelloch90,rajopadhye93}.
 

### 2011-06-23 papers/GPUScA-2011/programming-model.tex
Added intro to section.

@@
 
-A question that one may pose is ``\emph{Why Fortran and
-  not a more modern language like X?}''  The recent rise in interest
-in concurrency and parallelism at the language level driven by
-multicore CPUs and manycore accelerators has driven a number of new
-language developments, both as novel languages and extensions on
-existing ones.  For scientific users, new languages and language
-extensions to use novel new architectures present a challenge: how do
-developers effectively use them while avoiding rewriting code
-and potentially growing dependent on a transient technology that will
-vanish tomorrow?
+A question that one may pose is ``\emph{Why Fortran and not a more modern
+  language like X?}''  The recent rise in interest in concurrency and
+parallelism at the language level driven by multicore CPUs and manycore
+accelerators has driven a number of new language developments, both as novel
+languages and extensions on existing ones.  For many scientific users familiar
+with Fortran, new languages and language extensions to use novel new
+architectures present a challenge: how do they effectively use them while
+avoiding rewriting code and potentially growing dependent on a transient
+technology that will vanish tomorrow?
+
+In this section we present the Fortran subset employed in this paper.  This
+subsetting language will allow scientific programmers to stay within the Fortran
+language and yet have direct access to GPU hardware.  We first examine how this
+programming model relates to work in other languages.
 

### 2011-06-23 papers/GPUScA-2011/foropencl.tex
Modified to use ACM style.

@@
 
-\documentclass[10pt, conference, compsocconf]{IEEEtran}
+\documentclass{acm_proc_article-sp}
 
@@
 \usepackage{graphicx}
-\usepackage[cmex10]{amsmath}
-\usepackage{amssymb}
-\usepackage{array}
+%\usepackage[cmex10]{amsmath}
+%\usepackage{amssymb}
+%\usepackage{array}
 \usepackage{url}
@@
 
+\begin{document}
+
 \title{Exploiting Array Syntax in Fortran for Accelerator Programming}
 
-%%
-%% reorder appropriately later
-%%
-\author{\IEEEauthorblockN{Matthew J. Sottile}
-\IEEEauthorblockA{Galois, Inc.\\
-%421 SW 6th Ave. Suite 300 \\
-%Portland, OR 97204\\
-matt@galois.com}
-\and
-\IEEEauthorblockN{Craig E Rasmussen,\\
-                  Wayne N. Weseloh,\\
-                  Robert W. Robey}
-\IEEEauthorblockA{Los Alamos National Laboratory\\
-%CCS-7, MS B287\\
-%Los Alamos, NM\\
-\{crasmussen, weseloh, brobey\}@lanl.gov}
+\numberofauthors{6}
 
-\and
-\IEEEauthorblockN{Daniel Quinlan}
-\IEEEauthorblockA{Lawrence Livermore\\ National Laboratory\\
-dquinlan@llnl.gov}
-
-\and
-\IEEEauthorblockN{Jeffrey Overbey}
-\IEEEauthorblockA{University of Illinois\\at Urbana-Champaign\\
-overbey2@illinois.edu}
-
+\author{
+% 1st. author
+\alignauthor
+Matthew J. Sottile\\
+       \affaddr{Galois, Inc.}\\
+       \affaddr{421 SW 6th Ave. Suite 300}\\
+       \affaddr{Portland, OR 97204}\\
+       \email{matt@galois.com}
+% 2nd. author
+\alignauthor
+Craig E Rasmussen\\
+       \affaddr{Los Alamos National Laboratory}\\
+       \affaddr{CCS-7, MS B287}\\
+       \affaddr{Los Alamos, NM 87545}\\
+       \email{crasmussen@lanl.gov}
+% 3rd. author
+\alignauthor
+Wayne N. Weseloh\\
+       \affaddr{Los Alamos National Laboratory}\\
+       \affaddr{XTD-1, MS T082}\\
+       \affaddr{Los Alamos, NM 87545}\\
+       \email{weseloh@lanl.gov}
+\and  % use '\and' if you need 'another row' of author names
+% 4th. author
+\alignauthor
+Robert W. Robey\\
+       \affaddr{Los Alamos National Laboratory}\\
+       \affaddr{XCP-2, MS T086}\\
+       \affaddr{Los Alamos, NM 87545}\\
+       \email{brobey@lanl.gov}
+% 5th. author
+\alignauthor
+Daniel Quinlan\\
+       \affaddr{Lawrence Livermore National Laboratory}\\
+       \email{dquinlan@llnl.gov}
+% 6th. author
+\alignauthor
+Jeffrey Overbey\\
+       \affaddr{University of Illinois at Urbanna-Champaign}\\
+       \email{overbey2@illinois.edu}
 }
-
-\begin{document}
 
@@
 \end{abstract}
+
+% A category with the (minimum) three required fields
+\category{H.4}{Information Systems Applications}{Miscellaneous}
 
@@
 
+%ACKNOWLEDGMENTS are optional
+\section{Acknowledgments}
+This section is optional; it is a location for you
+to acknowledge grants, funding, editing assistance and
+what have you.  In the present case, for example, the
+authors would like to thank Gerald Murray of ACM for
+his help in codifying this \textit{Author's Guide}
+and the \textbf{.cls} and \textbf{.tex} files that it describes.
+
 %%TODO \cite{chamberlain04zpl, roth97stencils}
 
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,foropencl}
+\bibliographystyle{abbrv}
+\bibliography{foropencl}
 
+\balancecolumns
 \end{document}

### 2011-06-23 papers/GPUScA-2011/programming-model.tex
Changes on Thursday.

@@
 language and yet have direct access to GPU hardware.  We first examine how this
-programming model relates to work in other languages.
+programming model in Fortran relates to developments in other languages.
 
@@
 
-A number of previous efforts have exploited data-parallel programming
-at the language level to utilize novel architectures, particularly in
-previous decades during the reign of vector and massively parallel
-computers in the high performance computing world.  The origin of the
-array syntax that was adopted in Fortran 90 can be found in the APL
-language.  Fortran 90 differed from previous % \cite{iverson79apl}
-extensions of Fortran in that parallelism within whole-array
-operations was expressed at the expression level instead of via
-parallelism within explicit DO-loops (such as within IVTRAN for the
-Illiac IV).
+A number of previous efforts have exploited data-parallel programming at the
+language level to utilize novel architectures, particularly in previous decades
+during the reign of vector and massively parallel computers in high performance
+computing (HPC).  The origin of the array syntax adopted by Fortran in the 1990
+standard can be found in the APL language~\cite{iverson79apl}.  These additions
+to Fortran allowed parallelism to be expressed with whole-array operations at
+the expression level, instead of via parallelism within explicit DO-loops (such
+as within IVTRAN for the Illiac IV).
 
-The High Performance Fortran (HPF) extension of Fortran was
-proposed to add features to the language that would enhance the
-ability of compilers to emit fast parallel code for distributed and
-shared memory parallel computers\cite{koelbel94hpf}.  One of the
-notable additions to the language in HPF was syntax to specify the
-distribution of data structures amongst a set of parallel processors.
-HPF also introduced an alternative looping construct to the
-traditional DO-loop called {\tt FORALL} that was better suited for
-parallel compilation.  An additional keyword, {\tt INDEPENDENT}, was
-added to allow the programmer to indicate when the order of execution
-of the program (such as a sequence of loop iterations) can be flexible
-in order to allow parallel execution.  Interestingly, the parallelism
-features introduced in HPF did not exploit the new array features
-introduced in 1990 in any significant way, relying instead on explicit
-loop-based parallelism.  This was likely in order to support parallel
-programming that wasn't easily mapped onto a pure data-parallel model.
+The High Performance Fortran (HPF) extension of Fortran was proposed to add
+features to the language that would enhance the ability of compilers to emit
+fast parallel code for distributed and shared memory parallel
+computers\cite{koelbel94hpf}.  One of the notable additions to the language in
+HPF was syntax to specify the distribution of data structures amongst a set of
+parallel processors.  HPF also introduced an alternative looping construct to
+the traditional DO-loop called {\tt FORALL} that was better suited for parallel
+compilation.  An additional keyword, {\tt INDEPENDENT}, was added to allow the
+programmer to indicate when the loop contained no loop-order dependencies that
+allowed for parallel execution.  These constructs are similar to {\tt DO
+  CONCURRENT}, an addition to Fortran in 2008.  Interestingly, the parallelism
+features introduced in HPF did not exploit the new array features introduced in
+1990 in any significant way, relying instead on explicit loop-based parallelism.
+This restriction allowed the language to support parallel programming that
+wasn't easily mapped onto a pure data-parallel model.
 
-In some instances though, a purely data-parallel model is appropriate
-for part or all of the major computations within a program.  One of
-the systems where programmers relied heavily on higher level
-operations instead of explicit looping constructs was the Thinking
-Machines Connection Machine 5 (CM-5).  A common programming pattern
-used on the CM-5 that we exploit in this paper was to write
-whole-array operations from a global perspective in which computations
-are expressed in terms of operations over the entire array instead of
-a single local index.  The use of the array shift intrinsic functions
-(like {\tt CSHIFT}) were used to build computations in which arrays
-were combined by shifting the entire arrays instead of working based
-on local offsets from single indices.  A simple 1D example is one in
-which an element is replaced with the average of its own value with
-that of its two direct neighbors.  Ignoring boundary indices that wrap
-around, explicit indexing will result in a loop such as:
+In some instances though, a purely data-parallel model is appropriate for part
+or all of the major computations within a program.  One of the systems where
+programmers relied heavily on higher level operations instead of explicit
+looping constructs was the Thinking Machines Connection Machine 5 (CM-5).  A
+common programming pattern used on the CM-5 (that we exploit in this paper) was to
+write whole-array operations from a global perspective in which computations are
+expressed in terms of operations over the entire array instead of a single local
+index.  The use of the array shift intrinsic functions (like {\tt CSHIFT}) were
+used to build computations in which arrays were combined by shifting the entire
+arrays instead of working on local offsets based on single indices.  A simple
+1D example is one in which an element is replaced with the average of its own
+value and that of its two direct neighbors.  Ignoring boundary indices that
+wrap around, explicit indexing will result in a loop such as:
 
@@
 
-The whole-array model was attractive because it deferred
-responsibility for optimally implementing the computations to the
-compiler.  Instead of relying on a compiler to infer parallelism from
-a set of explicit loops, the choice for how to implement loops was
-left entirely up to the tool.  Unfortunately, this had two side
-effects that have limited broad acceptance of the whole-array
+The whole-array model was attractive because it deferred responsibility for
+optimally implementing the computations to the compiler.  Instead of relying on
+a compiler to infer parallelism from a set of explicit loops, the choice for how
+to implement loops was left entirely up to the tool.  Unfortunately, this had
+two side effects that have limited broad acceptance of the whole-array
 programming model in Fortran.  First, programmers must translate their
-algorithms into a set of global operations.  Finite difference
-stencils and similar computations are traditionally defined in terms
-of offsets from some central index.  Shifting, while conceptually
-analogous, can be awkward to think about for high dimensional stencils
-with many points.  Second, the semantics of these operations are such
-that all elements of an array operation are updated as if they were
-updated simultaneously.  In a program where the programmer explicitly
-manages arrays and loops, double buffering techniques and user managed
-temporaries are used to maintain these semantics.  When the compiler
-is responsible for managing this intermediate storage, it has
-historically proven that they are inefficient and generate code that
-requires far more temporary storage than really necessary.  This is
-not a flaw of the language constructs, but a sign of the lack of
-sophistication of the compilers with respect to their internal
-analysis to determine how to optimally generate this intermediate
+algorithms into a set of global operations.  Finite difference stencils and
+similar computations are traditionally defined in terms of offsets from some
+central index.  Shifting, while conceptually analogous, can be awkward to think
+about for high dimensional stencils with many points.  Second, the semantics of
+these operations are such that all elements of an array operation are updated as
+if they were updated simultaneously.  In a program where the programmer
+explicitly manages arrays and loops, double buffering techniques and user
+managed temporaries are used to maintain these semantics.  When the compiler is
+responsible for managing this intermediate storage, it has historically proven
+that they are inefficient and generate code that requires far more temporary
+storage than really necessary.  This is not a flaw of the language constructs,
+but a sign of the lack of sophistication of the compilers with respect to their
+internal analysis to determine how to optimally generate this intermediate
 storage.
 
-An interesting line of language research that grew out of the early work
-with HPF was that associated with the ZPL language work at the University
-of Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar
-global view of computation over arrays, but define their computations based
-on the local view of indices that participate in the update of each element of
-an array.
+An interesting line of language research that grew out of the early work with
+HPF was that associated with the ZPL language work at the University of
+Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar global
+view of computation over arrays, but define their computations based on regions,
+which provide a local view of the set of indices that participate in the update
+of each element of an array.  The Chapel parallel programming language being
+developed by Cray is directly influenced by HPF and
+ZPL~\cite{chamberlain_chapel}.
 
-The static analysis and source-to-source transformations used are very
-basic and only require the programmer to use a language subset that
-employs a data-parallel programming model.  In particular, it
-encourages use of four language features introduced in Fortran 90 ---
-array notation, shift functions, elemental functions, and pure
-procedures.  From these language constructs, we are able to easily
-transform them to a lower-level CUDA or OpenCL implementation.
+\subsection{Fortran Language Subset}
 
-%This paper uses four simple features of Fortran that form a language subset
-%that does not require any language extensions and can be easily transformed
-%to a lower-level implementation in either CUDA or OpenCL.
+The static analysis and source-to-source transformations used in this work are
+very basic and only require the programmer to use a language subset that employs
+a data-parallel programming model.  In particular, it encourages the use of
+array notation, elemental functions, and pure procedures.  From these language
+constructs, we are able to easily transform Fortran procedures to a lower-level
+OpenCL kernel implementation.
 
@@
 
-Fortran 90 introduced a rich array syntax that allows programmers to
-write statements in terms of whole arrays or subarrays, with
-data parallel operators to compute on the arrays.
-Array variables can be used in expressions
-based on whole-array operations.  For example, if {\tt A}, {\tt B},
-and {\tt C} are all arrays of the same rank and shape and {\tt s} is a
-scalar, then the statement
-
-%% Deleted by CER
-%%This has the benefit of avoiding explicit looping in the code and maintaining a high-level style that is closer to the original mathematical specification of the problem being solved.
-More importantly from a compilation perspective, this defers decisions about how to implement these whole-array operations to a compilation tool.
-Sophisticated loop analysis to identify parallelism within the sequential loop code are not necessary.
-When faced with a novel architecture such as modern GPUs that are ideally suited to data parallel programming models, the fit between Fortran arrays and these systems is quite clean.
-
-%% Deleted by CER
-%% A Fortran array combines a base pointer to memory with metadata that describes the array shape, stride and size.
-It should be noted that other languages that provide rich array data types and whole array operations may also be suitable targets for the transformations described in this paper.
+Fortran has a rich array syntax that allows programmers to write statements in
+terms of whole arrays or subarrays, with data parallel operators to compute on
+the arrays.  Array variables can be used in expressions based on whole-array
+operations.  For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays of the
+same rank and shape and {\tt s} is a scalar, then the statement
 
@@
 
-\noindent results in the element-wise sum of {\tt A} and {\tt s} times
-the elements of {\tt B} being stored in the corresponding elements of
-{\tt C}. The first element of {\tt C} will contain the value of the
-first element of {\tt A} added to the first element of {\tt c*B}.
-Note that no explicit iteration over array indices is needed and that
-the individual operators, plus, times, and assignment are applied by
-the compiler to individual elements of the arrays independently.  Thus
-the compiler is able to spread the computation in the example across
-any hardware threads under its control.
-
-%% NEW - CER
-%%% The use of array notation allows one to program in a data-parallel subset of Fortran.
-This style of programming makes use of array notation and pure and elemental functions to operate on array elements.
+\noindent results in the element-wise sum of {\tt A} and {\tt s} times the
+elements of {\tt B} being stored in the corresponding elements of {\tt C}. The
+first element of {\tt C} will contain the value of the first element of {\tt A}
+added to the first element of {\tt c*B}.  Note that no explicit iteration over
+array indices is needed and that the individual operators, plus, times, and
+assignment are applied by the compiler to individual elements of the arrays
+independently.  Thus the compiler is able to spread the computation in the
+example across any hardware threads under its control.
 
@@
 
-An elemental function consumes and produces scalar values, but can be applied
-to variables of array type such that the function is applied to each and every
+An elemental function consumes and produces scalar values, but can be applied to
+variables of array type such that the function is applied to each and every
 element of the array.  This allows programmers to avoid explicit looping and
-instead simply state that they intend a function to be applied to every
-element of an array in parallel, deferring the choice of implementation
-technique to the compiler.  Elemental functions are intended to be used for
-data parallel programming, and as a result must be side effect free and
-mandate an {\tt intent(in)} attribute for all arguments.
+instead simply state that they intend a function to be applied to every element
+of an array in parallel, deferring the choice of implementation technique to the
+compiler.  Elemental functions are intended to be used for data parallel
+programming, and as a result must be side effect free and mandate an {\tt
+intent(in)} attribute for all arguments.
 
-%% NEW - CER
-%%% Because elemental functions return scalar values and are free from side effects, the compiler is free to distribute the computation over any hardware processing elements available to it, such as the multiple cores and vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.
-
-%% NEW - CER
-For example, the basic array operation shown above could be refactored into
-an elemental function,
+For example, the basic array operation shown above could be refactored into an
+elemental function,
 
@@
 
-Note that while {\tt foo} is defined in terms of purely scalar
-quantities, it can be \emph{applied} to arrays as shown.  While this
-may seem like a trivial example, such simple functions may be composed
-with other elemental functions to perform powerful computations,
-especially when applied to arrays.  Our prototype tool transforms
-elemental functions to inline OpenCL functions.  Thus there is no
-penalty for usage of elemental functions and they provide a convenient
+Note that while {\tt foo} is defined in terms of purely scalar quantities, it
+can be \emph{applied} to arrays as shown.  While this may seem like a trivial
+example, such simple functions may be composed with other elemental functions to
+perform powerful computations, especially when applied to arrays.  Our prototype
+tool transforms elemental functions to inline OpenCL functions.  Thus there is
+no penalty for usage of elemental functions and they provide a convenient
 mechanism to express algorithms in simpler segments.
-
-%Also note that the plus and times operators shown in the data-statement
-%example can be seen as an elemental functions, as plus and times are defined
-%in terms of scalars but can be applied to whole arrays and can return an array
-%result.
 
@@
 
-Pure procedures, like elemental functions, must be free of side
-effects.  Unlike elemental functions that require arguments to have an
-{\tt intent(in)} attribute, they may change the contents of array
-arguments that are passed to them.  The absence of side effects
-removes ordering constraints that could restrict the freedom of the
-compiler to invoke pure functions out of order and possibly in parallel.
-Procedures and functions of this sort are also common in pure
-functional languages like Haskell, and are exploited by compilers in
-order to emit parallel code automatically due to their suitability for
-compiler-level analysis.
+Pure procedures, like elemental functions, must be free of side effects.  Unlike
+elemental functions that require arguments to have an {\tt intent(in)}
+attribute, they may change the contents of array arguments that are passed to
+them.  The absence of side effects removes ordering constraints that could
+restrict the freedom of the compiler to invoke pure functions out of order and
+possibly in parallel.  Procedures and functions of this sort are also common in
+pure functional languages like Haskell, and are exploited by compilers in order
+to emit parallel code automatically due to their suitability for compiler-level
+analysis.
 
-Since pure procedures don't have side effects they are candidates for
-running on accelerators in OpenCL.  Currently our prototype tool
-transforms pure procedures to OpenCL kernels that \emph{do not} call
-other procedures, except for elemental functions.
-
-\subsubsection*{Shift functions}
-
-Many array-based algorithms require the same operation to be performed
-on each element of the array using the value of that element and some small
-set of neighboring cells. As described above, this can often be done with
-shift operations.
-Fortran provides a set of shifting operators that allow programmers to
-define operations based on shifted arrays.  These intrinsic operators take
-an array, a dimension, and the amount by which it should be shifted (using
-the sign to indicate direction).  By defining operations on entire arrays
-based on a global view of them shifted relative to each other, programmers can
-avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
-analysis of the extent of the set of shifted arrays in a given expression
-allows analysis tools to determine the amount of temporary or buffer storage
-necessary to hold intermediate values during whole array operations.  With
-explicit loops, programmers must maintain this temporary storage manually.
-
-%% Deleted by CER
-%% Often programmers implement these operations that are local to each element within the inner-loop of a set of nested FOR- or DO-loops using offsets relative to the current array index.
-An alternative to this local-view of the algorithm is to take a global view and write the algorithm in terms of the whole array.
-For example, consider a 1D array in which we wish to subtract the $(i-1)$th element from the $i$th for all elements.
-One way to look at this is that we are subtracting the entire array shifted by one element from itself.
+Since pure procedures don't have side effects they are candidates for running on
+accelerators in OpenCL.  Currently our prototype tool transforms pure procedures
+to OpenCL kernels that \emph{do not} call other procedures, except for elemental
+functions.
 
@@
 
-Unlike OpenMP, we require \emph{limited} use of compiler directives, although
-two are used to enforce the semantics required by the programming
-model.  These directives are:
+Borrowing ideas from ZPL, we introduce a concept of a region to Fortran with a
+set of functions that allow programmers to work with subarrays in expressions.
+In Fortran, these functions return a copy of an existing array or array section.
+This is unlike ZPL, where regions are analogous to index sets and are used
+primarily for address resolution within an array without dictating storage
+related behavior.  The functions that we propose are similar in that they allow
+a programmer to deal with index regions that are meaningful to their algorithm,
+and automatically induce a halo (or ghost) cell pattern as needed in the
+implementation generated by the compiler, where the size of an array is
+increased to provide extra array elements surrounding the interior portion of
+the array.  This is critical to reducing the amount of code that the programmer
+is forced to write, as the halo cells are often where boundary errors and
+off-by-one errors may occur in code that is manually generated.
+
+Region functions are similar to the shift operator as they can be used to
+reference portions of the array that are shifted with respect to the interior
+portion.  However, unlike the shift operator, regions are not expressed in terms
+of boundary conditions and thus don't explicitly \emph{require} a knowledge of
+nor the application of boundary conditions locally.  Thus, as will be shown
+below, regions are more suitable for usage by OpenCL thread groups which access
+only local subsections of an array stored in global memory.
+
+ForOpenCL provides two new procedures that are defined in Fortran and are used
+in array-syntax operations.  Each procedure takes a integer array halo argument
+that specifies the number of ghost cells on either side of a region, for each
+dimension.  For example {\tt halo = [left, right, down, up]} specifies a halo
+for a two-dimensional region.  These functions are:
 
@@
 
-\item {\tt \!\$OFP CONTIGUOUS}: specifies that an dummy array variable is
-  contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.
+\item {\tt interior(array, halo)}: a pure function that returns a copy of
+  the interior portion of the array specified by halo.
 
-\item {\tt \!\$OFP KERNEL}: specifies that a pure subroutine can be
-  transformed to an OpenCL kernel.
+\item {\tt region(array, halo)}: a pure function that returns a copy of
+  the portion of the array specified by halo.
 
@@
 
-Borrowing ideas from ZPL, we introduce a concept of a region to
-Fortran with a set of functions that allow programmers to work with
-subarrays in expressions.  In Fortran, these functions return a copy
-of an existing array or array section.  This is unlike ZPL, where
-regions are analogous to index sets and are used primarily for address
-resolution within an array without dictating storage related behavior.
-The functions that we propose are similar in that they allow a
-programmer to deal with index regions that are meaningful to their
-algorithm, and automatically induce a halo (or ghost) cell pattern as
-needed in the implementation generated by the compiler, where the size
-of an array is increased to provide extra array elements surrounding
-the interior portion of the array.  This is critical to reducing the
-amount of code that the programmer is forced to write, as the halo cells
-are often where boundary errors and off-by-one errors may occur in code
-that is manually generated.
-
-Region functions are similar to the shift operator as they can be used
-to reference portions of the array that are shifted with respect to
-the interior portion.  However, unlike the shift operator, regions are
-not expressed in terms of boundary conditions and thus don't
-explicitly \emph{require} a knowledge of nor the application of
-boundary conditions locally.  Thus, as will be shown below, regions
-are more suitable for usage by OpenCL thread groups which access only
-local subsections of an array stored in global memory.
-
-Three new procedures (in addition to the intrinsic shift function) are
-defined in Fortran that are used in array-syntax operations.  Each
-procedure takes a integer array halo argument that specifies the
-number of ghost cells on either side of a region, for each dimension.
-For example {\tt halo = [left, right, down, up]} specifies a halo for
-a two-dimensional region.  These functions are:
-
-\begin{itemize}
-
-\item {\tt transfer\_halo(array, halo)}: an impure subroutine that
-  exchanges halo regions between nodes using MPI or Fortran
-  coarrays. Array is the memory stored on the node representing the
-  node-local portion of a virtual global array. Not used in this work.
-
-\item {\tt interior(array, halo)}: a pure function that returns a copy of
-  the interior portion of the array specified by halo. Array is
-  local to a node.
-
-\item {\tt region(array, halo)}: a pure function that returns a copy of
-  the portion of the array specified by halo.  Array is local to a node.
-
-\end{itemize}
-
-It should be noted that the two functions {\tt interior} and {\tt
-  region} are pure and thus can be called from within a pure kernel
-procedure.  These two functions are part of the language recognized by
-the compiler and though the two functions return a copy of a portion
-of an array \emph{semantically}, the compiler is not forced to
-actually make a copy and is free to enforce copy semantics through
-other means.
+It should be noted that the two functions {\tt interior} and {\tt region} are
+pure and thus can be called from within a pure kernel procedure.  These two
+functions are part of the language recognized by the compiler and though the two
+functions return a copy of a portion of an array \emph{semantically}, the
+compiler is not forced to actually make a copy and is free to enforce copy
+semantics through other means.  In addition to these two new functions,
+ForOpenCL provides the compiler directive, {\tt \!\$OFP KERNEL}, which specifies
+that a pure subroutine can be transformed to an OpenCL kernel.
 
@@
 
-%% MJS - mention that these features were designed for parallelism in
-%% the first place?
+Data parallelism has been called collection-oriented programming by
+Blelloch~\cite{blelloch90}.  As the {\tt cshift} function and the array-valued
+expressions all semantically return a value, this style of programming is also
+similar to functional programming (or value-oriented programming).
 
-%% Deleted by CER
-%% An example of this style of programming in Fortran is shown in
-%
-%{\small
-%\begin{verbatim}
-%  Bz = Bz &
-%     + dt*(cshift(Ex,dim=2,shift=+1)-Ex)/dy &
-%     - dt*(cshift(Ey,dim=1,shift=+1)-Ey)/dx
-%\end{verbatim}
-%}
+Unfortunately, this style of programming has never really caught on because when
+array syntax was first introduced in Fortran, performance of these features was
+relatively poor and thus programmers shied away from using array syntax (even
+recently, some are actively counseling against its usage because of performance
+issues~\cite{Levesque:SC08}).  Thus the Fortran community was caught in a
+classic ``chicken-and-egg'' conundrum: (1) programmers didn't use it because it
+was slow; and (2) compilers vendors didn't improve it because programmers didn't
+use it.  A goal of this paper is to demonstrate that parallel programs written
+in this style of Fortran are maintainable and can achieve good performance on
+accelerator architectures.
 
-%%This example is a solution to Maxwell's equations for the $z$ component of the magnetic field using Fortran array syntax.
-Note that there are no explicit loops in this example.
-The operators {\tt +}, {\tt -}, and {\tt *} are applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt Ex}, and {\tt Ey}, individually.
-
-%Blelloch~\cite{blelloch90,rajopadhye93}.
-
-Data parallelism has been called collection-oriented programming by
-Blelloch~\cite{blelloch90}.  As the {\tt cshift}
-function and the array-valued expressions all semantically return a
-value, this style of programming is also similar to functional
-programming (or value-oriented programming).
-
-%%~\cite{simonpeytonjones}).  The heart of the solution
-%to Maxwell's equations is the statement shown in
-%Listing~\ref{lst:dpexample} and five similar, simple equations.
-
-%% NEW - CER
-%%% The power in this notation is that the compiler is free to distribute the computation in these expressions over any hardware processing elements available to it, such as the vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.
-If the arrays are declared as coarrays, this includes spreading the computation over \emph{all of the nodes in   a cluster as well.}  In this case, communication occurs within the \texttt{cshift} functions, though the compiler is free to overlap communication and computation by scheduling the communication early and performing the computation on the interior of the arrays while waiting for the communication to complete.
-
-%% NEW - CER
-%Complete and very concise and elegant programs can be built with procedures
-%similar to the example shown above. To aid this effort, Fortran
-%supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
-%\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
-%functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
-%functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
-%\texttt{PRODUCT}, ...).  To this set we add region functions described above.
-
-%% Deleted by CER
-%% This style of programming meets the requirements we have set for a programming model for developing applications suitable for acceleration.
-It allows the programmer to program at a very-high level of abstraction while providing the compiler with maximum flexibility in targeting the application for a particular hardware architecture.
-The data parallel programming model simultaneously meets the seemingly conflicting goals of maintainability, portability, and performance.
-
-Unfortunately, this style of programming has never really caught on
-because when Fortran 90 was first introduced, performance of these
-features was relatively poor and thus programmers shied away from
-using array syntax (even recently, some are actively counseling against its
-usage because of performance issues~\cite{Levesque:SC08}).  Thus the
-Fortran community was caught in a classic ``chicken-and-egg''
-conundrum: (1) programmers didn't use it because it was slow; and (2)
-compilers vendors didn't improve it because programmers didn't use it.
-A goal of this paper is to demonstrate that parallel programs written
-in this style of Fortran are maintainable and can achieve good
-performance on accelerator architectures.
-
-\subsection{Current Limitations}
+\subsection{Restrictions}
 
 Only pure Fortran procedures are transformed into OpenCL kernels.  The
-programmer must explicitly call these kernels from Fortran using the
-ForOpenCL library described below.  It is also possible, using ROSE,
-to modify the calling site so that the entire program can be
-transformed, but this functionality is outside the scope of this
-paper.  Here we specifically examine transforming Fortran procedures
-to OpenCL kernels.  Because OpenCL support is relatively new to ROSE,
-some generated code must be modified.  For example, the {\tt
+programmer must explicitly call these kernels from Fortran using the ForOpenCL
+library described below.  It is also possible, using ROSE, to modify the calling
+site so that the entire program can be transformed, but this functionality is
+outside the scope of this paper.  Here we specifically examine transforming
+Fortran procedures to OpenCL kernels.  Because OpenCL support is relatively new
+to ROSE, some generated code must be modified.  For example, the {\tt
   \_\_global} attribute for kernel arguments was added by hand.
 
-It is assumed that memory for all arrays reside on the device.  The
-programmer must copy memory to and from the device.  In addition,
-array size (neglecting ghost cell regions) must be multiples of the
-global OpenCL kernel size.
-
-%% Duplicated CER
-%%Only elemental functions may be called from kernel functions.  These include Fortran functions that have an OpenCL analog and user-defined elemental functions.
+It is assumed that memory for all arrays reside on the device.  The programmer
+must copy memory to and from the device.  In addition, array size (neglecting
+ghost cell regions) must be multiples of the global OpenCL kernel size.
 
 A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
-pure and all array variables must be declared as contiguous.  A kernel
-procedure may not call other procedures except for limited intrinsic
-functions (primarily math), user-defined elemental functions, and the
-{\tt interior} and {\tt region} functions.  We plan to address
-non-contiguous arrays (such as those that result from strided access)
-by mapping array strides to gather/scatter-style memory accessors.
+pure and all array variables must be declared as contiguous.  A kernel procedure
+may not call other procedures except for limited intrinsic functions (primarily
+math), user-defined elemental functions, and the {\tt interior} and {\tt region}
+functions.  Future work will address non-contiguous arrays (such as those that result
+from strided access) by mapping array strides to gather/scatter-style memory
+accessors.

### 2011-06-26 papers/GPUScA-2011/shallow-water.tex
Nearing final version.

--- 2011-06-20 papers/shallow-water.tex
@@
 for simulating smooth shallow-water flows. In the following explanation,
-$U$ is the conserved state variable at the center of the cell . This
-state variable, $U=(h,hu,hv)$ in the first term in the equations
-above. $F$ is the flux quantity that crosses the boundary of the cell
+$U$ is the conserved state variable at the center of the cell. This
+state variable, $U$ $=(h,hu,hv)$ in the first term in the equations
+below. $F$ is the flux quantity that crosses the boundary of the cell
 and is subtracted from one cell and added to the other. The remaining
@@
 
-\subsection{Serial code examples}
+\subsection{Fortran implementation}
 
-The coding style to implement the serial version of the shallow water model
-is now shown.  The Fortran kernel procedure {\tt wave\_advance} 
-is declared as:
+Selected portions of the data-parallel implementation of the shallow water model
+are now shown.  This code serves as input to the ForOpenCL transformations
+described in the next section.  The interface for the Fortran kernel procedure {\tt
+  wave\_advance} is declared as:
 
@@
 \begin{verbatim}
-pure subroutine wave_advance(H,U,V,dx,dy,dt)
-  real, dimension(:,:),intent(inout):: H,U,V
+subroutine wave_advance(dx,dy,dt,H,U,V,oH,oU,oV)
+  !$OFP PURE, KERNEL   :: wave_advance
   real, intent(in)     :: dx,dy,dt
-  !$OFP CONTIGUOUS :: H,U,V
-  !$OFP KERNEL :: wave_advance
+  real, dimension(:,:) :: H,U,V,oH,oU,oV
+  contiguous  :: H,U,V,oH,oU,oV
+  intent(in)  :: H,U,V
+  intent(out) :: oH,oU,oV
+  target      :: oH,oU,oV
 end subroutine
@@
 \noindent
-where {\tt H, U}, and {\tt V} are state variables for height and
-$x$ and $y$ momentum respectively.  The \emph{OFP} compiler directives
-{\tt CONTIGUOUS} and {\tt KERNEL} are added because contiguous is Fortran
-2008 syntax (not in current compilers) and kernel is an \emph{OFP} extension.
+where {\tt dx, dy, dt} are differential quantities in space $x, y$ and time $t$,
+{\tt H, U}, and {\tt V} are state variables for the height and $x$ and $y$
+momentum respectively, and {\tt oH, oU, oV} are corresponding output arrays used
+in the double buffering scheme.  The \emph{OFP} compiler directive attributes
+{\tt PURE} and {\tt KERNEL} indicate that the procedure {\tt wave\_advance} is
+to be transformed as an OpenCL kernel and that it must be pure, other than for any
+pointers used to reference interior regions of the output arrays.
 
-Temporary arrays are required for the interior copies {\tt iH, iU}, and
-{\tt iV} of the state variables and for the quantities {\tt Hx, Hy, Ux,
-Vx}, and {\tt Vy} defined on cell faces.  These temporary arrays are declared as,
+Temporary arrays are required for the quantities {\tt Hx, Hy, Ux, Vx}, and {\tt
+  Vy,} that are defined on cell faces.  Also, the pointer variables, {\tt pH,
+  pU,} and {\tt pV}, are needed to access and update interior regions of the
+output arrays.  As these pointers are assigned to the arrays {\tt oH, oU, oV},
+these output arrays must have the target attribute, as shown in the interface
+above.  The temporary arrays and array pointers are declared as,
 
@@
 \begin{verbatim}
-real, allocatable, dimension(:,:) :: iH,iU,iV
-real, allocatable, dimension(:,:) :: Hx,Hy,Ux
-real, allocatable, dimension(:,:) :: Uy,Vx,Vy
+real, allocatable, dimension(:,:) :: Hx, Hy, Ux
+real, allocatable, dimension(:,:) :: Uy, Vx, Vy
+real, pointer,     dimension(:,:) :: pH, pU, pV
 \end{verbatim}
@@
 
-Halo variables for the interior and cell faces are
+Halo variables for the interior and the cell faces are declared and defined as
 
@@
 \begin{verbatim}
-integer, dimension(4) :: halo,face_lt,face_rt
-integer, dimension(4) :: face_up,face_dn
+integer, dimension(4) :: face_lt, face_rt, halo
+integer, dimension(4) :: face_up, face_dn
+
 halo    = [1,1,1,1]
@@
 \noindent
-Note that the halos for the four faces each have a 0 in the
-definition. Thus the returned array copy will have size that is larger
-than the interior regions that use {\tt [1,1,1,1]}.  This is because
-there is one more cell face quantity than there are cells in a given
+Note that the halo definitions for the four faces each have a 0 in the
+initialization. Thus the returned array copy will have a size that is larger
+than any interior region that uses the full halo {\tt [1,1,1,1]}.  This is
+because there is one more cell face quantity than there are cells in a given
 direction.
 
-The first Lax-Wendroff step updates state variables on the cell
-faces, for example,
+The first Lax-Wendroff step updates state variables on the cell faces.  Assignment
+statements like the following,
 
@@
 \begin{verbatim}
-Hx = 0.5*(region(H,face_lt)+ &
-          region(H,face_rt)) &
+Hx = 0.5*( region_cpy(H,face_lt) +   &
+           region_cpy(H,face_rt) )   &
      + (0.5*dt/dx)                   &
-   * (region(U,face_lt)-region(U,face_rt))
+     * (region_cpy(U,face_lt) - region_cpy(U,face_rt))
 \end{verbatim}
@@
 \noindent
-updates the $x$-direction height variable.  The second step then uses
-these face variables to update the interior region, for example,
+are used to calculate these quantities.  This equation updates the array for the
+height in the $x$-direction.  The second step then uses these face quantities to
+update the interior region, for example,
 
@@
 face_dn = [0,0,0,1];  face_up = [0,0,1,0]
-iH = iH + (dt/dx) * ( region(Ux, face_lt) - &
-                      region(Ux, face_rt) ) &
-        + (dt/dy) * ( region(Vy, face_dn) - &
-                      region(Vy, face_up) )
+
+pH = region_ptr(oH, halo)
+
+pH = region_cpy(H, halo)
+       + (dt/dx) * ( region_cpy(Ux, face_lt) -   &
+                     region_cpy(Ux, face_rt) )   &
+       + (dt/dy) * ( region_cpy(Vy, face_dn) -   &
+                     region_cpy(Vy, face_up) )
 \end{verbatim}
@@
 
-These simple code segments show how the shallow water model is
-implemented in standard Fortran (2003) using the data-parallel
-programming model described above.  The resulting code is simple,
-concise, and easy to understand.  However it does \emph{not}
-necessarily perform well because of the temporary array variables,
-especially those produced by the {\tt region} function.  This is
-generally true of algorithms that use Fortran shift functions as well,
-as some Fortran compilers (e.g., gfortran) do not generate optimal
-code for shifts.  We note (as shown below) that these temporary array
-copies are replaced by scalars in the transformed Fortran code so
-there is no performance penalty for using data-parallel statements as
-outlined.
+These simple code segments show how the shallow water model is implemented in
+standard Fortran using the data-parallel programming model described above.  The
+resulting code is simple, concise, and easy to understand.  However it does
+\emph{not} necessarily perform well because of the temporary array variables,
+especially those produced by the function {\tt region\_cpy}.  This is generally true
+of algorithms that use Fortran shift functions as well, as some Fortran
+compilers (e.g., gfortran) do not generate optimal code for shifts.  We note (as
+shown below) that these temporary array copies are replaced by scalars in the
+transformed Fortran code so there are no performance penalties for using
+data-parallel statements as outlined.  However, there is an increased memory
+cost due to the double buffering required by the kernel execution semantics.

### 2011-06-26 papers/GPUScA-2011/source-to-source.tex
Nearly final.

--- 2010-12-20 opencl/papers/source-to-source.tex
@@
 
-This section describes the transformations that take Fortran elemental and
-pure procedures as input and generate OpenCL kernels.
+This section provides an brief overview of the ForOpenCL transformations that take
+Fortran elemental and pure procedures as input and generate OpenCL code.  Recall
+that elemental functions are transformed to inline OpenCL functions and that
+subroutines with the {\tt PURE} and {\tt KERNEL} compiler directive attributes
+are transformed to OpenCL kernels.
 
@@
 
-OpenCL~\cite{opencl08} is an open language standard for developing
-applications for accelerators.
-The kernels
-are run by calling a C runtime library from the OpenCL host (normally the
-CPU).  Efforts to standardize a C++ runtime are underway and Fortran
-interfaces to the C runtime are described later.
+OpenCL~\cite{opencl08} is an open-language standard for developing applications
+targeted for GPUs, as well as for multi-threaded applications targeted for
+multi-core CPUs.  The kernels are run by calling a C runtime library from the
+OpenCL host (normally the CPU).  Efforts to standardize a C++ runtime are
+underway and Fortran interfaces to the C runtime (distributed in the ForOpenCL
+library).
 
@@
 groups are used to run an OpenCL kernel concurrently on several processor
-elements on the OpenCL device (often a GPU).  Consider a data-parallel
-statement written in terms of an elemental function as discussed above.  The
-act of running an OpenCL kernel can be thought of as having a particular
-thread assigned to each instance of the call to the elemental function as it
-is mapped across the arrays in the data-parallel statement.  In practice,
-these threads are packaged into thread groups when they are run on the device
-hardware.
+elements on the OpenCL device (often a GPU).  Consider a data-parallel statement
+written in terms of an elemental function as discussed above.  The act of
+running an OpenCL kernel can be thought of as having a particular thread
+assigned to each instance of the call to the elemental function as it is mapped
+across the arrays in the data-parallel statement.  In practice, these threads
+are packaged into thread groups when they are run on the device hardware.
 
-Device memory is separated hierarchically.  A thread instance has access to
-its own memory, thread groups to OpenCL local memory, and all thread groups
-have access to OpenCL global memory.  When multiple members of a thread group
-access the same memory elements (if {\tt region} or {\tt shift}
-functions are called), for performance reasons it is often best if global
-memory accessed by a thread group is copied into local memory.
+Device memory is separated hierarchically.  A thread instance has access to its
+own thread memory (normally a set of registers), threads in a thread group to
+OpenCL local memory, and all thread groups have access to OpenCL global memory.
+When multiple members of a thread group access the same memory elements (for
+example in the use of the {\tt region\_cpy} function in the calculation of face
+variable quantites shown above), for performance reasons it is often best that
+global memory accessed and shared by a thread group be copied into local memory.
 
@@
 
-We emphasize that the hierarchical distribution of memory used on the OpenCL
-device shown in Figure~\ref{fig:cl-memory} can be extended to include memory
-across MPI nodes as well.  In this case, the virtual global array is
-represented by the background layer (with its halo) and its partitions stored
-in the 4 MPI nodes shown in the foreground.
+We point out that the hierarchical distribution of memory used on the OpenCL
+device shown in Figure~\ref{fig:cl-memory} is similar to the distribution of
+memory across MPI nodes in an MPI application.  In the case of MPI, the
+virtual global array is represented by the background layer (with its halo) and
+partitions of the global array are stored in the 4 MPI nodes shown in the foreground.
 
-Halo regions are constrained semantically so that they can not be written to
-by an OpenCL kernel because the \emph{region} and \emph{interior} functions
-return copies of the global memory.  Thus once memory for a halo region has
-been transferred into global device memory by all of the host nodes running
-MPI (before the OpenCL kernel is run), memory is in a consistent state so that
-the kernels are free to read from global device memory.  Because the local
-memory is a copy, it functions as a software cache for the local thread group.
-Thus the compiler must insert OpenCL barriers at proper locations in the code
-to insure that all threads have written to the local memory cache before any
-thread can read from the cache.  On exit from a kernel, the local memory cache
-is copied back to global memory for all \emph{interior} regions leaving global
-memory in a consistent state again.
+Halo regions obtained via the {\tt region\_cpy} function (used with intent(in)
+arrays) are constrained semantically so that they can not be written to by an
+OpenCL kernel.  The {\tt region\_cpy} function returns a copy of the region of
+the array stored in global memory and places it in local memory shared by
+threads in a thread group.  Thus once memory for an array has been transferred
+into global device memory by the host (before the OpenCL kernel is run), memory
+is in a consistent state so that all kernel threads are free to read from global
+device memory.  Because the local memory is a copy, it functions as a software
+cache for the local thread group.  Thus the compiler must insert OpenCL barriers
+at proper locations in the code to insure that all threads have written to the
+local memory cache before any thread can start to read from the cache.  On exit
+from a kernel, any local memory explicitly stored in register variables by the
+compiler (memory accessed via the {\tt region\_cpy} function) is copied back to
+global memory for all intent(out) arrays.  Recall that a thread may only write to
+its own intent(out) array element, thus there are no race conditions when updating
+intent(out) arrays.
 
@@
 
-\section{Transformation examples}
+\subsection{Transformation examples}
 
-This section outlines the OpenCL equivalent syntax for portions of
-the Fortran shallow-water code described in
-Section~\ref{sec:shallow-water}.  The notation uses uppercase for
-arrays and lowercase for scalar quantities.  Interior variables are
-denoted by an {\tt i} preceding the array name.  For example, {\tt iH =
-  interior(H, halo)} is a local copy of interior region of array {\tt
-  H} (representing height) in the shallow water model.
+This section outlines the OpenCL equivalent syntax for portions of the Fortran
+shallow-water code described in Section~\ref{sec:shallow-water}.  The notation
+uses uppercase for arrays and lowercase for scalar quantities.  Variables
+temporarily storing quantities for updated output arrays (declared as pointers
+in Fortran) are denoted by a {\tt p} preceding the array name.  For example, the
+Fortran statement {\tt pH = region\_ptr(oH, halo)} is transformed as a scalar
+variable declaration representing a single element in the output array {\tt oH}.
 
-\subsubsection{Interior and region functions}
+\subsubsection{Region function}
 
-While the serial versions of the interior and regions functions return an
-array copy, in OpenCL, these functions return a scalar quantity based on the
+While the Fortran version of the {\tt region\_cpy} function semantically returns
+an array copy, in OpenCL this function returns a scalar quantity based on the
 location of a thread in a thread group and the relationship of its location to
-the array copy in local memory.  Because we assume there is a thread for every
-element in the interior, the array index is just the thread index adjusted for
-the halo.  Thus {\tt interior} and {\tt region} are just inline OpenCL
-functions provided by the ForOpenCL library.
+the array copy transferred to local memory.  Because we assume there is a thread
+for every element in the interior, the array index is just the thread index
+adjusted for the size of the halo.  Thus {\tt region\_cpy} is just an inline
+OpenCL function and is provided by the ForOpenCL library.
 
@@
 Fortran kernel procedures have direct correspondence with OpenCL equivalents.
-For example, {\tt wave\_advance} is transformed as {\tt \_\_kernel void
-  wave\_advance(\_\_ global float * H, ..., float dt);}.  The global state
-variables have local equivalents, e.g., {\tt \_\_local float
-  H\_local[MAX\_LOCAL\_SIZE]}, declared with the appropriate size and are
-copied to local memory by the compiler with inlined library functions.  The
+For example, the {\tt wave\_advance} interface declaration is transformed as
+
+{\small
+\begin{verbatim}
+__kernel void
+wave_advance(float dx, ..., __global float * H, ...);
+\end{verbatim}
+}
+
+The intent(in) arrays have local equivalents that are stored in local memory
+and are declared by, for example,
+{\small
+\begin{verbatim}
+__local float H_local[LOCAL_SIZE];
+\end{verbatim}
+}
+\noindent
+These local arrays are declared with the appropriate size and are
+copied to local memory by the compiler with an inlined library function.  The
 array temporaries defined on cell faces are declared similarly while interior
-variables are simple scalars, e.g., {\tt float iH;}.  Region variables cannot
-be scalar objects because regions may be shifted and thus \emph{shared} by
-threads within a thread group.
+pointer variables are simple scalars, e.g., {\tt float pH}.  Intent(in) array
+variables cannot be scalar objects because regions may be shifted and thus
+\emph{shared} by threads within a thread group.
 
@@
 Array syntax transforms nearly directly to OpenCL code.  For example, interior
-variables are particularly straightforward as they are scalar quantities in
+pointer variables are particularly straightforward as they are scalar quantities in
 OpenCL,
-
 {\small
 \begin{verbatim}
-iH = iH + (dt/dx) * (region(Ux, face_lt) - 
-                     region(Ux, face_rt) )
-        + (dt/dy) * (region(Vy, face_dn) - 
-                     region(Vy, face_up) );
+pH = region_cpy(H, halo)
+       + (dt/dx) * ( region_cpy(Ux, face_lt) -
+                     region_cpy(Ux, face_rt) )
+       + (dt/dy) * ( region_cpy(Vy, face_dn) -
+                     region_cpy(Vy, face_up) );
 \end{verbatim}
 }
-
-\noindent Region variables are more complicated because they are
-arrays.
-
+\noindent
+Allocated variables are more complicated because they are arrays.
 {\small
@@
 }
-
 \noindent

### 2011-06-26 papers/GPUScA-2011/performance.tex
Copied IPDPS 2011 paper to GPUScA 2011 workshop (in Galveston)
Near final version.

--- 2010-12-20 opencl/papers/performance.tex
@@
 
-Performance measurements were made comparing the transformed code with
-different versions of the serial shallow-water code.  The serial
-versions include two separate Fortran versions:
-one using data-parallel notation and the other using explicit looping
-constructs.
-We also compare with a hand-written OpenCL implementation
-that was optimized for local memory usage (no array temporaries).  The
-accelerated measurements were made using an NVIDIA Tesla C2050 (Fermi) cGPU with
-2.625 GB GDDR5 memory, and 448 processor cores.  The serial
-measurements were made using an Intel Xeon X5650 hexacore CPU with 96 GB of RAM
-running at 2.67 GHz.  The compilers were gfortran and gcc version 4.4.3
-with an optimization level of -O3.
+Performance measurements were made comparing the transformed code with different
+versions of the serial shallow-water code.  The serial versions included two
+separate Fortran versions: one using data-parallel notation and the other using
+explicit looping constructs.  We also compared with a hand-written OpenCL
+implementation that was optimized for local memory usage (no array temporaries).
+The accelerated measurements were made using an NVIDIA Tesla C2050 (Fermi) cGPU
+with 2.625 GB GDDR5 memory, and 448 processor cores.  The serial measurements
+were made using an Intel Xeon X5650 hexacore CPU with 96 GB of RAM running at
+2.67 GHz.  The compilers were gfortran and gcc version 4.4.3 with an
+optimization level of -O3.
 
@@
 
-The performance comparison is shown in Figure~\ref{fig:cl-performance} for
-varying array sizes (of the state variables).  The time represents an average
-of 100 iterations of the outer time-advance loop calling the OpenCL kernel.
-This tight loop keeps the OpenCL kernel supplied with threads to take
-advantage of potential latency hiding by the NVIDIA GPU.  Any serial code
-within this loop would reduce the measured values.
+%%The performance comparison is shown in Figure~\ref{fig:cl-performance} for varying array sizes (of the state variables).
+The time represents an average of 100 iterations of the outer time-advance loop calling the OpenCL kernel.
+This tight loop keeps the OpenCL kernel supplied with threads to take advantage of potential latency hiding by the NVIDIA GPU.
+Any serial code within this loop would reduce the measured values.
 
-As can be seen in Figure~\ref{fig:cl-performance}, the transformed code achieves very
-good results.  It is up to 40 times faster than the best serial Fortran code and
-within 25\% of the hand-optimized OpenCL kernel.
+Several timing measurements were made by varying the size of the array state
+variables.  An average time was obtained by executing 100 iterations of the outer
+time-advance loop that called the OpenCL kernel. This tight loop kept the OpenCL
+kernel supplied with threads to take advantage of potential latency hiding by
+the NVIDIA GPU.  Any serial code within this loop (not present in this study)
+would have reduced the measured values.
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3in]{cl-performance.pdf}
-\caption{Performance comparison for varying array size.}
-\label{fig:cl-performance}
-\end{figure}
+The transformed code achieved very good results.  In all instances, the
+performance of the transformed code was within 25\% of the hand-optimized OpenCL
+kernel.  Most of the extra performance of the hand-optimized code can be
+attributed to the absence of array temporaries and to packing the three state
+variables {\tt H, U}, and {\tt V} into a single vector datatype.
+
+While we did not have an OpenMP code for multi-core comparisons, the transformed
+OpenCL code on the NVIDIA C2050 was up to 40 times faster than the best serial
+Fortran code executing on the host CPU.
+
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=3in]{cl-performance.pdf}
+%\caption{Performance comparison for varying array size.}
+%\label{fig:cl-performance}
+%\end{figure}

### 2011-06-26 papers/GPUScA-2011/foropencl.tex
Version to ship to Matt.  Needs a couple of bib entries.

@@
 
-\title{Exploiting Array Syntax in Fortran for Accelerator Programming}
+\title{ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming}
 
@@
        \affaddr{Los Alamos National Laboratory}\\
-       \affaddr{XTD-1, MS T082}\\
-       \affaddr{Los Alamos, NM 87545}\\
+       %\affaddr{XTD-1, MS T082}\\
+       %\affaddr{Los Alamos, NM 87545}\\
        \email{weseloh@lanl.gov}
@@
        \affaddr{Los Alamos National Laboratory}\\
-       \affaddr{XCP-2, MS T086}\\
-       \affaddr{Los Alamos, NM 87545}\\
+       %\affaddr{XCP-2, MS T086}\\
+       %\affaddr{Los Alamos, NM 87545}\\
        \email{brobey@lanl.gov}
@@
   Emerging GPU architectures for high performance computing are well suited to a
-data-parallel programming model.  This paper presents a simple programming
-methodology that allows programmers to take advantage of these emerging systems.
-We use array constructs in Fortran 90 to show how this infrequently exploited,
-standardized language feature is easily transformed to lower-level accelerator
-code.  These transformations are based on a simple mapping from Fortran 90 to
-OpenCL.  We demonstrate, using a stencil code solving the shallow-water fluid
-equations, that the performance of our compiler-generated transformations is
-comparable with that of hand-optimized OpenCL code.
+  data-parallel programming model.  This paper presents preliminary work
+  examining a programming methodology that provides Fortran programmers with access
+  to these emerging systems.  We use array constructs in Fortran to
+  show how this infrequently exploited, standardized language feature is easily
+  transformed to lower-level accelerator code.  The transformations in ForOpenCL are based
+  on a simple mapping from Fortran to OpenCL.  We demonstrate, using a
+  stencil code solving the shallow-water fluid equations, that the performance
+  of the ForOpenCL compiler-generated transformations is comparable with that of
+  hand-optimized OpenCL code.
 \end{abstract}
@@
 % A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
+\category{D.3.3}{Language Constructs and Features}{Concurrent programming structures}
 
@@
 \section{Acknowledgments}
-This section is optional; it is a location for you
-to acknowledge grants, funding, editing assistance and
-what have you.  In the present case, for example, the
-authors would like to thank Gerald Murray of ACM for
-his help in codifying this \textit{Author's Guide}
-and the \textbf{.cls} and \textbf{.tex} files that it describes.
+This research was supported by the Departement of Energy, Office of Science.
 

### 2011-06-26 papers/GPUScA-2011/intro.tex
Version to ship to Matt.  Needs a couple of bib entries.

@@
 same neighborhood~\cite{hpcwire11manycore}.  Unfortunately the performance that
-many of the new accelerator architectures offer comes at a cost.
-Architectural changes are trending toward multiple heterogeneous cores and less
-of a reliance on superscalar instruction level parallelism and hardware managed
-memory hierarchies (such as traditional caches).  These changes place a heavy
-burden on application programmers as they are forced to adapt to these new
-systems.  An especially challenging problem is not only how to program to these
-new architectures (considering the massive scale of concurrency available), but
+many of the new accelerator architectures offer comes at a cost.  Architectural
+changes are trending toward multiple heterogeneous cores and less of a reliance
+on superscalar instruction level parallelism and hardware managed memory
+hierarchies (such as traditional caches).  These changes place a heavy burden on
+application programmers as they are attempt to adapt to these new systems.  An
+especially challenging problem is not only how to program to these new
+architectures --- considering the massive scale of concurrency available --- but
 also how to design programs that are portable across the changing landscape of
-computer architectures with unique memory systems and programming models.  How
-can a programmer easily write one program that can perform well on both a
-conventional multicore CPU and a GPU (or one of the emerging many-core
-architectures)?  A directive-based approach, such as OpenMP or the Accelerator
-programming model from the Portland Group~\cite{pgi10accelerator}, is one
-solution to this problem.
+computer architectures.  How does a programmer write one program that can
+perform well on both a conventional multicore CPU \emph{and} a GPU (or another
+of the emerging many-core architectures)?
 
-In this paper we take a somewhat different approach.  A common theme amongst the
-new processors is the emphasis on data-parallel programming.  This model is well
-suited to the emerging architectures that are based on either vector processing
-or massively parallel collections of simple cores.  The recent CUDA and OpenCL
-programming languages are are intended to support this programming model,
+A directive-based approach, such as OpenMP or the Accelerator programming model
+from the Portland Group~\cite{pgi10accelerator}, is one solution to this
+problem.  However, in this paper we take a somewhat different approach.  A
+common theme amongst the new processors is the emphasis on data-parallel
+programming.  This model is well suited to the emerging architectures that are
+based on either vector processing or massively parallel collections of simple
+cores.  The recent CUDA and OpenCL programming languages are are intended to
+support this programming model.
 
@@
 
-We demonstrate that the data-parallel features of Fortran map surprisingly well
-onto GPUs when transformed to OpenCL kernels.  These features include: array
-syntax (e.g., assignment and binary operators), array constructs like {\tt
-  WHERE} and {\tt DO CONCURRENT}, and the use of pure and elemental functions.
-In addition, we provide new functions that explicitly take advantage of the
-stencil geometry of the problem domain we are addressing.  Note that this subset
-of the Fortran language is implicitly parallel.  This programming model
-\emph{does not require explicit declaration of parallelism within the program.}
-In addition, programs are expressed using entirely standard Fortran so it can be
-compiled for and executed on a single core without concurrency.
+We demonstrate that a subset of Fortran map surprisingly well onto GPUs when
+transformed to OpenCL kernels.  This data-parallel subset includes: array syntax
+using assignment statements and binary operators, array constructs like {\tt
+  WHERE}, and the use of pure and elemental functions.  In addition, we provide
+new functions that explicitly take advantage of the stencil geometry of the
+problem domain we consider.  Note that this subset of the Fortran language is
+implicitly parallel.  This programming model \emph{does not require explicit
+  declaration of parallelism within the program.}  In addition, programs are
+expressed using entirely standard Fortran so it can be compiled for and executed
+on a single core without concurrency.
 
@@
 this task within ROSE, it is considered outside the scope of this paper.
-However, we do provide (via Fortran interfaces) a mechanism to call the C OpenCL
+However, ForOpenCL provides via Fortran interfaces a mechanism to call the C OpenCL
 runtime and enable Fortran programmers to access OpenCL kernels generated by the
-automatic transformations.
+supplied transformations.
 

### 2011-06-26 papers/GPUScA-2011/programming-model.tex
Version to ship to Matt.  Needs a couple of bib entries.

@@
 
-A question that one may pose is ``\emph{Why Fortran and not a more modern
-  language like X?}''  The recent rise in interest in concurrency and
-parallelism at the language level driven by multicore CPUs and manycore
-accelerators has driven a number of new language developments, both as novel
-languages and extensions on existing ones.  For many scientific users familiar
-with Fortran, new languages and language extensions to use novel new
-architectures present a challenge: how do they effectively use them while
-avoiding rewriting code and potentially growing dependent on a transient
-technology that will vanish tomorrow?
+A question that one may pose is ``\emph{Why choose Fortran and not a more modern
+  language like X for programming accelerator architectures?}''  The recent rise
+in interest in concurrency and parallelism at the language level driven by
+multicore CPUs and manycore accelerators has driven a number of new language
+developments, both as novel languages and extensions on existing ones.  However,
+for many scientific users with existing codes written in Fortran, new languages
+and language extensions to use novel new architectures present a challenge: how
+do they effectively use them while avoiding rewriting code and potentially
+growing dependent on a transient technology that will vanish tomorrow?
+In this paper we explore the constructs in Fortran that are particularly
+relevant to GPU architecures.
 
@@
 subsetting language will allow scientific programmers to stay within the Fortran
-language and yet have direct access to GPU hardware.  We first examine how this
-programming model in Fortran relates to developments in other languages.
+language and yet have direct access to GPU hardware.  We start by examining how this
+programming model relates to developments in other languages.
 
@@
 A number of previous efforts have exploited data-parallel programming at the
-language level to utilize novel architectures, particularly in previous decades
-during the reign of vector and massively parallel computers in high performance
-computing (HPC).  The origin of the array syntax adopted by Fortran in the 1990
+language level to utilize novel architectures.
+%%, particularly in previous decades during the reign of vector and massively parallel computers in high performance computing (HPC). 
+The origin of the array syntax adopted by Fortran in the 1990
 standard can be found in the APL language~\cite{iverson79apl}.  These additions
 to Fortran allowed parallelism to be expressed with whole-array operations at
-the expression level, instead of via parallelism within explicit DO-loops (such
-as within IVTRAN for the Illiac IV).
+the expression level, instead of via parallelism within explicit DO-loops (e.g.,
+IVTRAN for the Illiac IV).
 
@@
 
-Similar whole array shifting was used in higher dimensions for finite
-difference codes within the computational physics community, especially
-at Los Alamos for codes targeting the CM-5 system that resided there until the
-late 1990s.  A body of research in compilation of stencil-based codes
-that use shift operators targeting these systems is related to the
-work we present here~\cite{stencil-compiler}.
+\noindent
+Similar whole array shifting was used in higher dimensions for finite difference
+codes within the computational physics community for codes targeting the CM-5
+system.  Research in compilation of stencil-based codes that use shift operators
+targeting these systems is related to the work presented
+here~\cite{stencil-compiler}.
 
@@
 a compiler to infer parallelism from a set of explicit loops, the choice for how
-to implement loops was left entirely up to the tool.  Unfortunately, this had
+to implement loops was left entirely up to the tool.
+
+Unfortunately, this had
 two side effects that have limited broad acceptance of the whole-array
@@
 
-An interesting line of language research that grew out of the early work with
-HPF was that associated with the ZPL language work at the University of
-Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar global
-view of computation over arrays, but define their computations based on regions,
+An interesting line of language research that grew out of HPF was that
+associated with the ZPL language at the University of
+Washington~\cite{chamberlain04zpl} and Chapel, an HPCS language developed by
+Cray~\cite{chamberlain_chapel}.  In ZPL, programmers adopt a similar global view
+of computation over arrays, but define their computations based on regions,
 which provide a local view of the set of indices that participate in the update
-of each element of an array.  The Chapel parallel programming language being
-developed by Cray is directly influenced by HPF and
-ZPL~\cite{chamberlain_chapel}.
+of each element of an array.
 
@@
 
-The static analysis and source-to-source transformations used in this work are
-very basic and only require the programmer to use a language subset that employs
-a data-parallel programming model.  In particular, it encourages the use of
-array notation, elemental functions, and pure procedures.  From these language
-constructs, we are able to easily transform Fortran procedures to a lower-level
-OpenCL kernel implementation.
+The static analysis and source-to-source transformations used in this work
+require the programmer to use a language subset that employs a data-parallel
+programming model.  In particular, it encourages the use of array notation,
+elemental functions, and pure procedures.  From these language constructs, we
+are able to easily transform Fortran procedures to a lower-level OpenCL kernel
+implementation.
 
@@
 Fortran has a rich array syntax that allows programmers to write statements in
-terms of whole arrays or subarrays, with data parallel operators to compute on
+terms of whole arrays or subarrays, with data-parallel operators to compute on
 the arrays.  Array variables can be used in expressions based on whole-array
@@
 
-\noindent results in the element-wise sum of {\tt A} and {\tt s} times the
+\noindent
+results in the element-wise sum of {\tt A} and the product of {\tt s} times the
 elements of {\tt B} being stored in the corresponding elements of {\tt C}. The
@@
 of an array in parallel, deferring the choice of implementation technique to the
-compiler.  Elemental functions are intended to be used for data parallel
+compiler.  Elemental functions are intended to be used for data-parallel
 programming, and as a result must be side effect free and mandate an {\tt
@@
 Since pure procedures don't have side effects they are candidates for running on
-accelerators in OpenCL.  Currently our prototype tool transforms pure procedures
+accelerators in OpenCL.  Currently our prototype tool only transforms pure procedures
 to OpenCL kernels that \emph{do not} call other procedures, except for elemental
-functions.
+functions, either defined by the user or intrinsic to Fortran.
 
@@
 set of functions that allow programmers to work with subarrays in expressions.
-In Fortran, these functions return a copy of an existing array or array section.
+In Fortran, these functions return a copy of or a pointer to an existing array or array section.
 This is unlike ZPL, where regions are analogous to index sets and are used
@@
 implementation generated by the compiler, where the size of an array is
-increased to provide extra array elements surrounding the interior portion of
-the array.  This is critical to reducing the amount of code that the programmer
-is forced to write, as the halo cells are often where boundary errors and
-off-by-one errors may occur in code that is manually generated.
+explicitly increased to provide extra array elements surrounding the interior portion of
+the array.
+%%%This is critical to reducing the amount of code that the programmer is forced to write, as the halo cells are often where boundary errors and off-by-one errors may occur in code that is manually generated.
 
@@
 portion.  However, unlike the shift operator, regions are not expressed in terms
-of boundary conditions and thus don't explicitly \emph{require} a knowledge of
-nor the application of boundary conditions locally.  Thus, as will be shown
+of boundary conditions and thus don't explicitly \emph{require} a knowledge of,
+nor the application of, boundary conditions locally.  Thus, as will be shown
 below, regions are more suitable for usage by OpenCL thread groups which access
@@
 
-ForOpenCL provides two new procedures that are defined in Fortran and are used
+ForOpenCL provides two new functions that are defined in Fortran and are used
 in array-syntax operations.  Each procedure takes a integer array halo argument
@@
 
-\item {\tt interior(array, halo)}: a pure function that returns a copy of
+\item {\tt region\_cpy(array, halo)}: a pure function that returns a copy of
   the interior portion of the array specified by halo.
 
-\item {\tt region(array, halo)}: a pure function that returns a copy of
-  the portion of the array specified by halo.
+\item {\tt region\_ptr(array, halo)}: an impure function that returns a pointer 
+  to the portion of the array specified by halo.
 
@@
 
-It should be noted that the two functions {\tt interior} and {\tt region} are
-pure and thus can be called from within a pure kernel procedure.  These two
-functions are part of the language recognized by the compiler and though the two
-functions return a copy of a portion of an array \emph{semantically}, the
-compiler is not forced to actually make a copy and is free to enforce copy
-semantics through other means.  In addition to these two new functions,
-ForOpenCL provides the compiler directive, {\tt \!\$OFP KERNEL}, which specifies
-that a pure subroutine can be transformed to an OpenCL kernel.
+It should be noted that the function {\tt region\_cpy} is pure and thus can be
+called from within a pure kernel procedure, though it should be noted that {\tt
+region\_ptr} is impure because it aliases the array parameter.  However as will
+be shown below, the usage of {\tt region\_ptr} is constrained so that it does not
+introduce side effects in the functions that call it.  These two functions are
+part of the language recognized by the compiler and though {\tt region\_cpy}
+returns a copy of a portion of an array \emph{semantically}, the compiler is not
+forced to actually make a copy and is free to enforce copy semantics through
+other means.  In addition to these two new functions, ForOpenCL provides the
+compiler directive, {\tt \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine
+can be transformed to an OpenCL kernel and that the subroutine is pure
+except for calls to {\tt region\_ptr}.
 
@@
 There are several advantages to this style of programming using array
-syntax, shifts, regions, and pure and elemental functions:
+syntax, regions, and pure and elemental functions:
 
@@
 expressions all semantically return a value, this style of programming is also
-similar to functional programming (or value-oriented programming).
+similar to functional programming (or value-oriented programming).  It should be
+noted that the subsetting language we employ goes beyond pure data parallelism
+by the use of pure (other than calls to {\tt region\_ptr}) subroutines and
+not just elemental functions.
 
 Unfortunately, this style of programming has never really caught on because when
-array syntax was first introduced in Fortran, performance of these features was
+array syntax was first introduced in Fortran, performance of codes using these features was
 relatively poor and thus programmers shied away from using array syntax (even
@@
 use it.  A goal of this paper is to demonstrate that parallel programs written
-in this style of Fortran are maintainable and can achieve good performance on
-accelerator architectures.
+in this style of Fortran can achieve good performance on accelerator architectures.
 
@@
 
-Only pure Fortran procedures are transformed into OpenCL kernels.  The
-programmer must explicitly call these kernels from Fortran using the ForOpenCL
-library described below.  It is also possible, using ROSE, to modify the calling
+Only pure Fortran procedures are transformed into OpenCL kernels.  This restriction
+is lifted slightly to allow calls to {\tt region\_ptr} from within a kernel procedure.  The
+programmer must explicitly call these kernels using Fortran interfaces in the ForOpenCL
+library (described below).  It is also possible, using ROSE, to modify the calling
 site so that the entire program can be transformed, but this functionality is
@@
 
-A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
-pure and all array variables must be declared as contiguous.  A kernel procedure
-may not call other procedures except for limited intrinsic functions (primarily
-math), user-defined elemental functions, and the {\tt interior} and {\tt region}
-functions.  Future work will address non-contiguous arrays (such as those that result
-from strided access) by mapping array strides to gather/scatter-style memory
-accessors.
+Array variables within a kernel procedure (specified by the {\tt \!\$OFP PURE,
+  KERNEL} directive) must be declared as contiguous.  A kernel procedure may not
+call other procedures except for limited intrinsic functions (primarily math),
+user-defined elemental functions, and the {\tt region\_cpy} and {\tt region\_ptr}
+functions.  Future work will address non-contiguous arrays (such as those that
+result from strided access) by mapping array strides to gather/scatter-style
+memory accessors.
+
+Array parameters to a kernel procedure must be declared as either intent(in) or
+intent(out); they cannot be intent(inout).  A thread may read from an extended
+region about its local element (using the {\tt region\_cpy} function, but can
+only write to the single array element it owns.  If a variable were intent(inout),
+a thread could update its array element before another thread had read from that
+element.  This restriction requires double buffering techniques.

### 2011-06-26 papers/GPUScA-2011/shallow-water.tex
Version to ship to Matt.  Needs a couple of bib entries.

@@
 momentum respectively, and {\tt oH, oU, oV} are corresponding output arrays used
-in the double buffering scheme.  The \emph{OFP} compiler directive attributes
+in the double buffering scheme.  The \emph{OFP} compiler-directive attributes
 {\tt PURE} and {\tt KERNEL} indicate that the procedure {\tt wave\_advance} is

### 2011-06-27 papers/GPUScA-2011/intro.tex
wolfe bib entry, minor tweaks

@@
 implementations removes the burden of creating and maintaining multiple versions
-of architecuture specific code.
+of architecture specific code.
 
-The peak performance of the newer accelerator architectures can be substantial.
+The peak performance of these newer accelerator architectures can be substantial.
 Intel expects a teraflop for the SGEMM benchmark with their Knights Ferry
@@
 common theme amongst the new processors is the emphasis on data-parallel
-programming.  This model is well suited to the emerging architectures that are
+programming.  This model is well suited to architectures that are
 based on either vector processing or massively parallel collections of simple
@@
 The problem with OpenCL and CUDA is that they expose too much detail about the
-machine architecture to the programmer~\cite{wolfeSC10}.  The programmer is
+machine architecture to the programmer~\cite{wolfe08gpgpu}.  The programmer is
 responsible for explicitly managing memory (including the staging of data back

### 2011-06-27 papers/GPUScA-2011/programming-model.tex
spelling and ack

@@
 in interest in concurrency and parallelism at the language level driven by
-multicore CPUs and manycore accelerators has driven a number of new language
+multicore CPUs and many-core accelerators has driven a number of new language
 developments, both as novel languages and extensions on existing ones.  However,
@@
 In this paper we explore the constructs in Fortran that are particularly
-relevant to GPU architecures.
+relevant to GPU architectures.
 
 In this section we present the Fortran subset employed in this paper.  This
-subsetting language will allow scientific programmers to stay within the Fortran
+sub-setting language will allow scientific programmers to stay within the Fortran
 language and yet have direct access to GPU hardware.  We start by examining how this
@@
 to Fortran allowed parallelism to be expressed with whole-array operations at
-the expression level, instead of via parallelism within explicit DO-loops (e.g.,
+the expression level, instead of via parallelism within explicit DO-loops, as
+implemented in earlier variants of the language (e.g.,
 IVTRAN for the Illiac IV).

### 2011-06-27 papers/GPUScA-2011/source-to-source.tex
spelling and ack

@@
 example in the use of the {\tt region\_cpy} function in the calculation of face
-variable quantites shown above), for performance reasons it is often best that
+variable quantities shown above), for performance reasons it is often best that
 global memory accessed and shared by a thread group be copied into local memory.

### 2011-06-27 papers/GPUScA-2011/programming-model.tex
edits to fix typos

@@
 implementation generated by the compiler, where the size of an array is
-explicitly increased to provide extra array elements surrounding the interior portion of
-the array.
+implicitly increased to provide extra array elements surrounding the interior 
+portion of the array.
 %%%This is critical to reducing the amount of code that the programmer is forced to write, as the halo cells are often where boundary errors and off-by-one errors may occur in code that is manually generated.
@@
 ForOpenCL provides two new functions that are defined in Fortran and are used
-in array-syntax operations.  Each procedure takes a integer array halo argument
+in array-syntax operations.  Each function takes a integer array halo argument
 that specifies the number of ghost cells on either side of a region, for each
@@
 similar to functional programming (or value-oriented programming).  It should be
-noted that the subsetting language we employ goes beyond pure data parallelism
+noted that the sub-setting language we employ goes beyond pure data parallelism
 by the use of pure (other than calls to {\tt region\_ptr}) subroutines and
@@
 intent(out); they cannot be intent(inout).  A thread may read from an extended
-region about its local element (using the {\tt region\_cpy} function, but can
+region about its local element (using the {\tt region\_cpy} function), but can
 only write to the single array element it owns.  If a variable were intent(inout),

### 2011-06-27 papers/GPUScA-2011/performance.tex
performance table

@@
 Several timing measurements were made by varying the size of the array state
-variables.  An average time was obtained by executing 100 iterations of the outer
+variables.  The performance measurements are shown in Table~\ref{table:performance}.  An average time was obtained by executing 100 iterations of the outer
 time-advance loop that called the OpenCL kernel. This tight loop kept the OpenCL
@@
 would have reduced the measured values.
+
+\begin{table}
+\begin{center}
+	\begin{tabular}{|c|c|c|c|}
+	\hline Array width & F90 & GPU (16x8) & Speedup \\ \hline\hline
+	16   & 0.025 ms & 0.017 ms & 1.5  \\ \hline
+	32   & 0.086    & 0.02     & 4.3  \\ \hline
+	64   & 0.20     & 0.02     & 10.0 \\ \hline
+	128  & 0.76     & 0.036    & 21.1 \\ \hline
+	256  & 3.02     & 0.092    & 32.8 \\ \hline
+	512  & 12.1     & 0.32     & 37.8 \\ \hline
+	1024 & 49.5     & 1.22     & 40.6 \\ \hline
+	1280 & 77.7     & 1.89     & 41.1 \\ \hline
+	2048 & 199.1    &  4.82    & 41.3 \\ \hline
+	4096 & 794.7    & 19.29    & 41.2 \\ \hline
+	\end{tabular}
+\end{center}
+\caption{Performance measurements for the shallow-water code.  All times reported in milliseconds.}
+\label{table:performance}
+\end{table}
 

### 2011-06-27 papers/GPUScA-2011/programming-model.tex
lots of edits

@@
 
-\subsection{Comparison to Other Languages}
+\subsection{Comparison to Prior Fortran Work}
 
@@
 which provide a local view of the set of indices that participate in the update
-of each element of an array.
+of each element of an array.  A similar line of research in the functional
+language community has investigated array abstractions for expressing
+whole-array operations in the Haskell language in the REPA (regular, shape-polymorphic,
+parallel array) library~\cite{keller10repa}.
 
@@
 can be transformed to an OpenCL kernel and that the subroutine is pure
-except for calls to {\tt region\_ptr}.
+except for calls to {\tt region\_ptr}.  These directives are not strictly
+necessary for the technique described in this paper, but aid in automated
+identification of specific kernels to be transformed to OpenCL.  A directive-free
+implementation would require the transformation tool be provided the set of 
+kernels to work via a user defined list.
 

### 2011-06-27 papers/GPUScA-2011/intro.tex
spelling and ack
Final corrections from Craig on plane to Munich before submission.

@@
 hierarchies (such as traditional caches).  These changes place a heavy burden on
-application programmers as they are attempt to adapt to these new systems.  An
+application programmers as they work to adapt to these new systems.  An
 especially challenging problem is not only how to program to these new
@@
 based on either vector processing or massively parallel collections of simple
-cores.  The recent CUDA and OpenCL programming languages are are intended to
+cores.  The recent CUDA and OpenCL programming languages are intended to
 support this programming model.

### 2011-06-27 papers/GPUScA-2011/programming-model.tex
Final corrections from Craig on plane to Munich before submission.

@@
 and language extensions to use novel new architectures present a challenge: how
-do they effectively use them while avoiding rewriting code and potentially
+do programmers effectively use them while avoiding rewriting code and potentially
 growing dependent on a transient technology that will vanish tomorrow?
@@
 implicitly increased to provide extra array elements surrounding the interior 
-portion of the array.
+portion of the array.  It is important to note, however, that all memory allocated
+by the programmer must explicitly contain the extra array elements in the halo.
 %%%This is critical to reducing the amount of code that the programmer is forced to write, as the halo cells are often where boundary errors and off-by-one errors may occur in code that is manually generated.
@@
 of boundary conditions and thus don't explicitly \emph{require} a knowledge of,
-nor the application of, boundary conditions locally.  Thus, as will be shown
+nor the application of, boundary conditions locally (global boundary conditions must
+be explicitly provided by the programmer outside of calls to kernel procedures).
+Thus, as will be shown
 below, regions are more suitable for usage by OpenCL thread groups which access
@@
 ForOpenCL provides two new functions that are defined in Fortran and are used
-in array-syntax operations.  Each function takes a integer array halo argument
+in array-syntax operations.  Each function takes an integer array halo argument
 that specifies the number of ghost cells on either side of a region, for each
@@
   understand, and is usually substantially shorter.
-\item Semantically the intrinsic functions return arrays by value.
+\item Semantically the intrinsic function {\tt region\_cpy} returns an array by value.
   This is usually what the algorithm requires.
-\item Pure and elemental function are free from side effects, so it is
+\item Pure and elemental functions are free from side effects, so it is
   easier for a compiler to schedule the work to be done in parallel.

### 2011-06-27 papers/GPUScA-2011/source-to-source.tex
lots of edits
Final corrections from Craig on plane to Munich before submission.

@@
 OpenCL host (normally the CPU).  Efforts to standardize a C++ runtime are
-underway and Fortran interfaces to the C runtime (distributed in the ForOpenCL
-library).
+underway and Fortran interfaces to the C runtime are distributed in the ForOpenCL
+library.
 
@@
 partitions of the global array are stored in the 4 MPI nodes shown in the foreground.
+Our current and future work on this effort includes source-to-source transformations
+to generate MPI code in addition to OpenCL in order to deal with clusters of 
+nodes containing accelerators.  This work is outside the scope of this paper.
 
@@
 
+\vskip .4in
+
 \subsubsection{Region function}
@@
 For example, the {\tt wave\_advance} interface declaration is transformed as
-
 {\small
@@
 }
-
+\noindent
 The intent(in) arrays have local equivalents that are stored in local memory

### 2011-06-30 papers/GPUScA-2011/programming-model.tex
reference, and some compression

@@
 
-A question that one may pose is ``\emph{Why choose Fortran and not a more modern
-  language like X for programming accelerator architectures?}''  The recent rise
-in interest in concurrency and parallelism at the language level driven by
-multicore CPUs and many-core accelerators has driven a number of new language
-developments, both as novel languages and extensions on existing ones.  However,
-for many scientific users with existing codes written in Fortran, new languages
-and language extensions to use novel new architectures present a challenge: how
-do programmers effectively use them while avoiding rewriting code and potentially
-growing dependent on a transient technology that will vanish tomorrow?
-In this paper we explore the constructs in Fortran that are particularly
-relevant to GPU architectures.
+A question that one may pose is ``\emph{Why choose Fortran and not a
+  more modern language like X for programming accelerator
+  architectures?}''  The recent rise in interest in concurrency and
+parallelism at the language level driven by multicore CPUs and
+many-core accelerators has driven a number of new language
+developments, both as novel languages and extensions on existing ones.
+However, for many scientific users with existing codes written in
+Fortran, new languages and language extensions to use novel new
+architectures present a challenge: how do programmers effectively use
+them while avoiding rewriting code and potentially growing dependent
+on a transient technology that will vanish tomorrow?  In this paper we
+explore the constructs in Fortran that are particularly relevant to
+GPU architectures.
 
-In this section we present the Fortran subset employed in this paper.  This
-sub-setting language will allow scientific programmers to stay within the Fortran
-language and yet have direct access to GPU hardware.  We start by examining how this
-programming model relates to developments in other languages.
+In this section we present the Fortran subset employed in this paper.
+This sub-setting language will allow scientific programmers to stay
+within the Fortran language and yet have direct access to GPU
+hardware.  We start by examining how this programming model relates to
+developments in other languages.
 
@@
 
-A number of previous efforts have exploited data-parallel programming at the
-language level to utilize novel architectures.
-%%, particularly in previous decades during the reign of vector and massively parallel computers in high performance computing (HPC). 
-The origin of the array syntax adopted by Fortran in the 1990
-standard can be found in the APL language~\cite{iverson79apl}.  These additions
-to Fortran allowed parallelism to be expressed with whole-array operations at
-the expression level, instead of via parallelism within explicit DO-loops, as
-implemented in earlier variants of the language (e.g.,
-IVTRAN for the Illiac IV).
+A number of previous efforts have exploited data-parallel programming
+at the language level to utilize novel architectures.  The origin of
+the array syntax adopted by Fortran in the 1990 standard can be found
+in the APL language~\cite{iverson79apl}.  These additions to Fortran
+allowed parallelism to be expressed with whole-array operations at the
+expression level, instead of via parallelism within explicit DO-loops,
+as implemented in earlier variants of the language (e.g., IVTRAN for the
+Illiac IV).
 
-The High Performance Fortran (HPF) extension of Fortran was proposed to add
-features to the language that would enhance the ability of compilers to emit
-fast parallel code for distributed and shared memory parallel
-computers\cite{koelbel94hpf}.  One of the notable additions to the language in
-HPF was syntax to specify the distribution of data structures amongst a set of
-parallel processors.  HPF also introduced an alternative looping construct to
-the traditional DO-loop called {\tt FORALL} that was better suited for parallel
-compilation.  An additional keyword, {\tt INDEPENDENT}, was added to allow the
-programmer to indicate when the loop contained no loop-order dependencies that
-allowed for parallel execution.  These constructs are similar to {\tt DO
-  CONCURRENT}, an addition to Fortran in 2008.  Interestingly, the parallelism
-features introduced in HPF did not exploit the new array features introduced in
-1990 in any significant way, relying instead on explicit loop-based parallelism.
-This restriction allowed the language to support parallel programming that
-wasn't easily mapped onto a pure data-parallel model.
+The High Performance Fortran (HPF) extension of Fortran was proposed
+to add features to the language that would enhance the ability of
+compilers to emit fast parallel code for distributed and shared memory
+parallel computers\cite{koelbel94hpf}.  One of the notable additions
+to the language in HPF was syntax to specify the distribution of data
+structures amongst a set of parallel processors.  HPF also introduced
+an alternative looping construct to the traditional DO-loop called
+{\tt FORALL} that was better suited for parallel compilation.  An
+additional keyword, {\tt INDEPENDENT}, was added to allow the
+programmer to indicate when the loop contained no loop-order
+dependencies that allowed for parallel execution.  These constructs
+are similar to {\tt DO CONCURRENT}, an addition to Fortran in 2008.
 
-In some instances though, a purely data-parallel model is appropriate for part
-or all of the major computations within a program.  One of the systems where
-programmers relied heavily on higher level operations instead of explicit
-looping constructs was the Thinking Machines Connection Machine 5 (CM-5).  A
-common programming pattern used on the CM-5 (that we exploit in this paper) was to
-write whole-array operations from a global perspective in which computations are
-expressed in terms of operations over the entire array instead of a single local
-index.  The use of the array shift intrinsic functions (like {\tt CSHIFT}) were
-used to build computations in which arrays were combined by shifting the entire
-arrays instead of working on local offsets based on single indices.  A simple
-1D example is one in which an element is replaced with the average of its own
-value and that of its two direct neighbors.  Ignoring boundary indices that
-wrap around, explicit indexing will result in a loop such as:
+Interestingly, the parallelism features introduced in HPF did not
+exploit the new array features introduced in 1990 in any significant
+way, relying instead on explicit loop-based parallelism.  This
+restriction allowed the language to support parallel programming that
+wasn't easily mapped onto a pure data-parallel model.  The {\tt SHADOW}
+directive introduced in HPF-2, and the {\tt HALO} in HPF+~\cite{benkner99hpf}
+bear some similarity to the halo region concept that we discuss in this 
+paper.
+
+In some instances though, a purely data-parallel model is appropriate
+for part or all of the major computations within a program.  One of
+the systems where programmers relied heavily on higher level
+operations instead of explicit looping constructs was the Thinking
+Machines Connection Machine 5 (CM-5).  A common programming pattern
+used on the CM-5 (that we exploit in this paper) was to write
+whole-array operations from a global perspective in which computations
+are expressed in terms of operations over the entire array instead of
+a single local index.  The use of the array shift intrinsic functions
+(like {\tt CSHIFT}) were used to build computations in which arrays
+were combined by shifting the entire arrays instead of working on
+local offsets based on single indices.  A simple 1D example is one in
+which an element is replaced with the average of its own value and
+that of its two direct neighbors.  Ignoring boundary indices that wrap
+around, explicit indexing will result in a loop such as:
 
@@
 
-Unfortunately, this had
-two side effects that have limited broad acceptance of the whole-array
-programming model in Fortran.  First, programmers must translate their
-algorithms into a set of global operations.  Finite difference stencils and
-similar computations are traditionally defined in terms of offsets from some
-central index.  Shifting, while conceptually analogous, can be awkward to think
-about for high dimensional stencils with many points.  Second, the semantics of
-these operations are such that all elements of an array operation are updated as
-if they were updated simultaneously.  In a program where the programmer
-explicitly manages arrays and loops, double buffering techniques and user
-managed temporaries are used to maintain these semantics.  When the compiler is
-responsible for managing this intermediate storage, it has historically proven
-that they are inefficient and generate code that requires far more temporary
-storage than really necessary.  This is not a flaw of the language constructs,
-but a sign of the lack of sophistication of the compilers with respect to their
-internal analysis to determine how to optimally generate this intermediate
-storage.
+Unfortunately, this had two side effects that have limited broad
+acceptance of the whole-array programming model in Fortran.  First,
+programmers must translate their algorithms into a set of global
+operations.  Finite difference stencils and similar computations are
+traditionally defined in terms of offsets from some central index.
+Shifting, while conceptually analogous, can be awkward to think about
+for high dimensional stencils with many points.  Second, the semantics
+of these operations are such that all elements of an array operation
+are updated as if they were updated simultaneously.  In a program
+where the programmer explicitly manages arrays and loops, double
+buffering techniques and user managed temporaries are used to maintain
+these semantics.  Limited attention to optimizing memory usage due to
+this intermediate storage by compilers has led to these constructs seeing
+little adoption by programmers.
+
+%When the compiler is responsible for managing this
+%intermediate storage, it has historically proven that they are
+%inefficient and generate code that requires far more temporary storage
+%than really necessary.  
+%This is not a flaw of the language constructs,
+%but a sign of the lack of sophistication of the compilers with respect
+%to their internal analysis to determine how to optimally generate this
+%intermediate storage.
 

### 2011-08-30 papers/GPUScA-2011/intro.tex
updates based on reviewer comments and other fixes

@@
 programming model without sacrificing performance.  We suggest that existing
-data-parallel features in Fortran are well suited to applying automatic
+data-parallel features in Fortran are well-suited to applying automatic
 transformations that generate code specifically tuned for different hardware
@@
 on superscalar instruction level parallelism and hardware managed memory
-hierarchies (such as traditional caches).  These changes place a heavy burden on
+hierarchies (such as traditional caches).
+
+These changes place a heavy burden on
 application programmers as they work to adapt to these new systems.  An
@@
 computer architectures.  How does a programmer write one program that can
-perform well on both a conventional multicore CPU \emph{and} a GPU (or another
-of the emerging many-core architectures)?
+perform well on both a conventional multicore CPU \emph{and} a GPU (or any other emerging many-core architectures)?
 
@@
 common theme amongst the new processors is the emphasis on data-parallel
-programming.  This model is well suited to architectures that are
+programming.  This model is well-suited to architectures that are
 based on either vector processing or massively parallel collections of simple
@@
 to every new architecture and programming model that emerges.
+
+\newpage
 
@@
 (CFD).  The example described later in this paper is a simple shallow-water
-model in two dimensions using finite volume methods.
+model in two dimensions using finite volume methods.  Stencil-like patterns
+appear in a number of other contexts as well.  In image processing, they appear
+in convolution-based algorithms in which small kernels are convolved with an image
+to implement denoising, edge detection, and other common operators.  Similar stencil
+operators appear in general signal processing applications as well.  
+%% more?
 
@@
 version of the shallow-water code.
+
+We do not perform any additional analysis of the code to identify
+parallelism beyond that present in the data parallel operations that we
+focus on in this paper.  Additional program
+analysis methods may be investigated to study their applicability
+in future versions of this work.

### 2011-08-30 papers/GPUScA-2011/programming-model.tex
updates based on reviewer comments and other fixes

@@
   do i = 2,(n-1)
-    X(i) = (X(i-1) + X(i) + X(i+1)) / 3
+    Xnew(i) = (X(i-1) + X(i) + X(i+1)) / 3
   end do
+  PTR_SWAP(Xnew, X, tmp_ptr)
 \end{verbatim}
@@
 
-\noindent When shifts are employed, this can be expressed as:
+\noindent In this loop, it is necessary to manually implement a double buffering scheme
+in order to avoid mixing values computed in the current loop with values from a 
+prior execution of the loop. When shifts are employed, this can be expressed as:
 
@@
 \noindent
+The semantics of the shift intrinsic and operators like {\tt +} applied to the whole array make it
+unnecessary to manually implement the double buffering scheme found in the loop above.
 Similar whole array shifting was used in higher dimensions for finite difference
@@
 
-\subsubsection*{Elemental functions}
+\subsubsection*{Pure elemental functions}
 
@@
 of an array in parallel, deferring the choice of implementation technique to the
-compiler.  Elemental functions are intended to be used for data-parallel
+compiler.  Pure elemental functions are intended to be used for data-parallel
 programming, and as a result must be side effect free and mandate an {\tt
@@
 
-For example, the basic array operation shown above could be refactored into an
+For example, the basic array operation shown above could be refactored into a pure
 elemental function,
@@
 perform powerful computations, especially when applied to arrays.  Our prototype
-tool transforms elemental functions to inline OpenCL functions.  Thus there is
-no penalty for usage of elemental functions and they provide a convenient
+tool transforms pure elemental functions to inline OpenCL functions.  Thus there is
+no penalty for usage of pure elemental functions and they provide a convenient
 mechanism to express algorithms in simpler segments.
+
+It should be noted that in Fortran 2008,
+the concept of impure elemental functions was introduced.  This requires that
+elemental functions now need to be explicitly labelled as pure to ensure that they
+are side effect free.
 
@@
 
-Pure procedures, like elemental functions, must be free of side effects.  Unlike
-elemental functions that require arguments to have an {\tt intent(in)}
+Pure procedures, like pure elemental functions, must be free of side effects.  Unlike
+pure elemental functions that require arguments to have an {\tt intent(in)}
 attribute, they may change the contents of array arguments that are passed to
@@
 accelerators in OpenCL.  Currently our prototype tool only transforms pure procedures
-to OpenCL kernels that \emph{do not} call other procedures, except for elemental
+to OpenCL kernels that \emph{do not} call other procedures, except for pure elemental
 functions, either defined by the user or intrinsic to Fortran.
@@
   This is usually what the algorithm requires.
-\item Pure and elemental functions are free from side effects, so it is
+\item Pure elemental functions are free from side effects, so it is
   easier for a compiler to schedule the work to be done in parallel.

### 2011-08-30 papers/local-orientation/code/convolve/convolve.f90
Initial version.

@@ -1,0 +1,197 @@
+module convolve_mod
+   use ISO_C_BINDING
+
+   integer, parameter :: G_WIDTH = 1280
+
+   ! WARNING, must also change in convolve.cl
+   integer(c_int), parameter :: NPAD = 7
+
+   integer(c_int), parameter :: NXP = 1 + 2*NPAD
+   integer(c_int), parameter :: NYP = NXP
+
+   integer(c_size_t), parameter :: NX   = G_WIDTH
+   integer(c_size_t), parameter :: NY   = G_WIDTH
+   integer(c_size_t), parameter :: NXEX = NX + 2*NPAD
+   integer(c_size_t), parameter :: NYEX = NY + 2*NPAD
+   integer(c_size_t), parameter :: NXL  = 16
+   integer(c_size_t), parameter :: NYL  = 16
+
+   integer(c_size_t), parameter :: SIZE_ELEMENT = 4
+
+contains
+
+   subroutine convolve(S, Image, F)
+      implicit none
+      real :: S(NX,NY), Image(NXEX,NYEX), F(NXP,NYP)
+      real :: val
+      integer :: i, j, ip, jp
+
+      do j = 1, NY
+         do i = 1, NX
+            val = 0.0
+            do jp = 1, NYP
+               do ip = 1, NXP
+                  val = val + F(ip, jp)*Image(i+NPAD,j+NPAD)
+               end do
+            end do
+            S(i,j) = val
+         end do
+      end do
+   end subroutine convolve
+
+   subroutine init_filter(F)
+      implicit none
+      real :: F(NXP,NYP)
+      integer :: i
+
+      F = 0.0
+      F(:,NYP/2) = 1
+      F(NXP/2,:) = 1
+      F = F/sum(F)
+
+    end subroutine init_filter
+
+end module convolve_mod
+
+
+program test_convolve
+   use convolve_mod
+   use OpenCL
+   use Timer_mod
+   implicit none
+
+   integer :: status
+
+   real(c_float), target, dimension(NX+2*NPAD,NY+2*NPAD) :: I  ! image
+   real(c_float), target, dimension(NX,NY)               :: S  ! smoothed image
+   real(c_float), target, dimension(NXP,NYP)             :: F  ! filter
+
+   type(CLDevice) :: device
+   type(CLKernel) :: kernel
[…]

### 2011-08-30 papers/local-orientation/code/convolve/convolve_elemental.f90

@@ -1,0 +1,9 @@
+   CONCURRENT elemental subroutine convolve(S, Image, F, N)
+      real, intent(out) :: S
+      real, intent(in), EXTEND(0) :: Image(-N:N,-N:N)
+      real, intent(in), NON_ELEMENTAL :: F(-N:N,-N:N)
+      integer, intent(in) :: N
+
+      S = sum(F*Image)
+
+   end subroutine convolve

### 2011-08-30 papers/GPUScA-2011/performance.tex

@@
 
-While we did not have an OpenMP code for multi-core comparisons, the transformed
+While we did not have an OpenMP code for multicore comparisons, the transformed
 OpenCL code on the NVIDIA C2050 was up to 40 times faster than the best serial

### 2011-08-30 papers/GPUScA-2011/programming-model.tex

@@
   architectures?}''  The recent rise in interest in concurrency and
-parallelism at the language level driven by multicore CPUs and
+parallelism at the language level due to multicore CPUs and
 many-core accelerators has driven a number of new language
@@
 \noindent In this loop, it is necessary to manually implement a double buffering scheme
-in order to avoid mixing values computed in the current loop with values from a 
+in order to avoid mixing values computed in the current execution of the loop with values from a 
 prior execution of the loop. When shifts are employed, this can be expressed as:
@@
 programming model.  In particular, it encourages the use of array notation,
-elemental functions, and pure procedures.  From these language constructs, we
+pure elemental functions, and pure procedures.  From these language constructs, we
 are able to easily transform Fortran procedures to a lower-level OpenCL kernel
@@
 the concept of impure elemental functions was introduced.  This requires that
-elemental functions now need to be explicitly labelled as pure to ensure that they
+elemental functions now need to be explicitly labeled as pure to indicate that they
 are side effect free.
@@
 It should be noted that the function {\tt region\_cpy} is pure and thus can be
-called from within a pure kernel procedure, though it should be noted that {\tt
+called from within a pure kernel procedure, and {\tt
 region\_ptr} is impure because it aliases the array parameter.  However as will

### 2011-08-30 papers/local-orientation/code/convolve/convolve_elemental.f90
Changed name to SmoothedImage

@@ -1,9 +1,9 @@
-   CONCURRENT elemental subroutine convolve(S, Image, F, N)
-      real, intent(out) :: S
+   CONCURRENT elemental subroutine convolve(SmoothedImage, Image, F, N)
+      real, intent(out) :: SmoothedImage
       real, intent(in), EXTEND(0) :: Image(-N:N,-N:N)
       real, intent(in), NON_ELEMENTAL :: F(-N:N,-N:N)
       integer, intent(in) :: N
 
-      S = sum(F*Image)
+      SmoothedImage = sum(F*Image)
 
    end subroutine convolve

### 2011-08-30 papers/local-orientation/code/convolve/convolve_elemental.f90
Added pure attribute to subroutine

@@ -1,4 +1,4 @@
-   CONCURRENT elemental subroutine convolve(SmoothedImage, Image, F, N)
+   CONCURRENT pure elemental subroutine convolve(SmoothedImage, Image, F, N)
       real, intent(out) :: SmoothedImage
       real, intent(in), EXTEND(0) :: Image(-N:N,-N:N)
       real, intent(in), NON_ELEMENTAL :: F(-N:N,-N:N)

### 2011-10-24 checkers/implicitCast/compass.C
Added implicitCast checker, based on older noCast checker

@@ -1,0 +1,927 @@
+
+#if ROSE_MPI
+#include <mpi.h>
+#endif
+
+#ifdef PALETTE_USE_ROSE
+#include "compass.h"
+#else
+#include "rose.h"
+#include "compass.h"
+#endif
+
+
+//for exists
+#include "boost/filesystem/operations.hpp"
+//#include <rose.h>
+#include <sstream>
+#include <fstream>
+#include <iostream> 
+
+
+// Default setting for verbosity level (-1 is silent, and values greater then zero indicate different levels of verbosity)
+int  Compass::verboseSetting   = 0;
+bool Compass::compilerWarnings = false;
+bool Compass::compilerRemarks  = false;
+
+// DQ (1/17/2008): New Flymake mode
+bool Compass::UseFlymake       = false;
+
+//! Support for ToolGear XML viewer for output data when run as batch
+bool Compass::UseToolGear      = false; 
+std::string Compass::tguiXML;
+
+//! Support for outputting to SQLite database when run as batch
+bool Compass::UseDbOutput        = false; 
+std::string Compass::outputDbName;
+#ifdef HAVE_SQLITE3
+sqlite3x::sqlite3_connection Compass::con;
+#endif
+
+
+// TPS, needed for DEFUSE
+unsigned int Compass::global_arrsize=-1;
+unsigned int Compass::global_arrsizeUse=-1;
+unsigned int* Compass::def_values_global=NULL;
+unsigned int* Compass::use_values_global=NULL;
+bool Compass::quickSave=false;
+
+
+/******************************************************************
+ * MPI CODE TO RUN DEFUSE IN PARALLEL WITH MPI
+ ******************************************************************/
+
+#if ROSE_MPI
+int Compass::my_rank=0;
+int Compass::processes=0;
+
+
+void Compass::loadDFA(std::string name, SgProject* project) {
+  MPI_Status Stat;
+  int res[2];
+  res[0]=0;
+  res[1]=1;
+  bool done=false;
+  global_arrsize=-2; global_arrsizeUse=-2;
+  std::cerr << " Starting the load with " << processes << " processes. " << std::endl;
+  while (!done) {
+
+    Compass::sourceDefUsePrerequisite.load(project);
+    ROSE_ASSERT(Compass::sourceDefUsePrerequisite.done==true);
+    // The dataflow analysis must return at least one definition;
[…]

### 2011-10-24 checkers/implicitCast/compass.h

@@ -1,0 +1,401 @@
+#ifndef ROSE_COMPASS_H
+#define ROSE_COMPASS_H
+#if ROSE_MPI
+#undef SEEK_SET
+#undef SEEK_CUR
+#undef SEEK_END
+#include <mpi.h>
+#endif
+
+//#include <rose.h>
+#include "DefUseAnalysis_perFunction.h"
+#include "DefUseAnalysis.h"
+#include <boost/function.hpp>
+#include <boost/any.hpp>
+
+#if ROSE_MPI
+#include "functionLevelTraversal.h"
+#include "functionNames.h"
+#endif
+
+
+#ifdef HAVE_SQLITE3
+
+#include "sqlite3x.h"
+#include <boost/lexical_cast.hpp>
+#endif
+
+namespace Compass {
+  extern unsigned int global_arrsize;
+  extern unsigned int global_arrsizeUse;
+  extern unsigned int *def_values_global;
+  extern unsigned int *use_values_global;
+  typedef std::map< SgNode* , std::vector < std::pair <SgInitializedName* , SgNode*>  > > my_map; 
+  extern bool quickSave;
+
+  static std::set<std::string> pre;
+
+
+#if ROSE_MPI
+  //  struct timespec begin_time_node, end_time_node;
+  extern int my_rank, processes;
+
+  void loadDFA(std::string name, SgProject* project);
+  void saveDFA(std::string name, SgProject* project);
+
+  /****************************************************************************/
+  // (tps - 11Jul08) : needed for MPI run of Compass
+  void serializeDefUseResults(unsigned int *values,
+			      std::map< SgNode* , std::vector <std::pair  < SgInitializedName* , SgNode* > > > &defmap,
+			      std::map<SgNode*,unsigned int > &nodeMap);
+  void deserializeDefUseResults(unsigned int arrsize, DefUseAnalysis* defuse, unsigned int *values,
+				//std::map<unsigned int, SgNode* > &nodeMap, 
+				std::vector<SgNode* > &nodeMap, 
+				bool definition);
+
+  double timeDifference(struct timespec end, struct timespec begin);
+  inline void gettime(struct timespec &t) {
+    clock_gettime(CLOCK_PROCESS_CPUTIME_ID, &t);
+  }
+  class MemoryTraversal : public ROSE_VisitTraversal {
+  public:
+  MemoryTraversal():counter(0){}
+    unsigned int counter;
+    void visit ( SgNode* node );
+    //std::map <unsigned int, SgNode* > nodeMap;
[…]

### 2011-10-24 checkers/implicitCast/implicitCast.C

@@ -8,7 +8,7 @@
 #include "rose.h"
 #include "compass.h"
 
-extern const Compass::Checker* const noCastChecker;
+extern const Compass::Checker* const implicitCastChecker;
 
 // DQ (1/17/2009): Added declaration to match external defined in file:
 // rose/projects/compass/extensions/prerequisites/ProjectPrerequisite.h
@@ -18,11 +18,11 @@ Compass::ProjectPrerequisite Compass::projectPrerequisite;
 
 namespace CompassAnalyses
    { 
-     namespace NoCast
+     namespace ImplicitCast
         { 
         /*! \brief No Cast: Add your description here 
          */
-           extern const std::string checkerName= "NoCast";
+           extern const std::string checkerName= "ImplicitCast";
            extern const std::string shortDescription= "Implicit cast in Fortran code found: ";
            extern const std::string longDescription= "Finding statement with implicit cast in a Fortran code :";
 
@@ -61,23 +61,23 @@ namespace CompassAnalyses
         }
    }
 
-CompassAnalyses::NoCast::
+CompassAnalyses::ImplicitCast::
 CheckerOutput::CheckerOutput ( SgNode* node, const std::string & reason )
    : OutputViolationBase(node,checkerName,shortDescription+reason)
    {}
 
-CompassAnalyses::NoCast::Traversal::
+CompassAnalyses::ImplicitCast::Traversal::
 Traversal(Compass::Parameters inputParameters, Compass::OutputObject* output)
    : output(output)
    {
   // Initalize checker specific parameters here, for example: 
-  // YourParameter = Compass::parseInteger(inputParameters["NoCast.YourParameter"]);
+  // YourParameter = Compass::parseInteger(inputParameters["ImplicitCast.YourParameter"]);
 
 
    }
 
 void
-CompassAnalyses::NoCast::Traversal::
+CompassAnalyses::ImplicitCast::Traversal::
 visit(SgNode* node)
    { 
   // Implement your traversal here.  
@@ -147,20 +147,20 @@ visit(SgNode* node)
 // Checker main run function and metadata
 
 static void run(Compass::Parameters params, Compass::OutputObject* output) {
-  CompassAnalyses::NoCast::Traversal(params, output).run(Compass::projectPrerequisite.getProject());
+  CompassAnalyses::ImplicitCast::Traversal(params, output).run(Compass::projectPrerequisite.getProject());
 }
 
 // Remove this function if your checker is not an AST traversal
[…]

### 2011-10-24 checkers/implicitCast/implicitCastMain.C

@@ -1,0 +1,10 @@
+//
+// Do not modify this file
+//
+#include "rose.h"
+#include "compass.h"
+extern const Compass::Checker* const implicitCastChecker;
+extern const Compass::Checker* const myChecker = implicitCastChecker;
+
+#include "compass.C"
+#include "compassTestMain.C"

### 2011-10-24 checkers/implicitCast/implicitCastTest1.f90

@@ -1,0 +1,12 @@
+program test
+
+  INTEGER :: x
+  REAL    :: y
+  
+  y = 42.4
+  x = INT(y) + 4
+  
+  print *, x
+  print *, y
+
+end

### 2011-10-24 checkers/implicitSave/implicitSaveMain.C
add implicitSave checker

@@ -1,0 +1,10 @@
+//
+// Do not modify this file
+//
+#include "rose.h"
+#include "compass.h"
+extern const Compass::Checker* const implicitSaveChecker;
+extern const Compass::Checker* const myChecker = implicitSaveChecker;
+
+#include "compass.C"
+#include "compassTestMain.C"

### 2011-10-24 checkers/implicitSave/implicitSaveTest1.f90

@@ -1,0 +1,31 @@
+subroutine f
+  integer :: x = 4
+  print *, x
+  x = x + 1
+end subroutine f
+
+subroutine g
+  integer, save :: x = 4
+  print *, x
+  x = x +1
+end subroutine g
+
+subroutine h
+  integer :: x = 4
+  print *,x
+end subroutine
+
+program main
+  print *, "Testing f"
+  call f
+  call f
+  call f
+  print *, "Testing g"
+  call g
+  call g
+  call g
+  print *, "Testing h"
+  call h
+  call h
+  call h
+end program main

### 2011-10-27 papers/LOPe/code/convolve/variable_filter.f90
Commented out c_sizeof as doesnt work on with ifort.

@@ -1,0 +1,224 @@
+module variable_filter_mod
+   use FilterParams
+   use FilterType
+
+contains
+
+   subroutine convolve(S, Image, F)
+      implicit none
+      real :: S(NX,NY), Image(NXEX,NYEX)
+      type(FilterPatch) :: F(NX,NY)
+      real :: val
+      integer :: i, j, ip, jp
+
+      do j = 1, NY
+         do i = 1, NX
+            val = 0.0
+            do jp = -NPAD, NPAD
+               do ip = -NPAD, NPAD
+                  val = val + F(i,j)%p(ip,jp)*Image(i+NPAD,j+NPAD)
+               end do
+            end do
+            S(i,j) = val
+         end do
+      end do
+   end subroutine convolve
+
+   subroutine filter_bandwidth(S, Image, F)
+      implicit none
+      real :: S(NX,NY), Image(NXEX,NYEX)
+      type(FilterPatch) :: F(NX,NY)
+      real :: val
+      integer :: i, j, ip, jp
+
+      do j = 1, NY
+         do i = 1, NX
+            val = 0.0
+            do jp = -NPAD, NPAD
+               do ip = -NPAD, NPAD
+                  F(i,j)%p(ip,jp) = 1
+               end do
+            end do
+            S(i,j) = 1.0
+         end do
+      end do
+   end subroutine filter_bandwidth
+
+
+   subroutine init_filter(F)
+      implicit none
+      type(FilterPatch) :: F(NX,NY)
+      integer :: i, j
+
+      F(1,1)%p = 0.0
+      F(1,1)%p(:,0) = 1
+      F(1,1)%p(0,:) = 1
+      F(1,1)%p = F(1,1)%p/sum(F(1,1)%p)         
+
+      do j = 1, NY
+         do i = 1, NX
+            F(i,j)%p = F(1,1)%p
+         end do
+      end do
+
+      print *, nxp, nyp, NPAD
+      do i = -1, 1
+         print *, "F==", F(2,2)%p(i,-1:1)
+      end do
+      print *
+
+    end subroutine init_filter
+
+end module variable_filter_mod
+
+
+program test_variable_filter
+   use variable_filter_mod
+   use OpenCL
+   use Timer_mod
+   implicit none
+
+   integer :: status
+
[…]

### 2011-10-27 papers/local-orientation/code/C/elemental_add.c
Moved local-orientation directory to LOPe directory

@@ -1,45 +1,0 @@
-// Make extended syntax go away
-//
-#define ELEMENTAL
-#define PURE
-
-// Array size
-//
-#define N 100
-
-//
-// The function adds two input array elements and returns the output
-// as an output parameter.
-//
-
-void PURE ELEMENTAL elemental_add(const float * a, const float * b, float * c)
-{
-   *c = *a + *b;
-}
-
-int main(int argc, char* argv[])
-{
-   float A[N], B[N], C[N];
-
-   // initialize arrays...
-
-   // call elemental function
-
-   elemental_add(A, B, C);
-}
-
-
-int main_compiler_creates(int argc, char* argv[])
-{
-   int i;
-   float A[N], B[N], C[N];
-
-   // initialize arrays...
-
-   // inline elemental function
-   for (i = 0; i < N; i++) {
-      C[i] = A[i] + B[N];
-   }
-}
-
-

### 2011-10-27 papers/local-orientation/code/LIF_update_state.f90

@@ -1,111 +1,0 @@
-module types_mod
-   type LIF_params
-      real :: tauE, tauI, tauIB, tauVth
-   end type LIF_params
-   type uint4
-      integer :: val1
-   end type uint4
-end module types_mod
-
-pure subroutine &
-LIF_update_state(time, dt, &
-                 params, rnd, &
-                 V, Vth, &
-                 G_E, G_I, G_IB, &
-                 phiExc, phiInh, phiInhB, &
-                 R, activity)
-   !$OFP extended elemental :: LIF_update_state
-   use types_mod
-   implicit none
-   real, intent(in) :: time, dt
-   type(LIF_params), intent(in) :: params
-   type(uint4), intent(inout) :: rnd
-   real, intent(inout) :: V, V_th
-   real, intent(inout) :: G_E, G_I, G_IB
-   real, intent(inout) :: phiExc, phiInh, phiInhB
-   real, intent(inout) :: R, activity
-
-   real :: exp_tauE, exp_tauI, exp_tauIB, exp_tauVth, exp_tauRate
-   real :: dt_sec, GMAX
-
-   ! temporary arrays
-   real :: tauInf, VmemInf
-
-   ! local param variables
-   real :: tau, Vrest, VthRest, Vexc, Vinh, VinhB, deltaVth
-
-   exp_tauE    = exp(-dt/params%tauE)
-   exp_tauI    = exp(-dt/params%tauI)
-   exp_tauIB   = exp(-dt/params%tauIB)
-   exp_tauVth  = exp(-dt/params%tauVth)
-
-   dt_sec = .001 * dt
-   exp_tauRate = exp(-dt/params%tauRate)
-
-   GMAX = 10.0
-
-   !
-   ! start of LIF2_update_exact_linear
-   !
-
-   ! define local param variables
-   !
-   tau   = params%tau
-   Vexc  = params%Vexc
-   Vinh  = params%Vinh
-   VinhB = params%VinhB
-   Vrest = params%Vrest
-
-   VthRest  = params%VthRest
-   deltaVth = params%deltaVth
-
-   ! add noise
-   !
-
-   G_E  = phiExc  + G_E *exp_tauE
-   G_I  = phiInh  + G_I *exp_tauI
-   G_IB = phiInhB + G_IB*exp_tauIB
-   
-   !G_E  = (G_E  > GMAX) ? GMAX : G_E;
[…]

### 2011-10-27 papers/local-orientation/code/Retina_update_state.f90

@@ -1,12 +1,0 @@
-pure subroutine &
-Retina_update_state(time, dt, nx, ny, nf, nb, params, &
-                    phiExc, phiInh, activity)
-   !$OFP extended elemental :: Retina_update_state
-   implicit none
-   real, intent(in) :: time, dt
-   integer, intent(in) :: nx, ny, nf, nb
-   type(Retina_params), intent(in) :: params
-   type(C_PTR) :: params1
-   real, intent(inout) :: phiExc, phiInh, activity
-
-end subroutine Retina_update_state

### 2011-10-27 papers/local-orientation/code/advect.f90

@@ -1,81 +1,0 @@
-!
-! one dimensional advection with constant velocity
-!
-!         h_t + u_x = 0
-!
-! where u = ah
-!
-!
-!pure subroutine advect(H_big, halo, a, dx, dt)
-!   !$OFP LOCAL advect
-!   real, dimension(:), intent(inout) :: H_big   ! includes the halo region
-!   real, intent(in) :: a, dx, dt
-!   integer, intent(in) :: halo(2)
-!
-!   real :: u_west, u_east
-!   real :: iH, H(-1:1), U(-1:1)
-!   integer :: i
-!
-!   iH = local(H, halo)                      ! local variables are copies
-!    H = local_section(H_big, halo, [1,1])   ! take 3, 1 on either side
-!    U = a*H
-!
-!   ! Lax step (t+dt/2)
-!   !
-!   i = 0
-!   u_west = 0.5*a* ( H(i-1) + H(i  ) ) + (0.5*dt/dx) * ( U(i-1) - U(i  ) )
-!   u_east = 0.5*a* ( H(i  ) + H(i+1) ) + (0.5*dt/dx) * ( U(i  ) - U(i+1) )
-!
-!   ! second Lax-Wendroff step (t+dt)
-!   !
-!   iH = iH + (0.5*dt/dx) * (u_west - u_east)
-!
-!   ! local variable iH stored back to array H_big at exit
-!end subroutine advect
-
-
-!
-! implementation
-!
-subroutine advect_impl(H_big, halo, a, dx, dt)
-   implicit none
-   integer, parameter :: NX = 16
-   real, dimension(NX+2), intent(inout) :: H_big   ! includes the halo region
-   real, intent(in) :: a, dx, dt
-   integer, intent(in) :: halo(2)
-
-   real :: u_west, u_east, diff
-   real :: H(-1:1), U(-1:1)
-   integer :: i, iH
-
-   ! loop over interior
-   !
-   do i = 1, NX
-      iH = i + halo(1)
-       H = H_big(iH-1:iH+1)
-       U = a*H
-
-      ! Lax step (t+dt/2)
-      !
-      u_west = 0.5*a* ( H(-1) + H(0) ) - (0.5*dt/dx) * ( U(-1) - U(0) )
-      u_east = 0.5*a* ( H( 0) + H(1) ) - (0.5*dt/dx) * ( U( 0) - U(1) )
-
-      ! second Lax-Wendroff step (t+dt)
-      !
[…]

### 2011-10-27 papers/local-orientation/code/convolve.f90

@@ -1,51 +1,0 @@
-!
-! computes a convolution of V = w over A
-!
-! We use a local orientation
-!
-
-!
-! nHalf is half of odd filter size
-!
-subroutine filter_1d(V, A, w, nHalf)
-   use :: ForOpenCL
-   implicit none
-   real, dimension(:) :: V, A
-   integer :: nHalf, halo(2)
-   real :: A_local(-nHalf:nHalf), w(-nHalf:nHalf)
-
-   halo = nHalf
-   A_local = local_region(A, halo, halo)
-   local(V) = sum(w*A_local)
-
-end subroutine filter_1d
-
-subroutine filter_2d(V, A, w, nPad)
-   use :: ForOpenCL
-   implicit none
-   real, dimension(:,:) :: V, A
-   integer :: nPad, halo(4)
-   real :: A_local(-nPad:nPad,-nPad:nPad), w(-nPad:nPad,-nPad:nPad)
-
-   halo = nPad
-   A_local = local_region(A, halo, halo)
-   local(V) = sum(w*A_local)
-
-end subroutine filter_2d
-
-!
-! A's dimensions are scaled (expanded) relative to V
-! nHalf is half of odd filter size (A's dimension)
-!
-subroutine filter_with_expand_1d(V, A, w, nHalf, scale)
-   use :: ForOpenCL
-   implicit none
-   real, dimension(:) :: V, A
-   integer :: nHalf, scale, halo(2)
-   real :: A_local(-nHalf:nHalf), w(-nHalf:nHalf)
-
-   halo = nHalf
-   A_local = local_region(A, halo, halo, scale)
-   local(V) = sum(w*A_local)
-
-end subroutine filter_with_expand_1d

### 2011-10-27 papers/local-orientation/code/convolve/convolve.f90

@@ -1,191 +1,0 @@
-module convolve_mod
-   use FilterParams
-   use FilterType
-
-contains
-
-   subroutine convolve(S, Image, F)
-      implicit none
-      real :: S(NX,NY), Image(NXEX,NYEX)
-      type(FilterPatch) :: F
-      real :: val
-      integer :: i, j, ip, jp
-
-      do j = 1, NY
-         do i = 1, NX
-            val = 0.0
-            do jp = -NPAD, NPAD
-               do ip = -NPAD, NPAD
-                  val = val + F%p(ip,jp)*Image(i+NPAD,j+NPAD)
-               end do
-            end do
-            S(i,j) = val
-         end do
-      end do
-   end subroutine convolve
-
-   subroutine init_filter(F)
-      implicit none
-      type(FilterPatch) :: F
-      integer :: i
-
-      F%p = 0.0
-      F%p(:,0) = 1
-      F%p(0,:) = 1
-      F%p = F%p/sum(F%p)         
-
-      print *, nxp, nyp, NPAD
-      do i = -1, 1
-         print *, "F==", F%p(i,-1:1)
-      end do
-
-    end subroutine init_filter
-
-end module convolve_mod
-
-
-program test_convolve
-   use convolve_mod
-   use OpenCL
-   use Timer_mod
-   implicit none
-
-   integer :: status
-
-   real(c_float),     target  :: I(NXEX,NYEX)  ! image
-   real(c_float),     target  :: S(NX,NY)      ! smoothed image
-   type(FilterPatch), target  :: F             ! filter
-
-   type(CLDevice) :: device
-   type(CLKernel) :: kernel
-   type(CLBuffer) :: d_I, d_S, d_F
-
-   integer(c_size_t) :: nxLocal=NXL, nyLocal=NXL
-   integer(c_size_t) :: filter_mem_size    = NXP*NYP * SIZE_ELEMENT
-   integer(c_size_t) :: global_mem_size    = NX*NY*SIZE_ELEMENT
-   integer(c_size_t) :: global_ex_mem_size = NXEX*NYEX*SIZE_ELEMENT
-   integer(c_size_t) :: tile_mem_size      = (NXL+2*NPAD)*(NYL+2*NPAD) * SIZE_ELEMENT
-
-   type(CPUTimer) :: timer
-   real(c_double) :: cpu_time, gpu_time
-
-   integer :: nxGlobal=NX, nyGlobal=NY
[…]

### 2011-10-27 papers/local-orientation/code/convolve/convolve_concurrent.f90

@@ -1,17 +1,0 @@
-module Filter
-   integer, parameter :: N = 1
-contains
-
-   pure CONCURRENT subroutine convolve(Image, Filter, indices)
-      real, intent(in out), HALO(:,:) :: Image(:,:)
-      real, intent(in) :: Filter(-N:N,-N:N)
-      integer :: indices(2)
-
-      integer :: i = indices(1)
-      integer :: j = indices(2)
-
-      Image(i,j) = sum(Filter * Image(i-N:i+N,j-N:j+N))
-
-   end subroutine convolve
-
-end module Filter

### 2011-10-27 papers/local-orientation/code/convolve/convolve_elemental.f90

@@ -1,21 +1,0 @@
-module Filter
-
-   integer, parameter :: N
-
-   type :: FilterPatch
-      real :: p(-N:N,-N:N)
-   end type FilterPatch
-
-contains
-
-   pure elemental subroutine convolve(SmoothedImage, Image, Filter)
-      real, intent(out) :: SmoothedImage
-      real, intent(in), EXTEND :: Image(-N:N,-N:N)
-      type(FilterPatch), intent(in) :: Filter
-
-      SmoothedImage = sum(Filter%p * Image)
-
-   end subroutine convolve
-
-
-end module Filter

### 2011-10-27 papers/local-orientation/code/convolve/convolve_forall.f90

@@ -1,123 +1,0 @@
-module convolve_forall
-
-   integer, parameter :: N = 3  ! number of rows    in image
-   integer, parameter :: M = 3  ! number of columns in image
-   integer, parameter :: H = 1   ! size of halo
-
-   ! Number of rows in processor topology
-   ! This should be a command line argument somehow
-   integer, parameter :: PROC_ROWS = 4
-
-!$OFP real, allocatable, HALO(:,:) :: Image(:,:) [PROC_ROWS,*]
-      real, allocatable            :: Image(:,:)
-      real, allocatable            :: Filter(:,:)
-
-contains
-
-
-!===========================================================================
-! Description: Initializes the allocatable arrays.
-!
-! Note about semantics of LOPe: Would like to require all array variables
-! to be used in a concurrent procedure to be allocatable.  This allows
-! allocation for OpenCL buffers to be done at the same time.
-!
-! Note that the image is a coarray and the halo is allocated as part of the
-! array.  The halo is associated with the array but is not necessarily
-! made contiguous with the array proper.  The decision of how to lay out the
-! halo should be up to the compiler.  For example, all of the halo elements
-! may be placed in contiguous memory separate from the array for more
-! efficient memory copies of the halo.  The halo is assumed to be involved
-! in frequent communication between coarray images and between the OpenCL
-! host and device.
-!
-!===========================================================================
-
-   subroutine initialize
-!$OFP allocate( Image(N,M)[PROC_ROWS,*] HALO(H:H,H:H) )
-      allocate( Image(1-H:N+H,1-H:M+H) )
-      allocate( Filter(-H:H,-H:H) )
-
-      ! initialize filter and image arrays
-      !
-      Image  = 1.0
-      Filter = 2*1.0 / ((1+2*H)*(1+2*H))
-
-   end subroutine initialize
-
-   subroutine output_image
-      integer :: row
-      do row = 1-H, N+H
-         print *, Image(row,:)
-      end do
-   end subroutine
-
-
-!===========================================================================
-!
-! Kernel function that performs the convolution for the array element
-! specified by indices.
-!
-!===========================================================================
-
-!$OFP pure CONCURRENT subroutine convolve(Image, Filter, indices)
[…]

### 2011-10-27 papers/local-orientation/code/convolve/convolve_lope.f90

@@ -1,93 +1,0 @@
-module Convolve_LOPe
-
-   integer, parameter :: N = 64  ! number of rows    in image
-   integer, parameter :: M = 64  ! number of columns in image
-   integer, parameter :: H = 1   ! size of halo
-
-   ! Number of rows in processor topology
-   ! This should be a command line argument somehow
-   integer, parameter :: PROC_ROWS = 4
-
-   real, allocatable, HALO(:,:) :: Image(:,:) [*,*]    ! [*,*] is new syntax
-   real, allocatable            :: Filter(:,:)
-
-contains
-
-
-!===========================================================================
-! Description: Initializes the allocatable arrays.
-!
-! Note about semantics of LOPe: Would like to require all array variables
-! to be used in a concurrent procedure to be allocatable.  This allows
-! allocation for OpenCL buffers to be done at the same time.
-!
-! Note that the image is a coarray and the halo is allocated as part of the
-! array.  The halo is associated with the array but is not necessarily
-! made contiguous with the array proper.  The decision of how to lay out the
-! halo should be up to the compiler.  For example, all of the halo elements
-! may be placed in contiguous memory separate from the array for more
-! efficient memory copies of the halo.  The halo is assumed to be involved
-! in frequent communication between coarray images and between the OpenCL
-! host and device.
-!
-!===========================================================================
-
-   subroutine initialize
-      allocate( Image(N,M)[*,*] HALO([H,H],[H,H]) )
-      allocate( Filter(-H:H,-H:H) )
-
-      ! initialize filter and image arrays
-      !
-      Image  = 1.0
-      Filter = 1.0 / ((1+2*H)*(1+2*H))
-
-   end subroutine initialize
-
-   subroutine output_image
-      integer :: row
-      do row = 1-H, N+H
-         print *, Image(row,:)
-      end do
-   end subroutine
-
-
-!===========================================================================
-!
-! Kernel function that performs the convolution for the array element
-! specified by indices.
-!
-!===========================================================================
-
-pure CONCURRENT subroutine convolve(Image, Filter)
-
-   real, intent(in out), HALO(:,:) :: Image(0:,0:)
-   real, intent(in)                :: Filter(-H:H,-H:H)
-
-   Image(0,0) = sum(Filter * Image(-H:+H,-H:+H))
-
-end subroutine convolve
-
-end module Convolve_LOPe
-
-
-program convolve_image
[…]

### 2011-10-27 papers/local-orientation/code/convolve/filter_params.F90

@@ -1,23 +1,0 @@
-#include "npad.h"
-
-module FilterParams
-   use ISO_C_BINDING
-
-   ! WARNING, must also change in convolve.cl
-   integer(c_int), parameter :: NPAD = PADDING
-
-   integer, parameter :: G_WIDTH = 512
-
-   integer(c_int), parameter :: NXP = 1 + 2*NPAD
-   integer(c_int), parameter :: NYP = NXP
-
-   integer(c_size_t), parameter :: NX   = G_WIDTH
-   integer(c_size_t), parameter :: NY   = G_WIDTH
-   integer(c_size_t), parameter :: NXEX = NX + 2*NPAD
-   integer(c_size_t), parameter :: NYEX = NY + 2*NPAD
-   integer(c_size_t), parameter :: NXL  = 16
-   integer(c_size_t), parameter :: NYL  = 16
-
-   integer(c_size_t), parameter :: SIZE_ELEMENT = 4
-
-end module FilterParams

### 2011-10-27 papers/local-orientation/code/convolve/filter_type.f90

@@ -1,21 +1,0 @@
-module FilterType
-   use FilterParams
-
-   type :: FilterPatch
-      real :: p(-NPAD:NPAD,-NPAD:NPAD)
-   end type FilterPatch
-
-contains
-
-   pure elemental subroutine convolve_elemental(SmoothedImage, Image, Filter)
-      real, intent(out) :: SmoothedImage
-      !$OFP EXTEND :: Image(-N:N,-N:N)
-      real, intent(in) :: Image
-      type(FilterPatch), intent(in) :: Filter
-
-      ! SmoothedImage = sum(Filter%p * Image)
-      SmoothedImage = Filter%p(1,1) * Image
-
-   end subroutine convolve_elemental
-
-end module FilterType

### 2011-10-27 papers/local-orientation/code/convolve/npad.h

@@ -1,1 +1,0 @@
-#define PADDING 7

### 2011-10-27 papers/local-orientation/code/convolve/variable_filter.f90

@@ -1,224 +1,0 @@
-module variable_filter_mod
-   use FilterParams
-   use FilterType
-
-contains
-
-   subroutine convolve(S, Image, F)
-      implicit none
-      real :: S(NX,NY), Image(NXEX,NYEX)
-      type(FilterPatch) :: F(NX,NY)
-      real :: val
-      integer :: i, j, ip, jp
-
-      do j = 1, NY
-         do i = 1, NX
-            val = 0.0
-            do jp = -NPAD, NPAD
-               do ip = -NPAD, NPAD
-                  val = val + F(i,j)%p(ip,jp)*Image(i+NPAD,j+NPAD)
-               end do
-            end do
-            S(i,j) = val
-         end do
-      end do
-   end subroutine convolve
-
-   subroutine filter_bandwidth(S, Image, F)
-      implicit none
-      real :: S(NX,NY), Image(NXEX,NYEX)
-      type(FilterPatch) :: F(NX,NY)
-      real :: val
-      integer :: i, j, ip, jp
-
-      do j = 1, NY
-         do i = 1, NX
-            val = 0.0
-            do jp = -NPAD, NPAD
-               do ip = -NPAD, NPAD
-                  F(i,j)%p(ip,jp) = 1
-               end do
-            end do
-            S(i,j) = 1.0
-         end do
-      end do
-   end subroutine filter_bandwidth
-
-
-   subroutine init_filter(F)
-      implicit none
-      type(FilterPatch) :: F(NX,NY)
-      integer :: i, j
-
-      F(1,1)%p = 0.0
-      F(1,1)%p(:,0) = 1
-      F(1,1)%p(0,:) = 1
-      F(1,1)%p = F(1,1)%p/sum(F(1,1)%p)         
-
-      do j = 1, NY
-         do i = 1, NX
-            F(i,j)%p = F(1,1)%p
-         end do
-      end do
-
-      print *, nxp, nyp, NPAD
-      do i = -1, 1
-         print *, "F==", F(2,2)%p(i,-1:1)
-      end do
-      print *
-
-    end subroutine init_filter
-
-end module variable_filter_mod
-
-
-program test_variable_filter
-   use variable_filter_mod
-   use OpenCL
-   use Timer_mod
-   implicit none
-
-   integer :: status
-
[…]

### 2011-10-27 papers/local-orientation/code/elem_type.f90

@@ -1,44 +1,0 @@
-module CFD
-
-  type :: snode
-    real :: val
-  end type snode
-
-  type :: node
-    real :: val
-    class(node), pointer :: child(:)
-  end type node
-
-  type, extends(snode) :: cnode
-    class(node), pointer :: child(:)
-  end type cnode
-
-contains
-
-  real function myval(this)
-     class(cnode), intent(in) :: this
-     myval = this%val
-
-     select type(this)
-!     type is (snode)
-     type is (cnode)
-     end select
-
-  end function myval
-
-end module CFD
-
-pure elemental function value(a)
-   use CFD
-   real :: value
-   class(node), intent(in) :: a
-
-   if (associated(a%child)) then
-      value = sum(a%child(1:4)%val) / 4.0
-   else
-      value = a%val
-   end if
-   
-
-end function value
-

### 2011-10-27 papers/local-orientation/code/elemental_add.f90

@@ -1,36 +1,0 @@
-!
-! Eventual form of high-level representation.  This will first be turned
-! into syntax for OFP and ROSE and then, hopefully it can make its way
-! into the Fortran standard.
-!
-! pure elemental CONCURRENt subroutine elemental_add(A, B, C)
-!    TYPE(*), intent(COPYIN)  :: A, B
-!    TYPE(*), intent(COPYOUT) :: C
-!
-!   C = A + B   
-!
-!end subroutine elemental_add
-
-
-!
-! Note that this function is dimensionless.  The function parameters are scalars
-! but the compiler will generate code for arrays of a particular dimension based
-! on the actual parameters at the calling site.  Also note that the parameters
-! are typeless (see TYPE(*) above); the compiler will create an instance of
-! function according to the called type.
-!
-pure elemental subroutine elemental_add(A, B, C, s)
-!   !
-!   ! compiler directives that will eventually be turned into syntax
-!   ! 
-!   !$OFP CONCURRENT :: elemental_add
-!   !$OFP intent(COPYIN)   :: A, B
-!   !$OFP intent(COPYOUT)  :: C
-!   !
-!
-   real, intent(in), value :: A, B, s
-   real, intent(out) :: C
-
-   C = A + s*B   
-
-end subroutine elemental_add

### 2011-10-27 papers/local-orientation/code/filter.f90

@@ -1,39 +1,0 @@
-!
-! A 2x2 filter applied over an array 2x2 larger than destination
-!
-!  V = V + sum(w*A)
-!
-subroutine filter_2x2(nPad, V, A, w)
-  use :: ForOpenCL
-  implicit none
-  integer, intent(in) :: nPad
-  real, dimension(:,:) :: V, A  ! V 2x2 bigger
-  real, dimension(2,2) :: w
-
-  real, dimension(:,:), allocatable :: A_big
-
-  ! scatter A to array compatible with V
-  !
-  A_big = scatter(identity, A, 2)
-
-end subroutine filter_2x2
-
-
-!subroutine convolve(nPad, V, A, w)
-   ! nx = get_global_size(0), ny ... is default, otherwise explicit
-!   real, dimension(:,:) :: V, A, w
-!   asynchronous :: A     ! block on event?
-
-!   real, pointer, dimension(1+2*nPad,1+2*nPad) :: lA
-
-!   halo   = [nPad,nPad,nPad,nPad]
-!   filter = [nPad,nPad,nPad,nPad]
-
-   ! A is extended
-   ! iA => halo(A, aHalo)      ! k,kl indexes
-
-!   lA => halo(A, halo, local=filter)
-
-!   V = V + sum(w*lA)         ! this does convolution
-
-!end subroutine convolve

### 2011-10-27 papers/local-orientation/code/finite_difference.f90

@@ -1,20 +1,0 @@
-!
-! computes derivative of A using centered difference
-!
-! I believe A_prime(i) = ( A(i+1) - A(i-1) ) / (2*dx)
-!
-! We use a local orientation
-!
-subroutine finite_difference(A, A_prime, dx)
-   use :: ForOpenCL
-   implicit none
-   real, dimension(:) :: A, A_prime
-   real :: dx, A_local(-1:1)
-   real, pointer :: A_prime_local          ! this will go away
-
-   A_local = local_region(A, [1,1,1,1], halo=[1,1,1,1])   ! A needs halo specification
-   A_prime_local => local(A_prime)
-
-   A_prime_local = ( A_local(1) - A_local(-1) ) / (2*dx)
-
-end subroutine finite_difference

### 2011-10-27 papers/local-orientation/code/forall.f90

@@ -1,16 +1,0 @@
-program test_forall
-   implicit none
-   integer, parameter :: ng = 2
-   integer, parameter :: is = 1, ie = 10
-   integer, parameter :: km = 100
-
-   real, dimension(is-ng:ie+ng,km) :: delp
-   real, dimension(is:ie,km) :: dm
-
-   integer :: k, i
-
-   forall(k=1:km, i=is:ie)
-      dm(i,k) = delp(i,k-1)
-   end forall
-
-end program

### 2011-10-27 papers/local-orientation/code/heat/adv_time.f90

@@ -1,16 +1,0 @@
-!------------------------
-! Proposed syntax for the concurrent function adv_time
-!------------------------
-
-pure elemental function adv_time(T)
-   real, EXTEND, intent(in)  :: T(-1:1,-1:1)
-   real :: adv_time
-   real :: lhs1, lhs2
-
-   lhs1 = ( T(-1,0) - 2.0*T(0,0) + T(+1,0) ) / (dx*dx)
-   lhs2 = ( T(0,-1) - 2.0*T(0,0) + T(0,+1) ) / (dy*dy)
-
-   adv_time = (alpha * dt * (lhs2 + lhs2)) + T(0,0)
-
- end function adv_time
-    

### 2011-10-27 papers/local-orientation/code/heat/test_heat2d.f90

@@ -1,63 +1,0 @@
-!
-!-----  Fortran Programming Language with local extensions -----
-!
-! This file is distributed under an open source license by Los Alamos
-! National Security, LCC.  See the file License.txt (located in the
-! top level of the source distribution) for details.
-! 
-!-----
-!
-! Simplistic 2D heat transfer...
-! 
-!
-
-module heat_transfer
-
-   integer, parameter :: NPAD = 2
-   integer, parameter :: WIDTH  = 1024
-   integer, parameter :: HEIGHT = 1024
-   integer, parameter :: NX = WIDTH  + 2*NPAD
-   integer, parameter :: NY = HEIGHT + 2*NPAD
-
-   integer, parameter :: NTIME_STEPS = 1000
-   real,    parameter :: MAX_TEMP    = 100.0
-
-   interface
-      pure elemental function adv_time(T)
-         real, intent(in) :: T
-      end function adv_time
-   end interface
-
-end module heat_transfer
-
-
-program test_heat2d
-   use :: heat_transfer
-   implicit none
-   integer :: i
-
-   real, dimension(0:NX-1,0:NY-1) :: T1, T2
-  
-   real, parameter :: alpha = .001
-   real, parameter :: dx = 1.0 / WIDTH
-   real, parameter :: dy = 1.0 / HEIGHT
-   real, parameter :: dt = .01 * (alpha / 4.0) * ((1.0 / (dx*dx)) + (1.0 / (dy*dy)))
-
-   ! initialize cell temperature
-   !
-   T1 = MAX_TEMP
-   T2 = 0.0
-
-   T1(:,0)    = 0.0
-   T1(:,NY-1) = 0.0
-   T1(0,:)    = 0.0
-   T1(NX-1,:) = 0.0
-
-   ! time loop
-   !
-   do i = 1, NTIME_STEPS/2
-      T2(1:WIDTH,1:HEIGHT) = adv_time(T1(1:WIDTH,1:HEIGHT))
-      T1(1:WIDTH,1:HEIGHT) = adv_time(T2(1:WIDTH,1:HEIGHT))
-   end do
-
-end program test_heat2d

### 2011-10-27 papers/local-orientation/code/nh_core.F90

@@ -1,285 +1,0 @@
-  subroutine Riem_3D(ns, bdt, is, ie, js, je, ng, j, km, cp, gama, cappa, p3, dm2,    &
-                     pm2, w, dz2, pt, quick_p, c_core, ktop, iad)
-
-  integer, intent(in):: ns, is, ie, js, je, ng,  km, j
-  integer, intent(in):: iad      ! time step scheme 
-  integer, intent(in):: ktop     ! starting layer for non-hydrostatic dynamics
-                                 ! 1: All non-hydrostatic
-                                 ! 2: top sponge layer is hydrostatic
-  real,    intent(in):: bdt, cp, gama, cappa
-  real,    intent(in), dimension(is:ie,km):: dm2, pm2
-  logical, intent(in):: quick_p       ! fast algorithm for pressure
-  logical, intent(in):: c_core
-  real, intent(in  ) :: pt (is-ng:ie+ng,js-ng:je+ng,km)
-! IN/OUT:
-  real, intent(inout):: dz2(is:ie,km)
-  real, intent(inout)::   w(is-ng:ie+ng,js-ng:je+ng,km)
-  real, intent(out  )::  p3(is-ng:ie+ng,js-ng:je+ng,km+1)
-! --- Local 1D copyies -----
-#ifdef USE_2D
-  real, dimension(km,is:ie):: t2, p2, pt2
-#else
-  real, dimension(km):: c2, p2, pt2
-#endif
-  real, dimension(km):: r_p, r_n, rden, dz, dm, wm, dts, pdt
-  real, dimension(km+1):: m_bot, m_top, r_bot, r_top, time_left, pe1, pbar, wbar
-
-  real, parameter:: dzmx = 0.5*dz_max
-  real    :: dt, rdt, grg, z_frac, t_left
-  real    :: a1, b1, g2, rcp
-  real    :: seq(ns)       ! time stepping sequence
-  integer :: k2(km+1)
-  integer :: i, k, n, ke, kt, k0, k1, k3
-
-  call time_sequence( iad, ns, bdt, seq )
-
-  grg = gama * rdgas  
-  rcp = 1. / cp
-  rdt = 1. / bdt
-
-  if ( quick_p ) then
-       a1 = 0.5               ! a1=1 fully implicit
-       b1 = 1. - a1
-       g2 = -2.*gama
-  endif
-
-#ifdef USE_2D
-  do i=is,ie
-     do k=ktop,km
-        rden(k) = -rdgas*dm2(i,k)/dz2(i,k)
[…]

### 2011-10-27 papers/local-orientation/code/nh_core3d_forall.f90

@@ -1,80 +1,0 @@
-module constants_mod
-   real, parameter :: grav = 9.8      ! real value doesn't matter for now
-end module constants_mod
-
-
-module nh_core_forall_mod
-
-! Notes:
-! This module is used to test programming model concepts for
-! programming accelerated processors.  The central idea is for
-! the programming model to restrict the programmer so that
-! the compiler is able to generate efficient massively threaded
-! code.
-!
-
-   use constants_mod
-
-   implicit none
-   private
-
-   public Riem_3D_forall
-   real, parameter:: dz_max = -0.5               ! (meters)
-
-CONTAINS 
-
-  subroutine Riem_3D_forall(ns, bdt, is, ie, js, je, ng, km, cp, gama, cappa, p3, dm3,    &
-                            pm3, w, dz2, pt, quick_p, c_core, ktop, iad)
-!--------------------------------------------
-! !OUTPUT PARAMETERS
-! Output: gz: grav*height at edges
-!--------------------------------------------
-   implicit none
-  integer, intent(in):: ns, is, ie, js, je, ng,  km
-  integer, intent(in):: iad      ! time step scheme 
-  integer, intent(in):: ktop     ! starting layer for non-hydrostatic dynamics
-                                 ! 1: All non-hydrostatic
-                                 ! 2: top sponge layer is hydrostatic
-  real,    intent(in):: bdt, cp, gama, cappa
-  real,    intent(in), dimension(is:ie,js:je,km):: dm3, pm3
-  logical, intent(in):: quick_p       ! fast algorithm for pressure
-  logical, intent(in):: c_core
-  real, intent(in  ) :: pt (is-ng:ie+ng,js-ng:je+ng,km)
-! IN/OUT:
-  real, intent(inout):: dz2(is:ie,km)
-  real, intent(inout)::   w(is-ng:ie+ng,js-ng:je+ng,km)
-  real, intent(out  )::  p3(is-ng:ie+ng,js-ng:je+ng,km+1)
-
-! --- Local 1D copies -----
-  real, dimension(km):: c2, p2, pt2
-  real, dimension(km):: r_p, r_n, rden, dz, dm, wm, dts, pdt
-  real, dimension(km+1):: m_bot, m_top, r_bot, r_top, time_left, pe1, pbar, wbar
-
-! Local scalars:
-   integer :: i, j, k
-
-!
-! The first problem to solve is what is the dimension and size of
-! the forall loop.  This is important because it relates to the
[…]

### 2011-10-27 papers/local-orientation/code/nh_core_forall.f90

@@ -1,77 +1,0 @@
-module constants_mod
-   real, parameter :: grav = 9.8      ! real value doesn't matter for now
-end module constants_mod
-
-
-module nh_core_forall_mod
-
-! Notes:
-! This module is used to test programming model concepts for
-! programming accelerated processors.  The central idea is for
-! the programming model to restrict the programmer so that
-! the compiler is able to generate efficient massively threaded
-! code.
-!
-
-   use constants_mod
-
-   implicit none
-   private
-
-   public Riem_Solver_forall
-   real, parameter:: dz_max = -0.5               ! (meters)
-
-CONTAINS 
-
-  subroutine Riem_Solver_forall(dt, is, ie, js, je, km, ng, &
-                                akap, ptop, hs,             &
-                                delp, gz)
-!--------------------------------------------
-! !OUTPUT PARAMETERS
-! Output: gz: grav*height at edges
-!--------------------------------------------
-   implicit none
-   integer, intent(in):: is, ie, js, je, km, ng
-   real, intent(in):: dt         ! the BIG horizontal Lagrangian time step
-   real, intent(in):: akap, ptop
-   real, intent(in):: hs(is-ng:ie+ng,js-ng:je+ng)
-   real, intent(inout), dimension(is-ng:ie+ng,js-ng:je+ng,km):: delp
-   real, intent(out), dimension(is-ng:ie+ng,js-ng:je+ng,km+1):: gz
-
-! Local:
-   real, dimension(is:ie,km):: dm
-   real    :: gama, rgrav, ptk
-   integer :: i, j, k
-
-!
-! The first problem to solve is what is the dimension and size of
-! the forall loop.  This is important because it relates to the
-! number of threads running the problem.  It looks like this is
-! fundamentally a 3D problem.
-!
-
-   forall (j=js:je, k=1:km, i=is:ie)
-
-!
-! Scalar assignments, not a problem as these scalars
-! will have local storage and will be calculated for each thread.
-! So logically they are within the loop which leads to compiler
-! warnings.
-!
-       gama = 1./(1.-akap)
-      rgrav = 1./grav
-        ptk = ptop ** akap
-
-!
-! However, this is a problem as the arrays aren't the same size.
-! This means we have to rethink the problem.  We could treat
-! this as a 2D problem and call the kernel procedure for different
-! j values.
-!
-      dm(i,k) = delp(i,j,k)
-
-   end forall
-
-end subroutine Riem_Solver_forall
-
-end module nh_core_forall_mod

### 2011-10-27 papers/local-orientation/code/nh_core_simple.F90

@@ -1,497 +1,0 @@
-module nh_core_mod
-
-! Notes:
-! Using k_top=2 to treat the top layer hydrostatically so that delz will
-! be computed using hydrostatic balance (instead of the update by
-! advection of height using extrapolated winds at the model top)
-!
-! To do list:
-! include moisture effect in pt
-!------------------------------
-
-   use constants_mod
-
-   implicit none
-   private
-
-   public Riem_Solver
-   real, parameter:: dz_max = -0.5               ! (meters)
-
-CONTAINS 
-
-  subroutine Riem_Solver(dt,   is,   ie,   js, je, km, ng,    &
-                         akap, cp,   ptop, hs, peln, w,  delz, pt,  &
-                         delp, gz,   pkc, pk, pe, last_call, ip)
-!--------------------------------------------
-! !OUTPUT PARAMETERS
-! Ouput: gz: grav*height at edges
-!        pe: full     hydrostatic pressure
-!       pkc: full non-hydrostatic pressure
-!--------------------------------------------
-   implicit none
-   integer, intent(in):: is, ie, js, je, km, ng
-   integer, intent(in):: ip      ! ip==0 pkc is perturbation pressure
-   real, intent(in):: dt         ! the BIG horizontal Lagrangian time step
-   real, intent(in):: akap, cp, ptop
-   real, intent(in):: hs(is-ng:ie+ng,js-ng:je+ng)
-   logical, intent(in):: last_call
-   real, intent(inout), dimension(is-ng:ie+ng,js-ng:je+ng,km):: w, delp, pt
-   real, intent(inout):: delz(is:ie,js:je,km)
-   real, intent(out), dimension(is-ng:ie+ng,js-ng:je+ng,km+1):: gz, pkc
-   real, intent(out):: pk(is:ie,js:je,km+1)
-   real, intent(out):: pe(is-1:ie+1,km+1,js-1:je+1)
-   real, intent(out):: peln(is:ie,km+1,js:je)           ! ln(pe)
-! Local:
-  real, dimension(is:ie,km):: pm, dm, dz2
-  real :: pem(is:ie,km+1)
-  real gama, rgrav, ptk
-  integer i, j, k
-
[…]

### 2011-10-27 papers/local-orientation/code/petavision/recv_synaptic_input.f90

@@ -1,236 +1,0 @@
-module variable_filter_mod
-   use WeightParams
-   use WeightType
-
-contains
-
-   subroutine recv_synaptic_input(G, A, W)
-      implicit none
-      integer :: G(NXEX,NYEX)
-      real :: A(NX,NY)
-      type(WeightPatch) :: W(NX,NY)
-      integer :: i, j, ip, jp
-
-      ! assumes activity==1
-      do j = 1, NY
-         do i = 1, NX
-            if (A(i,j) > 0.0) then
-               do jp = -NPAD, NPAD
-                  do ip = -NPAD, NPAD
-                     G(i+NPAD+ip,j+NPAD+jp) = G(i+NPAD+ip,j+NPAD+jp) + W(i,j)%p(ip,jp)   ! *A(i,j)
-                  end do
-               end do
-            end if
-         end do
-      end do
-   end subroutine recv_synaptic_input
-
-   subroutine init_weights(W)
-      implicit none
-      type(WeightPatch) :: W(NX,NY)
-      integer :: i, j
-
-      W(1,1)%p = 0.0
-      W(1,1)%p(:,0) = 1
-      W(1,1)%p(0,:) = 1
-      W(1,1)%p = W(1,1)%p/sum(W(1,1)%p)         
-
-      do j = 1, NY
-         do i = 1, NX
-            W(i,j)%p = 1
-         end do
-      end do
-
-      print *, nxp, nyp, NPAD
-      do i = -1, 1
-         print *, "W==", W(2,2)%p(i,-1:1)
-      end do
-      print *
-
-   end subroutine init_weights
-
-   subroutine init_activity(A, limit)
-      implicit none
-      real :: A(NX,NY), limit
-
-      call random_number(A)
-      where (A > limit)
-         A = 1.0
-      else where
-         A = 0.0
-      end where
-
-      print *, A(1,:8)
-      print *, A(2,:8)
-      print *, A(3,:8)
-
-   end subroutine init_activity
-
-end module variable_filter_mod
-
-
-program test_recv_synaptic_input
-   use variable_filter_mod
-   use OpenCL
-   use Timer_mod
-   implicit none
-
-   integer :: status
-
-   real(c_float),     target  :: A(NX,NY)      ! pre-synaptic activity
[…]

### 2011-10-27 papers/local-orientation/code/petavision/test_atomics.f90

@@ -1,62 +1,0 @@
-program test_atomics
-   use OpenCL
-   implicit none
-
-   integer :: status
-
-   integer(c_int), target :: I
-
-   type(CLDevice) :: device
-   type(CLKernel) :: kernel
-   type(CLBuffer) :: d_I
-
-   integer, parameter :: NXL = 16
-   integer, parameter :: NYL = 16
-   integer, parameter :: NX = 64
-   integer, parameter :: NY = 64
-
-   integer(c_size_t) :: global_mem_size = 4
-   integer(c_size_t) :: nxLocal=NXL, nyLocal=NXL
-   integer(c_size_t) :: nxGlobal=NX, nyGlobal=NY
-   integer :: device_id
-
-   device_id = 0  ! 0=GPU, 1=CPU
-   status = init(device, device_id)
-
-   status = query(device)
-
-   call limitLocalSize(device, nxLocal, nyLocal)
-
-   ! initialize memory
-   !
-
-   I = 0;
-
-   ! create memory buffers
-   !
-   d_I = createBuffer(device, CL_MEM_COPY_HOST_PTR, global_mem_size, c_loc(I))
-
-   ! create the kernel
-   !
-   kernel = createKernel(device, &
-                         "test_atomics.cl" // C_NULL_CHAR, &
-                         "atomic_inc"      // C_NULL_CHAR)
-
-   ! add arguments
-   !
-   status = setKernelArgMem(kernel, 0, clMemObject(d_I))
-
-   ! run the kernel on the device
-   !
-
-   print *
-   status = run(kernel, nxGlobal, nyGlobal, nxLocal, nyLocal)
-   status = clFinish(kernel%commands)
-
-   ! get the results
-   !
-   status = readBuffer(d_I, c_loc(I), global_mem_size)
-   
-   print *, "results==", I
-
-end program test_atomics

### 2011-10-27 papers/local-orientation/code/petavision/weight_params.f90

@@ -1,21 +1,0 @@
-module WeightParams
-   use ISO_C_BINDING
-
-   ! WARNING, must also change in recv_synaptic_input.cl
-   integer(c_int), parameter :: NPAD = 3
-
-   integer, parameter :: G_WIDTH = 128
-
-   integer(c_int), parameter :: NXP = 1 + 2*NPAD
-   integer(c_int), parameter :: NYP = NXP
-
-   integer(c_size_t), parameter :: NX   = G_WIDTH
-   integer(c_size_t), parameter :: NY   = G_WIDTH
-   integer(c_size_t), parameter :: NXEX = NX + 2*NPAD
-   integer(c_size_t), parameter :: NYEX = NY + 2*NPAD
-   integer(c_size_t), parameter :: NXL  = 16
-   integer(c_size_t), parameter :: NYL  = 16
-
-   integer(c_size_t), parameter :: SIZE_ELEMENT = 4
-
-end module WeightParams

### 2011-10-27 papers/local-orientation/code/petavision/weight_type.f90

@@ -1,21 +1,0 @@
-module WeightType
-   use WeightParams
-
-   type :: WeightPatch
-      real :: p(-NPAD:NPAD,-NPAD:NPAD)
-   end type WeightPatch
-
-contains
-
-   pure elemental subroutine convolve_elemental(SmoothedImage, Image, Filter)
-      real, intent(out) :: SmoothedImage
-      !$OFP EXTEND :: Image(-N:N,-N:N)
-      real, intent(in) :: Image
-      type(WeightPatch), intent(in) :: Filter
-
-      ! SmoothedImage = sum(Filter%p * Image)
-      SmoothedImage = Filter%p(1,1) * Image
-
-   end subroutine convolve_elemental
-
-end module WeightType

### 2011-10-27 papers/local-orientation/code/reduce.f90

@@ -1,30 +1,0 @@
-!
-! I'm not sure about overloaded functions and function names for these.
-! There will potentiall be several versions, depending on scale factor.
-!
-
-! for now this reduction assumes a 2x2 expansion
-
-Module ForOpenCL
-
-interface
-
-   subroutine reduce(f, A_big, A)
-      real, dimension(:,:) :: A_big, A
-      interface
-         pure elemental function f(a, b)
-            real :: a, b, f
-            intent(in) :: a, b
-         end function
-      end interface
-   end subroutine reduce
-
-   subroutine scatter(A, A_big)
-      real, dimension(:,:) :: A, A_big
-   end subroutine scatter
-
-end interface
-
-end module ForOpenCL
-
-

### 2011-10-27 papers/local-orientation/code/shallow.f90

@@ -1,55 +1,0 @@
-module shallow_mod
-
-CONTAINS
-
-subroutine shallow(is, ie, js, je, ks, ke, velocity)
-!$OFP kernel procedure :: shallow
-   implicit none
-   integer :: is, ie, js, je, ks, ke
-   real, dimension(is:ie, js:je, ks:ke) :: velocity
-
-   integer :: i, j, k
-
-   DO CONCURRENT (i=is:ie, j=js:je, k=ks:ke)
-   BLOCK         
-
-      ! local variable declarations (mostly scalar)
-      !
-      real :: l_vel
-
-      ! load local variables (into registers) from arrays
-      !
-      l_vel = velocity(i,j,k)
-
-      ! do computation using local variables
-      !
-      l_vel = l_vel/2.0
-
-      ! store computed values back to arrays
-      !
-      velocity(i,j,k) = l_vel
-
-   END BLOCK
-   END DO CONCURRENT
-
-end subroutine shallow
-
-end module shallow_mod
-
-
-program test_shallow
-   use shallow_mod
-   implicit none
-
-   integer, parameter :: SIZE = 64   
-   real, dimension(SIZE, SIZE, SIZE) :: velocity
-   integer :: is, ie, js, je, ks, ke
-
-   is = 1;    js = 1;    ks = 1;
-   ie = SIZE; je = SIZE; ke = SIZE
-
-   velocity = 1.0
-
-   call shallow(is, ie, js, je, ks, ke, velocity)
-
-end program

### 2011-10-27 papers/local-orientation/code/test_advect.f90

@@ -1,47 +1,0 @@
-program test_advect
-   implicit none
-   integer, parameter :: NX = 16
-   integer :: i
-   real :: H(0:NX+1), a, dx, dt
-
-   interface
-       pure subroutine advect_impl(H_big, halo, a, dx, dt)
-          real, dimension(*), intent(inout) :: H_big   ! includes the halo region
-          real, intent(in) :: a, dx, dt
-          integer, intent(in) :: halo(2)
-       end subroutine
-   end interface
-
-   integer :: nLoops = 500
-
-   a  = 2.0
-   dx = 1.0
-   dt = 0.1 * (dx/a)
-
-   ! boundary conditions
-   !
-!   H(0)    = 1.0
-!   H(NX+1) = 0.0
-   H = 1.0
-
-   ! interior
-   !
-!   do i = 1, NX
-!      H(i) = 1.0 - (i/((NX+1)*dx))
-!   end do
-!   H(NX+1) = H(NX)    ! outflow boundary conditions
-
-   print *
-   print *, "initial conditions"
-   print *, H
-
-   do i = 1, nLoops
-      call advect_impl(H, [1,1], a, dx, dt)
-!      H(NX+1) = H(NX)    ! outflow boundary conditions
-   end do
-
-   print *
-   print *, "final conditions"
-   print *, H
-
-end program

### 2011-10-27 papers/local-orientation/codegen/ElementalTraversal.cpp

@@ -1,763 +1,0 @@
-#include <rose.h>
-#include "ElementalTraversal.hpp"
-#include "FortranAnalysis.hpp"
-
-ElementalTraversal::ElementalTraversal(int argc, char * argv[], const char * filename)
-   : FortranTraversal(filename)
-{
-   line = 0;
-   num_dims = 1;
-   num_arrays = 0;
-   array_list = new char * [1];
-   array_list[0] = NULL;
-
-   // look for -ofp:dims argument
-   //
-   for (int i = 0; i < argc; i++) {
-      char * str = argv[i];
-      if (strncmp(str, "-ofp:dims", 9) != 0) continue;
-      num_dims = atoi(&str[9]);
-      printf("found num_dims==%d str==%s\n", num_dims, str);
-      break;
-   }
-
-   open(filename);
-}
-
-void ElementalTraversal::open(const char * filename)
-{
-   this->fp = fopen(filename, "w");
-   if (fp == NULL) {
-      printf("ElementalTraversal:: ERROR in opening file %s\n", filename);
-   }
-   ROSE_ASSERT(fp != NULL);
-
-   fprintf(fp, "/*\n");
-   fprintf(fp, " * This kernel is automatically generated by Open Fortran Project tools.\n");
-   fprintf(fp, " */\n");
-   fprintf(fp, "\n");
-
-   fprintf(fp, "static inline int get_global_idx_1d()\n");
-   fprintf(fp, "{\n");
-   fprintf(fp, "   return get_global_id(0);\n");
-   fprintf(fp, "}\n\n");
-
-   fprintf(fp, "static inline int get_global_idx_2d()\n");
-   fprintf(fp, "{\n");
-   fprintf(fp, "   return get_global_id(0) + get_global_size(0)*get_global_id(1);\n");
-   fprintf(fp, "}\n");
-}
-
-void ElementalTraversal::visit(SgNode * node)
-{
-   switch (node->variantT())
-   {
-      case V_SgFunctionDefinition  : visit( (SgFunctionDefinition  *) node);  break;
-   }
-}
-
-#ifdef OBSOLETE
-void ElementalTraversal::visit(SgFunctionDeclaration * func_decl)
-{
-   char comma[2] = {0,0};
-
-   std::string func_name = func_decl->get_name();
-   std::string filename = func_name + ".f90";
-   filename = filename.substr(4,filename.size()-1);
-   filename[0] = tolower(filename[0]);
-
-   open(filename.c_str());
-
[…]

### 2011-10-27 papers/local-orientation/codegen/ElementalTraversal.hpp

@@ -1,57 +1,0 @@
-#ifndef ELEMENTAL_TRAVERSAL_HPP_
-#define ELEMENTAL_TRAVERSAL_HPP_
-
-#include "FortranTraversal.hpp"
-
-class ElementalTraversal : public FortranTraversal
-{
-
-public:
-   
-   ElementalTraversal(int argc, char * argv[], const char * filename);
-
-   virtual void visit(SgNode * node);
-
-   void open(const char * filename);
-
-   //   void visit(SgInterfaceBody * ibody);
-   //   void visit(SgFunctionDeclaration * func_decl);
-   void visit(SgFunctionDefinition  * func_def);
-
-   void unparseStmt(SgStatement * stmt);
-
-   void unparseExpr(SgExpression * expr);
-   void unparseBinaryOp(SgBinaryOp * op);
-   void unparseUnaryOp(SgUnaryOp * op);
-   void unparseFuncCallExpr(SgFunctionCallExp * expr);
-   void unparseVarRefExpr(SgVarRefExp * var);
-
-   void unparseVarDecl(SgVariableDeclaration * var_decl, SgFunctionDefinition * func_def);
-   void unparseVarDef(SgVariableDeclaration * var_decl, bool isLocal=false);
-   void unparseParamDef(SgVariableDeclaration * var_decl);
-   void unparseFunctionCall(SgFunctionDefinition * func_def);
-
-   void unparseFuncParam(SgInitializedName * name, SgFunctionDefinition * func_def);
-   void unparseFuncParamReturn(SgFunctionDeclaration * func_decl, const char * comma);
-   void unparseVarDeclListForScalars(SgVariableDeclaration * var_decl, bool isLocal);
-   void unparseVarDeclListForArrays(SgVariableDeclaration * var_decl, bool isLocal);
-   void unparseIntVarDeclStmt(SgVariableDeclaration * var_decl, bool isLocal=false);
-   void unparseFloatVarDeclStmt(SgVariableDeclaration * var_decl, bool isLocal=false);
-
-   void unparseType(SgType *);
-
-   void unparseArrayDescs(SgFunctionDeclaration * func_decl);
-   void unparseIndexVars(SgFunctionDeclaration * func_decl);
-
-
-protected:
-
-   int line;
-   int num_dims;
-
-   int num_arrays;
-   char ** array_list;
-
-};
-
-#endif // ELEMENTAL_TRAVERSAL_HPP_

### 2011-10-27 papers/local-orientation/codegen/FortranAnalysis.cpp

@@ -1,363 +1,0 @@
-#include "FortranAnalysis.hpp"
-
-#define DEBUG_PRINT
-
-//
-// Helper class
-//
-class HaloRefSearch : public AstSimpleProcessing
-{
-public:
-   HaloRefSearch();
-   virtual void visit(SgNode * node);
-   static bool findHaloRef(SgExprStatement * expr_stmt);
-   bool found;
-};
-
-bool HaloRefSearch::findHaloRef(SgExprStatement * expr_stmt)
-{
-   HaloRefSearch search;   
-   SgExpression * expr = expr_stmt->get_expression();
-
-   //   SgBinaryOp * bin_op = isSgBinaryOp(expr_stmt->get_expression());
-   //   if (bin_op == NULL) {
-   //      expr = expr_stmt->get_expression();
-   //   }
-
-   search.traverse(expr, preorder);
-   return search.found;
-}
-
-HaloRefSearch::HaloRefSearch()
-{
-   this->found = false;
-}
-
-void HaloRefSearch::visit(SgNode * node)
-{
-   switch (node->variantT()) {
-   case V_SgVarRefExp:
-      SgVarRefExp * var = isSgVarRefExp(node);
-      if (var->get_symbol()->getAttribute("halo_attr") != NULL) {
-         found = true;
-      }
-      break;
-   }
-}
-
-
-static int split_string(const char * arg, char *** list)
-{
-   int i = 0;
-   int count = 0;
-
-   char * str = strdup(arg);
-   char * chrp = str;
-
-   if (*chrp != '\0') count = 1;
-
-   // count the number of strings in the list
-   while (*chrp != '\0') {
-      if (*chrp++ == ' ') count += 1;
-   } 
-
-   if (count == 0) return count;
-
-   *list = new char * [count];
-         
-   chrp = str;
-   (*list)[i++] = str;
-
-   // replace ' ' by '\0' to terminate strings in place
-   while (*chrp != '\0') {
-      if (*chrp == ' ') {
-         *chrp = '\0';
-         (*list)[i++] = chrp + 1;
-      } 
-      chrp += 1;
-   } 
-
-   return count;
-}
-
-
-FortranAnalysis::FortranAnalysis(int argc, char * argv[], SgGlobal * scope)
-{
-   this->src_global_scope = scope;
-   this->last_halo_symbol = NULL;
-   this->first_array_dummy_ref = NULL;
-
-   this->num_arrays = 0;
[…]

### 2011-10-27 papers/local-orientation/codegen/FortranAnalysis.hpp

@@ -1,59 +1,0 @@
-#include <rose.h>
-
-using namespace SageBuilder;
-using namespace SageInterface;
-
-class FortranAnalysis : public AstSimpleProcessing
-{
-
-public:
-   
-   FortranAnalysis(int argc, char * argv[], SgGlobal * scope);
-   virtual ~FortranAnalysis();
-
-   virtual void visit(SgNode * node);
-   virtual void atTraversalEnd();
-
-   void visit(SgFunctionDeclaration * func_decl);
-   void visit(SgFunctionDefinition  * func_def);
-   void visit(SgVariableDeclaration * var_decl);
-   void visit(SgFunctionCallExp     * func_call_exp);
-   void visit(SgExprStatement       * expr_stmt);
-
-   // helper functions
-   //
-
-   bool matchRegionAssignment(SgExprStatement * expr_stmt);
-
-   bool isFunctionArg(SgInitializedName * name);
-   bool isRegionSelector(SgInitializedName * var);
-
-   bool isElementalArrayType(SgInitializedName * iname);
-   bool isElementalArrayType(SgFunctionDeclaration * func_decl);
-
-   bool hasArrayDescriptor(SgInitializedName * iname);
-
-   static bool isDummyVariable(SgSymbol * sym);
-   static bool hasLocalVariable(SgVariableDeclaration * var_decl, SgFunctionDefinition * func_def);
-
-   static const char * getAttributeValue(SgSymbol * sym, std::string attr_name);
-
-   // accessor function
-   //
-
-   SgSymbol * lastHaloSymbol()    {return last_halo_symbol;}
-
-protected:
-
-   SgGlobal * src_global_scope;   
-   SgBasicBlock * src_block;
-   SgFunctionDeclaration * src_func_decl;
-   SgSymbol * last_halo_symbol;
-   SgSymbol * first_array_dummy_ref;
-
-   int num_arrays;
-   char ** array_list;
-
-   int num_descs;
-   char ** desc_list;
-};

### 2011-10-27 papers/local-orientation/codegen/FortranTraversal.cpp

@@ -1,114 +1,0 @@
-#include <rose.h>
-#include "FortranTraversal.hpp"
-
-bool FortranTraversal::isErrorParam(SgInitializedName * name)
-{
-   SgType * type = name->get_typeptr();  // TODO - check type
-   if (name->get_name() == "ierror") return true;
-   return false;
-}
-
-bool FortranTraversal::isErrorReturn(SgVariableDeclaration * var_decl)
-{
-   SgTypeModifier & typemod = var_decl->get_declarationModifier().get_typeModifier();
-
-   SgInitializedNamePtrList & vars = var_decl->get_variables();
-   SgInitializedNamePtrList::iterator it = vars.begin();
-   SgType * type = (*it)->get_typeptr();
-
-   if (isSgTypeInt(type) && typemod.isOptional() && typemod.isIntent_out()) {
-      if ((*it)->get_name() == "ierror") return true;
-      return false;
-   }
-
-   return false;
-}
-
-bool FortranTraversal::isChoiceBuffer(SgVariableDeclaration * var_decl)
-{
-   SgTypeModifier & typemod = var_decl->get_declarationModifier().get_typeModifier();
-
-   SgInitializedNamePtrList & vars = var_decl->get_variables();
-   SgInitializedNamePtrList::iterator it = vars.begin();
-   SgType * type = (*it)->get_typeptr();
-
-   if (isSgArrayType(type)) {
-      SgArrayType * array_type = isSgArrayType(type);
-      type = array_type->get_base_type();
-      if (isSgNamedType(type) != NULL) {
-         if (isSgNamedType(type)->get_name() == "MPI_Choice") return true;
-      }
-      return false;
-   }
-   return false;
-}
-
-bool FortranTraversal::isArrayOfRequests(SgVariableDeclaration * var_decl)
-{
-   SgTypeModifier & typemod = var_decl->get_declarationModifier().get_typeModifier();
-
-   SgInitializedNamePtrList & vars = var_decl->get_variables();
-   SgInitializedNamePtrList::iterator it = vars.begin();
-   SgType * type = (*it)->get_typeptr();
-
-   if (isSgArrayType(type)) {
-      SgArrayType * array_type = isSgArrayType(type);
-      type = array_type->get_base_type();
-      if (isSgNamedType(type) && (*it)->get_name() == "array_of_requests") {
[…]

### 2011-10-27 papers/local-orientation/codegen/FortranTraversal.hpp

@@ -1,34 +1,0 @@
-#ifndef FORTRAN_TRAVERSAL_HPP_
-#define FORTRAN_TRAVERSAL_HPP_
-
-#include <rose.h>
-#include <stdio.h>
-
-using namespace SageBuilder;
-using namespace SageInterface;
-
-class FortranTraversal : public AstSimpleProcessing
-{
-
-public:
-   
-   FortranTraversal(const char * filename);
-   virtual void visit(SgNode * node) = 0;
-   virtual void atTraversalEnd();
-
-   static bool isErrorParam(SgInitializedName * name);
-   static bool isErrorReturn(SgVariableDeclaration * var_decl);
-   static bool isChoiceBuffer(SgVariableDeclaration * var_decl);
-   static bool isArrayOfRequests(SgVariableDeclaration * var_decl);
-   static bool isArrayOfStatuses(SgVariableDeclaration * var_decl);
-
-   void unparseOptionalModifier(SgTypeModifier & typemod);
-   void unparseValueModifier(SgTypeModifier & typemod);
-   void unparseIntentModifier(SgTypeModifier & typemod);
-
-protected:
-
-   FILE * fp;  // output file
-};
-
-#endif // FORTRAN_TRAVERSAL_HPP_

### 2011-10-27 papers/local-orientation/codegen/build_elemental_cl.cpp

@@ -1,41 +1,0 @@
-#include "FortranAnalysis.hpp"
-#include "ElementalTraversal.hpp"
-#include <stdio.h>
-
-int main(int argc, char ** argv)
-{
-   FortranAnalysis * analysis;
-   ElementalTraversal * traversal;
-
-   if (argc < 2) {
-      printf("usage: build_elemental_cl -c -rose:skip_syntax_check elemental_add.f90 ");
-      printf("-ofp:<array_list>\n");
-   }      
-
-   SgProject* project = frontend(argc, argv);
-   ROSE_ASSERT(project != NULL);
-
-   SgSourceFile * src_file = isSgSourceFile((*project)[0]);
-   ROSE_ASSERT(src_file);
-	
-   analysis = new FortranAnalysis(argc, argv, src_file->get_globalScope());
-   traversal = new ElementalTraversal(argc, argv, "template.c");
-
-   //src_file->set_outputFormat(SgFile::e_free_form_output_format);
-   //src_file->set_outputLanguage(SgFile::e_C_output_language);
-
-   SgDeclarationStatementPtrList & decls = src_file->get_globalScope()->get_declarations();
-   SgDeclarationStatementPtrList::iterator it_decls;
-   for (it_decls = decls.begin(); it_decls != decls.end(); it_decls++) {
-
-      if (isSgFunctionDeclaration(*it_decls) != NULL) {
-         analysis->traverse(*it_decls, preorder);
-         traversal->traverse(*it_decls, preorder);
-      }
-
-   }
-
-   delete analysis;
-   delete traversal;
-   return 0;
-}

### 2011-10-27 papers/local-orientation/codegen/template.c

@@ -1,24 +1,0 @@
-/*
- * This kernel is automatically generated by Open Fortran Project tools.
- */
-
-static inline int get_global_idx_1d()
-{
-   return get_global_id(0);
-}
-
-static inline int get_global_idx_2d()
-{
-   return get_global_id(0) + get_global_size(0)*get_global_id(1);
-}
-
-__kernel void elemental_add (
-    __global float4 * A,
-    __global float4 * B,
-    __global float4 * C )
-{
-
-   int idx = get_global_idx_2d();
-
-   C[idx] = A[idx] + B[idx];
-}

### 2011-10-27 papers/LOPe/program-restructuring.tex
Moved local-orientation directory to LOPe directory

--- 2011-04-09 opencl/paper-local/programming-model.tex
@@
 \end{itemize}
-
-Programming Models
-------------------
-
-Fortran currently supports:
-
-1.
-Array syntax: e.g., C = A + B, where A, B, C are arrays.
-Note that this is implicitly a loop structure, but that no loop indices need be provided.
-Also the programmer need not specify where in memory these arrays reside.
-Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardward threads).
-This is a simple example and it is a research question as to what compiler directives would be useful for memory placement and other directions to the compiler for efficient code generation.
-
-2.
-Pure procedures:  Fortran has syntax for specifying procedures that have no side effects during execution.
-Specifying code that is side-effect free code is important information to provide to the compiler so that it can generate efficient multi-threaded code.
-
-3.
-Pure elemental procedures: Fortran has syntax for specifying procedures that take only scalar arguments, but may be applied across array elements.
-Elemental procedures are ideal for writing code to be executed within a hardware thread.
-They resemble OpenCL kernels, but are simpler because they leave all indexing up to the compiler.
-
-We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.
-This additional syntax is provided by functions that return a copy of a small region of memory surrounding an array element (as seen within an elemental procedure) and with functions for thread synchronization.
-This additional syntax will allow pure procedures to perform stencil and other convolution-like operations on a copy of memory, synchronize, then store the computed results back to the array element associated with the given thread.
-
-In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.
-This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.
-Fortran also has excellent facilities for interoperability with C so that programming in a mixed language invironment is easily accomplished, including interoperability with native Fortran arrays.

### 2011-10-27 papers/LOPe/code/C/elemental_add.c

@@ -1,0 +1,45 @@
+// Make extended syntax go away
+//
+#define ELEMENTAL
+#define PURE
+
+// Array size
+//
+#define N 100
+
+//
+// The function adds two input array elements and returns the output
+// as an output parameter.
+//
+
+void PURE ELEMENTAL elemental_add(const float * a, const float * b, float * c)
+{
+   *c = *a + *b;
+}
+
+int main(int argc, char* argv[])
+{
+   float A[N], B[N], C[N];
+
+   // initialize arrays...
+
+   // call elemental function
+
+   elemental_add(A, B, C);
+}
+
+
+int main_compiler_creates(int argc, char* argv[])
+{
+   int i;
+   float A[N], B[N], C[N];
+
+   // initialize arrays...
+
+   // inline elemental function
+   for (i = 0; i < N; i++) {
+      C[i] = A[i] + B[N];
+   }
+}
+
+

### 2011-10-27 papers/LOPe/code/LIF_update_state.f90

@@ -1,0 +1,111 @@
+module types_mod
+   type LIF_params
+      real :: tauE, tauI, tauIB, tauVth
+   end type LIF_params
+   type uint4
+      integer :: val1
+   end type uint4
+end module types_mod
+
+pure subroutine &
+LIF_update_state(time, dt, &
+                 params, rnd, &
+                 V, Vth, &
+                 G_E, G_I, G_IB, &
+                 phiExc, phiInh, phiInhB, &
+                 R, activity)
+   !$OFP extended elemental :: LIF_update_state
+   use types_mod
+   implicit none
+   real, intent(in) :: time, dt
+   type(LIF_params), intent(in) :: params
+   type(uint4), intent(inout) :: rnd
+   real, intent(inout) :: V, V_th
+   real, intent(inout) :: G_E, G_I, G_IB
+   real, intent(inout) :: phiExc, phiInh, phiInhB
+   real, intent(inout) :: R, activity
+
+   real :: exp_tauE, exp_tauI, exp_tauIB, exp_tauVth, exp_tauRate
+   real :: dt_sec, GMAX
+
+   ! temporary arrays
+   real :: tauInf, VmemInf
+
+   ! local param variables
+   real :: tau, Vrest, VthRest, Vexc, Vinh, VinhB, deltaVth
+
+   exp_tauE    = exp(-dt/params%tauE)
+   exp_tauI    = exp(-dt/params%tauI)
+   exp_tauIB   = exp(-dt/params%tauIB)
+   exp_tauVth  = exp(-dt/params%tauVth)
+
+   dt_sec = .001 * dt
+   exp_tauRate = exp(-dt/params%tauRate)
+
+   GMAX = 10.0
+
+   !
+   ! start of LIF2_update_exact_linear
+   !
+
+   ! define local param variables
+   !
+   tau   = params%tau
+   Vexc  = params%Vexc
+   Vinh  = params%Vinh
+   VinhB = params%VinhB
+   Vrest = params%Vrest
+
+   VthRest  = params%VthRest
+   deltaVth = params%deltaVth
+
+   ! add noise
+   !
+
+   G_E  = phiExc  + G_E *exp_tauE
+   G_I  = phiInh  + G_I *exp_tauI
+   G_IB = phiInhB + G_IB*exp_tauIB
+   
+   !G_E  = (G_E  > GMAX) ? GMAX : G_E;
[…]

### 2011-10-27 papers/LOPe/code/convolve/convolve_elemental.f90

@@ -1,0 +1,21 @@
+module Filter
+
+   integer, parameter :: N
+
+   type :: FilterPatch
+      real :: p(-N:N,-N:N)
+   end type FilterPatch
+
+contains
+
+   pure elemental subroutine convolve(SmoothedImage, Image, Filter)
+      real, intent(out) :: SmoothedImage
+      real, intent(in), EXTEND :: Image(-N:N,-N:N)
+      type(FilterPatch), intent(in) :: Filter
+
+      SmoothedImage = sum(Filter%p * Image)
+
+   end subroutine convolve
+
+
+end module Filter

### 2011-10-27 papers/LOPe/code/convolve/convolve_forall.f90

@@ -1,0 +1,123 @@
+module convolve_forall
+
+   integer, parameter :: N = 3  ! number of rows    in image
+   integer, parameter :: M = 3  ! number of columns in image
+   integer, parameter :: H = 1   ! size of halo
+
+   ! Number of rows in processor topology
+   ! This should be a command line argument somehow
+   integer, parameter :: PROC_ROWS = 4
+
+!$OFP real, allocatable, HALO(:,:) :: Image(:,:) [PROC_ROWS,*]
+      real, allocatable            :: Image(:,:)
+      real, allocatable            :: Filter(:,:)
+
+contains
+
+
+!===========================================================================
+! Description: Initializes the allocatable arrays.
+!
+! Note about semantics of LOPe: Would like to require all array variables
+! to be used in a concurrent procedure to be allocatable.  This allows
+! allocation for OpenCL buffers to be done at the same time.
+!
+! Note that the image is a coarray and the halo is allocated as part of the
+! array.  The halo is associated with the array but is not necessarily
+! made contiguous with the array proper.  The decision of how to lay out the
+! halo should be up to the compiler.  For example, all of the halo elements
+! may be placed in contiguous memory separate from the array for more
+! efficient memory copies of the halo.  The halo is assumed to be involved
+! in frequent communication between coarray images and between the OpenCL
+! host and device.
+!
+!===========================================================================
+
+   subroutine initialize
+!$OFP allocate( Image(N,M)[PROC_ROWS,*] HALO(H:H,H:H) )
+      allocate( Image(1-H:N+H,1-H:M+H) )
+      allocate( Filter(-H:H,-H:H) )
+
+      ! initialize filter and image arrays
+      !
+      Image  = 1.0
+      Filter = 2*1.0 / ((1+2*H)*(1+2*H))
+
+   end subroutine initialize
+
+   subroutine output_image
+      integer :: row
+      do row = 1-H, N+H
+         print *, Image(row,:)
+      end do
+   end subroutine
+
+
+!===========================================================================
+!
+! Kernel function that performs the convolution for the array element
+! specified by indices.
+!
+!===========================================================================
+
+!$OFP pure CONCURRENT subroutine convolve(Image, Filter, indices)
[…]

### 2011-10-27 papers/LOPe/code/convolve/convolve_lope.f90

@@ -1,0 +1,93 @@
+module Convolve_LOPe
+
+   integer, parameter :: N = 64  ! number of rows    in image
+   integer, parameter :: M = 64  ! number of columns in image
+   integer, parameter :: H = 1   ! size of halo
+
+   ! Number of rows in processor topology
+   ! This should be a command line argument somehow
+   integer, parameter :: PROC_ROWS = 4
+
+   real, allocatable, HALO(:,:) :: Image(:,:) [*,*]    ! [*,*] is new syntax
+   real, allocatable            :: Filter(:,:)
+
+contains
+
+
+!===========================================================================
+! Description: Initializes the allocatable arrays.
+!
+! Note about semantics of LOPe: Would like to require all array variables
+! to be used in a concurrent procedure to be allocatable.  This allows
+! allocation for OpenCL buffers to be done at the same time.
+!
+! Note that the image is a coarray and the halo is allocated as part of the
+! array.  The halo is associated with the array but is not necessarily
+! made contiguous with the array proper.  The decision of how to lay out the
+! halo should be up to the compiler.  For example, all of the halo elements
+! may be placed in contiguous memory separate from the array for more
+! efficient memory copies of the halo.  The halo is assumed to be involved
+! in frequent communication between coarray images and between the OpenCL
+! host and device.
+!
+!===========================================================================
+
+   subroutine initialize
+      allocate( Image(N,M)[*,*] HALO([H,H],[H,H]) )
+      allocate( Filter(-H:H,-H:H) )
+
+      ! initialize filter and image arrays
+      !
+      Image  = 1.0
+      Filter = 1.0 / ((1+2*H)*(1+2*H))
+
+   end subroutine initialize
+
+   subroutine output_image
+      integer :: row
+      do row = 1-H, N+H
+         print *, Image(row,:)
+      end do
+   end subroutine
+
+
+!===========================================================================
+!
+! Kernel function that performs the convolution for the array element
+! specified by indices.
+!
+!===========================================================================
+
+pure CONCURRENT subroutine convolve(Image, Filter)
+
+   real, intent(in out), HALO(:,:) :: Image(0:,0:)
+   real, intent(in)                :: Filter(-H:H,-H:H)
+
+   Image(0,0) = sum(Filter * Image(-H:+H,-H:+H))
+
+end subroutine convolve
+
+end module Convolve_LOPe
+
+
+program convolve_image
[…]

### 2011-10-27 papers/LOPe/code/convolve/filter_type.f90

@@ -1,0 +1,21 @@
+module FilterType
+   use FilterParams
+
+   type :: FilterPatch
+      real :: p(-NPAD:NPAD,-NPAD:NPAD)
+   end type FilterPatch
+
+contains
+
+   pure elemental subroutine convolve_elemental(SmoothedImage, Image, Filter)
+      real, intent(out) :: SmoothedImage
+      !$OFP EXTEND :: Image(-N:N,-N:N)
+      real, intent(in) :: Image
+      type(FilterPatch), intent(in) :: Filter
+
+      ! SmoothedImage = sum(Filter%p * Image)
+      SmoothedImage = Filter%p(1,1) * Image
+
+   end subroutine convolve_elemental
+
+end module FilterType

### 2011-10-27 papers/LOPe/code/convolve/npad.h

@@ -1,0 +1,1 @@
+#define PADDING 7

### 2011-10-27 papers/LOPe/code/elemental_add.f90

@@ -1,0 +1,36 @@
+!
+! Eventual form of high-level representation.  This will first be turned
+! into syntax for OFP and ROSE and then, hopefully it can make its way
+! into the Fortran standard.
+!
+! pure elemental CONCURRENt subroutine elemental_add(A, B, C)
+!    TYPE(*), intent(COPYIN)  :: A, B
+!    TYPE(*), intent(COPYOUT) :: C
+!
+!   C = A + B   
+!
+!end subroutine elemental_add
+
+
+!
+! Note that this function is dimensionless.  The function parameters are scalars
+! but the compiler will generate code for arrays of a particular dimension based
+! on the actual parameters at the calling site.  Also note that the parameters
+! are typeless (see TYPE(*) above); the compiler will create an instance of
+! function according to the called type.
+!
+pure elemental subroutine elemental_add(A, B, C, s)
+!   !
+!   ! compiler directives that will eventually be turned into syntax
+!   ! 
+!   !$OFP CONCURRENT :: elemental_add
+!   !$OFP intent(COPYIN)   :: A, B
+!   !$OFP intent(COPYOUT)  :: C
+!   !
+!
+   real, intent(in), value :: A, B, s
+   real, intent(out) :: C
+
+   C = A + s*B   
+
+end subroutine elemental_add

### 2011-10-27 papers/LOPe/code/filter.f90

@@ -1,0 +1,39 @@
+!
+! A 2x2 filter applied over an array 2x2 larger than destination
+!
+!  V = V + sum(w*A)
+!
+subroutine filter_2x2(nPad, V, A, w)
+  use :: ForOpenCL
+  implicit none
+  integer, intent(in) :: nPad
+  real, dimension(:,:) :: V, A  ! V 2x2 bigger
+  real, dimension(2,2) :: w
+
+  real, dimension(:,:), allocatable :: A_big
+
+  ! scatter A to array compatible with V
+  !
+  A_big = scatter(identity, A, 2)
+
+end subroutine filter_2x2
+
+
+!subroutine convolve(nPad, V, A, w)
+   ! nx = get_global_size(0), ny ... is default, otherwise explicit
+!   real, dimension(:,:) :: V, A, w
+!   asynchronous :: A     ! block on event?
+
+!   real, pointer, dimension(1+2*nPad,1+2*nPad) :: lA
+
+!   halo   = [nPad,nPad,nPad,nPad]
+!   filter = [nPad,nPad,nPad,nPad]
+
+   ! A is extended
+   ! iA => halo(A, aHalo)      ! k,kl indexes
+
+!   lA => halo(A, halo, local=filter)
+
+!   V = V + sum(w*lA)         ! this does convolution
+
+!end subroutine convolve

### 2011-10-27 papers/LOPe/code/finite_difference.f90

@@ -1,0 +1,20 @@
+!
+! computes derivative of A using centered difference
+!
+! I believe A_prime(i) = ( A(i+1) - A(i-1) ) / (2*dx)
+!
+! We use a local orientation
+!
+subroutine finite_difference(A, A_prime, dx)
+   use :: ForOpenCL
+   implicit none
+   real, dimension(:) :: A, A_prime
+   real :: dx, A_local(-1:1)
+   real, pointer :: A_prime_local          ! this will go away
+
+   A_local = local_region(A, [1,1,1,1], halo=[1,1,1,1])   ! A needs halo specification
+   A_prime_local => local(A_prime)
+
+   A_prime_local = ( A_local(1) - A_local(-1) ) / (2*dx)
+
+end subroutine finite_difference

### 2011-10-27 papers/LOPe/code/forall.f90

@@ -1,0 +1,16 @@
+program test_forall
+   implicit none
+   integer, parameter :: ng = 2
+   integer, parameter :: is = 1, ie = 10
+   integer, parameter :: km = 100
+
+   real, dimension(is-ng:ie+ng,km) :: delp
+   real, dimension(is:ie,km) :: dm
+
+   integer :: k, i
+
+   forall(k=1:km, i=is:ie)
+      dm(i,k) = delp(i,k-1)
+   end forall
+
+end program

### 2011-10-27 papers/LOPe/code/heat/test_heat2d.f90

@@ -1,0 +1,63 @@
+!
+!-----  Fortran Programming Language with local extensions -----
+!
+! This file is distributed under an open source license by Los Alamos
+! National Security, LCC.  See the file License.txt (located in the
+! top level of the source distribution) for details.
+! 
+!-----
+!
+! Simplistic 2D heat transfer...
+! 
+!
+
+module heat_transfer
+
+   integer, parameter :: NPAD = 2
+   integer, parameter :: WIDTH  = 1024
+   integer, parameter :: HEIGHT = 1024
+   integer, parameter :: NX = WIDTH  + 2*NPAD
+   integer, parameter :: NY = HEIGHT + 2*NPAD
+
+   integer, parameter :: NTIME_STEPS = 1000
+   real,    parameter :: MAX_TEMP    = 100.0
+
+   interface
+      pure elemental function adv_time(T)
+         real, intent(in) :: T
+      end function adv_time
+   end interface
+
+end module heat_transfer
+
+
+program test_heat2d
+   use :: heat_transfer
+   implicit none
+   integer :: i
+
+   real, dimension(0:NX-1,0:NY-1) :: T1, T2
+  
+   real, parameter :: alpha = .001
+   real, parameter :: dx = 1.0 / WIDTH
+   real, parameter :: dy = 1.0 / HEIGHT
+   real, parameter :: dt = .01 * (alpha / 4.0) * ((1.0 / (dx*dx)) + (1.0 / (dy*dy)))
+
+   ! initialize cell temperature
+   !
+   T1 = MAX_TEMP
+   T2 = 0.0
+
+   T1(:,0)    = 0.0
+   T1(:,NY-1) = 0.0
+   T1(0,:)    = 0.0
+   T1(NX-1,:) = 0.0
+
+   ! time loop
+   !
+   do i = 1, NTIME_STEPS/2
+      T2(1:WIDTH,1:HEIGHT) = adv_time(T1(1:WIDTH,1:HEIGHT))
+      T1(1:WIDTH,1:HEIGHT) = adv_time(T2(1:WIDTH,1:HEIGHT))
+   end do
+
+end program test_heat2d

### 2011-10-27 papers/LOPe/code/nh_core.F90

@@ -1,0 +1,285 @@
+  subroutine Riem_3D(ns, bdt, is, ie, js, je, ng, j, km, cp, gama, cappa, p3, dm2,    &
+                     pm2, w, dz2, pt, quick_p, c_core, ktop, iad)
+
+  integer, intent(in):: ns, is, ie, js, je, ng,  km, j
+  integer, intent(in):: iad      ! time step scheme 
+  integer, intent(in):: ktop     ! starting layer for non-hydrostatic dynamics
+                                 ! 1: All non-hydrostatic
+                                 ! 2: top sponge layer is hydrostatic
+  real,    intent(in):: bdt, cp, gama, cappa
+  real,    intent(in), dimension(is:ie,km):: dm2, pm2
+  logical, intent(in):: quick_p       ! fast algorithm for pressure
+  logical, intent(in):: c_core
+  real, intent(in  ) :: pt (is-ng:ie+ng,js-ng:je+ng,km)
+! IN/OUT:
+  real, intent(inout):: dz2(is:ie,km)
+  real, intent(inout)::   w(is-ng:ie+ng,js-ng:je+ng,km)
+  real, intent(out  )::  p3(is-ng:ie+ng,js-ng:je+ng,km+1)
+! --- Local 1D copyies -----
+#ifdef USE_2D
+  real, dimension(km,is:ie):: t2, p2, pt2
+#else
+  real, dimension(km):: c2, p2, pt2
+#endif
+  real, dimension(km):: r_p, r_n, rden, dz, dm, wm, dts, pdt
+  real, dimension(km+1):: m_bot, m_top, r_bot, r_top, time_left, pe1, pbar, wbar
+
+  real, parameter:: dzmx = 0.5*dz_max
+  real    :: dt, rdt, grg, z_frac, t_left
+  real    :: a1, b1, g2, rcp
+  real    :: seq(ns)       ! time stepping sequence
+  integer :: k2(km+1)
+  integer :: i, k, n, ke, kt, k0, k1, k3
+
+  call time_sequence( iad, ns, bdt, seq )
+
+  grg = gama * rdgas  
+  rcp = 1. / cp
+  rdt = 1. / bdt
+
+  if ( quick_p ) then
+       a1 = 0.5               ! a1=1 fully implicit
+       b1 = 1. - a1
+       g2 = -2.*gama
+  endif
+
+#ifdef USE_2D
+  do i=is,ie
+     do k=ktop,km
[…]

### 2011-10-27 papers/LOPe/code/petavision/weight_type.f90

@@ -1,0 +1,21 @@
+module WeightType
+   use WeightParams
+
+   type :: WeightPatch
+      real :: p(-NPAD:NPAD,-NPAD:NPAD)
+   end type WeightPatch
+
+contains
+
+   pure elemental subroutine convolve_elemental(SmoothedImage, Image, Filter)
+      real, intent(out) :: SmoothedImage
+      !$OFP EXTEND :: Image(-N:N,-N:N)
+      real, intent(in) :: Image
+      type(WeightPatch), intent(in) :: Filter
+
+      ! SmoothedImage = sum(Filter%p * Image)
+      SmoothedImage = Filter%p(1,1) * Image
+
+   end subroutine convolve_elemental
+
+end module WeightType

### 2011-10-27 papers/LOPe/code/reduce.f90

@@ -1,0 +1,30 @@
+!
+! I'm not sure about overloaded functions and function names for these.
+! There will potentiall be several versions, depending on scale factor.
+!
+
+! for now this reduction assumes a 2x2 expansion
+
+Module ForOpenCL
+
+interface
+
+   subroutine reduce(f, A_big, A)
+      real, dimension(:,:) :: A_big, A
+      interface
+         pure elemental function f(a, b)
+            real :: a, b, f
+            intent(in) :: a, b
+         end function
+      end interface
+   end subroutine reduce
+
+   subroutine scatter(A, A_big)
+      real, dimension(:,:) :: A, A_big
+   end subroutine scatter
+
+end interface
+
+end module ForOpenCL
+
+

### 2011-10-27 papers/LOPe/codegen/ElementalTraversal.cpp

@@ -1,0 +1,763 @@
+#include <rose.h>
+#include "ElementalTraversal.hpp"
+#include "FortranAnalysis.hpp"
+
+ElementalTraversal::ElementalTraversal(int argc, char * argv[], const char * filename)
+   : FortranTraversal(filename)
+{
+   line = 0;
+   num_dims = 1;
+   num_arrays = 0;
+   array_list = new char * [1];
+   array_list[0] = NULL;
+
+   // look for -ofp:dims argument
+   //
+   for (int i = 0; i < argc; i++) {
+      char * str = argv[i];
+      if (strncmp(str, "-ofp:dims", 9) != 0) continue;
+      num_dims = atoi(&str[9]);
+      printf("found num_dims==%d str==%s\n", num_dims, str);
+      break;
+   }
+
+   open(filename);
+}
+
+void ElementalTraversal::open(const char * filename)
+{
+   this->fp = fopen(filename, "w");
+   if (fp == NULL) {
+      printf("ElementalTraversal:: ERROR in opening file %s\n", filename);
+   }
+   ROSE_ASSERT(fp != NULL);
+
+   fprintf(fp, "/*\n");
+   fprintf(fp, " * This kernel is automatically generated by Open Fortran Project tools.\n");
+   fprintf(fp, " */\n");
+   fprintf(fp, "\n");
+
+   fprintf(fp, "static inline int get_global_idx_1d()\n");
+   fprintf(fp, "{\n");
+   fprintf(fp, "   return get_global_id(0);\n");
+   fprintf(fp, "}\n\n");
+
+   fprintf(fp, "static inline int get_global_idx_2d()\n");
+   fprintf(fp, "{\n");
+   fprintf(fp, "   return get_global_id(0) + get_global_size(0)*get_global_id(1);\n");
+   fprintf(fp, "}\n");
+}
+
+void ElementalTraversal::visit(SgNode * node)
+{
+   switch (node->variantT())
+   {
+      case V_SgFunctionDefinition  : visit( (SgFunctionDefinition  *) node);  break;
+   }
+}
+
+#ifdef OBSOLETE
+void ElementalTraversal::visit(SgFunctionDeclaration * func_decl)
+{
+   char comma[2] = {0,0};
+
+   std::string func_name = func_decl->get_name();
+   std::string filename = func_name + ".f90";
+   filename = filename.substr(4,filename.size()-1);
+   filename[0] = tolower(filename[0]);
+
+   open(filename.c_str());
+
[…]

### 2011-10-27 papers/LOPe/codegen/ElementalTraversal.hpp

@@ -1,0 +1,57 @@
+#ifndef ELEMENTAL_TRAVERSAL_HPP_
+#define ELEMENTAL_TRAVERSAL_HPP_
+
+#include "FortranTraversal.hpp"
+
+class ElementalTraversal : public FortranTraversal
+{
+
+public:
+   
+   ElementalTraversal(int argc, char * argv[], const char * filename);
+
+   virtual void visit(SgNode * node);
+
+   void open(const char * filename);
+
+   //   void visit(SgInterfaceBody * ibody);
+   //   void visit(SgFunctionDeclaration * func_decl);
+   void visit(SgFunctionDefinition  * func_def);
+
+   void unparseStmt(SgStatement * stmt);
+
+   void unparseExpr(SgExpression * expr);
+   void unparseBinaryOp(SgBinaryOp * op);
+   void unparseUnaryOp(SgUnaryOp * op);
+   void unparseFuncCallExpr(SgFunctionCallExp * expr);
+   void unparseVarRefExpr(SgVarRefExp * var);
+
+   void unparseVarDecl(SgVariableDeclaration * var_decl, SgFunctionDefinition * func_def);
+   void unparseVarDef(SgVariableDeclaration * var_decl, bool isLocal=false);
+   void unparseParamDef(SgVariableDeclaration * var_decl);
+   void unparseFunctionCall(SgFunctionDefinition * func_def);
+
+   void unparseFuncParam(SgInitializedName * name, SgFunctionDefinition * func_def);
+   void unparseFuncParamReturn(SgFunctionDeclaration * func_decl, const char * comma);
+   void unparseVarDeclListForScalars(SgVariableDeclaration * var_decl, bool isLocal);
+   void unparseVarDeclListForArrays(SgVariableDeclaration * var_decl, bool isLocal);
+   void unparseIntVarDeclStmt(SgVariableDeclaration * var_decl, bool isLocal=false);
+   void unparseFloatVarDeclStmt(SgVariableDeclaration * var_decl, bool isLocal=false);
+
+   void unparseType(SgType *);
+
+   void unparseArrayDescs(SgFunctionDeclaration * func_decl);
+   void unparseIndexVars(SgFunctionDeclaration * func_decl);
+
+
+protected:
+
+   int line;
+   int num_dims;
+
+   int num_arrays;
+   char ** array_list;
+
+};
+
+#endif // ELEMENTAL_TRAVERSAL_HPP_

### 2011-10-27 papers/LOPe/codegen/FortranAnalysis.cpp

@@ -1,0 +1,363 @@
+#include "FortranAnalysis.hpp"
+
+#define DEBUG_PRINT
+
+//
+// Helper class
+//
+class HaloRefSearch : public AstSimpleProcessing
+{
+public:
+   HaloRefSearch();
+   virtual void visit(SgNode * node);
+   static bool findHaloRef(SgExprStatement * expr_stmt);
+   bool found;
+};
+
+bool HaloRefSearch::findHaloRef(SgExprStatement * expr_stmt)
+{
+   HaloRefSearch search;   
+   SgExpression * expr = expr_stmt->get_expression();
+
+   //   SgBinaryOp * bin_op = isSgBinaryOp(expr_stmt->get_expression());
+   //   if (bin_op == NULL) {
+   //      expr = expr_stmt->get_expression();
+   //   }
+
+   search.traverse(expr, preorder);
+   return search.found;
+}
+
+HaloRefSearch::HaloRefSearch()
+{
+   this->found = false;
+}
+
+void HaloRefSearch::visit(SgNode * node)
+{
+   switch (node->variantT()) {
+   case V_SgVarRefExp:
+      SgVarRefExp * var = isSgVarRefExp(node);
+      if (var->get_symbol()->getAttribute("halo_attr") != NULL) {
+         found = true;
+      }
+      break;
+   }
+}
+
+
+static int split_string(const char * arg, char *** list)
+{
+   int i = 0;
+   int count = 0;
+
+   char * str = strdup(arg);
+   char * chrp = str;
+
+   if (*chrp != '\0') count = 1;
+
+   // count the number of strings in the list
+   while (*chrp != '\0') {
+      if (*chrp++ == ' ') count += 1;
+   } 
+
+   if (count == 0) return count;
+
+   *list = new char * [count];
+         
+   chrp = str;
+   (*list)[i++] = str;
+
+   // replace ' ' by '\0' to terminate strings in place
+   while (*chrp != '\0') {
+      if (*chrp == ' ') {
+         *chrp = '\0';
+         (*list)[i++] = chrp + 1;
+      } 
+      chrp += 1;
+   } 
+
+   return count;
+}
+
+
+FortranAnalysis::FortranAnalysis(int argc, char * argv[], SgGlobal * scope)
+{
+   this->src_global_scope = scope;
+   this->last_halo_symbol = NULL;
+   this->first_array_dummy_ref = NULL;
+
+   this->num_arrays = 0;
[…]

### 2011-10-27 papers/LOPe/codegen/FortranAnalysis.hpp

@@ -1,0 +1,59 @@
+#include <rose.h>
+
+using namespace SageBuilder;
+using namespace SageInterface;
+
+class FortranAnalysis : public AstSimpleProcessing
+{
+
+public:
+   
+   FortranAnalysis(int argc, char * argv[], SgGlobal * scope);
+   virtual ~FortranAnalysis();
+
+   virtual void visit(SgNode * node);
+   virtual void atTraversalEnd();
+
+   void visit(SgFunctionDeclaration * func_decl);
+   void visit(SgFunctionDefinition  * func_def);
+   void visit(SgVariableDeclaration * var_decl);
+   void visit(SgFunctionCallExp     * func_call_exp);
+   void visit(SgExprStatement       * expr_stmt);
+
+   // helper functions
+   //
+
+   bool matchRegionAssignment(SgExprStatement * expr_stmt);
+
+   bool isFunctionArg(SgInitializedName * name);
+   bool isRegionSelector(SgInitializedName * var);
+
+   bool isElementalArrayType(SgInitializedName * iname);
+   bool isElementalArrayType(SgFunctionDeclaration * func_decl);
+
+   bool hasArrayDescriptor(SgInitializedName * iname);
+
+   static bool isDummyVariable(SgSymbol * sym);
+   static bool hasLocalVariable(SgVariableDeclaration * var_decl, SgFunctionDefinition * func_def);
+
+   static const char * getAttributeValue(SgSymbol * sym, std::string attr_name);
+
+   // accessor function
+   //
+
+   SgSymbol * lastHaloSymbol()    {return last_halo_symbol;}
+
+protected:
+
+   SgGlobal * src_global_scope;   
+   SgBasicBlock * src_block;
+   SgFunctionDeclaration * src_func_decl;
+   SgSymbol * last_halo_symbol;
+   SgSymbol * first_array_dummy_ref;
+
+   int num_arrays;
+   char ** array_list;
+
+   int num_descs;
+   char ** desc_list;
+};

### 2011-10-27 papers/LOPe/codegen/FortranTraversal.cpp

@@ -1,0 +1,114 @@
+#include <rose.h>
+#include "FortranTraversal.hpp"
+
+bool FortranTraversal::isErrorParam(SgInitializedName * name)
+{
+   SgType * type = name->get_typeptr();  // TODO - check type
+   if (name->get_name() == "ierror") return true;
+   return false;
+}
+
+bool FortranTraversal::isErrorReturn(SgVariableDeclaration * var_decl)
+{
+   SgTypeModifier & typemod = var_decl->get_declarationModifier().get_typeModifier();
+
+   SgInitializedNamePtrList & vars = var_decl->get_variables();
+   SgInitializedNamePtrList::iterator it = vars.begin();
+   SgType * type = (*it)->get_typeptr();
+
+   if (isSgTypeInt(type) && typemod.isOptional() && typemod.isIntent_out()) {
+      if ((*it)->get_name() == "ierror") return true;
+      return false;
+   }
+
+   return false;
+}
+
+bool FortranTraversal::isChoiceBuffer(SgVariableDeclaration * var_decl)
+{
+   SgTypeModifier & typemod = var_decl->get_declarationModifier().get_typeModifier();
+
+   SgInitializedNamePtrList & vars = var_decl->get_variables();
+   SgInitializedNamePtrList::iterator it = vars.begin();
+   SgType * type = (*it)->get_typeptr();
+
+   if (isSgArrayType(type)) {
+      SgArrayType * array_type = isSgArrayType(type);
+      type = array_type->get_base_type();
+      if (isSgNamedType(type) != NULL) {
+         if (isSgNamedType(type)->get_name() == "MPI_Choice") return true;
+      }
+      return false;
+   }
+   return false;
+}
+
+bool FortranTraversal::isArrayOfRequests(SgVariableDeclaration * var_decl)
+{
+   SgTypeModifier & typemod = var_decl->get_declarationModifier().get_typeModifier();
+
+   SgInitializedNamePtrList & vars = var_decl->get_variables();
+   SgInitializedNamePtrList::iterator it = vars.begin();
+   SgType * type = (*it)->get_typeptr();
+
+   if (isSgArrayType(type)) {
+      SgArrayType * array_type = isSgArrayType(type);
+      type = array_type->get_base_type();
+      if (isSgNamedType(type) && (*it)->get_name() == "array_of_requests") {
[…]

### 2011-10-27 papers/LOPe/codegen/FortranTraversal.hpp

@@ -1,0 +1,34 @@
+#ifndef FORTRAN_TRAVERSAL_HPP_
+#define FORTRAN_TRAVERSAL_HPP_
+
+#include <rose.h>
+#include <stdio.h>
+
+using namespace SageBuilder;
+using namespace SageInterface;
+
+class FortranTraversal : public AstSimpleProcessing
+{
+
+public:
+   
+   FortranTraversal(const char * filename);
+   virtual void visit(SgNode * node) = 0;
+   virtual void atTraversalEnd();
+
+   static bool isErrorParam(SgInitializedName * name);
+   static bool isErrorReturn(SgVariableDeclaration * var_decl);
+   static bool isChoiceBuffer(SgVariableDeclaration * var_decl);
+   static bool isArrayOfRequests(SgVariableDeclaration * var_decl);
+   static bool isArrayOfStatuses(SgVariableDeclaration * var_decl);
+
+   void unparseOptionalModifier(SgTypeModifier & typemod);
+   void unparseValueModifier(SgTypeModifier & typemod);
+   void unparseIntentModifier(SgTypeModifier & typemod);
+
+protected:
+
+   FILE * fp;  // output file
+};
+
+#endif // FORTRAN_TRAVERSAL_HPP_

### 2011-10-27 papers/LOPe/codegen/build_elemental_cl.cpp

@@ -1,0 +1,41 @@
+#include "FortranAnalysis.hpp"
+#include "ElementalTraversal.hpp"
+#include <stdio.h>
+
+int main(int argc, char ** argv)
+{
+   FortranAnalysis * analysis;
+   ElementalTraversal * traversal;
+
+   if (argc < 2) {
+      printf("usage: build_elemental_cl -c -rose:skip_syntax_check elemental_add.f90 ");
+      printf("-ofp:<array_list>\n");
+   }      
+
+   SgProject* project = frontend(argc, argv);
+   ROSE_ASSERT(project != NULL);
+
+   SgSourceFile * src_file = isSgSourceFile((*project)[0]);
+   ROSE_ASSERT(src_file);
+	
+   analysis = new FortranAnalysis(argc, argv, src_file->get_globalScope());
+   traversal = new ElementalTraversal(argc, argv, "template.c");
+
+   //src_file->set_outputFormat(SgFile::e_free_form_output_format);
+   //src_file->set_outputLanguage(SgFile::e_C_output_language);
+
+   SgDeclarationStatementPtrList & decls = src_file->get_globalScope()->get_declarations();
+   SgDeclarationStatementPtrList::iterator it_decls;
+   for (it_decls = decls.begin(); it_decls != decls.end(); it_decls++) {
+
+      if (isSgFunctionDeclaration(*it_decls) != NULL) {
+         analysis->traverse(*it_decls, preorder);
+         traversal->traverse(*it_decls, preorder);
+      }
+
+   }
+
+   delete analysis;
+   delete traversal;
+   return 0;
+}

### 2011-10-27 papers/LOPe/codegen/template.c

@@ -1,0 +1,24 @@
+/*
+ * This kernel is automatically generated by Open Fortran Project tools.
+ */
+
+static inline int get_global_idx_1d()
+{
+   return get_global_id(0);
+}
+
+static inline int get_global_idx_2d()
+{
+   return get_global_id(0) + get_global_size(0)*get_global_id(1);
+}
+
+__kernel void elemental_add (
+    __global float4 * A,
+    __global float4 * B,
+    __global float4 * C )
+{
+
+   int idx = get_global_idx_2d();
+
+   C[idx] = A[idx] + B[idx];
+}

### 2011-10-27 papers/LOPe/intro.tex
Initial version.

It has been hard historically to create parallel programs.  The responsibility
(and difficulty) of creating \emph{correct} parallel programs can be viewed as
being spread between the programmer and the compiler.
Ideally we would like to have parallel languages that make it easy for the
programmer to express correct parallel programs (and conversely it should be difficult to express
incorrect parallel programs).  Unfortunately, many current languages and
standards place all of the responsibility on the user.  The best example of this
are programs written using MPI (Message Passing Interface), where the programmer
expresses all parallelism in terms of calls to library routines and the serial C or Fortran
compiler knows nothing about the parallel semantics of the combined language
plus MPI library.

For the purpose of discussion, we define a degree of difficulty term called the
joint Parallel Complexity Product (PCP),

    PCP = code\_complexity X compiler\_complexity

and suggest that PCP is roughly a constant over time, say the level of
complexity of writing an MPI program today, MPI\_PCP.

Of course one would hope that over time code complexity goes down as better
languages allow compilers to take on a greater share of the parallel complexity
burden.  This is somewhat true of UPC and Coarray Fortran.  With the PGAS
extensions, C and Fortran compilers are now aware of parallelism and now
generate message passing code that had been handled by the MPI library.  In some
instances the compiler is able to perform optimizations that it is not able to
do with a library based scheme like MPI \cite{CrayPGASPaper}

%%% Would like to say something here about HPC as a language based solution to reducing code complexity

However, in large part these languages are largely syntatic sugar for message
passing and do not provide a substatial decrease in code complexity
\cite{CAF_MPI_PGAS_COMPARISON}.  Fortunately, skilled programmers in the HPC
community have become accustomed to the level of complexity in an MPI program
and are welcome to considering the simplifications of HPC and Coarray Fortran.

The problem for programmers is that hardware is changing in ways that increase
the level of on-chip parallelism.  For ultimate performance, programmers must now
account for huge new levels of on-node parallelism at the same time they account
for off-node parallelism.  Thus PCP has suddenly increased with PCP = PCP\_Multi\_Core >> PCP\_MPI.
Since languages have not evolved that allow the compiler to take up this increase, unfortunately the
complexity for a programmer has dramatically increased.

A reasonable solution given todays language limitations is to use MPI (or a PGAS
language) \emph{plus} OpenMP for on-chip parallelism.  This is the solution proposed
by Cray and PGI \cite{BOF_SC10} (currently Chapel does not target GPUs\cite{Brad?}).
Other choices for expressing on-chip parallelism are OpenCL\cite{OPENCL} and NVIDIA's CUDA.

In this paper we examine a language-based paradigm that allows the compiler to
take on a larger portion of the PCP burden.  Anyone programming in OpenCL or
CUDA is aware that explicit loop structures over array elements in a serial
program are removed and replaced by a kernel program that is run over all of
the

### 2011-10-27 papers/LOPe/motivation.tex
Initial version.

\section{Motivation}

\subsection{Coarray Fortran Comparison}

Below is the motivational example from the original Numrich and Reic paper.
{NOTE: This should sortof/kindof show how it would work between nodes?  Next
section shows OpenMP example for parallelism at the thread level.}

{\small
\begin{verbatim}

subroutine laplace (nrow,ncol,u)
   integer, intent(in) real, intent(inout) real integer
   ! that refers to the current image
   left = me-1
   if (me == 1) left = ncol
   right = me + 1
   if (me == ncol) right = 1
   call sync_all( (/left,right/) ) ! Wait if left and right
   ! have not already reached here
   new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
   call sync_all( (/left,right/) )
   u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow) end subroutine laplace

\end{verbatim}
}

Below is the equivalent implementation in LOPe.

{\small
\begin{verbatim}

pure CONCURRENT subroutine laplace(U)
   real, intent(inout), HALO(-1:*:1,-1:*:1) :: U(0:,0:)

   U(0:0) = U(-1,0) + U(+1,0) + U(0,-1) + U(0,+1) - 3*U(0,0)

end subroutine laplace

\end{verbatim}
}

NOTE: Following not a criticism of CAF as CAF is a general purpose parallel
programming language and LOPe only pertains to the halo parallel pattern.

\subsubsection{Improvements}
\begin{itemize}

\item
LOPe requires that the implementation of the algorithm to be separate from the
call to effect the halo transfer.  Removing boundary condition specification
(e.g., the cyclic boundary conditions implemented in the CAF example) from the
algorithm allows the boundary conditions to be changed and without changing
algorithm code.

\item
LOPe applies the transfer of halo memory across multiple (possibly) levels of
memory with the intrinsic {\tt transfer\_halo} function (not shown here).  Thus
the LOPe algorithm can be run on a machine with many interconnected nodes, each
containing hybrid processor cores.  The coarray example can only be run on
multiple nodes (called images in Fortran) without accelerator cores.

\item
The algorithm implementation is separate from user-specified synchronization,
e.g., {\tt call sync\_all}.  In LOPe, synchronization is subsummed in the
semantics of the CONCURRENT attribute and the {\tt transfer\_halo} function.

\item
The algorithm implementation is separate from any specification as to where the
array memory is located.  The CAF example explicity denotes where memory is
located with the {\tt [left]} and {\tt [right]} syntax where left and right
specifiy a processor topology.

\item
The algorithm implementation is separate from any specification as to where the
algorithm is to be executed.  The CAF example explicity denotes where a statement is to
executed with control flow construct like {\tt if (me == 1)}.

\item
The LOPe implementation is easier to understand and frequently follows the
mathematical algorithm directly.  For example, the CAF example adds 4 neighbors
plus the center value to make the implementation with direct remote coarray
access possible, while the LOPe example is able to implement the same algorithm
with fewer operations by adding 4 neighbors (not including the center array
element) and then only subtracting 3 center values.

\item
The semantics of LOPe makes explicit management of array temporaries (e.g., {\tt
  u} and {\tt new\_u} by programmers unnecessary (though still possible).
Because in LOPe the halo region is a language construct, the compiler is better
able to manage temporary buffers than users on the target hardware platform.

\end{itemize}

\subsubsection{Errors that are constrained by the language}
\begin{itemize}

\item
A programmer is not able to store data to the halo region.  If this were
allowed, one thread could overwrite another threads data at undefined times.
The compiler is able to catch this class of error.

\item
A programmer can't make indexing errors in a concurrent routine by going out of
bounds of the array plus halo memory.  The compiler is able to catch this class of
error at compile time unless the halo shape is assumed (e.g., HALO(:,:)).

\item
A programmer is not able to cause race conditions by forgetting to create and
use temporary arrays properly.  In LOPe it is the comilers responsibility to
store data in temporary memory.

\item
A programmer can't make synchronization errors as synchronization is implicit in
the CONCURRENT attribute.  A thread running a CONCURRENT procedure is provided
with a copy of it's array element (plus halo) that is consistent with the state
of memory at the time of invocation of procedure.  Stores to an individual
thread's array element (by that thread) is never visible to other threads.
It is possible to relax this restraint with a halo synchronization intrisic.
However, LOPE encourages the creation of small functions and lets the compiler
fuse the procedures together and provide the necessary synchronization.

\end{itemize}

Please note that this example is somewhat unfair because in practice CAF is usually
refactored in a locally oriented way by so that communication and
synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.

This is the normal way that large complex MPI and CAF programs are implemented, i.e., using
halo cells.  LOPe proposes to formalize this common pattern into the Fortran
language providing the compiler information in order to spread computation over
more hardware resources and reduce complexity for the programmer.

\subsection{OpenMP Comparision}

The coarray Fortran example shows how LOPe

### 2011-10-27 papers/LOPe/code/convolve/omp.f90
Initial version.

@@ -1,0 +1,57 @@
+PROGRAM VEC_ADD_DO
+   use Timer_mod
+   type(CPUTimer) :: timer
+   real(c_double) :: cpu_time
+
+   integer, parameter :: N = 100000
+   integer, parameter :: CHUNKSIZE = 10000
+
+   INTEGER CHUNK, I
+   REAL A(N), B(N), C(N)
+
+!  Some initializations
+   do I = 1, N
+      A(I) = I * 1.0
+      B(I) = A(I)
+   end do
+   CHUNK = CHUNKSIZE
+        
+   C = 2*A + 3*B
+   print *, C(10)   
+
+   call init(timer)
+   call start(timer)
+
+!   do I = 1, N
+!      C(I) = A(I) + B(I)
+!   enddo
+   C = A + B
+
+   call stop(timer)
+   cpu_time = elapsed_time(timer)
+
+   print *, "   host time    ==   ", real(cpu_time), " msec"
+
+   call init(timer)
+   call start(timer)
+
+!!$OMP PARALLEL SHARED(A,B,C,CHUNK) PRIVATE(I)
+!$OMP PARALLEL
+
+!!$OMP DO SCHEDULE(DYNAMIC,CHUNK)
+!$OMP DO
+   do I = 1, N
+      C(I) = A(I) + B(I)
+   end do
+
+!$OMP END DO
+!!$OMP END DO NOWAIT
+
+!$OMP END PARALLEL
+
+   call stop(timer)
+
+   cpu_time = elapsed_time(timer)
+   print *, "    omp time    ==   ", real(cpu_time), " msec"
+
+end program

### 2011-10-27 papers/LOPe/code/convolve/file_io.f90
Initial version.

@@ -1,0 +1,61 @@
+module file_io
+   use :: iso_c_binding
+
+interface
+   integer(C_INT) function image_size(filename, nx, ny) &
+      bind(c,name="image_size")
+      use :: iso_c_binding
+      character(C_CHAR), intent(in) :: filename(*)
+      integer(C_INT), intent(out) :: nx, ny
+    end function image_size
+
+    integer(C_INT) function print_header(filename) &
+       bind(c,name="print_header")
+       use :: iso_c_binding
+       character(C_CHAR), intent(in) :: filename(*)
+     end function print_header
+
+    integer(C_INT) function read_image(filename, image) &
+       bind(c,name="read_image")
+       use :: iso_c_binding
+       character(C_CHAR), intent(in) :: filename(*)
+       integer(C_CHAR), intent(out) :: image(*)
+    end function read_image
+  
+    subroutine read_image_file(filename, nx, ny, image) &
+       bind(c,name="read_image_file")
+       use :: iso_c_binding
+       character(C_CHAR), intent(in) :: filename(*)
+       real(C_FLOAT), intent(out) :: image(*)
+       integer, value, intent(in) :: nx, ny
+    end subroutine read_image_file
+
+    subroutine write_image_file(filename, nx, ny, image) &
+       bind(c,name="write_image_file")
+       use :: iso_c_binding
+       character(C_CHAR), intent(in) :: filename(*)
+       real(C_FLOAT), intent(in) :: image(*)
+       integer, value, intent(in) :: nx, ny
+    end subroutine write_image_file
+end interface
+
+end module file_io
+
+!program main
+!  use file_io
+!  implicit none
+!
+!  integer(C_INT) :: size, nx, ny
+!  integer(C_CHAR), allocatable :: image(:)
+!
+!  size = image_size("lena-sjooblom.pvp", nx, ny)
+!  print *, " image size is", size, nx, ny
+!
+!  allocate(image(size))
+!
+!  size = read_image("lena-sjooblom.pvp", image)
+!  print *, " read size is ", size
+!
+!  print *, image(1:10)
+!
+!end program

### 2011-10-27 checkers/multipleDefinitions/tests/cases/multipleDefinitionsTest1.f90
multipleDefinitons: add a checker for declarations with multiple definitions

@@ -1,0 +1,8 @@
+program test
+
+  integer :: x, y
+
+
+  print *, x
+  print *, y
+end program

### 2011-10-27 papers/LOPe/local.tex

--- 2011-06-26 papers/GPUScA-2011/foropencl.tex
@@
-%
-% I'm adding an outline here so I can see how everything fits together
-
-%
-% \title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
-%
-
-%
-% \abstract
-%
-
-%
-% \section{Introduction}
-%    \subsection{Approach}
-%    \subsection{Why Fortran?}
-%    \subsection{Comparison to Other Languages}
-%
-
-%
-% \section{Programming model}
-%    \subsection{Fortran Syntax}   
-%       \subsubsection*{Array notation}
-%       \subsubsection*{Elemental functions}
-%       \subsubsection*{Pure procedures}
-%       \subsubsection*{Shift functions}
-%       \subsubsection*{Regions}
-%    \subsection{New Functions}
-%    \subsection{Parallelism}
-%    \subsection{Limitations}
-%
-
-%
-% \section{Shallow Water Model}
-%    \subsection{Equations}
-%
-
-%
-% \section{Source-To-Source Transformations}
-%    \subsection{ForOpenCL}
-%       \subsubsection{array syntax}
-%       \subsubsection{where construct}
-%    \subsection{New functions}
-%       \subsubsection{region}
-%    \subsection{Static Analysis}
-%       \subsubsection{Analysis not required}
-%    \subsection{Simplifying Assumptions}
-%
-
-%
-% \section{Performance}
-%
-
-%
-% \section{Conclusions}
-%
-
 \documentclass{acm_proc_article-sp}
@@
 
-\title{ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming}
+\title{Locally-Oriented Programming: A stunningly simple programming model for GPU programming in Fortran and C++}
 
-\numberofauthors{6}
+\numberofauthors{4}
 
@@
 \alignauthor
+Craig E Rasmussen\\
+       \affaddr{Los Alamos National Laboratory}\\
+%       \affaddr{CCS-7, MS B287}\\
+       \affaddr{Los Alamos, NM 87545}\\
+       \email{crasmussen@lanl.gov}
+% 2nd. author
+\alignauthor
 Matthew J. Sottile\\
        \affaddr{Galois, Inc.}\\
-       \affaddr{421 SW 6th Ave. Suite 300}\\
+%       \affaddr{421 SW 6th Ave. Suite 300}\\
        \affaddr{Portland, OR 97204}\\
-       \email{matt@galois.com}
-% 2nd. author
-\alignauthor
-Craig E Rasmussen\\
-       \affaddr{Los Alamos National Laboratory}\\
-       \affaddr{CCS-7, MS B287}\\
-       \affaddr{Los Alamos, NM 87545}\\
-       \email{crasmussen@lanl.gov}
+       \email{mjsottile@computer.org}
 % 3rd. author
 \alignauthor
-Wayne N. Weseloh\\
-       \affaddr{Los Alamos National Laboratory}\\
-       %\affaddr{XTD-1, MS T082}\\
-       %\affaddr{Los Alamos, NM 87545}\\
-       \email{weseloh@lanl.gov}
-\and  % use '\and' if you need 'another row' of author names
+Dan Nagle\\
+       \email{dannagle@verizon.net}
 % 4th. author
-\alignauthor
-Robert W. Robey\\
-       \affaddr{Los Alamos National Laboratory}\\
-       %\affaddr{XCP-2, MS T086}\\
-       %\affaddr{Los Alamos, NM 87545}\\
-       \email{brobey@lanl.gov}
-% 5th. author
 \alignauthor
@@
        \email{dquinlan@llnl.gov}
-% 6th. author
-\alignauthor
-Jeffrey Overbey\\
-       \affaddr{University of Illinois at Urbanna-Champaign}\\
-       \email{overbey2@illinois.edu}
 }
@@
 Emerging GPU architectures for high performance computing are well suited to a
-  data-parallel programming model.  This paper presents preliminary work
-  examining a programming methodology that provides Fortran programmers with access
-  to these emerging systems.  We use array constructs in Fortran to
-  show how this infrequently exploited, standardized language feature is easily
-  transformed to lower-level accelerator code.  The transformations in ForOpenCL are based
-  on a simple mapping from Fortran to OpenCL.  We demonstrate, using a
-  stencil code solving the shallow-water fluid equations, that the performance
-  of the ForOpenCL compiler-generated transformations is comparable with that of
-  hand-optimized OpenCL code.
+data-parallel programming model.  However, programming for these architectures
+currently requires learning a new language, e.g., OpenCL or CUDA, or employing
+OpenMP compiler directives.  This paper presents preliminary work examining how
+programming with a local orientation can be employed in Fortran (or using a
+special array library in C++) to provide simple access GPU architectures.  A
+locally-oriented programming model is especially useful for stencil-based codes
+or algorithms requiring the application of convolution kernels.  In this
+programming model, a programmer codes the algorithm {\it only} for a single
+array element, but has read-only access to a small sub-array surrounding the
+given array element, so that a stencil can be applied.  We demonstrate how a
+locally-oriented programming model can be adopted using source-to-source program
+transformations in conjuction with a compiler-based infrastructure like ROSE.
 \end{abstract}
@@
 % A category with the (minimum) three required fields
-\category{D.3.3}{Language Constructs and Features}{Concurrent programming structures}
+%\category{D.3.3}{Language Constructs and Features}{Concurrent programming structures}
 
@@
 the data-parallel model attractive.  The increasing complexity of
-todays applications (especially in light of the increasing complexity
+today's applications (especially in light of the increasing complexity
 of the hardware) and the need for portability across architectures
@@
 \section{Acknowledgments}
-This research was supported by the Departement of Energy, Office of Science.
+This work was supported in part by the Department of Energy Office of Science,
+Advanced Scientific Computing Research.
 
@@
 \bibliographystyle{abbrv}
-\bibliography{foropencl}
+\bibliography{local}
 

### 2011-11-02 papers/AMR/code/wave_kern_calc.f90
Initial version.

@@ -1,0 +1,152 @@
+pure function calcDelta(dx)
+   integer, parameter :: l=-1, r=+1
+   real, intent(in) :: dx(l:r)
+   real :: calcDeltaX(l:r)
+
+   calcDeltaX(l) = dx(l) + dx(0)
+   calcDeltaX(0) = dx(0)
+   calcDeltaX(r) = dx(0) + dx(r)
+end function calcDelta
+
+pure function calcVarDist(Var,Dist)
+   integer, parameter :: l=-1, r=+1
+   real, intent(in) :: Var(l:r), Dist(l:r)
+   real :: calcVarDist(l:r)
+
+   calcVarDist(l) = dx(l) + dx(0)
+   calcVarDist(0) = dx(0)
+   calcVarDist(r) = dx(0) + dx(r)
+end function calcDelta
+
+
+
+pure concurrent &
+subroutine calc_one_cycle(H, U, V, deltaT)
+   real(C_DOUBLE), halo(1:*:1,1:*:1), halo_fn(amr_copy), &
+                   intent(in out), dimension(0:,0:) :: H, U, V
+   real(C_DOUBLE), intent(in) :: deltaT
+
+
+!                 const int    ncells,   // 0  Total number of cells.
+!                 const int    levmx,    // 1  Maximum level
+!        __global const real  *H_old,    // 2
+!        __global const real  *U_old,    // 3
+!        __global const real  *V_old,    // 4
+!        __global       real  *H_new,    // 5
+!        __global       real  *U_new,    // 6
+!        __global       real  *V_new,    // 7
+!        __global const int   *nlft,     // 8  Array of left neighbors.
+!        __global const int   *nrht,     // 9  Array of right neighbors.
+!        __global const int   *ntop,     // 10  Array of bottom neighbors.
+!        __global const int   *nbot,     // 11  Array of top neighbors.
+!        __global const int   *level,    // 12  Array of level information.
+!                 const real   deltaT,   // 14  Size of time step.
[…]

### 2011-11-02 papers/LOPe/intro.tex

@@
 CUDA is aware that explicit loop structures over array elements in a serial
-program are removed and replaced by a kernel program that is run over all of
-the
+program are removed and replaced by a kernel program that is run over all of the
+elements of the input arrays.  We propose Locally Orientated Programming
+extensions (LOPe) to the Fortran and C languages that formally adopt this
+programming model.

### 2011-11-02 papers/LOPe/motivation.tex

@@
 \section{Motivation}
+
+The LOPe programming model restricts the programmer --- while implementing a stencil
+algorithm --- to a local view of the array index space.  Within a concurrent function,
+only the local array element plus a small halo region is visible to the programmer.
+This reduces complexity by all boundary conditions and processor topology information
+from the algorithm implementation.  This reduction in complexity reduces programming
+errors.  While developing the convolution example described later in the paper we
+made an indexing error in applying the 2D stencil loops in the standard serial 
+Fortran test implementation.  This error required over 3 hours of programming time
+to repair.  First the error had to be isolated to the function implementing the
+convolution (it was first thought to be in the complicated tiff image output routine
+as the convolution code was ``thought'' to be too simple to wrong.  Then the index
+error has to be understood.  As will be seen, LOPe makes it more difficult to make
+these errors as Fortran array intrics can be used.  In some instance, the restricted
+semantics of LOPe allows the compiler to catch errors (e.g., some errors involving
+race conditions).
 

### 2011-11-15 papers/IJCSE/foropencl.tex
starting point for journal paper due Dec. 12

--- 2011-06-26 papers/GPUScA-2011/foropencl.tex
@@
        \affaddr{Galois, Inc.}\\
-       \affaddr{421 SW 6th Ave. Suite 300}\\
+%       \affaddr{421 SW 6th Ave. Suite 300}\\
        \affaddr{Portland, OR 97204}\\
-       \email{matt@galois.com}
+       \email{mjsottile@computer.org}
 % 2nd. author
@@
        \affaddr{Los Alamos National Laboratory}\\
-       \affaddr{CCS-7, MS B287}\\
+%       \affaddr{CCS-7, MS B287}\\
        \affaddr{Los Alamos, NM 87545}\\

### 2011-11-15 papers/IJCSE/shallow-water.tex
starting point for journal paper due Dec. 12

--- 2011-06-26 papers/GPUScA-2011/shallow-water.tex
@@
 resulting code is simple, concise, and easy to understand.  However it does
-\emph{not} necessarily perform well because of the temporary array variables,
+\emph{not} necessarily perform well when compiled for a traditional sequential
+system because of suboptimal use of temporary array variables,
 especially those produced by the function {\tt region\_cpy}.  This is generally true

### 2011-11-15 papers/IJCSE/source-to-source.tex
starting point for journal paper due Dec. 12

--- 2011-06-27 papers/GPUScA-2011/source-to-source.tex
@@
 
-This section provides an brief overview of the ForOpenCL transformations that take
-Fortran elemental and pure procedures as input and generate OpenCL code.  Recall
-that elemental functions are transformed to inline OpenCL functions and that
-subroutines with the {\tt PURE} and {\tt KERNEL} compiler directive attributes
-are transformed to OpenCL kernels.
+This section provides an brief overview of the ForOpenCL
+transformations that take Fortran elemental and pure procedures as
+input and generate OpenCL code.  Elemental functions are
+transformed to inline OpenCL functions and subroutines with the
+{\tt PURE} and {\tt KERNEL} compiler directive attributes are
+transformed to OpenCL kernels.
 

### 2011-11-16 papers/IJCSE/rose-transform.tex
starting new subsection

\subsection{Kernel transformation}

In addition to the transformation of the code at call sites to copy
data back and forth from the device and invoke the OpenCL kernel, we
must also provide transformations of the Fortran kernel logic to the
equivalent OpenCL C code.  This transformation is fairly
straightforward, and maps a restricted subset of Fortran directly to
its C equivalent.  The transformation is based on a regular traversal
of the ROSE Sage AST that represents the Fortran code.  The visitor
used in the traversal handles five types of Sage AST Nodes: {\tt
  SgAllocateStatment}, {\tt SgFunctionDeclaration},
{\tt SgVariableDeclaration}, {\tt SgFunctionCallExp}, and
{\tt SgExprStatement}.  These encompass the primary activities that are
commonly encountered in computational kernels: allocation, function invocation,
variable and function declaration, and expressions that correspond to
arithmetic operations.

### 2011-11-27 papers/IJCSE/style/IJSISE-1(3-4)-04Tuithungetal.tex
style sheet from journal web site

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass{doublecol-new}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{natbib,stfloats}
\def\newblock{\hskip .11em plus.33em minus.07em}

\theoremstyle{TH}{
\newtheorem{lemma}{Lemma}[section]
\newtheorem{theorem}{Theorem}
\newtheorem{corrolary}{Corrolary}
\newtheorem{conjecture}[lemma]{Conjecture}
\newtheorem{proposition}[lemma]{Proposition}
\newtheorem{claim}[lemma]{Claim}
\newtheorem{stheorem}[lemma]{Wrong Theorem}
}

\theoremstyle{THrm}{
\newtheorem{definition}{Definition}[section]
\newtheorem{question}{Question}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{scheme}{Scheme}
}

\theoremstyle{THhit}{
\newtheorem{case}{Case}[section]
}

\makeatletter
\def\theequation{\arabic{equation}}
\makeatother

%%%%%%%%%%%%%%%%%
\begin{document}%
%%%%%%%%%%%%%%%%%

\thispagestyle{plain}

\setcounter{page}{197}

\LRH{T. Tuithung et~al.}

\RRH{Motion Compensated JPEG2000 based video compression algorithms}

\VOL{1}

\ISSUE{3/4}

\PUBYEAR{2008}

%\BottomCatch

\title{Motion Compensated JPEG2000 based video compression algorithms}

\authorA{T. Tuithung}

\affA{Department of Computer Science and Engineering,\\
 Indian Institute of Technology Kharagpur,\\
 Kharagpur 721302, India\\
Fax: 91-3222-282206/282700\\ E-mail: themri@cse.iitkgp.ernet.in}

\authorB{S.K. Ghosh*}
\affB{School of Information Technology,\\
 Indian Institute of Technology Kharagpur,\\
 Kharagpur 721302, India\\
Fax: 91-3222-282206/282700\\ E-mail: skg@iitkgp.ac.in\\ {*}Corresponding
author}

\authorC{Jayanta Mukherjee}

\affC{Department of Computer Science and Engineering,\\
 Indian Institute of Technology Kharagpur,\\
 Kharagpur 721302, India\\
Fax: 91-3222-282206/282700\\ E-mail: jay@cse.iitkgp.ernet.in}

\begin{abstract}
A new approach of Motion-Compensated JPEG2000 (MCJ2K) video compression is
proposed in this paper. It uses simplified MPEG-2 fundamentals along with
JPEG2000 encoding for compression of the image frames. MCJ2K generates
differential images using motion compensation algorithms and store the motion
vectors in a customised header. The~images are then compressed using JPEG2000
based scheme. MCJ2K have been studied on QCIF and CIF formated videos for
various rate control mechanism. The proposed scheme produced encouraging results
compared to standard MPEG-2 and MJPEG2000 schemes, especially at higher bit
rates.
\end{abstract}

\KEYWORD{video compression; JPEG2000; MJPEG2000; motion compensation; motion
vectors; rate-control.}

\REF{to this paper should be made as follows: Tuithung,~T.,
Ghosh,~S.K. and Mukherjee,~J. (2008) `Motion Compensated JPEG2000
based video compression algorithms', {\it Int. J. Signal and Imaging
Systems Engineering}, Vol.~1, Nos.~3/4, pp.197--212.}

\begin{bio}
T. Tuithung did her BTech in Computer Science and Engineering and MTech in
Information Technology from North Eastern Regional Institute of Science and and
Technology (NERIST), Itanagar, India. Currently she is doing her PhD in Computer
Science and Engineering at Indian Institute of Technology, Kharagpur, India. Her
areas of interest are image processing, video compression, algorithms.\vs{8}

\noindent S.K. Ghosh has done his PhD in the year 2002 from Indian
Institute of Technology (IIT), Kharagpur, India from the Department
of Computer Science and Engineering. He~is~presently working as an
Associate Professor in the School of Information Technology, Indian
Institute of Technology, Kharagpur. Prior to IIT Kharagpur, he~worked
for Indian Space Research Organisation (ISRO), Department of Space,
Government of India, for about eight years in the field of Satellite
Remote Sensing and GIS. His research interest includes image
processing, computer networking, education technology, remote sensing
and~GIS.\vs{8}

\noindent Jayanta Mukherjee received his BTech, MTech, and PhD
Degrees in Electronics and Electrical Communication Engineering from
the Indian Institute of Technology (IIT), Kharagpur in 1985, 1987,
and 1990, respectively. He is presently a Professor in the Department
of Computer Science and Engineering, IIT Kharagpur. He~was a
Humboldt\break
\end{bio}

\maketitle\vfill\pagebreak

\maketitle

\thispagestyle{headings}

\noindent\parbox{42.4pc}{\leftskip 12.75pc\NINE Research Fellow at
the Technical University of Munich in Germany for one year in 2002.
He also has held short term visiting positions at the University of
California, Santa Barbara, University of Southern California, and the
National University of Singapore. His research interests are in image
processing, pattern recognition, computer graphics, multimedia
systems and medical informatics. He~has published over 100~papers in
international journals and conference proceedings in these areas.
He~served in various technical program committees of international
conferences in related areas. He~received the Young Scientist Award
from the Indian National Science Academy in 1992. He~is~a~Senior
Member of the IEEE. He is a fellow of the Indian National Academy of
Engineering~(INAE).\vs{12}}

\noindent\rule{42.5pc}{1.5pt}

\maketitle

%\vspace*{-9pt}
\section{Introduction}

Over the years Discrete Cosine Transform (DCT) based encoding algorithms such
as, MPEG-2 \citep{1,2}, MPEG-4 \citep{3}, H.264 \citep{4,5} etc.
 have been proposed for video compression.
The core compression algorithm used in such algorithms  is the baseline JPEG
\citep{6} encoding algorithm. A~typical block diagram outlining these approaches
are shown in Figure~\ref{fig1}.

\noindent
 All of these approaches have the following
commonalities:

\begin{BL}
\item Use of block DCT.

\item The video stream is segmented into a Group Of Pictures
(GOP), each
       starting with an Intra (I) frame, which is encoded by the usual static
       image compression algorithm such as block DCT based base-line JPEG
       compression algorithm.

\item Block-wise (or Macroblock-wise) motion compensated error
prediction.
       Prediction may be made from previous and future frames. However, in
       our simplified framework we have considered only predictions from the
       previous frame. This type of predicted error frame is termed as a `P' %(P frames)
       frame. For each predictive frame, there is a set of motion vectors and
       corresponding error blocks, which are also further compressed using
  again the block-DCT based compression scheme, almost similar to baseline
       JPEG compression algorithm.

\item In the above schemes, typical size of a DCT block is $8
\times 8$ and that  of a macro-block for motion compensation is $16 \times 16$.
Colour images are represented in $Y-C_r-C_b$ colour space and usually $C_r$ and
$C_b$ components are subsampled at a ratio of $4:1:1$ before the application of
DCT.
\end{BL}
However one should note that there are  variations on the motion compensated
prediction algorithms, block and macro block sizes, intra-block prediction,
object based video streaming, rate control strategies etc. among these
approaches.

\enlargethispage{7pt}

Other than DCT based approaches, wavelet based approaches are also
increasingly being pursued in recent years. Specially, wavelet
\vadjust{\vspace*{-9pt}\pagebreak\vspace*{11.9pc}}based static image
compression schemes were found to be having better rate distortion
performance over the DCT based JPEG \citep{6} compression scheme.
Subsequently, the JPEG2000 \citep{7,8,9,10} standard has incorporated
wavelet based static image compression scheme. In~wavelet based approaches
usually 3-D wavelets \citep{11,12} are applied to~a~GOP. In~some work
\citep{13,14}, motion compensation is used to improve the performance.
However the usage of 3D wavelets over the GOP causes temporal artifacts
(equivalent to jittering etc.) at its boundaries \citep{15}. In~our
proposed algorithm \citep{16}, we~restrict ourselves to 2-D wavelets only.
The scheme is similar to the DCT based schemes, where Intra frames
($I$-frame) and motion compensated Predicted frames ($P$-frame) are
encoded using similar 2-D transform domain techniques. Like MPEG
\citep{17}, in our proposed algorithm we have used JPEG2000 (instead of
JPEG as the core compression technique in MPEG) towards this. It~may be
noted that JPEG2000 has already been used in the MJPEG2000 \citep{18}
video compression standard, where all the frames are intra frames and no
motion compensation is performed. The~MJPEG2000 usually provides superior
performance over MPEG-2 at very high bit rates. Our proposed scheme is a
generalisation over MJPEG2000, as~it allows motion compensation. In~many
cases, the proposed approach \citep{19} is found to be having better rate
distortion performance than the MJPEG2000 at very high bit rates. There
are a few advantages on using JPEG2000 over the JPEG-like schemes for
encoding prediction errors which have been adopted in the proposed
algorithm:\vs{-4}

\begin{NLh}{4}
\item[1] Since the wavelet transform can be applied over the
entire image, a JPEG2000
       image does not exhibit the blocky artifacts common in highly compressed
       traditional JPEG images.

\item[2] Scalability and multi-resolution representation make
JPEG2000 as ideal
       choice for scalable video encoding scheme.

\item[3] Precise frame wise rate control is possible while using
JPEG2000. This is significant in the context of errors being modelled as
{\it Gaussian distribution} \citep{21}. It~may be noted that MPEG-2 adopts
macroblock wise rate control (for~$P$~frames),  which makes error encoding
algorithm little different than the standard
       baseline JPEG encoding scheme used for $I$~frames. In~our proposed
       approach, rate control mechanisms
       for both $I$ and $P$~frames remain same, which makes the
encoder much simpler for
       implementation.

\item[4] The other advantage of JPEG2000 scheme is that the scope
of making the        scheme lossless or semi-lossless by using reversible
wavelets. This makes
        the proposed approach very attractive for the higher
bit rate video transmission,
        specially the medical videos, where physicians demand very high quality
        of the decompressed video.\vs{-2}
\end{NLh}

\noindent
 The proposed algorithm has the following distinct
features:\vs{-2}

\begin{NLh}{3}
\item[1] It uses similar motion compensation algorithm that is
used in MPEG-2. However, in the present work, we have demonstrated results using
only full pixel motion compensation and also using only predictive frames
($P$~frames).

\item[2] To make the scheme semi-lossless, we have used the
feed-back from the  compressed video stream in the prediction of errors. We~have
also used reversible wavelets for encoding error frames.

\item[3] Precise rate control is possible by allocating fixed bits
for $I$ and $P$~frames. For efficient rate distortion performance, we~have
adopted  dynamic rate control and distortion adjustment strategies.\vs{-2}
\end{NLh}
The paper is organised in the following sections. In~Section~\ref{base_algo} we
provide the basic outline of the proposed scheme and its performance on fixed
bit allocation. We~discuss about the rate and distortion control strategies in
Section~\ref{rate_control}, which make the proposed  scheme more efficient.
Section~\ref{discussion} presents a brief overall discussion of the performance
of the scheme. Finally, the conclusion is drawn in Section~\ref{conclusion}.
Experimental results and graphs are provided in the respective sections.
Further, we have considered the videos in YUV colour model for CIF and QCIF
formats throughout this paper.

\vspace*{-0.5pc}
\section{Base compression scheme}\label{base_algo}

In our scheme, we have used JPEG2000 for compressing the intra frames
or $I$~frames as well as the error frames obtained through motion
compensation. The $I$~frame compression is the usual JPEG2000 still
image compression. However, while compressing error frames
($P$~or~$B$), error values are computed for each macroblock of $Y$
($16\times 16$ pixels for each block), $U$ and $V$ ($8\times
 8$ blocks for each of $U$ and~$V$) components by estimating motion
vectors with respect to the previous (or/and next) frame(s). Motion compensation
is made over the decompressed previous frames. Then the whole error frame is
subjected to the JPEG2000 compression scheme. A~typical block diagram of the
scheme is shown in Figure~\ref{fig2}.

\noindent As JPEG2000 works with precise rate control (within a range of
target bit rates), initially we have used fixed bit per pixel allocation
for both $I$ and $P$~frames. A~parameter $\alpha$ has been considered
indicating the relative proportions (fractions) of bit investment on $I$
and $P$~frames. Let $R$ be the target rate in bits  per second (bps), $G$
be the GOP (with single $I$~frame and $(G-1)$ $P$~frames) and $f$ be the
frame per second. Let $h$ and $w$ be the height and width of a frame. The
bit per pixels for $I$~frame ($\beta_I$) and for $P$ frames ($\beta_P$)
are given~by:\vs{-2}
\begin{eqnarray}
\beta_I&=&\alpha\cdot \frac{R\cdot G}{f\cdot w\cdot h}\label{eqn_bpp_I}\\
\beta_P&=&(1-\alpha)\cdot \frac{R\cdot G}{f\cdot w\cdot h\cdot
(G-1)}.\label{eqn_bpp_P}
\end{eqnarray}
For studying the performance of the compression schemes, we have used the
Peak Signal to Noise Ratio (PSNR) measure. Let $I_s(x,y)$ denote the pixel
value at $(x,y)$ for the $s$th component ($s \in \{Y,U,V\}$) of the
original frame and $I'_s(x,y)$ denote the corresponding recovered value
after decompression. Then the PSNR is defined~as:\vs{4}
\begin{eqnarray}
\hspace*{-9pt}
\h{PSNR}&=&20\log_{10}\kern-2pt\left(\kern-2pt\frac{255}{\sqrt{\frac{\Sigma_{\forall
s \in \{Y,U,V\}}\Sigma_{\forall x}\Sigma_{\forall
y}(I_s(x,y)-I^{'}_s(x,y))^2}{\frac{3}{2}\cdot h\cdot
w}}}\kern-2pt\right)\kern-3pt.\nonumber\\[-2pt] \label{eqn_psnr}
\end{eqnarray}

\vspace*{-4pt} \noindent For the entire video, the {\it average} PSNR
value of all the frames has been considered. The effect of using
reversible and irreversible wavelets has also been studied. The JPEG2000
compression has been implemented using the {\it
Kakadu}\endnote{http://www.kakadusoftware.com} software. The number of
levels of decomposition for $I$~frame is taken as 5 and for $P$~frame the
value is~0. These values are chosen empirically. It~can~be observed that
the use of reversible wavelets leads to higher PSNR values at higher bit
rate. Experimental results have been shown in Figures~\ref{fig3}
and~\ref{fig4}, with the proposed MCJ2K using reversible wavelets.\vs{-1}

In Figures~\ref{fig3} and~\ref{fig4} we present the PSNR vs. \h{{\it bit-rate}}
performances for the proposed schemes for different values of $\alpha$ using
reversible and irreversible wavelets on QCIF and CIF videos respectively. The
graph show that the MCJ2K using reversible wavelets provides higher PSNR values
over MCJ2K with irreversible wavelets at high bit-rates. The PSNR values at
typical \h{bit-rates} using reversible wavelets with different values
of~$\alpha$ for QCIF and CIF videos, are shown in Tables~\ref{brqcif}
and~\ref{brcif} respectively. It~is observed that lower values of $\alpha$
performs better at high bit-rate. In Tables~\ref{cmpbrqcif} and~\ref{cmpbrcif}
the PSNR values are compared with those obtained by MPEG-2 and MJPEG2000 for
QCIF and CIF videos respectively. One may observe that the use of feedback from
the compressed stream improves the performance remarkably. The use of feedback
makes the lossless reconstruction of $P$ possible. In~fact, the PSNR values of
recovery of error frames and recovery of $P$~frame become equivalent in such
cases. This fact is stated in the Lemma~\ref{lemma_error_psnr}.\vs{2}

\begin{lemma}\label{lemma_error_psnr}
The PSNR value of the error recovery remains as same as the PSNR value of the
decompressed frame, when errors are computed from the previous decompressed
frame.\vs{2}
\end{lemma}

\proof{Proof} Let $\widehat{I}_{\rm prev}^{(d)}$ be the motion compensated
values from the decompressed previous frame.  Let $I_{\rm curr}$ be the original
current frame. Then the error values for the current frame are
obtained~as:\vs{2}
\begin{equation}
 E=I_{\rm curr}-\widehat{I}_{\rm prev}^{(d)}.
\end{equation}

\begin{table*}[!t]%Tab5
\caption{Results on measures of skewness and curtosis of error frames for {\it
Container. cif} video at 2449~Kbps}\label{curtcontainer}
{\NINE\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccccc@{}} \toprule
{\it Frame component} & {\it Avg\_Skewness} & {\it Avg\_Curtosis} & $\sigma$
{\it Skewness} & $\sigma$ {\it Curtosis} & {\it Min\_Curtosis}\\ \midrule Y &
\p{$-$}0.45 & 41.82 & 1.53  & 42.29 &9.53\\ U & $-$0.49 & \p{0}7.90  & 1.55 &
47.17 & 3.05\\
 V & $-$0.40 & \p{0}6.88  & 0.51  & 6.53 & 4.01\\
\botrule
\end{tabular*}}{}
\vspace*{4pt}
\end{table*}

\begin{table*}[!t]%Tab6
\caption{Results on measures of skewness and curtosis of error frames for {\it
News.cif} video at 2640~Kbps}\label{curtnews}
{\NINE\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccccc@{}} \toprule
{\it Frame component} & {\it Avg\_Skewness} & {\it Avg\_Curtosis} & $\sigma$
{\it Skewness} & $\sigma$ {\it Curtosis} & {\it Min\_Curtosis}\\ \midrule Y &
\p{$-$}0.39 & 96.08 & 1.98  & 46.56 & 5.90\\ U & $-$0.52 & 48.04  & 1.41 & 49.52
& 4.38\\
 V & $-$0.01 & 57.19  & 1.42 & 54.97 & 4.31\\
\botrule
\end{tabular*}}{}
\end{table*}

\begin{table*}[!t]%Tab7
\caption{Results on measures of skewness and curtosis of error frames for {\it
Foreman.cif} video at 3043~Kbps}\label{curtforeman}
{\NINE\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccccc@{}} \toprule
{\it Frame component} & {\it Avg\_Skewness} & {\it Avg\_Curtosis} & $\sigma$
{\it Skewness} & $\sigma$ {\it Curtosis} & {\it Min\_Curtosis}\\ \midrule Y &
$-$0.03 & 17.20 & 0.70  & 10.42 & 4.75\\
 U & $-$1.16 & 47.16  & 4.58  &165.36\p{0} & 2.79\\
V & \p{$-$}0.11 & 22.83  & 1.34  & 28.01 & 2.91\\ \botrule
\end{tabular*}}{}
\end{table*}

\noindent After decompression, let the error values be $E^{(d)}$. Hence,
the $I_{\rm curr}^{(d)}$ after decompression could be expressed~as:
\begin{equation}
I_{\rm curr}^{(d)}=\widehat{I}_{\rm prev}^{(d)}+E^{(d)}.
\end{equation}
Hence,\vs{-3}
\begin{eqnarray}
|E-E^{(d)}| & = & |(I_{\rm curr}-\widehat{I}_{\rm
prev}^{(d)})-E^{(d)}|\nonumber\\
 &=& |(I_{\rm curr}-\widehat{I}_{\rm prev}^{(d)})-(I_{\rm curr}^{(d)}-\widehat{I}_{\rm prev}^{(d)})|\nonumber\\
  &=&  |I_{\rm curr}-I_{\rm curr}^{(d)}|.
\end{eqnarray}
Hence the PSNR values for error recovery and frame recovery are same.\vs{-3}
\endproof

\noindent From the above lemma, we can observe that if the error recovery is
{\it lossless}, the decompressed frame is also {\it lossless}. This motivates us
to use the reversible wavelets. We~have observed that usage of reversible
wavelet slightly improves the scheme at the higher \h{bit-rate}. However, as the
dynamic ranges of errors increase by two folds, we have scaled the values by
$\frac{1}{2}$ before compression. For this reason, our scheme behaves as a
semi-lossless scheme at higher bit rates.

\subsection{Gaussian modelling of error frames}

Lemma~\ref{lemma_error_psnr} is also significant as it~converts the problem of
compressing individual frames to the compression of error frames. It~would be
interesting to study how best these errors are modelled by a Gaussian
distribution. In~that case we would be able to use the well known rate
distortion relationship of encoding of Gaussian distributed values, as described
by the following\break lemma.

\begin{lemma}\label{lemma_RD_gauss}
 The values of a random variable following Gaussian
distribution with the standard
 deviation $\sigma$ would require   $R(D)$ bits per symbol at a distortion
 {\rm (}the
 Mean Square Error {\rm (}MSE{\rm ))} of $D$ where $R(D)$
  holds the following relationship
with~$D$
\begin{equation}
R(D) = \begin{cases}
         \dfrac{1}{2} \log_2 \dfrac{\sigma^2}{D} &\quad 0 \le D \le
                    \sigma^2\\[1pc]
  0  &\quad D > \sigma^2\end{cases}.
\end{equation}
\end{lemma}

\proof{Proof} Refer \cite{20}.
\endproof

\noindent
 It was observed that in most cases the distributions
of error values do not  pass the significance levels (of~1\%) of {\it
Chi-Square} ($\chi^2$) {\it Goodness
 of Fit test} \citep{21} or the {\it Kolmogrov-Smirinov Test} \citep{21}.
 However, the distributions are
 observed as symmetric and highly peaked around its mean.
Skewness and curtosis are measures of asymmetry and sharpness of the height
relative to a normal distribution respectively.
In~Tables~\ref{curtcontainer}--\ref{curtforeman}, we~provide the average
skewness and average curtosis measures (along with their standard deviations).
They indicate that the Lemma~\ref{lemma_RD_gauss} would provide an upper bound
on the rate for a given distortion. We~have made use of this fact in designing
the rate control strategy.

\subsection{Varying macroblock sizes}

We have also studied the effect of different macroblock size on the performance
of the proposed scheme. If~errors are generated by larger macroblocks, it~is
expected the standard deviations of the error frames would be higher, on the
other hand lower macroblocks should have smaller standard deviations. But there
are increasing overheads on encoding the motion vectors for lower macroblock
sizes. If~we~assume the bit assignment on encoding full pixel motion vectors is
same (say $mv_b$ for each motion vector, in our case $mv_b = 8$~bits), a
macroblock of size $b_w \times b_h$ would require $\big\lceil \frac{w\cdot
h}{b_w\cdot b_h\vphantom{\sum_{1_1}}}\big\rceil$ number of motion vectors.
If~$\sigma_i$, $i=1,2,\dots,N$ is the standard deviation for the $i$th error
frames the average bit requirement $E_b$ for an error frame is given~by:\vs{-3}
\begin{equation}
E_b= \frac{1}{N}\sum_{i=1}^{N}\log_2(\sigma_i)+\frac{mv_b\cdot\big\lceil
\frac{w\cdot h}{b_w\cdot b_h}\big\rceil}{w\cdot h}.
\end{equation}
In Table~\ref{Avgbit}, we have shown the theoretical average bit
requirements for varying macroblock size. In~Figure~\ref{fig5}, PSNR vs.
rate curves are drawn for varying macroblock sizes. It~is empirically
observed that the macroblock size $16 \times 16$ yield better performance
empirically. Theoretical bounds on the average bit requirement are also
lower for different videos in such case. In~Figure~\ref{fig5}(a),
macroblock size $32 \times 32$ is showing better performance while in
Figure~\ref{fig5}(b) and (c), macroblock size $16 \times 16$ shows better
performance at lower bit-rates. Moreover \h{MPEG-2} also uses a macroblock
size of $16 \times 16$ and since our scheme uses the fundamentals of
\h{MPEG-2}, thus the macroblock size $16 \times 16$ is chosen for
implementing our scheme.

\begin{table}%%Tab8
\def\arraystretch{1.1}
\caption{Average bit requirement for Y-component with different macroblock
size ($8\times8$, $16\times16$ and $32\times32$) at higher bit-rate for
CIF videos (Container, News, Foreman)}\label{Avgbit}
{\NINE\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}lccc@{}}
\toprule {\it Video} & {\it Vb\_8} & {\it Vb\_16} & {\it Vb\_32}\\
\midrule Container & 0.80 & 0.72 & {\bf 0.70}\\
 News & {\bf 0.95} & 1.05 & 1.24\\
 Foreman & {\bf 1.82} & 1.85 & 1.99\\
\botrule
\end{tabular*}}
{}
\end{table}

\subsection{Varying length of GOP}

We have also observed the effect of variation of the length of GOP. In our
scheme, as propagation of errors is arrested by the feedback from the compressed
stream, the effect on variation of the GOP length is less (refer
Figure~\ref{fig6}). In~fact, the scheme works better for larger GOP length, as
long as, there is a gain in rate for encoding a $P$ frame compared to that of an
Intra encoding. However, larger GOP will also cause, increasing values of
standard deviations for the error frames at its trailers. This puts a
restriction on the length of the GOP. This leads us to design a compression
scheme for adaptive GOP. If~the standard deviation of an error frame exceeds a
threshold (called as $\sigma$th in our work) the GOP is initialised.
In~Figure~\ref{fig7}, we~have demonstrated the performances for adaptive GOPs.
In~Table~6, we~have also provided the average lengths of GOP at
varying~$\sigma$th.\vs{-3}

\section{Rate control} \label{rate_control}

As discussed earlier, the use of JPEG2000 provides an opportunity for precise
rate control. Moreover, the Gaussian nature of error frames motivates us to use
the following lemma for the rate control of the MCJ2K scheme with fixed GOP.

\begin{lemma}\label{lemma_rt_gauss}
Let $\overline{\beta_P}$ be the average bit per pixel for encoding a $P$ frame.
Let $\sigma_i$, $i=1,2,\dots,(G-1)$ be the standard deviation of the $i$th
$P$~frame of a GOP of the length~$G$. Then, the optimal bit per pixel at which
the $i$th $P$~frame would be encoded is given~by:
\begin{equation}\label{eqn_bpp_optimal}
\beta_{P_i}=\overline{\beta_P}+ \frac{1}{2} \log_2 \frac{\sigma^2}{\sigma^{*^2}}
\end{equation}
where
\begin{equation}\label{eqn_sigma_gm}
\sigma^*= \big(\Pi_{i=1}^{(G-1)} \sigma_i\big)^{\frac{1}{(G-1)}}
\end{equation}
with the optimum rate allocation the $i$th frame distortion $D_i$ becomes,
\begin{equation}
D_i=\sigma^{*^2} 2^{-2\overline{\beta_P}}.
\end{equation}
\end{lemma}

\proof{Proof} Refer \cite{22}.\vs{-1}
\endproof

\subsection{Rate control of the MCJ2K}

The above lemma motivates us to design the following rate control algorithm for
the MCJ2K.\vs{-1}

\vspace*{9pt} \noindent
 Algorithm {\it Rate\_Control\_MCJ2K}

\vspace*{8pt}
 \noindent
{\it Encoding Parameters}: Length of GOP ($G$), relative bit investment for
$I$~frame ($\alpha$), rate~($R$).\vs{-1}

\begin{NLh}{3}
\item[1] Determine bit per pixel for encoding $I$ frame
($\beta_I$) and average
 bpp for encoding a $P$ frame ($\overline{\beta_P}$) using
 equations~(\ref{eqn_bpp_I}) and~(\ref{eqn_bpp_P}).

\item[2] Compute the geometric mean of the standard
deviations (refer equation~(\ref{eqn_sigma_gm}))  for the next $GOP$ from the
current one and use it for the purpose of~optimal bit assignment following the
(refer equation~(\ref{eqn_bpp_optimal})).

\item[3] Compute the bpp for $i$th $P$ frame ($\beta_{P_i}$)
following~the Lemma~\ref{lemma_rt_gauss} and encode them\break accordingly.
\end{NLh}
\vspace*{0.3pc} \noindent End {\it Rate\_Control\_MCJ2K}

\vspace*{8pt} \noindent
 In this algorithm, $P$ frames for the first GOP are
encoded using the average bpp as computed from (refer
equation~(\ref{eqn_bpp_P})). The geometric mean of the standard deviations of
all the $P$~frames is also computed and used for encoding $P$~frames of the next
GOP using (refer equation~(\ref{eqn_bpp_optimal})). The process is repeated for
all the subsequent GOPs in the same way. The performance is shown in the
Figure~\ref{fig7}. From the Table~\ref{brcif}, it is observed that $\alpha=50\%$
gives better PSNR value. As such $\sigma$ value at 5 and $\alpha$ at 50\% are
typically chosen for studying the performance of {\it Rate\_Control\_MCJ2K}.

At lower bit-rate, the achieved bit-rate is almost as same as the given target
bit-rate but at higher bit-rate the achieved bit-rate gets saturated even if the
target bit-rate is increased (refer Figure~\ref{fig8}).

\subsection{Distortion control of the MCJ2K}\label{distortion_control}

Interestingly, the MCJ2K scheme is more suitable for controlling distortion or
quality of the decompressed video. In~this case, given a target PSNR~$\Omega$,
the MCJ2K attempts to encode at the optimal rate for achieving the target rate.
The
 distortion $D$ is computed from $\Omega$ as follows:
\begin{equation}\label{eqn_D_PSNR}
 D=255^2\times 10^{-\frac{\Omega}{10}}.
\end{equation}
Then following the Lemma~\ref{lemma_RD_gauss}, $P$ frames are encoded. The
performance  of this strategy with different bpp's used for encoding $I$~frames
is shown  in Figure~\ref{fig9}. The achieved PSNR values given the target PSNR
values are also
 shown in the Figure~\ref{fig10}.

However, one may observe that there is a considerable gap between the achieved
PSNR values and the target PSNR values. To~reduce these gaps, we have adjusted
the target PSNR values after encoding of each frame and measuring its PSNR with
the original frame. Let the achieved PSNR value for the $i$th frame
be~$\Omega^{(d)}$.
 Then the target PSNR value for the $(i+1)$th frame is computed~as:\vs{-3}
\begin{equation}
    \Omega^{(i+1)}=2\Omega^{(i)}-\Omega^{(d)}.
\end{equation}
For the first $P$ frame of a GOP, the target PSNR is initialised to~$\Omega$.
The PSNR vs. rate curves with this distortion control strategy (referred as {\it
Quality\_Control\_MCJ2K})
 are shown in the Figure~\ref{fig11}. In~Figure~\ref{fig12},
 achieved PSNR values against the target
 PSNR values are also shown. It~may be noted that due to the
PSNR feedback, the achieved  bit-rate gets saturated faster as compared to the
scheme without using this feedback (refer Figure~\ref{fig9}).\vs{-3}

\section{Discussion}\label{discussion}

In the {\it base compression} scheme of MCJ2K (refer Section~\ref{base_algo}),
we~have experimented the bit-investment on $I$ and $P$~frames, using a
parameter~$\alpha$. It~was found empirically that lower value of $\alpha$
($\alpha = 0.5$), produces better PSNR value at the higher bit-rate. We~also
experimented on using both reversible and irreversible wavelets and found that
reversible wavelets outperforms irreversible wavelets (refer
Lemma~\ref{lemma_error_psnr}).

%\vfill\eject

Using Gaussian modelling of error frames (refer Lemma~\ref{lemma_RD_gauss}),
MCJ2K scheme is subjected to varying length of GOP base on a threshold
value~$\sigma$. We~also found that GOP size increases with higher value of
$\sigma$ (refer Table~\ref{Avggop}).

\def\thetable{9}
\begin{table}[h]%%Tab9
\def\arraystretch{1.1}
\caption{Average length of GOP for different $\sigma$th for CIF
videos}\label{Avggop}
{\NINE\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}lcc@{}} \toprule {\it
Video} & $\sigma={\it 4.0}$ & $\sigma={\it 5.0}$\\ \midrule Container & 150 &
300\\
 News & \p{0}30 & \p{0}75\\
  Foreman & \p{00}5 & \p{0}10\\
\botrule
\end{tabular*}}{}
\vspace*{5pt}
\end{table}

\noindent In {\it rate control} of MCJ2K (refer Section~\ref{rate_control}),
Gaussian modelling of error frames along with $\alpha$ (the relative
bit-investment for $I$~frame), has been implemented and found that it performs
better than our base compression scheme at higher bit-rate.

In {\it distortion control} (quality control without feedback) of MCJ2K
scheme (refer Section~\ref{distortion_control}), given a target
PSNR~$\gamma$, the proposed MCJ2K algorithm attempts to encode at the
optimal rate for achieving the target rate using equation~(11) and
Lemma~\ref{lemma_RD_gauss}. We~also~have quality control with feedback in
which the target PSNR values are adjusted after encoding each frame (refer
equation~(12)). The effect of having feedback can be observed from
Figures~\ref{fig10} and~\ref{fig12}.

From the Table~\ref{res1}, it can be observed that quality control with
feedback scheme of MCJ2K produces better PSNR values than that of quality
control without feedback and rate control schemes of MCJ2K
(refer~Table~\ref{res1}). However, it~may be noted that quality control
with feedback saturates faster than both \h{quality-control} without
feedback and rate control schemes of MCJ2K (refer Table~\ref{res2}). It~is
also worth mentioning that due to lower bound constraints, in both types
of quality control schemes (with and without feedback), the scheme works
from a certain high bit-rate only as compared to the rate control scheme
of MCJ2K (refer Figures~\ref{fig7},~\ref{fig9} and~\ref{fig11}).

Finally, in optimal bit allocation strategy, the MCJ2K scheme adjusts
$\alpha$ for bit-investment for both the $I$ and $P$~frames.
It~behaves like MCJ2K rate control scheme at lower bit-rate but at
the higher bit-rate it tends to behave like MJPEG2000, where the GOP
size tends to become one. Hence this scheme works better than MPEG-2
at very high bit-rate (refer Figure~\ref{fig13}).

\section{Conclusion}\label{conclusion}
\vspace*{6pt}

In this paper we have  proposed a novel video compression algorithms which
use JPEG2000 as the core compression engine for compressing {\it intra}
frames as well as the motion compensated {\it prediction} error frames.
The rate and quality control algorithms for the MCJ2K scheme have been
developed and the MCJ2K  is found to perform better than MPEG-2 and
MJPEG2000 at higher bit-rates for many  cases. Extensive experimentations
have been carried out and the results are compared with standard schemes.
It~is also found that MCJ2K performs better for low motion videos as
compared to high motion videos. The proposed MCJ2K scheme can find its
applications like HDTV video conferencing, medical video transmission
etc., where high quality video is required, even at the cost of high
bit-rate.

\begin{thebibliography}{10}

\bibitem[\protect\citeauthoryear{Antonini et~al.}{1992}]{10}
Antonini, M., Barlaud, M., Mathieu,~P. and Daubechies,~I. (1992) `Image coding
using wavelet transform', {\it IEEE  Transactions on Image Processing}, Vol.~1,
No.~2, \h{pp.205--222}.

%\enlargethispage{-2pt}

\bibitem[\protect\citeauthoryear{Dufaux and Ebrahimi}{2003}]{18}
Dufaux, F. and Ebrahimi, T. (2003) `Motion JPEG2000 for wireless applications',
{\it Signal Processing Laboratory (EPLF)}, {\it Proceedings of First
International JPEG 2000 Workshop, Motion Analysis and Image Sequence
Processing}, IEEE, Lugano, Switzerland, July, pp.1--5.% \tc{AUTHOR PLEASE SUPPLY
%PAGE LOCATION AND PAGE RANGE.}

\bibitem[\protect\citeauthoryear{Gray and Neuhoff}{1998}]{22}
Gray, R.M. and Neuhoff, D.L. (1998) `Quantization: invited paper', {\it IEEE
Transactions on Information Theory}, Vol.~44, No.~6, pp.2325--2383.

\bibitem[\protect\citeauthoryear{He~and Mitra}{2005}]{20}
He, Z. and Mitra, S.K. (2005) `From rate-distortion analysis to
resource-distortion analysis', {\it IEEE Transaction on Circuit and
System for Video Technology}, Vol.~5, No.~3, pp.6--18.

\bibitem[\protect\citeauthoryear{LeGall}{1991}]{17}
LeGall, D. (1991) `MPEG: a video compression standard for multimedia
applications', {\it Communication of ACM}, Vol.~34, No.~4, pp.46--58.

\bibitem[\protect\citeauthoryear{Leung and Taubman}{2005}]{13}
Leung, R. and Taubman, D. (2005) `Transform and embedded coding techniques for
maximum efficiency and random accessibility in 3-D scalable compression', {\it
IEEE Transactions on Image Processing}, Vol.~14, No.~14, pp.1632--1646.

\bibitem[\protect\citeauthoryear{Liang et~al.}{2005}]{15}
Liang,~J., Tu,~C. and Tran,~T.D. (2005) `Optimal block boundary
pre/postfiltering for wavelet-based image and video compression', {\it IEEE
Transactions on Image Processing}, Vol.~14, No.~12, pp.2151--2158.

\bibitem[\protect\citeauthoryear{Montgomery et~al.}{2001}]{21}
Montgomery, D.C., Runger, G.C. and Hubele,~N.F. (2001) {\it Engineering
Statistics}, 2nd~ed., John~Wiley, New~York.

\bibitem[\protect\citeauthoryear{Rath et~al.}{2006}]{16}
Rath,~A., Tuithung,~T., Ghosh, S.K. and Mukherjee,~J. (2006)
`Compressing videos using JPEG2000', {\it Proceedings of the 3rd
Workshop on Computer Vision, Graphics and Image Processing}, 12--13th
January, Hyderabad, Andra~Pradesh, India, pp.93--97.

\bibitem[\protect\citeauthoryear{Senda}{1995}]{1}
Senda, Y. (1995) `A~simplified motion estimation using an approximation for the
MPEG-2 real time encoder', {\it International Conference on Acoustics, Speech
and Signal Processing}, Detroit, MI, USA, pp.3221--3224. %\tc{AUTHOR PLEASE
%SUPPLY LOCATION.}

\bibitem[\protect\citeauthoryear{Sikora}{1997}]{3}
Sikora, T. (1997) `The MPEG-4 video standard verification model', {\it IEEE
Transactions on Circuit and System for Video Technology}, Vol.~7, pp.19--31.

\bibitem[\protect\citeauthoryear{Skodras et~al.}{2000}]{9}
Skodras, A.N., Christopoulos, C.A. and Ebrahimi,~T. (2000) `JPEG2000:
the upcoming still image compression standard', {\it Proceeding of
the 11th Portuguese Conf. on Pattern Recognition (RECPA00D~20:
Invited paper)}, \h{11--12th} May, Prto, Portugal,  pp.359--366.

\bibitem[\protect\citeauthoryear{Tudor}{1995}]{2}
Tudor, P.N. (1995) `MPEG-2 video compression', {\it Electronics and
Communication Engineering Journal}, pp.257--264.

\bibitem[\protect\citeauthoryear{Tuithung et~al.}{2006}]{19}
Tuithung, T., Sinha, D., Ghosh,~S.K. and Mukherjee,~J. (2006)
`Bit-investment policy of MCJ2K: a~new video codec', {\it Proceedings
of the 1st IEEE International Conference on Signal and Image
Processing}, \h{7--9th}~December, Hubli, Karnataka, India,
pp.648--652.

\bibitem[\protect\citeauthoryear{Taubman and Marcellin}{2002}]{7}
Taubman, D. and Marcellin, M.W. (2002) {\it JPEG 2000: Image Compression
Fundamentals, Standards and Practice}, Kluwer, MA, Norwell.

\bibitem[\protect\citeauthoryear{Taubman}{1999}]{8}
Taubman, D. (1999) `High performance scalable image compression with EBCOT',
{\it Proceeding of IEEE International Conf. on Image Processing (ICIP99)},
Tokyo, Japan, Vol.~3, pp.344--348.

\bibitem[\protect\citeauthoryear{Taubman and Zakhor}{1994}]{12}
Taubman, D. and Zakhor, A. (1994) `Multirate 3-D subband coding of video', {\it
IEEE Transactions on Image Processing}, Vol.~3, No.~5, pp.572--589.

\bibitem[\protect\citeauthoryear{Wallace}{1991}]{6}
Wallace, G.K. (1991) `The JPEG still picture compression standard', {\it
Communication of ACM}, Vol.~34, No.~4, pp.30--44.

\bibitem[\protect\citeauthoryear{Wiegand et~al.}{2003}]{4}
Wiegand, T., Sullivan, G.J., Bjontegaard,~G. and Luthra,~A. (2003) `Overview of
the H.264/AVC video coding standard', {\it IEEE Transaction on Circuit and
System for Video Technology}, Vol.~13, pp.560--576.

\bibitem[\protect\citeauthoryear{Wiegand and Sullivan}{2005}]{5}
Wiegand, T. and Sullivan, G.J. (2005) `Video compression-from concepts to the
H.264/AVC standard', {\it Proceeding of the IEEE}, Vol.~93, No.~1, pp.18--31.
%\tc{AUTHOR PLEASE SUPPLY PAGE RANGE.}

\bibitem[\protect\citeauthoryear{Xu~et~al.}{1994}]{11}
Xu, J., Xiong, Z., Li, S. and Zhang,~Y. (1994) `3-D embedded subband coding with
optimal truncation (3-D~Escot)', {\it J.~Appl. Comput. Harmon. Anal.}, Vol.~10,
pp.290--315.

\bibitem[\protect\citeauthoryear{Xu et~al.}{2002}]{14}
Xu, J., Xiong, Z., Li, S. and Zhang,~Y.Q. (2002) \h{`Memory-constrained} 3-D
wavelet transform for video coding without boundary effects', {\it IEEE
Transaction on Circuit and System for Video Technology}, Vol.~12, No.~9,
\h{pp.812--818}.

\end{thebibliography}

\def\notesname{Note}

\theendnotes

%\section*{Query}
%
%\tc{AQ1: AUTHOR PLEASE CITE FIGURE 8 IN TEXT.}

\end{document}

### 2011-11-27 papers/IJCSE/foropencl.tex
edits

@@
 
-\documentclass{acm_proc_article-sp}
+\documentclass{doublecol-new}
 
@@
 
+\LRH{M. Sottile et~al.}
+
+\RRH{ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming}
+
 \title{ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming}
 
-\numberofauthors{6}
+% 1st. author
+\authorA{Matthew J. Sottile}
+\affA{Galois, Inc.\\
+ Portland, OR 97204\\
+ E-mail: mjsottile@computer.org}
 
-\author{
-% 1st. author
-\alignauthor
-Matthew J. Sottile\\
-       \affaddr{Galois, Inc.}\\
-%       \affaddr{421 SW 6th Ave. Suite 300}\\
-       \affaddr{Portland, OR 97204}\\
-       \email{mjsottile@computer.org}
 % 2nd. author
-\alignauthor
-Craig E Rasmussen\\
-       \affaddr{Los Alamos National Laboratory}\\
-%       \affaddr{CCS-7, MS B287}\\
-       \affaddr{Los Alamos, NM 87545}\\
-       \email{crasmussen@lanl.gov}
+\authorB{Craig E Rasmussen}
+\affB{Los Alamos National Laboratory\\
+ Los Alamos, NM 87545\\
+ E-mail: crasmussen@lanl.gov}
+
 % 3rd. author
-\alignauthor
-Wayne N. Weseloh\\
-       \affaddr{Los Alamos National Laboratory}\\
-       %\affaddr{XTD-1, MS T082}\\
-       %\affaddr{Los Alamos, NM 87545}\\
-       \email{weseloh@lanl.gov}
-\and  % use '\and' if you need 'another row' of author names
+\authorC{Wayne N. Weseloh}
+\affC{Los Alamos National Laboratory\\
+ Los Alamos, NM 87545\\
+ E-mail: weseloh@lanl.gov}
+
 % 4th. author
-\alignauthor
-Robert W. Robey\\
-       \affaddr{Los Alamos National Laboratory}\\
-       %\affaddr{XCP-2, MS T086}\\
-       %\affaddr{Los Alamos, NM 87545}\\
-       \email{brobey@lanl.gov}
+\authorD{Robert W. Robey}
+\affD{Los Alamos National Laboratory\\
+ Los Alamos, NM 87545\\
+ E-mail: brobey@lanl.gov}
+
 % 5th. author
-\alignauthor
-Daniel Quinlan\\
-       \affaddr{Lawrence Livermore National Laboratory}\\
-       \email{dquinlan@llnl.gov}
+\authorE{Daniel Quinlan}
+\affE{Lawrence Livermore National Laboratory\\
+ Livermore, CA 94550\\
+ E-Mail: dquinlan@llnl.gov}
+
 % 6th. author
-\alignauthor
-Jeffrey Overbey\\
-       \affaddr{University of Illinois at Urbanna-Champaign}\\
-       \email{overbey2@illinois.edu}
-}
+\authorF{Jeffrey Overbey}
+\affF{University of Illinois at Urbanna-Champaign\\
+ Champaign, IL 61820\\
+ E-mail: overbey2@illinois.edu}
 
@@
 
-\balancecolumns
 \end{document}

### 2011-11-27 papers/IJCSE/programming-model.tex
edits

--- 2011-08-30 papers/GPUScA-2011/programming-model.tex
@@
 Washington~\cite{chamberlain04zpl} and Chapel, an HPCS language developed by
-Cray~\cite{chamberlain_chapel}.  In ZPL, programmers adopt a similar global view
+Cray~\cite{chamberlainchapel}.  In ZPL, programmers adopt a similar global view
 of computation over arrays, but define their computations based on regions,

### 2011-11-27 papers/IJCSE/foropencl.tex
more edits

@@
 \affA{Galois, Inc.\\
- Portland, OR 97204\\
+% Portland, OR 97204\\
  E-mail: mjsottile@computer.org}
@@
 \affB{Los Alamos National Laboratory\\
- Los Alamos, NM 87545\\
+% Los Alamos, NM 87545\\
  E-mail: crasmussen@lanl.gov}
@@
 \affC{Los Alamos National Laboratory\\
- Los Alamos, NM 87545\\
+% Los Alamos, NM 87545\\
  E-mail: weseloh@lanl.gov}
@@
 \affD{Los Alamos National Laboratory\\
- Los Alamos, NM 87545\\
+% Los Alamos, NM 87545\\
  E-mail: brobey@lanl.gov}
@@
 \affE{Lawrence Livermore National Laboratory\\
- Livermore, CA 94550\\
+% Livermore, CA 94550\\
  E-Mail: dquinlan@llnl.gov}
@@
 \affF{University of Illinois at Urbanna-Champaign\\
- Champaign, IL 61820\\
+% Champaign, IL 61820\\
  E-mail: overbey2@illinois.edu}

### 2011-11-27 papers/IJCSE/foropencl.tex
edits

@@
 
+%% bio
+\begin{bio}
+M. Sottile completed his PhD in 2006 from the University of New Mexico
+from the Department of Electrical and Computer Engineering.  He is
+presently working as a researcher at Galois, Inc. in Portland, Oregon
+leading research work efforts in scientific and high performance computing.
+His research interests include programming languages, compilation techniques,
+computational physics and applied mathematics.
+
+\noindent Craig has a bio.
+
+\noindent Wayne has a bio.
+
+\noindent Bob has a bio.
+
+\noindent Dan has a bio.
+
+\noindent Jeff has a bio.
+\end{bio}
+
 \maketitle

### 2011-11-27 papers/IJCSE/rose-transform.tex
edits

@@
-\subsection{Kernel transformation}
+\subsection{Fortran to OpenCL code conversion}
 
-In addition to the transformation of the code at call sites to copy
-data back and forth from the device and invoke the OpenCL kernel, we
-must also provide transformations of the Fortran kernel logic to the
-equivalent OpenCL C code.  This transformation is fairly
-straightforward, and maps a restricted subset of Fortran directly to
-its C equivalent.  The transformation is based on a regular traversal
+Programmers using ForOpenCL write their code in standard Fortran, and
+a sequence of code transformations are applied to generate equivalent
+code in C-based OpenCL.  These transformations are fairly
+straightforward, and map a restricted subset of Fortran directly to
+the C equivalent.  The transformation is based on a regular traversal
 of the ROSE Sage AST that represents the Fortran code.  The visitor
 used in the traversal handles five types of Sage AST Nodes: {\tt
-  SgAllocateStatment}, {\tt SgFunctionDeclaration},
-{\tt SgVariableDeclaration}, {\tt SgFunctionCallExp}, and
-{\tt SgExprStatement}.  These encompass the primary activities that are
-commonly encountered in computational kernels: allocation, function invocation,
-variable and function declaration, and expressions that correspond to
-arithmetic operations.
+  SgAllocateStatment}, {\tt SgFunctionDeclaration}, {\tt
+  SgVariableDeclaration}, {\tt SgFunctionCallExp}, and {\tt
+  SgExprStatement}.  These encompass the primary activities that are
+commonly encountered in computational kernels: allocation, function
+invocation, variable and function declaration, and expressions that
+correspond to arithmetic operations.
+
+In order to perform the language conversion, two ROSE {\tt SgSourceFile}
+objects are created -- one corresponding to the input Fortran code, and
+one corresponding to the output OpenCL code.  The transformation performs
+a traversal of the input Fortran code during which AST nodes are inserted
+into the OpenCL code to represent the equivalent code in the output language.
+The input Fortran code is the unmodified code that defines the pure
+elemental functions and related code that represent the computations to be
+performed on the accelerator.  The output is seeded with a template C
+source file ({\tt cl\_template.c}) that comes pre-defined with a set of
+helper functions that are used by the generated OpenCL code.  These
+generic pre-defined functions are primarily related to memory indexing (such
+as mapping 2D logical array indices to physical 1D addresses).  Our current
+prototyle includes a limited set of helper functions sufficient to demonstrate
+the shallow water model code, but could conceivably be expanded to include
+higher dimensional arrays with little effort.

### 2011-11-29 papers/IJCSE/abstract.tex
fixedits

--- 2011-11-27 papers/IJCSE/foropencl.tex
@@
-%
-% I'm adding an outline here so I can see how everything fits together
-
-%
-% \title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
-%
-
-%
-% \abstract
-%
-
-%
-% \section{Introduction}
-%    \subsection{Approach}
-%    \subsection{Why Fortran?}
-%    \subsection{Comparison to Other Languages}
-%
-
-%
-% \section{Programming model}
-%    \subsection{Fortran Syntax}   
-%       \subsubsection*{Array notation}
-%       \subsubsection*{Elemental functions}
-%       \subsubsection*{Pure procedures}
-%       \subsubsection*{Shift functions}
-%       \subsubsection*{Regions}
-%    \subsection{New Functions}
-%    \subsection{Parallelism}
-%    \subsection{Limitations}
-%
-
-%
-% \section{Shallow Water Model}
-%    \subsection{Equations}
-%
-
-%
-% \section{Source-To-Source Transformations}
-%    \subsection{ForOpenCL}
-%       \subsubsection{array syntax}
-%       \subsubsection{where construct}
-%    \subsection{New functions}
-%       \subsubsection{region}
-%    \subsection{Static Analysis}
-%       \subsubsection{Analysis not required}
-%    \subsection{Simplifying Assumptions}
-%
-
-%
-% \section{Performance}
-%
-
-%
-% \section{Conclusions}
-%
-
-\documentclass{doublecol-new}
-
-\usepackage{cite}
-\usepackage{graphicx}
-%\usepackage[cmex10]{amsmath}
-%\usepackage{amssymb}
-%\usepackage{array}
-\usepackage{url}
-\usepackage{listings}
-
-\lstdefinestyle{FortranLike}{float,frame=lines,language=Fortran,commentstyle=\ttfamily,basicstyle=\ttfamily}
-\lstdefinestyle{CLike}{float,frame=lines,language=C,commentstyle=\ttfamily,basicstyle=\ttfamily}
-\lstdefinestyle{NoFloatCLike}{frame=lines,language=C,commentstyle=\ttfamily,basicstyle=\ttfamily}
-
-\begin{document}
-
-\LRH{M. Sottile et~al.}
-
-\RRH{ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming}
-
-\title{ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming}
-
-% 1st. author
-\authorA{Matthew J. Sottile}
-\affA{Galois, Inc.\\
-% Portland, OR 97204\\
- E-mail: mjsottile@computer.org}
-
-% 2nd. author
-\authorB{Craig E Rasmussen}
-\affB{Los Alamos National Laboratory\\
-% Los Alamos, NM 87545\\
- E-mail: crasmussen@lanl.gov}
-
-% 3rd. author
-\authorC{Wayne N. Weseloh}
-\affC{Los Alamos National Laboratory\\
-% Los Alamos, NM 87545\\
- E-mail: weseloh@lanl.gov}
-
-% 4th. author
-\authorD{Robert W. Robey}
-\affD{Los Alamos National Laboratory\\
-% Los Alamos, NM 87545\\
- E-mail: brobey@lanl.gov}
-
-% 5th. author
-\authorE{Daniel Quinlan}
-\affE{Lawrence Livermore National Laboratory\\
-% Livermore, CA 94550\\
- E-Mail: dquinlan@llnl.gov}
-
-% 6th. author
-\authorF{Jeffrey Overbey}
-\affF{University of Illinois at Urbanna-Champaign\\
-% Champaign, IL 61820\\
- E-mail: overbey2@illinois.edu}
-
-\maketitle
-
 \begin{abstract}
-  Emerging GPU architectures for high performance computing are well suited to a
-  data-parallel programming model.  This paper presents preliminary work
-  examining a programming methodology that provides Fortran programmers with access
-  to these emerging systems.  We use array constructs in Fortran to
-  show how this infrequently exploited, standardized language feature is easily
-  transformed to lower-level accelerator code.  The transformations in ForOpenCL are based
-  on a simple mapping from Fortran to OpenCL.  We demonstrate, using a
-  stencil code solving the shallow-water fluid equations, that the performance
-  of the ForOpenCL compiler-generated transformations is comparable with that of
-  hand-optimized OpenCL code.
+  Emerging GPU architectures for high performance computing are well
+  suited to a data-parallel programming model.  This paper presents
+  preliminary work examining a programming methodology that provides
+  Fortran programmers with access to these emerging systems.  We use
+  array constructs in Fortran to show how this infrequently exploited,
+  standardized language feature is easily transformed to lower-level
+  accelerator code.  The transformations in ForOpenCL are based on a
+  simple mapping from Fortran to OpenCL.  We demonstrate, using a
+  stencil code solving the shallow-water fluid equations, that the
+  performance of the ForOpenCL compiler-generated transformations is
+  comparable with that of hand-optimized OpenCL code.
 \end{abstract}
-
-% A category with the (minimum) three required fields
-%\category{D.3.3}{Language Constructs and Features}{Concurrent programming structures}
-
-\input{intro}
-\input{programming-model}
-\input{shallow-water}
-\input{source-to-source}
-%\input{static-analysis}
-\input{performance}
-
-\section{Conclusions}
-
-The sheer complexity of programming for clusters of many or multi-core
-processors with tens of millions threads of execution makes the simplicity of
-the data-parallel model attractive.  The increasing complexity of
-today's applications (especially in light of the increasing complexity
-of the hardware) and the need for portability across architectures
-make a higher-level and simpler programming model like data-parallel
-attractive.
-
-The goal of this work has been to exploit source-to-source transformations that
-allow programmers to develop and maintain programs at a high-level of
-abstraction, without coding to a specific hardware architecture.
-Furthermore these transformations allow multiple hardware architectures
-to be targeted without changing the high-level source.  It also removes the
-necessity for application programmers to understand details of the accelerator
-architecture or to know OpenCL.
-
-%ACKNOWLEDGMENTS are optional
-\section{Acknowledgments}
-This work was supported in part by the Department of Energy Office of Science,
-Advanced Scientific Computing Research.
-
-%%TODO \cite{chamberlain04zpl, roth97stencils}
-
-\bibliographystyle{abbrv}
-\bibliography{foropencl}
-
-\end{document}

### 2011-11-29 papers/IJCSE/foropencl.tex
fixedits

@@
-%
-% I'm adding an outline here so I can see how everything fits together
+%%
+\documentclass{doublecol-new}
+%%
 
-%
-% \title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
-%
-
-%
-% \abstract
-%
-
-%
-% \section{Introduction}
-%    \subsection{Approach}
-%    \subsection{Why Fortran?}
-%    \subsection{Comparison to Other Languages}
-%
-
-%
-% \section{Programming model}
-%    \subsection{Fortran Syntax}   
-%       \subsubsection*{Array notation}
-%       \subsubsection*{Elemental functions}
-%       \subsubsection*{Pure procedures}
-%       \subsubsection*{Shift functions}
-%       \subsubsection*{Regions}
-%    \subsection{New Functions}
-%    \subsection{Parallelism}
-%    \subsection{Limitations}
-%
-
-%
-% \section{Shallow Water Model}
-%    \subsection{Equations}
-%
-
-%
-% \section{Source-To-Source Transformations}
-%    \subsection{ForOpenCL}
-%       \subsubsection{array syntax}
-%       \subsubsection{where construct}
-%    \subsection{New functions}
-%       \subsubsection{region}
-%    \subsection{Static Analysis}
-%       \subsubsection{Analysis not required}
-%    \subsection{Simplifying Assumptions}
-%
-
-%
-% \section{Performance}
-%
-
-%
-% \section{Conclusions}
-%
-
-\documentclass{doublecol-new}
-
+%\usepackage{natbib, stfloats}
 \usepackage{cite}
@@
 \begin{document}
+
+\thispagestyle{plain}
 
@@
 
-\begin{abstract}
-  Emerging GPU architectures for high performance computing are well suited to a
-  data-parallel programming model.  This paper presents preliminary work
-  examining a programming methodology that provides Fortran programmers with access
-  to these emerging systems.  We use array constructs in Fortran to
-  show how this infrequently exploited, standardized language feature is easily
-  transformed to lower-level accelerator code.  The transformations in ForOpenCL are based
-  on a simple mapping from Fortran to OpenCL.  We demonstrate, using a
-  stencil code solving the shallow-water fluid equations, that the performance
-  of the ForOpenCL compiler-generated transformations is comparable with that of
-  hand-optimized OpenCL code.
-\end{abstract}
+\input{abstract}
 
@@
 
-\bibliographystyle{abbrv}
+%\bibliographystyle{abbrv}
+\bibliographystyle{plain}
 \bibliography{foropencl}

### 2011-12-08 papers/IJCSE/rose-transform.tex
edits

@@
 
-Programmers using ForOpenCL write their code in standard Fortran, and
+Programmers using ForOpenCL write kernel code in standard Fortran, and
 a sequence of code transformations are applied to generate equivalent
@@
 of the ROSE Sage AST that represents the Fortran code.  The visitor
-used in the traversal handles five types of Sage AST Nodes: {\tt
+used in the traversal handles four types of Sage AST Nodes: {\tt
   SgAllocateStatment}, {\tt SgFunctionDeclaration}, {\tt
-  SgVariableDeclaration}, {\tt SgFunctionCallExp}, and {\tt
-  SgExprStatement}.  These encompass the primary activities that are
-commonly encountered in computational kernels: allocation, function
-invocation, variable and function declaration, and expressions that
-correspond to arithmetic operations.
+  SgVariableDeclaration}, and {\tt SgExprStatement}.  These encompass
+the primary activities that are commonly encountered in computational
+kernels: allocation, variable and function declaration, and
+expressions that correspond to arithmetic operations.  Allocate
+statements are mapped to local tile variables in the kernel.  The
+function declarations for the pure elemental Fortran functions are
+mapped to OpenCL kernel function declarations.  Variable declarations and
+expressions based on primitive (non-record) numerical types are mapped
+to their direct equivalent on the C side.
+
+Our current prototype contains a visitor for {\tt SgFunctionCallExp}
+nodes, but no conversion is performed.  We plan to provide support to
+map numerical intrinsic functions to the equivalent math primitives
+provided by OpenCL.  In many cases, this is a simple one to one
+mapping from Fortran intrinsic to OpenCL primitive.  The more complex
+case is invocation of user-defined functions from within kernel, which
+will require additional analysis by the ForOpenCL conversion tool to
+perform inlining of the called code into the kernel when appropriate.
 
@@
 prototyle includes a limited set of helper functions sufficient to demonstrate
-the shallow water model code, but could conceivably be expanded to include
+the 2D shallow water model code, but could conceivably be expanded to include
 higher dimensional arrays with little effort.

### 2011-12-08 papers/IJCSE/intro.tex
edits
fixed (?) bibliography

--- 2011-08-30 papers/GPUScA-2011/intro.tex
@@
 processor while the performance of the M2090 NVIDIA Tesla processor is in the
-same neighborhood~\cite{hpcwire11manycore}.  Unfortunately the performance that
+same neighborhood~\citep{hpcwire11manycore}.  Unfortunately the performance that
 many of the new accelerator architectures offer comes at a cost.  Architectural
@@
 A directive-based approach, such as OpenMP or the Accelerator programming model
-from the Portland Group~\cite{pgi10accelerator}, is one solution to this
+from the Portland Group~\citep{pgi10accelerator}, is one solution to this
 problem.  However, in this paper we take a somewhat different approach.  A
@@
 The problem with OpenCL and CUDA is that they expose too much detail about the
-machine architecture to the programmer~\cite{wolfe08gpgpu}.  The programmer is
+machine architecture to the programmer~\citep{wolfe08gpgpu}.  The programmer is
 responsible for explicitly managing memory (including the staging of data back
@@
 to every new architecture and programming model that emerges.
-
-\newpage
 

### 2011-12-08 papers/IJCSE/programming-model.tex
edits
fixed (?) bibliography

@@
 the array syntax adopted by Fortran in the 1990 standard can be found
-in the APL language~\cite{iverson79apl}.  These additions to Fortran
+in the APL language~\citep{iverson79apl}.  These additions to Fortran
 allowed parallelism to be expressed with whole-array operations at the
@@
 compilers to emit fast parallel code for distributed and shared memory
-parallel computers\cite{koelbel94hpf}.  One of the notable additions
+parallel computers\citep{koelbel94hpf}.  One of the notable additions
 to the language in HPF was syntax to specify the distribution of data
@@
 wasn't easily mapped onto a pure data-parallel model.  The {\tt SHADOW}
-directive introduced in HPF-2, and the {\tt HALO} in HPF+~\cite{benkner99hpf}
+directive introduced in HPF-2, and the {\tt HALO} in HPF+~\citep{benkner99hpf}
 bear some similarity to the halo region concept that we discuss in this 
@@
 \noindent
-The semantics of the shift intrinsic and operators like {\tt +} applied to the whole array make it
-unnecessary to manually implement the double buffering scheme found in the loop above.
-Similar whole array shifting was used in higher dimensions for finite difference
-codes within the computational physics community for codes targeting the CM-5
-system.  Research in compilation of stencil-based codes that use shift operators
-targeting these systems is related to the work presented
-here~\cite{stencil-compiler}.
+The semantics of the shift intrinsic and operators like {\tt +}
+applied to the whole array make it unnecessary to manually implement
+the double buffering scheme found in the loop above.  Similar whole
+array shifting was used in higher dimensions for finite difference
+codes within the computational physics community for codes targeting
+the CM-5 system.  Research in compilation of stencil-based codes that
+use shift operators targeting these systems is related to the work
+presented here~\citep{stencil-compiler}.
 
-The whole-array model was attractive because it deferred responsibility for
-optimally implementing the computations to the compiler.  Instead of relying on
-a compiler to infer parallelism from a set of explicit loops, the choice for how
-to implement loops was left entirely up to the tool.
+The whole-array model was attractive because it deferred
+responsibility for optimally implementing the computations to the
+compiler.  Instead of relying on a compiler to infer parallelism from
+a set of explicit loops, the choice for how to implement loops was
+left entirely up to the tool.
 
@@
 these semantics.  Limited attention to optimizing memory usage due to
-this intermediate storage by compilers has led to these constructs seeing
-little adoption by programmers.
+this intermediate storage by compilers has led to these constructs
+seeing little adoption by programmers.
 
@@
 associated with the ZPL language at the University of
-Washington~\cite{chamberlain04zpl} and Chapel, an HPCS language developed by
-Cray~\cite{chamberlainchapel}.  In ZPL, programmers adopt a similar global view
+Washington~\citep{chamberlain04zpl} and Chapel, an HPCS language developed by
+Cray~\citep{chamberlainchapel}.  In ZPL, programmers adopt a similar global view
 of computation over arrays, but define their computations based on regions,
@@
 whole-array operations in the Haskell language in the REPA (regular, shape-polymorphic,
-parallel array) library~\cite{keller10repa}.
+parallel array) library~\citep{keller10repa}.
 
@@
 
-Region functions are similar to the shift operator as they can be used to
-reference portions of the array that are shifted with respect to the interior
-portion.  However, unlike the shift operator, regions are not expressed in terms
-of boundary conditions and thus don't explicitly \emph{require} a knowledge of,
-nor the application of, boundary conditions locally (global boundary conditions must
-be explicitly provided by the programmer outside of calls to kernel procedures).
-Thus, as will be shown
-below, regions are more suitable for usage by OpenCL thread groups which access
-only local subsections of an array stored in global memory.
+Region functions are similar to the shift operator as they can be used
+to reference portions of the array that are shifted with respect to
+the interior portion.  However, unlike the shift operator, regions are
+not expressed in terms of boundary conditions and thus don't
+explicitly \emph{require} a knowledge of, nor the application of,
+boundary conditions locally (global boundary conditions must be
+explicitly provided by the programmer outside of calls to kernel
+procedures).  Thus, as will be shown below, regions are more suitable
+for usage by OpenCL thread groups which access only local subsections
+of an array stored in global memory.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.5in]{halo.pdf}
+\caption{An illustration of three simple halo regions around a single central
+index.}
+\label{fig:halos}
+\end{figure}
 
@@
 dimension.  For example {\tt halo = [left, right, down, up]} specifies a halo
-for a two-dimensional region.  These functions are:
+for a two-dimensional region.  Example halo regions are illustrated in
+Figure~\ref{fig:halos}.  These functions are:
 
@@
 Data parallelism has been called collection-oriented programming by
-Blelloch~\cite{blelloch90}.  As the {\tt cshift} function and the array-valued
+Blelloch~\citep{blelloch90}.  As the {\tt cshift} function and the array-valued
 expressions all semantically return a value, this style of programming is also
@@
 recently, some are actively counseling against its usage because of performance
-issues~\cite{Levesque:SC08}).  Thus the Fortran community was caught in a
+issues~\citep{Levesque:SC08}).  Thus the Fortran community was caught in a
 classic ``chicken-and-egg'' conundrum: (1) programmers didn't use it because it

### 2011-12-08 papers/IJCSE/shallow-water.tex
fixed (?) bibliography

@@
 
-The numerical code used for this work is from a presentation at the
-NM Supercomputing Challenge~\cite{Robey07}.
-The algorithm solves the standard 2D shallow water equations. This
-algorithm is typical of a wide range of modeling equations based on
-conservation laws such as compressible fluid dynamics (CFD), elastic
-material waves, acoustics, electromagnetic waves and even traffic
-flow~\cite{Leveque02}. For the shallow water problem there are
-three equations with one based on conservation of mass and the other
-two on conservation of momentum.
+The numerical code used for this work is from a presentation at the NM
+Supercomputing Challenge~\citep{Robey07}.  The algorithm solves the
+standard 2D shallow water equations. This algorithm is typical of a
+wide range of modeling equations based on conservation laws such as
+compressible fluid dynamics (CFD), elastic material waves, acoustics,
+electromagnetic waves and even traffic flow~\citep{Leveque02}. For the
+shallow water problem there are three equations with one based on
+conservation of mass and the other two on conservation of momentum.
 

### 2011-12-09 papers/IJCSE/programming-model.tex
edits

@@
 
-Borrowing ideas from ZPL, we introduce a concept of a region to Fortran with a
-set of functions that allow programmers to work with subarrays in expressions.
-In Fortran, these functions return a copy of or a pointer to an existing array or array section.
-This is unlike ZPL, where regions are analogous to index sets and are used
-primarily for address resolution within an array without dictating storage
-related behavior.  The functions that we propose are similar in that they allow
-a programmer to deal with index regions that are meaningful to their algorithm,
-and automatically induce a halo (or ghost) cell pattern as needed in the
-implementation generated by the compiler, where the size of an array is
-implicitly increased to provide extra array elements surrounding the interior 
-portion of the array.  It is important to note, however, that all memory allocated
-by the programmer must explicitly contain the extra array elements in the halo.
-%%%This is critical to reducing the amount of code that the programmer is forced to write, as the halo cells are often where boundary errors and off-by-one errors may occur in code that is manually generated.
+Borrowing ideas from ZPL, we introduce a concept of a region to
+Fortran with a set of functions that allow programmers to work with
+subarrays in expressions.  In Fortran, these functions return a copy
+of or a pointer to an existing array or array section.  This is unlike
+ZPL, where regions are analogous to index sets and are used primarily
+for address resolution within an array without dictating storage
+related behavior.  The functions that we propose are similar in that
+they allow a programmer to deal with index regions that are meaningful
+to their algorithm, and automatically induce a halo (or ghost) cell
+pattern as needed in the implementation generated by the compiler,
+where the size of an array is implicitly increased to provide extra
+array elements surrounding the interior portion of the array.  The
+goal is to reduce the amount of code that the programmer is forced to
+write, as the halo cells are often where boundary errors and
+off-by-one errors may occur in code that is manually generated.  It is
+important to note, however, that all memory allocated by the
+programmer must explicitly contain the extra array elements in the
+halo.
 
@@
 Data parallelism has been called collection-oriented programming by
-Blelloch~\citep{blelloch90}.  As the {\tt cshift} function and the array-valued
-expressions all semantically return a value, this style of programming is also
-similar to functional programming (or value-oriented programming).  It should be
-noted that the sub-setting language we employ goes beyond pure data parallelism
-by the use of pure (other than calls to {\tt region\_ptr}) subroutines and
-not just elemental functions.
+Blelloch~\citep{blelloch90}.  As the {\tt cshift} function and the
+array-valued expressions all semantically return a value, this style
+of programming is also similar to functional programming (or
+value-oriented programming).  It should be noted that the sub-setting
+language we employ goes beyond pure data parallelism by the use of
+pure (other than calls to {\tt region\_ptr}) subroutines and not just
+elemental functions.
 
-Unfortunately, this style of programming has never really caught on because when
-array syntax was first introduced in Fortran, performance of codes using these features was
-relatively poor and thus programmers shied away from using array syntax (even
-recently, some are actively counseling against its usage because of performance
-issues~\citep{Levesque:SC08}).  Thus the Fortran community was caught in a
-classic ``chicken-and-egg'' conundrum: (1) programmers didn't use it because it
-was slow; and (2) compilers vendors didn't improve it because programmers didn't
-use it.  A goal of this paper is to demonstrate that parallel programs written
-in this style of Fortran can achieve good performance on accelerator architectures.
+Unfortunately, this style of programming has never really caught on
+because when array syntax was first introduced in Fortran, performance
+of codes using these features was relatively poor and thus programmers
+shied away from using array syntax (even recently, some are actively
+counseling against its usage because of performance
+issues~\citep{Levesque:SC08}).  Thus the Fortran community was caught
+in a classic ``chicken-and-egg'' conundrum: (1) programmers didn't use
+it because it was slow; and (2) compilers vendors didn't improve it
+because programmers didn't use it.  A goal of this paper is to
+demonstrate that parallel programs written in this style of Fortran
+can achieve good performance on accelerator architectures.
 
@@
 
-Only pure Fortran procedures are transformed into OpenCL kernels.  This restriction
-is lifted slightly to allow calls to {\tt region\_ptr} from within a kernel procedure.  The
-programmer must explicitly call these kernels using Fortran interfaces in the ForOpenCL
-library (described below).  It is also possible, using ROSE, to modify the calling
-site so that the entire program can be transformed, but this functionality is
-outside the scope of this paper.  Here we specifically examine transforming
-Fortran procedures to OpenCL kernels.  Because OpenCL support is relatively new
-to ROSE, some generated code must be modified.  For example, the {\tt
+Only pure Fortran procedures are transformed into OpenCL kernels.
+This restriction is lifted slightly to allow calls to {\tt
+  region\_ptr} from within a kernel procedure.  The programmer must
+explicitly call these kernels using Fortran interfaces in the
+ForOpenCL library (described below).  It is also possible, using ROSE,
+to modify the calling site so that the entire program can be
+transformed, but this functionality is outside the scope of this
+paper.  Here we specifically examine transforming Fortran procedures
+to OpenCL kernels.  Because OpenCL support is relatively new to ROSE,
+some generated code must be modified.  For example, the {\tt
   \_\_global} attribute for kernel arguments was added by hand.
 
-It is assumed that memory for all arrays reside on the device.  The programmer
-must copy memory to and from the device.  In addition, array size (neglecting
-ghost cell regions) must be multiples of the global OpenCL kernel size.
+It is assumed that memory for all arrays reside on the device.  The
+programmer must copy memory to and from the device.  In addition,
+array size (neglecting ghost cell regions) must be multiples of the
+global OpenCL kernel size.
 
-Array variables within a kernel procedure (specified by the {\tt \!\$OFP PURE,
-  KERNEL} directive) must be declared as contiguous.  A kernel procedure may not
-call other procedures except for limited intrinsic functions (primarily math),
-user-defined elemental functions, and the {\tt region\_cpy} and {\tt region\_ptr}
-functions.  Future work will address non-contiguous arrays (such as those that
-result from strided access) by mapping array strides to gather/scatter-style
-memory accessors.
+Array variables within a kernel procedure (specified by the {\tt
+  \!\$OFP PURE, KERNEL} directive) must be declared as contiguous.  A
+kernel procedure may not call other procedures except for limited
+intrinsic functions (primarily math), user-defined elemental
+functions, and the {\tt region\_cpy} and {\tt region\_ptr} functions.
+Future work will address non-contiguous arrays (such as those that
+result from strided access) by mapping array strides to
+gather/scatter-style memory accessors.
 
-Array parameters to a kernel procedure must be declared as either intent(in) or
-intent(out); they cannot be intent(inout).  A thread may read from an extended
-region about its local element (using the {\tt region\_cpy} function), but can
-only write to the single array element it owns.  If a variable were intent(inout),
-a thread could update its array element before another thread had read from that
+Array parameters to a kernel procedure must be declared as either
+intent(in) or intent(out); they cannot be intent(inout).  A thread may
+read from an extended region about its local element (using the {\tt
+  region\_cpy} function), but can only write to the single array
+element it owns.  If a variable were intent(inout), a thread could
+update its array element before another thread had read from that
 element.  This restriction requires double buffering techniques.

### 2011-12-09 papers/IJCSE/intro.tex
edits to replace footnotes with cites

@@
 kernels.  We use the ROSE compiler
-infrastructure\footnote{\url{http://www.rosecompiler.org/}} to develop these
+infrastructure~\citep{rose-web} to develop these
 transformations.  ROSE uses the Open Fortran
-Parser\footnote{\url{http://fortran-parser.sf.net/}} to parse Fortran 2008
+Parser~\citep{ofp:sf} to parse Fortran 2008
 syntax and can generate C-based OpenCL.  Since ROSE's intermediate

### 2011-12-09 papers/IJCSE/foropencl.tex
edits
fixed (?) bibliography
edits to anonymize and put abstract in correct place

@@
 
-%\usepackage{natbib, stfloats}
+\usepackage{natbib, stfloats}
 \usepackage{cite}
@@
 
-\LRH{M. Sottile et~al.}
+%% ANONYMIZED
+%\LRH{M. Sottile et~al.}
+\LRH{ }
 
@@
 
+%% ANONYMIZED
+\authorA{Author names removed for review}
 % 1st. author
-\authorA{Matthew J. Sottile}
-\affA{Galois, Inc.\\
+%\authorA{Matthew J. Sottile}
+%\affA{Galois, Inc.\\
 % Portland, OR 97204\\
- E-mail: mjsottile@computer.org}
+% E-mail: mjsottile@computer.org}
 
 % 2nd. author
-\authorB{Craig E Rasmussen}
-\affB{Los Alamos National Laboratory\\
+%\authorB{Craig E Rasmussen}
+%\affB{Los Alamos National Laboratory\\
 % Los Alamos, NM 87545\\
- E-mail: crasmussen@lanl.gov}
+% E-mail: crasmussen@lanl.gov}
 
 % 3rd. author
-\authorC{Wayne N. Weseloh}
-\affC{Los Alamos National Laboratory\\
+%\authorC{Wayne N. Weseloh}
+%\affC{Los Alamos National Laboratory\\
 % Los Alamos, NM 87545\\
- E-mail: weseloh@lanl.gov}
+% E-mail: weseloh@lanl.gov}
 
 % 4th. author
-\authorD{Robert W. Robey}
-\affD{Los Alamos National Laboratory\\
+%\authorD{Robert W. Robey}
+%\affD{Los Alamos National Laboratory\\
 % Los Alamos, NM 87545\\
- E-mail: brobey@lanl.gov}
+% E-mail: brobey@lanl.gov}
 
 % 5th. author
-\authorE{Daniel Quinlan}
-\affE{Lawrence Livermore National Laboratory\\
+%\authorE{Daniel Quinlan}
+%\affE{Lawrence Livermore National Laboratory\\
 % Livermore, CA 94550\\
- E-Mail: dquinlan@llnl.gov}
+% E-Mail: dquinlan@llnl.gov}
 
 % 6th. author
-\authorF{Jeffrey Overbey}
-\affF{University of Illinois at Urbanna-Champaign\\
+%\authorF{Jeffrey Overbey}
+%\affF{National Center for Supercomputing Applications\\
 % Champaign, IL 61820\\
- E-mail: overbey2@illinois.edu}
+% E-mail: overbey2@illinois.edu}
+
+\input{abstract}
 
 %% bio
-\begin{bio}
-M. Sottile completed his PhD in 2006 from the University of New Mexico
-from the Department of Electrical and Computer Engineering.  He is
-presently working as a researcher at Galois, Inc. in Portland, Oregon
-leading research work efforts in scientific and high performance computing.
-His research interests include programming languages, compilation techniques,
-computational physics and applied mathematics.
+%\begin{bio}
+%M. Sottile completed his PhD in 2006 from the University of New Mexico
+%from the Department of Electrical and Computer Engineering.  He is
+%presently working as a researcher at Galois, Inc. in Portland, Oregon
+%leading research work efforts in scientific and high performance computing.
+%His research interests include programming languages, compilation techniques,
+%computational physics and applied mathematics.
 
-\noindent Craig has a bio.
+%\noindent Craig has a bio.
 
-\noindent Wayne has a bio.
+%\noindent Wayne has a bio.
 
-\noindent Bob has a bio.
+%\noindent Bob has a bio.
 
-\noindent Dan has a bio.
+%\noindent Dan has a bio.
 
-\noindent Jeff has a bio.
-\end{bio}
+%\noindent Jeff has a bio.
+%\end{bio}
 
 \maketitle
-
-\input{abstract}
 
@@
 %\bibliographystyle{abbrv}
-\bibliographystyle{plain}
+\newcommand{\newblock}{}
+\bibliographystyle{plainnat}
 \bibliography{foropencl}

### 2011-12-11 papers/IJCSE/intro.tex
edits to intro

@@
 
-This paper presents a compiler-level approach for targeting a single program to
-multiple, and possibly fundamentally different, processor architectures.  This
-technique allows the application programmer to adopt a single, high-level
-programming model without sacrificing performance.  We suggest that existing
-data-parallel features in Fortran are well-suited to applying automatic
-transformations that generate code specifically tuned for different hardware
-architectures using low-level programming models such as OpenCL.  For algorithms
-that can be easily expressed in terms of whole array, data-parallel operations,
-writing code in Fortran and transforming it automatically to specific low-level
-implementations removes the burden of creating and maintaining multiple versions
-of architecture specific code.
+This paper presents a compiler-level approach for targeting a single
+program to multiple, and possibly fundamentally different, processor
+architectures.  This technique allows the application programmer to
+adopt a single, high-level programming model without sacrificing
+performance.  We suggest that existing data-parallel features in
+Fortran are well-suited to applying automatic transformations that
+generate code specifically tuned for different hardware architectures
+using low-level programming models such as OpenCL.  For algorithms
+that can be easily expressed in terms of whole array, data-parallel
+operations, writing code in Fortran and transforming it automatically
+to specific low-level implementations removes the burden of creating
+and maintaining multiple versions of architecture specific code.
 
-The peak performance of these newer accelerator architectures can be substantial.
-Intel expects a teraflop for the SGEMM benchmark with their Knights Ferry
-processor while the performance of the M2090 NVIDIA Tesla processor is in the
-same neighborhood~\citep{hpcwire11manycore}.  Unfortunately the performance that
-many of the new accelerator architectures offer comes at a cost.  Architectural
-changes are trending toward multiple heterogeneous cores and less of a reliance
-on superscalar instruction level parallelism and hardware managed memory
-hierarchies (such as traditional caches).
+The peak performance of these newer accelerator architectures can be
+substantial.  Intel expects a teraflop for the SGEMM benchmark with
+their Knights Ferry processor while the performance of the M2090
+NVIDIA Tesla processor is in the same
+neighborhood~\citep{hpcwire11manycore}.  Unfortunately the performance
+that many of the new accelerator architectures offer comes at a cost.
+Architectural changes are trending toward multiple heterogeneous cores
+and less of a reliance on superscalar instruction level parallelism
+and hardware managed memory hierarchies (such as traditional caches).
 
-These changes place a heavy burden on
-application programmers as they work to adapt to these new systems.  An
-especially challenging problem is not only how to program to these new
-architectures --- considering the massive scale of concurrency available --- but
-also how to design programs that are portable across the changing landscape of
-computer architectures.  How does a programmer write one program that can
-perform well on both a conventional multicore CPU \emph{and} a GPU (or any other emerging many-core architectures)?
+These changes place a heavy burden on application programmers as they
+work to adapt to these new systems.  An especially challenging problem
+is not only how to program to these new architectures --- considering
+the massive scale of concurrency available --- but also how to design
+programs that are portable across the changing landscape of computer
+architectures.  How does a programmer write one program that can
+perform well on both a conventional multicore CPU \emph{and} a GPU (or
+any other emerging many-core architectures)?
 
-A directive-based approach, such as OpenMP or the Accelerator programming model
-from the Portland Group~\citep{pgi10accelerator}, is one solution to this
-problem.  However, in this paper we take a somewhat different approach.  A
-common theme amongst the new processors is the emphasis on data-parallel
-programming.  This model is well-suited to architectures that are
-based on either vector processing or massively parallel collections of simple
-cores.  The recent CUDA and OpenCL programming languages are intended to
-support this programming model.
+A directive-based approach, such as OpenMP or the Accelerator
+programming model from the Portland Group~\citep{pgi10accelerator} and
+now generalized in the OpenACC standard~\citep{openacc11std}, is one
+solution to this problem.  However, in this paper we take a somewhat
+different approach not requiring explicit declaration of parallelism,
+instead relying on implicitly parallel subsets of the Fortran
+language.  A common theme amongst the new processors is the emphasis
+on data-parallel programming.  This model is well-suited to
+architectures that are based on either vector processing or massively
+parallel collections of simple cores.  The recent CUDA and OpenCL
+programming languages are intended to support this programming model.
 
-The problem with OpenCL and CUDA is that they expose too much detail about the
-machine architecture to the programmer~\citep{wolfe08gpgpu}.  The programmer is
-responsible for explicitly managing memory (including the staging of data back
-and forth between the host CPU and the accelerator device) and specifically
-taking into account architectural differences (such as whether the architecture
-contains vector units).  While these languages have been attractive as a method
-for early adopters to utilize these new architectures, they are less attractive
-to programmers who do not have the time or resources to manually port their code
-to every new architecture and programming model that emerges.
+The problem with OpenCL and CUDA is that they expose too much detail
+about the machine architecture to the programmer~\citep{wolfe08gpgpu}.
+The programmer is responsible for explicitly managing memory
+(including the staging of data back and forth between the host CPU and
+the accelerator device) and specifically taking into account
+architectural differences (such as whether the architecture contains
+vector units).  While these languages have been attractive as a method
+for early adopters to utilize these new architectures, they are less
+attractive to programmers who do not have the time or resources to
+manually port their code to every new architecture and programming
+model that emerges.
 
@@
 
-We demonstrate that a subset of Fortran map surprisingly well onto GPUs when
-transformed to OpenCL kernels.  This data-parallel subset includes: array syntax
-using assignment statements and binary operators, array constructs like {\tt
-  WHERE}, and the use of pure and elemental functions.  In addition, we provide
-new functions that explicitly take advantage of the stencil geometry of the
-problem domain we consider.  Note that this subset of the Fortran language is
-implicitly parallel.  This programming model \emph{does not require explicit
-  declaration of parallelism within the program.}  In addition, programs are
-expressed using entirely standard Fortran so it can be compiled for and executed
-on a single core without concurrency.
+We demonstrate that a subset of Fortran maps surprisingly well onto
+GPUs when transformed to OpenCL kernels.  This data-parallel subset
+includes: array syntax using assignment statements and binary
+operators, array constructs like {\tt WHERE}, and the use of pure and
+elemental functions.  In addition, we provide new functions that
+explicitly take advantage of the stencil geometry of the problem
+domain we consider.  Note that this subset of the Fortran language is
+implicitly parallel.  This programming model \emph{does not require
+  explicit declaration of parallelism within the program.}  In
+addition, programs are expressed using entirely standard Fortran so it
+can be compiled for and executed on a single core without concurrency.
 
-Transformations are supplied that provide a mechanism for converting Fortran
-procedures written in the Fortran subset described in this paper to OpenCL
-kernels.  We use the ROSE compiler
-infrastructure~\citep{rose-web} to develop these
-transformations.  ROSE uses the Open Fortran
-Parser~\citep{ofp:sf} to parse Fortran 2008
+Transformations are supplied that provide a mechanism for converting
+Fortran procedures written in the Fortran subset described in this
+paper to OpenCL kernels.  We use the ROSE compiler
+infrastructure~\citep{rose-web} to develop these transformations.
+ROSE uses the Open Fortran Parser~\citep{ofp:sf} to parse Fortran 2008
 syntax and can generate C-based OpenCL.  Since ROSE's intermediate
-representation (IR) was constructed to represent multiple languages, it is
-relatively straightforward to transform high-level Fortran IR nodes to C OpenCL
-nodes.  This work is also applicable to transformations to vendor-specific
-languages, similar to OpenCL, such as the NVIDIA CUDA language.
+representation (IR) was constructed to represent multiple languages,
+it is relatively straightforward to transform high-level Fortran IR
+nodes to C OpenCL nodes.  This work is also applicable to
+transformations to vendor-specific languages, similar to OpenCL, such
+as the NVIDIA CUDA language.
 
 Transformations for arbitrary Fortran procedures are not attempted.
-Furthermore, a mechanism to transform the calling site to automatically invoke
-OpenCL kernels is not provided at this time.  While it is possible to accomplish
-this task within ROSE, it is considered outside the scope of this paper.
-However, ForOpenCL provides via Fortran interfaces a mechanism to call the C OpenCL
-runtime and enable Fortran programmers to access OpenCL kernels generated by the
-supplied transformations.
+Furthermore, a mechanism to transform the calling site to
+automatically invoke OpenCL kernels is not provided at this time.
+While it is possible to accomplish this task within ROSE, it is
+considered outside the scope of this paper.  However, ForOpenCL
+provides via Fortran interfaces a mechanism to call the C OpenCL
+runtime and enable Fortran programmers to access OpenCL kernels
+generated by the supplied transformations.
 
-We study the automatic transformations for an application example that is
-typical of stencil codes that update array elements according to a fixed
-pattern.  Stencil codes are often employed in applications based on
-finite-difference or finite-volume methods in computational fluid dynamics
-(CFD).  The example described later in this paper is a simple shallow-water
-model in two dimensions using finite volume methods.  Stencil-like patterns
-appear in a number of other contexts as well.  In image processing, they appear
-in convolution-based algorithms in which small kernels are convolved with an image
-to implement denoising, edge detection, and other common operators.  Similar stencil
-operators appear in general signal processing applications as well.  
+We study the automatic transformations for an application example that
+is typical of stencil codes that update array elements according to a
+fixed pattern.  Stencil codes are often employed in applications based
+on finite-difference or finite-volume methods in computational fluid
+dynamics (CFD).  The example described later in this paper is a simple
+shallow-water model in two dimensions using finite volume methods.
+Stencil-like patterns appear in a number of other contexts as well.
+In image processing, they appear in convolution-based algorithms in
+which small kernels are convolved with an image to implement
+denoising, edge detection, and other common operators.  Similar
+stencil operators appear in general signal processing applications as
+well.
 %% more?

### 2011-12-11 papers/IJCSE/foropencl.tex
more edits from bob

@@
 of the hardware) and the need for portability across architectures
-make a higher-level and simpler programming model like data-parallel
-attractive.
+make a higher-level and simpler programming model like the data-parallel
+model attractive.
 

### 2011-12-11 papers/IJCSE/programming-model.tex
more edits from bob

@@
 
-It should be noted that the function {\tt region\_cpy} is pure and thus can be
-called from within a pure kernel procedure, and {\tt
-region\_ptr} is impure because it aliases the array parameter.  However as will
-be shown below, the usage of {\tt region\_ptr} is constrained so that it does not
-introduce side effects in the functions that call it.  These two functions are
-part of the language recognized by the compiler and though {\tt region\_cpy}
-returns a copy of a portion of an array \emph{semantically}, the compiler is not
-forced to actually make a copy and is free to enforce copy semantics through
-other means.  In addition to these two new functions, ForOpenCL provides the
-compiler directive, {\tt \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine
-can be transformed to an OpenCL kernel and that the subroutine is pure
-except for calls to {\tt region\_ptr}.  These directives are not strictly
-necessary for the technique described in this paper, but aid in automated
-identification of specific kernels to be transformed to OpenCL.  A directive-free
-implementation would require the transformation tool be provided the set of 
-kernels to work via a user defined list.
+It should be noted that the function {\tt region\_cpy} is pure and
+thus can be called from within a pure kernel procedure, and {\tt
+  region\_ptr} is impure because it aliases the array parameter.
+However as will be shown below, the usage of {\tt region\_ptr} is
+constrained so that it does not introduce side effects in the
+functions that call it.  These two functions are part of the language
+recognized by the compiler and though {\tt region\_cpy} returns a copy
+of a portion of an array \emph{semantically}, the compiler is not
+forced to actually make a copy and is free to enforce copy semantics
+through other means within the pure constraint.  In addition to these
+two new functions, ForOpenCL provides the compiler directive, {\tt
+  \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine can be
+transformed to an OpenCL kernel and that the subroutine is pure except
+for calls to {\tt region\_ptr}.  These directives are not strictly
+necessary for the technique described in this paper, but aid in
+automated identification of specific kernels to be transformed to
+OpenCL.  A directive-free implementation would require the
+transformation tool be provided the set of kernels to work via a user
+defined list.
 

### 2011-12-11 papers/IJCSE/shallow-water.tex
more edits from bob

@@
 can be ignored. A fixed time step is used for simplicity though it
-must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$ to fulfill the CFL
-condition.
+must be less than
+
+\begin{eqnarray*}
+dt \leqq \frac{dx}{\left(\left(\sqrt{gh}+|u|\right) + \left(\sqrt{gh}+|v|\right)\right)}
+\end{eqnarray*}
+
+\noindent to fulfill the CFL condition.
 
@@
 $U$ is the conserved state variable at the center of the cell. This
-state variable, $U$ $=(h,hu,hv)$ in the first term in the equations
+state variable, $U$ $=\{h,hu,hv\}$ in the first term in the equations
 below. $F$ is the flux quantity that crosses the boundary of the cell

### 2011-12-12 papers/IJCSE/response_to_reviews.txt
text to include with journal submission re: changes since workshop paper

Dear Reviewers-

This paper is an extended and updated version of the paper by the same title
that was submitted to the GPUScA 2011 workshop.  The first round of reviews
that were for the workshop resulted in us making a set of changes to add
detail in addressing the scope of work that the paper does and does not
cover.  Specifically, we state that the transformation technique descriped
in the paper is focused on the pure elementatal functions that are
turned into OpenCL kernels, and do not perform loop or other analyses that
are used by other methods to infer what loop bodies should be turned into
kernels.  We also provided updated information to reflect the most
recent Fortran 2008 standard treatment of pure and elemental functions.

The revised version of the GPUScA paper contains the following changes in
addition to those made based on the reviewer comments from the workshop:

- Added a subsection that details the ROSE-based transformation that does
  the actual Fortran to OpenCL conversion of the kernel.  This was not
  present in the workshop paper due to space constraints.  Based on
  comments during the presentation of the paper and discussions with
  readers, we believe that this makes the paper more complete.

- We added an additional figure to illustrate the halo concept to augment
  the discussion in the text.  This was again based on feedback that we
  received from readers in order to provide clarity with the additional
  space in the journal submission.

- Updated reference set to include recent activity in similar efforts,
  specifically the OpenACC work that was presented at Supercomputing 2011.

- Revision of the CFL condition presented in the shallow water model
  to address a missing term that, while unrelated to the transformation,
  is important for correctly describing the shallow water example.

- Numerous tweaks and minor text edits for clarification and presentation.

- Editing to formatting to match the LaTeX style for the journal.

Thank you for your time to review this updated version of the paper.

### 2011-12-12 papers/IJCSE/rose-transform.tex
final edits from craig

@@
 used in the traversal handles four types of Sage AST Nodes: {\tt
-  SgAllocateStatment}, {\tt SgFunctionDeclaration}, {\tt
+  SgAllocateStatement}, {\tt SgFunctionDeclaration}, {\tt
   SgVariableDeclaration}, and {\tt SgExprStatement}.  These encompass
@@
 as mapping 2D logical array indices to physical 1D addresses).  Our current
-prototyle includes a limited set of helper functions sufficient to demonstrate
+prototype includes a limited set of helper functions sufficient to demonstrate
 the 2D shallow water model code, but could conceivably be expanded to include

### 2012-01-05 papers/IJCSE/programming-model.tex
one more edit from Bob

@@
 compilers to emit fast parallel code for distributed and shared memory
-parallel computers\citep{koelbel94hpf}.  One of the notable additions
+parallel computers~\citep{koelbel94hpf}.  One of the notable additions
 to the language in HPF was syntax to specify the distribution of data
@@
 are updated as if they were updated simultaneously.  In a program
-where the programmer explicitly manages arrays and loops, double
-buffering techniques and user managed temporaries are used to maintain
-these semantics.  Limited attention to optimizing memory usage due to
-this intermediate storage by compilers has led to these constructs
-seeing little adoption by programmers.
+where the programmer explicitly manages arrays and loops (rather than
+the compiler), double buffering techniques and user managed
+temporaries are used to maintain these semantics.  Limited attention
+to optimizing memory usage due to this intermediate storage by
+compilers has led to these constructs seeing little adoption by
+programmers.
 
@@
 two new functions, ForOpenCL provides the compiler directive, {\tt
-  \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine can be
+  \!\$OFP PURE, CONCURRENT}, which specifies that a pure subroutine can be
 transformed to an OpenCL kernel and that the subroutine is pure except
@@
 array size (neglecting ghost cell regions) must be multiples of the
-global OpenCL kernel size.
+global OpenCL kernel size.  The explicit memory copy and array sizes
+are constraints of this current work and not of the possible future
+enhancements.
 
 Array variables within a kernel procedure (specified by the {\tt
-  \!\$OFP PURE, KERNEL} directive) must be declared as contiguous.  A
+  \!\$OFP PURE, CONCURRENT} directive) must be declared as contiguous.  A
 kernel procedure may not call other procedures except for limited

### 2012-01-05 papers/IJCSE/shallow-water.tex

@@
 half-step advanced in time and space on each face, using loops on
-the faces.\begin{eqnarray*}
+the faces.
+
+\begin{eqnarray*}
 U_{i+\frac{1}{2},j}^{n+\frac{1}{2}} & = & (U_{i+1,j}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle x}\left(F_{i+1,j}^{n}-F_{i,j}^{n}\right)\\
-U_{i,j+\frac{1}{2}}^{n+\frac{1}{2}} & = & (U_{i,j+1}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle y}\left(F_{i,j+1}^{n}-F_{i,j}^{n}\right)\end{eqnarray*}
+U_{i,j+\frac{1}{2}}^{n+\frac{1}{2}} & = & (U_{i,j+1}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle y}\left(F_{i,j+1}^{n}-F_{i,j}^{n}\right)
+\end{eqnarray*}
 
 The second step uses the estimated values from step 1 to compute the
-values at the next time step in a dimensionally unsplit loop.\[
-U_{i,j}^{n+1}=U_{i,j}^{n}-\frac{\triangle t}{\triangle x}(F_{i+\frac{1}{2},j}^{n+\frac{1}{2}}-F_{i-\frac{1}{2},j}^{n+\frac{1}{2}})-\frac{\triangle t}{\triangle y}(F_{i,j+\frac{1}{2}}^{n+\frac{1}{2}}-F_{i,j-\frac{1}{2}}^{n+\frac{1}{2}})\]
+values at the next time step in a dimensionally unsplit loop.
+
+\begin{eqnarray*}
+U_{i,j}^{n+1} = U_{i,j}^{n} &-&\frac{\triangle t}{\triangle
+  x}(F_{i+\frac{1}{2},j}^{n+\frac{1}{2}}-F_{i-\frac{1}{2},j}^{n+\frac{1}{2}})\\
+&-& \frac{\triangle
+  t}{\triangle
+  y}(F_{i,j+\frac{1}{2}}^{n+\frac{1}{2}}-F_{i,j-\frac{1}{2}}^{n+\frac{1}{2}})
+\end{eqnarray*}
 

### 2012-01-05 papers/IJCSE/source-to-source.tex
edits

@@
 transformations that take Fortran elemental and pure procedures as
-input and generate OpenCL code.  Elemental functions are
-transformed to inline OpenCL functions and subroutines with the
-{\tt PURE} and {\tt KERNEL} compiler directive attributes are
-transformed to OpenCL kernels.
+input and generate OpenCL code.  Elemental functions are transformed
+to inline OpenCL functions and subroutines with the {\tt PURE} and
+{\tt KERNEL} compiler directive attributes are transformed to OpenCL
+kernels.
 
@@
 
-OpenCL~\cite{opencl08} is an open-language standard for developing applications
-targeted for GPUs, as well as for multi-threaded applications targeted for
-multi-core CPUs.  The kernels are run by calling a C runtime library from the
-OpenCL host (normally the CPU).  Efforts to standardize a C++ runtime are
-underway and Fortran interfaces to the C runtime are distributed in the ForOpenCL
-library.
+OpenCL~\citep{opencl08} is an open-language standard for developing
+applications targeted for GPUs, as well as for multi-threaded
+applications targeted for multi-core CPUs.  The kernels are run by
+calling a C runtime library from the OpenCL host (normally the CPU).
+Efforts to standardize a C++ runtime are underway and Fortran
+interfaces to the C runtime are distributed in the ForOpenCL library.
 
-An important concept in OpenCL is that of a thread and a thread group.  Thread
-groups are used to run an OpenCL kernel concurrently on several processor
-elements on the OpenCL device (often a GPU).  Consider a data-parallel statement
-written in terms of an elemental function as discussed above.  The act of
-running an OpenCL kernel can be thought of as having a particular thread
-assigned to each instance of the call to the elemental function as it is mapped
-across the arrays in the data-parallel statement.  In practice, these threads
-are packaged into thread groups when they are run on the device hardware.
+An important concept in OpenCL is that of a thread and a thread group.
+Thread groups are used to run an OpenCL kernel concurrently on several
+processor elements on the OpenCL device (often a GPU).  Consider a
+data-parallel statement written in terms of an elemental function as
+discussed above.  The act of running an OpenCL kernel can be thought
+of as having a particular thread assigned to each instance of the call
+to the elemental function as it is mapped across the arrays in the
+data-parallel statement.  In practice, these threads are packaged into
+thread groups when they are run on the device hardware.
 
-Device memory is separated hierarchically.  A thread instance has access to its
-own thread memory (normally a set of registers), threads in a thread group to
-OpenCL local memory, and all thread groups have access to OpenCL global memory.
-When multiple members of a thread group access the same memory elements (for
-example in the use of the {\tt region\_cpy} function in the calculation of face
-variable quantities shown above), for performance reasons it is often best that
-global memory accessed and shared by a thread group be copied into local memory.
+Device memory is separated hierarchically.  A thread instance has
+access to its own thread memory (normally a set of registers), threads
+in a thread group to OpenCL local memory, and all thread groups have
+access to OpenCL global memory.  When multiple members of a thread
+group access the same memory elements (for example in the use of the
+{\tt region\_cpy} function in the calculation of face variable
+quantities shown above), for performance reasons it is often best that
+global memory accessed and shared by a thread group be copied into
+local memory.
 
-The \emph{region} and \emph{halo} constructs easily map onto the OpenCL memory
-hierarchy.  A schematic of this mapping is shown in Figure~\ref{fig:cl-memory}
-for a two-dimensional array with a 2x2 array of 4 thread groups.  The memory
-for the array and its halo are stored in global memory on the device as shown
-in the background layer of the figure.  The array copy in local memory is
-shown in the foreground divided into 4 \emph{local} tiles that partition the
-array.  Halo regions in global memory are shown in dark gray and halo regions
+The \emph{region} and \emph{halo} constructs easily map onto the
+OpenCL memory hierarchy.  A schematic of this mapping is shown in
+Figure~\ref{fig:cl-memory} for a two-dimensional array with a 2x2
+array of 4 thread groups.  The memory for the array and its halo are
+stored in global memory on the device as shown in the background layer
+of the figure.  The array copy in local memory is shown in the
+foreground divided into 4 \emph{local} tiles that partition the array.
+Halo regions in global memory are shown in dark gray and halo regions
 in local memory are shown in light gray.
@@
 
-We point out that the hierarchical distribution of memory used on the OpenCL
-device shown in Figure~\ref{fig:cl-memory} is similar to the distribution of
-memory across MPI nodes in an MPI application.  In the case of MPI, the
-virtual global array is represented by the background layer (with its halo) and
-partitions of the global array are stored in the 4 MPI nodes shown in the foreground.
-Our current and future work on this effort includes source-to-source transformations
-to generate MPI code in addition to OpenCL in order to deal with clusters of 
-nodes containing accelerators.  This work is outside the scope of this paper.
+We point out that the hierarchical distribution of memory used on the
+OpenCL device shown in Figure~\ref{fig:cl-memory} is similar to the
+distribution of memory across MPI nodes in an MPI application.  In the
+case of MPI, the virtual global array is represented by the background
+layer (with its halo) and partitions of the global array are stored in
+the 4 MPI nodes shown in the foreground.  Our current and future work
+on this effort includes source-to-source transformations to generate
+MPI code in addition to OpenCL in order to deal with clusters of nodes
+containing accelerators.  This work is outside the scope of this
+paper.
 
-Halo regions obtained via the {\tt region\_cpy} function (used with intent(in)
-arrays) are constrained semantically so that they can not be written to by an
-OpenCL kernel.  The {\tt region\_cpy} function returns a copy of the region of
-the array stored in global memory and places it in local memory shared by
-threads in a thread group.  Thus once memory for an array has been transferred
-into global device memory by the host (before the OpenCL kernel is run), memory
-is in a consistent state so that all kernel threads are free to read from global
-device memory.  Because the local memory is a copy, it functions as a software
-cache for the local thread group.  Thus the compiler must insert OpenCL barriers
-at proper locations in the code to insure that all threads have written to the
-local memory cache before any thread can start to read from the cache.  On exit
-from a kernel, any local memory explicitly stored in register variables by the
-compiler (memory accessed via the {\tt region\_cpy} function) is copied back to
-global memory for all intent(out) arrays.  Recall that a thread may only write to
-its own intent(out) array element, thus there are no race conditions when updating
-intent(out) arrays.
+Halo regions obtained via the {\tt region\_cpy} function (used with
+intent(in) arrays) are constrained semantically so that they can not
+be written to by an OpenCL kernel.  The {\tt region\_cpy} function
+returns a copy of the region of the array stored in global memory and
+places it in local memory shared by threads in a thread group.  Thus
+once memory for an array has been transferred into global device
+memory by the host (before the OpenCL kernel is run), memory is in a
+consistent state so that all kernel threads are free to read from
+global device memory.  Because the local memory is a copy, it
+functions as a software cache for the local thread group.  Thus the
+compiler must insert OpenCL barriers at proper locations in the code
+to insure that all threads have written to the local memory cache
+before any thread can start to read from the cache.  On exit from a
+kernel, any local memory explicitly stored in register variables by
+the compiler (memory accessed via the {\tt region\_cpy} function) is
+copied back to global memory for all intent(out) arrays.  Recall that
+a thread may only write to its own intent(out) array element, thus
+there are no race conditions when updating intent(out) arrays.
 
@@
 
-This section outlines the OpenCL equivalent syntax for portions of the Fortran
-shallow-water code described in Section~\ref{sec:shallow-water}.  The notation
-uses uppercase for arrays and lowercase for scalar quantities.  Variables
-temporarily storing quantities for updated output arrays (declared as pointers
-in Fortran) are denoted by a {\tt p} preceding the array name.  For example, the
-Fortran statement {\tt pH = region\_ptr(oH, halo)} is transformed as a scalar
-variable declaration representing a single element in the output array {\tt oH}.
-
-\vskip .4in
+This section outlines the OpenCL equivalent syntax for portions of the
+Fortran shallow-water code described in
+Section~\ref{sec:shallow-water}.  The notation uses uppercase for
+arrays and lowercase for scalar quantities.  Variables temporarily
+storing quantities for updated output arrays (declared as pointers in
+Fortran) are denoted by a {\tt p} preceding the array name.  For
+example, the Fortran statement {\tt pH = region\_ptr(oH, halo)} is
+transformed as a scalar variable declaration representing a single
+element in the output array {\tt oH}.
 
@@
 
-While the Fortran version of the {\tt region\_cpy} function semantically returns
-an array copy, in OpenCL this function returns a scalar quantity based on the
-location of a thread in a thread group and the relationship of its location to
-the array copy transferred to local memory.  Because we assume there is a thread
-for every element in the interior, the array index is just the thread index
-adjusted for the size of the halo.  Thus {\tt region\_cpy} is just an inline
-OpenCL function and is provided by the ForOpenCL library.
+While the Fortran version of the {\tt region\_cpy} function
+semantically returns an array copy, in OpenCL this function returns a
+scalar quantity based on the location of a thread in a thread group
+and the relationship of its location to the array copy transferred to
+local memory.  Because we assume there is a thread for every element
+in the interior, the array index is just the thread index adjusted for
+the size of the halo.  Thus {\tt region\_cpy} is just an inline OpenCL
+function and is provided by the ForOpenCL library.
 
@@
 OpenCL language standard.
+
+\input{rose-transform}

### 2012-03-13 papers/IJCSE/foropencl.tex
keywords
edits for final version

@@
 %% ANONYMIZED
-%\LRH{M. Sottile et~al.}
-\LRH{ }
+\LRH{M. Sottile et~al.}
+%\LRH{ }
 
@@
 %% ANONYMIZED
-\authorA{Author names removed for review}
+%\authorA{Author names removed for review}
 % 1st. author
-%\authorA{Matthew J. Sottile}
-%\affA{Galois, Inc.\\
-% Portland, OR 97204\\
-% E-mail: mjsottile@computer.org}
+\authorA{Matthew J. Sottile}
+\affA{Galois, Inc.\\
+ Portland, OR 97204\\
+ E-mail: mjsottile@computer.org}
 
 % 2nd. author
-%\authorB{Craig E Rasmussen}
-%\affB{Los Alamos National Laboratory\\
-% Los Alamos, NM 87545\\
-% E-mail: crasmussen@lanl.gov}
+\authorB{Craig E Rasmussen}
+\affB{Los Alamos National Laboratory\\
+ Los Alamos, NM 87545\\
+ E-mail: crasmussen@lanl.gov}
 
 % 3rd. author
-%\authorC{Wayne N. Weseloh}
-%\affC{Los Alamos National Laboratory\\
-% Los Alamos, NM 87545\\
-% E-mail: weseloh@lanl.gov}
+\authorC{Wayne N. Weseloh}
+\affC{Los Alamos National Laboratory\\
+ Los Alamos, NM 87545\\
+ E-mail: weseloh@lanl.gov}
 
 % 4th. author
-%\authorD{Robert W. Robey}
-%\affD{Los Alamos National Laboratory\\
-% Los Alamos, NM 87545\\
-% E-mail: brobey@lanl.gov}
+\authorD{Robert W. Robey}
+\affD{Los Alamos National Laboratory\\
+ Los Alamos, NM 87545\\
+ E-mail: brobey@lanl.gov}
 
 % 5th. author
-%\authorE{Daniel Quinlan}
-%\affE{Lawrence Livermore National Laboratory\\
-% Livermore, CA 94550\\
-% E-Mail: dquinlan@llnl.gov}
+\authorE{Daniel Quinlan}
+\affE{Lawrence Livermore National Laboratory\\
+ Livermore, CA 94550\\
+ E-Mail: dquinlan@llnl.gov}
 
 % 6th. author
-%\authorF{Jeffrey Overbey}
-%\affF{National Center for Supercomputing Applications\\
-% Champaign, IL 61820\\
-% E-mail: overbey2@illinois.edu}
+\authorF{Jeffrey Overbey}
+\affF{National Center for Supercomputing Applications\\
+ Urbana, IL 61801\\
+ E-mail: overbey2@illinois.edu}
 
 \input{abstract}
+
+\KEYWORD{Fortran; OpenCL; Accelerator programming; GPU programming; code generation.}
 
@@
 %ACKNOWLEDGMENTS are optional
+
 \section{Acknowledgments}
@@
 
+%% this was only required for review purposes - remove in final
+%\newpage
+%
+%\input{note-to-reviewers}
+
 \end{document}

### 2012-04-16 papers/IJCSE/intro.tex

@@
 While it is possible to accomplish this task within ROSE, it is
-considered outside the scope of this paper.  However, ForOpenCL
-provides via Fortran interfaces a mechanism to call the C OpenCL
-runtime and enable Fortran programmers to access OpenCL kernels
-generated by the supplied transformations.
+considered outside the scope of this paper.  ForOpenCL is a set of
+tools providing: 1. Fortran interfaces to the OpenCL runtime library;
+2. A wrapper library providing a high-level interface to these direct
+calls to C, thus simplifying calling OpenCL kernels for
+Fortran Programmers; 3. A code generation facility for transforming
+Fortran pure elemental functions to OpenCL kernels.
+%However, ForOpenCL
+%provides via Fortran interfaces a mechanism to call the C OpenCL
+%runtime and enable Fortran programmers to access OpenCL kernels
+%generated by the supplied transformations.
 

### 2012-04-16 papers/IJCSE/note-to-reviewers.tex

\section{Notes to reviewers for IJCSE submission}

This section was added based on instructions for submission to the SI from
the GPUScA program committee and will NOT appear in the final version of the
paper.

This paper is an extended and updated version of the paper by the same title
that was submitted to the GPUScA 2011 workshop.  The first round of reviews
that were for the workshop resulted in us making a set of changes to add
detail in addressing the scope of work that the paper does and does not
cover.  Specifically, we state that the transformation technique descriped
in the paper is focused on the pure elementatal functions that are
turned into OpenCL kernels, and do not perform loop or other analyses that
are used by other methods to infer what loop bodies should be turned into
kernels.  We also provided updated information to reflect the most
recent Fortran 2008 standard treatment of pure and elemental functions.

The revised version of the GPUScA paper contains the following changes in
addition to those made based on the reviewer comments from the workshop:

- Added a subsection that details the ROSE-based transformation that does
  the actual Fortran to OpenCL conversion of the kernel.  This was not
  present in the workshop paper due to space constraints.  Based on
  comments during the presentation of the paper and discussions with
  readers, we believe that this makes the paper more complete.

- We added an additional figure to illustrate the halo concept to augment
  the discussion in the text.  This was again based on feedback that we
  received from readers in order to provide clarity with the additional
  space in the journal submission.

- Updated reference set to include recent activity in similar efforts,
  specifically the OpenACC work that was presented at Supercomputing 2011.

- Revision of the CFL condition presented in the shallow water model
  to address a missing term that, while unrelated to the transformation,
  is important for correctly describing the shallow water example.

- Numerous tweaks and minor text edits for clarification and presentation.

- Editing to formatting to match the LaTeX style for the journal.

Thank you for your time to review this updated version of the paper.

### 2012-05-30 papers/LOPe/programming-model.tex
Added a paragraph on memory management.

--- 2011-04-09 opencl/paper-local/programming-model.tex
@@
 
+%% A note from a conversation with Matt regarding memory management
+%%
+\subsection{Memory Management}
+
+The Fortran language has much tighter restrictions on aliasing than does C.
+So unless a variable has the pointer or target attribute, it cannot be aliased.
+Thus the compiler is able to agressively optimized for memory movement between
+the CPU and accelerator.  However, because the design philosophy of coarrays
+is that memory transfer between images can be expensive, the programmer must
+explicitly transfer memory between images with explicit syntax with square
+bracket notation, i.e., $a[1] = a[2]$.  So we allow the compiler to manage
+memory within an image but require the use of {\tt halo_exchange} for the
+transfer of halo memory between images.
+
 \subsection{Code Restructuring Tools}

### 2015-01-22 papers/LOPe/local.tex
Starting work on EuroPar version.

@@
 
-\title{Locally-Oriented Programming: A stunningly simple programming model for GPU programming in Fortran and C++}
+\title{Locally-Oriented Programming: A simple programming model for stencil computations}
 
@@
 \alignauthor
-Daniel Quinlan\\
-       \affaddr{Lawrence Livermore National Laboratory}\\
-       \email{dquinlan@llnl.gov}
+Soren Rasmussen\\
+       \affaddr{Utah State University}\\
+       \email{soren.rasmussen@aggiemail.usu.edu}
 }
@@
 \begin{abstract}
-Emerging GPU architectures for high performance computing are well suited to a
-data-parallel programming model.  However, programming for these architectures
-currently requires learning a new language, e.g., OpenCL or CUDA, or employing
-OpenMP compiler directives.  This paper presents preliminary work examining how
-programming with a local orientation can be employed in Fortran (or using a
-special array library in C++) to provide simple access GPU architectures.  A
-locally-oriented programming model is especially useful for stencil-based codes
-or algorithms requiring the application of convolution kernels.  In this
-programming model, a programmer codes the algorithm {\it only} for a single
-array element, but has read-only access to a small sub-array surrounding the
-given array element, so that a stencil can be applied.  We demonstrate how a
-locally-oriented programming model can be adopted using source-to-source program
-transformations in conjuction with a compiler-based infrastructure like ROSE.
+Some stuff...
+
+%%Emerging GPU architectures for high performance computing are well suited to a data-parallel programming model.
+However, programming for these architectures currently requires learning a new language, e.g., OpenCL or CUDA, or employing OpenMP compiler directives.
+This paper presents preliminary work examining how programming with a local orientation can be employed in Fortran (or using a special array library in C++) to provide simple access GPU architectures.
+A locally-oriented programming model is especially useful for stencil-based codes or algorithms requiring the application of convolution kernels.
+In this programming model, a programmer codes the algorithm {\it only} for a single array element, but has read-only access to a small sub-array surrounding the given array element, so that a stencil can be applied.
+We demonstrate how a locally-oriented programming model can be adopted using source-to-source program transformations in conjuction with a compiler-based infrastructure like ROSE.
 \end{abstract}
@@
 
-\input{intro}
+%\input{intro}
+%\input{motivation}
 \input{programming-model}
-\input{shallow-water}
-\input{source-to-source}
+%\input{program-restructuring}
 %\input{static-analysis}
-\input{performance}
+%\input{performance}
 
@@
 
-The sheer complexity of programming for clusters of many or multi-core
-processors with tens of millions threads of execution makes the simplicity of
-the data-parallel model attractive.  The increasing complexity of
-today's applications (especially in light of the increasing complexity
-of the hardware) and the need for portability across architectures
-make a higher-level and simpler programming model like data-parallel
-attractive.
+Some stuff...
 
-The goal of this work has been to exploit source-to-source transformations that
-allow programmers to develop and maintain programs at a high-level of
-abstraction, without coding to a specific hardware architecture.
-Furthermore these transformations allow multiple hardware architectures
-to be targeted without changing the high-level source.  It also removes the
-necessity for application programmers to understand details of the accelerator
-architecture or to know OpenCL.
+%%The sheer complexity of programming for clusters of many or multi-core processors with tens of millions threads of execution makes the simplicity of the data-parallel model attractive.
+The increasing complexity of today's applications (especially in light of the increasing complexity of the hardware) and the need for portability across architectures make a higher-level and simpler programming model like data-parallel attractive.
+
+%%The goal of this work has been to exploit source-to-source transformations that allow programmers to develop and maintain programs at a high-level of abstraction, without coding to a specific hardware architecture.
+Furthermore these transformations allow multiple hardware architectures to be targeted without changing the high-level source.
+It also removes the necessity for application programmers to understand details of the accelerator architecture or to know OpenCL.
 
@@
 
-\bibliographystyle{abbrv}
-\bibliography{local}
+%\bibliographystyle{abbrv}
+%\bibliography{local}
 

### 2015-01-22 papers/LOPe/program-restructuring.tex
Escaped # symbol.

@@
 to expose parallelism (a Fortran 2008 addition) or by using similar constructs
-available in OpenMP like, {\tt #pragma omp parallel for}.
+available in OpenMP like, {\tt \#pragma omp parallel for}.
 

### 2015-01-22 papers/LOPe/programming-model.tex
Starting work on EuroPar version.

--- 2011-04-09 opencl/paper-local/programming-model.tex
@@
-\section{Program Restructuring}
-
-Major challenges face application programmers on exescale platforms.  Among
-these challenges are of course performance, but in addition, programmers are
-also challenged by the need for their applications to be maintainable and
-portable across a wide variety of hardware platforms.  Unfortunately,
-performance often is at odds with maintainability and portability.  For example,
-hardware accelerators like GPUs are rapidly evolving and require separate coding
-for each GPU vendor and even for a specific vendor's range of models.  A
-synthetic cognition application at LANL has six different software versions just
-to support a subset of the possible hardware combinations that exist today.
-
-To meet these challenges we propose three main thrust areas:
-
-1. Guidance to users on the choice of a programming model for their
-applications.  This guidance will be based on an extensive analysis by proposal
-team members on the models that would work well on existing and future exoscale
-hardware architectures along with input from the user communities.  A high-level
-programming model will provide application programmers with the ability to
-incrementally raise the level of abstraction within their applications.  Only by
-raising the level of abstraction can codes be retargeted by compilers across the
-expected range of exoscale platforms.  Discussions with scientists in various
-scientific disciplines (e.g., atmospheric science) suggest that programmers
-would prefer that their codes just work \emph{as is} on exoscale platforms,
-primarily because of the huge effort envolved in recoding a major scientific
-code.  However experience on porting codes to LANL's petascale Roadrunner
-computer suggests this may not be entirely possible, at least on early
-encarnations of exoscale platforms.  Every code ported to Roadrunner was
-completely rewritten to perform well on Roadrunner's many-core processors.  It
-is our goal to provide user-guided refactoring tools to incrementally modify
-codes in order to ease the porting effort for programmers, without a complete
-hand recoding as was required for the initial applications that were ported to
-Roadrunner.
-
-2. Once a user or user community has chosen a programming model, tools can be
-provided that restructure an application to adopt the chosen programming model.
-These code refactorings need not be complicated.  For example, OpenMP uses
-compiler directives to provide high-level information to the compiler.  Thus
-analysis and refactoring tools guiding users in porting codes to an OpenMP
-threading model would be appropriate.  Other tools supporting a data parallel
-programming model would require a different set of tools.
-
-3. Finally, given functions (or code segments) that have been restructured to
-take advantage the chosen, high-level programming model, compiler tools are
-needed to generate code for different system architectures.  We intent to
-develop tools that are complimentary to and exploit a hardware vendor's C, C++,
-OpenCL, and Fortran compilers.  We propose to develop source-to-source
-transformational tools that take in high-level code and produce an intermidiate
-code that can be further optimized by the vendor's backend compiler.
-
-\subsection{Programming Models}
-
-While it is impossible to predict the exact system architecture that an
-exoscale system will have, we can anticipate the general direction that a
-programming model must support.  We anticipate the need to support over 100K
-nodes supported by a message passing communication system.  Each node can be
-expected to contain a few 10s of heterogeneous cores with the possibility of
-order 1K additional simple cores like those seen in existing GPUs.  Thus the
-programming model must support this hierarchical distribution of
-memory and processing elements.
-
-Replicated programs using message passing with MPI is the current standard for
-parallel programming using distributed nodes.  Parallel programmers have become
-used to this environment and we expect to support this programming environment.  While
-OpenMP is gathering acceptance, unfortunately there is currently not a
-widely-adopted programming model that exploits on-node parallelism like MPI does
-for distributed nodes.  Two evolving cross-platform standards for developing
-multi-threaded applications are OpenMP and OpenCL.  We intend to develop tools
-that support both.  OpenMP as a way to express parallelism directly within an
-application and OpenCL as an intermediate language generated by our
-source-to-source compiler tools.  OpenCL is too low-level a language to achieve our
-code-maintainability goal but supports a data-parallel programming model as a
-high level of abstraction.
-
-\subsection{Source-Code Abstractions}
-
-Since a majority of scientific codes are written in Fortran, a large effort will
-be given to supporting the Fortran programming language.  Fortunately Fortran is
-an ideal language for programming at a high level of abstraction, especially for
-procedures chosen as targets for threaded architectures like GPUs, as it has a
-well-developed array abstraction as part of the base language.  While not always
-exploited in existing Fortran applications, most of these abstractions are
-available in todays compilers (although not currently targeted for GPUs).
-Fortran provides a data-parallel programming model much like CUDA and is well
-suited for intermedicate code generation to, for example, OpenCL (or CUDA if it
-becomes a standard supported by multiple vendors).  Many of these abstractions
-can be extended to C and C++ by array abstractions and compiler directives.
-Currently Intel is working on adding array extensions to their C compiler.
-
-In Fortran a user can choose between several source-code abstractions for
-exploiting parallelism.  The goal of this proposal is to provide tools transform
-these abstractions to low-level code for final compilation.  Fortran supports a
-data-parallel programming model with array syntax, e.g., C = A + B, where A, B,
-C are arrays.  Or the user can explicitly use loop constructs like DO CONCURRENT
-to expose parallelism (a Fortran 2008 addition) or by using similar constructs
-available in OpenMP like, {\tt #pragma omp parallel for}.
+\section{Programming Model}
 
@@
 
-We highlight the Fortran elemental source-code abstraction because it provides a
-local orientation similar to that of OpenCL or CUDA kernel functions that were
-designed to exploit streaming, highly-threaded architectures like GPUs.
-Elemental functions are pure in that they are guaranteed to be free of side
-effects such as I/O.  Consider a convolution where a 3x3 filter is applied to
-individual pixel elements in a photograph to average of blur the original photograph.
-A convolution is similar to stencil operations in computational fluid dynamics that
-are used to obtain spatial derivatives of state variables.  Simple code is shown
-with extentions to Fortran shown in capital letters.  The extensions could be
-provided to C or Fortran functions via compiler directives rather than explicit 
-language syntax.
+%%We highlight the Fortran elemental source-code abstraction because it provides a local orientation similar to that of OpenCL or CUDA kernel functions that were designed to exploit streaming, highly-threaded architectures like GPUs.
+Elemental functions are pure in that they are guaranteed to be free of side effects such as I/O.
+Consider a convolution where a 3x3 filter is applied to individual pixel elements in a photograph to average of blur the original photograph.
+A convolution is similar to stencil operations in computational fluid dynamics that are used to obtain spatial derivatives of state variables.
+Simple code is shown with extentions to Fortran shown in capital letters.
+The extensions could be provided to C or Fortran functions via compiler directives rather than explicit language syntax.
+
+Consider a simple average over a five point stencil in two dimensions.  The algorithm updates the
+stencil interior point $A(0,0)$ as shown in the following:
 
 \begin{verbatim}
-    CONCURRENT real elemental function convolve(a, filter)
-       real, intent(in) :: a, filter(3,3)
-       real :: a_region(3,3)
+  pure CONCURRENT subroutine average(A)
+     real, HALO(:,:) :: A
 
-       a_region = HALO(a, [1,1,1,1])
-       convolve = sum( filter*a_region )
+     A(0,0) =              A(+1,0)                &
+               + A(0,-1) + A( 0,0)   +  A(0,+1)   &
+                         + A(-1,0)
+
+     A(0,0) = (1./5.) * A(0,0)
 
@@
 
-The {\tt convolve} function is called with
+The
+
 \begin{verbatim}
-    A = convolve(A, filter)
+   real, allocatable, HALO(-1:*:1) :: A(:)[:], B(:)[:]
+
+   !! Allocate local memory
+   !  -------------------------------------------
+   allocate(A(N)[*], B(N)[*])
 \end{verbatim}
-where {\tt A} and {\tt Avg} are arrays, but where the {\tt a} argument in the
-{\tt convolve} definition is declared as a scalar.  The {\tt HALO} function
-returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.  The
-{\tt convolve} function is free of race conditions because of the copy semantics
-of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute,
-whereby no output variables can be updated before all threads have completed
-execution.  In addition, while any thread may load from an extended region
-about \emph{its} element with the {\tt HALO} function, it may only store into
-its own elemental location.
 
-\subsection{Code Restructuring Tools}
+\begin{verbatim}
+   device = get_subimage(1)          ! THIS_IMAGE() if device unavailable
 
-This example is provided in some detail as it highlights several of the tools
-that are useful in aiding the user in the task of restructuring source code.
-The goal of the restructuring is to transform it into a high-level, maintainable
-form, that can be automatically transformed into a source form that can
-ultimately be compiled into an executable on a particular hardware platform.  We
-assume that the starting point is a traditional serial code using loop
-constructs; the convolution example would require four nested loops.
+   !! Allocate memory on subimage if available
+   !  ----------------------------------------
+   if (device /= THIS_IMAGE()) then
+      allocate(U1h(0:J)  [device], HALO=(-1:*:1))   [[device]]
+      allocate(U2h(0:J/2)[device], HALO=(-1:*:1))   [[device]]
+   end if
 
-\begin{itemize}
+   !! Initialize and transfer memory
+   !  ------------------------------
+   do j1h = 0, J
+      U1h(j1h) = j1h
+   end do
+   U1h[device] = U1h
 
-\item \emph{Code outliner.}  This tool would replace a block of code chosen by
-the user with a function and a call to the function.  In the above example the
-loops over the array elements would be moved to the elemental {\tt convolve}
-function and replaced by a call to the newly created function.  The tool may
-require input from the user as it completes its task (size of ghost/halo region
-for example).
+   do concurrent (j2h = 1:J/2-1)  [[device]]
+      call restrict(U1h(2*j2h)[device], U2h(j2h)[device])
+   end do
 
-\item \emph{User modifications.} At this point the user may want to restructure
-the outlined function {\tt convolve} by hand, by replacing the loops over the
-filter with array syntax as shown in the example.
+   do concurrent (j2h = 1:J/2-1)  [[device]]
+      call restrict(U1h(2*j2h)[device], U2h(j2h)[device], [0.25, 0.5, 0.25])
+   end do
 
-\item \emph{Insertion of memory placement hints.}  Any compiler generated code
-will have to be conservative regarding where memory resides, whether on the
-accelerator device or on the host CPU.  For the above example, memory for {\tt
-A} will need to be copied to and from the device on entry and exit of function
-{\tt convolve}.  However, memory for {\tt A} may already reside on the device
-from a previous function invocation and therefore need not be copied.  Users may
-provide hints to the compiler as to whether to copy the memory and if so, where
-to begin the memory copy so that it can be done asynchronously with other work.
-Tools will be created to guide the user in making this placement based on
-dependence analysis.
+   deallocate(U1h, U2h)
 
-\item \emph{Insertion of MPI communication.}  Memory placement hints can also be
-used so that calls to MPI can be inserted into the code for the communication of
-ghost cells between nodes.  Users may start with a version of the code that
-already uses MPI, but the same information needed to generate OpenCL kernels can
-be used to generate MPI communication; off-node memory is just another level in
-the memory hierarchy.
+   if (device /= THIS_IMAGE()) then
+      deallocate(U1h[device], U2h[device])   [[device]]
+   end if
 
-%%% NOTE, I can provide a figure here showing the memory hierarchy.
+end program
+\end{verbatim}
 
-\item \emph{Insertion of OpenMP directives.}  As an alternative to outlining a
-function (the opposite of inlining) a tool will be created to restructure user
-code by inserting OpenMP directives to indicate that loop iterations can be
-performed concurrently, for example.  This tool will provide guidance to the
-user as to where it may be appropriate to place these directives (and why) but
-the ultimate responsibility for where and what directives to be used is a user's
-responsibility.
+%%where {\tt A} and {\tt Avg} are arrays, but where the {\tt a} argument in the {\tt convolve} definition is declared as a scalar.
+The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.
+The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.
+In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.
 
-\item \emph{Replace with DO CONCURRENT.}  Another alternative to code outlining
-is to replace loops in Fortran with the DO CONCURRENT construct.  As with
-compiler directives, this restructuring would require user input and can also
-suggest loops that are candidates for converting.
+\subsection{Memory Management}
 
-\end{itemize}
+This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.
+It describes what is the compiler's responsibility and which is the user's responsibility.
 
-At this point it is assumed that the user has finished restructuring the
-original source into a form such that that parallelism is exposed to the
-source-to-source compiler through language syntax or compiler directives.  This
-is the high-level, maintainable, and portable source form that is stored in CVS
-repository.  Tools are also required that can take the high-level source and
-retargets it to a specific hardware platform.  These tools produce a lower-level
-source, through source-to-source transformations, that a vendor's compiler will
-use to generate executable code.  While described as separate tools, they will
-likely be aggragated into a single tool.
+%% A note from a conversation with Matt regarding memory management
+%%
 
-The transformations may produce code for following targets (a combination of
-language and hardware platform): 1. Serial version to be run on a single core.
-This is necessary because some language constructs like DO CONCURRENT may not be
-available for a period of time; 2. OpenCL version optimized for GPUs; 3. OpenCL
-for optimzed for multiple CPU cores; and 4. OpenMP version for final compilation
-with an OpenMP compiler.
+%The Fortran language has much tighter restrictions on aliasing than does C.
+%So unless a variable has the pointer or target attribute, it cannot be aliased.
+%Thus the compiler is able to agressively optimized for memory movement between
+%the CPU and accelerator.  However, because the design philosophy of coarrays
+%is that memory transfer between images can be expensive, the programmer must
+%explicitly transfer memory between images with explicit syntax with square
+%bracket notation, i.e., $a[1] = a[2]$.  So we allow the compiler to manage
+%memory within an image but require the use of {\tt halo\_exchange} for the
+%transfer of halo memory between images.
 
-\begin{itemize}
+%Fortran currently supports:
 
-\item \emph{Thread initialization.} This tool generates code for calls to the
-thread runtime library to initialize all thread devices and thread memory.
-
-\item \emph{Code inliner.} This tool inlines calls to elemental functions,
-effectively undoing the effect of the outliner. It is not needed for Fortran as
-the compiler already performs this task.
-
-\item \emph{Thread execution.} This tool generates code for execution of an
-elemental or kernel function by a thread runtime library like OpenCL.
-
-\item \emph{Memory transfer.}  This tool generates code for all memory transfer
-and synchronization points.
-
-\end{itemize}
-
-Programming Models
-------------------
-
-Fortran currently supports:
-
-1.
+%1.
 Array syntax: e.g., C = A + B, where A, B, C are arrays.
@@
 
-2.
+%2.
 Pure procedures:  Fortran has syntax for specifying procedures that have no side effects during execution.
@@
 
-3.
+%3.
 Pure elemental procedures: Fortran has syntax for specifying procedures that take only scalar arguments, but may be applied across array elements.
@@
 
-We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.
+%We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.
 This additional syntax is provided by functions that return a copy of a small region of memory surrounding an array element (as seen within an elemental procedure) and with functions for thread synchronization.
@@
 
-In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.
+%In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.
 This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.

### 2015-01-23 papers/LOPe/motivation.tex
Added first paragraph to motivation section.

@@
 \section{Motivation}
+
+The concepts in this paper were motivated by problems encountered during
+the development of PetaVision, a C++ software framework designed for simulating
+large networks of spiking neurons in the visual cortex of primates.  PetaVision
+was designed to run on massively parallel hardware architectures and a primitive
+version of PetaVision achieved over a Petaflop/s of sustained single precision
+performance on Roadrunner, the fastest computer in the world for a brief
+period of time.
+
+Describe neural layers, spiking neurons, synaptic connections, weights,
+and plasticity.  Reference Heubel.
+
+Give a simple 1D (cartoon) example of connections within a neural network in
+Figure 1.  Describe the size of visual cortex in terms of the number of neurons,
+the number of synaptic connections, and performance.  Reference SC Gorden Bell
+paper from IBM.
 

### 2015-01-23 papers/LOPe/local.tex
Moving to Springer LNCS style guide.

--- 2011-10-27 papers/LOPe/local.tex
@@
-\documentclass{acm_proc_article-sp}
+%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This is the LaTeX source for the instructions to authors using
+% the LaTeX document class 'llncs.cls' for contributions to
+% the Lecture Notes in Computer Sciences series.
+% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
+%
+% It may be used as a template for your own input - copy it
+% to a new file with a new name and use it as the basis
+% for your article.
+%
+% NB: the document class 'llncs' has its own and detailed documentation, see
+% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\usepackage{cite}
+\documentclass[runningheads,a4paper]{llncs}
+
+\usepackage{amssymb}
+\setcounter{tocdepth}{3}
 \usepackage{graphicx}
-%\usepackage[cmex10]{amsmath}
-%\usepackage{amssymb}
-%\usepackage{array}
+
 \usepackage{url}
-\usepackage{listings}
-
-\lstdefinestyle{FortranLike}{float,frame=lines,language=Fortran,commentstyle=\ttfamily,basicstyle=\ttfamily}
-\lstdefinestyle{CLike}{float,frame=lines,language=C,commentstyle=\ttfamily,basicstyle=\ttfamily}
-\lstdefinestyle{NoFloatCLike}{frame=lines,language=C,commentstyle=\ttfamily,basicstyle=\ttfamily}
+\urldef{\mailsa}\path|rasmus@uoregon.edu, mjsottile@computer.org,|
+\urldef{\mailsb}\path|dannagle@verizon.net, soren.rasmussen@aggiemail.usu.edu|    
+\newcommand{\keywords}[1]{\par\addvspace\baselineskip
+\noindent\keywordname\enspace\ignorespaces#1}
 
@@
 
-\title{Locally-Oriented Programming: A stunningly simple programming model for GPU programming in Fortran and C++}
+\mainmatter  % start of an individual contribution
 
-\numberofauthors{4}
+% first the title is needed
+\title{Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations}
 
-\author{
-% 1st. author
-\alignauthor
-Craig E Rasmussen\\
-       \affaddr{Los Alamos National Laboratory}\\
-%       \affaddr{CCS-7, MS B287}\\
-       \affaddr{Los Alamos, NM 87545}\\
-       \email{crasmussen@lanl.gov}
-% 2nd. author
-\alignauthor
-Matthew J. Sottile\\
-       \affaddr{Galois, Inc.}\\
-%       \affaddr{421 SW 6th Ave. Suite 300}\\
-       \affaddr{Portland, OR 97204}\\
-       \email{mjsottile@computer.org}
-% 3rd. author
-\alignauthor
-Dan Nagle\\
-       \email{dannagle@verizon.net}
-% 4th. author
-\alignauthor
-Daniel Quinlan\\
-       \affaddr{Lawrence Livermore National Laboratory}\\
-       \email{dquinlan@llnl.gov}
-}
+% a short form should be given in case it is too long for the running head
+\titlerunning{Locally-Oriented Programming}
 
+\author{Craig Rasmussen%
+\thanks{The authors would like to thank Robert Robey and Wayne Weseloh
+at Los Alamos National Laborabory for several stimulating
+conversations with respect to programming models.}%
+\and Matthew Sottile\and Dan Nagle\and Soren Rasmussen}
+
+\authorrunning{Locally-Oriented Programming}
+% (feature abused for this document to repeat the title also on left hand pages)
+
+% the affiliations are given next; don't give your e-mail address
+% unless you accept that it will be published
+\institute{441 McKenzie Hall,\\
+5246 University of Oregon, Eugene, Oregon, USA\\
+\mailsa\\
+\mailsb\\
+\url{https://openfortran.uoregon.edu}}
+
+%
+% NB: a more complex sample for affiliations and the mapping to the
+% corresponding authors can be found in the file "llncs.dem"
+% (search for the string "\mainmatter" where a contribution starts).
+% "llncs.dem" accompanies the document class "llncs.cls".
+%
+
+%\toctitle{Lecture Notes in Computer Science}
+%\tocauthor{Authors' Instructions}
 \maketitle
@@
 \begin{abstract}
-Emerging GPU architectures for high performance computing are well suited to a
-data-parallel programming model.  However, programming for these architectures
-currently requires learning a new language, e.g., OpenCL or CUDA, or employing
-OpenMP compiler directives.  This paper presents preliminary work examining how
-programming with a local orientation can be employed in Fortran (or using a
-special array library in C++) to provide simple access GPU architectures.  A
-locally-oriented programming model is especially useful for stencil-based codes
-or algorithms requiring the application of convolution kernels.  In this
-programming model, a programmer codes the algorithm {\it only} for a single
-array element, but has read-only access to a small sub-array surrounding the
-given array element, so that a stencil can be applied.  We demonstrate how a
-locally-oriented programming model can be adopted using source-to-source program
-transformations in conjuction with a compiler-based infrastructure like ROSE.
+
+Emerging accelerator architectures for high performance computing are well
+suited to a data-parallel programming model.  However, programming for
+these architectures currently requires learning a new language, e.g.,
+OpenCL or CUDA, or employing OpenMP compiler directives.  This paper
+presents preliminary work examining how programming with a local
+orientation can be employed in Fortran (or using a special array
+library in C++) to provide simple access to accelerator architectures.  A
+locally-oriented programming model is especially useful for
+stencil-based codes or algorithms requiring the application of
+convolution kernels.  In this programming model, a programmer codes
+the algorithm \emph{only} for a single array element, but has read-only
+access to a small sub-array surrounding the given array element, so
+that a stencil can be applied.  We demonstrate how a locally-oriented
+programming model can be adopted using source-to-source program
+transformations in conjuction with a compiler-based infrastructure
+like ROSE.
+
+\keywords{stencil compilers, distributed memory parallelism}
 \end{abstract}
 
-% A category with the (minimum) three required fields
-%\category{D.3.3}{Language Constructs and Features}{Concurrent programming structures}
+\section{Introduction}
 
-\input{intro}
-\input{programming-model}
-\input{shallow-water}
-\input{source-to-source}
-%\input{static-analysis}
-\input{performance}
+You are strongly encouraged to use \LaTeXe{} for the
+preparation of your camera-ready manuscript together with the
+corresponding Springer class file \verb+llncs.cls+. Only if you use
+\LaTeXe{} can hyperlinks be generated in the online version
+of your manuscript.
 
-\section{Conclusions}
+The \LaTeX{} source of this instruction file for \LaTeX{} users may be
+used as a template. This is
+located in the ``authors'' subdirectory in
+\url{ftp://ftp.springer.de/pub/tex/latex/llncs/latex2e/instruct/} and
+entitled \texttt{typeinst.tex}. There is a separate package for Word 
+users. Kindly send the final and checked source
+and PDF files of your paper to the Contact Volume Editor. This is
+usually one of the organizers of the conference. You should make sure
+that the \LaTeX{} and the PDF files are identical and correct and that
+only one version of your paper is sent. It is not possible to update
+files at a later stage. Please note that we do not need the printed
+paper.
 
-The sheer complexity of programming for clusters of many or multi-core
-processors with tens of millions threads of execution makes the simplicity of
-the data-parallel model attractive.  The increasing complexity of
-today's applications (especially in light of the increasing complexity
-of the hardware) and the need for portability across architectures
-make a higher-level and simpler programming model like data-parallel
-attractive.
+We would like to draw your attention to the fact that it is not possible
+to modify a paper in any way, once it has been published. This applies
+to both the printed book and the online version of the publication.
+Every detail, including the order of the names of the authors, should
+be checked before the paper is sent to the Volume Editors.
 
-The goal of this work has been to exploit source-to-source transformations that
-allow programmers to develop and maintain programs at a high-level of
-abstraction, without coding to a specific hardware architecture.
-Furthermore these transformations allow multiple hardware architectures
-to be targeted without changing the high-level source.  It also removes the
-necessity for application programmers to understand details of the accelerator
-architecture or to know OpenCL.
+\subsection{Checking the PDF File}
 
-%ACKNOWLEDGMENTS are optional
-\section{Acknowledgments}
-This work was supported in part by the Department of Energy Office of Science,
-Advanced Scientific Computing Research.
+Kindly assure that the Contact Volume Editor is given the name and email
+address of the contact author for your paper. The Contact Volume Editor
+uses these details to compile a list for our production department at
+SPS in India. Once the files have been worked upon, SPS sends a copy of
+the final pdf of each paper to its contact author. The contact author is
+asked to check through the final pdf to make sure that no errors have
+crept in during the transfer or preparation of the files. This should
+not be seen as an opportunity to update or copyedit the papers, which is
+not possible due to time constraints. Only errors introduced during the
+preparation of the files will be corrected.
 
-%%TODO \cite{chamberlain04zpl, roth97stencils}
+This round of checking takes place about two weeks after the files have
+been sent to the Editorial by the Contact Volume Editor, i.e., roughly
+seven weeks before the start of the conference for conference
+proceedings, or seven weeks before the volume leaves the printer's, for
+post-proceedings. If SPS does not receive a reply from a particular
+contact author, within the timeframe given, then it is presumed that the
+author has found no errors in the paper. The tight publication schedule
+of LNCS does not allow SPS to send reminders or search for alternative
+email addresses on the Internet.
 
-\bibliographystyle{abbrv}
-\bibliography{local}
+In some cases, it is the Contact Volume Editor that checks all the final
+pdfs. In such cases, the authors are not involved in the checking phase.
 
-\balancecolumns
+\subsection{Additional Information Required by the Volume Editor}
+
+If you have more than one surname, please make sure that the Volume Editor
+knows how you are to be listed in the author index.
+
+\subsection{Copyright Forms}
+
+The copyright form may be downloaded from the ``For Authors"
+(Information for LNCS Authors) section of the LNCS Website:
+\texttt{www.springer.com/lncs}. Please send your signed copyright form
+to the Contact Volume Editor, either as a scanned pdf or by fax or by
+courier. One author may sign on behalf of all of the other authors of a
+particular paper. Digital signatures are acceptable.
+
+\section{Paper Preparation}
+
+Springer provides you with a complete integrated \LaTeX{} document class
+(\texttt{llncs.cls}) for multi-author books such as those in the LNCS
+series. Papers not complying with the LNCS style will be reformatted.
+This can lead to an increase in the overall number of pages. We would
+therefore urge you not to squash your paper.
+
+Please always cancel any superfluous definitions that are
+not actually used in your text. If you do not, these may conflict with
+the definitions of the macro package, causing changes in the structure
+of the text and leading to numerous mistakes in the proofs.
+
+If you wonder what \LaTeX{} is and where it can be obtained, see the
+``\textit{LaTeX project site}'' (\url{http://www.latex-project.org})
+and especially the webpage ``\textit{How to get it}''
+(\url{http://www.latex-project.org/ftp.html}) respectively.
+
+When you use \LaTeX\ together with our document class file,
+\texttt{llncs.cls},
+your text is typeset automatically in Computer Modern Roman (CM) fonts.
+Please do
+\emph{not} change the preset fonts. If you have to use fonts other
+than the preset fonts, kindly submit these with your files.
+
+Please use the commands \verb+\label+ and \verb+\ref+ for
+cross-references and the commands \verb+\bibitem+ and \verb+\cite+ for
+references to the bibliography, to enable us to create hyperlinks at
+these places.
+
+For preparing your figures electronically and integrating them into
+your source file we recommend using the standard \LaTeX{} \verb+graphics+ or
+\verb+graphicx+ package. These provide the \verb+\includegraphics+ command.
+In general, please refrain from using the \verb+\special+ command.
+
+Remember to submit any further style files and
+fonts you have used together with your source files.
+
+\subsubsection{Headings.}
+
+Headings should be capitalized
+(i.e., nouns, verbs, and all other words
+except articles, prepositions, and conjunctions should be set with an
+initial capital) and should,
+with the exception of the title, be aligned to the left.
+Words joined by a hyphen are subject to a special rule. If the first
+word can stand alone, the second word should be capitalized.
+
+Here are some examples of headings: ``Criteria to Disprove
+Context-Freeness of Collage Language", ``On Correcting the Intrusion of
+Tracing Non-deterministic Programs by Software", ``A User-Friendly and
+Extendable Data Distribution System", ``Multi-flip Networks:
+Parallelizing GenSAT", ``Self-determinations of Man".
+
+\subsubsection{Lemmas, Propositions, and Theorems.}
+
+The numbers accorded to lemmas, propositions, and theorems, etc. should
+appear in consecutive order, starting with Lemma 1, and not, for
+example, with Lemma 11.
+
+\subsection{Figures}
+
+For \LaTeX\ users, we recommend using the \emph{graphics} or \emph{graphicx}
+package and the \verb+\includegraphics+ command.
+
+Please check that the lines in line drawings are not
+interrupted and are of a constant width. Grids and details within the
+figures must be clearly legible and may not be written one on top of
+the other. Line drawings should have a resolution of at least 800 dpi
+(preferably 1200 dpi). The lettering in figures should have a height of
+2~mm (10-point type). Figures should be numbered and should have a
+caption which should always be positioned \emph{under} the figures, in
+contrast to the caption belonging to a table, which should always appear
+\emph{above} the table; this is simply achieved as matter of sequence in
+your source.
+
+Please center the figures or your tabular material by using the \verb+\centering+
+declaration. Short captions are centered by default between the margins
+and typeset in 9-point type (Fig.~\ref{fig:example} shows an example).
+The distance between text and figure is preset to be about 8~mm, the
+distance between figure and caption about 6~mm.
+
+To ensure that the reproduction of your illustrations is of a reasonable
+quality, we advise against the use of shading. The contrast should be as
+pronounced as possible.
+
+If screenshots are necessary, please make sure that you are happy with
+the print quality before you send the files.
+\begin{figure}
+\centering
+\includegraphics[height=6.2cm]{eijkel2}
+\caption{One kernel at $x_s$ (\emph{dotted kernel}) or two kernels at
+$x_i$ and $x_j$ (\textit{left and right}) lead to the same summed estimate
+at $x_s$. This shows a figure consisting of different types of
+lines. Elements of the figure described in the caption should be set in
+italics, in parentheses, as shown in this sample caption.}
+\label{fig:example}
+\end{figure}
+
+Please define figures (and tables) as floating objects. Please avoid
+using optional location parameters like ``\verb+[h]+" for ``here".
+
+\paragraph{Remark 1.}
+
+In the printed volumes, illustrations are generally black and white
+(halftones), and only in exceptional cases, and if the author is
+prepared to cover the extra cost for color reproduction, are colored
+pictures accepted. Colored pictures are welcome in the electronic
+version free of charge. If you send colored figures that are to be
+printed in black and white, please make sure that they really are
+legible in black and white. Some colors as well as the contrast of
+converted colors show up very poorly when printed in black and white.
+
+\subsection{Formulas}
+
+Displayed equations or formulas are centered and set on a separate
+line (with an extra line or halfline space above and below). Displayed
+expressions should be numbered for reference. The numbers should be
+consecutive within each section or within the contribution,
+with numbers enclosed in parentheses and set on the right margin --
+which is the default if you use the \emph{equation} environment, e.g.,
+\begin{equation}
+  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
+  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt \;  .
+\end{equation}
+
+Equations should be punctuated in the same way as ordinary
+text but with a small space before the end punctuation mark.
+
+\subsection{Footnotes}
+
+The superscript numeral used to refer to a footnote appears in the text
+either directly after the word to be discussed or -- in relation to a
+phrase or a sentence -- following the punctuation sign (comma,
+semicolon, or period). Footnotes should appear at the bottom of
+the
+normal text area, with a line of about 2~cm set
+immediately above them.\footnote{The footnote numeral is set flush left
+and the text follows with the usual word spacing.}
+
+\subsection{Program Code}
+
+Program listings or program commands in the text are normally set in
+typewriter font, e.g., CMTT10 or Courier.
+
+\medskip
+
+\noindent
+{\it Example of a Computer Program}
+\begin{verbatim}
+program Inflation (Output)
+  {Assuming annual inflation rates of 7%, 8%, and 10%,...
+   years};
+   const
+     MaxYears = 10;
+   var
+     Year: 0..MaxYears;
+     Factor1, Factor2, Factor3: Real;
+   begin
+     Year := 0;
+     Factor1 := 1.0; Factor2 := 1.0; Factor3 := 1.0;
+     WriteLn('Year  7% 8% 10%'); WriteLn;
+     repeat
+       Year := Year + 1;
+       Factor1 := Factor1 * 1.07;
+       Factor2 := Factor2 * 1.08;
+       Factor3 := Factor3 * 1.10;
+       WriteLn(Year:5,Factor1:7:3,Factor2:7:3,Factor3:7:3)
+     until Year = MaxYears
+end.
+\end{verbatim}
+%
+\noindent
+{\small (Example from Jensen K., Wirth N. (1991) Pascal user manual and
+report. Springer, New York)}
+
+\subsection{Citations}
+
+For citations in the text please use
+square brackets and consecutive numbers: \cite{jour}, \cite{lncschap},
+\cite{proceeding1} -- provided automatically
+by \LaTeX 's \verb|\cite| \dots\verb|\bibitem| mechanism.
+
+\subsection{Page Numbering and Running Heads}
+
+There is no need to include page numbers. If your paper title is too
+long to serve as a running head, it will be shortened. Your suggestion
+as to how to shorten it would be most welcome.
+
+\section{LNCS Online}
+
+The online version of the volume will be available in LNCS Online.
+Members of institutes subscribing to the Lecture Notes in Computer
+Science series have access to all the pdfs of all the online
+publications. Non-subscribers can only read as far as the abstracts. If
+they try to go beyond this point, they are automatically asked, whether
+they would like to order the pdf, and are given instructions as to how
+to do so.
+
+Please note that, if your email address is given in your paper,
+it will also be included in the meta data of the online version.
+
+\section{BibTeX Entries}
+
+The correct BibTeX entries for the Lecture Notes in Computer Science
+volumes can be found at the following Website shortly after the
+publication of the book:
+\url{http://www.informatik.uni-trier.de/~ley/db/journals/lncs.html}
+
+\subsubsection*{Acknowledgments.} The heading should be treated as a
+subsubsection heading and should not be assigned a number.
+
+\section{The References Section}\label{references}
+
+In order to permit cross referencing within LNCS-Online, and eventually
+between different publishers and their online databases, LNCS will,
+from now on, be standardizing the format of the references. This new
+feature will increase the visibility of publications and facilitate
+academic research considerably. Please base your references on the
+examples below. References that don't adhere to this style will be
+reformatted by Springer. You should therefore check your references
+thoroughly when you receive the final pdf of your paper.
+The reference section must be complete. You may not omit references.
+Instructions as to where to find a fuller version of the references are
+not permissible.
+
+We only accept references written using the latin alphabet. If the title
+of the book you are referring to is in Russian or Chinese, then please write
+(in Russian) or (in Chinese) at the end of the transcript or translation
+of the title.
+
+The following section shows a sample reference list with entries for
+journal articles \cite{jour}, an LNCS chapter \cite{lncschap}, a book
+\cite{book}, proceedings without editors \cite{proceeding1} and
+\cite{proceeding2}, as well as a URL \cite{url}.
+Please note that proceedings published in LNCS are not cited with their
+full titles, but with their acronyms!
+
+\begin{thebibliography}{4}
+
+\bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular
+Subsequences. J. Mol. Biol. 147, 195--197 (1981)
+
+\bibitem{lncschap} May, P., Ehrlich, H.C., Steinke, T.: ZIB Structure Prediction Pipeline:
+Composing a Complex Biological Workflow through Web Services. In: Nagel,
+W.E., Walter, W.V., Lehner, W. (eds.) Euro-Par 2006. LNCS, vol. 4128,
+pp. 1148--1158. Springer, Heidelberg (2006)
+
+\bibitem{book} Foster, I., Kesselman, C.: The Grid: Blueprint for a New Computing
+Infrastructure. Morgan Kaufmann, San Francisco (1999)
+
+\bibitem{proceeding1} Czajkowski, K., Fitzgerald, S., Foster, I., Kesselman, C.: Grid
+Information Services for Distributed Resource Sharing. In: 10th IEEE
+International Symposium on High Performance Distributed Computing, pp.
+181--184. IEEE Press, New York (2001)
+
+\bibitem{proceeding2} Foster, I., Kesselman, C., Nick, J., Tuecke, S.: The Physiology of the
+Grid: an Open Grid Services Architecture for Distributed Systems
+Integration. Technical report, Global Grid Forum (2002)
+
+\bibitem{url} National Center for Biotechnology Information, \url{http://www.ncbi.nlm.nih.gov}
+
+\end{thebibliography}
+
+\section*{Appendix: Springer-Author Discount}
+
+LNCS authors are entitled to a 33.3\% discount off all Springer
+publications. Before placing an order, the author should send an email, 
+giving full details of his or her Springer publication,
+to \url{orders-HD-individuals@springer.com} to obtain a so-called token. This token is a
+number, which must be entered when placing an order via the Internet, in
+order to obtain the discount.
+
+\section{Checklist of Items to be Sent to Volume Editors}
+Here is a checklist of everything the volume editor requires from you:
+
+\begin{itemize}
+\settowidth{\leftmargin}{{\Large$\square$}}\advance\leftmargin\labelsep
+\itemsep8pt\relax
+\renewcommand\labelitemi{{\lower1.5pt\hbox{\Large$\square$}}}
+
+\item The final \LaTeX{} source files
+\item A final PDF file
+\item A copyright form, signed by one author on behalf of all of the
+authors of the paper.
+\item A readme giving the name and email address of the
+corresponding author.
+\end{itemize}
 \end{document}

### 2015-01-24 papers/LOPe/local.tex
Working on paper layout and program models section primarily.

--- 2011-10-27 papers/LOPe/local.tex
@@
 
-\title{Locally-Oriented Programming: A stunningly simple programming model for GPU programming in Fortran and C++}
+\mainmatter  % start of an individual contribution
 
-\numberofauthors{4}
+% first the title is needed
+\title{Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations
+       on Multi-Level Distributed Memory Architectures}
 
-\author{
-% 1st. author
-\alignauthor
-Craig E Rasmussen\\
-       \affaddr{Los Alamos National Laboratory}\\
-%       \affaddr{CCS-7, MS B287}\\
-       \affaddr{Los Alamos, NM 87545}\\
-       \email{crasmussen@lanl.gov}
-% 2nd. author
-\alignauthor
-Matthew J. Sottile\\
-       \affaddr{Galois, Inc.}\\
-%       \affaddr{421 SW 6th Ave. Suite 300}\\
-       \affaddr{Portland, OR 97204}\\
-       \email{mjsottile@computer.org}
-% 3rd. author
-\alignauthor
-Dan Nagle\\
-       \email{dannagle@verizon.net}
-% 4th. author
-\alignauthor
-Daniel Quinlan\\
-       \affaddr{Lawrence Livermore National Laboratory}\\
-       \email{dquinlan@llnl.gov}
-}
+% a short form should be given in case it is too long for the running head
+\titlerunning{Locally-Oriented Programming}
 
+\author{Craig Rasmussen%
+\thanks{The authors would like to thank Robert Robey and Wayne Weseloh
+at Los Alamos National Laborabory for several stimulating
+conversations with respect to programming models.}%
+\and Matthew Sottile\and Dan Nagle\and Soren Rasmussen}
+
+\authorrunning{Locally-Oriented Programming}
+% (feature abused for this document to repeat the title also on left hand pages)
+
+% the affiliations are given next; don't give your e-mail address
+% unless you accept that it will be published
+\institute{441 McKenzie Hall,\\
+5246 University of Oregon, Eugene, Oregon, USA\\
+\mailsa\\
+\mailsb\\
+\url{https://openfortran.uoregon.edu}}
+
+%
+% NB: a more complex sample for affiliations and the mapping to the
+% corresponding authors can be found in the file "llncs.dem"
+% (search for the string "\mainmatter" where a contribution starts).
+% "llncs.dem" accompanies the document class "llncs.cls".
+%
+
+%\toctitle{Lecture Notes in Computer Science}
+%\tocauthor{Authors' Instructions}
 \maketitle
@@
 \begin{abstract}
-Emerging GPU architectures for high performance computing are well suited to a
-data-parallel programming model.  However, programming for these architectures
-currently requires learning a new language, e.g., OpenCL or CUDA, or employing
-OpenMP compiler directives.  This paper presents preliminary work examining how
-programming with a local orientation can be employed in Fortran (or using a
-special array library in C++) to provide simple access GPU architectures.  A
-locally-oriented programming model is especially useful for stencil-based codes
-or algorithms requiring the application of convolution kernels.  In this
-programming model, a programmer codes the algorithm {\it only} for a single
-array element, but has read-only access to a small sub-array surrounding the
-given array element, so that a stencil can be applied.  We demonstrate how a
-locally-oriented programming model can be adopted using source-to-source program
-transformations in conjuction with a compiler-based infrastructure like ROSE.
+
+Emerging hybrid accelerator architectures for high performance computing are often well suited for
+usage of a data-parallel programming model.  However, programmers for these emerging architectures
+face a steep learning curve that frequently requires learning a new language, e.g., OpenCL or CUDA,
+or employing OpenMP compiler directives.  Furthermore, the distributed (and frequently multi-level)
+nature of the memory organization of clusters of these architectures provides an additional level of
+complexity for the programmer.  This paper presents preliminary work examining how programming with
+a local orientation can be employed to provide simple access to accelerator architectures.  A
+locally-oriented programming model is especially useful for stencil-based codes or algorithms
+requiring the application of convolution kernels.  In this programming model, a programmer codes the
+algorithm by modifying \emph{only a single array element}, but has read-only access to a small
+sub-array surrounding the given array element, so that a stencil can be applied.  We demonstrate how
+a locally-oriented programming model can be adopted as an embedded, domain-specific language using
+source-to-source program transformations.
+
+\keywords{domain specific language, stencil compilers, distributed memory parallelism}
 \end{abstract}
 
-% A category with the (minimum) three required fields
-%\category{D.3.3}{Language Constructs and Features}{Concurrent programming structures}
+\input{motivation}
+\input{programming-model}
+\input{appendix-a}
 
-\input{intro}
-\input{programming-model}
-\input{shallow-water}
-\input{source-to-source}
-%\input{static-analysis}
-\input{performance}
-
-\section{Conclusions}
-
-The sheer complexity of programming for clusters of many or multi-core
-processors with tens of millions threads of execution makes the simplicity of
-the data-parallel model attractive.  The increasing complexity of
-today's applications (especially in light of the increasing complexity
-of the hardware) and the need for portability across architectures
-make a higher-level and simpler programming model like data-parallel
-attractive.
-
-The goal of this work has been to exploit source-to-source transformations that
-allow programmers to develop and maintain programs at a high-level of
-abstraction, without coding to a specific hardware architecture.
-Furthermore these transformations allow multiple hardware architectures
-to be targeted without changing the high-level source.  It also removes the
-necessity for application programmers to understand details of the accelerator
-architecture or to know OpenCL.
-
-%ACKNOWLEDGMENTS are optional
-\section{Acknowledgments}
+\subsection{Acknowledgments}
 This work was supported in part by the Department of Energy Office of Science,
@@
 
-%%TODO \cite{chamberlain04zpl, roth97stencils}
-
-\bibliographystyle{abbrv}
-\bibliography{local}
-
-\balancecolumns
 \end{document}

### 2015-01-24 papers/LOPe/motivation.tex
Working on paper layout and program models section primarily.

--- 2011-10-27 papers/LOPe/motivation.tex
@@
 
-\subsection{Coarray Fortran Comparison}
+The concepts in this paper were motivated by problems encountered during
+the development of PetaVision, a C++ software framework designed for simulating
+large networks of spiking neurons in the visual cortex of primates.  PetaVision
+was designed to run on massively parallel hardware architectures and a primitive
+version of PetaVision achieved over a Petaflop/s of sustained single precision
+performance on Roadrunner, the fastest computer in the world for a brief
+period of time.
 
-Below is the motivational example from the original Numrich and Reic paper.
-{NOTE: This should sortof/kindof show how it would work between nodes?  Next
-section shows OpenMP example for parallelism at the thread level.}
+TODO: Describe neural layers, spiking neurons, synaptic connections, weights,
+and plasticity. Cite Heubel.
 
-{\small
-\begin{verbatim}
-
-subroutine laplace (nrow,ncol,u)
-   integer, intent(in) real, intent(inout) real integer
-   ! that refers to the current image
-   left = me-1
-   if (me == 1) left = ncol
-   right = me + 1
-   if (me == ncol) right = 1
-   call sync_all( (/left,right/) ) ! Wait if left and right
-   ! have not already reached here
-   new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
-   call sync_all( (/left,right/) )
-   u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow) end subroutine laplace
-
-\end{verbatim}
-}
-
-Below is the equivalent implementation in LOPe.
-
-{\small
-\begin{verbatim}
-
-pure CONCURRENT subroutine laplace(U)
-   real, intent(inout), HALO(-1:*:1,-1:*:1) :: U(0:,0:)
-
-   U(0:0) = U(-1,0) + U(+1,0) + U(0,-1) + U(0,+1) - 3*U(0,0)
-
-end subroutine laplace
-
-\end{verbatim}
-}
-
-NOTE: Following not a criticism of CAF as CAF is a general purpose parallel
-programming language and LOPe only pertains to the halo parallel pattern.
-
-\subsubsection{Improvements}
-\begin{itemize}
-
-\item
-LOPe requires that the implementation of the algorithm to be separate from the
-call to effect the halo transfer.  Removing boundary condition specification
-(e.g., the cyclic boundary conditions implemented in the CAF example) from the
-algorithm allows the boundary conditions to be changed and without changing
-algorithm code.
-
-\item
-LOPe applies the transfer of halo memory across multiple (possibly) levels of
-memory with the intrinsic {\tt transfer\_halo} function (not shown here).  Thus
-the LOPe algorithm can be run on a machine with many interconnected nodes, each
-containing hybrid processor cores.  The coarray example can only be run on
-multiple nodes (called images in Fortran) without accelerator cores.
-
-\item
-The algorithm implementation is separate from user-specified synchronization,
-e.g., {\tt call sync\_all}.  In LOPe, synchronization is subsummed in the
-semantics of the CONCURRENT attribute and the {\tt transfer\_halo} function.
-
-\item
-The algorithm implementation is separate from any specification as to where the
-array memory is located.  The CAF example explicity denotes where memory is
-located with the {\tt [left]} and {\tt [right]} syntax where left and right
-specifiy a processor topology.
-
-\item
-The algorithm implementation is separate from any specification as to where the
-algorithm is to be executed.  The CAF example explicity denotes where a statement is to
-executed with control flow construct like {\tt if (me == 1)}.
-
-\item
-The LOPe implementation is easier to understand and frequently follows the
-mathematical algorithm directly.  For example, the CAF example adds 4 neighbors
-plus the center value to make the implementation with direct remote coarray
-access possible, while the LOPe example is able to implement the same algorithm
-with fewer operations by adding 4 neighbors (not including the center array
-element) and then only subtracting 3 center values.
-
-\item
-The semantics of LOPe makes explicit management of array temporaries (e.g., {\tt
-  u} and {\tt new\_u} by programmers unnecessary (though still possible).
-Because in LOPe the halo region is a language construct, the compiler is better
-able to manage temporary buffers than users on the target hardware platform.
-
-\end{itemize}
-
-\subsubsection{Errors that are constrained by the language}
-\begin{itemize}
-
-\item
-A programmer is not able to store data to the halo region.  If this were
-allowed, one thread could overwrite another threads data at undefined times.
-The compiler is able to catch this class of error.
-
-\item
-A programmer can't make indexing errors in a concurrent routine by going out of
-bounds of the array plus halo memory.  The compiler is able to catch this class of
-error at compile time unless the halo shape is assumed (e.g., HALO(:,:)).
-
-\item
-A programmer is not able to cause race conditions by forgetting to create and
-use temporary arrays properly.  In LOPe it is the comilers responsibility to
-store data in temporary memory.
-
-\item
-A programmer can't make synchronization errors as synchronization is implicit in
-the CONCURRENT attribute.  A thread running a CONCURRENT procedure is provided
-with a copy of it's array element (plus halo) that is consistent with the state
-of memory at the time of invocation of procedure.  Stores to an individual
-thread's array element (by that thread) is never visible to other threads.
-It is possible to relax this restraint with a halo synchronization intrisic.
-However, LOPE encourages the creation of small functions and lets the compiler
-fuse the procedures together and provide the necessary synchronization.
-
-\end{itemize}
-
-Please note that this example is somewhat unfair because in practice CAF is usually
-refactored in a locally oriented way by so that communication and
-synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
-
-This is the normal way that large complex MPI and CAF programs are implemented, i.e., using
-halo cells.  LOPe proposes to formalize this common pattern into the Fortran
-language providing the compiler information in order to spread computation over
-more hardware resources and reduce complexity for the programmer.
-
-\subsection{OpenMP Comparision}
-
-The coarray Fortran example shows how LOPe
+TODO: Give a simple 1D (cartoon) example of connections within a neural network in
+Figure 1.  Describe the size of visual cortex in terms of the number of neurons,
+the number of synaptic connections, and performance.  Reference SC Gorden Bell
+paper from IBM.

### 2015-01-24 papers/LOPe/programming-model.tex
Working on paper layout and program models section primarily.

@@
 \section{Programming Model}
+
+The LOPe (for Locally-Oriented Programming extensions) programming model restricts the programmer ---
+when implementing a stencil algorithm --- to a local view of the array index space.  Within a LOPe
+function, only the local array element (plus a small halo region surrounding the local element) is
+visible to the programmer.  This serves to reduce complexity by removing all boundary conditions and
+processor topology information from the implementation of the algorithm.
+
+%%This reduction in complexity reduces programming errors.
+While developing the convolution example described later in the paper we made an indexing error in applying the 2D stencil loops in the standard serial Fortran test implementation.
+This error required over 3 hours of programming time to repair.
+First the error had to be isolated to the function implementing the convolution (it was first thought to be in the complicated tiff image output routine as the convolution code was ``thought'' to be too simple to wrong.
+Then the index error has to be understood.
+As will be seen, LOPe makes it more difficult to make these errors as Fortran array intrics can be used.
+In some instance, the restricted semantics of LOPe allows the compiler to catch errors (e.g., some errors involving race conditions).
+
+LOPe was implemented as an embedded domain specific language (DSL).  Fortran was chosen as the base
+language for LOPe, although in principle, the same techniques could be applied to language such as C
+or C++ (in conjuction with some form of array extensions such as those proposed by Intel).  Fortran
+was chosen because it already provides a rich array-based syntax.  Readers who are unfamiliar with
+Fortran syntax may wish to consult Appendix A for a brief description of Fortran notation.
+
+\subsection{LOPe Syntax Extensions}
+
+There are only a few syntax additions required for LOPe.  In the code examples describing these
+additions (and throughout this paper as well), DSL syntax additions are highlighted by the usage of
+capitalization for all new keywords.
+
+\subsubsection{HALO.}
+
+The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
+or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
+called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
+an array boundary, so that the array may be referenced using indicies that fall ``outside'' of the
+domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler can
+exploit this knowledge.  For example, a halo region can be explicitly provided in a declaration
+statement of the form,
+\begin{verbatim}
+  real, allocatable, dimension(:), HALO(1:*:1) :: A
+\end{verbatim}
+This statement indicates that array \texttt{A} is one dimensional, will be allocated later when it
+is provided with a specific size, and has a halo region of 1 surrounding the array on both sides.
+The notation \texttt{M:*:N} is to be read as describing a halo region of size \texttt{M} to the
+left, size \texttt{N} to the right, and with an arbitrary size in the middle (denoted by the $*$
+symbol).  When used to describe a formal parameter of a function, the \texttt{HALO} keyword may
+``assumed'' (from the actual argument by the compiler) as is the following statement,
+\begin{verbatim}
+  real, HALO(:,:) :: U
+\end{verbatim}
+which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
+whose size is ``assumed'' by the compiler from the explicit halo size of the actual array argument
+(provided at the calling site of the function).  In LOPe, there is no need in this instance for a
+repetitive \texttt{dimension(:,:)} specification as it is inferred by the \texttt{HALO}
+specification.
+
+\subsubsection{CONCURRENT.}
+
+The second keyword employed by LOPe (although in this case extended) is \texttt{CONCURRENT}.  The
+\texttt{concurrent} keyword already exists in the form of a \texttt{do concurrent} loop construct (a
+Fortran 2008 feature).  By using \texttt{do concurrent}, the programmer asserts that specific
+iterations of the loop body may be executed by the compiler in \emph{any order;} even
+\emph{concurrently.}  For example, the following,
+\begin{verbatim}
+  pure CONCURRENT subroutine Laplace(U)
+     real, HALO(:,:) :: U
+     U(0,0) =                 U(0,+1)               &
+              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
+                          +   U(0,-1)
+  end subroutine Laplace
+\end{verbatim}
+defines a function that can be used as part of an iterative solution of Laplace's equation in two
+dimensions.  A function with the \texttt{pure} and \texttt{CONCURRENT} attributes may be called from
+with a \texttt{do concurrent} loop.  One should imagine that this function is called \emph{for each}
+\texttt{(i,j)} iterate of the loop.
+
+\subsubsection{LOPe index notation.}
+In the \texttt{Laplace} example shown above, the \texttt{(0,0)} array element is the \emph{local} 
+array element and only the local element may be modified.
+
+\input{coarray-comparison}
 
@@
 stencil interior point $A(0,0)$ as shown in the following:
-
-\begin{verbatim}
-  pure CONCURRENT subroutine average(A)
-     real, HALO(:,:) :: A
-
-     A(0,0) =              A(+1,0)                &
-               + A(0,-1) + A( 0,0)   +  A(0,+1)   &
-                         + A(-1,0)
-
-     A(0,0) = (1./5.) * A(0,0)
-
-  end subroutine
-\end{verbatim}
 
@@
 This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.
-It describes what is the compiler's responsibility and which is the user's responsibility.
+It describes what is the compiler's responsibility and what is the programmer's responsibility.
 

### 2015-01-24 papers/LOPe/appendix-a.tex
Initial version.

\section*{Appendix: A Fortran Primer}

This section provides a brief description of Fortran syntax for readers unfamiliar with Fortran syntax.

### 2015-01-24 papers/LOPe/coarray-comparison.tex
Initial version.

\subsection{Comparison to Coarray Fortran}

Below is the motivational example from the original Numrich and Reic paper.
{NOTE: This should sortof/kindof show how it would work between nodes?  Next
section shows OpenMP example for parallelism at the thread level.}

{\small
\begin{verbatim}

subroutine laplace (nrow,ncol,u)
   integer, intent(in) real, intent(inout) real integer
   ! that refers to the current image
   left = me-1
   if (me == 1) left = ncol
   right = me + 1
   if (me == ncol) right = 1
   call sync_all( (/left,right/) ) ! Wait if left and right
   ! have not already reached here
   new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
   call sync_all( (/left,right/) )
   u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
end subroutine laplace

\end{verbatim}
}

Below is the equivalent implementation in LOPe.

{\small
\begin{verbatim}

pure CONCURRENT subroutine laplace(U)
   real, intent(inout), HALO(-1:*:1,-1:*:1) :: U(0:,0:)

   U(0:0) = U(-1,0) + U(+1,0) + U(0,-1) + U(0,+1) - 3*U(0,0)

end subroutine laplace

\end{verbatim}
}

NOTE: Following not a criticism of CAF as CAF is a general purpose parallel
programming language and LOPe only pertains to the halo parallel pattern.

\subsubsection{Improvements}
\begin{itemize}

\item
LOPe requires that the implementation of the algorithm to be separate from the
call to effect the halo transfer.  Removing boundary condition specification
(e.g., the cyclic boundary conditions implemented in the CAF example) from the
algorithm allows the boundary conditions to be changed and without changing
algorithm code.

\item
LOPe applies the transfer of halo memory across multiple (possibly) levels of
memory with the intrinsic {\tt transfer\_halo} function (not shown here).  Thus
the LOPe algorithm can be run on a machine with many interconnected nodes, each
containing hybrid processor cores.  The coarray example can only be run on
multiple nodes (called images in Fortran) without accelerator cores.

\item
The algorithm implementation is separate from user-specified synchronization,
e.g., {\tt call sync\_all}.  In LOPe, synchronization is subsummed in the
semantics of the CONCURRENT attribute and the {\tt transfer\_halo} function.

\item
The algorithm implementation is separate from any specification as to where the
array memory is located.  The CAF example explicity denotes where memory is
located with the {\tt [left]} and {\tt [right]} syntax where left and right
specifiy a processor topology.

\item
The algorithm implementation is separate from any specification as to where the
algorithm is to be executed.  The CAF example explicity denotes where a statement is to
executed with control flow construct like {\tt if (me == 1)}.

\item
The LOPe implementation is easier to understand and frequently follows the
mathematical algorithm directly.  For example, the CAF example adds 4 neighbors
plus the center value to make the implementation with direct remote coarray
access possible, while the LOPe example is able to implement the same algorithm
with fewer operations by adding 4 neighbors (not including the center array
element) and then only subtracting 3 center values.

\item
The semantics of LOPe makes explicit management of array temporaries (e.g., {\tt
  u} and {\tt new\_u} by programmers unnecessary (though still possible).
Because in LOPe the halo region is a language construct, the compiler is better
able to manage temporary buffers than users on the target hardware platform.

\end{itemize}

\subsubsection{Errors that are constrained by the language}
\begin{itemize}

\item
A programmer is not able to store data to the halo region.  If this were
allowed, one thread could overwrite another threads data at undefined times.
The compiler is able to catch this class of error.

\item
A programmer can't make indexing errors in a concurrent routine by going out of
bounds of the array plus halo memory.  The compiler is able to catch this class of
error at compile time unless the halo shape is assumed (e.g., HALO(:,:)).

\item
A programmer is not able to cause race conditions by forgetting to create and
use temporary arrays properly.  In LOPe it is the comilers responsibility to
store data in temporary memory.

\item
A programmer can't make synchronization errors as synchronization is implicit in
the CONCURRENT attribute.  A thread running a CONCURRENT procedure is provided
with a copy of it's array element (plus halo) that is consistent with the state
of memory at the time of invocation of procedure.  Stores to an individual
thread's array element (by that thread) is never visible to other threads.
It is possible to relax this restraint with a halo synchronization intrisic.
However, LOPE encourages the creation of small functions and lets the compiler
fuse the procedures together and provide the necessary synchronization.

\end{itemize}

Please note that this example is somewhat unfair because in practice CAF is usually
refactored in a locally oriented way by so that communication and
synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.

This is the normal way that large complex MPI and CAF programs are implemented, i.e., using
halo cells.  LOPe proposes to formalize this common pattern into the Fortran
language providing the compiler information in order to spread computation over
more hardware resources and reduce complexity for the programmer.

### 2015-01-24 papers/LOPe/coarray-extensions.tex
Initial version.

--- 2015-01-22 papers/LOPe/programming-model.tex
@@
-\section{Programming Model}
-
-\subsection{Locally-Oriented Programming}
+\section{Coarray Fortran Extensions}
 
@@
 
-\begin{verbatim}
-  pure CONCURRENT subroutine average(A)
-     real, HALO(:,:) :: A
-
-     A(0,0) =              A(+1,0)                &
-               + A(0,-1) + A( 0,0)   +  A(0,+1)   &
-                         + A(-1,0)
-
-     A(0,0) = (1./5.) * A(0,0)
-
-  end subroutine
-\end{verbatim}
-
 The
@@
 \begin{verbatim}
-   real, allocatable, HALO(-1:*:1) :: A(:)[:], B(:)[:]
+   real, allocatable, HALO(1:*:1,1:*:1) :: U(:,:)[:]
 
-   !! Allocate local memory
+   !! Allocate normal coarray memory
    !  -------------------------------------------
-   allocate(A(N)[*], B(N)[*])
+   allocate(U(M,N)[*])
 \end{verbatim}
@@
    if (device /= THIS_IMAGE()) then
-      allocate(U1h(0:J)  [device], HALO=(-1:*:1))   [[device]]
-      allocate(U2h(0:J/2)[device], HALO=(-1:*:1))   [[device]]
+      allocate(U[device], HALO_SRC=U)   [[device]]
    end if
@@
 
-   do concurrent (j2h = 1:J/2-1)  [[device]]
-      call restrict(U1h(2*j2h)[device], U2h(j2h)[device])
+   do i = 1, nLoops
+      do concurrent (i=1:M, j=1:N)   [[device]]
+         call Laplace(U(i,j)[device])
+      end do
+      call HALO_TRANSFER(U, BC=CYCLIC)
    end do
 
-   do concurrent (j2h = 1:J/2-1)  [[device]]
-      call restrict(U1h(2*j2h)[device], U2h(j2h)[device], [0.25, 0.5, 0.25])
-   end do
-
-   deallocate(U1h, U2h)
-
+   deallocate(U)
    if (device /= THIS_IMAGE()) then
-      deallocate(U1h[device], U2h[device])   [[device]]
+      deallocate(U[device])   [[device]]
    end if

### 2015-01-24 papers/LOPe/coarray-comparison.tex
First cut at programming-model (mostly all there) stage.

@@
 
-Below is the motivational example from the original Numrich and Reic paper.
-{NOTE: This should sortof/kindof show how it would work between nodes?  Next
-section shows OpenMP example for parallelism at the thread level.}
+LOPe provides a purely \emph{local} viewpoint; the programmer is only provided read and write access
+to the local array element and write access to a small halo region surrounding the local element.
+There is simply \emph{no} mechanism provided for the programmer to even know \emph{where} the local
+element is in the context of the broader array.  On a distributed memory architecture, the halo
+elements may not even be physically located on the same processor.  If executed on a cluster
+containing hybrid processing elements (e.g. GPUs), the halo elements may be as far as three hops
+away: one to get to the host processor and another two to get to memory on the hybrid processor
+executing on another distributed memory node.  LOPe provides a complete separation between algorithm
+development and memory management and synchronization (between memory copies of the same logical
+array region covered by halos).  By explicitly describing the existance and size of an array's halo
+region, the compiler is provided with enough information to manage most of the hard and detailed work
+involved in memory transfer and synchronization.
 
-{\small
-\begin{verbatim}
+Additionally, the semantics of the LOPe execution model (on which more will be said later)
+remove even the \emph{possibility} of race conditions developing during execution of a concurrent
+procedure.
 
-subroutine laplace (nrow,ncol,u)
-   integer, intent(in) real, intent(inout) real integer
-   ! that refers to the current image
-   left = me-1
+We emphasize some of these advantages by compairing the \texttt{Laplace} implementation shown above
+with the implementation of the same algorithm from the original Numrich and Reid paper describing
+coarrays in Fortran.  We should point out that this comparison is somewhat unfair because
+Numrich and Reid were describing the advantages of coarray notation for transferring memory
+on distributed memory architectures, not how ideally to use coarrays within a major code.  But
+this example serves to highlight the advantages of LOPe as described.  In the coarray example
+shown below type declarations have been removed to save space:
+{\small \begin{verbatim}
+subroutine Laplace (nrow,ncol,U)
+   left = me-1     ! me refers to the current image
    if (me == 1) left = ncol
@@
    if (me == ncol) right = 1
-   call sync_all( (/left,right/) ) ! Wait if left and right
-   ! have not already reached here
+   call sync_all( [left,right] ) ! Wait if left and right have
+                                 ! not already reached here
    new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
-   call sync_all( (/left,right/) )
+   call sync_all( [left,right] )
    u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
-end subroutine laplace
+end subroutine
+\end{verbatim}}
 
-\end{verbatim}
-}
+The advantages of LOPe over this example are now described.  Please note the the following
+is not a criticism of Coarray Fortran (CAF) as CAF is a general purpose parallel programming
+language and LOPe only pertains to the halo pattern useful in stencil-based algorithms.
 
-Below is the equivalent implementation in LOPe.
-
-{\small
-\begin{verbatim}
-
-pure CONCURRENT subroutine laplace(U)
-   real, intent(inout), HALO(-1:*:1,-1:*:1) :: U(0:,0:)
-
-   U(0:0) = U(-1,0) + U(+1,0) + U(0,-1) + U(0,+1) - 3*U(0,0)
-
-end subroutine laplace
-
-\end{verbatim}
-}
-
-NOTE: Following not a criticism of CAF as CAF is a general purpose parallel
-programming language and LOPe only pertains to the halo parallel pattern.
-
-\subsubsection{Improvements}
+%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.
+Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
+\subsubsection{LOPe advantages.}
 \begin{itemize}
@@
 \item
-LOPe requires that the implementation of the algorithm to be separate from the
-call to effect the halo transfer.  Removing boundary condition specification
-(e.g., the cyclic boundary conditions implemented in the CAF example) from the
-algorithm allows the boundary conditions to be changed and without changing
-algorithm code.
+LOPe requires that the implementation of the algorithm to be separate from the call to
+effect the halo transfer.  Removing boundary condition specification (e.g., the cyclic
+boundary conditions implemented in the CAF example) from the algorithm allows the boundary
+conditions to be changed and without changing algorithm code.
 
 \item
-LOPe applies the transfer of halo memory across multiple (possibly) levels of
-memory with the intrinsic {\tt transfer\_halo} function (not shown here).  Thus
-the LOPe algorithm can be run on a machine with many interconnected nodes, each
-containing hybrid processor cores.  The coarray example can only be run on
-multiple nodes (called images in Fortran) without accelerator cores.
+LOPe applies the transfer of halo memory across multiple (possibly) levels of memory with
+the LOPe intrinsic \texttt{TRANSFER\_HALO} function (described later).  Thus the LOPe
+algorithm can be run on a machine with many interconnected nodes, each containing hybrid
+processor cores.  The coarray example can only be run on multiple nodes (called images in
+Fortran) without accelerator cores.
 
 \item
-The algorithm implementation is separate from user-specified synchronization,
-e.g., {\tt call sync\_all}.  In LOPe, synchronization is subsummed in the
-semantics of the CONCURRENT attribute and the {\tt transfer\_halo} function.
+The algorithm implementation is separate from user-specified synchronization, e.g.,
+\texttt{call sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the
+\texttt{CONCURRENT} attribute and the \texttt{TRANSFER\_HALO} function described later.
 
 \item
-The algorithm implementation is separate from any specification as to where the
-array memory is located.  The CAF example explicity denotes where memory is
-located with the {\tt [left]} and {\tt [right]} syntax where left and right
-specifiy a processor topology.
+The algorithm implementation is separate from any specification as to where the array
+memory is located.  The CAF example explicity denotes where memory is located with the
+\texttt{[left]} and \texttt{[right]} syntax where left and right specifiy a processor
+topology.
 
 \item
-The algorithm implementation is separate from any specification as to where the
-algorithm is to be executed.  The CAF example explicity denotes where a statement is to
-executed with control flow construct like {\tt if (me == 1)}.
+The algorithm implementation is separate from any specification as to where the algorithm
+is to be executed.  The CAF example explicity denotes where a statement is to executed
+with control flow construct like \texttt{if (me == 1)}.
 
 \item
-The LOPe implementation is easier to understand and frequently follows the
-mathematical algorithm directly.  For example, the CAF example adds 4 neighbors
-plus the center value to make the implementation with direct remote coarray
-access possible, while the LOPe example is able to implement the same algorithm
-with fewer operations by adding 4 neighbors (not including the center array
-element) and then only subtracting 3 center values.
+The LOPe implementation is easier to understand and frequently follows the mathematical
+algorithm directly.  For example, the CAF example adds 4 neighbors plus the center value
+to make the implementation with direct remote coarray access possible, while the LOPe
+example is able to implement the same algorithm with fewer operations by adding 4
+neighbors (not including the center array element) and then only subtracting 3 center
+values.
 
 \item
-The semantics of LOPe makes explicit management of array temporaries (e.g., {\tt
-  u} and {\tt new\_u} by programmers unnecessary (though still possible).
-Because in LOPe the halo region is a language construct, the compiler is better
-able to manage temporary buffers than users on the target hardware platform.
+The semantics of LOPe makes explicit management of array temporaries (e.g., \texttt{U} and
+\texttt{new\_U} by programmers unnecessary (though still possible).  Because in LOPe the
+halo region is a language construct, the compiler is better able to manage temporary
+buffers than users on the target hardware platform.
 
@@
 
-\subsubsection{Errors that are constrained by the language}
+\subsubsection{Errors that are constrained by the language.}
 \begin{itemize}
@@
 \item
-A programmer is not able to store data to the halo region.  If this were
-allowed, one thread could overwrite another threads data at undefined times.
-The compiler is able to catch this class of error.
+A programmer is not able to store data to the halo region.  If this were allowed, one
+thread could overwrite another threads data at undefined times.  The compiler is able to
+catch this class of error.
 
 \item
-A programmer can't make indexing errors in a concurrent routine by going out of
-bounds of the array plus halo memory.  The compiler is able to catch this class of
-error at compile time unless the halo shape is assumed (e.g., HALO(:,:)).
+A programmer can't make indexing errors in a concurrent routine by going out of bounds of
+the array plus halo memory.  The compiler is able to catch this class of error at compile
+time as long as compile-time constants are used to specify halo sizes.
 
@@
 A programmer can't make synchronization errors as synchronization is implicit in
-the CONCURRENT attribute.  A thread running a CONCURRENT procedure is provided
+the \texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided
 with a copy of it's array element (plus halo) that is consistent with the state
 of memory at the time of invocation of procedure.  Stores to an individual
-thread's array element (by that thread) is never visible to other threads.
-It is possible to relax this restraint with a halo synchronization intrisic.
-However, LOPE encourages the creation of small functions and lets the compiler
-fuse the procedures together and provide the necessary synchronization.
+thread's local array element (by that thread) is never visible to other threads.
+LOPE encourages the creation of small functions and lets the compiler
+fuse the procedures together to provide the necessary synchronization.
 
 \end{itemize}
-
-Please note that this example is somewhat unfair because in practice CAF is usually
-refactored in a locally oriented way by so that communication and
-synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
-
-This is the normal way that large complex MPI and CAF programs are implemented, i.e., using
-halo cells.  LOPe proposes to formalize this common pattern into the Fortran
-language providing the compiler information in order to spread computation over
-more hardware resources and reduce complexity for the programmer.

### 2015-01-24 papers/LOPe/programming-model.tex
First cut at programming-model (mostly all there) stage.

@@
 
-The second keyword employed by LOPe (although in this case extended) is \texttt{CONCURRENT}.  The
-\texttt{concurrent} keyword already exists in the form of a \texttt{do concurrent} loop construct (a
-Fortran 2008 feature).  By using \texttt{do concurrent}, the programmer asserts that specific
-iterations of the loop body may be executed by the compiler in \emph{any order;} even
-\emph{concurrently.}  For example, the following,
+The second keyword employed by LOPe is \texttt{CONCURRENT}.  The \texttt{concurrent} keyword already
+exists in the form of a \texttt{do concurrent} loop construct (a Fortran 2008 feature).  By using
+\texttt{do concurrent}, the programmer asserts that specific iterations of the loop body may be
+executed by the compiler in \emph{any order;} even \emph{concurrently.}  For example, the following,
 \begin{verbatim}
@@
 with a \texttt{do concurrent} loop.  One should imagine that this function is called \emph{for each}
-\texttt{(i,j)} iterate of the loop.
+\texttt{(i,j)} iterate of the loop.  An example of this usage will be given later in the text.
 
 \subsubsection{LOPe index notation.}
-In the \texttt{Laplace} example shown above, the \texttt{(0,0)} array element is the \emph{local} 
-array element and only the local element may be modified.
+In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
+array element and only the local element may be modified.  The default zero-based indexing scheme
+for the local array element is somewhat different that normal Fortran notation in that by default,
+array indices start a 1.  For LOPe, this default convention is modified; indices for arrays declared
+with the \texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in
+one dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
+\texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
+the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local) and
+\texttt{U(0,-1)} and \texttt{U(0,+1)} (bottom and top of local).  The geometric positioning of the
+array elements can be seen by the arrangement of the expressions on the right-hand side of the
+\texttt{Laplace} example.
 
@@
 
-\subsection{Locally-Oriented Programming}
-
-%%We highlight the Fortran elemental source-code abstraction because it provides a local orientation similar to that of OpenCL or CUDA kernel functions that were designed to exploit streaming, highly-threaded architectures like GPUs.
-Elemental functions are pure in that they are guaranteed to be free of side effects such as I/O.
-Consider a convolution where a 3x3 filter is applied to individual pixel elements in a photograph to average of blur the original photograph.
-A convolution is similar to stencil operations in computational fluid dynamics that are used to obtain spatial derivatives of state variables.
-Simple code is shown with extentions to Fortran shown in capital letters.
-The extensions could be provided to C or Fortran functions via compiler directives rather than explicit language syntax.
-
-Consider a simple average over a five point stencil in two dimensions.  The algorithm updates the
-stencil interior point $A(0,0)$ as shown in the following:
-
-The
-
-\begin{verbatim}
-   real, allocatable, HALO(-1:*:1) :: A(:)[:], B(:)[:]
-
-   !! Allocate local memory
-   !  -------------------------------------------
-   allocate(A(N)[*], B(N)[*])
-\end{verbatim}
-
-\begin{verbatim}
-   device = get_subimage(1)          ! THIS_IMAGE() if device unavailable
-
-   !! Allocate memory on subimage if available
-   !  ----------------------------------------
-   if (device /= THIS_IMAGE()) then
-      allocate(U1h(0:J)  [device], HALO=(-1:*:1))   [[device]]
-      allocate(U2h(0:J/2)[device], HALO=(-1:*:1))   [[device]]
-   end if
-
-   !! Initialize and transfer memory
-   !  ------------------------------
-   do j1h = 0, J
-      U1h(j1h) = j1h
-   end do
-   U1h[device] = U1h
-
-   do concurrent (j2h = 1:J/2-1)  [[device]]
-      call restrict(U1h(2*j2h)[device], U2h(j2h)[device])
-   end do
-
-   do concurrent (j2h = 1:J/2-1)  [[device]]
-      call restrict(U1h(2*j2h)[device], U2h(j2h)[device], [0.25, 0.5, 0.25])
-   end do
-
-   deallocate(U1h, U2h)
-
-   if (device /= THIS_IMAGE()) then
-      deallocate(U1h[device], U2h[device])   [[device]]
-   end if
-
-end program
-\end{verbatim}
-
-%%where {\tt A} and {\tt Avg} are arrays, but where the {\tt a} argument in the {\tt convolve} definition is declared as a scalar.
-The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.
-The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.
-In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.
-
-\subsection{Memory Management}
-
-This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.
-It describes what is the compiler's responsibility and what is the programmer's responsibility.
-
-%% A note from a conversation with Matt regarding memory management
-%%
-
-%The Fortran language has much tighter restrictions on aliasing than does C.
-%So unless a variable has the pointer or target attribute, it cannot be aliased.
-%Thus the compiler is able to agressively optimized for memory movement between
-%the CPU and accelerator.  However, because the design philosophy of coarrays
-%is that memory transfer between images can be expensive, the programmer must
-%explicitly transfer memory between images with explicit syntax with square
-%bracket notation, i.e., $a[1] = a[2]$.  So we allow the compiler to manage
-%memory within an image but require the use of {\tt halo\_exchange} for the
-%transfer of halo memory between images.
-
-%Fortran currently supports:
-
-%1.
-Array syntax: e.g., C = A + B, where A, B, C are arrays.
-Note that this is implicitly a loop structure, but that no loop indices need be provided.
-Also the programmer need not specify where in memory these arrays reside.
-Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardward threads).
-This is a simple example and it is a research question as to what compiler directives would be useful for memory placement and other directions to the compiler for efficient code generation.
-
-%2.
-Pure procedures:  Fortran has syntax for specifying procedures that have no side effects during execution.
-Specifying code that is side-effect free code is important information to provide to the compiler so that it can generate efficient multi-threaded code.
-
-%3.
-Pure elemental procedures: Fortran has syntax for specifying procedures that take only scalar arguments, but may be applied across array elements.
-Elemental procedures are ideal for writing code to be executed within a hardware thread.
-They resemble OpenCL kernels, but are simpler because they leave all indexing up to the compiler.
-
-%We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.
-This additional syntax is provided by functions that return a copy of a small region of memory surrounding an array element (as seen within an elemental procedure) and with functions for thread synchronization.
-This additional syntax will allow pure procedures to perform stencil and other convolution-like operations on a copy of memory, synchronize, then store the computed results back to the array element associated with the given thread.
-
-%In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.
-This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.
-Fortran also has excellent facilities for interoperability with C so that programming in a mixed language invironment is easily accomplished, including interoperability with native Fortran arrays.
+Note that the use of halo cells is the normal way that large and complex MPI and CAF
+programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
+language thus providing the compiler information in order to spread computation over more
+hardware resources and reduce complexity for the programmer.

### 2015-01-26 papers/LOPe/coarray-extensions.tex
Working on subimage description.

@@
 
-Consider a simple average over a five point stencil in two dimensions.  The algorithm updates the
-stencil interior point $A(0,0)$ as shown in the following:
+Consider the \texttt{Laplace} concurrent function whose implementation was provided above.
+In this section we demonstrate how this function can be called in a context that allows
+full access to all of the enterior elements of the array, as well as array element within
+the halo boundary region.  Topics highlighted in this section are: 1. Distributed memory
+array allocation; 2. The placement or location(s) (possibly multiple) of the memory
+allocated for the distributed array; 3. Remote memory transfer; and 4. Remote execution.
+This description is within the context of a Domain Specific Language (DSL) extending
+Fortran 2008.  These specific extensions are called CAFe, (Coarray Fortran extensions).
 
-The
+We start with the declaration of an array (to be allocated later),
+\begin{verbatim}
+  real, allocatable, HALO(1:*:1,1:*:1), dimension(:,:), codimensions(:,:) :: U
+\end{verbatim}
+with an explicit halo size and with two local dimensions (rank) and two distributed memory
+codimensions (corank).  The corank of the array is chosen to be identical to the rank of
+the array so that the logical process topology aligns is a way that allows a natural
+halo exchange between logically neighboring processes.  For example, if my process
+location is \texttt{[pcol,prow]}, then the right-hand halo for my local array \texttt{U}
+can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol-1,prow]} where the size
+and cosize of \texttt{U} are given by the allocation statement,
+\begin{verbatim}
+  allocate(U(0:M+1,0:N+1)[MP,*])
+\end{verbatim}
 
+This allocation statement specifies that (given the 1 element halo size provided earlier for
+\texttt{U}) the local, left halo is \texttt{U(0,:)}, the right is \texttt{U(M+1,:)},
+the bottom is \texttt{U(:,0)} and the top is \texttt{U(:,N+1)}, leaving the enterior
+region to be \texttt{U(1:M,1:N)}.  In this allocation statement the total number of
+process rows \texttt{NP} can be obtained at runtime and is not a compile time parameter
+because the actual number of participating processes (in Fortran called images) is
+variable, depending on how many processes are requested at startup.  In this discussion,
+it is \emph{assumed} that there are no holes in the logical processor topology, thus
+$MP*NP = P$, where $P$ is the total number of participating processes (images).
+
+So far --- except for the \texttt{HALO} attribute --- this is totally standard Fortran (2008).  For CAFe,
+we extend the concept of Fortran images to be hierachical.  By this we mean that each image \emph{may}
+have a subimage (or subimages), but this subimage is not visible to other normal Fortran images.
+Subimages also execute differently than normal images.  Subimages are task based while images all
+execute a single program but with different (multiple) data (SPMD).  A task can be given to a subimage,
+but execution on the subimage terminates once the task is finished.  Memory on a subimage, however, is
+permanent and must be specifically allocated and deallocated.
+
+One obtains a subimage by executing,
 \begin{verbatim}
-   real, allocatable, HALO(1:*:1,1:*:1) :: U(:,:)[:]
-
-   !! Allocate normal coarray memory
-   !  -------------------------------------------
-   allocate(U(M,N)[*])
+   device = get_subimage(1)          ! THIS_IMAGE() if device unavailable
 \end{verbatim}
@@
 \begin{verbatim}
-   device = get_subimage(1)          ! THIS_IMAGE() if device unavailable
 

### 2015-01-26 papers/LOPe/coarray-extensions.tex
Almost finished with CAFe part.

@@
 we extend the concept of Fortran images to be hierachical.  By this we mean that each image \emph{may}
-have a subimage (or subimages), but this subimage is not visible to other normal Fortran images.
+have a subimage (or subimages), but this subimage is not visible to other regular Fortran images.
 Subimages also execute differently than normal images.  Subimages are task based while images all
 execute a single program but with different (multiple) data (SPMD).  A task can be given to a subimage,
-but execution on the subimage terminates once the task is finished.  Memory on a subimage, however, is
-permanent and must be specifically allocated and deallocated.
+but execution on the subimage terminates once the task is finished.  Memory on a subimage is
+permanent, however, and must be specifically allocated and deallocated.
 
-One obtains a subimage by executing,
+One obtains a subimage by executing a function call,
 \begin{verbatim}
-   device = get_subimage(1)          ! THIS_IMAGE() if device unavailable
+   device = get_subimage(1)
 \end{verbatim}
+where the integer argument representing one of attached hardware devices.  If the function
+fails (e.g., the requested device is unavailable) it returns the image number,
+\texttt{this\_image()}, of the process that is executing the current program.  Returning
+the current image allows program execution to proceed correctly even if there are no
+attached devices.
 
+Once a subimage is obtained, memory on the device may be allocated,
 \begin{verbatim}
-
-   !! Allocate memory on subimage if available
-   !  ----------------------------------------
-   if (device /= THIS_IMAGE()) then
+  if (device /= this_image()) then
     allocate(U[device], HALO_SRC=U)   [[device]]
   end if
+\end{verbatim}
+There are four points to note regarding this memory allocation: 1. Memory is only
+allocated if a subimage has been obtained; 2. The location where memory is allocated is
+denoted by the regular coarray notation \texttt{U[device]}; 3. The allocated size and halo
+attribute of the new array are assumed from the previously allocated local array
+\texttt{U} via the notation \texttt{HALO\_SRC=U} (using \texttt{HALO\_SRC} will also copy
+the memory to the device); and finally 4. The allocation itself is \emph{executed} on the
+subimage device with the notation \texttt{[[device]]}.
 
-   !! Initialize and transfer memory
-   !  ------------------------------
-   do j1h = 0, J
-      U1h(j1h) = j1h
-   end do
-   U1h[device] = U1h
+Fortran uses square bracket notation \texttt{[image]} to specify on what process the
+memory reference is physically located.  Square brackets are a visual clue to the
+programmer that the memory reference may be remote and therefore potentially suffer a
+performance penalty.  CAFe extends this by employing double-bracket notation to indicate
+possibly \emph{remote subimage execution}.
 
-   do i = 1, nLoops
+Execution of the Laplace task (defined previously) is done using the Fortran \texttt{do
+concurrent} construct as shown below,
+\begin{verbatim}
+   do while (.not. converged)
       do concurrent (i=1:M, j=1:N)   [[device]]
@@
    end do
+\end{verbatim}
+There are several points that require highlighting: 1. Iteration occurs over the interior
+of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop occurs on the
+specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
+in any order, even \emph{concurrently}; 4. The \texttt{local element} of the array (as
+described above in reference to the definition of the \texttt{CONCURRENT} procedure
+\texttt{Laplace} is given by the indices \texttt{(i,j)}; 5. Location of memory for the
+task is to be taken from the subimage as noted by \texttt{[device]}; and 6. Transfer of
+all requisite halo regions is effected by the call to the new LOPe intrinsic function
+\texttt{HALO\_TRANSFER()}.
 
+The transfer of halo memory is necessary after each completion of the do concurrent loop.
+Completion of a do concurrent construct indicates that all executing threads (if a threading
+model is used by the compiler) have completed and that all memory on the executing subimage is
+in a consistent state.
+
+Finally, deallocation of memory is similar to memory allocation:
+\begin{verbatim}
    deallocate(U)
-   if (device /= THIS_IMAGE()) then
+   if (device /= this_image()) then
       deallocate(U[device])   [[device]]
    end if
-
-end program
 \end{verbatim}

### 2015-01-27 papers/LOPe/coarray-extensions.tex
Finished second reading of programming model.

@@
 
-The transfer of halo memory is necessary after each completion of the do concurrent loop.
-Completion of a do concurrent construct indicates that all executing threads (if a threading
-model is used by the compiler) have completed and that all memory on the executing subimage is
-in a consistent state.
+Note that a transfer of halo memory is necessary after each completion of the do concurrent loop.
+This must be done in order for the halo region of a coarray on a given process to be consistent
+with the corresponding interior of the coarray on a logically neighboring process.
+
+%% However, recall that with the LOPe programming model only the local element is mutable; thus, the halo region of an array needs update.
 
@@
 
-%%where {\tt A} and {\tt Avg} are arrays, but where the {\tt a} argument in the {\tt convolve} definition is declared as a scalar.
-The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.
+\subsection{Execution Symantics and Memory Management}
+
+This section describes the hierarchical memory layout and how memory consistency between
+the 3-levels of memory is maintained.  It describes what is the compiler's responsibility
+and what is the programmer's responsibility.
+
+%% Completion of a do concurrent construct indicates that all executing threads (if a threading model is used by the compiler) have completed and that all memory on the executing subimage is in a consistent state.
+
+%%The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.
 The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.
 In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.
-
-\subsection{Memory Management}
-
-This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.
-It describes what is the compiler's responsibility and what is the programmer's responsibility.
 
@@
 Fortran also has excellent facilities for interoperability with C so that programming in a mixed language invironment is easily accomplished, including interoperability with native Fortran arrays.
+
+\input{coarray-comparison}
+
+Note that the use of halo cells is the normal way that large and complex MPI and CAF
+programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
+language thus providing the compiler information in order to spread computation over more
+hardware resources and reduce complexity for the programmer.

### 2015-01-27 papers/LOPe/local.tex
First cut at programming-model (mostly all there) stage.
Finished second reading of programming model.

@@
 
-Emerging hybrid accelerator architectures for high performance computing are often well suited for
-usage of a data-parallel programming model.  However, programmers for these emerging architectures
-face a steep learning curve that frequently requires learning a new language, e.g., OpenCL or CUDA,
-or employing OpenMP compiler directives.  Furthermore, the distributed (and frequently multi-level)
-nature of the memory organization of clusters of these architectures provides an additional level of
-complexity for the programmer.  This paper presents preliminary work examining how programming with
-a local orientation can be employed to provide simple access to accelerator architectures.  A
-locally-oriented programming model is especially useful for stencil-based codes or algorithms
-requiring the application of convolution kernels.  In this programming model, a programmer codes the
-algorithm by modifying \emph{only a single array element}, but has read-only access to a small
-sub-array surrounding the given array element, so that a stencil can be applied.  We demonstrate how
-a locally-oriented programming model can be adopted as an embedded, domain-specific language using
-source-to-source program transformations.
+Emerging hybrid accelerator architectures for high performance computing are often suited for the
+use of a data-parallel programming model.  Unfortunately, programmers for these emerging
+architectures face a steep learning curve that frequently requires learning a new language, e.g.,
+OpenCL or CUDA, or employing OpenMP compiler directives.  Furthermore, the distributed (and
+frequently multi-level) nature of the memory organization of clusters of these architectures
+provides an additional level of complexity for the programmer.  This paper presents preliminary work
+examining how programming with a local orientation can be employed to provide simple access to
+accelerator architectures.  A locally-oriented programming model is especially useful for the
+solution of algorithms requiring the application of a convolution kernel.  In this programming
+model, a programmer codes the algorithm by modifying \emph{only a single array element} (called the
+local element), but has read-only access to a small sub-array surrounding the local element so that
+a stencil can be applied.  We demonstrate how a locally-oriented programming model can be adopted as
+an embedded, domain-specific language using source-to-source program transformations.
 
@@
 \input{programming-model}
+\input{coarray-extensions}
 \input{appendix-a}

### 2015-01-27 papers/LOPe/motivation.tex
Finished second reading of programming model.

@@
 was designed to run on massively parallel hardware architectures and a primitive
-version of PetaVision achieved over a Petaflop/s of sustained single precision
+version of PetaVision achieved over a Petaflop of sustained single precision
 performance on Roadrunner, the fastest computer in the world for a brief
@@
 
-TODO: Give a simple 1D (cartoon) example of connections within a neural network in
+TODO: If space allows, give a simple 1D example of connections within a neural network in
 Figure 1.  Describe the size of visual cortex in terms of the number of neurons,
@@
 paper from IBM.
+
+TODO: Explain difficulties encountered.

### 2015-01-27 papers/LOPe/programming-model.tex
Finished second reading of programming model.

@@
 
-The LOPe (for Locally-Oriented Programming extensions) programming model restricts the programmer ---
-when implementing a stencil algorithm --- to a local view of the array index space.  Within a LOPe
-function, only the local array element (plus a small halo region surrounding the local element) is
-visible to the programmer.  This serves to reduce complexity by removing all boundary conditions and
-processor topology information from the implementation of the algorithm.
+The LOPe programming model (for Locally-Oriented Programming extensions) restricts the programmer to
+a local view of the index space of an array.  Within a LOPe function, only a single array element (called
+the local element) is mutable.  In addition, a small halo region surrounding the local element is
+visible to the programmer, but this region is immutable.  Restricting the programmer to a local
+index space serves to reduce complexity by removing all boundary-value and processor-topology
+information from the implementation of the algorithm.
 
@@
 
-There are only a few syntax additions required for LOPe.  In the code examples describing these
-additions (and throughout this paper as well), DSL syntax additions are highlighted by the usage of
-capitalization for all new keywords.
+There are only a few syntax additions required for a LOPe program.  These additions include syntax for
+describing halo regions and concurrent procedures.  In all of the code examples that follow, DSL
+syntax additions are highlighted by the usage of capitalization for keywords that are either new or
+that acquire new usage.
 
-\subsubsection{HALO.}
+\subsubsection{Halo regions.}
 
@@
 called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
-an array boundary, so that the array may be referenced using indicies that fall ``outside'' of the
+an array boundary, so that an array may be referenced using indices that fall ``outside'' of the
 domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler can
-exploit this knowledge.  For example, a halo region can be explicitly provided in a declaration
-statement of the form,
+exploit this knowledge.  For example, a halo region can be declared with a statement of the form,
 \begin{verbatim}
@@
 \end{verbatim}
-This statement indicates that array \texttt{A} is one dimensional, will be allocated later when it
-is provided with a specific size, and has a halo region of 1 surrounding the array on both sides.
-The notation \texttt{M:*:N} is to be read as describing a halo region of size \texttt{M} to the
-left, size \texttt{N} to the right, and with an arbitrary size in the middle (denoted by the $*$
-symbol).  When used to describe a formal parameter of a function, the \texttt{HALO} keyword may
-``assumed'' (from the actual argument by the compiler) as is the following statement,
+This statement indicates that \texttt{A} is a rank one array, will be allocated later (when it will
+be provided with a specific size), and has a halo region of one element surrounding the array on
+either side.  Notation such as \texttt{M:*:N} is to be read as describing a halo region of size
+\texttt{M} to the left, size \texttt{N} to the right, and with an arbitrary size in the middle
+(denoted by the $*$ symbol).  When used to describe a formal parameter of a function, the size may
+be ``assumed'' (denoted by a ``:'' symbol substituted in place of an explicit size) as is the
+following,
 \begin{verbatim}
@@
 which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
-whose size is ``assumed'' by the compiler from the explicit halo size of the actual array argument
-(provided at the calling site of the function).  In LOPe, there is no need in this instance for a
-repetitive \texttt{dimension(:,:)} specification as it is inferred by the \texttt{HALO}
+whose size is obtained by the compiler from the explicit halo size of the actual array argument
+provided at the calling site of the function.  In LOPe, there is no need (in this instance) for a
+repetitive \texttt{dimension(:,:)} specification as it is inferred from the \texttt{HALO}
 specification.
 
-\subsubsection{CONCURRENT.}
+\subsubsection{Concurrent functions.}
 
-The second keyword employed by LOPe is \texttt{CONCURRENT}.  The \texttt{concurrent} keyword already
-exists in the form of a \texttt{do concurrent} loop construct (a Fortran 2008 feature).  By using
-\texttt{do concurrent}, the programmer asserts that specific iterations of the loop body may be
-executed by the compiler in \emph{any order;} even \emph{concurrently.}  For example, the following,
+The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
+of a \texttt{do} \texttt{concurrent} loop construct.  By using this construct, the programmer
+asserts that specific iterations of the loop body may be executed by the compiler in \emph{any
+  order;} even \emph{concurrently.}
+%%LOPe allows the \texttt{concurrent} attribute to be attached to a function declaration so that the function can be called from within a concurrent loop without fear of race conditions.
+For example, the following,
 \begin{verbatim}
@@
 defines a function that can be used as part of an iterative solution of Laplace's equation in two
-dimensions.  A function with the \texttt{pure} and \texttt{CONCURRENT} attributes may be called from
-with a \texttt{do concurrent} loop.  One should imagine that this function is called \emph{for each}
-\texttt{(i,j)} iterate of the loop.  An example of this usage will be given later in the text.
+dimensions.  LOPe allows a function with the \texttt{pure} and \texttt{concurrent} attributes to be
+called from within a \texttt{do} \texttt{concurrent} loop body.  One should imagine that this
+function is called \emph{for each} \texttt{(i,j)} index of the interior of the array \texttt{U}.
+(An example of this usage will be given later in the text.)  Note that this usage introduces a race
+condition as new values of elements of \texttt{U} are created on the left-hand side of the
+assignment statement that may use \emph{new or old} values of \texttt{U} on the right-hand side.
+LOPe requires the compiler to guarantee that race conditions won't occur by using, e.g.,
+double-buffering techniques as needed.
 
@@
 In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
-array element and only the local element may be modified.  The default zero-based indexing scheme
-for the local array element is somewhat different that normal Fortran notation in that by default,
-array indices start a 1.  For LOPe, this default convention is modified; indices for arrays declared
-with the \texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in
-one dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
+array element and only the local element may be modified.  This zero-based indexing for the
+local-array element is different than conventional Fortran, where by default, array indices start at
+1.  For LOPe, this default convention is modified; indices for arrays declared with the
+\texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in one
+dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
 \texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
-the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local) and
-\texttt{U(0,-1)} and \texttt{U(0,+1)} (bottom and top of local).  The geometric positioning of the
-array elements can be seen by the arrangement of the expressions on the right-hand side of the
-\texttt{Laplace} example.
-
-\input{coarray-comparison}
-
-Note that the use of halo cells is the normal way that large and complex MPI and CAF
-programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
-language thus providing the compiler information in order to spread computation over more
-hardware resources and reduce complexity for the programmer.
+the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local, respectively) and
+\texttt{U(0,-1)} and \texttt{U(0,+1)} (bottom and top of local, respectively).  The geometric
+positioning of the array elements can be seen by examining the arrangement of the expressions on the
+right-hand side of the \texttt{Laplace} example.

### 2015-01-27 papers/LOPe/appendix-a.tex
Finished second reading of programming model.
First complete draft of sections 2 and 3.

@@
-\section*{Appendix: A Fortran Primer}
+\section*{Appendix: A Fortran Primer for the C Programmer}
 
-This section provides a brief description of Fortran syntax for readers unfamiliar with Fortran syntax.
+This section provides a brief description of Fortran syntax for readers unfamiliar with Fortran
+syntax.  The most common source of confusion is array notation.  Fortran uses parentheses
+\texttt{()} for both functions arguments and array indexing; thus the confusing expression
+\texttt{foo(i,j)} can either be an array reference or a function call depending on context.  Where C
+reserves square brackets \texttt{[]} to denote an array reference, Fortran uses square brackets to
+reference coarray memory on a remote, distributed memory process (called an image in Fortran).
+
+Otherwise, Fortran syntax is usually straightforward though it may appear strange to the uninitiated
+programmer.  A subroutine is a function with no return value and must be called in a statement
+rather than as an expression.  Simple syntatic elements are: \texttt{!} the line comment symbol;
+\texttt{\&} the line continuation symbol; and \texttt{::} is syntatic sugar used for the separation
+of elements, e.g., separating type attributes from the variable list in a type declaration
+statement.

### 2015-01-27 papers/LOPe/coarray-comparison.tex
First complete draft of sections 2 and 3.

@@
 array region covered by halos).  By explicitly describing the existance and size of an array's halo
-region, the compiler is provided with enough information to manage most of the hard and detailed work
-involved in memory transfer and synchronization.
+region, the compiler is provided with enough information to manage most of the hard and detailed
+work involved in memory transfer and synchronization.  Additionally, the semantics of the LOPe
+execution model remove the possibility of race conditions developing during execution of a
+concurrent procedure.
 
-Additionally, the semantics of the LOPe execution model (on which more will be said later)
-remove even the \emph{possibility} of race conditions developing during execution of a concurrent
-procedure.
-
-We emphasize some of these advantages by compairing the \texttt{Laplace} implementation shown above
-with the implementation of the same algorithm from the original Numrich and Reid paper describing
-coarrays in Fortran.  We should point out that this comparison is somewhat unfair because
-Numrich and Reid were describing the advantages of coarray notation for transferring memory
-on distributed memory architectures, not how ideally to use coarrays within a major code.  But
-this example serves to highlight the advantages of LOPe as described.  In the coarray example
-shown below type declarations have been removed to save space:
+We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
+implementation of the same algorithm from the original Numrich and Reid paper describing coarrays in
+Fortran.  We should point out that this comparison is somewhat unfair, because Numrich and Reid were
+introducing coarray notation for transferring memory on distributed memory architectures, not
+demonstrating how ideally to use coarrays within a large application.
+%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.
+Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
+However this example serves
+to highlight some of the advantages of LOPe and CAFe as introduced above.  Note that in the coarray
+example shown below, type declarations have been removed to save space:
 {\small \begin{verbatim}
@@
     if (me == ncol) right = 1
-   call sync_all( [left,right] ) ! Wait if left and right have
-                                 ! not already reached here
-   new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
+    call sync_all( [left,right] )   ! Wait for left and right
+    new_u(1:nrow) = new_u(1:nrow) + u(1:nrow)[left] + u(1:nrow)[right]
     call sync_all( [left,right] )
@@
 
-The advantages of LOPe over this example are now described.  Please note the the following
-is not a criticism of Coarray Fortran (CAF) as CAF is a general purpose parallel programming
-language and LOPe only pertains to the halo pattern useful in stencil-based algorithms.
+The advantages of LOPe and CAFe over this example are now described.  Please note that the following
+is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and
+LOPe only pertains to the halo pattern useful in stencil-based algorithms.
 
-%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.
-Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
 \subsubsection{LOPe advantages.}
@@
 \item
-LOPe requires that the implementation of the algorithm to be separate from the call to
-effect the halo transfer.  Removing boundary condition specification (e.g., the cyclic
-boundary conditions implemented in the CAF example) from the algorithm allows the boundary
-conditions to be changed and without changing algorithm code.
+LOPe requires the implementation of the algorithm to be separate from the call to effect the halo
+transfer.  Removing boundary condition specification (e.g., the cyclic boundary conditions
+implemented in the CAF example) from the algorithm allows the boundary conditions to be changed
+without changing algorithm code.
 
 \item
-LOPe applies the transfer of halo memory across multiple (possibly) levels of memory with
-the LOPe intrinsic \texttt{TRANSFER\_HALO} function (described later).  Thus the LOPe
-algorithm can be run on a machine with many interconnected nodes, each containing hybrid
-processor cores.  The coarray example can only be run on multiple nodes (called images in
-Fortran) without accelerator cores.
+LOPe applies the transfer of halo memory across possibly multiple levels of memory with the LOPe
+intrinsic \texttt{TRANSFER\_HALO} function.  Thus the LOPe algorithm can be run on a machine with
+many interconnected nodes, each containing hybrid processor cores.  The coarray example can only be
+run on multiple nodes (called images in Fortran) without accelerator cores.
 
 \item
-The algorithm implementation is separate from user-specified synchronization, e.g.,
-\texttt{call sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the
-\texttt{CONCURRENT} attribute and the \texttt{TRANSFER\_HALO} function described later.
+Algorithm implementation is separate from user-specified synchronization, e.g., \texttt{call
+  sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the \texttt{CONCURRENT}
+attribute and the \texttt{TRANSFER\_HALO}.
 
@@
 \item
-The semantics of LOPe makes explicit management of array temporaries (e.g., \texttt{U} and
-\texttt{new\_U} by programmers unnecessary (though still possible).  Because in LOPe the
+The semantics of LOPe makes explicit management of array temporaries (e.g., \texttt{u} and
+\texttt{new\_u} by the programmer unnecessary (though still possible).  Because in LOPe the
 halo region is a language construct, the compiler is better able to manage temporary
@@
 \item
-A programmer is not able to store data to the halo region.  If this were allowed, one
-thread could overwrite another threads data at undefined times.  The compiler is able to
-catch this class of error.
+A programmer is not able to store data to the halo region during execution of a LOPe concurrent
+function.  If this were allowed, one thread could overwrite another threads data at undefined times.
+The compiler is able to catch this class of error.
 
 \item
-A programmer can't make indexing errors in a concurrent routine by going out of bounds of
-the array plus halo memory.  The compiler is able to catch this class of error at compile
-time as long as compile-time constants are used to specify halo sizes.
+A programmer can't make indexing errors in a concurrent routine by going out of bounds of the array
+plus halo memory.  The compiler is able to catch this class of error at compile time as long as
+compile-time constants are used to specify halo sizes.
 
 \item
-A programmer is not able to cause race conditions by forgetting to create and
-use temporary arrays properly.  In LOPe it is the comilers responsibility to
-store data in temporary memory.
+A programmer is not able to cause race conditions by forgetting to create and use temporary arrays
+properly.  In LOPe it is the comilers responsibility to store data in temporary memory.
 
 \item
-A programmer can't make synchronization errors as synchronization is implicit in
-the \texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided
-with a copy of it's array element (plus halo) that is consistent with the state
-of memory at the time of invocation of procedure.  Stores to an individual
-thread's local array element (by that thread) is never visible to other threads.
-LOPE encourages the creation of small functions and lets the compiler
-fuse the procedures together to provide the necessary synchronization.
+A programmer can't make synchronization errors as synchronization is implicit in the
+\texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided with a copy of
+it's array element (plus halo) that is consistent with the state of memory \emph{at the time of
+  invocation of the procedure}.  Stores to an individual thread's local array element (by that
+thread) is never visible to other threads.  LOPE encourages the creation of small functions and lets
+the compiler fuse the functions together to improve performance and to provide necessary
+synchronization.
 

### 2015-01-27 papers/LOPe/coarray-extensions.tex
First complete draft of sections 2 and 3.

@@
 
-Consider the \texttt{Laplace} concurrent function whose implementation was provided above.
-In this section we demonstrate how this function can be called in a context that allows
-full access to all of the enterior elements of the array, as well as array element within
-the halo boundary region.  Topics highlighted in this section are: 1. Distributed memory
-array allocation; 2. The placement or location(s) (possibly multiple) of the memory
-allocated for the distributed array; 3. Remote memory transfer; and 4. Remote execution.
-This description is within the context of a Domain Specific Language (DSL) extending
-Fortran 2008.  These specific extensions are called CAFe, (Coarray Fortran extensions).
+Consider the \texttt{Laplace} concurrent function discussed above.  In this section we demonstrate
+how this function can be called in a context that allows full access to all of the interior elements
+of the array, as well as array elements within the logically exterior, halo boundary region.  Topics
+highlighted in this section are:
+1. Distributed memory array allocation;
+2. Explicit memory placement;
+3. Remote memory transfer; and
+4. Remote execution.
+This description is within the context of a Domain Specific Language (DSL), again extending Fortran
+2008.  As shorthand, these specific extensions are referred to as CAFe, for Coarray Fortran extensions.
+
+\subsection{CAFe Example}
 
@@
 \begin{verbatim}
-  real, allocatable, HALO(1:*:1,1:*:1), dimension(:,:), codimensions(:,:) :: U
+  real, allocatable, HALO(1:*:1,1:*:1),          &
+        dimension(:,:), codimensions(:,:) :: U
 \end{verbatim}
@@
 location is \texttt{[pcol,prow]}, then the right-hand halo for my local array \texttt{U}
-can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol-1,prow]} where the size
+can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size
 and cosize of \texttt{U} are given by the allocation statement,
 \begin{verbatim}
-  allocate(U(0:M+1,0:N+1)[MP,*])
+  allocate( U(0:M+1,0:N+1)[MP,*] )
 \end{verbatim}
 
-This allocation statement specifies that (given the 1 element halo size provided earlier for
-\texttt{U}) the local, left halo is \texttt{U(0,:)}, the right is \texttt{U(M+1,:)},
-the bottom is \texttt{U(:,0)} and the top is \texttt{U(:,N+1)}, leaving the enterior
-region to be \texttt{U(1:M,1:N)}.  In this allocation statement the total number of
-process rows \texttt{NP} can be obtained at runtime and is not a compile time parameter
-because the actual number of participating processes (in Fortran called images) is
-variable, depending on how many processes are requested at startup.  In this discussion,
-it is \emph{assumed} that there are no holes in the logical processor topology, thus
-$MP*NP = P$, where $P$ is the total number of participating processes (images).
+This allocation statement specifies (given the one element halo size provided earlier for
+\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the
+bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the enterior region
+\texttt{U(1:M,1:N)}.  In this allocation statement, the total number of process rows \texttt{NP} can
+be obtained at runtime, but may not be explicitly provided (according to CAF rules) because the
+actual number of participating processes (in Fortran called images) is variable, depending on how
+many processes are requested at startup.  In this discussion, it is \emph{assumed} that there are no
+holes in the logical processor topology, thus $MP*NP = P$, where $P$ is the total number of
+participating processes (images).
 
@@
 \end{verbatim}
-There are four points to note regarding this memory allocation: 1. Memory is only
-allocated if a subimage has been obtained; 2. The location where memory is allocated is
-denoted by the regular coarray notation \texttt{U[device]}; 3. The allocated size and halo
-attribute of the new array are assumed from the previously allocated local array
-\texttt{U} via the notation \texttt{HALO\_SRC=U} (using \texttt{HALO\_SRC} will also copy
-the memory to the device); and finally 4. The allocation itself is \emph{executed} on the
-subimage device with the notation \texttt{[[device]]}.
+There are four points to note regarding this memory allocation: 1. Memory is only allocated if a
+subimage has been obtained; 2. The location where memory is allocated is denoted by regular coarray
+notation \texttt{U[device]}; 3. The allocated size and halo attribute of the new array are assumed
+from the previously allocated local array \texttt{U} via the notation \texttt{HALO\_SRC=U} (using
+\texttt{HALO\_SRC} will also initially copy \texttt{U} to the subimage); and finally 4. The
+allocation itself is \emph{executed} on the subimage device with the notation \texttt{[[device]]}.
 
@@
 
-Execution of the Laplace task (defined previously) is done using the Fortran \texttt{do
-concurrent} construct as shown below,
+Execution of the \texttt{Laplace} task (defined previously) is done using the Fortran \texttt{do}
+\texttt{concurrent} construct as shown,
 \begin{verbatim}
@@
       do concurrent (i=1:M, j=1:N)   [[device]]
-         call Laplace(U(i,j)[device])
+         call Laplace( U(i,j)[device] )
       end do
@@
 There are several points that require highlighting: 1. Iteration occurs over the interior
-of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop occurs on the
+of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
 specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
-in any order, even \emph{concurrently}; 4. The \texttt{local element} of the array (as
-described above in reference to the definition of the \texttt{CONCURRENT} procedure
-\texttt{Laplace} is given by the indices \texttt{(i,j)}; 5. Location of memory for the
-task is to be taken from the subimage as noted by \texttt{[device]}; and 6. Transfer of
+in any order, even \emph{concurrently}; 4. The local element of the array (as
+defined above in reference to the definition of the concurrent procedure
+\texttt{Laplace}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
+task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
+execution of the loop body before further execution of the program proceeds; and 6. Transfer of
 all requisite halo regions is effected by the call to the new LOPe intrinsic function
-\texttt{HALO\_TRANSFER()}.
+\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
+complete the halo transfer before program execution continues.
 
@@
 
-Finally, deallocation of memory is similar to memory allocation:
+Finally, memory for the entire array \texttt{U} can be copied from the subimage device
+with the statement,
+\begin{verbatim}
+  U = U[device]
+\end{verbatim}
+add memory deallocation is similar to memory allocation:
 \begin{verbatim}
@@
 
-\subsection{Execution Symantics and Memory Management}
+%%\subsection{Execution Symantics and Memory Management}
 
-This section describes the hierarchical memory layout and how memory consistency between
-the 3-levels of memory is maintained.  It describes what is the compiler's responsibility
-and what is the programmer's responsibility.
+%%This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.
+It describes what is the compiler's responsibility and what is the programmer's responsibility.
 
@@
 programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
-language thus providing the compiler information in order to spread computation over more
-hardware resources and reduce complexity for the programmer.
+language allowing the compiler access to this information in order to spread computation over more
+hardware resources, improve performance, and to reduce complexity for the programmer.

### 2015-01-27 papers/LOPe/appendix-a.tex
Completion for coauthors review of sections 2 and 3.

@@
 \texttt{\&} the line continuation symbol; and \texttt{::} is syntatic sugar used for the separation
-of elements, e.g., separating type attributes from the variable list in a type declaration
+of syntatic elements, e.g., separating type attributes from the variable list in a type declaration
 statement.

### 2015-01-27 papers/LOPe/coarray-comparison.tex
Completion for coauthors review of sections 2 and 3.

@@
 executing on another distributed memory node.  LOPe provides a complete separation between algorithm
-development and memory management and synchronization (between memory copies of the same logical
+development and memory management (synchronization between memory copies of the same logical
 array region covered by halos).  By explicitly describing the existance and size of an array's halo
@@
 We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
-implementation of the same algorithm from the original Numrich and Reid paper describing coarrays in
-Fortran.  We should point out that this comparison is somewhat unfair, because Numrich and Reid were
-introducing coarray notation for transferring memory on distributed memory architectures, not
-demonstrating how ideally to use coarrays within a large application.
+implementation of the same algorithm from the original Numrich and Reid paper first describing
+coarrays in Fortran.  We should point out that this comparison is somewhat unfair, because Numrich
+and Reid were introducing coarray notation for transferring memory on distributed memory
+architectures, not demonstrating how ideally one should use coarrays within a large application.
 %%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.
@@
 is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and
-LOPe only pertains to the halo pattern useful in stencil-based algorithms.
+LOPe only pertains to the halo pattern most useful in the implementation of stencil-based algorithms.
 
@@
 many interconnected nodes, each containing hybrid processor cores.  The coarray example can only be
-run on multiple nodes (called images in Fortran) without accelerator cores.
+run on multiple nodes without accelerator cores.
 
@@
   sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the \texttt{CONCURRENT}
-attribute and the \texttt{TRANSFER\_HALO}.
+attribute and the \texttt{TRANSFER\_HALO} function call.
 
@@
 \item
-The LOPe implementation is easier to understand and frequently follows the mathematical
-algorithm directly.  For example, the CAF example adds 4 neighbors plus the center value
-to make the implementation with direct remote coarray access possible, while the LOPe
-example is able to implement the same algorithm with fewer operations by adding 4
-neighbors (not including the center array element) and then only subtracting 3 center
-values.
+The LOPe implementation is easier to understand and frequently follows the mathematical algorithm
+directly.  For example, the CAF implementation of Numrich and Reid adds 4 neighbors plus the center
+value to make the implementation with direct remote coarray access possible, while the LOPe example
+is able to implement the same algorithm with one statement and no intervening synchronization.
 
@@
 A programmer is not able to cause race conditions by forgetting to create and use temporary arrays
-properly.  In LOPe it is the comilers responsibility to store data in temporary memory.
+properly.  In LOPe it is the compilers responsibility to store data in temporary memory.
 

### 2015-01-27 papers/LOPe/coarray-extensions.tex
Completion for coauthors review of sections 2 and 3.

@@
 Consider the \texttt{Laplace} concurrent function discussed above.  In this section we demonstrate
-how this function can be called in a context that allows full access to all of the interior elements
-of the array, as well as array elements within the logically exterior, halo boundary region.  Topics
-highlighted in this section are:
-1. Distributed memory array allocation;
-2. Explicit memory placement;
-3. Remote memory transfer; and
-4. Remote execution.
-This description is within the context of a Domain Specific Language (DSL), again extending Fortran
-2008.  As shorthand, these specific extensions are referred to as CAFe, for Coarray Fortran extensions.
+how this function can be called in the normal context of a program, one that allows full access to
+all of the interior elements of the array, as well as array elements within the logically exterior,
+halo-boundary region.  Topics highlighted in this section are: 1. Distributed memory array
+allocation; 2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
+description is within the context of a Domain Specific Language (DSL), again extending Fortran 2008.
+As shorthand, these specific extensions are referred to as CAFe, for Coarray Fortran extensions.
 
@@
   real, allocatable, HALO(1:*:1,1:*:1),          &
-        dimension(:,:), codimensions(:,:) :: U
+        dimension(:,:), codimension(:,:) :: U
 \end{verbatim}
@@
 codimensions (corank).  The corank of the array is chosen to be identical to the rank of
-the array so that the logical process topology aligns is a way that allows a natural
+the array so that the logical process topology aligns in a way that allows a natural
 halo exchange between logically neighboring processes.  For example, if my process
@@
 \texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the
-bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the enterior region
+bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the interior region
 \texttt{U(1:M,1:N)}.  In this allocation statement, the total number of process rows \texttt{NP} can
-be obtained at runtime, but may not be explicitly provided (according to CAF rules) because the
+be obtained at runtime, but \emph{may not} be explicitly provided (according to CAF rules) because the
 actual number of participating processes (in Fortran called images) is variable, depending on how
@@
 where the integer argument representing one of attached hardware devices.  If the function
-fails (e.g., the requested device is unavailable) it returns the image number,
-\texttt{this\_image()}, of the process that is executing the current program.  Returning
+fails (e.g., the requested device is unavailable) it returns the image number
+\texttt{this\_image()} of the process that is executing the current program.  Returning
 the current image allows program execution to proceed correctly even if there are no
@@
 
-Fortran uses square bracket notation \texttt{[image]} to specify on what process the
+Fortran uses square bracket notation, e.g. \texttt{[image]}, to specify on what process the
 memory reference is physically located.  Square brackets are a visual clue to the
@@
 
-Execution of the \texttt{Laplace} task (defined previously) is done using the Fortran \texttt{do}
-\texttt{concurrent} construct as shown,
+Execution of the \texttt{Laplace} task is done using the Fortran \texttt{do}
+\texttt{concurrent} construct:
 \begin{verbatim}

### 2015-01-27 papers/LOPe/local.tex
Completion for coauthors review of sections 2 and 3.

@@
 
-\author{Craig Rasmussen%
-\thanks{The authors would like to thank Robert Robey and Wayne Weseloh
-at Los Alamos National Laborabory for several stimulating
-conversations with respect to programming models.}%
+\author{Craig Rasmussen
 \and Matthew Sottile\and Dan Nagle\and Soren Rasmussen}
@@
 
-\subsection{Acknowledgments}
-This work was supported in part by the Department of Energy Office of Science,
-Advanced Scientific Computing Research.
+\section*{Acknowledgments}
+This work was supported in part by the Department of Energy Office of Science, Advanced Scientific
+Computing Research.  The authors would also like to thank Robert Robey and Wayne Weseloh at Los
+Alamos National Laborabory for several stimulating conversations with respect to programming models.
 

### 2015-01-27 papers/LOPe/programming-model.tex
Completion for coauthors review of sections 2 and 3.

@@
 visible to the programmer, but this region is immutable.  Restricting the programmer to a local
-index space serves to reduce complexity by removing all boundary-value and processor-topology
-information from the implementation of the algorithm.
+index space serves to reduce complexity by separating all boundary-value and processor-topology
+concerns from the implementation of the algorithm.
 
@@
 was chosen because it already provides a rich array-based syntax.  Readers who are unfamiliar with
-Fortran syntax may wish to consult Appendix A for a brief description of Fortran notation.
+Fortran syntax may wish to consult Appendix A, for a brief description of Fortran notation.
 
@@
 
-\subsubsection{Halo regions.}
-
+\textbf{Halo regions.}
 The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
@@
 an array boundary, so that an array may be referenced using indices that fall ``outside'' of the
-domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler can
-exploit this knowledge.  For example, a halo region can be declared with a statement of the form,
+logical domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler
+can exploit this knowledge.  For example, a halo region can be declared with a statement of the form,
 \begin{verbatim}
@@
 be provided with a specific size), and has a halo region of one element surrounding the array on
-either side.  Notation such as \texttt{M:*:N} is to be read as describing a halo region of size
-\texttt{M} to the left, size \texttt{N} to the right, and with an arbitrary size in the middle
-(denoted by the $*$ symbol).  When used to describe a formal parameter of a function, the size may
-be ``assumed'' (denoted by a ``:'' symbol substituted in place of an explicit size) as is the
-following,
+either side.  Notation such as \texttt{M:*:N} is to be read as describing a halo region of
+\texttt{M} elements to the left, \texttt{N} elements to the right, and with an arbitrary number of
+elements in the middle (denoted by the $*$ symbol).  When used to describe a formal parameter of a
+function, the halo size may be ``assumed'' (denoted by a ``:'' symbol substituted in place of an
+explicit size) as is the following,
 \begin{verbatim}
@@
 provided at the calling site of the function.  In LOPe, there is no need (in this instance) for a
-repetitive \texttt{dimension(:,:)} specification as it is inferred from the \texttt{HALO}
+repetitive \texttt{dimension(:,:)} specification, as it is inferred from the \texttt{HALO}
 specification.
 
-\subsubsection{Concurrent functions.}
-
+\textbf{Concurrent functions.}
 The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
@@
 
-\subsubsection{LOPe index notation.}
+\textbf{LOPe index notation.}
 In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
@@
 the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local, respectively) and
-\texttt{U(0,-1)} and \texttt{U(0,+1)} (bottom and top of local, respectively).  The geometric
+\texttt{U(0,-1)} and \texttt{U(0,+1)} (below and above of local, respectively).  The geometric
 positioning of the array elements can be seen by examining the arrangement of the expressions on the

### 2015-01-28 papers/LOPe/coarray-comparison.tex
Changes as suggested by Daniel.

@@
 LOPe provides a purely \emph{local} viewpoint; the programmer is only provided read and write access
-to the local array element and write access to a small halo region surrounding the local element.
+to the local array element and read access to a small halo region surrounding the local element.
 There is simply \emph{no} mechanism provided for the programmer to even know \emph{where} the local
@@
 \texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided with a copy of
-it's array element (plus halo) that is consistent with the state of memory \emph{at the time of
+it's local array element plus halo that is consistent with the state of memory \emph{at the time of
   invocation of the procedure}.  Stores to an individual thread's local array element (by that

### 2015-01-28 papers/LOPe/coarray-extensions.tex
Changes as suggested by Daniel.

@@
   real, allocatable, HALO(1:*:1,1:*:1),          &
-        dimension(:,:), codimension(:,:) :: U
+        dimension(:,:), codimension[:,:] :: U
 \end{verbatim}
@@
 the array so that the logical process topology aligns in a way that allows a natural
-halo exchange between logically neighboring processes.  For example, if my process
-location is \texttt{[pcol,prow]}, then the right-hand halo for my local array \texttt{U}
+halo exchange between logically neighboring processes.  For example, if the process
+location is \texttt{[pcol,prow]}, then the right-hand halo for the local array \texttt{U}
 can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size
@@
 actual number of participating processes (in Fortran called images) is variable, depending on how
-many processes are requested at startup.  In this discussion, it is \emph{assumed} that there are no
+many processes are requested at program startup.  In this discussion, it is \emph{assumed} that there are no
 holes in the logical processor topology, thus $MP*NP = P$, where $P$ is the total number of

### 2015-01-31 papers/LOPe/code/tests/halos_2d.h
Normalizing to 3D version.

@@ -33,7 +33,7 @@
 #define HALO2(NAME,SIDE)   ( (NAME ## _H2 ## SIDE ## _) )
 
 // 1D index for the array NAME and offset OFF
-#define IDX1(NAME,OFF)     ( (OFF) + (K1_) + (NAME ## _H1L_) )
+#define IDX1(NAME,OFF1)    ( (OFF1) + (K1_) + (NAME ## _H1L_) )
 
 // 2D index for the array NAME and offset OFF
 #define IDX2(NAME,OFF1,OFF2) (  ((OFF1) + (K1_) + (NAME ## _H1L_)) * (NAME ## _S1_) \

### 2015-01-31 papers/LOPe/code/tests/elemental_add_2d.f90
Fixing bug found when implementing 3D version.

@@ -18,7 +18,7 @@ program elemental_add_2d
    integer(c_size_t), parameter :: NHY = 1
 
    integer(c_size_t), parameter ::  SIZE_FLOAT = 4
-   integer(c_size_t), parameter :: N_HALO_ELEM = NX*(2*NHX) + NY*(2*NHY)
+   integer(c_size_t), parameter :: N_HALO_ELEM = (2*NHX)*NY + NX*(2*NHY)
    integer(c_size_t), parameter ::   HALO_SIZE = N_HALO_ELEM*SIZE_FLOAT
 
    ! work group size

### 2015-01-31 papers/LOPe/code/tests/elemental_add_3d.f90
Initial version.

@@ -1,0 +1,169 @@
+program elemental_add_3d
+!...TODO-GENERATE
+   use ForOpenCL
+   use Parallel_Halo , only : Context, Parallel_Start, Parallel_End, Parallel_Topology
+   use Parallel      , only : Halo_Exchange3D
+!...TODO-END-GENERATE
+   use Timer_mod
+   implicit none
+
+   ! layer size
+   integer(c_size_t), parameter :: NX  = 16
+   integer(c_size_t), parameter :: NY  = 4
+   integer(c_size_t), parameter :: NZ  = 4
+
+!...TODO-GENERATE
+   integer :: cl_status_
+
+   integer(c_size_t), parameter :: NHX = 1
+   integer(c_size_t), parameter :: NHY = 1
+   integer(c_size_t), parameter :: NHZ = 1
+
+   integer(c_size_t), parameter ::  SIZE_FLOAT = 4
+   integer(c_size_t), parameter :: N_HALO_ELEM = (2*NHX)*NY*NZ + NX*(2*NHY)*NZ + NX*NY*(2*NHZ)
+   integer(c_size_t), parameter ::   HALO_SIZE = N_HALO_ELEM*SIZE_FLOAT
+
+   ! work group size
+   integer(c_size_t), parameter :: NXL = 32
+   integer(c_size_t), parameter :: NYL = 8
+   integer(c_size_t), parameter :: NZL = 8
+   integer(c_size_t) :: nxGlobal, nyGlobal, nzGlobal, nxLocal, nyLocal, nzLocal
+
+   Type(Context) :: aContext
+   Integer       :: ndims, dims(3)
+!...TODO-END-GENERATE
+
+   real(c_float), target, allocatable, dimension(:,:,:) :: A, B, C
+
+   type(CPUTimer) :: timer
+   real(c_double) :: h_time
+
+   integer :: device, i, j, k, nLoops = 1, nWarm = 1
+
+!...TODO-GENERATE
+   real(c_float), allocatable :: in_C_H_(:), out_C_H_(:)
+
+   type(CLDevice) :: cl_device_
+   type(CLKernel) :: kernel
+   type(CLBuffer) :: cl_A_, cl_B_, cl_C_
+   type(CLBuffer) :: cl_C_H_
+
[…]

### 2015-01-31 papers/LOPe/local.tex
Changes as suggested by Daniel.
Fixed Dan's email address.
Slight modification to keyword list.

@@
 \urldef{\mailsa}\path|rasmus@uoregon.edu, mjsottile@computer.org,|
-\urldef{\mailsb}\path|dannagle@verizon.net, soren.rasmussen@aggiemail.usu.edu|    
+\urldef{\mailsb}\path|dangle@ucar.edu, soren.rasmussen@aggiemail.usu.edu|
 \newcommand{\keywords}[1]{\par\addvspace\baselineskip
@@
 \author{Craig Rasmussen
-\and Matthew Sottile\and Dan Nagle\and Soren Rasmussen}
+\and Matthew Sottile\and Daniel Nagle\and Soren Rasmussen}
 
@@
 
-\keywords{domain specific language, stencil compilers, distributed memory parallelism}
+\keywords{domain specific language, stencil compiler, distributed memory parallelism}
 \end{abstract}

### 2015-01-31 papers/LOPe/code/tests/halos_3d.h
Initial version.

@@ -1,0 +1,58 @@
+#define K1_ (get_global_id(0))
+#define K2_ (get_global_id(1))
+#define K3_ (get_global_id(2))
+
+#define N1_ (16)
+#define N2_ (4)
+#define N3_ (4)
+
+#define A_H1L_ 1
+#define A_H1R_ 1
+#define A_H2L_ 1
+#define A_H2R_ 1
+#define A_H3L_ 1
+#define A_H3R_ 1
+#define A_S1_  1
+#define A_S2_  (A_H1L_ + A_H1R_ + N1_)
+#define A_S3_  ?
+
+#define B_H1L_ 1
+#define B_H1R_ 1
+#define B_H2L_ 1
+#define B_H2R_ 1
+#define B_H3L_ 1
+#define B_H3R_ 1
+#define B_S1_  1
+#define B_S2_  (B_H1L_ + B_H1R_ + N1_)
+#define B_S3_  ?
+
+#define C_H1L_ 1
+#define C_H1R_ 1
+#define C_H2L_ 1
+#define C_H2R_ 1
+#define C_H3L_ 1
+#define C_H3R_ 1
+#define C_S1_  1
+#define C_S2_  (C_H1L_ + C_H1R_ + N1_)
+#define C_S3_  ?
+
+// First dimension (Left/Right) halo size for the given SIDE of the array NAME
+#define HALO1(NAME,SIDE)   ( (NAME ## _H1 ## SIDE ## _) )
+
+// Second dimension (Bottom/Top) halo size for the given SIDE of the array NAME
+#define HALO2(NAME,SIDE)   ( (NAME ## _H2 ## SIDE ## _) )
+
+// Third dimension (Front/Back) halo size for the given SIDE of the array NAME
+#define HALO2(NAME,SIDE)   ( (NAME ## _H3 ## SIDE ## _) )
+
+// 1D index for the array NAME and offset OFF
+#define IDX1(NAME,OFF1)    ( (OFF1) + (K1_) + (NAME ## _H1L_) )
+
+// 2D index for the array NAME and offset OFF
+#define IDX2(NAME,OFF1,OFF2) (  ((OFF1) + (K1_) + (NAME ## _H1L_)) * (NAME ## _S1_) \
+                              + ((OFF2) + (K2_) + (NAME ## _H2L_)) * (NAME ## _S2_)  )
+
+// 3D index for the array NAME and offset OFF
[…]

### 2015-01-31 papers/LOPe/coarray-comparison.tex
first pass of edits

@@
 \item
-A programmer is not able to store data to the halo region during execution of a LOPe concurrent
-function.  If this were allowed, one thread could overwrite another threads data at undefined times.
-The compiler is able to catch this class of error.
+A programmer is not able to store data to the halo region during
+execution of a LOPe concurrent function.  If this were allowed, one
+thread could overwrite another threads data at undefined times.  The
+compiler is able to catch this class of data race error.
 
@@
 \item
-A programmer can't make synchronization errors as synchronization is implicit in the
-\texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided with a copy of
-it's local array element plus halo that is consistent with the state of memory \emph{at the time of
-  invocation of the procedure}.  Stores to an individual thread's local array element (by that
-thread) is never visible to other threads.  LOPE encourages the creation of small functions and lets
-the compiler fuse the functions together to improve performance and to provide necessary
+A programmer can't make synchronization errors as synchronization is
+implicit in the \texttt{CONCURRENT} attribute\todo{clarify what sync
+  errors are prevented.  restricted concurrency model === less room
+  for errors.}.  A thread running a concurrent procedure is provided
+with a copy of it's local array element plus halo that is consistent
+with the state of memory \emph{at the time of invocation of the
+  procedure}.  Stores to an individual thread's local array element
+(by that thread) is never visible to other threads.  LOPE encourages
+the creation of small functions and lets the compiler fuse the
+functions together to improve performance and to provide necessary
 synchronization.

### 2015-01-31 papers/LOPe/coarray-extensions.tex
first pass of edits

@@
 
-Consider the \texttt{Laplace} concurrent function discussed above.  In this section we demonstrate
-how this function can be called in the normal context of a program, one that allows full access to
-all of the interior elements of the array, as well as array elements within the logically exterior,
-halo-boundary region.  Topics highlighted in this section are: 1. Distributed memory array
-allocation; 2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
-description is within the context of a Domain Specific Language (DSL), again extending Fortran 2008.
-As shorthand, these specific extensions are referred to as CAFe, for Coarray Fortran extensions.
+Consider the \texttt{Laplace} concurrent function discussed above.  In
+this section we demonstrate how this function can be called in the
+normal context of a program, one that allows full access to all of the
+interior elements of the array, as well as array elements within the
+logically exterior, halo-boundary region.  Topics highlighted in this
+section are: 1. Distributed memory array allocation; 2. Explicit
+memory placement; 3. Remote memory transfer; and 4. Remote execution.
+This description is within the context of an extension of the Fortran
+2008 language.  As shorthand, these specific extensions are referred
+to as CAFe, for Coarray Fortran extensions.
 
@@
 \end{verbatim}
-with an explicit halo size and with two local dimensions (rank) and two distributed memory
-codimensions (corank).  The corank of the array is chosen to be identical to the rank of
-the array so that the logical process topology aligns in a way that allows a natural
-halo exchange between logically neighboring processes.  For example, if the process
-location is \texttt{[pcol,prow]}, then the right-hand halo for the local array \texttt{U}
-can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size
-and cosize of \texttt{U} are given by the allocation statement,
+
+with an explicit halo size and with two local dimensions (rank) and
+two distributed memory codimensions (corank).  The corank of the array
+is chosen to be identical to the rank of the array so that the logical
+process topology aligns in a way that allows a natural halo exchange
+between logically neighboring processes.\todo{What if corank and rank are not the same?}  For example, if the process
+location is \texttt{[pcol,prow]}, then the right-hand halo for the
+local array \texttt{U} can be obtained by the assignment
+\texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
+\texttt{U} are given by the allocation statement,
+
 \begin{verbatim}
@@
 
-This allocation statement specifies (given the one element halo size provided earlier for
-\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the
-bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the interior region
-\texttt{U(1:M,1:N)}.  In this allocation statement, the total number of process rows \texttt{NP} can
-be obtained at runtime, but \emph{may not} be explicitly provided (according to CAF rules) because the
-actual number of participating processes (in Fortran called images) is variable, depending on how
-many processes are requested at program startup.  In this discussion, it is \emph{assumed} that there are no
-holes in the logical processor topology, thus $MP*NP = P$, where $P$ is the total number of
+This allocation statement specifies (given the one element halo size
+provided earlier for \texttt{U}) that the left halo column is
+\texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the bottom row
+is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the
+interior region \texttt{U(1:M,1:N)}.  In this allocation statement,
+the total number of process rows \texttt{NP} can be obtained at
+runtime, but \emph{may not} be explicitly provided (according to CAF
+rules) because the actual number of participating processes (in
+Fortran called images) is variable, depending on how many processes
+are requested at program startup.  In this discussion, it is
+\emph{assumed} that there are no holes in the logical processor
+topology, thus $MP*NP = P$, where $P$ is the total number of
 participating processes (images).
 
-So far --- except for the \texttt{HALO} attribute --- this is totally standard Fortran (2008).  For CAFe,
-we extend the concept of Fortran images to be hierachical.  By this we mean that each image \emph{may}
-have a subimage (or subimages), but this subimage is not visible to other regular Fortran images.
-Subimages also execute differently than normal images.  Subimages are task based while images all
-execute a single program but with different (multiple) data (SPMD).  A task can be given to a subimage,
-but execution on the subimage terminates once the task is finished.  Memory on a subimage is
-permanent, however, and must be specifically allocated and deallocated.
+\subsection{Subimages}
+
+\todo{subimages are critical and deserve their own subsection}
+
+So far --- except for the \texttt{HALO} attribute --- this is totally
+standard Fortran (2008).  For CAFe, we extend the concept of Fortran
+images to be hierachical.  By this we mean that each image \emph{may}
+have a subimage (or subimages), but this subimage is not visible to
+other regular Fortran images.  Subimages also execute differently than
+normal images.  Subimages are task based while images all execute a
+single program but with different (multiple) data (SPMD).  A task can
+be given to a subimage, but execution on the subimage terminates once
+the task is finished.  Memory on a subimage is permanent, however, and
+must be specifically allocated and deallocated.
 
@@
 \end{verbatim}
-where the integer argument representing one of attached hardware devices.  If the function
-fails (e.g., the requested device is unavailable) it returns the image number
-\texttt{this\_image()} of the process that is executing the current program.  Returning
-the current image allows program execution to proceed correctly even if there are no
-attached devices.
+
+where the integer argument representing one of attached hardware
+devices\todo{mention the relation of subimages and hardware earlier in
+  defn. of subimage}.  If the function fails (e.g., the requested
+device is unavailable) it returns the image number
+\texttt{this\_image()} of the process that is executing the current
+program.  Returning the current image allows program execution to
+proceed correctly even if there are no attached devices.
 
@@
 
-Finally, memory for the entire array \texttt{U} can be copied from the subimage device
-with the statement,
+Finally, memory for the entire array \texttt{U} can be copied from the
+subimage device with the statement,
+
 \begin{verbatim}
@@
 \end{verbatim}
-add memory deallocation is similar to memory allocation:
+
+add\todo{Add?  is that typo?} memory deallocation is similar to memory allocation:
+
 \begin{verbatim}

### 2015-01-31 papers/LOPe/motivation.tex
Completion for coauthors review of sections 2 and 3.
first pass of edits

@@
 
+\todo{This section should be rewritten and remove the petavision stuff}
 The concepts in this paper were motivated by problems encountered during
@@
 
-TODO: If space allows, give a simple 1D example of connections within a neural network in
-Figure 1.  Describe the size of visual cortex in terms of the number of neurons,
+TODO: If space allows, give a simple 1D figure of connections within a neural network.
+Describe the size of visual cortex in terms of the number of neurons,
 the number of synaptic connections, and performance.  Reference SC Gorden Bell

### 2015-01-31 papers/LOPe/programming-model.tex
first pass of edits

@@
 
-The LOPe programming model (for Locally-Oriented Programming extensions) restricts the programmer to
-a local view of the index space of an array.  Within a LOPe function, only a single array element (called
-the local element) is mutable.  In addition, a small halo region surrounding the local element is
-visible to the programmer, but this region is immutable.  Restricting the programmer to a local
-index space serves to reduce complexity by separating all boundary-value and processor-topology
-concerns from the implementation of the algorithm.
+The LOPe programming model (for Locally-Oriented Programming
+extensions) restricts the programmer to a local view of the index
+space of an array.  Within a LOPe function, only a single array
+element (called the local element) is mutable.  In addition, a small
+halo region surrounding the local element is visible to the
+programmer, but this region is immutable.  Restricting the programmer
+to a local index space serves to reduce complexity by separating all
+data- and task-decomposition concerns from the implementation of the
+element-level array calculations.
 
@@
 
-LOPe was implemented as an embedded domain specific language (DSL).  Fortran was chosen as the base
-language for LOPe, although in principle, the same techniques could be applied to language such as C
-or C++ (in conjuction with some form of array extensions such as those proposed by Intel).  Fortran
-was chosen because it already provides a rich array-based syntax.  Readers who are unfamiliar with
-Fortran syntax may wish to consult Appendix A, for a brief description of Fortran notation.
+LOPe is a domain specific language (DSL) implemented as a small extension
+to standard Fortran 2008\todo{What version of Fortran does it extend?}.
+Fortran was chosen as the base language for LOPe, although in
+principle, the same techniques could be applied to language such as C
+or C++ (in conjuction with some form of array extensions such as those
+proposed by Intel).  Fortran was chosen because it already provides a
+rich array-based syntax.  Readers who are unfamiliar with Fortran
+syntax may wish to consult Appendix A, for a brief description of
+Fortran notation.
+
+\subsection{Related work}
+
+LOPe builds upon prior work studying how to map Fortran to accelerator
+programming models like OpenCL.  In the ForOpenCL project~\cite{foropencl}
+we exploited Fortran pure and elemental functions to express the core
+data parallel kernels of an OpenCL-based program.  In practice, array
+calculations for a given index $i,j$ will require read-only access to
+a local neighborhood of size $[M,N]$ around $i,j$.  The LOPe work introduces
+a mechanism for representing these neighborhoods as array declaration
+type annotations.
+
+This work is based on concepts explored in the ZPL programming
+language~\cite{zpl1,zpl2} in which the programmer can define regions and 
+operators that are applied over the index sets corresponding to the regions.
+For example, the following ZPL code implements the same stencil as the
+LOPe code in Section~\ref{sec:where-is-jacobi-example}:
+
+\begin{verbatim}
+ZPL jacobi example here.
+\end{verbatim}
+
+This approach as demonstrated by ZPL is quite powerful for compilation
+purposes since it provides a clean decoupling of the operators applied
+over an array from the decomposition of that array over a potentially
+complex distributed memory hierarchy.
 
@@
 
-There are only a few syntax additions required for a LOPe program.  These additions include syntax for
-describing halo regions and concurrent procedures.  In all of the code examples that follow, DSL
-syntax additions are highlighted by the usage of capitalization for keywords that are either new or
-that acquire new usage.
+There are only a few syntax additions required for a LOPe program.
+These additions include syntax for describing halo regions and
+concurrent procedures.  In all of the code examples that follow,
+syntax additions are highlighted by the usage of capitalization for
+keywords that are either new or that acquire new usage.
 
 \textbf{Halo regions.}
-The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
-or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
-called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
-an array boundary, so that an array may be referenced using indices that fall ``outside'' of the
-logical domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler
-can exploit this knowledge.  For example, a halo region can be declared with a statement of the form,
+
+The principle semantic element of LOPe is the concept of a halo
+region.  A halo is an ``artificial'' or ``virtual'' region surrounding
+an array that contains boundary-value information.  Halo (also called
+ghost-cell) regions are commonly employed to unify array indexing
+schemes in the vicinity of an array boundary, so that an array may be
+referenced using indices that fall ``outside'' of the logical domain
+of the array.  In LOPe, the halo region is given explicit syntax so
+that the compiler can exploit this information for purposes of memory
+allocation, data replication and thread synchronization.  For example,
+a halo region can be declared with a statement of the form,
+
 \begin{verbatim}
@@
 \end{verbatim}
-This statement indicates that \texttt{A} is a rank one array, will be allocated later (when it will
-be provided with a specific size), and has a halo region of one element surrounding the array on
-either side.  Notation such as \texttt{M:*:N} is to be read as describing a halo region of
-\texttt{M} elements to the left, \texttt{N} elements to the right, and with an arbitrary number of
-elements in the middle (denoted by the $*$ symbol).  When used to describe a formal parameter of a
-function, the halo size may be ``assumed'' (denoted by a ``:'' symbol substituted in place of an
-explicit size) as is the following,
+
+This statement indicates that \texttt{A} is a rank one array, will be
+allocated later (when it will be provided with a specific size), and
+has a halo region of one element surrounding the array on either side.
+Notation such as \texttt{M:*:N} is to be read as describing a halo
+region of \texttt{M} elements to the left, \texttt{N} elements to the
+right, and with an arbitrary number of elements in the middle (denoted
+by the $*$ symbol).  When used to describe a formal parameter of a
+function, the halo size may be ``assumed'' (denoted by a ``:'' symbol
+substituted in place of an explicit size) as is the following,
+
 \begin{verbatim}
@@
 \end{verbatim}
-which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
-whose size is obtained by the compiler from the explicit halo size of the actual array argument
-provided at the calling site of the function.  In LOPe, there is no need (in this instance) for a
-repetitive \texttt{dimension(:,:)} specification, as it is inferred from the \texttt{HALO}
-specification.
+
+which in this instance indicates that \texttt{U} is a two-dimensional
+array, with a halo region whose size is obtained by the compiler from
+the explicit halo size of the actual array argument provided at the
+calling site of the function.  In LOPe, there is no need (in this
+instance) for a repetitive \texttt{dimension(:,:)} specification, as
+it is inferred from the \texttt{HALO} specification.\todo{This paragraph is a bit weird to read.}
 
@@
 \end{verbatim}
-defines a function that can be used as part of an iterative solution of Laplace's equation in two
-dimensions.  LOPe allows a function with the \texttt{pure} and \texttt{concurrent} attributes to be
-called from within a \texttt{do} \texttt{concurrent} loop body.  One should imagine that this
-function is called \emph{for each} \texttt{(i,j)} index of the interior of the array \texttt{U}.
-(An example of this usage will be given later in the text.)  Note that this usage introduces a race
-condition as new values of elements of \texttt{U} are created on the left-hand side of the
-assignment statement that may use \emph{new or old} values of \texttt{U} on the right-hand side.
-LOPe requires the compiler to guarantee that race conditions won't occur by using, e.g.,
-double-buffering techniques as needed.
+
+defines a function that can be used as part of an iterative solution
+of Laplace's equation in two dimensions.  LOPe allows a function with
+the \texttt{pure} and \texttt{concurrent} attributes to be called from
+within a \texttt{do} \texttt{concurrent} loop body.\todo{Worth pointing out that pure implies no side effects, concurrent implies no dependencies between iterations?}  One should
+imagine that this function is called \emph{for each} \texttt{(i,j)}
+index of the interior of the array \texttt{U}.  (An example of this
+usage will be given later in the text.)  Note that this usage
+introduces a race condition as new values of elements of \texttt{U}
+are created on the left-hand side of the assignment statement that may
+use \emph{new or old} values of \texttt{U} on the right-hand side.
+LOPe requires the compiler to guarantee that race conditions won't
+occur by using, e.g., double-buffering techniques as needed.
 
 \textbf{LOPe index notation.}
-In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
-array element and only the local element may be modified.  This zero-based indexing for the
-local-array element is different than conventional Fortran, where by default, array indices start at
-1.  For LOPe, this default convention is modified; indices for arrays declared with the
-\texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in one
-dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
-\texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
-the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local, respectively) and
-\texttt{U(0,-1)} and \texttt{U(0,+1)} (below and above of local, respectively).  The geometric
-positioning of the array elements can be seen by examining the arrangement of the expressions on the
-right-hand side of the \texttt{Laplace} example.
+
+In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array
+element is the \emph{local} array element and only the local element
+may be modified.  This zero-based indexing for the local-array element
+is different than conventional Fortran, where by default, array
+indices start at 1.  The use of zero-based indexing gives a clean
+symmetry for indices on either side of the central element at zero.
+For LOPe, this default convention is modified; indices for arrays
+declared with the \texttt{HALO} attribute (within the context of a
+\texttt{CONCURRENT} procedure) range (in one dimension) from
+\texttt{-M} to \texttt{+N} where the halo attribute has been declared
+as \texttt{M:*:N} by the originating type declaration.  The other
+array elements (in the halo region of the example) are
+\texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local,
+respectively) and \texttt{U(0,-1)} and \texttt{U(0,+1)} (below and
+above of local, respectively).  The geometric positioning of the array
+elements can be seen by examining the arrangement of the expressions
+on the right-hand side of the \texttt{Laplace} example.

### 2015-02-04 papers/LOPe/intro.tex
Many changes to serve as a candidate for an introduction.

@@
-It has been hard historically to create parallel programs.  The responsibility
-(and difficulty) of creating \emph{correct} parallel programs can be viewed as
-being spread between the programmer and the compiler.
-Ideally we would like to have parallel languages that make it easy for the
+\section{Introduction}
+
+Historically it has been hard to create parallel programs.  The responsibility (and difficulty) of
+creating \emph{correct} parallel programs can be viewed as being spread between the programmer and
+the compiler.  Ideally we would like to have parallel languages that make it easy for the
 programmer to express correct parallel programs (and conversely it should be difficult to express
-incorrect parallel programs).  Unfortunately, many current languages and
-standards place all of the responsibility on the user.  The best example of this
-are programs written using MPI (Message Passing Interface), where the programmer
-expresses all parallelism in terms of calls to library routines and the serial C or Fortran
-compiler knows nothing about the parallel semantics of the combined language
-plus MPI library.
+incorrect parallel programs).  Unfortunately, many current languages and standards place all of the
+responsibility on the user.  The best example of this are programs written using MPI (Message
+Passing Interface), where the programmer expresses all parallelism in terms of calls to library
+routines and the serial C or Fortran compiler knows nothing about the parallel semantics of the
+combined language plus MPI library.
 
-For the purpose of discussion, we define a degree of difficulty term called the
-joint Parallel Complexity Product (PCP),
+For the purpose of discussion, we define a degree of difficulty term called the joint Parallel
+Complexity Product (PCP),
 
-    PCP = code\_complexity X compiler\_complexity
+\begin{verbatim}
+    PCP = code_complexity X compiler_complexity
+\end{verbatim}
+and suggest that PCP is roughly a constant over time, say the level of complexity of writing an MPI
+program today, PCP\_MPI.
 
-and suggest that PCP is roughly a constant over time, say the level of
-complexity of writing an MPI program today, MPI\_PCP.
-
-Of course one would hope that over time code complexity goes down as better
-languages allow compilers to take on a greater share of the parallel complexity
-burden.  This is somewhat true of UPC and Coarray Fortran.  With the PGAS
-extensions, C and Fortran compilers are now aware of parallelism and now
-generate message passing code that had been handled by the MPI library.  In some
-instances the compiler is able to perform optimizations that it is not able to
-do with a library based scheme like MPI \cite{CrayPGASPaper}
+Of course one would hope that over time code complexity goes down as better languages allow
+compilers to take on a greater share of the parallel complexity burden.  This is somewhat true of
+Unified Parallel C (UPC) and Coarray Fortran.  With the PGAS extensions, C and Fortran compilers
+are now aware of parallelism and now generate message passing code that previously had been handled
+by the MPI library.  In some instances the compiler is able to perform optimizations that it is not
+able to do with a library-based scheme like MPI \cite{preissl:2011:SC}.
 
@@
 
-However, in large part these languages are largely syntatic sugar for message
-passing and do not provide a substatial decrease in code complexity
-\cite{CAF_MPI_PGAS_COMPARISON}.  Fortunately, skilled programmers in the HPC
-community have become accustomed to the level of complexity in an MPI program
-and are welcome to considering the simplifications of HPC and Coarray Fortran.
+However, in large part these languages are largely syntatic sugar for message passing and do not
+provide a substatial decrease in code complexity when application performance is a
+goal\cite{hasert:2011:EuroMPI}.  While skilled programmers in the HPC community have become
+accustomed to the level of complexity of an MPI program,
+%%and yet are welcome to considering the simplifications of UPC and Coarray Fortran.
+the problem for programmers is that hardware is changing in ways that increase the level of on-chip
+parallelism.  Indeed, the current generation of machines could be the last to provide homogeneous
+multi-core parallelism\cite{JORSjors.aw}.  For ultimate performance, programmers must envision
+coding for heterogeneous architectures 
+%%hyper-parallelism
+with huge new levels of on-node parallelism, at the same time they account for off-node
+parallelism.  Thus PCP has suddenly increased with PCP $=$ PCP\_MPI $+$ PCP\_Multi\_Core $>>$
+PCP\_MPI.  Since languages have not evolved that allow the compiler to take up this increase, the
+complexity for a programmer has necessarily dramatically increased.  Unfortunately, considering the
+complexity of large multiphysics code frameworks, it could take several person years to retarget an
+application for \emph{each} new heterogeneous platform\cite{DBLP:journals/corr/DubeyS13}.
 
-The problem for programmers is that hardware is changing in ways that increase
-the level of on-chip parallelism.  For ultimate performance, programmers must now
-account for huge new levels of on-node parallelism at the same time they account
-for off-node parallelism.  Thus PCP has suddenly increased with PCP = PCP\_Multi\_Core >> PCP\_MPI.
-Since languages have not evolved that allow the compiler to take up this increase, unfortunately the
-complexity for a programmer has dramatically increased.
+A reasonable solution given todays language limitations is to use MPI for distributed memory
+parallelism \emph{plus} OpenMP for on-chip parallelism.  This hybrid approach is taken by two
+successful, large-scale AMR multiphysics code frameworks, FLASH and Chombo; currently however,
+neither of these codes currently support hardware accelerators\cite{DBLP:journals/corr/DubeyS13}.
+%%This is the solution proposed by Cray and PGI \cite{BOF_SC10} (currently Chapel does not target GPUs\cite{Brad?}).
+Other choices for expressing on-chip parallelism are OpenCL\cite{OPENCL} and NVIDIA's CUDA.
+Unfortunately, achieving high performance using either of these two languages can be a daunting
+challenge and there are no guarantees that either language is suitable for future hardware
+generations.
 
-A reasonable solution given todays language limitations is to use MPI (or a PGAS
-language) \emph{plus} OpenMP for on-chip parallelism.  This is the solution proposed
-by Cray and PGI \cite{BOF_SC10} (currently Chapel does not target GPUs\cite{Brad?}).
-Other choices for expressing on-chip parallelism are OpenCL\cite{OPENCL} and NVIDIA's CUDA.
-
-In this paper we examine a language-based paradigm that allows the compiler to
-take on a larger portion of the PCP burden.  Anyone programming in OpenCL or
-CUDA is aware that explicit loop structures over array elements in a serial
-program are removed and replaced by a kernel program that is run over all of the
-elements of the input arrays.  We propose Locally Orientated Programming
-extensions (LOPe) to the Fortran and C languages that formally adopt this
-programming model.
+In this paper we examine a language-based paradigm that allows the compiler to take on a larger
+portion of the PCP burden.  Anyone programming in OpenCL or CUDA is aware that explicit loop
+structures over array elements in a serial program are removed and replaced by a kernel program
+that is run over all of the elements of the input arrays.  We propose Locally Orientated
+Programming extensions (LOPe) to the Fortran and potentially C languages that formally adopt this
+programming model.  The LOPe programming model explicitly separates numerical complexity from
+parallel complexity --- leaving much of the staging and assembling of data to the compiler --- and
+is applicable to the solution of stencil-based algorithms that provide data access patterns that
+are regular and spatially local.

### 2015-02-04 papers/LOPe/local.tex
first pass of edits
Added bibliography and replaced motivation with an introduction.

@@
 \usepackage{graphicx}
-
+\usepackage{todonotes}
 \usepackage{url}
-\urldef{\mailsa}\path|rasmus@uoregon.edu, mjsottile@computer.org,|
+\urldef{\mailsa}\path|rasmus@uoregon.edu, mjsottile@gmail.com,|
 \urldef{\mailsb}\path|dangle@ucar.edu, soren.rasmussen@aggiemail.usu.edu|
@@
 \noindent\keywordname\enspace\ignorespaces#1}
+
+\usepackage{cite}
 
@@
 
-\input{motivation}
+%\input{motivation}
+\input{intro}
 \input{programming-model}
 \input{coarray-extensions}
-\input{appendix-a}
 
-\section*{Acknowledgments}
+\bibliographystyle{plain}
+\bibliography{local}{}
+
+%\begin{thebibliography}{4}
+%\end{thebibliography}
+
+\subsubsection*{Acknowledgments.}
 This work was supported in part by the Department of Energy Office of Science, Advanced Scientific
@@
 
+\input{appendix-a}
+
 \end{document}

### 2015-02-04 papers/LOPe/programming-model.tex
Replied to \todo comment from Matt.

@@
 LOPe is a domain specific language (DSL) implemented as a small extension
-to standard Fortran 2008\todo{What version of Fortran does it extend?}.
+to standard Fortran 2008\todo{What version of Fortran does it extend? Not sure what you mean as F2008 is explicitly mentioned?}.
 Fortran was chosen as the base language for LOPe, although in

### 2015-02-05 papers/LOPe/intro.tex

@@
 
-Historically it has been hard to create parallel programs.  The responsibility (and difficulty) of
-creating \emph{correct} parallel programs can be viewed as being spread between the programmer and
-the compiler.  Ideally we would like to have parallel languages that make it easy for the
-programmer to express correct parallel programs (and conversely it should be difficult to express
-incorrect parallel programs).  Unfortunately, many current languages and standards place all of the
-responsibility on the user.  The best example of this are programs written using MPI (Message
-Passing Interface), where the programmer expresses all parallelism in terms of calls to library
-routines and the serial C or Fortran compiler knows nothing about the parallel semantics of the
-combined language plus MPI library.
+Historically it has been hard to create parallel programs.  The
+responsibility (and difficulty) of creating \emph{correct} parallel
+programs can be viewed as being spread between the programmer and the
+compiler.  Ideally we would like to have parallel languages that make
+it easy for the programmer to express correct parallel programs (and
+conversely it should be difficult to express incorrect parallel
+programs).  Unfortunately, many current languages and standards place
+all of the responsibility on the user.  The best example of this are
+programs written using MPI (Message Passing Interface), where the
+programmer expresses all parallelism in terms of calls to library
+routines and the serial C or Fortran compiler knows nothing about the
+parallel semantics of the combined language plus MPI library.
 
-For the purpose of discussion, we define a degree of difficulty term called the joint Parallel
-Complexity Product (PCP),
-
-\begin{verbatim}
-    PCP = code_complexity X compiler_complexity
-\end{verbatim}
-and suggest that PCP is roughly a constant over time, say the level of complexity of writing an MPI
-program today, PCP\_MPI.
-
-Of course one would hope that over time code complexity goes down as better languages allow
-compilers to take on a greater share of the parallel complexity burden.  This is somewhat true of
-Unified Parallel C (UPC) and Coarray Fortran.  With the PGAS extensions, C and Fortran compilers
-are now aware of parallelism and now generate message passing code that previously had been handled
-by the MPI library.  In some instances the compiler is able to perform optimizations that it is not
-able to do with a library-based scheme like MPI \cite{preissl:2011:SC}.
+Of course one would hope that over time code complexity goes down as
+better languages allow compilers to take on a greater share of the
+parallel complexity burden.  Multithreaded code written using OpenMP
+can be significantly simpler than the corresponding code in which the
+programmer explicitly manages a pool of threads themselves.  This is
+somewhat true for distributed memory parallelism using Unified
+Parallel C (UPC) and Coarray Fortran.  With the PGAS extensions, C and
+Fortran compilers are now aware of parallelism and now generate
+message passing code that previously had been handled by the MPI
+library.  In some instances the compiler is able to perform
+optimizations that it is not able to do with a library-based scheme
+like MPI \cite{preissl:2011:SC}.
 
@@
 
-However, in large part these languages are largely syntatic sugar for message passing and do not
-provide a substatial decrease in code complexity when application performance is a
-goal\cite{hasert:2011:EuroMPI}.  While skilled programmers in the HPC community have become
-accustomed to the level of complexity of an MPI program,
-%%and yet are welcome to considering the simplifications of UPC and Coarray Fortran.
-the problem for programmers is that hardware is changing in ways that increase the level of on-chip
-parallelism.  Indeed, the current generation of machines could be the last to provide homogeneous
-multi-core parallelism\cite{JORSjors.aw}.  For ultimate performance, programmers must envision
-coding for heterogeneous architectures 
-%%hyper-parallelism
-with huge new levels of on-node parallelism, at the same time they account for off-node
-parallelism.  Thus PCP has suddenly increased with PCP $=$ PCP\_MPI $+$ PCP\_Multi\_Core $>>$
-PCP\_MPI.  Since languages have not evolved that allow the compiler to take up this increase, the
-complexity for a programmer has necessarily dramatically increased.  Unfortunately, considering the
-complexity of large multiphysics code frameworks, it could take several person years to retarget an
-application for \emph{each} new heterogeneous platform\cite{DBLP:journals/corr/DubeyS13}.
+However, in large part these languages are largely syntatic sugar for
+message passing and do not provide a substantial decrease in code
+complexity when application performance is a
+goal\cite{hasert:2011:EuroMPI}.  While skilled programmers in the HPC
+community have become accustomed to the level of complexity of an MPI
+program, the problem for programmers is that hardware is changing in
+ways that increase the level of on-chip parallelism.  Indeed, the
+current generation of machines could be the last to provide
+homogeneous multi-core parallelism\cite{JORSjors.aw}.  For ultimate
+performance, programmers must envision coding for heterogeneous
+architectures with huge new levels of on-node parallelism, at the same
+time they account for off-node parallelism.
+Since languages have not evolved that allow the compiler to take up
+this increase, the complexity for a programmer has necessarily
+dramatically increased.  Unfortunately, considering the complexity of
+large multiphysics code frameworks, it could take several person years
+to retarget an application for \emph{each} new heterogeneous
+platform\cite{DBLP:journals/corr/DubeyS13}.
 
@@
 
-In this paper we examine a language-based paradigm that allows the compiler to take on a larger
-portion of the PCP burden.  Anyone programming in OpenCL or CUDA is aware that explicit loop
-structures over array elements in a serial program are removed and replaced by a kernel program
-that is run over all of the elements of the input arrays.  We propose Locally Orientated
-Programming extensions (LOPe) to the Fortran and potentially C languages that formally adopt this
-programming model.  The LOPe programming model explicitly separates numerical complexity from
-parallel complexity --- leaving much of the staging and assembling of data to the compiler --- and
-is applicable to the solution of stencil-based algorithms that provide data access patterns that
-are regular and spatially local.
+In this paper we examine a language-based paradigm that allows the
+compiler to take on a larger portion of the code complexity burden.
+Anyone programming in OpenCL or CUDA is aware that explicit loop
+structures over array elements in a serial program are removed and
+replaced by a kernel program that is run over all of the elements of
+the input arrays.  We propose \emph{Locally Orientated Programming
+extensions} (LOPe) to the Fortran and potentially C languages that
+formally adopt this programming model.  The LOPe programming model
+explicitly separates numerical complexity from parallel complexity ---
+leaving much of the staging and assembling of data to the compiler ---
+and is applicable to the solution of stencil-based algorithms that
+provide data access patterns that are regular and spatially local.

### 2015-02-05 papers/LOPe/conclusions.tex
Added text and notes.

\section{Conclusions}

Fortran is a general-purpose programming language and as such it provides limited facilities for
expressing concepts useful for stencil operations. For example, halo regions must be expressed in
terms of the existing syntax of the language as there is no way to specify that the ``interior'' of
an array is in any way special from an ``outside'' region.  By not providing support for stencils in
the language, the programmer must make specific choices regarding the order of access to the data
and in the order of operations on the data.  These choices often hide the opportunity for
optimizations by the compiler \cite{dubey:2014:wosc}.  If the compiler had knowledge of the symantic
intent of halo regions, it could reorder operations so that border regions were computed
\emph{before} interior regions, allowing the the transfer of data in halo regions to overlap with
computations on the interior.

Just as Coarray Fortran originally extended Fortran to include domain specific knowledge (parallel
computation) replacing MPI library routines\cite{numeric:reid}, LOPe seeks to extend Fortran by
providing domain specific knowledge of stencil operations.

CAFe extensions ...
* subimages

Specifically, LOPe and CAFe together provide:
* A local view of stencil operations on data.  This allows a complete separation of the implementation of a stencil algorithm with data access patterns.
* Memory placement via allocation routines allow the specification of the allocation location.
* Task execution placement with [[ ]] specifying which subimage is to execute a particular operation.
* Memory exchange via the \texttt{HALO\_TRANSFER} intrinsic procedure.
* Ability to run correctly even if subimages are not available.  Similar to OpenMP.
* Ability to associate arrays on subimages with coarrays on host thus allowing compiler to effect the halo transfers between the multiple memory levels.
* Halo abstraction that allows for more flexibility for the compiler to retarget for different architectures.
* A simple way to express variables as ``thread private'' as all local variables declared in a \texttt{CONCURRENT} procedure are private to the calling procedure.  This is distinguished from OpenMP.

\subsubsection{Benefits.}
* extends the programmer
* restricts the programmer
* no race conditions
* Abstractions provide ample opportunities for compiler optimizations.

\subsubsection{Limitations.}
* no composition useful for nonlinear dynamics\cite{coella}
* difficult (but not impossible) to implement algoriths that perform operations between arrays of
differing sizes (multigrid implementations for example).

By implementing LOPe
we have demonstrated that LOPe can be ... It remains to history to see if LOPe is general purpose
enough to be included in Fortran or if it should remain as a special purpose preprocessing tool.

### 2015-02-05 papers/LOPe/transformations.tex
Added very little text.

\section{LOPe Implementation}

This section describes how LOPe extensions to Fortran have been implemented.  It also describes how
source-to-source transformations have been implemented via rewrite rules and strategies in
StrategoXT.

\subsection{Halo Transfer Performance}

Amortization of halo transfer and kernel

### 2015-02-06 papers/LOPe/transformations.tex
Beta rough draft.

@@
-\section{LOPe Implementation}
+\section{LOPe and CAFe Implementation}
 
@@
 
+\subsection{Source-to-Source Transformation Infrastructure}
+
+1. Parsing to produce an AST nodes representated as ATerms.
+2. Transformations on the AST nodes represented as rewrite rules and strategies in Stratego/XT
+3. Pretty printing of transformed nodes using the pretty printing facilites of BoxLib.
+
+\subsection{Transformations for Concurrent Procedures}
+
+A key component of the LOPe transformations is the targeting of a LOPe \texttt{CONCURRENT}
+procedure for a particular hardware (and software) architecture.  The execution target can be one
+of several choices, including serial execution by the current process (and hardware) via inlining
+of the function, parallel execution by inlining with OpenMP compiler directives, or parallel
+execution by a heterogeneous processing elements with a language like OpenCL, for example.
+
+For this work we have developed rewrite strategies and rules in Stratego/XT to rewrite Fortran AST
+nodes to C AST nodes (extended with necessary OpenCL keywords).  The C AST ATerms have mostly a
+one-to-one correspondence with Fortran terms.  A \texttt{CONCURRENT} procedure is transformed to a
+OpenCL kernel (with the \texttt{\_\_kernel} attribute), Fortran formal parameters are transformed
+to C parameters (with a direct mapping of types), and local Fortran variable declarations are rewritten
+as C declarations.  Similarly, Fortran executable statements are rewritten as C executable AST nodes.
+Mostly the only complication is mapping the LOPe local array index view to the global C index space.
+This translation is facilitated by the Fortran symbol table that stores array shape and halo
+size information.
+
+Once all rewriting passes have been made to transform from LOPe Fortran ATerms to C ATerms, the
+C ATerms are output as an OpenCL file using a C pretty printer.
+
+\subsection{Transformations at the Call Site}
+
+Transformations at a LOPe procedure call site are more difficult, though technically straight forward.
+The Fortran function call must be transformed to a call to run the OpenCL kernel (produced as described
+above).  This is facilitated by the use of the ForOpenCL library that provides Fortran bindings to the
+OpenCL runtime\cite{foropencl}.  But the usage requires the declaration of extra variables,
+allocation of memory on the OpenCL device (subimage), transfer of memory, marshalling of kernel
+arguments, and synchronization.
+
+These transformations are accomplished using several rewrite stages using Stratego/XT strategies:
+1. A symbol table is produced in the first pass to store information related to arrays including,
+array shape, halo size, and allocation status;
+2. Additional variables are declared that are used to maintain the OpenCL runtime state, including
+the OpenCL device (as obtained from a \texttt{GET\_SUBIMAGE} call), the OpenCL kernel (produced as
+described above), and OpenCL variables used to marshall kernel arguments; and
+3. All CAFe code related to subimage usage is lowered to standard Fortran with calls to the
+ForOpenCL library as needed.  The latter step includes the allocation of OpenCL device memory,
+transfer of memory to and from the OpenCL device and the executing Fortran process (image) as
+indicated in the program source by the single bracket \texttt{[subimage]} notation, marshalling
+of kernel argument, running of the OpenCL kernel, and synchronization.
+
+Though not yet available, similar rewrite strategies are planned for targeting programming models
+other than OpenCL including parallel execution with OpenMP directives.  In addition, simple serial
+execution with function inlining (if desired) will be performed by the regular Fortran compiler once
+all CAFe and LOPe code has been lowered to standard Fortran.
+
 \subsection{Halo Transfer Performance}
 
-Amortization of halo transfer and kernel
+A thorough examination of potential performance gains using LOPe for parallelization of code is beyond
+the scope of this paper.  The emphasis of this work is to explain the syntax and execution symantics
+of a LOPe application.  However, preliminary work using the StrategoXT rewrite system as described indicates
+that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).
+
+The current implementation does \emph{not} take advantage of optimization strategies such as
+prefetching of array tiles (including halos) into OpenCL local memory or of loop unrolling on the
+OpenCL device.  Neither does it take advantage of the potential of LOPe \texttt{CONCURRENT} function
+fusion along with execution and synchronization strategies to overlap communication with
+computation.
+
+Since many scientific codes are dominated by memory performance, including and especially stencil algorithms
+as they typically only involve a computation on a small locally central array element and a small overlapping
+halo region.  Stencil operations frequently do not contain enough floating point operations per memory load
+to allow for floating point performation to operate at peak (though this is entirely application and domain
+specific).  Thus we illustrate the \emph{potential} for performance by noting the latency and throughput
+performance of an attached GPU inconjuction with MPI distributed memory performance associated with halo
+transfer in Table 1.
+
+The results in Table 1 indicate that the primary bottleneck in using accelerators attached to the
+TODO bus using OpenCL is likely to be the latency in transferring memory to and from the device for
+distributed memory clusters of only a few nodes exchanging halo data using MPI.

### 2015-02-07 papers/LOPe/coarray-extensions.tex
Fixed one of Matt's \todo comments.
Paper lenght shortening begun.

@@
 \end{verbatim}
-
 with an explicit halo size and with two local dimensions (rank) and
@@
 \texttt{U} are given by the allocation statement,
-
-\begin{verbatim}
-  allocate( U(0:M+1,0:N+1)[MP,*] )
-\end{verbatim}
-
+\texttt{allocate(U(0:M+1,0:N+1)[MP,*])}.
 This allocation statement specifies (given the one element halo size
@@
 
-\todo{subimages are critical and deserve their own subsection}
-
 So far --- except for the \texttt{HALO} attribute --- this is totally
@@
 normal images.  Subimages are task based while images all execute a
-single program but with different (multiple) data (SPMD).  A task can
+Single Program but with different (Multiple) Data (SPMD).  A task can
 be given to a subimage, but execution on the subimage terminates once
@@
 One obtains a subimage by executing a function call,
-\begin{verbatim}
-   device = get_subimage(1)
-\end{verbatim}
-
+\texttt{device = GET\_SUBIMAGE(1)}
 where the integer argument representing one of attached hardware
@@
 
-Execution of the \texttt{Laplace} task is done using the Fortran \texttt{do}
+Execution of the \texttt{Laplace} task is done using the \texttt{do}
 \texttt{concurrent} construct:
@@
 subimage device with the statement,
+\texttt{U = U[device]}, 
+and memory deallocation (not shown) is similar to memory allocation.
 
-\begin{verbatim}
-  U = U[device]
-\end{verbatim}
-
-add\todo{Add?  is that typo?} memory deallocation is similar to memory allocation:
-
-\begin{verbatim}
-   deallocate(U)
-   if (device /= this_image()) then
-      deallocate(U[device])   [[device]]
-   end if
-\end{verbatim}
+%%\begin{verbatim}
+%%   deallocate(U)
+%%   if (device /= this_image()) then
+%%      deallocate(U[device])   [[device]]
+%%   end if
+%%\end{verbatim}
 

### 2015-02-07 papers/LOPe/intro.tex
Paper lenght shortening begun.

@@
 compiler.  Ideally we would like to have parallel languages that make
-it easy for the programmer to express correct parallel programs (and
+it easy for the programmer to express correct parallel programs --- and
 conversely it should be difficult to express incorrect parallel
-programs).  Unfortunately, many current languages and standards place
+programs.  Unfortunately, many current languages and standards place
 all of the responsibility on the user.  The best example of this are
@@
 somewhat true for distributed memory parallelism using Unified
-Parallel C (UPC) and Coarray Fortran.  With the PGAS extensions, C and
+Parallel C (UPC) and Coarray Fortran.  With these PGAS extensions, C and
 Fortran compilers are now aware of parallelism and now generate
@@
 compiler to take on a larger portion of the code complexity burden.
-Anyone programming in OpenCL or CUDA is aware that explicit loop
+Anyone programming in OpenCL is aware that explicit loop
 structures over array elements in a serial program are removed and

### 2015-02-07 papers/LOPe/local.tex
Added sections on transformations and conclusions.
Paper lenght shortening begun.

@@
 \usepackage{url}
-\urldef{\mailsa}\path|rasmus@uoregon.edu, mjsottile@gmail.com,|
+%%\urldef{\mailsa}\path|rasmus@uoregon.edu, mjsottile@gmail.com,|
+\urldef{\mailsa}\path|rasmus@uoregon.edu|
 \urldef{\mailsb}\path|dangle@ucar.edu, soren.rasmussen@aggiemail.usu.edu|
@@
 % unless you accept that it will be published
-\institute{441 McKenzie Hall,\\
-5246 University of Oregon, Eugene, Oregon, USA\\
-\mailsa\\
-\mailsb\\
-\url{https://openfortran.uoregon.edu}}
+%%\institute{441 McKenzie Hall,\\
+%%5246 University of Oregon, Eugene, Oregon, USA\\
+%%\mailsa\\
+%%\mailsb\\
+%%\url{https://openfortran.uoregon.edu}}
+
+\institute{University of Oregon, Eugene, Oregon, USA\\
+\mailsa
+}
 
@@
 Emerging hybrid accelerator architectures for high performance computing are often suited for the
-use of a data-parallel programming model.  Unfortunately, programmers for these emerging
-architectures face a steep learning curve that frequently requires learning a new language, e.g.,
-OpenCL or CUDA, or employing OpenMP compiler directives.  Furthermore, the distributed (and
-frequently multi-level) nature of the memory organization of clusters of these architectures
-provides an additional level of complexity for the programmer.  This paper presents preliminary work
-examining how programming with a local orientation can be employed to provide simple access to
+use of a data-parallel programming model.  Unfortunately, programmers of these
+architectures face a steep learning curve that frequently requires learning a new language (e.g.,
+OpenCL). Furthermore, the distributed (and
+frequently multi-level) nature of the memory organization of clusters of these machines
+provides an additional level of complexity.  This paper presents preliminary work
+examining how programming with a local orientation can be employed to provide simpler access to
 accelerator architectures.  A locally-oriented programming model is especially useful for the
-solution of algorithms requiring the application of a convolution kernel.  In this programming
+solution of algorithms requiring the application of a stencil or convolution kernel.  In this programming
 model, a programmer codes the algorithm by modifying \emph{only a single array element} (called the
-local element), but has read-only access to a small sub-array surrounding the local element so that
-a stencil can be applied.  We demonstrate how a locally-oriented programming model can be adopted as
-an embedded, domain-specific language using source-to-source program transformations.
+local element), but has read-only access to a small sub-array surrounding the local element.
+We demonstrate how a locally-oriented programming model can be adopted as
+a language extension using source-to-source program transformations.
 
@@
 \input{coarray-extensions}
+\input{transformations}
+\input{conclusions}
 

### 2015-02-07 papers/LOPe/programming-model.tex
Paper lenght shortening begun.

--- 2015-01-31 papers/LOPe/programming-model.tex
@@
 
-The LOPe programming model (for Locally-Oriented Programming
-extensions) restricts the programmer to a local view of the index
+The LOPe programming model
+restricts the programmer to a local view of the index
 space of an array.  Within a LOPe function, only a single array
@@
 LOPe is a domain specific language (DSL) implemented as a small extension
-to standard Fortran 2008\todo{What version of Fortran does it extend?}.
+to the Fortran 2008 standard.
 Fortran was chosen as the base language for LOPe, although in
 principle, the same techniques could be applied to language such as C
-or C++ (in conjuction with some form of array extensions such as those
-proposed by Intel).  Fortran was chosen because it already provides a
+or C++.  Fortran was chosen because it provides a
 rich array-based syntax.  Readers who are unfamiliar with Fortran
@@
 \end{verbatim}
-
 This approach as demonstrated by ZPL is quite powerful for compilation
@@
 
-\textbf{Halo regions.}
-
+\subsubsection{Halo regions.}
 The principle semantic element of LOPe is the concept of a halo
@@
 \end{verbatim}
-
 This statement indicates that \texttt{A} is a rank one array, will be
@@
 \end{verbatim}
-
 which in this instance indicates that \texttt{U} is a two-dimensional
@@
 
-\textbf{Concurrent functions.}
+\subsubsection{Concurrent functions.}
 The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
@@
 \end{verbatim}
-
 defines a function that can be used as part of an iterative solution
@@
 
-\textbf{LOPe index notation.}
+\subsubsection{LOPe index notation.}
 

### 2015-02-07 papers/LOPe/transformations.tex
Paper lenght shortening begun.

@@
 
-\subsection{Halo Transfer Performance}
+%%\subsection{Halo Transfer Performance}
 
-A thorough examination of potential performance gains using LOPe for parallelization of code is beyond
-the scope of this paper.  The emphasis of this work is to explain the syntax and execution symantics
-of a LOPe application.  However, preliminary work using the StrategoXT rewrite system as described indicates
-that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).
+%%A thorough examination of potential performance gains using LOPe for parallelization of code is beyond the scope of this paper.
+The emphasis of this work is to explain the syntax and execution symantics of a LOPe application.
+However, preliminary work using the StrategoXT rewrite system as described indicates that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).
 
-The current implementation does \emph{not} take advantage of optimization strategies such as
-prefetching of array tiles (including halos) into OpenCL local memory or of loop unrolling on the
-OpenCL device.  Neither does it take advantage of the potential of LOPe \texttt{CONCURRENT} function
-fusion along with execution and synchronization strategies to overlap communication with
-computation.
+%%The current implementation does \emph{not} take advantage of optimization strategies such as prefetching of array tiles (including halos) into OpenCL local memory or of loop unrolling on the OpenCL device.
+Neither does it take advantage of the potential of LOPe \texttt{CONCURRENT} function fusion along with execution and synchronization strategies to overlap communication with computation.
 
-Since many scientific codes are dominated by memory performance, including and especially stencil algorithms
-as they typically only involve a computation on a small locally central array element and a small overlapping
-halo region.  Stencil operations frequently do not contain enough floating point operations per memory load
-to allow for floating point performation to operate at peak (though this is entirely application and domain
-specific).  Thus we illustrate the \emph{potential} for performance by noting the latency and throughput
-performance of an attached GPU inconjuction with MPI distributed memory performance associated with halo
-transfer in Table 1.
+%%Since many scientific codes are dominated by memory performance, including and especially stencil algorithms as they typically only involve a computation on a small locally central array element and a small overlapping halo region.
+Stencil operations frequently do not contain enough floating point operations per memory load to allow for floating point performation to operate at peak (though this is entirely application and domain specific).
+Thus we illustrate the \emph{potential} for performance by noting the latency and throughput performance of an attached GPU inconjuction with MPI distributed memory performance associated with halo transfer in Table 1.
 
-The results in Table 1 indicate that the primary bottleneck in using accelerators attached to the
-TODO bus using OpenCL is likely to be the latency in transferring memory to and from the device for
-distributed memory clusters of only a few nodes exchanging halo data using MPI.
+%%The results in Table 1 indicate that the primary bottleneck in using accelerators attached to the TODO bus using OpenCL is likely to be the latency in transferring memory to and from the device for distributed memory clusters of only a few nodes exchanging halo data using MPI.

### 2015-02-07 papers/LOPe/intro.tex
Beta-version of text.

@@
 somewhat true for distributed memory parallelism using Unified
-Parallel C (UPC) and Coarray Fortran.  With these PGAS extensions, C and
+Parallel C (UPC) and Coarray Fortran.  With these PGAS
+(Partitioned Global Address Space) extensions, C and
 Fortran compilers are now aware of parallelism and now generate
@@
 %%This is the solution proposed by Cray and PGI \cite{BOF_SC10} (currently Chapel does not target GPUs\cite{Brad?}).
-Other choices for expressing on-chip parallelism are OpenCL\cite{OPENCL} and NVIDIA's CUDA.
+Other choices for expressing on-chip parallelism are OpenCL~\cite{opencl11} and NVIDIA's CUDA.
 Unfortunately, achieving high performance using either of these two languages can be a daunting

### 2015-02-07 papers/LOPe/programming-model.tex
Beta-version of text.

@@
 Fortran was chosen as the base language for LOPe, although in
-principle, the same techniques could be applied to language such as C
+principle, the same techniques could be applied to languages such as C
 or C++.  Fortran was chosen because it provides a
@@
 LOPe builds upon prior work studying how to map Fortran to accelerator
-programming models like OpenCL.  In the ForOpenCL project~\cite{foropencl}
-we exploited Fortran pure and elemental functions to express the core
-data parallel kernels of an OpenCL-based program.  In practice, array
+programming models like OpenCL.  In the ForOpenCL project~\cite{Sottile:2013:FTE:2441516.2441520}
+we exploited Fortran's pure and elemental functions to express
+data-parallel kernels of an OpenCL-based program.  In practice, array
 calculations for a given index $i,j$ will require read-only access to
-a local neighborhood of size $[M,N]$ around $i,j$.  The LOPe work introduces
+a local neighborhood of size $[M,N]$ around $i,j$.  LOPe extends this work by introducing
 a mechanism for representing these neighborhoods as array declaration
@@
 
-This work is based on concepts explored in the ZPL programming
-language~\cite{zpl1,zpl2} in which the programmer can define regions and 
-operators that are applied over the index sets corresponding to the regions.
-For example, the following ZPL code implements the same stencil as the
-LOPe code in Section~\ref{sec:where-is-jacobi-example}:
+ForOpenCL was based on concepts explored in the ZPL programming language~\cite{chamberlain04zpl} in
+which the programmer can define regions and operators that are applied over the index sets
+corresponding to the sub-array regions.  This approach is quite powerful for
+compilation purposes since it provides a clean decoupling of the operators applied over an array
+from the decomposition of that array over a potentially complex distributed memory hierarchy.
+However, unlike the ZPL operations on entire sub-arrays, LOPe expresses operations based on
+a \emph{single} local array-element.
 
-\begin{verbatim}
-ZPL jacobi example here.
-\end{verbatim}
-This approach as demonstrated by ZPL is quite powerful for compilation
-purposes since it provides a clean decoupling of the operators applied
-over an array from the decomposition of that array over a potentially
-complex distributed memory hierarchy.
+%
+% For space reasons a ZPL example is not shown and text reworded accordingly
+%
+%%For example, the following ZPL code implements the same stencil as the LOPe code in Fig. 1:
+
+%%\begin{verbatim}
+%%ZPL jacobi example here.
+%%\end{verbatim}
+%%This approach as demonstrated by ZPL is quite powerful for compilation purposes since it provides a clean decoupling of the operators applied over an array from the decomposition of that array over a potentially complex distributed memory hierarchy.
 
@@
 These additions include syntax for describing halo regions and
-concurrent procedures.  In all of the code examples that follow,
-syntax additions are highlighted by the usage of capitalization for
+concurrent procedures.  In code examples that follow,
+language additions are highlighted by the usage of capitalization for
 keywords that are either new or that acquire new usage.
@@
 \subsubsection{Halo regions.}
-The principle semantic element of LOPe is the concept of a halo
-region.  A halo is an ``artificial'' or ``virtual'' region surrounding
+The principle semantic element of LOPe is the concept of a halo.
+A halo is an ``artificial'' or ``virtual'' region surrounding
 an array that contains boundary-value information.  Halo (also called
 ghost-cell) regions are commonly employed to unify array indexing
-schemes in the vicinity of an array boundary, so that an array may be
+schemes in the vicinity of an array boundary so that an array may be
 referenced using indices that fall ``outside'' of the logical domain
@@
 This statement indicates that \texttt{A} is a rank one array, will be
-allocated later (when it will be provided with a specific size), and
+allocated later, and
 has a halo region of one element surrounding the array on either side.
-Notation such as \texttt{M:*:N} is to be read as describing a halo
-region of \texttt{M} elements to the left, \texttt{N} elements to the
-right, and with an arbitrary number of elements in the middle (denoted
-by the $*$ symbol).  When used to describe a formal parameter of a
-function, the halo size may be ``assumed'' (denoted by a ``:'' symbol
-substituted in place of an explicit size) as is the following,
+The halo notation \texttt{M:*:N} specifies a halo
+of \texttt{M} elements to the left, \texttt{N} elements to the
+right, and an arbitrary number of ``interior'' array elements.
+When used to describe a formal parameter of a
+function, such as the type-declaration statement, \texttt{real, HALO(:,:) :: U},
+the halo size is inferred by the compiler
+from the actual array argument provided at the
+calling site of the function.
 
-\begin{verbatim}
-  real, HALO(:,:) :: U
-\end{verbatim}
-which in this instance indicates that \texttt{U} is a two-dimensional
-array, with a halo region whose size is obtained by the compiler from
-the explicit halo size of the actual array argument provided at the
-calling site of the function.  In LOPe, there is no need (in this
-instance) for a repetitive \texttt{dimension(:,:)} specification, as
-it is inferred from the \texttt{HALO} specification.\todo{This paragraph is a bit weird to read.}
+%%In LOPe, there is no need (in this instance) for a repetitive \texttt{dimension(:,:)} specification, as it is inferred from the \texttt{HALO} specification.
 
 \subsubsection{Concurrent functions.}
-The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
-of a \texttt{do} \texttt{concurrent} loop construct.  By using this construct, the programmer
-asserts that specific iterations of the loop body may be executed by the compiler in \emph{any
-  order;} even \emph{concurrently.}
-%%LOPe allows the \texttt{concurrent} attribute to be attached to a function declaration so that the function can be called from within a concurrent loop without fear of race conditions.
-For example, the following,
+
+The second keyword employed by LOPe is \texttt{concurrent} which already exists in the form of a
+\texttt{do} \texttt{concurrent} loop, whereby the programmer asserts that specific
+iterations of the loop body may be executed by the compiler in \emph{any order,} even
+\emph{concurrently.}  LOPe allows a function with the attributes \texttt{pure} (assertion of no
+side effects) and \texttt{concurrent} (assertion of no dependencies between iterations) to be
+called from within a \texttt{do} \texttt{concurrent} loop.  An example of a LOPe function is
+shown in Fig. 1 and an example calling this function will be provided later in the text.  One
+should imagine that a LOPe function is called \emph{for each} \texttt{i,j} index of the interior of
+the array \texttt{U}.  Note that this usage introduces a race condition as new values of
+elements of \texttt{U} are created on the left-hand side of the assignment statement that may use
+\emph{new or old} values of \texttt{U} on the right-hand side.  LOPe requires the compiler to
+guarantee that race conditions won't occur by using, e.g., double-buffering techniques as needed.
+
+\vspace{-.1in}
+
+\begin{figure}
 \begin{verbatim}
-  pure CONCURRENT subroutine Laplace(U)
+           pure CONCURRENT subroutine Laplacian(U)}
                real, HALO(:,:) :: U
@@
                                     +   U(0,-1)
-  end subroutine Laplace
+           end subroutine Laplacian
 \end{verbatim}
-defines a function that can be used as part of an iterative solution
-of Laplace's equation in two dimensions.  LOPe allows a function with
-the \texttt{pure} and \texttt{concurrent} attributes to be called from
-within a \texttt{do} \texttt{concurrent} loop body.\todo{Worth pointing out that pure implies no side effects, concurrent implies no dependencies between iterations?}  One should
-imagine that this function is called \emph{for each} \texttt{(i,j)}
-index of the interior of the array \texttt{U}.  (An example of this
-usage will be given later in the text.)  Note that this usage
-introduces a race condition as new values of elements of \texttt{U}
-are created on the left-hand side of the assignment statement that may
-use \emph{new or old} values of \texttt{U} on the right-hand side.
-LOPe requires the compiler to guarantee that race conditions won't
-occur by using, e.g., double-buffering techniques as needed.
+\vspace{-.1in}
+\caption{A LOPe function implementing a Laplacian kernel in two dimensions.}
+\end{figure}
+
+\vspace{-.3in}
 
@@
 
-In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array
-element is the \emph{local} array element and only the local element
-may be modified.  This zero-based indexing for the local-array element
-is different than conventional Fortran, where by default, array
-indices start at 1.  The use of zero-based indexing gives a clean
-symmetry for indices on either side of the central element at zero.
-For LOPe, this default convention is modified; indices for arrays
-declared with the \texttt{HALO} attribute (within the context of a
-\texttt{CONCURRENT} procedure) range (in one dimension) from
-\texttt{-M} to \texttt{+N} where the halo attribute has been declared
-as \texttt{M:*:N} by the originating type declaration.  The other
-array elements (in the halo region of the example) are
-\texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local,
-respectively) and \texttt{U(0,-1)} and \texttt{U(0,+1)} (below and
-above of local, respectively).  The geometric positioning of the array
-elements can be seen by examining the arrangement of the expressions
-on the right-hand side of the \texttt{Laplace} example.
+In the \texttt{Laplacian} example the \texttt{U(0,0)} array element is the \emph{local} array
+element and only the local element may be modified.  This zero-based indexing for the local-array
+element differs from conventional Fortran, where by default, array indices start at 1.  The use of
+zero-based indexing gives a clean symmetry for indices on either side of the central element at
+zero. The other array elements are in the halo region and are \texttt{U(-1,0)} and
+\texttt{U(+1,0)} (left and right of local, respectively) and \texttt{U(0,-1)} and \texttt{U(0,+1)}
+(below and above of local).  The geometric positioning of the array elements can be
+seen by examining the arrangement of the expressions on the right-hand side of Fig. 1.

### 2015-02-08 papers/LOPe/transformations.tex
A few minor changes before finishing up.

@@
 
-This section describes how LOPe extensions to Fortran have been implemented.  It also describes how
-source-to-source transformations have been implemented via rewrite rules and strategies in
-StrategoXT.
+This section briefly describes how LOPe extensions to Fortran have been implemented as
+source-to-source transformations via rewrite rules and strategies in StrategoXT.
 
-\subsection{Source-to-Source Transformation Infrastructure}
-
-1. Parsing to produce an AST nodes representated as ATerms.
-2. Transformations on the AST nodes represented as rewrite rules and strategies in Stratego/XT
+Transformations are accomplished in three phases:
+1. Parsing to produce an AST nodes representated as ATerms;
+2. Transformations on the AST nodes represented as rewrite rules and strategies in Stratego/XT; and
 3. Pretty printing of transformed nodes using the pretty printing facilites of BoxLib.
@@
 nodes to C AST nodes (extended with necessary OpenCL keywords).  The C AST ATerms have mostly a
-one-to-one correspondence with Fortran terms.  A \texttt{CONCURRENT} procedure is transformed to a
-OpenCL kernel (with the \texttt{\_\_kernel} attribute), Fortran formal parameters are transformed
-to C parameters (with a direct mapping of types), and local Fortran variable declarations are rewritten
-as C declarations.  Similarly, Fortran executable statements are rewritten as C executable AST nodes.
-Mostly the only complication is mapping the LOPe local array index view to the global C index space.
-This translation is facilitated by the Fortran symbol table that stores array shape and halo
-size information.
+one-to-one correspondence with Fortran terms.  A \texttt{CONCURRENT} procedure is transformed to an
+OpenCL kernel, Fortran formal parameters are transformed to C parameters (with a direct mapping of
+types), and local Fortran variable declarations are rewritten as C declarations.  Similarly,
+Fortran executable statements are rewritten as C executable AST nodes.  Mostly the only
+complication is mapping the LOPe local array index view to the global C index space.  This
+translation is facilitated by a Fortran symbol table that stores array shape and halo size
+information.
 

### 2015-02-08 papers/LOPe/coarray-comparison.tex
Beta-version of text.

--- 2015-01-28 papers/LOPe/coarray-comparison.tex
@@
 We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
-implementation of the same algorithm from the original Numrich and Reid paper first describing
-coarrays in Fortran.  We should point out that this comparison is somewhat unfair, because Numrich
-and Reid were introducing coarray notation for transferring memory on distributed memory
-architectures, not demonstrating how ideally one should use coarrays within a large application.
+implementation of the same algorithm from the original Numrich and Reid
+paper\cite{Numrich:1998:CFP:289918.289920} first describing coarrays in Fortran.  We should point
+out that this comparison is somewhat unfair, because Numrich and Reid were introducing coarray
+notation for transferring memory on distributed memory architectures, not demonstrating how ideally
+one should use coarrays within a large application.
+
 %%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.
@@
     if (me == ncol) right = 1
-    call sync_all( [left,right] )   ! Wait for left and right
+    call sync all    ! Wait for left and right
     new_u(1:nrow) = new_u(1:nrow) + u(1:nrow)[left] + u(1:nrow)[right]
-    call sync_all( [left,right] )
+    call sync all
     u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
@@
 
-The advantages of LOPe and CAFe over this example are now described.  Please note that the following
-is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and
-LOPe only pertains to the halo pattern most useful in the implementation of stencil-based algorithms.
+%%The advantages of LOPe and CAFe over this example are now described.
+Please note that the following is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and LOPe only pertains to the halo pattern most useful in the implementation of stencil-based algorithms.
 
-\subsubsection{LOPe advantages.}
+\subsection{LOPe Advantages}
+
 \begin{itemize}
@@
 Algorithm implementation is separate from user-specified synchronization, e.g., \texttt{call
-  sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the \texttt{CONCURRENT}
+  sync\_all}.  In LOPe, synchronization is subsumed in the semantics of the \texttt{CONCURRENT}
 attribute and the \texttt{TRANSFER\_HALO} function call.
@@
 is to be executed.  The CAF example explicity denotes where a statement is to executed
-with control flow construct like \texttt{if (me == 1)}.
+with a control flow construct like \texttt{if (me == 1)}.
 
@@
 
-\subsubsection{Errors that are constrained by the language.}
+\subsection{LOPe Constraints}
+
+Contraints provided by the LOPe language extensions allow the compiler to catch several classes of
+errors that otherwise would be the programmer's responsibility:
+
 \begin{itemize}
@@
 A programmer is not able to store data to the halo region during execution of a LOPe concurrent
-function.  If this were allowed, one thread could overwrite another threads data at undefined times.
-The compiler is able to catch this class of error.
+function.  Neither are stores to the local element, followed by a read from the halo region of the
+same variable allowed.  If these were allowed, one thread could overwrite another threads data at
+undefined times.
 
@@
 A programmer can't make indexing errors in a concurrent routine by going out of bounds of the array
-plus halo memory.  The compiler is able to catch this class of error at compile time as long as
-compile-time constants are used to specify halo sizes.
+plus halo memory.
 
@@
 A programmer is not able to cause race conditions by forgetting to create and use temporary arrays
-properly.  In LOPe it is the compilers responsibility to store data in temporary memory.
+properly.
 
 \item
-A programmer can't make synchronization errors as synchronization is implicit in the
-\texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided with a copy of
-it's local array element plus halo that is consistent with the state of memory \emph{at the time of
-  invocation of the procedure}.  Stores to an individual thread's local array element (by that
-thread) is never visible to other threads.  LOPE encourages the creation of small functions and lets
-the compiler fuse the functions together to improve performance and to provide necessary
+A programmer can't make synchronization errors in calls to LOPe functions as synchronization is
+implicit in the \texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided
+with a copy of it's local array element plus halo that is consistent with the state of memory
+\emph{at the time of invocation of the procedure}.  Stores to an individual thread's local array
+element (by that thread) are never visible to other threads.  Normal coarray programs (like MPI)
+require explicit synchronization to ensure that processes arrive at the same program location
+before a memory read occurs (for example).  LOPE encourages the creation of small functions and
+lets the compiler fuse the functions together to improve performance and to provide necessary
 synchronization.

### 2015-02-08 papers/LOPe/coarray-extensions.tex
Beta-version of text.

@@
 
-Consider the \texttt{Laplace} concurrent function discussed above.  In
-this section we demonstrate how this function can be called in the
-normal context of a program, one that allows full access to all of the
-interior elements of the array, as well as array elements within the
-logically exterior, halo-boundary region.  Topics highlighted in this
-section are: 1. Distributed memory array allocation; 2. Explicit
-memory placement; 3. Remote memory transfer; and 4. Remote execution.
-This description is within the context of an extension of the Fortran
-2008 language.  As shorthand, these specific extensions are referred
-to as CAFe, for Coarray Fortran extensions.
+Consider the \texttt{Laplacian} concurrent function in Fig. 1.  In this section we demonstrate how
+this function can be called in the normal context of a program, one that allows full access to all
+of the interior elements of the array, as well as array elements within the logically exterior,
+halo-boundary region.  Topics highlighted in this section are: 1. distributed memory array
+allocation; 2. explicit memory placement; 3. remote memory transfer; and 4. remote execution.  This
+description is within the context of extensions to Fortran; as shorthand, these extensions are
+referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to previous work
+extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
+
+\subsection{Subimages}
+
+We introduce the important new concept of a CAFe subimage.  Fortran images are a collection of
+distributed memory processes that all run the same program (image).  LOPe extends the concept of a
+Fortran image by allowing images to be hierachical.  By this we mean that each image \emph{may}
+have a subimage (or subimages), but this subimage is not visible to other regular Fortran images.
+Subimages also execute differently than normal images and may execute on different non-homogeneous
+hardware, .e.g, an attached accelerator device.  Subimages are task based while images all execute
+a Single Program but with different (Multiple) Data (SPMD).  A task can be given to a subimage, but
+execution on the subimage terminates once the task is finished.  Memory on a subimage is permanent,
+however, and must be specifically allocated and deallocated.
+
+One obtains a subimage by executing the new LOPe function call, \texttt{device = GET\_SUBIMAGE(1)},
+where the integer argument represents an attached hardware device (or a separate process).
+If the function fails (e.g., the requested device is unavailable) it returns the image number
+\texttt{this\_image()} of the process that is executing the current program.  Returning the current
+image allows program execution to proceed correctly even if there are no attached devices.
 
@@
 
-We start with the declaration of an array (to be allocated later),
+We start with the declaration of an array with an explicit halo size and with two local
+dimensions (rank) and two distributed memory codimensions (corank),
 \begin{verbatim}
-  real, allocatable, HALO(1:*:1,1:*:1),          &
-        dimension(:,:), codimension[:,:] :: U
+   real, allocatable, dimension(:,:), codimension[:,:]      &
+         HALO(1:*:1,1:*:1) :: U
 \end{verbatim}
-with an explicit halo size and with two local dimensions (rank) and
-two distributed memory codimensions (corank).  The corank of the array
-is chosen to be identical to the rank of the array so that the logical
-process topology aligns in a way that allows a natural halo exchange
-between logically neighboring processes.\todo{What if corank and rank are not the same?}  For example, if the process
-location is \texttt{[pcol,prow]}, then the right-hand halo for the
-local array \texttt{U} can be obtained by the assignment
+The corank of the array is chosen to be identical to the rank of the array so that the logical
+process topology aligns in a way that allows a natural halo exchange between logically neighboring
+processes (this could not occur if corank and rank are not the same).  For example, if the process
+location is \texttt{[pcol,prow]}, then the right-hand halo for the local array \texttt{U} can be
+obtained by the assignment
 \texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
@@
 \texttt{allocate(U(0:M+1,0:N+1)[MP,*])}.
-This allocation statement specifies (given the one element halo size
-provided earlier for \texttt{U}) that the left halo column is
-\texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the bottom row
-is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the
-interior region \texttt{U(1:M,1:N)}.  In this allocation statement,
-the total number of process rows \texttt{NP} can be obtained at
-runtime, but \emph{may not} be explicitly provided (according to CAF
-rules) because the actual number of participating processes (in
-Fortran called images) is variable, depending on how many processes
-are requested at program startup.  In this discussion, it is
-\emph{assumed} that there are no holes in the logical processor
-topology, thus $MP*NP = P$, where $P$ is the total number of
-participating processes (images).
+This allocation statement specifies (given the one element halo size provided earlier for
+\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)},
+the bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the interior region
+\texttt{U(1:M,1:N)}.
 
-\subsection{Subimages}
+In this allocation statement, the total number of process columns $NP$ can be obtained at runtime,
+but \emph{may not} be explicitly provided (according to CAF rules) because the actual number of
+participating processes (in Fortran called images) is variable, depending on how many processes are
+requested at program startup.  In this discussion, it is \emph{assumed} that there are no holes in
+the logical processor topology, thus $MP*NP = P$, where $MP$ is the number of process rows and $P$
+is the total number of participating processes (images).
 
-So far --- except for the \texttt{HALO} attribute --- this is totally
-standard Fortran (2008).  For CAFe, we extend the concept of Fortran
-images to be hierachical.  By this we mean that each image \emph{may}
-have a subimage (or subimages), but this subimage is not visible to
-other regular Fortran images.  Subimages also execute differently than
-normal images.  Subimages are task based while images all execute a
-Single Program but with different (Multiple) Data (SPMD).  A task can
-be given to a subimage, but execution on the subimage terminates once
-the task is finished.  Memory on a subimage is permanent, however, and
-must be specifically allocated and deallocated.
-
-One obtains a subimage by executing a function call,
-\texttt{device = GET\_SUBIMAGE(1)}
-where the integer argument representing one of attached hardware
-devices\todo{mention the relation of subimages and hardware earlier in
-  defn. of subimage}.  If the function fails (e.g., the requested
-device is unavailable) it returns the image number
-\texttt{this\_image()} of the process that is executing the current
-program.  Returning the current image allows program execution to
-proceed correctly even if there are no attached devices.
-
-Once a subimage is obtained, memory on the device may be allocated,
+Once a subimage is obtained, memory on the device can be allocated,
 \begin{verbatim}
@@
 subimage has been obtained; 2. The location where memory is allocated is denoted by regular coarray
-notation \texttt{U[device]}; 3. The allocated size and halo attribute of the new array are assumed
+notation \texttt{U[device]}; 3. The allocated size and halo attribute of the new array are obtained
 from the previously allocated local array \texttt{U} via the notation \texttt{HALO\_SRC=U} (using
@@
 
-Execution of the \texttt{Laplace} task is done using the \texttt{do}
+Execution of the \texttt{Laplacian} task is done using the \texttt{do}
 \texttt{concurrent} construct:
@@
       do concurrent (i=1:M, j=1:N)   [[device]]
-         call Laplace( U(i,j)[device] )
+         call Laplacian( U(i,j)[device] )
       end do
@@
 defined above in reference to the definition of the concurrent procedure
-\texttt{Laplace}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
+\texttt{Laplacian}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
 task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
@@
 with the corresponding interior of the coarray on a logically neighboring process.
-
-%% However, recall that with the LOPe programming model only the local element is mutable; thus, the halo region of an array needs update.
-
 Finally, memory for the entire array \texttt{U} can be copied from the
@@
 
-Note that the use of halo cells is the normal way that large and complex MPI and CAF
-programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
-language allowing the compiler access to this information in order to spread computation over more
-hardware resources, improve performance, and to reduce complexity for the programmer.
+%
+% This could be part of conclusions
+%
+
+%%Note that the use of halo cells is the normal way that large and complex MPI and CAF programs are implemented.
+LOPe proposes to formalize this common pattern into the Fortran language allowing the compiler access to this information in order to spread computation over more hardware resources, improve performance, and to reduce complexity for the programmer.

### 2015-02-09 papers/LOPe/transformations.tex
Finished beta-level changes.

@@
 This section briefly describes how LOPe extensions to Fortran have been implemented as
-source-to-source transformations via rewrite rules and strategies in StrategoXT.
+source-to-source transformations via rewrite rules (for expressing basic transformations) and
+rewriting strategies (for controlling the application of the rewrite rules).  A LOPe file is
+transformed to Fortran and OpenCL files through generative programming techniques using
+Stratego/XT tools\cite{Bravenboer200852} and is accomplished in three phases: (1) parsing to
+produce a LOPe Abstract Syntax Tree (AST) represented in the Annotated Term
+Format (ATerm\cite{DBLP:journals/spe/BrandJKO00}); (2) transformations of LOPe AST nodes to Fortran
+and C AST nodes; and (3) pretty-printing to the base Fortran and OpenCL languages.
 
-Transformations are accomplished in three phases:
-1. Parsing to produce an AST nodes representated as ATerms;
-2. Transformations on the AST nodes represented as rewrite rules and strategies in Stratego/XT; and
-3. Pretty printing of transformed nodes using the pretty printing facilites of BoxLib.
+The foundation of LOPe is the syntax definition of the base language expressed in
+SDF (Syntax Definition Formalism) as part of the Open Fortran Project (OFP)\cite{OFP:git:url}.
+LOPe is defined in a separate SDF module that extends the Fortran 2008 language standard with 15
+context-free syntax rules.  Parsing is implemented in Stratego/XT by a scannerless generalized-LR
+parser and the conversion of transformed AST nodes to text is accomplished with simple
+pretty-printing rules.
 
@@
 
-A key component of the LOPe transformations is the targeting of a LOPe \texttt{CONCURRENT}
-procedure for a particular hardware (and software) architecture.  The execution target can be one
-of several choices, including serial execution by the current process (and hardware) via inlining
+A key component of code generation for LOPe is the targeting of a LOPe \texttt{CONCURRENT}
+procedure for a particular hardware architecture.  The execution target can be one
+of several choices, including serial execution by the current process via inlining
 of the function, parallel execution by inlining with OpenMP compiler directives, or parallel
-execution by a heterogeneous processing elements with a language like OpenCL, for example.
+execution by heterogeneous processing elements with a language like OpenCL.
 
-For this work we have developed rewrite strategies and rules in Stratego/XT to rewrite Fortran AST
+For this work we have developed rewrite rules and strategies in Stratego/XT to rewrite Fortran AST
 nodes to C AST nodes (extended with necessary OpenCL keywords).  The C AST ATerms have mostly a
-one-to-one correspondence with Fortran terms.  A \texttt{CONCURRENT} procedure is transformed to an
-OpenCL kernel, Fortran formal parameters are transformed to C parameters (with a direct mapping of
-types), and local Fortran variable declarations are rewritten as C declarations.  Similarly,
-Fortran executable statements are rewritten as C executable AST nodes.  Mostly the only
-complication is mapping the LOPe local array index view to the global C index space.  This
+one-to-one correspondence with Fortran terms: a \texttt{CONCURRENT} procedure is transformed to an
+OpenCL kernel; Fortran formal parameters are transformed to C parameters (with a direct mapping of
+types); and local Fortran variable declarations are rewritten as C declarations.  Similarly,
+Fortran executable statements are rewritten as C statements.  The only
+minor complication is mapping the LOPe local, array index view to the global C index space.  This
 translation is facilitated by a Fortran symbol table that stores array shape and halo size
@@
 
-Once all rewriting passes have been made to transform from LOPe Fortran ATerms to C ATerms, the
-C ATerms are output as an OpenCL file using a C pretty printer.
+\subsection{Transformations of the Calling Site}
 
-\subsection{Transformations at the Call Site}
-
-Transformations at a LOPe procedure call site are more difficult, though technically straight forward.
-The Fortran function call must be transformed to a call to run the OpenCL kernel (produced as described
-above).  This is facilitated by the use of the ForOpenCL library that provides Fortran bindings to the
-OpenCL runtime\cite{foropencl}.  But the usage requires the declaration of extra variables,
-allocation of memory on the OpenCL device (subimage), transfer of memory, marshalling of kernel
-arguments, and synchronization.
+Transformations of a LOPe procedure call site are more difficult, though technically straight
+forward.  The Fortran function call must be transformed to a call to run the OpenCL kernel
+(generated as described above).  This is facilitated by use of the ForOpenCL library which
+provides Fortran bindings to the OpenCL runtime\cite{Sottile:2013:FTE:2441516.2441520}.  However,
+this usage requires the declaration of extra variables, allocation of memory on the OpenCL device
+(subimage), transfer of memory, marshalling of kernel arguments, and synchronization.
 
 These transformations are accomplished using several rewrite stages using Stratego/XT strategies:
-1. A symbol table is produced in the first pass to store information related to arrays including,
+(1) a symbol table is produced in the first pass to store information related to arrays including,
 array shape, halo size, and allocation status;
-2. Additional variables are declared that are used to maintain the OpenCL runtime state, including
-the OpenCL device (as obtained from a \texttt{GET\_SUBIMAGE} call), the OpenCL kernel (produced as
-described above), and OpenCL variables used to marshall kernel arguments; and
-3. All CAFe code related to subimage usage is lowered to standard Fortran with calls to the
-ForOpenCL library as needed.  The latter step includes the allocation of OpenCL device memory,
-transfer of memory to and from the OpenCL device and the executing Fortran process (image) as
-indicated in the program source by the single bracket \texttt{[subimage]} notation, marshalling
-of kernel argument, running of the OpenCL kernel, and synchronization.
+(2) additional variables are declared that are used to maintain the OpenCL runtime state, including
+the OpenCL device, the OpenCL kernel, and OpenCL variables used to marshall kernel arguments; and
+(3) all CAFe code related to subimage usage is desugared to standard Fortran with calls to the
+ForOpenCL library as needed.
+%%The latter step includes the allocation of OpenCL device memory, transfer of memory to and from the OpenCL device, marshalling of kernel arguments, running of the OpenCL kernel, and synchronization.
 
@@
 execution with function inlining (if desired) will be performed by the regular Fortran compiler once
-all CAFe and LOPe code has been lowered to standard Fortran.
+all CAFe and LOPe code has been desugared (lowered) to standard Fortran.
 

### 2015-02-09 papers/LOPe/appendix-a.tex
Spelling fixes.

--- 2015-01-27 papers/LOPe/appendix-a.tex
@@
 programmer.  A subroutine is a function with no return value and must be called in a statement
-rather than as an expression.  Simple syntatic elements are: \texttt{!} the line comment symbol;
-\texttt{\&} the line continuation symbol; and \texttt{::} is syntatic sugar used for the separation
-of elements, e.g., separating type attributes from the variable list in a type declaration
+rather than as an expression.  Simple syntactic elements are: \texttt{!} the line comment symbol;
+\texttt{\&} the line continuation symbol; and \texttt{::} is syntactic sugar used for the separation
+of syntactic elements, e.g., separating type attributes from the variable list in a type declaration
 statement.

### 2015-02-09 papers/LOPe/coarray-comparison.tex
Spelling fixes.

@@
 development and memory management (synchronization between memory copies of the same logical
-array region covered by halos).  By explicitly describing the existance and size of an array's halo
+array region covered by halos).  By explicitly describing the existence and size of an array's halo
 region, the compiler is provided with enough information to manage most of the hard and detailed
@@
 
+There are several advantages to LOPe, including:
+
 \begin{itemize}
@@
 The algorithm implementation is separate from any specification as to where the array
-memory is located.  The CAF example explicity denotes where memory is located with the
-\texttt{[left]} and \texttt{[right]} syntax where left and right specifiy a processor
+memory is located.  The CAF example explicitly denotes where memory is located with the
+\texttt{[left]} and \texttt{[right]} syntax where left and right specify a processor
 topology.
@@
 The algorithm implementation is separate from any specification as to where the algorithm
-is to be executed.  The CAF example explicity denotes where a statement is to executed
+is to be executed.  The CAF example explicitly denotes where a statement is to executed
 with a control flow construct like \texttt{if (me == 1)}.
@@
 
-Contraints provided by the LOPe language extensions allow the compiler to catch several classes of
+Constraints provided by the LOPe language extensions allow the compiler to catch several classes of
 errors that otherwise would be the programmer's responsibility:

### 2015-02-09 papers/LOPe/conclusions.tex
Moved a previous paragraph for possible inclusion in conclusions.
Beta-level changes.

@@
 expressing concepts useful for stencil operations. For example, halo regions must be expressed in
-terms of the existing syntax of the language as there is no way to specify that the ``interior'' of
-an array is in any way special from an ``outside'' region.  By not providing support for stencils in
-the language, the programmer must make specific choices regarding the order of access to the data
-and in the order of operations on the data.  These choices often hide the opportunity for
-optimizations by the compiler \cite{dubey:2014:wosc}.  If the compiler had knowledge of the symantic
-intent of halo regions, it could reorder operations so that border regions were computed
-\emph{before} interior regions, allowing the the transfer of data in halo regions to overlap with
-computations on the interior.
+terms of the existing syntax of the language and there is no way to specify that the ``interior''
+of an array is in any way special from an ``outside'' region.  By not providing support for
+stencils in the language, the programmer must make specific choices regarding data-access patterns
+and the order of operations on the data.  These choices often hide the opportunity for
+optimizations by the compiler \cite{Dubey:2014:SSC:2686745.2686756}.  For example, if the compiler
+had knowledge of the semantic intent of halo regions, it could reorder operations so that border
+regions were computed \emph{before} interior regions, allowing the transfer of data in halo regions
+to overlap with computations on the interior.
 
 Just as Coarray Fortran originally extended Fortran to include domain specific knowledge (parallel
-computation) replacing MPI library routines\cite{numeric:reid}, LOPe seeks to extend Fortran by
+computation) replacing MPI library routines\cite{Numrich:1998:CFP:289918.289920}, LOPe seeks to extend Fortran by
 providing domain specific knowledge of stencil operations.
+Specifically, LOPe and CAFe together provide:
+(1) a local view of stencil operations on data that allows a complete separation of the implementation of a stencil
+algorithm with data-access patterns;
+(2) memory placement via allocation routines that allow the specification of allocation location;
+(3) task execution placement with double-bracket syntax specifying which subimage is to execute a particular operation; and
+(4) memory exchange via the \texttt{TRANSFER\_HALO} intrinsic procedure.
 
-CAFe extensions ...
-* subimages
+%
+% This could be part of conclusions
+%
 
-Specifically, LOPe and CAFe together provide:
-* A local view of stencil operations on data.  This allows a complete separation of the implementation of a stencil algorithm with data access patterns.
-* Memory placement via allocation routines allow the specification of the allocation location.
-* Task execution placement with [[ ]] specifying which subimage is to execute a particular operation.
-* Memory exchange via the \texttt{HALO\_TRANSFER} intrinsic procedure.
-* Ability to run correctly even if subimages are not available.  Similar to OpenMP.
-* Ability to associate arrays on subimages with coarrays on host thus allowing compiler to effect the halo transfers between the multiple memory levels.
-* Halo abstraction that allows for more flexibility for the compiler to retarget for different architectures.
-* A simple way to express variables as ``thread private'' as all local variables declared in a \texttt{CONCURRENT} procedure are private to the calling procedure.  This is distinguished from OpenMP.
+%%Note that the use of halo cells is the normal way that large and complex MPI and CAF programs are implemented.
 
 \subsubsection{Benefits.}
-* extends the programmer
-* restricts the programmer
-* no race conditions
-* Abstractions provide ample opportunities for compiler optimizations.
+LOPe proposes to formalize the common, halo software pattern in language syntax, thus providing the
+compiler with access to halo information in order to spread computation over more hardware
+resources, improve performance, and to reduce complexity for the programmer.  Furthermore, LOPe
+semantics provide important \emph{language restrictions} that remove the possibility of race conditions
+that occur when multiple threads have write access to overlapping data regions.
 
 \subsubsection{Limitations.}
-* no composition useful for nonlinear dynamics\cite{coella}
-* difficult (but not impossible) to implement algoriths that perform operations between arrays of
-differing sizes (multigrid implementations for example).
+LOPe only supports regular structured grids through Fortran multi-dimensional arrays and a
+corresponding multi-dimensional processor layout.  It does not allow the composability of stencils
+required by non-linear physics operators, nor does it provide automatic support for the storage of
+intermediate results resulting from multiple intermediate update steps.  Adaptive Mesh Refinement
+(AMR) Shift Calculus provides a generalized abstraction that addresses many of these concerns
+(see\cite{Dubey:2014:SSC:2686745.2686756} and references therein).  However, it may be possible to
+support AMR in LOPe through locally-structured grid methods based on the work of Berger and
+Oliger\cite{colella2007performance}.  In this instance, LOPe could be used to update the regular
+array regions in each rectangular patch.
 
-By implementing LOPe
-we have demonstrated that LOPe can be ... It remains to history to see if LOPe is general purpose
-enough to be included in Fortran or if it should remain as a special purpose preprocessing tool.
+\vspace{.1in}
+
+By implementing LOPe we have demonstrated that LOPe can be used to easily and succinctly code
+the stencil algorithms that are common to many areas of science, and furthermore, that LOPe is
+suitable for transformation to languages like OpenCL that support heterogeneous computing.  It
+remains to history to ascertain if LOPe is sufficiently general purpose to be included in a
+general-purpose programming language or if it is better suited to remain as a DSL and to be used as a
+special-purpose preprocessing tool.

### 2015-02-09 papers/LOPe/intro.tex
Spelling fixes.

@@
 
-However, in large part these languages are largely syntatic sugar for
+However, in large part these languages are largely syntactic sugar for
 message passing and do not provide a substantial decrease in code

### 2015-02-09 papers/LOPe/programming-model.tex
Removed appendix.
Spelling fixes.
Fixes to problems found by Kristi.

@@
 Then the index error has to be understood.
-As will be seen, LOPe makes it more difficult to make these errors as Fortran array intrics can be used.
+As will be seen, LOPe makes it more difficult to make these errors as Fortran array intrinsics can be used.
 In some instance, the restricted semantics of LOPe allows the compiler to catch errors (e.g., some errors involving race conditions).
 
-LOPe is a domain specific language (DSL) implemented as a small extension
-to the Fortran 2008 standard.
-Fortran was chosen as the base language for LOPe, although in
-principle, the same techniques could be applied to languages such as C
-or C++.  Fortran was chosen because it provides a
-rich array-based syntax.  Readers who are unfamiliar with Fortran
-syntax may wish to consult Appendix A, for a brief description of
-Fortran notation.
+LOPe is a domain specific language (DSL) implemented as a small extension to the Fortran 2008
+standard.  Fortran was chosen as the base language for LOPe because it provides a rich array-based
+syntax.  Although, in principle, the same techniques could be applied to languages such as C or C++.
+
+%%Readers who are unfamiliar with Fortran syntax may wish to consult Appendix A, for a brief description of Fortran notation.
 

### 2015-02-10 papers/LOPe/intro.tex
Fixes to problems found by Kristi.

--- 2015-02-07 papers/LOPe/intro.tex
@@
 
-However, in large part these languages are largely syntatic sugar for
+However, in large part these languages are mostly syntactic sugar for
 message passing and do not provide a substantial decrease in code
@@
 
-A reasonable solution given todays language limitations is to use MPI for distributed memory
+A reasonable solution given today's language limitations is to use MPI for distributed memory
 parallelism \emph{plus} OpenMP for on-chip parallelism.  This hybrid approach is taken by two
 successful, large-scale AMR multiphysics code frameworks, FLASH and Chombo; currently however,
-neither of these codes currently support hardware accelerators\cite{DBLP:journals/corr/DubeyS13}.
+neither of these codes support hardware accelerators\cite{DBLP:journals/corr/DubeyS13}.
 %%This is the solution proposed by Cray and PGI \cite{BOF_SC10} (currently Chapel does not target GPUs\cite{Brad?}).

### 2015-02-10 papers/LOPe/coarray-extensions.tex
Spelling fixes.
Fixes to problems found by Kristi.

@@
 A convolution is similar to stencil operations in computational fluid dynamics that are used to obtain spatial derivatives of state variables.
-Simple code is shown with extentions to Fortran shown in capital letters.
+Simple code is shown with extensions to Fortran shown in capital letters.
 The extensions could be provided to C or Fortran functions via compiler directives rather than explicit language syntax.
@@
 distributed memory processes that all run the same program (image).  LOPe extends the concept of a
-Fortran image by allowing images to be hierachical.  By this we mean that each image \emph{may}
+Fortran image by allowing images to be hierarchical.  By this we mean that each image \emph{may}
 have a subimage (or subimages), but this subimage is not visible to other regular Fortran images.
@@
 In this allocation statement, the total number of process columns $NP$ can be obtained at runtime,
-but \emph{may not} be explicitly provided (according to CAF rules) because the actual number of
+but \emph{may not} be explicitly provided (according to Coarray Fortran (CAF) rules) because the actual number of
 participating processes (in Fortran called images) is variable, depending on how many processes are
@@
 
-%%\subsection{Execution Symantics and Memory Management}
+%%\subsection{Execution Semantics and Memory Management}
 
@@
 %So unless a variable has the pointer or target attribute, it cannot be aliased.
-%Thus the compiler is able to agressively optimized for memory movement between
+%Thus the compiler is able to aggressively optimized for memory movement between
 %the CPU and accelerator.  However, because the design philosophy of coarrays
@@
 Also the programmer need not specify where in memory these arrays reside.
-Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardward threads).
+Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardware threads).
 This is a simple example and it is a research question as to what compiler directives would be useful for memory placement and other directions to the compiler for efficient code generation.
@@
 This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.
-Fortran also has excellent facilities for interoperability with C so that programming in a mixed language invironment is easily accomplished, including interoperability with native Fortran arrays.
+Fortran also has excellent facilities for interoperability with C so that programming in a mixed language environment is easily accomplished, including interoperability with native Fortran arrays.
 

### 2015-02-10 papers/LOPe/coarray-extensions.tex
tweaks

@@
 
-We introduce the important new concept of a CAFe subimage.  Fortran images are a collection of
+We must first introduce the important new concept of a \emph{CAFe subimage}.  Fortran images are a collection of
 distributed memory processes that all run the same program (image).  LOPe extends the concept of a
@@
 Subimages also execute differently than normal images and may execute on different non-homogeneous
-hardware, .e.g, an attached accelerator device.  Subimages are task based while images all execute
+hardware, e.g., an attached accelerator device.  Subimages are task based while images all execute
 a Single Program but with different (Multiple) Data (SPMD).  A task can be given to a subimage, but
 execution on the subimage terminates once the task is finished.  Memory on a subimage is permanent,
-however, and must be specifically allocated and deallocated.
+however, and must be explicitly allocated and deallocated.
 
@@
 \texttt{Laplacian}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
-task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
-execution of the loop body before further execution of the program proceeds; and 6. Transfer of
+task is to be taken from the subimage as noted by \texttt{[device]}; 6. All threads must finish
+execution of the loop body before further execution of the program proceeds; and 7. Transfer of
 all requisite halo regions is effected by the call to the new LOPe intrinsic function

### 2016-01-05 papers/CAFe/HIPS-paper/transformations.tex
Initial version.

--- 2015-02-09 papers/LOPe/transformations.tex
@@
 
-\subsection{Transformations of the Calling Site}
+\subsection{Transformations at the Calling Site}
 
@@
 the OpenCL device, the OpenCL kernel, and OpenCL variables used to marshall kernel arguments; and
-(3) all CAFe code related to subimage usage is desugared to standard Fortran with calls to the
+(3) all CAFe code related to subimage usage is desugared (lowered) to standard Fortran with calls to the
 ForOpenCL library as needed.
@@
 execution with function inlining (if desired) will be performed by the regular Fortran compiler once
-all CAFe and LOPe code has been desugared (lowered) to standard Fortran.
+all CAFe and LOPe code has been desugared to standard Fortran.
 
@@
 %%A thorough examination of potential performance gains using LOPe for parallelization of code is beyond the scope of this paper.
-The emphasis of this work is to explain the syntax and execution symantics of a LOPe application.
-However, preliminary work using the StrategoXT rewrite system as described indicates that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).
+The emphasis of this work is to explain the syntax and execution semantics of a LOPe application.
+However, preliminary work using the Stratego/XT rewrite system as described indicates that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).
 
@@
 %%Since many scientific codes are dominated by memory performance, including and especially stencil algorithms as they typically only involve a computation on a small locally central array element and a small overlapping halo region.
-Stencil operations frequently do not contain enough floating point operations per memory load to allow for floating point performation to operate at peak (though this is entirely application and domain specific).
-Thus we illustrate the \emph{potential} for performance by noting the latency and throughput performance of an attached GPU inconjuction with MPI distributed memory performance associated with halo transfer in Table 1.
+Stencil operations frequently do not contain enough floating point operations per memory load to allow for floating point performance to operate at peak (though this is entirely application and domain specific).
+Thus we illustrate the \emph{potential} for performance by noting the latency and throughput performance of an attached GPU in conjunction with MPI distributed memory performance associated with halo transfer in Table 1.
 

### 2016-01-06 papers/CAFe/HIPS-paper/coarray-comparison.tex

--- 2015-02-09 papers/LOPe/coarray-comparison.tex
@@
 
-We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
+We emphasize some of these advantages by comparing the \texttt{Laplacian} implementation in Fig. 1 with the
 implementation of the same algorithm from the original Numrich and Reid
@@
 
-There are several advantages to LOPe, including:
+%%There are several advantages to LOPe over the CAF example, including:
+A comparison of Fig. 1 to the CAF example suggests the following advantages:
 
@@
 LOPe requires the implementation of the algorithm to be separate from the call to effect the halo
-transfer.  Removing boundary condition specification (e.g., the cyclic boundary conditions
-implemented in the CAF example) from the algorithm allows the boundary conditions to be changed
+transfer.  Removing boundary condition specification
+%%(e.g., the cyclic boundary conditions implemented in the CAF example)
+from the algorithm allows the boundary conditions to be changed
 without changing algorithm code.
@@
 intrinsic \texttt{TRANSFER\_HALO} function.  Thus the LOPe algorithm can be run on a machine with
-many interconnected nodes, each containing hybrid processor cores.  The coarray example can only be
-run on multiple nodes without accelerator cores.
+many interconnected nodes, each containing hybrid processor cores.
+%%The coarray example can only be run on multiple nodes without accelerator cores.
 
@@
 The algorithm implementation is separate from any specification as to where the algorithm
-is to be executed.  The CAF example explicitly denotes where a statement is to executed
-with a control flow construct like \texttt{if (me == 1)}.
+is to be executed.  The CAF example explicitly denotes where a statement is to be executed
+with the control flow construct \texttt{if (me == 1)}.
 
@@
 The semantics of LOPe makes explicit management of array temporaries (e.g., \texttt{u} and
-\texttt{new\_u} by the programmer unnecessary (though still possible).  Because in LOPe the
+\texttt{new\_u}) by the programmer unnecessary.
+%%(though still possible).
+Because in LOPe the
 halo region is a language construct, the compiler is better able to manage temporary
@@
 implicit in the \texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided
-with a copy of it's local array element plus halo that is consistent with the state of memory
+with a copy of its local array element plus halo that is consistent with the state of memory
 \emph{at the time of invocation of the procedure}.  Stores to an individual thread's local array

### 2016-01-06 papers/CAFe/HIPS-paper/programming-model.tex
Saving partial changes.

--- 2015-02-09 papers/LOPe/programming-model.tex
@@
 
-There are only a few syntax additions required for a LOPe program.
+There are only a few syntactic additions required for a LOPe program.
 These additions include syntax for describing halo regions and
@@
 keywords that are either new or that acquire new usage.
-
-\subsubsection{Halo regions.}
-The principle semantic element of LOPe is the concept of a halo.
-A halo is an ``artificial'' or ``virtual'' region surrounding
-an array that contains boundary-value information.  Halo (also called
-ghost-cell) regions are commonly employed to unify array indexing
-schemes in the vicinity of an array boundary so that an array may be
-referenced using indices that fall ``outside'' of the logical domain
-of the array.  In LOPe, the halo region is given explicit syntax so
-that the compiler can exploit this information for purposes of memory
-allocation, data replication and thread synchronization.  For example,
-a halo region can be declared with a statement of the form,
-
-\begin{verbatim}
-  real, allocatable, dimension(:), HALO(1:*:1) :: A
-\end{verbatim}
-This statement indicates that \texttt{A} is a rank one array, will be
-allocated later, and
-has a halo region of one element surrounding the array on either side.
-The halo notation \texttt{M:*:N} specifies a halo
-of \texttt{M} elements to the left, \texttt{N} elements to the
-right, and an arbitrary number of ``interior'' array elements.
-When used to describe a formal parameter of a
-function, such as the type-declaration statement, \texttt{real, HALO(:,:) :: U},
-the halo size is inferred by the compiler
-from the actual array argument provided at the
-calling site of the function.
-
-%%In LOPe, there is no need (in this instance) for a repetitive \texttt{dimension(:,:)} specification, as it is inferred from the \texttt{HALO} specification.
-
-\subsubsection{Concurrent functions.}
-
-The second keyword employed by LOPe is \texttt{concurrent} which already exists in the form of a
-\texttt{do} \texttt{concurrent} loop, whereby the programmer asserts that specific
-iterations of the loop body may be executed by the compiler in \emph{any order,} even
-\emph{concurrently.}  LOPe allows a function with the attributes \texttt{pure} (assertion of no
-side effects) and \texttt{concurrent} (assertion of no dependencies between iterations) to be
-called from within a \texttt{do} \texttt{concurrent} loop.  An example of a LOPe function is
-shown in Fig. 1 and an example calling this function will be provided later in the text.  One
-should imagine that a LOPe function is called \emph{for each} \texttt{i,j} index of the interior of
-the array \texttt{U}.  Note that this usage introduces a race condition as new values of
-elements of \texttt{U} are created on the left-hand side of the assignment statement that may use
-\emph{new or old} values of \texttt{U} on the right-hand side.  LOPe requires the compiler to
-guarantee that race conditions won't occur by using, e.g., double-buffering techniques as needed.
 
@@
 \vspace{-.3in}
-
-\subsubsection{LOPe index notation.}
-
-In the \texttt{Laplacian} example the \texttt{U(0,0)} array element is the \emph{local} array
-element and only the local element may be modified.  This zero-based indexing for the local-array
-element differs from conventional Fortran, where by default, array indices start at 1.  The use of
-zero-based indexing gives a clean symmetry for indices on either side of the central element at
-zero. The other array elements are in the halo region and are \texttt{U(-1,0)} and
-\texttt{U(+1,0)} (left and right of local, respectively) and \texttt{U(0,-1)} and \texttt{U(0,+1)}
-(below and above of local).  The geometric positioning of the array elements can be
-seen by examining the arrangement of the expressions on the right-hand side of Fig. 1.

### 2016-01-07 papers/CAFe/code/OpenCL/poisson-multigrid_3d_cl.F90
Working on 3D implementation of Restrict, added some notes about dimensions

@@ -1,8 +1,8 @@
 #define DUMP_OUTPUT
 #undef DO_HALO_EXCHANGE
 #undef DO_PROLONGATE
-#undef DO_RESTRICT
-#define DO_RELAX
+#define DO_RESTRICT
+#undef DO_RELAX
 
 PROGRAM PoissonMultigrid
 USE ForOpenCL
@@ -13,7 +13,7 @@ IMPLICIT NONE
 REAL, PARAMETER :: w = (2.0/3.0)
 INTEGER, PARAMETER :: N = 32
 INTEGER, PARAMETER :: M = 32
-INTEGER, PARAMETER :: L = 16
+INTEGER, PARAMETER :: L = 32
 INTEGER, PARAMETER :: NP = 2
 INTEGER, PARAMETER :: MP = 2
 INTEGER, PARAMETER :: LP = 2
@@ -38,7 +38,7 @@ INTEGER(KIND=cl_int) :: cl_status__
 INTEGER(KIND=c_size_t) :: cl_size__
 INTEGER(KIND=c_size_t) :: cl_gwo__(3)
 INTEGER(KIND=c_size_t) :: cl_gws__(3)
-INTEGER(KIND=c_size_t) :: cl_lws__(3) = [16,8,1]
+INTEGER(KIND=c_size_t) :: cl_lws__(3) = [16,8,8]
 
 !! Device id
 !
@@ -108,7 +108,7 @@ cl_status__ = writeBuffer(cl_V1h_,C_LOC(V1h),cl_size__)
 #ifdef DUMP_OUTPUT
 CALL Textual_Output_3D(N,M,L,V1h,"1h_0")
 #endif
-#define DO_RELAX
+
 !! level 1h
 !
 DO t = 1, nsteps
@@ -125,6 +125,8 @@ DO t = 1, nsteps
   cl_gws__ = focl_global_size(1,cl_lws__,cl_gws__,[1,1,1])
   cl_gws__ = focl_global_size(3,cl_lws__,cl_gws__,[((N+1-(-1))+1),((M+1-(-1))+1),((L+1-(-1))+1)])
   cl_gws__ = focl_global_size(3,cl_lws__,cl_gws__,[((N+1-(-1))+1),((M+1-(-1))+1),((L+1-(-1))+1)])
+cl_gws__ = [32,24,24]
+print *, "------------------------------"
 print *, cl_gwo__
 print *, cl_gws__
[…]

### 2016-01-07 papers/CAFe/HIPS-paper/coarray-extensions.tex

--- 2015-02-10 papers/LOPe/coarray-extensions.tex
@@
 
-%%We highlight the Fortran elemental source-code abstraction because it provides a local orientation similar to that of OpenCL or CUDA kernel functions that were designed to exploit streaming, highly-threaded architectures like GPUs.
-Elemental functions are pure in that they are guaranteed to be free of side effects such as I/O.
-Consider a convolution where a 3x3 filter is applied to individual pixel elements in a photograph to average of blur the original photograph.
-A convolution is similar to stencil operations in computational fluid dynamics that are used to obtain spatial derivatives of state variables.
-Simple code is shown with extensions to Fortran shown in capital letters.
-The extensions could be provided to C or Fortran functions via compiler directives rather than explicit language syntax.
-
-Consider the \texttt{Laplacian} concurrent function in Fig. 1.  In this section we demonstrate how
-this function can be called in the normal context of a program, one that allows full access to all
-of the interior elements of the array, as well as array elements within the logically exterior,
-halo-boundary region.  Topics highlighted in this section are: 1. distributed memory array
-allocation; 2. explicit memory placement; 3. remote memory transfer; and 4. remote execution.  This
-description is within the context of extensions to Fortran; as shorthand, these extensions are
-referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to previous work
-extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
+Topics highlighted in this section are: 1. Distributed memory array allocation;
+2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
+description is within the context of extensions to Fortran; as shorthand, these extensions
+are referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to
+previous work extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
 
@@
 
-We must first introduce the important new concept of a \emph{CAFe subimage}.  Fortran images are a collection of
-distributed memory processes that all run the same program (image).  LOPe extends the concept of a
-Fortran image by allowing images to be hierarchical.  By this we mean that each image \emph{may}
-have a subimage (or subimages), but this subimage is not visible to other regular Fortran images.
-Subimages also execute differently than normal images and may execute on different non-homogeneous
-hardware, e.g., an attached accelerator device.  Subimages are task based while images all execute
-a Single Program but with different (Multiple) Data (SPMD).  A task can be given to a subimage, but
-execution on the subimage terminates once the task is finished.  Memory on a subimage is permanent,
-however, and must be explicitly allocated and deallocated.
+We must first introduce the important new concept of a \emph{CAFe subimage}.  Fortran
+images are a collection of distributed memory processes that all run the same program
+(image).  CAFe extends the concept of a Fortran image by allowing images to be
+hierarchical.  By this we mean that each image \emph{may} have a subimage (or subimages),
+but this subimage is not visible to other regular Fortran images.  Subimages also execute
+differently than normal images and may execute on different non-homogeneous hardware,
+e.g., an attached accelerator device.  Subimages are task based while images all execute a
+Single Program but with different (Multiple) Data (SPMD).  A task can be given to a
+subimage, but execution on the subimage terminates once the task is finished.  Memory on a
+subimage is permanent, however, and must be explicitly allocated and deallocated.
 
-One obtains a subimage by executing the new LOPe function call, \texttt{device = GET\_SUBIMAGE(1)},
-where the integer argument represents an attached hardware device (or a separate process).
-If the function fails (e.g., the requested device is unavailable) it returns the image number
-\texttt{this\_image()} of the process that is executing the current program.  Returning the current
-image allows program execution to proceed correctly even if there are no attached devices.
+One obtains a subimage by executing a new CAFe function,
+\begin{verbatim}
+   device = GET_SUBIMAGE(device_id)
+\end{verbatim}
+where the integer argument represents an attached hardware device (or a separate process
+or a team of threads).  If the function fails (e.g., the requested device is unavailable)
+it returns the local image number of the process that is executing the
+current program, obtained by a call to the Fortran 2008 function \texttt{this\_image()}.
+Returning the current image (similar to MPI rank) allows program execution to proceed
+correctly even if there are no attached devices.
+
+Having obtained an subimage, we now consider how a subimage can be used within a program
+to allocate memory, transfer memory, and execute tasks.
+
+\subsection{Distributed Memory Array Declaration}
+
+In Fortran, arrays that are visible to other program images must be declared with the
+codimension attribute, for example,
+\begin{verbatim}
+ real, allocatable :: U(:,:)[:], V(:,:)[:]
+\end{verbatim}
+declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two
+and a corank of 1.  In Fortran, square brackets \texttt{[ ]} are used to denote operations
+(possibly expensive) on distributed memory where parentheses \texttt{( )} are used to
+denote array elements; square brackets denote array location.
+
+When declared thusly, coarrays must be allocated on all program images.  CAFe extends this
+declaration to include memory that \emph{may} be allocated on a subimage, but does not
+need to be allocated unless it is specifically used by a task executing on a subimage.
+
+\subsection{Explicit Memory Placement}
+
+As discussed, CAFe allows coarray memory to be conditionally allocated on a subimage.
+It should be allocated conditionally because the requested subimage in a texttt{GET\_SUBIMAGE}
+call may not be available.  In addition, not all coarrays need be allocated on each
+subimage.  For example, the following code segment allocates memory for coarrays
+\texttt{U} and \texttt{V} on all images, but only coarrays \texttt{U} is allocated on
+the subimage specified by \texttt{device}:
+\begin{verbatim}
+   allocate(U(M,N)[*], V(M,N)[*])
+   if (device \= this_image()) then
+     allocate(U(M,N)[*])  [[device]]
+   end if
+\end{verbatim}
+where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement is
+mandatory and allocates the declared coarrays on all program images.  The use of the
+\texttt{*} symbol is used for the last corank dimension in CAF (in this case there is only
+one) to allow for the number of program images to be a runtime parameter, which can be
+ascertained with a \texttt{num\_images()} function call.  If \texttt{device} is
+available, its numeric value will be different from \texttt{this\_image()} and thus
+coarray \texttt{U} will also be allocated on the device.  Note that placement of the
+memory is specified by the use of double square bracket notation \texttt{[[ ]]}.  Like
+single square brackets, double square brackets indicate something special (and possibly
+expensive) is to occur, in this case possibly remote memory allocation.
+
+\subsection{Remote Memory Transfer}
+
+Once memory is allocated on the device, it can be initialized by copying memory from the
+hosting image to the device.  This is explicitly done with normal CAF syntax.  For example,
+\begin{verbatim}
+  U[device] = 3.14
+\end{verbatim}
+assigns 3.14 to all of the array elements on \texttt{device}.  Specific array elements
+can also be transferred, 
+\begin{verbatim}
+  U(1,:) = U(1,:)[device]
+  U(M,:) = U(M,:)[device]
+\end{verbatim}
+where the first and last columns of \texttt{U} are copied from the device to the hosting image.
+
+CAFe restricts memory transfer between a subimage and its hosting image only. Transferring
+memory to another images subimage is not allowed.  In addition, memory transfer can only be
+initiated by the hosting image; code executing on a subimage cannot use coarray notation to
+transfer memory.
+
+\subsection{Remote Execution}
+
+The final CAFe concept introduced is that of remote execution.  As discussed, in Fortran 2008
+all images (similar to MPI ranks) execute the same program.  The only way for an image
+to execute something different is to conditionally execute a block of code based on explicitly
+checking that an executing image's rank (\texttt{this\_image()}) meets some established criteria.
+There exists no mechanism for one image to execute a block of code or a procedure on another
+image.
+
+CAFe allows images to execute tasks on hosted subimages using standard procedure calls.
+For example,
 
@@
 
-We start with the declaration of an array with an explicit halo size and with two local
-dimensions (rank) and two distributed memory codimensions (corank),
+We start with the declaration of an array \texttt{U} with two local dimensions (rank) and
+two distributed memory codimensions (corank).
 \begin{verbatim}
-   real, allocatable, dimension(:,:), codimension[:,:]      &
-         HALO(1:*:1,1:*:1) :: U
+   real, allocatable, :: U(:,:)[:,:]
 \end{verbatim}
@@
 execution of the loop body before further execution of the program proceeds; and 7. Transfer of
-all requisite halo regions is effected by the call to the new LOPe intrinsic function
+all requisite halo regions is effected by the call to the new CAFe intrinsic function
 \texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must

### 2016-01-07 papers/CAFe/HIPS-paper/intro.tex

--- 2015-02-10 papers/LOPe/intro.tex
@@
 \section{Introduction}
+
+A goal of high-level programming languages should be to \emph{allow} a programmer to
+develop software for a specific hardware architecture, without actually \emph{exposing} hardware
+features.  In fact, allowing low-level hardware details to be controlled by the language is actually
+a barrier to software performance portability.  The path forward for high-level
+programming languages should be to allow a programmer to provide more symantic information
+about the intent of the programmer, thus allowing the compiler more freedom to retarget
+applications to new architectures.
+
+This becomes critical as the scientific community tries to achieve exascale performance.
+Energy is a major concern, thus, heterogenous archtectures with many simple, low-energy cores.
+
+There is a plethera of options available to an HPC programmer beyond just using a serial
+language with MPI.  New parallel languages are under development, for example, Chapel and
+the PGAS languages including Coarray Fortran (CAF).  Options for heterogeneous computing
+include usage of OpenMP, OpenACC, CUDA, CUDA Fortran, and OpenCL for programming attached
+accelerator devices.
+
+A reasonable solution given today's language limitations is to use MPI for distributed memory
+parallelism \emph{plus} OpenMP for on-chip parallelism.  This hybrid approach is taken by two
+successful, large-scale AMR multiphysics code frameworks, FLASH and Chombo; currently however,
+neither of these codes support hardware accelerators\cite{DBLP:journals/corr/DubeyS13}.
+%%This is the solution proposed by Cray and PGI \cite{BOF_SC10} (currently Chapel does not target GPUs\cite{Brad?}).
+Other choices for expressing on-chip parallelism are OpenCL~\cite{opencl11} and NVIDIA's CUDA.
+Unfortunately, achieving high performance using either of these two languages can be a daunting
+challenge and there are no guarantees that either language is suitable for future hardware
+generations.
+
+OpenACC directives and features:
+\begin{verbatim}
+● acc_copy(in|out)( ptr, bytes )
+● acc_create( ptr, bytes )
+● acc_delete( ptr, bytes )
+● acc_is_present( ptr, bytes )
+● acc_update_(device|local)( ptr, bytes )
+● acc_deviceptr( ptr )
+● acc_hostptr( devptr )
+● acc_[un]map_data( devptr, hostptr, bytes )
+● acc_memcpy_(to|from)_device
+\end{verbatim}
+
+With the addition of coarrays in the 2008 standard, Fortran is now a \emph{parallel}
+language.  Originally called Coarray Fortran (CAF) by its
+authors\cite{Numrich:1998:CFP:289918.289920}, parallism in Fortran is similar to MPI in
+that all cooperating CAF processes (called images) execute a Single Program but with
+different (Multiple) Data (SPMD).  Key concepts in CAF are parallel execution, distributed
+memory allocation, and remote memory transfer.  Parallel execution and memory allocation
+are straightforward as each program is replicated $NP$ times with each program image
+allocating its own memory locally.  Special syntax is provided to indicate which arrays
+are visible between distributed memory processes and remote memory transfer is indicated
+with new array syntax using square brackets, for example,
+\begin{verbatim}
+   U(1) = U(1)[3]
+\end{verbatim}
+copies memory from the first element of \texttt{U} residing on image number 3 into
+the corresponding element of the \emph{local} array \texttt{U}.
+
+Since each CAF process executes the same program, there are no good avenues for exploiting
+task-based parallelism.  Neither is there a mechanism for dynamic coarray memory
+allocation, nor a mechanism for the remote execution of functions.  In CAF, like MPI, the
+programmer often combines distributed memory processes running the same program image with a
+shared memory threaded programming model using OpenMP (or OpenACC).
+
+In this paper we examine...
+
+Topics highlighted in this section are: 1. Distributed memory array allocation;
+2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
+description is within the context of extensions to Fortran; as shorthand, these extensions
+are referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to
+previous work extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
 
@@
 
-A reasonable solution given today's language limitations is to use MPI for distributed memory
-parallelism \emph{plus} OpenMP for on-chip parallelism.  This hybrid approach is taken by two
-successful, large-scale AMR multiphysics code frameworks, FLASH and Chombo; currently however,
-neither of these codes support hardware accelerators\cite{DBLP:journals/corr/DubeyS13}.
-%%This is the solution proposed by Cray and PGI \cite{BOF_SC10} (currently Chapel does not target GPUs\cite{Brad?}).
-Other choices for expressing on-chip parallelism are OpenCL~\cite{opencl11} and NVIDIA's CUDA.
-Unfortunately, achieving high performance using either of these two languages can be a daunting
-challenge and there are no guarantees that either language is suitable for future hardware
-generations.
-
 In this paper we examine a language-based paradigm that allows the

### 2016-01-07 papers/CAFe/HIPS-paper/coarray-extensions.tex
Added text about remote execution.

@@
 
-The final CAFe concept introduced is that of remote execution.  As discussed, in Fortran 2008
-all images (similar to MPI ranks) execute the same program.  The only way for an image
-to execute something different is to conditionally execute a block of code based on explicitly
-checking that an executing image's rank (\texttt{this\_image()}) meets some established criteria.
-There exists no mechanism for one image to execute a block of code or a procedure on another
-image.
+The final CAFe concept introduced is that of remote execution.  As discussed above, in
+Fortran all CAF images (similar to MPI ranks) execute the same program.  The only way for
+an image to execute something different is to conditionally execute a block of code based
+on explicitly checking that an executing image's rank (\texttt{this\_image()}) meets some
+established criterion.  There currently exists no mechanism for one image to execute a
+block of code or a procedure on another image.
 
@@
 For example,
+\begin{verbatim}
+  call relax(U[device])  [[device]]
+\end{verbatim}
+executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
+on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
+whereby the double square brackets notation indicates \emph{where} the function will execute.
+Funtions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
+keyword \texttt{pure}) and scalar formal parameters must be passed by value (i.e., declared with the
+\texttt{value} attribute).
+
+CAFe expands on CAF execution semantics by providing a task-based mechanism, where tasks
+are defined in terms of pure Fortran procedures.  Since the procedures must be declared
+as pure, they cannot perform any I/O or call any impure procedures; nor can they initiate
+any communication between the subimage and the hosting image.
+
+CAFe subimage tasks can also be defined by the body of a Fortran do concurrent construct
+and executed on a subimage.  For example, the code segment,
+\begin{verbatim}
+do concurrent(i=1:N-1)  [[device]]
+ begin block
+   real :: w = 0.67
+   S(i) = (1-w)*T(i) + w*(T(i-1)+T(i+1)/2)
+ end block
+end do
+\end{verbatim}
+will execute all of the code within the \texttt{do concurrent} construct on device.  In
+Fortran, the execution semantics of a \texttt{do concurrent} loop is that the loop
+iterates can be executed concurrently in any order.
+
+Note the similarities between this example and the OpenMP pragma \texttt{!\$omp parallel do,
+private(w)}.  The use of the Fortran block construct to declare private
+variables allows a natural and clear way to distinguish between variables that are private
+to a ``thread'' (although in the example above, the ``private'' distinction for the
+variable \texttt{w} is unnecessary because it is a constant for all loop iterations).
+(NOTE FOR CRAIG: need to read the standard to see if a block construct can be used in this
+way. ALSO, there is a lot more work here to think about how do concurrent merges with
+OpenMP and OpenACC with regards to synchronization, barriers, teams, atomics,
+, critical sections, ...)
 

### 2016-01-08 papers/CAFe/HIPS-paper/intro.tex

--- 2015-02-10 papers/LOPe/intro.tex
@@
 \section{Introduction}
+
+A goal of high-level programming languages should be to \emph{allow} a programmer to
+develop software for a specific hardware architecture, without actually \emph{exposing} hardware
+features.  In fact, allowing low-level hardware details to be controlled by the programmer is actually
+a barrier to software performance portability: the explicit use of low-level hardware details 
+leads to code becoming excessively specialized to a given target system.  The path forward for high-level
+programming languages should be to allow a programmer to provide more semantic information
+about their \emph{intent}, thus allowing the compiler more freedom to choose how to instantiate this
+intent in a specific implementation in order to retarget applications to new architectures.
+
+Performance portability (or, even simply portability) has been a perpetual problem in computing,
+especially within the high performance computing community in which relatively new and immature
+emerging technologies are constantly being tried and tested.  As processor architectures hit limitations due
+to physical and manufacturing constraints, we are seeing a trend towards more diversity in
+processor designs than were experienced for the bulk of the 1990s and 2000s.  Portability, especially
+performance portability, is becoming more challenging as moving to a new CPU may entail more than simple
+tuning if the architecture is fundamentally different than previous ones.
+
+A number of options are available to an HPC programmer beyond just using a serial
+language with MPI or OpenMP.  New parallel languages have been under development for over a decade, 
+for example, Chapel and the PGAS languages including Coarray Fortran (CAF).  Options for 
+heterogeneous computing
+include usage of OpenACC, CUDA, CUDA Fortran, and OpenCL for programming attached
+accelerator devices.  Each of these choices provides the programmer with abstractions over parallel
+systems that expose different levels of detail about the specific systems to compile to, and as such,
+entail different degrees of effort to design for and realize performance portability.
+
+A reasonable solution given today's language limitations is to use MPI for distributed memory
+parallelism \emph{plus} OpenMP for on-chip parallelism.  This hybrid approach is taken by two
+successful, large-scale AMR multiphysics code frameworks, FLASH and Chombo; currently however,
+neither of these codes support hardware accelerators\cite{DBLP:journals/corr/DubeyS13}.
+%%This is the solution proposed by Cray and PGI \cite{BOF_SC10} (currently Chapel does not target GPUs\cite{Brad?}).
+Other choices for expressing on-chip parallelism are OpenCL~\cite{opencl11} and NVIDIA's CUDA.
+Unfortunately, achieving high performance using either of these two languages can be a daunting
+challenge and there are no guarantees that either language is suitable for future hardware
+generations.
+
+% OpenACC directives and features:
+%\begin{verbatim}
+%● acc_copy(in|out)( ptr, bytes )
+%● acc_create( ptr, bytes )
+%● acc_delete( ptr, bytes )
+%● acc_is_present( ptr, bytes )
+%● acc_update_(device|local)( ptr, bytes )
+%● acc_deviceptr( ptr )
+%● acc_hostptr( devptr )
+%● acc_[un]map_data( devptr, hostptr, bytes )
+%● acc_memcpy_(to|from)_device
+%\end{verbatim}
+
+With the addition of coarrays in the 2008 standard, Fortran is now a \emph{parallel}
+language.  Originally called Coarray Fortran (CAF) by its
+authors\cite{Numrich:1998:CFP:289918.289920}, parallism in Fortran is similar to MPI in
+that all cooperating CAF processes (called \emph{images}) execute a Single Program but with
+different (Multiple) Data (SPMD).  Key concepts in CAF are parallel execution, distributed
+memory allocation, and remote memory transfer.  Parallel execution and memory allocation
+are straightforward as each program is replicated $NP$ times with each program image
+allocating its own memory locally.  Special syntax is provided to indicate which arrays
+are visible between distributed memory processes and remote memory transfer is indicated
+with new array syntax using square brackets, for example,
+\begin{verbatim}
+   U(1) = U(1)[3]
+\end{verbatim}
+copies memory from the first element of \texttt{U} residing on image number 3 into
+the corresponding element of the \emph{local} array \texttt{U}.  The fundamental programming
+model is similar to that provided by message passing systems.
+
+Since each CAF process executes the same program, there are no good avenues for exploiting
+task-based parallelism.  Neither is there a mechanism for dynamic coarray memory
+allocation, nor a mechanism for the remote execution of functions.  In CAF, like MPI, the
+programmer often combines distributed memory processes running the same program image with a
+shared memory threaded programming model using OpenMP (or OpenACC).  This model breaks down
+when dealing with architectures that are not isomorphic to the model for which MPI and
+relatives were designed -- a distributed collection of homogenous processing elements in
+which each processing element supports a small number of parallel threads of execution.  For
+example, an HPC system that makes use of traditional multicore CPUs as well as GPGPU accelerators
+breaks this model and, to effectively program it, currently requires a mixture of programming
+systems in a single application.
+
+In this paper we propose CAFe: Co-Array Fortran Extensions.  These extensions to CAF introduce
+relatively minor additions to the existing coarray portion of the Fortran language.  They
+include the concept of subimages in order to nest coarray images, semantics for distributed
+array declarations, memory placement, remote memory transfer, and remote task execution.  CAFe
+builds upon the existing standardized coarray features of Fortran.  We will introduce the
+CAFe extensions that are proposed to define their semantics, demonstrate their use in simple
+examples, and provide experimental results demonstrating their performance in a prototype implementation.
+CAFe is complementary to
+previous work extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
+
+%Topics highlighted in this section are: 1. Distributed memory array allocation;
+%2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
+%description is within the context of extensions to Fortran; as shorthand, these extensions
+%are referred to as CAFe, for Coarray Fortran extensions.
+
+%%% MJS: stopped editing here 1/7/2016
 

### 2016-01-09 papers/CAFe/HIPS-paper/multigrid.tex
Initial version.

--- 2015-02-10 papers/LOPe/coarray-extensions.tex
@@
-\section{Coarray Fortran Extensions}
+\section{Multigrid Laplacian Implementation}
 
-%%We highlight the Fortran elemental source-code abstraction because it provides a local orientation similar to that of OpenCL or CUDA kernel functions that were designed to exploit streaming, highly-threaded architectures like GPUs.
-Elemental functions are pure in that they are guaranteed to be free of side effects such as I/O.
-Consider a convolution where a 3x3 filter is applied to individual pixel elements in a photograph to average of blur the original photograph.
-A convolution is similar to stencil operations in computational fluid dynamics that are used to obtain spatial derivatives of state variables.
-Simple code is shown with extensions to Fortran shown in capital letters.
-The extensions could be provided to C or Fortran functions via compiler directives rather than explicit language syntax.
+This section describes the Laplacian multigrid example.
 
-Consider the \texttt{Laplacian} concurrent function in Fig. 1.  In this section we demonstrate how
-this function can be called in the normal context of a program, one that allows full access to all
-of the interior elements of the array, as well as array elements within the logically exterior,
-halo-boundary region.  Topics highlighted in this section are: 1. distributed memory array
-allocation; 2. explicit memory placement; 3. remote memory transfer; and 4. remote execution.  This
-description is within the context of extensions to Fortran; as shorthand, these extensions are
-referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to previous work
-extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
-
-\subsection{Subimages}
-
-We must first introduce the important new concept of a \emph{CAFe subimage}.  Fortran images are a collection of
-distributed memory processes that all run the same program (image).  LOPe extends the concept of a
-Fortran image by allowing images to be hierarchical.  By this we mean that each image \emph{may}
-have a subimage (or subimages), but this subimage is not visible to other regular Fortran images.
-Subimages also execute differently than normal images and may execute on different non-homogeneous
-hardware, e.g., an attached accelerator device.  Subimages are task based while images all execute
-a Single Program but with different (Multiple) Data (SPMD).  A task can be given to a subimage, but
-execution on the subimage terminates once the task is finished.  Memory on a subimage is permanent,
-however, and must be explicitly allocated and deallocated.
-
-One obtains a subimage by executing the new LOPe function call, \texttt{device = GET\_SUBIMAGE(1)},
-where the integer argument represents an attached hardware device (or a separate process).
-If the function fails (e.g., the requested device is unavailable) it returns the image number
-\texttt{this\_image()} of the process that is executing the current program.  Returning the current
-image allows program execution to proceed correctly even if there are no attached devices.
-
-\subsection{CAFe Example}
-
-We start with the declaration of an array with an explicit halo size and with two local
-dimensions (rank) and two distributed memory codimensions (corank),
+We start with the declaration of an array \texttt{U} with two local dimensions (rank) and
+two distributed memory codimensions (corank).
 \begin{verbatim}
-   real, allocatable, dimension(:,:), codimension[:,:]      &
-         HALO(1:*:1,1:*:1) :: U
+   real, allocatable, :: U(:,:)[:,:]
 \end{verbatim}
@@
 %%\end{verbatim}
-
-%%\subsection{Execution Semantics and Memory Management}
-
-%%This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.
-It describes what is the compiler's responsibility and what is the programmer's responsibility.
-
-%% Completion of a do concurrent construct indicates that all executing threads (if a threading model is used by the compiler) have completed and that all memory on the executing subimage is in a consistent state.
-
-%%The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.
-The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.
-In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.
-
-%% A note from a conversation with Matt regarding memory management
-%%
-
-%The Fortran language has much tighter restrictions on aliasing than does C.
-%So unless a variable has the pointer or target attribute, it cannot be aliased.
-%Thus the compiler is able to aggressively optimized for memory movement between
-%the CPU and accelerator.  However, because the design philosophy of coarrays
-%is that memory transfer between images can be expensive, the programmer must
-%explicitly transfer memory between images with explicit syntax with square
-%bracket notation, i.e., $a[1] = a[2]$.  So we allow the compiler to manage
-%memory within an image but require the use of {\tt halo\_exchange} for the
-%transfer of halo memory between images.
-
-%Fortran currently supports:
-
-%1.
-Array syntax: e.g., C = A + B, where A, B, C are arrays.
-Note that this is implicitly a loop structure, but that no loop indices need be provided.
-Also the programmer need not specify where in memory these arrays reside.
-Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardware threads).
-This is a simple example and it is a research question as to what compiler directives would be useful for memory placement and other directions to the compiler for efficient code generation.
-
-%2.
-Pure procedures:  Fortran has syntax for specifying procedures that have no side effects during execution.
-Specifying code that is side-effect free code is important information to provide to the compiler so that it can generate efficient multi-threaded code.
-
-%3.
-Pure elemental procedures: Fortran has syntax for specifying procedures that take only scalar arguments, but may be applied across array elements.
-Elemental procedures are ideal for writing code to be executed within a hardware thread.
-They resemble OpenCL kernels, but are simpler because they leave all indexing up to the compiler.
-
-%We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.
-This additional syntax is provided by functions that return a copy of a small region of memory surrounding an array element (as seen within an elemental procedure) and with functions for thread synchronization.
-This additional syntax will allow pure procedures to perform stencil and other convolution-like operations on a copy of memory, synchronize, then store the computed results back to the array element associated with the given thread.
-
-%In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.
-This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.
-Fortran also has excellent facilities for interoperability with C so that programming in a mixed language environment is easily accomplished, including interoperability with native Fortran arrays.
-
-\input{coarray-comparison}
-
-%
-% This could be part of conclusions
-%
-
-%%Note that the use of halo cells is the normal way that large and complex MPI and CAF programs are implemented.
-LOPe proposes to formalize this common pattern into the Fortran language allowing the compiler access to this information in order to spread computation over more hardware resources, improve performance, and to reduce complexity for the programmer.

### 2016-01-09 papers/CAFe/HIPS-paper/performance.tex
Initial version.

\section{Performance Measurements}

This section shows and compares performance measurements for: 1. Memory bandwidth usage on GPU; 2. Computaional time on GPU; 3. Halo Communication time (total plus breakdown to and from device and MPI); 4. Total time scaling on ACISS; 5. Comparison with OpenACC (hopefully).

### 2016-01-09 papers/CAFe/HIPS-paper/coarray-extensions.tex
Moved old LOPe/CAFe example to multigrid.tex (requires lots of editing).

@@
 
+Deallocation of subimage memory is similar to allocation,
+\begin{verbatim}
+   if (device \= this_image()) then
+     deallocate(U)  [[device]]
+   end if
+\end{verbatim}
+
 \subsection{Remote Memory Transfer}
@@
 
-\subsection{CAFe Example}
+\subsection{Integration with Coarray Standard Features}
 
-We start with the declaration of an array \texttt{U} with two local dimensions (rank) and
-two distributed memory codimensions (corank).
-\begin{verbatim}
-   real, allocatable, :: U(:,:)[:,:]
-\end{verbatim}
-The corank of the array is chosen to be identical to the rank of the array so that the logical
-process topology aligns in a way that allows a natural halo exchange between logically neighboring
-processes (this could not occur if corank and rank are not the same).  For example, if the process
-location is \texttt{[pcol,prow]}, then the right-hand halo for the local array \texttt{U} can be
-obtained by the assignment
-\texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
-\texttt{U} are given by the allocation statement,
-\texttt{allocate(U(0:M+1,0:N+1)[MP,*])}.
-This allocation statement specifies (given the one element halo size provided earlier for
-\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)},
-the bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the interior region
-\texttt{U(1:M,1:N)}.
+Want to address: Barriers and teams primarily
 
-In this allocation statement, the total number of process columns $NP$ can be obtained at runtime,
-but \emph{may not} be explicitly provided (according to Coarray Fortran (CAF) rules) because the actual number of
-participating processes (in Fortran called images) is variable, depending on how many processes are
-requested at program startup.  In this discussion, it is \emph{assumed} that there are no holes in
-the logical processor topology, thus $MP*NP = P$, where $MP$ is the number of process rows and $P$
-is the total number of participating processes (images).
+%\input{coarray-comparison}
 
-Once a subimage is obtained, memory on the device can be allocated,
-\begin{verbatim}
-   if (device /= this_image()) then
-      allocate(U[device], HALO_SRC=U)   [[device]]
-   end if
-\end{verbatim}
-There are four points to note regarding this memory allocation: 1. Memory is only allocated if a
-subimage has been obtained; 2. The location where memory is allocated is denoted by regular coarray
-notation \texttt{U[device]}; 3. The allocated size and halo attribute of the new array are obtained
-from the previously allocated local array \texttt{U} via the notation \texttt{HALO\_SRC=U} (using
-\texttt{HALO\_SRC} will also initially copy \texttt{U} to the subimage); and finally 4. The
-allocation itself is \emph{executed} on the subimage device with the notation \texttt{[[device]]}.
+%%A team of images is a set of images that can readily execute independently of other images (from N2007)
 
-Fortran uses square bracket notation, e.g. \texttt{[image]}, to specify on what process the
-memory reference is physically located.  Square brackets are a visual clue to the
-programmer that the memory reference may be remote and therefore potentially suffer a
-performance penalty.  CAFe extends this by employing double-bracket notation to indicate
-possibly \emph{remote subimage execution}.
-
-Execution of the \texttt{Laplacian} task is done using the \texttt{do}
-\texttt{concurrent} construct:
-\begin{verbatim}
-   do while (.not. converged)
-      do concurrent (i=1:M, j=1:N)   [[device]]
-         call Laplacian( U(i,j)[device] )
-      end do
-      call HALO_TRANSFER(U, BC=CYCLIC)
-   end do
-\end{verbatim}
-There are several points that require highlighting: 1. Iteration occurs over the interior
-of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
-specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
-in any order, even \emph{concurrently}; 4. The local element of the array (as
-defined above in reference to the definition of the concurrent procedure
-\texttt{Laplacian}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
-task is to be taken from the subimage as noted by \texttt{[device]}; 6. All threads must finish
-execution of the loop body before further execution of the program proceeds; and 7. Transfer of
-all requisite halo regions is effected by the call to the new CAFe intrinsic function
-\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
-complete the halo transfer before program execution continues.
-
-Note that a transfer of halo memory is necessary after each completion of the do concurrent loop.
-This must be done in order for the halo region of a coarray on a given process to be consistent
-with the corresponding interior of the coarray on a logically neighboring process.
-Finally, memory for the entire array \texttt{U} can be copied from the
-subimage device with the statement,
-\texttt{U = U[device]}, 
-and memory deallocation (not shown) is similar to memory allocation.
-
-%%\begin{verbatim}
-%%   deallocate(U)
-%%   if (device /= this_image()) then
-%%      deallocate(U[device])   [[device]]
-%%   end if
-%%\end{verbatim}
+%%An allocated allocatable coarray is established in the team in which it was allocated
 
@@
 
-\input{coarray-comparison}
-
 %

### 2016-01-09 papers/CAFe/HIPS-paper/coarray-extensions.tex
Second pass through CAFe syntax extensions.

@@
 
-We must first introduce the important new concept of a \emph{CAFe subimage}.  Fortran
-images are a collection of distributed memory processes that all run the same program
-(image).  CAFe extends the concept of a Fortran image by allowing images to be
-hierarchical.  By this we mean that each image \emph{may} have a subimage (or subimages),
-but this subimage is not visible to other regular Fortran images.  Subimages also execute
-differently than normal images and may execute on different non-homogeneous hardware,
-e.g., an attached accelerator device.  Subimages are task based while images all execute a
-Single Program but with different (Multiple) Data (SPMD).  A task can be given to a
-subimage, but execution on the subimage terminates once the task is finished.  Memory on a
-subimage is permanent, however, and must be explicitly allocated and deallocated.
+We first introduce the important new concept of a \emph{CAFe subimage}.  Fortran images
+are a collection of distributed memory processes that all run the same program.  CAFe
+extends the concept of a Fortran image by allowing images to be hierarchical; by this we
+mean that an image \emph{may} host a subimage (or several subimages).  It is important to
+note that a subimage is not visible to Fortran images other than its hosting image.
+Subimages also execute differently than normal images and may even execute on different
+non-homogeneous hardware, e.g., an attached accelerator device.  Subimages are task based
+while images all execute a Single Program but with different (Multiple) Data (SPMD).  A
+task can be given to a subimage, but execution on the subimage terminates once the task is
+finished.  Memory on a subimage is permanent, however, and must be explicitly allocated
+and deallocated.
 
-One obtains a subimage by executing a new CAFe function,
+A programmer requests a subimage by executing the new CAFe function,
 \begin{verbatim}
@@
 \end{verbatim}
-where the integer argument represents an attached hardware device (or a separate process
-or a team of threads).  If the function fails (e.g., the requested device is unavailable)
-it returns the local image number of the process that is executing the
-current program, obtained by a call to the Fortran 2008 function \texttt{this\_image()}.
-Returning the current image (similar to MPI rank) allows program execution to proceed
-correctly even if there are no attached devices.
+where the integer argument represents an attached hardware device (or perhaps a separate
+process or a team of threads; it is the compilers responsibility to determine precisely
+what a subimage means in terms of the software and hardware environment under its
+control).  If the function fails (e.g., the requested device is unavailable) it returns
+the local image number of the process that is executing the current program, obtained by a
+call to the Fortran 2008 function \texttt{this\_image()}.  Returning the current image if
+the \texttt{GET\_SUBIMAGE} fails, allows program execution to proceed correctly even if
+there are no attached devices.
 
-Having obtained an subimage, we now consider how a subimage can be used within a program
-to allocate memory, transfer memory, and execute tasks.
+Please note that throughout this paper, text appearing in all caps (such as
+\texttt{GET\_SUBIMAGE}) indicates new syntax or functionality extended by CAFe.
+
+Once a variable representing a subimage has been obtained, it can be used within a program
+to allocate memory, transfer memory, and execute tasks.  This functionality is described below.
 
@@
 \end{verbatim}
-declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two
-and a corank of 1.  In Fortran, square brackets \texttt{[ ]} are used to denote operations
+declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two and a
+corank of 1.  In Fortran, square brackets \texttt{[ ]} are used to indicate operations
 (possibly expensive) on distributed memory where parentheses \texttt{( )} are used to
-denote array elements; square brackets denote array location.
+select specific array elements; square brackets, on the other hand, denote the
+\emph{location} of the array or array elements selected.
 
-When declared thusly, coarrays must be allocated on all program images.  CAFe extends this
-declaration to include memory that \emph{may} be allocated on a subimage, but does not
-need to be allocated unless it is specifically used by a task executing on a subimage.
+When declared thusly, coarrays must subsequently be allocated on all program images.  With
+CAFe, this declaration also allows memory to be allocated on a subimage, however memory
+allocation on a subimage is not required unless it is specifically used by some task
+executing that will be executed on the subimage.
 
@@
 As discussed, CAFe allows coarray memory to be conditionally allocated on a subimage.
-It should be allocated conditionally because the requested subimage in a texttt{GET\_SUBIMAGE}
+It should be allocated conditionally because the requested subimage in a \texttt{GET\_SUBIMAGE}
 call may not be available.  In addition, not all coarrays need be allocated on each
 subimage.  For example, the following code segment allocates memory for coarrays
-\texttt{U} and \texttt{V} on all images, but only coarrays \texttt{U} is allocated on
+\texttt{U} and \texttt{V} on all images, but only coarray \texttt{U} is allocated on
 the subimage specified by \texttt{device}:
@@
 \end{verbatim}
-where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement is
-mandatory and allocates the declared coarrays on all program images.  The use of the
-\texttt{*} symbol is used for the last corank dimension in CAF (in this case there is only
-one) to allow for the number of program images to be a runtime parameter, which can be
-ascertained with a \texttt{num\_images()} function call.  If \texttt{device} is
+where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement
+allocates coarrays \texttt{U} and \texttt{V} on all program images.  The \texttt{*} symbol
+in the allocation is required for the last corank dimension in CAF (in this case there is
+only one) to allow for the number of program images to be a runtime parameter, which can
+be ascertained with a \texttt{num\_images()} function call.  If \texttt{device} is
 available, its numeric value will be different from \texttt{this\_image()} and thus
@@
 \end{verbatim}
-assigns 3.14 to all of the array elements on \texttt{device}.  Specific array elements
-can also be transferred, 
+assigns 3.14 to all of the array elements of \texttt{U} located on \texttt{device}.
+Specific array elements can also be transferred, 
 \begin{verbatim}
@@
 
-CAFe restricts memory transfer between a subimage and its hosting image only. Transferring
-memory to another images subimage is not allowed.  In addition, memory transfer can only be
-initiated by the hosting image; code executing on a subimage cannot use coarray notation to
-transfer memory.
+CAFe restricts memory transfer between a subimage and its hosting image only; transferring
+memory to a subimage hosted by another image is not allowed.  In addition, memory transfer
+can only be initiated by the hosting image.  Thus code executing on a subimage cannot use
+coarray notation to select memory on another subimage nor on its host.
 
@@
 
-The final CAFe concept introduced is that of remote execution.  As discussed above, in
-Fortran all CAF images (similar to MPI ranks) execute the same program.  The only way for
+The final CAFe concept introduced is that of remote execution.  As discussed above,
+all CAF images (similar to MPI ranks) execute the same program.  The only way for
 an image to execute something different is to conditionally execute a block of code based
@@
 
-CAFe allows images to execute tasks on hosted subimages using standard procedure calls.
-For example,
+However, CAFe allows images to execute tasks on hosted subimages using standard procedure
+calls.  For example,
 \begin{verbatim}
@@
 on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
-whereby the double square brackets notation indicates \emph{where} the function will execute.
+whereby the double square bracket notation indicates \emph{where} the function will execute.
 Funtions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
@@
 
-CAFe subimage tasks can also be defined by the body of a Fortran do concurrent construct
-and executed on a subimage.  For example, the code segment,
+In addition, CAFe subimage tasks may also be defined by the body of a Fortran do
+concurrent construct and executed on a subimage.  For example, the code segment,
 \begin{verbatim}
@@
 \end{verbatim}
-will execute all of the code within the \texttt{do concurrent} construct on device.  In
-Fortran, the execution semantics of a \texttt{do concurrent} loop is that the loop
-iterates can be executed concurrently in any order.
+will execute all of the code within the \texttt{do concurrent} construct on \texttt{device}.
 
-Note the similarities between this example and the OpenMP pragma \texttt{!\$omp parallel do,
-private(w)}.  The use of the Fortran block construct to declare private
-variables allows a natural and clear way to distinguish between variables that are private
-to a ``thread'' (although in the example above, the ``private'' distinction for the
-variable \texttt{w} is unnecessary because it is a constant for all loop iterations).
+In Fortran, the execution semantics of a \texttt{do concurrent} loop is that the loop
+iterates can be executed concurrently in any order.  Note the similarities between this
+example and the OpenMP pragma \texttt{!\$omp parallel do, private(w)}.  The use of the
+Fortran block construct to declare private variables allows a natural and clear way to
+distinguish between variables that are private to a ``thread'' (although in the example
+above, the ``private'' distinction for the variable \texttt{w} is unnecessary because it
+is a constant for all loop iterations).
+
 (NOTE FOR CRAIG: need to read the standard to see if a block construct can be used in this
 way. ALSO, there is a lot more work here to think about how do concurrent merges with
-OpenMP and OpenACC with regards to synchronization, barriers, teams, atomics,
-, critical sections, ...)
+OpenMP and OpenACC with regards to synchronization, barriers, teams, atomics, , critical
+sections, ...)
 

### 2016-01-09 papers/CAFe/HIPS-paper/coarray-extensions.tex
Finished first cut by adding to subsection on integration with standard CAF.

@@
 will execute all of the code within the \texttt{do concurrent} construct on \texttt{device}.
+Execution of a \texttt{do concurrent} loop on a subimage acts \emph{as if} the loop were
+extracted as a outlined function and the subsequent task executed on a subimage.  Thus
+there is an implicit barrier at the end of the loop that blocks the hosting image
+from continuing execution until \emph{all} of the iterates of the loop have finished.
 
-In Fortran, the execution semantics of a \texttt{do concurrent} loop is that the loop
+In Fortran, the execution semantics of a \texttt{do concurrent} loop is that loop
 iterates can be executed concurrently in any order.  Note the similarities between this
@@
 above, the ``private'' distinction for the variable \texttt{w} is unnecessary because it
-is a constant for all loop iterations).
+is a constant over all loop iterations).
 
+The comparison with OpenMP also suggests that subimage execution may not be restricted to
+a single hardware execution unit, but may indeed be executed by a team of threads in a
+single (shared) memory address space.  This association with a team of threads
+will be exploited below when considering a team of images.
+
+\begin{comment}
   (NOTE FOR CRAIG: need to read the standard to see if a block construct can be used in this
@@
 sections, ...)
+\end{comment}
 
-\subsection{Integration with Coarray Standard Features}
+\subsection{Integration with Standard Coarray Features}
 
-Want to address: Barriers and teams primarily
+The new CAFe features introduced above work in concert with the existing 2008 Fortran
+coarray constructs and with the new parallel features to appear in the next Fortran
+standard\cite{TS:18508} (referred to here as Fortran 2015).  This section describes
+in more detail how the proposed extensions of CAFe interact with the parallel
+language features of Fortran 2015.
 
-%\input{coarray-comparison}
+\subsubsection{Fortran 2008 Standard}
 
-%%A team of images is a set of images that can readily execute independently of other images (from N2007)
+Perhaps the most important of the 2008 parallel constructs are those involving the
+control of image execution, including \texttt{sync all} and \texttt{sync images}
+statements.  A \texttt{sync all} statement implements a barrier synchronizing all CAF
+images while a \texttt{sync images} statement synchronizes a specified set of images.
+Neither of these synchronization primitives affect tasks potentially executing concurrently
+on a subimage.  Once subimage tasks begin execution they run to completion from the perspective
+of the hosting image.
 
-%%An allocated allocatable coarray is established in the team in which it was allocated
+Other features of 2008 include the \texttt{critical} construct and the \texttt{lock} and
+\texttt{unlock} statements.  The \texttt{critical} construct limits the execution of a
+block of code to one image at a time, while lock variables provide a facility for atomic
+operations.  Both of these features are defined in terms of normal CAF images so subimages
+do not directly participate unless the parent (or host) of a subimage does.
 
-%%\subsection{Execution Semantics and Memory Management}
+\subsubsection{Fortran 2015 Standard}
 
-%%This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.
-It describes what is the compiler's responsibility and what is the programmer's responsibility.
+The most important concept introduced by the 2015 standard is that of a team of images.
+However, subimages may not be included within a team of standard CAF images.  However, if one
+considers the comparison of the execution of a \texttt{do concurrent} block on a subimage
+with a team of OpenMP threads (as described above), the subimage could be considered as
+a team of one member that \emph{may} be executing its tasks on a team of shared memory threads.
+This allows allows the use of a \texttt{sync team} statement to synchronize with
+possibly executing threads.
 
-%% Completion of a do concurrent construct indicates that all executing threads (if a threading model is used by the compiler) have completed and that all memory on the executing subimage is in a consistent state.
+Thus we introduce another CAFe function, \texttt{GET\_SUBIMAGE\_TEAM()}, that returns a
+variable representing a fully-formed (established) team consisting of only one team
+member, the current subimage.  If this function is called outside the context of a task
+executing on a subimage it will return only a team consisting of the current image,
+\texttt{this\_image()}.  Synchronization across possible subimage threads is then possible
+using the statement
+\begin{verbatim}
+  sync team (GET_SUBIMAGE_TEAM())
+\end{verbatim}
+This allows implementation of the technique of double buffering within a subimage task
+so that all of the \emph{potential} subimage threads will have completed operating on the
+temporary buffer before it is copied back to the primary array variable.
 
-%%The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.
-The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.
-In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.
+Fortran 2015 also introduces the concept of events.  An event is a Fortran type that
+can be posted to by one image and waited for by another image.  CAFe relaxes the
+constraint that events must be coarrays so that events can be used locally within
+a single image to control concurrent execution of multiple subimage tasks.  For example,
+the following code segment shows how two tasks can be run concurrently on two separate
+subimages:
+\begin{verbatim}
+ type(event_type) :: evt
+ call update_ocn [[dev1, WITH_EVENT=evt]]
+ call update_atm [[dev2, WITH_EVENT=evt]]
+ event wait (evt, until_count=2)
+\end{verbatim}
 
-%% A note from a conversation with Matt regarding memory management
-%%
+In this example, the syntax \texttt{[[dev1, WITH\_EVENT=evt]]} indicates that the called
+subroutine is to be run on \texttt{dev1} and that the event \texttt{evt} will receive
+notification (an implicit posting) once the task has completed.  The \texttt{event wait}
+statement will cause execution on the hosting image to wait until \emph{both} tasks have
+completed.  One may even run multiple tasks on the \emph{same} device using events,
+however unless the subimage hardware allows concurrent execution of multiple tasks, only
+one task will be executed at a time.
 
-%The Fortran language has much tighter restrictions on aliasing than does C.
-%So unless a variable has the pointer or target attribute, it cannot be aliased.
-%Thus the compiler is able to aggressively optimized for memory movement between
-%the CPU and accelerator.  However, because the design philosophy of coarrays
-%is that memory transfer between images can be expensive, the programmer must
-%explicitly transfer memory between images with explicit syntax with square
-%bracket notation, i.e., $a[1] = a[2]$.  So we allow the compiler to manage
-%memory within an image but require the use of {\tt halo\_exchange} for the
-%transfer of halo memory between images.
+Fortran 2015 also provides new intrinsic procedures for atomic and collective operations.
+At this time we do not consider atomic operations for usage with the CAFe extensions.
+Atomic variables and associated synchronization constructs place too great of requirements
+on the executing software and hardware environment.  For example, the Open Community
+Runtime environment does not provide for atomic variables as the restriction
+helps OCR to better scale on a wider range of parallel computers.
 
-%Fortran currently supports:
-
-%1.
-Array syntax: e.g., C = A + B, where A, B, C are arrays.
-Note that this is implicitly a loop structure, but that no loop indices need be provided.
-Also the programmer need not specify where in memory these arrays reside.
-Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardware threads).
-This is a simple example and it is a research question as to what compiler directives would be useful for memory placement and other directions to the compiler for efficient code generation.
-
-%2.
-Pure procedures:  Fortran has syntax for specifying procedures that have no side effects during execution.
-Specifying code that is side-effect free code is important information to provide to the compiler so that it can generate efficient multi-threaded code.
-
-%3.
-Pure elemental procedures: Fortran has syntax for specifying procedures that take only scalar arguments, but may be applied across array elements.
-Elemental procedures are ideal for writing code to be executed within a hardware thread.
-They resemble OpenCL kernels, but are simpler because they leave all indexing up to the compiler.
-
-%We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.
-This additional syntax is provided by functions that return a copy of a small region of memory surrounding an array element (as seen within an elemental procedure) and with functions for thread synchronization.
-This additional syntax will allow pure procedures to perform stencil and other convolution-like operations on a copy of memory, synchronize, then store the computed results back to the array element associated with the given thread.
-
-%In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.
-This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.
-Fortran also has excellent facilities for interoperability with C so that programming in a mixed language environment is easily accomplished, including interoperability with native Fortran arrays.
-
-%
-% This could be part of conclusions
-%
-
-%%Note that the use of halo cells is the normal way that large and complex MPI and CAF programs are implemented.
-LOPe proposes to formalize this common pattern into the Fortran language allowing the compiler access to this information in order to spread computation over more hardware resources, improve performance, and to reduce complexity for the programmer.
+Collective operations, however, could potentially be very useful and fit nicely within
+CAFe concepts.  For example, a team consisting of hosted subimages could be formed and
+then data broadcasted to team members and reduction operations such as \texttt{co\_add}
+performed.  However since all team members of the current team must participate in
+collective operations \emph{and} since CAFe restricts subimages from being part of the
+same team with regular images, it makes it impossible for an image to participate with a
+team of its hosted subimages in collective operations.  So for now, collectives are
+excluded from CAFe.

### 2016-01-10 papers/CAFe/HIPS-paper/transformations.tex
Removed LOPe from section heading.
Completed cut at first cut including removing of references to LOPe.

@@
-\section{LOPe and CAFe Implementation}
+\section{CAFe Implementation}
 
-This section briefly describes how LOPe extensions to Fortran have been implemented as
-source-to-source transformations via rewrite rules (for expressing basic transformations) and
-rewriting strategies (for controlling the application of the rewrite rules).  A LOPe file is
-transformed to Fortran and OpenCL files through generative programming techniques using
-Stratego/XT tools\cite{Bravenboer200852} and is accomplished in three phases: (1) parsing to
-produce a LOPe Abstract Syntax Tree (AST) represented in the Annotated Term
-Format (ATerm\cite{DBLP:journals/spe/BrandJKO00}); (2) transformations of LOPe AST nodes to Fortran
-and C AST nodes; and (3) pretty-printing to the base Fortran and OpenCL languages.
+This section briefly describes how CAFe extensions to Fortran have been implemented as
+source-to-source transformations via rewrite rules (for expressing basic transformations)
+and rewriting strategies (for controlling the application of the rewrite rules).  A CAFe
+file is transformed to Fortran and OpenCL files through generative programming techniques
+using Stratego/XT tools\cite{Bravenboer200852} and is accomplished in three phases: (1)
+parsing to produce a CAFe Abstract Syntax Tree (AST) represented in the Annotated Term
+Format (ATerm\cite{DBLP:journals/spe/BrandJKO00}); (2) transformations of CAFe AST nodes
+to Fortran and C AST nodes; and finally (3) pretty-printing to the base Fortran and OpenCL
+languages.
 
-The foundation of LOPe is the syntax definition of the base language expressed in
-SDF (Syntax Definition Formalism) as part of the Open Fortran Project (OFP)\cite{OFP:git:url}.
-LOPe is defined in a separate SDF module that extends the Fortran 2008 language standard with 15
-context-free syntax rules.  Parsing is implemented in Stratego/XT by a scannerless generalized-LR
-parser and the conversion of transformed AST nodes to text is accomplished with simple
-pretty-printing rules.
+The foundation of CAFe is the syntax definition of the base language expressed in SDF
+(Syntax Definition Formalism) as part of the Open Fortran Project (OFP)\cite{OFP:git:url}.
+CAFe is defined in a separate SDF module that extends the Fortran 2008 language standard
+with only 11 context-free syntax rules.  Parsing is implemented in Stratego/XT by a
+scannerless generalized-LR parser and the conversion of transformed AST nodes to text is
+accomplished with simple pretty-printing rules.
 
@@
 
-A key component of code generation for LOPe is the targeting of a LOPe \texttt{CONCURRENT}
-procedure for a particular hardware architecture.  The execution target can be one
-of several choices, including serial execution by the current process via inlining
-of the function, parallel execution by inlining with OpenMP compiler directives, or parallel
-execution by heterogeneous processing elements with a language like OpenCL.
+A key component of code generation for CAFe is the targeting of a Fortran pure procedure
+or a \texttt{do concurrent} construct for execution on a particular hardware architecture.
+The execution target can be one of several choices, including serial execution by the
+current program image, parallel execution by inlining with OpenMP compiler directives, or
+parallel execution by heterogeneous processing elements with a language like OpenCL.
 
-For this work we have developed rewrite rules and strategies in Stratego/XT to rewrite Fortran AST
+We have developed rewrite rules and strategies in Stratego/XT to rewrite Fortran AST
 nodes to C AST nodes (extended with necessary OpenCL keywords).  The C AST ATerms have mostly a
-one-to-one correspondence with Fortran terms: a \texttt{CONCURRENT} procedure is transformed to an
+one-to-one correspondence with Fortran terms: a pure Fortran procedure is transformed to an
 OpenCL kernel; Fortran formal parameters are transformed to C parameters (with a direct mapping of
@@
 Fortran executable statements are rewritten as C statements.  The only
-minor complication is mapping the LOPe local, array index view to the global C index space.  This
+minor complication is mapping the CAFe local, array index view to the global C index space.  This
 translation is facilitated by a Fortran symbol table that stores array shape and halo size
-information.
+information.  However, these transformation are not entirely completed and some of the
+OpenCL kernels used in the example section have been coded by hand.
 
@@
 
-Transformations of a LOPe procedure call site are more difficult, though technically straight
-forward.  The Fortran function call must be transformed to a call to run the OpenCL kernel
-(generated as described above).  This is facilitated by use of the ForOpenCL library which
-provides Fortran bindings to the OpenCL runtime\cite{Sottile:2013:FTE:2441516.2441520}.  However,
-this usage requires the declaration of extra variables, allocation of memory on the OpenCL device
-(subimage), transfer of memory, marshalling of kernel arguments, and synchronization.
+Transformations of a CAFe procedure call site are more difficult, though technically
+straight forward.  The Fortran function call must be transformed to a call to run the
+OpenCL kernel (generated as described above).  This is facilitated by use of the ForOpenCL
+library which provides Fortran bindings to the OpenCL
+runtime\cite{Sottile:2013:FTE:2441516.2441520}.  However, this usage requires the
+declaration of extra variables, allocation of memory on the OpenCL device (subimage),
+transfer of memory, marshalling of kernel arguments, and synchronization.
 
-These transformations are accomplished using several rewrite stages using Stratego/XT strategies:
-(1) a symbol table is produced in the first pass to store information related to arrays including,
-array shape, halo size, and allocation status;
-(2) additional variables are declared that are used to maintain the OpenCL runtime state, including
-the OpenCL device, the OpenCL kernel, and OpenCL variables used to marshall kernel arguments; and
-(3) all CAFe code related to subimage usage is desugared (lowered) to standard Fortran with calls to the
-ForOpenCL library as needed.
-%%The latter step includes the allocation of OpenCL device memory, transfer of memory to and from the OpenCL device, marshalling of kernel arguments, running of the OpenCL kernel, and synchronization.
+These transformations are accomplished using several rewrite stages using Stratego/XT
+strategies:
+(1) a symbol table is produced in the first pass to store information related to arrays
+including, array shape and allocation status;
+(2) additional variables are declared that are used to maintain the OpenCL runtime state,
+including the OpenCL device, the OpenCL kernel, and OpenCL variables used to marshall
+kernel arguments; and
+(3) all CAFe code related to subimage usage is desugared (lowered) to standard Fortran
+with calls to the ForOpenCL library as needed.
+The latter step includes the allocation of OpenCL device memory, transfer of memory to and
+from the OpenCL device, marshalling of kernel arguments, running of the OpenCL kernel, and
+synchronization.
 
-Though not yet available, similar rewrite strategies are planned for targeting programming models
-other than OpenCL including parallel execution with OpenMP directives.  In addition, simple serial
-execution with function inlining (if desired) will be performed by the regular Fortran compiler once
-all CAFe and LOPe code has been desugared to standard Fortran.
-
-%%\subsection{Halo Transfer Performance}
-
-%%A thorough examination of potential performance gains using LOPe for parallelization of code is beyond the scope of this paper.
-The emphasis of this work is to explain the syntax and execution semantics of a LOPe application.
-However, preliminary work using the Stratego/XT rewrite system as described indicates that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).
-
-%%The current implementation does \emph{not} take advantage of optimization strategies such as prefetching of array tiles (including halos) into OpenCL local memory or of loop unrolling on the OpenCL device.
-Neither does it take advantage of the potential of LOPe \texttt{CONCURRENT} function fusion along with execution and synchronization strategies to overlap communication with computation.
-
-%%Since many scientific codes are dominated by memory performance, including and especially stencil algorithms as they typically only involve a computation on a small locally central array element and a small overlapping halo region.
-Stencil operations frequently do not contain enough floating point operations per memory load to allow for floating point performance to operate at peak (though this is entirely application and domain specific).
-Thus we illustrate the \emph{potential} for performance by noting the latency and throughput performance of an attached GPU in conjunction with MPI distributed memory performance associated with halo transfer in Table 1.
-
-%%The results in Table 1 indicate that the primary bottleneck in using accelerators attached to the TODO bus using OpenCL is likely to be the latency in transferring memory to and from the device for distributed memory clusters of only a few nodes exchanging halo data using MPI.
+Though not yet available, similar rewrite strategies are planned for targeting programming
+models other than OpenCL including parallel execution with OpenMP and OpenACC directives.
+However, as CAFe is designed, simple serial execution --- with respect to subimage tasks
+--- is automatically provided by the regular CAF compiler.

### 2016-01-10 papers/CAFe/HIPS-paper/performance.tex
Completed cut at this section, primarily removing references to LOPe.

@@
 
-This section shows and compares performance measurements for: 1.
-Memory bandwidth usage on GPU; 2.
-Computaional time on GPU; 3.
-Halo Communication time (total plus breakdown to and from device and MPI); 4.
-Total time scaling on ACISS; 5.
-Comparison with OpenACC (hopefully).
+It is important to note that the emphasis of this work is to explore and explain the new
+syntax and execution semantics of a CAFe application.  A thorough examination of potential
+performance gains (if any) using CAFe for parallelization of code is beyond the scope of
+this paper.  The primary purpose of CAFe is to combine the parallel features of Fortran coarrays
+--- executing a \emph{single} program --- with concurrent execution of separate tasks executing on
+potentially heterogeneous hardware.  However, the relative performance of computation on
+a cluster of GPUs compared with the necessary communication of halo information is
+of interest; especially considering complete exchance of halo data between multiple
+coarray images \emph{and} between each individual host image and its subimage (an attached
+OpenCL device).
+
+%%However, preliminary work using the Stratego/XT rewrite system as described indicates that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).
+
+This section shows and compares performance measurements for: 1. Memory bandwidth usage on
+GPU; 2. Computaional time on GPU; 3. Halo Communication time (total plus breakdown to and
+from device and MPI); 4. Total time scaling on ACISS; 5. Comparison with OpenACC
+(hopefully).
+
+\subsection{Halo Transfer Performance}
+
+The current implementation does \emph{not} take advantage of optimization strategies such
+as prefetching of array tiles (including halos) into OpenCL local memory or of loop
+unrolling on the OpenCL device.
+
+%%either does it take advantage of the potential of CAFe to overlap communication with computation (for example, computing on
+
+\begin{comment}
+Since many scientific codes are dominated by memory performance, including and especially
+stencil algorithms as they typically only involve a computation on a small locally central
+array element and a small overlapping halo region.  Stencil operations frequently do not
+contain enough floating point operations per memory load to allow for floating point
+performance to operate at peak (though this is entirely application and domain specific).
+Thus we illustrate the \emph{potential} for performance by noting the latency and
+throughput performance of an attached GPU in conjunction with MPI distributed memory
+performance associated with halo transfer in Table 1.
+\end{comment}
+
+\begin{comment}
+The results in Table 1 indicate that the primary bottleneck in using accelerators attached
+to the TODO bus using OpenCL is likely to be the latency in transferring memory to and
+from the device for distributed memory clusters of only a few nodes exchanging halo data
+using MPI.
+\end{comment}

### 2016-01-11 papers/CAFe/HIPS-paper/coarray-extensions.tex
Reduced font size of verbatim text.

@@
 A programmer requests a subimage by executing the new CAFe function,
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 where the integer argument represents an attached hardware device (or perhaps a separate
@@
 codimension attribute, for example,
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two and a
@@
 the subimage specified by \texttt{device}:
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement
@@
 Deallocation of subimage memory is similar to allocation,
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 
@@
 hosting image to the device.  This is explicitly done with normal CAF syntax.  For example,
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 assigns 3.14 to all of the array elements of \texttt{U} located on \texttt{device}.
 Specific array elements can also be transferred, 
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 where the first and last columns of \texttt{U} are copied from the device to the hosting image.
@@
 calls.  For example,
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
@@
 concurrent construct and executed on a subimage.  For example, the code segment,
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 will execute all of the code within the \texttt{do concurrent} construct on \texttt{device}.
@@
 using the statement
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 This allows implementation of the technique of double buffering within a subimage task
@@
 subimages:
+\small
 \begin{verbatim}

### 2016-01-11 papers/CAFe/HIPS-paper/intro.tex
Reduced font size of verbatim text.

@@
 with new array syntax using square brackets, for example,
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 copies memory from the first element of \texttt{U} residing on image number 3 into

### 2016-01-11 papers/CAFe/HIPS-paper/multigrid.tex
Small change suggesting description of 1D solution rather than 2D.
Added description of relaxation loop.

@@
 
-This section describes the Laplacian multigrid example.
+In this section a one-dimensional solution to Laplace's equation is described using CAFe.
+The solution uses multigrid techniques to speedup convergence.
 
-We start with the declaration of an array \texttt{U} with two local dimensions (rank) and
-two distributed memory codimensions (corank).
+Array allocation, initialization, using Fortran and CAFe contructs as shown above.
+
+\small
 \begin{verbatim}
-   real, allocatable, :: U(:,:)[:,:]
+ type(event_type) :: evt
 \end{verbatim}
-The corank of the array is chosen to be identical to the rank of the array so that the logical
-process topology aligns in a way that allows a natural halo exchange between logically neighboring
-processes (this could not occur if corank and rank are not the same).  For example, if the process
-location is \texttt{[pcol,prow]}, then the right-hand halo for the local array \texttt{U} can be
-obtained by the assignment
-\texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
-\texttt{U} are given by the allocation statement,
-\texttt{allocate(U(0:M+1,0:N+1)[MP,*])}.
-This allocation statement specifies (given the one element halo size provided earlier for
-\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)},
-the bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the interior region
-\texttt{U(1:M,1:N)}.
+\normalsize
 
-In this allocation statement, the total number of process columns $NP$ can be obtained at runtime,
-but \emph{may not} be explicitly provided (according to Coarray Fortran (CAF) rules) because the actual number of
-participating processes (in Fortran called images) is variable, depending on how many processes are
-requested at program startup.  In this discussion, it is \emph{assumed} that there are no holes in
-the logical processor topology, thus $MP*NP = P$, where $MP$ is the number of process rows and $P$
-is the total number of participating processes (images).
+\small
+\begin{verbatim}
+ do i = 1, nsteps                            1
+   call relax (N, V1h[dev], Buf[dev])        &
+                    [[dev, EVENT=evt]]       2
+   call relax_boundary (N, V1h)              3
+
+   wait event (evt, until_count=i)           4
+
+   V1h(0)[dev] = V1h(0)                      5
+   V1h(N)[dev] = V1h(N)                      6
+
+   V1h(  1) = V1h(  1) [dev]                 7
+   V1h(N-1) = V1h(N-1) [dev]                 8
+
+   sync all                                  9
+
+   V1h( -1) = V1h(N-1) [left]               10
+   V1h(N+1) = V1h(  1) [right]              11
+ end do                                     12
+
+ call restrict(N,V1h[dev],V2h[dev]) [[dev]] 13
+\end{verbatim}
+\normalsize
 
 Once a subimage is obtained, memory on the device can be allocated,
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 There are four points to note regarding this memory allocation: 1. Memory is only allocated if a
@@
 
-Execution of the \texttt{Laplacian} task is done using the \texttt{do}
-\texttt{concurrent} construct:
-\begin{verbatim}
-   do while (.not. converged)
-      do concurrent (i=1:M, j=1:N)   [[device]]
-         call Laplacian( U(i,j)[device] )
-      end do
-      call HALO_TRANSFER(U, BC=CYCLIC)
-   end do
-\end{verbatim}
-There are several points that require highlighting: 1. Iteration occurs over the interior
-of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
-specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
-in any order, even \emph{concurrently}; 4. The local element of the array (as
-defined above in reference to the definition of the concurrent procedure
-\texttt{Laplacian}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
-task is to be taken from the subimage as noted by \texttt{[device]}; 6. All threads must finish
-execution of the loop body before further execution of the program proceeds; and 7. Transfer of
-all requisite halo regions is effected by the call to the new CAFe intrinsic function
-\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
-complete the halo transfer before program execution continues.
-
 Note that a transfer of halo memory is necessary after each completion of the do concurrent loop.
@@
 and memory deallocation (not shown) is similar to memory allocation.
-
-%%\begin{verbatim}
-%%   deallocate(U)
-%%   if (device /= this_image()) then
-%%      deallocate(U[device])   [[device]]
-%%   end if
-%%\end{verbatim}

### 2016-01-11 papers/CAFe/HIPS-paper/intro.tex
Added a couple of paragraphs on OCR.

@@
 
-% OpenACC directives and features:
-%\begin{verbatim}
-%● acc_copy(in|out)( ptr, bytes )
-%● acc_create( ptr, bytes )
-%● acc_delete( ptr, bytes )
-%● acc_is_present( ptr, bytes )
-%● acc_update_(device|local)( ptr, bytes )
-%● acc_deviceptr( ptr )
-%● acc_hostptr( devptr )
-%● acc_[un]map_data( devptr, hostptr, bytes )
-%● acc_memcpy_(to|from)_device
-%\end{verbatim}
+Another option is to abandon the stricture of message passing within a procedural
+programming language altogether.  The Open Community Runtime (OCR) project provides an
+asynchronous task-based runtime designed for machines with high-core count, such as can be
+expected in any exascale hardware design\cite{OCR:SC:2012}.  The project's goal is to
+create a framework and reference implementation to help developers explore programming
+methods to improve the power efficiency, programmability, and reliability of HPC
+applications while maintaining application performance.
+
+OCR helps the application developer with the complex process of writing multi-core
+applications.  It does this by providing facilities for creating and executing a set of
+tasks.  The OCR runtime schedules the execution of these tasks based on events which
+embody dataflow and code flow dependencies; when all of the dependencies for a task are
+satisfied, the task can be run.
+
+However, it cannot be emphasized too strongly that like using the OpenCL programming
+environment, actually creating real scientific applications based on OCR is a high hurdle
+for the application developer.  A goal of this paper is to examine ways in
+which high-level syntax based on Fortran coarrays can be utilized by compiler-based tools
+to \emph{target} programming environments like OpenCL or OCR --- thus freeing the developer from
+the task of frequently porting the application to new complex environments.  The hope is
+that high-level syntax can hide the complexities of the underlying software and hardware.
 
@@
 with new array syntax using square brackets, for example,
-\small
+%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%\normalsize
 copies memory from the first element of \texttt{U} residing on image number 3 into
@@
 
+\begin{comment}
 Historically it has been hard to create parallel programs.  The
@@
 provide data access patterns that are regular and spatially local.
+
+\end{comment}
+
+% OpenACC directives and features:
+%\begin{verbatim}
+%● acc_copy(in|out)( ptr, bytes )
+%● acc_create( ptr, bytes )
+%● acc_delete( ptr, bytes )
+%● acc_is_present( ptr, bytes )
+%● acc_update_(device|local)( ptr, bytes )
+%● acc_deviceptr( ptr )
+%● acc_hostptr( devptr )
+%● acc_[un]map_data( devptr, hostptr, bytes )
+%● acc_memcpy_(to|from)_device
+%\end{verbatim}

### 2016-01-11 papers/CAFe/code/OpenCL/plot-1h-error.py
Initial version

@@ -1,0 +1,49 @@
+#!/usr/bin/env python
+"""
+Example: simple line plot.
+Show how to make and save a simple line plot with labels, title and grid
+"""
+import numpy
+import pylab
+import re
+import string
+
+x = []
+v = []
+f = open('output_1h_0.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+x = []
+v = []
+f = open('output_1h_mid.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+x = []
+v = []
+f = open('output_1h_end.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+pylab.xlabel('x')
+pylab.xlim([0, 1])
+pylab.ylabel('v')
+pylab.ylim([-1, 1])
+pylab.title('1h Error')
+pylab.grid(True)
+pylab.savefig('poisson.png')
+
+pylab.show()

### 2016-01-11 papers/CAFe/code/OpenCL/plot-2h-error.py

@@ -1,0 +1,48 @@
+#!/usr/bin/env python
+"""
+Example: simple line plot.
+Show how to make and save a simple line plot with labels, title and grid
+"""
+import numpy
+import pylab
+import re
+import string
+
+x = []
+v = []
+f = open('output_2h_0.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+x = []
+v = []
+f = open('output_2h_mid.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+x = []
+v = []
+f = open('output_2h_end.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+pylab.xlabel('x')
+pylab.ylabel('v')
+pylab.ylim([-1,1])
+pylab.title('2h Error')
+pylab.grid(True)
+pylab.savefig('poisson.png')
+
+pylab.show()

### 2016-01-11 papers/CAFe/code/OpenCL/plot-8h-error.py

@@ -1,0 +1,48 @@
+#!/usr/bin/env python
+"""
+Example: simple line plot.
+Show how to make and save a simple line plot with labels, title and grid
+"""
+import numpy
+import pylab
+import re
+import string
+
+x = []
+v = []
+f = open('output_8h_0.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+x = []
+v = []
+f = open('output_8h_mid.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+x = []
+v = []
+f = open('output_8h_end.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[1]))
+   v.append(float(words[2]))
+f.close()
+pylab.plot(x, v)
+
+pylab.xlabel('x')
+pylab.ylabel('v')
+pylab.ylim([-1,1])
+pylab.title('8h Error')
+pylab.grid(True)
+pylab.savefig('poisson.png')
+
+pylab.show()

### 2016-01-11 papers/CAFe/code/OpenCL/plot-error-time.py

@@ -1,0 +1,27 @@
+#!/usr/bin/env python
+"""
+Example: simple line plot.
+Show how to make and save a simple line plot with labels, title and grid
+"""
+import numpy
+import pylab
+import re
+import string
+
+x = []
+e = []
+f = open('error_time.dat', 'r')
+for line in f:
+   words = re.split(' *', string.strip(line))
+   x.append(float(words[0]))
+   e.append(float(words[1]))
+f.close()
+pylab.plot(x, e)
+
+pylab.xlabel('time')
+pylab.ylabel('max error')
+pylab.title('Error')
+pylab.grid(True)
+pylab.savefig('poisson.png')
+
+pylab.show()

### 2016-01-13 papers/CAFe/HIPS-paper/intro.tex
minor tweaks

@@
 language.  Originally called Coarray Fortran (CAF) by its
-authors\cite{Numrich:1998:CFP:289918.289920}, parallism in Fortran is similar to MPI in
+authors\cite{Numrich:1998:CFP:289918.289920}, parallelism in Fortran is similar to MPI in
 that all cooperating CAF processes (called \emph{images}) execute a Single Program but with
@@
 
-In this paper we propose CAFe: Co-Array Fortran Extensions.  These extensions to CAF introduce
+In this paper we propose CAFe: \emph{Co-Array Fortran Extensions}.  These extensions to CAF introduce
 relatively minor additions to the existing coarray portion of the Fortran language.  They

### 2016-01-14 papers/CAFe/HIPS-paper/intro.tex
Added better citations to the OCR section.

--- 2016-01-11 papers/CAFe/HIPS-paper/intro.tex
@@
 Another option is to abandon the stricture of message passing within a procedural
-programming language altogether.  The Open Community Runtime (OCR) project provides an
-asynchronous task-based runtime designed for machines with high-core count, such as can be
-expected in any exascale hardware design\cite{OCR:SC:2012}.  The project's goal is to
-create a framework and reference implementation to help developers explore programming
-methods to improve the power efficiency, programmability, and reliability of HPC
-applications while maintaining application performance.
+programming language altogether.  The Open Community Runtime (OCR)
+project\cite{OCR:wiki:url} provides an asynchronous task-based runtime designed for
+machines with high-core count\cite{Dokulil20151453}, such as can be expected in any
+exascale hardware design.  The project's goal is to create a framework and reference
+implementation to help developers explore programming methods to improve the power
+efficiency, programmability, and reliability of HPC applications while maintaining
+application performance.
 

### 2016-01-14 papers/CAFe/HIPS-paper/coarray-extensions.tex

@@
 previous work extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
+Please note that throughout this paper, text appearing in all caps (such as
+\texttt{GET\_SUBIMAGE}) indicates new syntax or functionality extended by CAFe.
 
@@
 A programmer requests a subimage by executing the new CAFe function,
+
 \small
@@
 \normalsize
-where the integer argument represents an attached hardware device (or perhaps a separate
+
+\noindent where the integer argument represents an attached hardware device (or perhaps a separate
 process or a team of threads; it is the compilers responsibility to determine precisely
@@
 call to the Fortran 2008 function \texttt{this\_image()}.  Returning the current image if
-the \texttt{GET\_SUBIMAGE} fails, allows program execution to proceed correctly even if
+the \texttt{GET\_SUBIMAGE} call fails, allows program execution to proceed correctly even if
 there are no attached devices.
-
-Please note that throughout this paper, text appearing in all caps (such as
-\texttt{GET\_SUBIMAGE}) indicates new syntax or functionality extended by CAFe.
 

### 2016-01-14 papers/CAFe/HIPS-paper/multigrid.tex
Added more better :-) description of the relaxation loop on the finest grid level.

@@
 
-In this section a one-dimensional solution to Laplace's equation is described using CAFe.
-The solution uses multigrid techniques to speedup convergence.
+In this section a one-dimensional solution to Laplace's equation is described using CAFe
+syntax.  The solution uses multigrid techniques to improve the rate of convergence
+of iterative methods.  We decribe the implementation in terms of a one-dimensional
+problem for simplicity.
 
-Array allocation, initialization, using Fortran and CAFe contructs as shown above.
+\subsection{Multigrid Algorithm}
 
+The multigrid algorithm uses a series of successively coarser grids to interate
+approximage solution obtained from the preceeding finer grid.  Higher frequency error modes are
+damped out as the solution processes up the grid hierarchy.  At the top of the
+hierarchy, an exact solution is obtained and then this solution is iterated and
+propogated down the grid hierarchy form coarser to finer grids.  After a few sweeps
+up and down the hierarchy the solution with have been obtained on the desired
+original grid.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.5in]{Fig1}
+\caption{Two multigrid levels (N,N/2) on one image showing the two cells that are shared with
+  neighbors on the left and the right and computed redundantly.  The fine level grid (N)
+  also shows the two cells that must be exchanged between neighbors.  Arrows represent the direction
+  of data exchange from the perspective of the local image.}
+\label{fig_grids}
+\end{figure}
+
+Two levels of the grid hierarchy are shown in \ref{fig_grids} running on a processing computing on
+an interior portion of the grid; additional images are computing on regions to the left and right of
+the region shown in the figure.  The finist grid (size N) is shown
+in the bottom of the figure with cell indices running from (-1:N+1).  Cells 0 and N are shared
+between image neighbors to the left and right (indicated by the light blue coloring and the verticle
+dashed lines.  These shared cells are computed redundantly on each program image.
+
+The red cells in the figure are computed concurrently on all subimages (one per image).
+This distribution of computation over the hierarchical domain of images and subimages
+requires communication of boundary regions between the various computational resources.
+The communication is shown by arrows in the grid level N of the figure.  Data in cells 1
+and 7 must be transferred from the subimage (upwardly pointing arrows) and the shared cell
+computations must be copied to the subimage (downwardly pointing arrows) The horizontal
+arrows representing data computed and subsequently transferred from the left and right
+neighbors.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.5in]{Fig2}
+\caption{Restriction of fine grid to coarse grid. Arrows show the points used in
+         mapping onto the coarse grid.}
+\label{fig_restric}
+\end{figure}
+
+The second grid level N/2 is also shown in Figure \ref{fig_grids}.  Data on this grid is obtained
+by simple interpolation from information on grid level N as shown in Figure \ref{fig_restric}.
+We call this interpolation a restriction and refer the reader to previous work on the multigrid
+technique for more information\cite{multigrid_methods}.  The approximate solution on the N/2
+grid is relaxed a few times and then passed on the grid level N/4 for further iteration.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.5in]{Fig3}
+\caption{Prolongation of coarse grid to fine grid. Arrows show the points used to
+         interpolate onto the fine grid.}
+\label{fig_prolongate}
+\end{figure}
+
+At the top of the gird hierachy there are sufficiently few cells that an exact solution can be
+obtained (perhaps by a non-interpretive method).  This solution is then propogated down the
+grid hierarchy as shown in Figure\ref{fig_prolongate}.
+
+\subsection{Implementation of the Relaxation Step}
+
+The multigrid algorithm has been implemented in Fortran using the CAFe syntax described in
+the previous section.  Much of the implementation is straighforward in terms of CAFe are
+is only described briefly.
+
+Data coarrays at each grid level are declared, allocated, and initialized on each program image.
+These coarrays are then copied to the subimages by the parent, hosting images.  Below we show
+the implementation of the relaxation loop on grid level N.  Each iteration step incorporates
+concurrent computation on an image and its hosted subimage.  In order to execute concurrently,
+we must declare a event variable to allow concurrent execution and then synchronization once
+\emph{both} computation ond the subimage \emph{and} the image have completed.
+
+The event variable is declared as,
 \small
@@
 \normalsize
-
+After initialization, this event variable has its count variable set to 0 and is incremented
+each time the subimage task completes during an interation of the code segment shown below:
 \small
@@
    V1h(N+1) = V1h(  1) [right]              11
- end do                                     12
 
- call restrict(N,V1h[dev],V2h[dev]) [[dev]] 13
+   sync all                                 12
+ end do                                     13
 \end{verbatim}
@@
 
-Once a subimage is obtained, memory on the device can be allocated,
-\small
-\begin{verbatim}
-   if (device /= this_image()) then
-      allocate(U[device], HALO_SRC=U)   [[device]]
-   end if
-\end{verbatim}
-\normalsize
+This example shows the first relaxation steps on the finest grid level N.  The initial
+guess (provided by the coarray \texttt{V1h}) is relaxed \texttt{nsteps} times as
+specified by the loop beginning at statement 1.  The relaxation is computed on
+the subimage \texttt{dev} by the call to the \texttt{relax} procedure at statement 2.
+This call uses coarray memory located on the subimage as specified by the arguments
+\texttt{V1h[dev]} and \texttt{Buf[dev]}, where the second coarray is used for temporary
+storage.
+
+Since an event \texttt{evt} has been supplied to the call in statement 2, the program
+continues execution without waiting for the remote task on \texttt{dev} to complete.  The
+count associated with the event will be increased by an implicit post to the event once
+the task has finished executing.  Thus the task started by statement 2 will execute
+concurrently with the \texttt{relax\_boundary} call at statement 3 --- the first call
+executes on the subimage \texttt{dev} while the second call executes on the local image
+\texttt{this\_image()}.
+
+Most of the computation is accomplised by the subimage executing the \texttt{relax}
+procedure (not shown) operating over the interior corresponding to fine grid indices (1:N-1).
+Concurrently, \texttt{relax\_boundary} operates on the boundary indices 0 and N
+(note that memory for the boundary computation in located on local image as
+the coarray \texttt{V1h} is not coindexed using square bracket notation in statement 3.
+
+Once relaxation of the boundary has completed, the program image will wait for the event
+indicating that the subimage task has finished executing relaxation on the interior (statement
+4).  The \texttt{until\_count} for the event is increased by 1 each iteration of the loop,
+so the program waits until the event counter equals the iterate counter \texttt{i}.
+
+Once the wait has completed, communication of the halo regions between the subimage and
+the hosting image can begin.  Shared boundary cells at 0 and N are copied \emph{to} the
+subimage by the execution of statements 5 and 6 and interior regions at cells 1 and N-1 are
+copied \emph{from} the subimage by statements 7 and 8.  Boundary cells are exchanged (copied)
+from neighboring images by statements 10 and 11.  To ensure that no race conditions are
+introduced by the exchange of information between images, explicit synchronization is
+accomplished by the execution of statements 9 and 12.
+
+\begin{comment}
 There are four points to note regarding this memory allocation: 1. Memory is only allocated if a
@@
 and memory deallocation (not shown) is similar to memory allocation.
+\end{comment}

### 2016-01-17 papers/CAFe/HIPS-paper/conclusions.tex
Removed extra TeX vspace commands.
Rewriting with may improvements and deletions of LOPe.

--- 2015-02-09 papers/LOPe/conclusions.tex
@@
 
-Fortran is a general-purpose programming language and as such it provides limited facilities for
-expressing concepts useful for stencil operations. For example, halo regions must be expressed in
-terms of the existing syntax of the language and there is no way to specify that the ``interior''
-of an array is in any way special from an ``outside'' region.  By not providing support for
-stencils in the language, the programmer must make specific choices regarding data-access patterns
-and the order of operations on the data.  These choices often hide the opportunity for
-optimizations by the compiler \cite{Dubey:2014:SSC:2686745.2686756}.  For example, if the compiler
-had knowledge of the semantic intent of halo regions, it could reorder operations so that border
-regions were computed \emph{before} interior regions, allowing the transfer of data in halo regions
-to overlap with computations on the interior.
+Fortran is a general-purpose programming language and is extremely important to the
+scientific and high performance computing (HPC) community.  While new code development is
+often in languages other than Fortran, such as C++, Fortran usage in terms of CPU hours on
+HPC platforms remains as high as ??? \cite{Yellick::language:breakdown}.  Fortran will be
+with us well into the future as the lifetime of important scientific applications is
+measured in decades and Fortran usage dominates in scientific communitities that are
+critical to understanding the future of life on our planet such as the climate community.
 
-Just as Coarray Fortran originally extended Fortran to include domain specific knowledge (parallel
-computation) replacing MPI library routines\cite{Numrich:1998:CFP:289918.289920}, LOPe seeks to extend Fortran by
-providing domain specific knowledge of stencil operations.
-Specifically, LOPe and CAFe together provide:
-(1) a local view of stencil operations on data that allows a complete separation of the implementation of a stencil
-algorithm with data-access patterns;
-(2) memory placement via allocation routines that allow the specification of allocation location;
-(3) task execution placement with double-bracket syntax specifying which subimage is to execute a particular operation; and
-(4) memory exchange via the \texttt{TRANSFER\_HALO} intrinsic procedure.
+However significant challenges arise as hardware adapts to the end of Moore's
+Law\cite{exascale:workshop:2011}.  The revolution occurring in the hardware architecture
+has completely altered the landscape for scientific computing as a code designed for
+distributed memory bulk synchronous model of parallelism may actually run slower on
+advanced hardware\cite{Dubey:2014:SSC:2686745.2686756}.
+
+Parallelism was added to Fortran with the addition of coarrays in the 2008
+standard\cite{fortran:2008} to enable the migration of codes as hardware evolves.
+Significant language features were left out of this early version of coarrays as
+identified by \emph{Mellor-Crummy et al}\cite{mellor-crummey:2009:caf2}.  Many of these features
+have been adopted in what is likely to be called the 2015 standard of the language\cite{fortran:2015},
+such as teams of program images and improved synchronization constructs like events.
+However, the Fortran coarray model fails to address heterogeneity in hardware as is already
+being seen in HPC computing platforms with the adoption of distributed nodes consisting of
+general purpose CPUs with attached accelerator devices like GPUs, for example, in the early
+RoadRunner Machine at Los Alamos\cite{RoadRunner} and new machines arriving such as Blue
+Waters\cite{blue:waters}.
+
+Current
+
+In this preliminary work we have examined ....
+In no way does this represent a thorough examination of these potential features.
+
+CAFe introduces the concept of subimages executing on a hosting image.  CAFe includes
+\item
+  Dynamic creation of subimages.
+\item
+  Dynamic memory allocation and placement on a subimage.
+\item
+  Explicit addressing of memory on subimages using standard coarray notation.  It follows
+  the Partitioned Global Address Space model (PGAS), although with subimages the address
+  space is hierahical in the sense that the memory partioning only allows memory exchange
+  between a subimage and its hosting image as initiated by the hosting image.
+\item
+  Task creation and execution on subimages using extended coarray syntax with double
+  square brackets \testtt{[[ ]]}.  Task mays be defined in terms of standard \emph{pure}
+  Fortran procedures and task execution continues without interuption until completion.
+  In addition a Fortran 2015 event may be notified upon completion of the task to allow
+  concurrent execution on the subimage and on the hosting image.
+\item
+  CAFe integrates Fortran 2015 features to allow relatively simple programs to be written
+  that employ all of the heterogenous components expected on exascale platforms.
+
+\subsection{Limitations}
+
+Possible criticism of this work includes the following:
+\item
+  The source-to-source transformations that have been developed to implement CAFe
+  are rudimentory and have not been optimized.  For example, memory transfers have
+  not been aggregated into larger messages to improve performance.  Memory transfer
+  is also blocking so that prefetching of data cannot be employed via techniques
+  envolving the reordering of statements (code motion) that effectively employed in
+  CAF compilers.
+\item
+  The Laplacian example is in reality a ``toy'' problem.  Real multiphysics codes
+  employ hundreds of more variables and would have entirely different performance
+  characteristics because of the increased pressure on cache and register usage.
+\item
+  Existing scientific codes often separate communication into separate modules from
+  the computational kernels.  This makes adoption of CAF itself slow (because these
+  MPI modules already work fine) and thus additional extensions like CAFe also
+  would be slow to adopt. However CAFe constructs can be used in an incremental
+  fashion making it easier to adopt
+\item
+  In reality CAF (and therefore CAFe) are relatively low-level parallel programming
+  languages that require the programmer to \emph{explicitly} control the placement
+  of memory, communication, and synchronization.  Perhaps higher level languages like
+  Chapel would be better to adopt as they could provide an easier programming model
+  to the programmer and possibly more avenues for compiler optimization.
+\item
+  Thorough examination of performance capabilities on real scientific applications
+  (or at least scientific ``mini-apps'') has not been attempted.
+\item
+  Programming models like that proposed by CAFe already exist in the form of a combination
+  of MPI and OpenMP (or OpenACC) programs.  The advantage of CAFe --- by being a purely
+  language construct --- is that it opens up the possibility of compiler optimizations
+  that would otherwise be impossible with a library approach intermixed with compiler directives.
+
+\subsection{Future Work}
+
+However, especially with the integration of task-based parallelism with synchronization
+events arriving in Fortran 2015, the possibilities are intriguing.  Tasks plus the
+dependencies represented by events suggest a correspondence with the Open Community
+Runtime (OCR) libraries.  In OCR, the entire program must be broken into a set of tasks
+represented by a directed acycic graph (DAG).  In OCR, tasks are available for executing
+when all of their input dependencies are satisfied, such occurs when computation of a new
+interation of data elements have been updated and other upstream tasks have completed.
+
+Tools to automatically convert existing codes to run under the OCR model faces challenges.  For
+example, the hundreds of MPI functions would have to be semantically understood by the tools,
+integrated with existing OpenACC directives (for example), a DAG generaged for the complete
+program, and then transformed into an OCR program.
+
+An intriguing \emph{possibility} exists for automatic conversion of CAFe codes to OCR exists:
+\item
+  CAF codes have well defined boundaries (called segment boundaries) indicating where new
+  OCR tasks must be created.
+\item
+  CAF has many fewer library routines (primarily collectives) that must be converted.
+\item
+  CAF has events which already define some of the dependencies necessary for the creation
+  of OCR tasks.
+\item
+  CAFe introduces the notion of task based parallelism (in addition to the existing
+  Fortran features like \texttt{do concurrent}.
+
+In the future we hope to examine \emph{if} the possibility exists for CAFe codes to be
+automatically converted to programs employing the Open Community Runtime.
+
+\begin{comment}
+These choices often hide the opportunity for
+optimizations by the compiler \cite{Dubey:2014:SSC:2686745.2686756}.
+\end{comment}
 
@@
 
-\vspace{.1in}
-
 By implementing LOPe we have demonstrated that LOPe can be used to easily and succinctly code

### 2016-01-17 papers/CAFe/HIPS-paper/intro.tex
Added comment sections from Heroux.

@@
 \section{Introduction}
+
+\begin{comment}
+Fortran is essential for exascale programming.
+The explicit policy of non-overlapping arrays (enabling compilers to more easily vectorize and parallelize automatically), simple loop syntax, longstanding support for robust real and complex arithmetic, along with the vast quantity of encoded scientific knowledge mean Fortran will remain the primary development language for many domain scientists who write software.
+Even as clean-slate Fortran development efforts decline, the value of the legacy Fortran software base and its ongoing refactoring and extension will be essential for decades to come.
+
+Execution environments must undergo more substantial changes than programming languages and environments.
+Current runtime systems are very good at concurrent execution, but are not designed for lightweight threading (except GPUs) or locality-aware task mapping.
+Thread-scalable computing will require better, more transparent on-node thread parallel environments.
+Fortunately, we see much progress lately, underneath existing programming languages and environments.
+
+\emph{Adding Tasking is Critical}
+
+Although programming and execution environments are evolving smoothly toward exascale capabilities, we do have disruptive changes ahead.
+Most scalable parallel applications today have simple data and work decompositions: Each MPI rank owns a static portion of large data objects, e.g., a subdomain of a large distributed global domain, and each rank executes its code sequentially (potentially vectorizing), or with modest thread-parallel capabilities.
+This approach works on existing NUMA multicore systems by assigning multiple MPI ranks to a node and using OpenMP across a handful of cores, but performance using this approach is not sustainable as core counts continue to increase.
+
+Tasking, with work granularities sufficiently large to make effective use of one or a few cores, must be added to most applications in order to sustain performance improvement as concurrency demands increase.
+Specifically, tasking requires one or more levels of additional decomposition (at least logically) of data objects, e.g., create multiple patches or tiles from each MPI subdomain, and assign tasks to execute concurrently on these patches.
+Within a single shared memory node, tasks can in principle cooperate closely, executing dataflow patterns, sharing data and otherwise collaborating in lightweight
+
+\cite{HPCWire:Heroux}
+\end{comment}
 
@@
 with new array syntax using square brackets, for example,
-%\small
 \begin{verbatim}
@@
 \end{verbatim}
-%\normalsize
 copies memory from the first element of \texttt{U} residing on image number 3 into

### 2016-01-18 papers/CAFe/HIPS-paper/intro.tex
Added text on synchonization

@@
 Since each CAF process executes the same program, there are no good avenues for exploiting
-task-based parallelism.  Neither is there a mechanism for dynamic coarray memory
+task-based parallelism.  Neither is there a adequate mechanism for dynamic coarray memory
 allocation, nor a mechanism for the remote execution of functions.  In CAF, like MPI, the
@@
 include the concept of subimages in order to nest coarray images, semantics for distributed
-array declarations, memory placement, remote memory transfer, and remote task execution.  CAFe
+array declarations, memory placement, remote memory transfer, and remote task execution
+and synchronization.  CAFe
 builds upon the existing standardized coarray features of Fortran.  We will introduce the
-CAFe extensions that are proposed to define their semantics, demonstrate their use in simple
+proposed CAFe extensions, define their semantics, demonstrate their use in simple
 examples, and provide experimental results demonstrating their performance in a prototype implementation.

### 2016-01-18 papers/CAFe/HIPS-paper/conclusions.tex
Fixed itemize stuff
edits to conclusion

@@
 such as teams of program images and improved synchronization constructs like events.
+
+%%
+%% MJS - this stuff is so familiar these days, no need to cite specific machine examples
+%%
+%, for example, in the early
+%RoadRunner Machine at Los Alamos\cite{RoadRunner} and new machines arriving such as Blue
+%Waters\cite{blue:waters}.
+
 However, the Fortran coarray model fails to address heterogeneity in hardware as is already
 being seen in HPC computing platforms with the adoption of distributed nodes consisting of
-general purpose CPUs with attached accelerator devices like GPUs, for example, in the early
-RoadRunner Machine at Los Alamos\cite{RoadRunner} and new machines arriving such as Blue
-Waters\cite{blue:waters}.
+general purpose CPUs with attached accelerator devices like GPUs.  To address this shortcoming
+of the coarray model, CAFe introduces the concept of subimages executing on a hosting image.  
+CAFe includes:
 
-Current
-
-In this preliminary work we have examined ....
-In no way does this represent a thorough examination of these potential features.
-
-CAFe introduces the concept of subimages executing on a hosting image.  CAFe includes
+\begin{itemize}
 \item
@@
   Task creation and execution on subimages using extended coarray syntax with double
-  square brackets \testtt{[[ ]]}.  Task mays be defined in terms of standard \emph{pure}
+  square brackets \texttt{[[ ]]}.  Task mays be defined in terms of standard \emph{pure}
   Fortran procedures and task execution continues without interuption until completion.
-  In addition a Fortran 2015 event may be notified upon completion of the task to allow
+  In addition, a Fortran 2015 event may be notified upon completion of the task to allow
   concurrent execution on the subimage and on the hosting image.
@@
   that employ all of the heterogenous components expected on exascale platforms.
+\end{itemize}
 
@@
 
-Possible criticism of this work includes the following:
+In this paper we have defined CAFe and described its semantics and relation to the 2008 and
+2015 Fortran standards.  Possible criticism of this work includes the following:
+
+\begin{itemize}
 \item
@@
   characteristics because of the increased pressure on cache and register usage.
-\item
-  Existing scientific codes often separate communication into separate modules from
-  the computational kernels.  This makes adoption of CAF itself slow (because these
-  MPI modules already work fine) and thus additional extensions like CAFe also
-  would be slow to adopt. However CAFe constructs can be used in an incremental
-  fashion making it easier to adopt
-\item
-  In reality CAF (and therefore CAFe) are relatively low-level parallel programming
-  languages that require the programmer to \emph{explicitly} control the placement
-  of memory, communication, and synchronization.  Perhaps higher level languages like
-  Chapel would be better to adopt as they could provide an easier programming model
-  to the programmer and possibly more avenues for compiler optimization.
+%\item
+%  Existing scientific codes often separate communication into separate modules from
+%  the computational kernels.  This makes adoption of CAF itself slow (because these
+%  MPI modules already work fine) and thus additional extensions like CAFe also
+%  would be slow to adopt. However CAFe constructs can be used in an incremental
+%  fashion making it easier to adopt
+%\item
+%  In reality CAF (and therefore CAFe) are relatively low-level parallel programming
+%  languages that require the programmer to \emph{explicitly} control the placement
+%  of memory, communication, and synchronization.  
 \item
   Thorough examination of performance capabilities on real scientific applications
-  (or at least scientific ``mini-apps'') has not been attempted.
+  (or at least scientific ``mini-apps'') has not been attempted.  Such a study of both
+  CAFe and CAF itself is necessary to demonstrate the coarray model as a viable alternative
+  for application codes with complexity above that of an idealized mini-app.
 \item
@@
   that would otherwise be impossible with a library approach intermixed with compiler directives.
+\end{itemize}
 
@@
 An intriguing \emph{possibility} exists for automatic conversion of CAFe codes to OCR exists:
+\begin{itemize}
 \item
@@
   Fortran features like \texttt{do concurrent}.
+\end{itemize}
 

### 2016-01-18 papers/CAFe/HIPS-paper/multigrid.tex
typo fixing

@@
 
-The multigrid algorithm uses a series of successively coarser grids to interate
-approximage solution obtained from the preceeding finer grid.  Higher frequency error modes are
-damped out as the solution processes up the grid hierarchy.  At the top of the
-hierarchy, an exact solution is obtained and then this solution is iterated and
-propogated down the grid hierarchy form coarser to finer grids.  After a few sweeps
-up and down the hierarchy the solution with have been obtained on the desired
-original grid.
+The multigrid algorithm uses a series of successively coarser grids to
+iteratively approximate the solution obtained from the preceeding
+finer grid.  Higher frequency error modes are damped out as the
+solution processes up the grid hierarchy.  At the top of the
+hierarchy, an exact solution is obtained and then this solution is
+iterated and propogated down the grid hierarchy form coarser to finer
+grids.  After a few sweeps up and down the hierarchy the solution with
+have been obtained on the desired original grid.
 
@@
 
-Two levels of the grid hierarchy are shown in \ref{fig_grids} running on a processing computing on
+Two levels of the grid hierarchy are shown in \ref{fig_grids} running on a processor computing on
 an interior portion of the grid; additional images are computing on regions to the left and right of
@@
 in the bottom of the figure with cell indices running from (-1:N+1).  Cells 0 and N are shared
-between image neighbors to the left and right (indicated by the light blue coloring and the verticle
-dashed lines.  These shared cells are computed redundantly on each program image.
+between image neighbors to the left and right (indicated by the light blue coloring and the vertical
+dashed lines).  These shared cells are computed redundantly on each program image.
 
@@
 and 7 must be transferred from the subimage (upwardly pointing arrows) and the shared cell
-computations must be copied to the subimage (downwardly pointing arrows) The horizontal
-arrows representing data computed and subsequently transferred from the left and right
+computations must be copied to the subimage (downwardly pointing arrows). The horizontal
+arrows represent data computed and subsequently transferred from the left and right
 neighbors.
@@
 
-At the top of the gird hierachy there are sufficiently few cells that an exact solution can be
-obtained (perhaps by a non-interpretive method).  This solution is then propogated down the
+At the top of the grid hierarchy there are sufficiently few cells that an exact solution can be
+obtained (perhaps by a non-iterative method).  This solution is then propagated down the
 grid hierarchy as shown in Figure\ref{fig_prolongate}.
@@
 
-The multigrid algorithm has been implemented in Fortran using the CAFe syntax described in
-the previous section.  Much of the implementation is straighforward in terms of CAFe are
-is only described briefly.
-
-Data coarrays at each grid level are declared, allocated, and initialized on each program image.
-These coarrays are then copied to the subimages by the parent, hosting images.  Below we show
-the implementation of the relaxation loop on grid level N.  Each iteration step incorporates
-concurrent computation on an image and its hosted subimage.  In order to execute concurrently,
-we must declare a event variable to allow concurrent execution and then synchronization once
-\emph{both} computation ond the subimage \emph{and} the image have completed.
+The multigrid algorithm has been implemented in Fortran using the CAFe
+syntax described in the previous section. Data coarrays at each grid
+level are declared, allocated, and initialized on each program image.
+These coarrays are then copied to the subimages by the parent, hosting
+images.  Below we show the implementation of the relaxation loop on
+grid level N.  Each iteration step incorporates concurrent computation
+on an image and its hosted subimage.  In order to execute
+concurrently, we must declare a event variable to allow concurrent
+execution and then synchronization once \emph{both} computation ond
+the subimage \emph{and} the image have completed.
 

### 2016-01-18 papers/CAFe/HIPS-paper/intro.tex
Removed over emphasis on OCR.

@@
 
-Another option is to abandon the stricture of message passing within a procedural
-programming language altogether.  The Open Community Runtime (OCR)
+Another potential option is to abandon the stricture of message passing within a procedural
+programming language altogether.  For example, the Open Community Runtime (OCR)
 project\cite{OCR:wiki:url} provides an asynchronous task-based runtime designed for
@@
 
+\begin{comment}
 OCR helps the application developer with the complex process of writing multi-core
@@
 satisfied, the task can be run.
+\end{comment}
 
-However, it cannot be emphasized too strongly that like using the OpenCL programming
-environment, actually creating real scientific applications based on OCR is a high hurdle
-for the application developer.  A goal of this paper is to examine ways in
-which high-level syntax based on Fortran coarrays can be utilized by compiler-based tools
-to \emph{target} programming environments like OpenCL or OCR --- thus freeing the developer from
-the task of frequently porting the application to new complex environments.  The hope is
-that high-level syntax can hide the complexities of the underlying software and hardware.
+It cannot be emphasized too strongly that using the OpenCL programming language and
+libraries (or programming for the OCR environment) presents a high hurdle for the
+application developer creating real scientific applications.  A goal of this paper is to
+examine ways in which high-level syntax based on Fortran coarrays can be utilized by
+compiler-based tools to \emph{target} programming environments like OpenCL, thus
+freeing the developer from the task of frequently porting the application to new complex
+environments.  The hope is that high-level syntax can hide the complexities of the
+underlying hardware \emph{and software.}
 
@@
 builds upon the existing standardized coarray features of Fortran.  We will introduce the
-proposed CAFe extensions, define their semantics, demonstrate their use in simple
-examples, and provide experimental results demonstrating their performance in a prototype implementation.
+proposed CAFe extensions, define their semantics, demonstrate their use in a simple
+example, and provide experimental results demonstrating the relative
+performance of computation and communication in a prototype implementation executing
+in a heterogeneous environment.
 CAFe is complementary to

### 2016-01-18 papers/CAFe/HIPS-paper/coarray-extensions.tex
Fixed font problems.

@@
 
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 
@@
 codimension attribute, for example,
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two and a
@@
 the subimage specified by \texttt{device}:
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement
@@
 Deallocation of subimage memory is similar to allocation,
-\small
+%%\small
 \begin{verbatim}
@@
 hosting image to the device.  This is explicitly done with normal CAF syntax.  For example,
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 assigns 3.14 to all of the array elements of \texttt{U} located on \texttt{device}.
 Specific array elements can also be transferred, 
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 where the first and last columns of \texttt{U} are copied from the device to the hosting image.
@@
 calls.  For example,
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
@@
 concurrent construct and executed on a subimage.  For example, the code segment,
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 will execute all of the code within the \texttt{do concurrent} construct on \texttt{device}.
@@
 using the statement
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 This allows implementation of the technique of double buffering within a subimage task
@@
 subimages:
-\small
+%%\small
 \begin{verbatim}

### 2016-01-18 papers/CAFe/HIPS-paper/conclusions.tex
Finished second pass through the conclusions.

@@
 scientific and high performance computing (HPC) community.  While new code development is
-often in languages other than Fortran, such as C++, Fortran usage in terms of CPU hours on
-HPC platforms remains as high as ??? \cite{Yellick::language:breakdown}.  Fortran will be
-with us well into the future as the lifetime of important scientific applications is
-measured in decades and Fortran usage dominates in scientific communitities that are
-critical to understanding the future of life on our planet such as the climate community.
+often in languages other than Fortran (such as C++), surveys show that Fortran usage is
+still high\cite{prabhu2011survey}.  Fortran will be with us well into the future as the
+lifetime of important scientific applications is measured in decades and Fortran usage
+dominates in scientific communitities that are critical to understanding the future of
+life on our planet such as the climate community.
 
 However significant challenges arise as hardware adapts to the end of Moore's
-Law\cite{exascale:workshop:2011}.  The revolution occurring in the hardware architecture
+Law\cite{exascale:workshop:2011}.  The revolution occurring in hardware architecture
 has completely altered the landscape for scientific computing as a code designed for
@@
   Explicit addressing of memory on subimages using standard coarray notation.  It follows
-  the Partitioned Global Address Space model (PGAS), although with subimages the address
+  the Partitioned Global Address Space (PGAS) model, although with subimages the address
   space is hierahical in the sense that the memory partioning only allows memory exchange
@@
   Fortran procedures and task execution continues without interuption until completion.
-  In addition, a Fortran 2015 event may be notified upon completion of the task to allow
+  In addition, a Fortran 2015 event type may be notified upon completion of the task to allow
   concurrent execution on the subimage and on the hosting image.
@@
 
-In this paper we have defined CAFe and described its semantics and relation to the 2008 and
+In this paper we have defined CAFe and described its semantics and relationship to the 2008 and
 2015 Fortran standards.  Possible criticism of this work includes the following:
@@
   are rudimentory and have not been optimized.  For example, memory transfers have
-  not been aggregated into larger messages to improve performance.  Memory transfer
+  not been aggregated into larger message blocks to improve performance.  Memory transfer
   is also blocking so that prefetching of data cannot be employed via techniques
-  envolving the reordering of statements (code motion) that effectively employed in
+  envolving the reordering of statements (code motion) that are effectively employed in
   CAF compilers.
@@
   The Laplacian example is in reality a ``toy'' problem.  Real multiphysics codes
-  employ hundreds of more variables and would have entirely different performance
+  employ hundreds (or more) of variables and would have entirely different performance
   characteristics because of the increased pressure on cache and register usage.
@@
 
-However, especially with the integration of task-based parallelism with synchronization
-events arriving in Fortran 2015, the possibilities are intriguing.  Tasks plus the
+In spite of the limitations discussed, 
+adoption of the task-based parallelism of CAFe and the synchronization
+event types in Fortran 2015 suggests intriguing possibilities.  Tasks plus the
 dependencies represented by events suggest a correspondence with the Open Community
@@
 represented by a directed acycic graph (DAG).  In OCR, tasks are available for executing
-when all of their input dependencies are satisfied, such occurs when computation of a new
-interation of data elements have been updated and other upstream tasks have completed.
+when all of their input dependencies are satisfied, such as occurs when computation of a new
+iteration of data elements have been updated and other upstream tasks have been completed.
 
-Tools to automatically convert existing codes to run under the OCR model faces challenges.  For
+Tools to automatically convert existing applications to run under the OCR model faces challenges.  For
 example, the hundreds of MPI functions would have to be semantically understood by the tools,
@@
 \item
-  CAF has events which already define some of the dependencies necessary for the creation
+  CAF has events that already define some of the dependencies necessary for the creation
   of OCR tasks.
@@
   CAFe introduces the notion of task based parallelism (in addition to the existing
-  Fortran features like \texttt{do concurrent}.
+  Fortran constructs like \texttt{do concurrent}.
 \end{itemize}
@@
 In the future we hope to examine \emph{if} the possibility exists for CAFe codes to be
-automatically converted to programs employing the Open Community Runtime.
+automatically converted to an entirely differently programming paradigm like
+the Open Community Runtime.
 
@@
 
-%
-% This could be part of conclusions
-%
-
-%%Note that the use of halo cells is the normal way that large and complex MPI and CAF programs are implemented.
-
-\subsubsection{Benefits.}
-LOPe proposes to formalize the common, halo software pattern in language syntax, thus providing the
-compiler with access to halo information in order to spread computation over more hardware
-resources, improve performance, and to reduce complexity for the programmer.  Furthermore, LOPe
-semantics provide important \emph{language restrictions} that remove the possibility of race conditions
-that occur when multiple threads have write access to overlapping data regions.
-
+\begin{comment}
 \subsubsection{Limitations.}
@@
 array regions in each rectangular patch.
-
-By implementing LOPe we have demonstrated that LOPe can be used to easily and succinctly code
-the stencil algorithms that are common to many areas of science, and furthermore, that LOPe is
-suitable for transformation to languages like OpenCL that support heterogeneous computing.  It
-remains to history to ascertain if LOPe is sufficiently general purpose to be included in a
-general-purpose programming language or if it is better suited to remain as a DSL and to be used as a
-special-purpose preprocessing tool.
+\end{comment}

### 2016-01-18 papers/CAFe/HIPS-paper/coarray-extensions.tex
spelling

@@
 whereby the double square bracket notation indicates \emph{where} the function will execute.
-Funtions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
+Functions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
 keyword \texttt{pure}) and scalar formal parameters must be passed by value (i.e., declared with the

### 2016-01-18 papers/CAFe/HIPS-paper/conclusions.tex
spelling

@@
 lifetime of important scientific applications is measured in decades and Fortran usage
-dominates in scientific communitities that are critical to understanding the future of
+dominates in scientific communities that are critical to understanding the future of
 life on our planet such as the climate community.
@@
   the Partitioned Global Address Space (PGAS) model, although with subimages the address
-  space is hierahical in the sense that the memory partioning only allows memory exchange
+  space is hierarchical in the sense that the memory partitioning only allows memory exchange
   between a subimage and its hosting image as initiated by the hosting image.
@@
   square brackets \texttt{[[ ]]}.  Task mays be defined in terms of standard \emph{pure}
-  Fortran procedures and task execution continues without interuption until completion.
+  Fortran procedures and task execution continues without interruption until completion.
   In addition, a Fortran 2015 event type may be notified upon completion of the task to allow
@@
   The source-to-source transformations that have been developed to implement CAFe
-  are rudimentory and have not been optimized.  For example, memory transfers have
+  are rudimentary and have not been optimized.  For example, memory transfers have
   not been aggregated into larger message blocks to improve performance.  Memory transfer
   is also blocking so that prefetching of data cannot be employed via techniques
-  envolving the reordering of statements (code motion) that are effectively employed in
+  involving the reordering of statements (code motion) that are effectively employed in
   CAF compilers.
@@
 Runtime (OCR) libraries.  In OCR, the entire program must be broken into a set of tasks
-represented by a directed acycic graph (DAG).  In OCR, tasks are available for executing
+represented by a directed acyclic graph (DAG).  In OCR, tasks are available for executing
 when all of their input dependencies are satisfied, such as occurs when computation of a new
@@
 example, the hundreds of MPI functions would have to be semantically understood by the tools,
-integrated with existing OpenACC directives (for example), a DAG generaged for the complete
+integrated with existing OpenACC directives (for example), a DAG generated for the complete
 program, and then transformed into an OCR program.

### 2016-01-18 papers/CAFe/HIPS-paper/multigrid.tex
spelling

@@
 syntax.  The solution uses multigrid techniques to improve the rate of convergence
-of iterative methods.  We decribe the implementation in terms of a one-dimensional
+of iterative methods.  We describe the implementation in terms of a one-dimensional
 problem for simplicity.
@@
 The multigrid algorithm uses a series of successively coarser grids to
-iteratively approximate the solution obtained from the preceeding
+iteratively approximate the solution obtained from the preceding
 finer grid.  Higher frequency error modes are damped out as the
@@
 hierarchy, an exact solution is obtained and then this solution is
-iterated and propogated down the grid hierarchy form coarser to finer
+iterated and propagated down the grid hierarchy form coarser to finer
 grids.  After a few sweeps up and down the hierarchy the solution with
@@
 an interior portion of the grid; additional images are computing on regions to the left and right of
-the region shown in the figure.  The finist grid (size N) is shown
+the region shown in the figure.  The finest grid (size N) is shown
 in the bottom of the figure with cell indices running from (-1:N+1).  Cells 0 and N are shared
@@
 obtained (perhaps by a non-iterative method).  This solution is then propagated down the
-grid hierarchy as shown in Figure\ref{fig_prolongate}.
+grid hierarchy as shown in Figure \ref{fig_prolongate}.
 
@@
 concurrently, we must declare a event variable to allow concurrent
-execution and then synchronization once \emph{both} computation ond
+execution and then synchronization once \emph{both} computation and
 the subimage \emph{and} the image have completed.
@@
 After initialization, this event variable has its count variable set to 0 and is incremented
-each time the subimage task completes during an interation of the code segment shown below:
+each time the subimage task completes during an iteration of the code segment shown below:
 \small
@@
 
-Most of the computation is accomplised by the subimage executing the \texttt{relax}
+Most of the computation is accomplished by the subimage executing the \texttt{relax}
 procedure (not shown) operating over the interior corresponding to fine grid indices (1:N-1).

### 2016-01-18 papers/CAFe/HIPS-paper/conclusions.tex
edits

@@
 
-However significant challenges arise as hardware adapts to the end of Moore's
+However, significant challenges arise as hardware adapts to the end of Moore's
 Law\cite{exascale:workshop:2011}.  The revolution occurring in hardware architecture
-has completely altered the landscape for scientific computing as a code designed for
+has completely altered the landscape for scientific computing as a code designed for the
 distributed memory bulk synchronous model of parallelism may actually run slower on
@@
 
-Tools to automatically convert existing applications to run under the OCR model faces challenges.  For
-example, the hundreds of MPI functions would have to be semantically understood by the tools,
-integrated with existing OpenACC directives (for example), a DAG generated for the complete
-program, and then transformed into an OCR program.
+%%%
+%%% MJS: I cut the following out - conversion of MPI -> OCR is not in scope for us, and 
+%%% we don't want to argue one way or the other about that topic.  Stay focused on CAF/CAFe.
+%%%
+%Tools to automatically convert existing applications to run under the OCR model faces challenges.  For
+%example, the hundreds of MPI functions would have to be semantically understood by the tools,
+%integrated with existing OpenACC directives (for example), a DAG generated for the complete
+%program, and then transformed into an OCR program.
 

### 2016-01-18 papers/CAFe/HIPS-paper/coarray-extensions.tex
Near final pass.

@@
 Topics highlighted in this section are: 1. Distributed memory array allocation;
-2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
+2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution
+and synchronization.  This
 description is within the context of extensions to Fortran; as shorthand, these extensions
@@
 declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two and a
-corank of 1.  In Fortran, square brackets \texttt{[ ]} are used to indicate operations
+corank of one.  In Fortran, square brackets \texttt{[ ]} are used to indicate operations
 (possibly expensive) on distributed memory where parentheses \texttt{( )} are used to
@@
 \texttt{U} and \texttt{V} on all images, but only coarray \texttt{U} is allocated on
-the subimage specified by \texttt{device}:
+the subimage specified by the variable \texttt{device}:
 %%\small
@@
    allocate(U(M,N)[*], V(M,N)[*])
-   if (device \= this_image()) then
+   if (device /= this_image()) then
      allocate(U(M,N)[*])  [[device]]
@@
 \begin{verbatim}
-   if (device \= this_image()) then
+   if (device /= this_image()) then
      deallocate(U)  [[device]]
@@
 %%\normalsize
-executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
+executes the subroutine relax, on the subimage \texttt{device}, using coarray \texttt{U} \emph{located}
 on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
@@
 a single hardware execution unit, but may indeed be executed by a team of threads in a
-single (shared) memory address space.  This association with a team of threads
-will be exploited below when considering a team of images.
+single (shared) memory address space.  This association with a team of threads will be
+exploited below when considering a team of subimages.
 
@@
 The most important concept introduced by the 2015 standard is that of a team of images.
-However, subimages may not be included within a team of standard CAF images.  However, if one
+Subimages may not be included within a team of standard CAF images.  However, if one
 considers the comparison of the execution of a \texttt{do concurrent} block on a subimage
@@
 a team of one member that \emph{may} be executing its tasks on a team of shared memory threads.
-This allows allows the use of a \texttt{sync team} statement to synchronize with
+This allows the use of a \texttt{sync team} statement to synchronize with
 possibly executing threads.
 
-Thus we introduce another CAFe function, \texttt{GET\_SUBIMAGE\_TEAM()}, that returns a
+Thus we introduce another CAFe intrinsic function, \texttt{GET\_SUBIMAGE\_TEAM()}, that returns a
 variable representing a fully-formed (established) team consisting of only one team
@@
 In this example, the syntax \texttt{[[dev1, WITH\_EVENT=evt]]} indicates that the called
-subroutine is to be run on \texttt{dev1} and that the event \texttt{evt} will receive
+subroutine is to be run on \texttt{dev1} and that the event variable \texttt{evt} will receive
 notification (an implicit posting) once the task has completed.  The \texttt{event wait}
@@
 CAFe concepts.  For example, a team consisting of hosted subimages could be formed and
-then data broadcasted to team members and reduction operations such as \texttt{co\_add}
+then data be broadcasted to team members and reduction operations such as \texttt{co\_add}
 performed.  However since all team members of the current team must participate in

### 2016-01-18 papers/CAFe/HIPS-paper/multigrid.tex
Near final pass.

@@
 iterated and propagated down the grid hierarchy form coarser to finer
-grids.  After a few sweeps up and down the hierarchy the solution with
+grids.  After a few sweeps up and down the hierarchy the solution will
 have been obtained on the desired original grid.
@@
 
-Two levels of the grid hierarchy are shown in \ref{fig_grids} running on a processor computing on
+Two levels of the grid hierarchy are shown in Fig. \ref{fig_grids} running on a processor computing on
 an interior portion of the grid; additional images are computing on regions to the left and right of
@@
 technique for more information\cite{multigrid_methods}.  The approximate solution on the N/2
-grid is relaxed a few times and then passed on the grid level N/4 for further iteration.
+grid is relaxed a few times and then passed on to grid level N/4 for further iteration.
 
@@
 on an image and its hosted subimage.  In order to execute
-concurrently, we must declare a event variable to allow concurrent
-execution and then synchronization once \emph{both} computation and
+concurrently, we must declare an event variable to allow concurrent
+execution and then synchronization once computation on \emph{both} 
 the subimage \emph{and} the image have completed.
@@
 The event variable is declared as,
-\small
+%%\small
 \begin{verbatim}
@@
 \end{verbatim}
-\normalsize
+%%\normalsize
 After initialization, this event variable has its count variable set to 0 and is incremented
@@
 Concurrently, \texttt{relax\_boundary} operates on the boundary indices 0 and N
-(note that memory for the boundary computation in located on local image as
+(note that memory for the boundary computation is located on local image as
 the coarray \texttt{V1h} is not coindexed using square bracket notation in statement 3.

### 2016-01-18 papers/CAFe/HIPS-paper/performance.tex
Near final pass.

@@
 a cluster of GPUs compared with the necessary communication of halo information is
-of interest; especially considering complete exchance of halo data between multiple
+of interest, especially considering that a complete exchance of halo data involves communication
+between multiple
 coarray images \emph{and} between each individual host image and its subimage (an attached
@@
 
-This section shows and compares performance measurements for: 1. Memory bandwidth usage on
-GPU; 2. Computaional time on GPU; 3. Halo Communication time (total plus breakdown to and
-from device and MPI); 4. Total time scaling on ACISS; 5. Comparison with OpenACC
-(hopefully).
+This section shows and compares performance measurements for: 1. Computational time on the GPU
+of the interior domain;
+2. Computational time on the CPU of the shared boundary regions; and 3. Halo communication time.
 
-\subsection{Halo Transfer Performance}
+%%\subsection{Halo Transfer Performance}
 
 The current implementation does \emph{not} take advantage of optimization strategies such
-as prefetching of array tiles (including halos) into OpenCL local memory or of loop
-unrolling on the OpenCL device.
+as prefetching of array tiles (including halos) into OpenCL local memory.
 

### 2016-01-18 papers/CAFe/HIPS-paper/coarray-extensions.tex
Repeated sentence, sounded better in introduction.
a vs. an

@@
 description is within the context of extensions to Fortran; as shorthand, these extensions
-are referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to
-previous work extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
+are referred to as CAFe, for Coarray Fortran extensions.
 Please note that throughout this paper, text appearing in all caps (such as
@@
 Execution of a \texttt{do concurrent} loop on a subimage acts \emph{as if} the loop were
-extracted as a outlined function and the subsequent task executed on a subimage.  Thus
+extracted as an outlined function and the subsequent task executed on a subimage.  Thus
 there is an implicit barrier at the end of the loop that blocks the hosting image

### 2016-01-18 papers/CAFe/HIPS-paper/conclusions.tex
final changes
Fixed a few typos.

@@
 
-An intriguing \emph{possibility} exists for automatic conversion of CAFe codes to OCR exists:
+An intriguing \emph{possibility} exists for automatic conversion of CAFe codes to OCR:
 \begin{itemize}
@@
   CAFe introduces the notion of task based parallelism (in addition to the existing
-  Fortran constructs like \texttt{do concurrent}.
+  Fortran constructs like \texttt{do concurrent}).
 \end{itemize}
@@
 In the future we hope to examine \emph{if} the possibility exists for CAFe codes to be
-automatically converted to an entirely differently programming paradigm like
-the Open Community Runtime.
+automatically converted to an entirely different programming paradigm like
+the Open Community Runtime.  Given the change in architectures being considered for
+future HPC platforms it is likely that similar, novel runtime models will be proposed,
+and we believe that CAFe provides a viable programming model for compiler-based mapping to
+these new systems.
 

### 2016-01-18 papers/CAFe/HIPS-paper/conclusions.tex
Fixed last remaining references.

@@
 However, significant challenges arise as hardware adapts to the end of Moore's
-Law\cite{exascale:workshop:2011}.  The revolution occurring in hardware architecture
+Law\cite{ashby2010opportunities}.  The revolution occurring in hardware architecture
 has completely altered the landscape for scientific computing as a code designed for the
@@
 Parallelism was added to Fortran with the addition of coarrays in the 2008
-standard\cite{fortran:2008} to enable the migration of codes as hardware evolves.
+standard\cite{Reid:2008:NFF:1408643.1408645} to enable the migration of codes as hardware evolves.
 Significant language features were left out of this early version of coarrays as

### 2016-01-18 papers/CAFe/HIPS-paper/coarray-extensions.tex
Added comment at end that perhaps could be used in final version regarding domain specific languages (for D-TEC purposes).

@@
 excluded from CAFe.
+
+\begin{comment}
+It is noted that these extensions to Fortran have been described in terms
+explicit language syntax modifying the base language.  CAFe could equivalently have
+been described in terms of compiler directives.  For example, subimage task execution could be
+written as
+\begin{verbatim}
+ call relax(U[device]) !$OFP [[device]]
+\end{verbatim}
+which would allow a standard CAF compiler to compile CAFe programs.  This gives
+the flavor of a domain specific language.
+\end{comment}

### 2016-01-18 papers/CAFe/HIPS-paper/coarray-extensions.tex
Replaced begin block with block to make it real Fortran.
Modified as Dan suggested: variables can be coarrays, not just arrays.

@@
 
-In Fortran, arrays that are visible to other program images must be declared with the
+In Fortran, variables that are visible to other program images must be declared with the
 codimension attribute, for example,
@@
 do concurrent(i=1:N-1)  [[device]]
- begin block
+ block
    real :: w = 0.67

### 2016-01-19 papers/CAFe/HIPS-paper/performance.tex
Added performance table and text.

@@
 --- executing a \emph{single} program --- with concurrent execution of separate tasks executing on
-potentially heterogeneous hardware.  However, the relative performance of computation on
-a cluster of GPUs compared with the necessary communication of halo information is
-of interest, especially considering that a complete exchance of halo data involves communication
-between multiple
-coarray images \emph{and} between each individual host image and its subimage (an attached
-OpenCL device).
+potentially heterogeneous hardware.
 
-%%However, preliminary work using the Stratego/XT rewrite system as described indicates that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).
+\begin{comment}
+However, the relative performance of computation on a cluster of GPUs compared with the
+necessary communication of halo information is of interest, especially considering that a
+complete exchance of halo data involves communication between multiple coarray images
+\emph{and} between each individual host image and its subimage (an attached OpenCL
+device).
+\end{comment}
 
-This section shows and compares performance measurements for: 1. Computational time on the GPU
-of the interior domain;
-2. Computational time on the CPU of the shared boundary regions; and 3. Halo communication time.
+However the relative performance of computation on the interior of a three-dimensional
+grid, performed by the GPU, compared with the time required to compute on boundary planes by the
+CPU is of interest.  Table 1 shows average execution time for a relaxation step on the GPU
+(column 2) and on the CPU (column 3) in milliseconds, where $N$ (column 1) is the number of
+cells in one dimension of a 3D cube; time for the exchange of halo information
+between the GPU and the CPU for each iteration is also shown in column 4.
 
-%%\subsection{Halo Transfer Performance}
+\begin{table}[]
+\centering
+\caption{Size of cube and time in milliseconds}
+\label{table1}
+\begin{tabular}{rrrrr}
+$N$   &   GPU	 &  CPU         &   Comm    \\
+      &   	 &              &           \\
+16    &  0.02    &   0.01	&   0.07    \\
+32    &  0.02    &   0.03       &   0.10    \\
+64    &  0.04    &   0.12       &   0.27    \\
+128   &  0.42    &   0.47	&   0.81    \\
+256   &  4.59    &   1.97	&   2.99    \\
+512   & 43.08    &   12.51	&  16.49    \\
+\end{tabular}
+\end{table}
 
+Note that all three times in Table 1 are roughly equivalent for $N=128$.  Also note that
+the communication and the computational times on the CPU are roughly equivalent and scale
+the same with increasing $N$.  This is not surprising as both involve only the surface
+elements of the cube.  However, computational time on the GPU grows much faster, O($N^3$),
+as it is computing on the interior of the cube.
+
+\begin{comment}
 The current implementation does \emph{not} take advantage of optimization strategies such
-as prefetching of array tiles (including halos) into OpenCL local memory.
-
-%%either does it take advantage of the potential of CAFe to overlap communication with computation (for example, computing on
+as prefetching of array tiles (including halos) into OpenCL local memory.  Neither does it
+take advantage of the potential of CAFe to overlap communication with computation (for
+example, computing on ...
+\end{comment}
 

### 2016-01-20 papers/CAFe/HIPS-paper/conclusions.tex
Fixed spelling error.

@@
 Significant language features were left out of this early version of coarrays as
-identified by \emph{Mellor-Crummy et al}\cite{mellor-crummey:2009:caf2}.  Many of these features
+identified by \emph{Mellor-Crummy et al.}\cite{mellor-crummey:2009:caf2}.  Many of these features
 have been adopted in what is likely to be called the 2015 standard of the language\cite{fortran:2015},

### 2016-02-01 papers/CAFe/HIPS-paper/intro.tex
tweaks

@@
 
-Another potential option is to abandon the stricture of message passing within a procedural
+Another potential option is to abandon the structure of message passing within a procedural
 programming language altogether.  For example, the Open Community Runtime (OCR)
@@
 environments.  The hope is that high-level syntax can hide the complexities of the
-underlying hardware \emph{and software.}
+underlying hardware \emph{and software} by delegating some optimization tasks to the
+compiler instead of the programmer.
+
+%% MJS: this seems like a natural place for a subsection break, but it looks ugly
+%%      in the IEEE two column style
 
@@
 Since each CAF process executes the same program, there are no good avenues for exploiting
-task-based parallelism.  Neither is there a adequate mechanism for dynamic coarray memory
+task-based parallelism.  Neither is there an adequate mechanism for dynamic coarray memory
 allocation, nor a mechanism for the remote execution of functions.  In CAF, like MPI, the

### 2016-02-01 papers/CAFe/HIPS-paper/performance.tex

@@
 necessary communication of halo information is of interest, especially considering that a
-complete exchance of halo data involves communication between multiple coarray images
+complete exchange of halo data involves communication between multiple coarray images
 \emph{and} between each individual host image and its subimage (an attached OpenCL

### 2016-02-01 papers/CAFe/HIPS-paper/transformations.tex

@@
 strategies:
-(1) a symbol table is produced in the first pass to store information related to arrays
-including, array shape and allocation status;
+(1) a symbol table is produced in the first pass to store information related to arrays,
+including array shape and allocation status;
 (2) additional variables are declared that are used to maintain the OpenCL runtime state,

### 2016-02-01 papers/Dijkstra/OpenCL/dijkstra_cl.F90
Add '' to blank print statements so the file would compile

@@ -99,7 +99,7 @@ DO WHILE(.NOT. done)
 END DO 
 
 IF (debug) THEN
-  PRINT *, 
+  PRINT *, ""
   DO i = 1, NX
      DO j = 1, NY
         DO k = 1, NZ
@@ -109,7 +109,7 @@ IF (debug) THEN
   END DO 
 END IF
 
-PRINT *, 
+PRINT *, ""
 PRINT *, "Sweep/reduce time for N=", NX*NY*NZ, real(time_sweep), real(time_reduce)
 
 DEALLOCATE(U,TT,Changed,Offset)

### 2016-03-07 papers/CAFe/HIPS-paper/coarray-extensions.tex
Added subsection on alternative syntax.

@@
 
-\begin{comment}
-It is noted that these extensions to Fortran have been described in terms
+\subsubsection{Alternative Syntax}
+
+It is noted that these extensions to Fortran have been described in terms of
 explicit language syntax modifying the base language.  CAFe could equivalently have
@@
 \begin{verbatim}
- call relax(U[device]) !$OFP [[device]]
+   call relax(U[dev1]) !$OFP [[dev1]]
 \end{verbatim}
-which would allow a standard CAF compiler to compile CAFe programs.  This gives
-the flavor of a domain specific language.
-\end{comment}
+which would allow a standard CAF compiler to compile CAFe programs.  This syntax usage
+would make the CAFe constructs described here an embedded Domain Specific Language.  In a
+future version of the implementation of CAFe, we intend to also support compiler directives
+in this form.
+
+Alternative syntax for remote execution may be preferable to the double square bracket notation.
+While the syntax \texttt{[[ ]]} makes it clear that the programmer intends \emph{remote} execution
+to take place, perhaps syntax closer to standard Fortran is better, especially if
+standardization of CAFe were to occur.  For example, memory placement could be requested by a
+slight modification of the allocation statement
+\begin{verbatim}
+   allocate(U(M,N)[*], DEVICE=dev1)
+\end{verbatim}
+and similarly remote execution via a subroutine call could be replaced by
+\begin{verbatim}
+   call relax(U[dev1]), DEVICE=dev1
+\end{verbatim}
+However, while the alternative syntax for memory allocation appears very natural, using 
+the syntax \texttt{,DEVICE=dev1} to denote remote execution is problematic when applied to
+Fortran function calls.  For example, the double square bracket notation in
+\begin{verbatim}
+   y = sqrt( sin(x)[[dev1]] * 2 )
+\end{verbatim}
+cannot be replaced with \texttt{,DEVICE=dev1}.
